\chapter{All 206 examples of 1pl in these texts are found in the following sentences:}

\ea \label{ZB.7}
\textit{ginmē dewrū hāminī wehārī temāmīyo} \\ 
\gll ∅-gin-mē dewr=ū hāmin-ī wehār-ī temāmīye-o \\ 
 \textsc{ind-}fall\textsc{.prs}\textsc{-1pl} around\textsc{\textsc{=ez.gen}} summer\textsc{.m}\textsc{-obl}\textsc{.m} spring\textsc{.m}\textsc{-obl}\textsc{.m} finish\textsc{.prs}\textsc{-3sg} \\ 
\glt `It was\footnote{Lit. We fell} near the end of summer, [or] spring.'
\z 
 
\ea \label{ZB.15}
\textit{māčo, ēšew wāro tekbīrmā mekirīyo} \\ 
\gll m-āč-o ēšew ∅-wār-o tekbīr=mā me-kir-īye-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} tonight \textsc{ind-}rain\textsc{-3sg} solution\textsc{=1pl} \textsc{neg-}do.pass-pass\textsc{-3sg} \\ 
\glt `He said, ‘It’s going to rain tonight; We can’t handle (it)\footnote{Lit. Our solution won’t be done.} '
\z 
 
\ea \label{ZB.33}
\textit{yeženekē biřē pīyāke māčo, melāmā pey’} \\ 
\gll yeženī-ekē biřē pīyā-eke m-āč-o me-l-ā=mā pey \\ 
 woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} a\_while man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}go\textsc{.prs}\textsc{-3pl}\textsc{=1pl} for \\ 
\glt `The woman (or) the man said several times, ‘It's not possible for us [to take the babies]. \footnote{they don't go for us}’'
\z 
 
\ea \label{ZB.46}
\textit{sāɫē ter mēnēwe, māčo, bilmē bizānmē tekbīrī ētir řāwēžmā čēš bo’ āđē, wiškesāɫīn?} \\ 
\gll sāɫe-ē ter m-ē-nē=we m-āč-o bi-l-mē bi-zān-mē tekbīr-ī ētir řāwēž=mā čēš ∅-b-o āđē wiškesāɫī=n \\ 
 year\textsc{.f}\textsc{-dir}\textsc{.pl} other \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} advice\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} conjecture\textsc{=1pl} what \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} 3pl\textsc{.dir} drought\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The following year, they went back. He (the father) said, ‘We shall go [and] see what is ahead of us with this period of drought.’'
\z 
 
\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.9}
\textit{vātiš, qurbān luāymē pār yā činnē sāɫē čēweɫter luāymē pey sārāɫū ērānī, koy sān} \\ 
\gll vāt=iš qurbān luā-ymē pār yā činnē sāɫe-ē čēweɫter luā-ymē pey sārāɫ=ū ērān-ī ko=y sān \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} go\textsc{.pst} \textsc{1pl} last\_year either some\textsc{.pl} year.f\textsc{-indf} ago go\textsc{.pst} \textsc{1pl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} where=ez.\textsc{attr} \textsc{pn} \\ 
\glt `He said, ‘Sir, we went, say, last year or a few years ago, we went to the Saral region in Iran, to the San mountain.'
\z 
 
\ea \label{ZQ.10}
\textit{čowe sārāɫū ērānīwe gēɫāymēwe} \\ 
\gll č=o=we sārāɫ=ū ērān-ī=we gēɫā-ymē=we \\ 
 in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=compl} turn\textsc{.pst} \textsc{1pl}\textsc{=compl} \\ 
\glt `On the way back, we returned from the Saral region of Iran.'
\z 
 
\ea \label{ZQ.11}
\textit{wextū ī jimāymārā āmāymēwe āmāymē kelēwe} \\ 
\gll wext=ū ī jimāy=mā=rā āmā-ymē=we āmā-īmē kel-ēwe \\ 
 time\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} move\textsc{.inf}\textsc{=\textsc{1pl}}\textsc{=\textsc{povb}} come\textsc{.pst} \textsc{1pl}\textsc{=compl} come\textsc{.pst} \textsc{1pl} mountain\textsc{-indf} \\ 
\glt `On our return (journey), we went\footnote{Lit. came.}  to a mountain. '
\z 
 
\ea \label{ZQ.16}
\textit{duē, be gīrmā wārd čenū ženīū čenū heywānīū dewɫetīū bē āwīū nāřihet e pāwe} \\ 
\gll duē be gīr=mā wārd čenū ženī=ū čenū heywān-ī=ū dewɫet-ī=ū bē āwī=ū nāřihet e=pāwe \\ 
 two by hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} woman.f=and with\textsc{.ez.gen} animal\textsc{.m}\textsc{-obl}\textsc{.m}=and animal\textsc{-obl}\textsc{.m}=and come\textsc{.imp}\textsc{.\textsc{2sg}} water.f=and sad in=there \\ 
\glt `I was stuck there with my wife, livestock, a lack of water, and a lack of comfort. '
\z 
 
\ea \label{ZQ.18}
\textit{āđmā tehwēɫ gērtū} \\ 
\gll āđ=mā tehwēɫ gērt=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=\textsc{1pl}} delivery\textsc{.m} grab\textsc{.pst}\textsc{=ez.gen} \\ 
\glt `We got hold of him.'
\z 
 
\ea \label{ZQ.21}
\textit{ēmeyč gīrmā wārd čenū her wī} \\ 
\gll ēme=yč gīr=mā wārd čenū her wē-ī \\ 
 \textsc{1pl}\textsc{=add} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} just two\textsc{-obl}\textsc{.m} \\ 
\glt `We were stuck with both babies.'
\z 
 
\ea \label{ZQ.22}
\textit{yošā kerdmā bāboɫē ārdmā} \\ 
\gll yo=šā kerd-∅=mā bāboɫē ārd-∅=mā \\ 
 one\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1pl}} swaddling bring\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `We rolled one in clothing [and] took him.'
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{ZQ.27}
\textit{luāymēwe pey germīyānū ʕērāqī} \\ 
\gll luā-ymē=we pey germīyān=ū ʕērāq-ī \\ 
 go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We went back to the Garmiyan region of Iraq.'
\z 
 
\ea \label{ZQ.28}
\textit{sāɫē.. luāymērā oge,} \\ 
\gll sāɫe-ē luā-ymē=rā oge \\ 
 year.f\textsc{-indf} go\textsc{.pst} \textsc{1pl}\textsc{=\textsc{povb}} there \\ 
\glt `A year [passed since] we went back there.'
\z 
 
\ea \label{ZQ.29}
\textit{āne ke berdmā šiš māngē menn} \\ 
\gll āne ke berd-∅=mā šiš mānge-ē menn \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{rel} take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{1pl} six month.f\textsc{-dir}\textsc{.pl} remain\textsc{.pst} \\ 
\glt `The one whom we took [with us] lived for six months.'
\z 
 
\ea \label{ZQ.33}
\textit{vātiš, qurbān wehārē wāwey luāymēwe pey sārāɫū ērānī} \\ 
\gll vāt=iš qurbān wehār-ē wāwey luā-ymē=we pey sārāɫ=ū ērān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} spring\textsc{.m}\textsc{-indf} again go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘Sir, [The next] spring, again, we went back to the Saral region of Iran.'
\z 
 
\ea \label{ZQ.34}
\textit{her pā henāhur luāymēwe} \\ 
\gll her p=ā hen-ā=hur luā-ymē=we \\ 
 just at=\textsc{dem.dist} side\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} go\textsc{.pst} \textsc{1pl}\textsc{=compl} \\ 
\glt `We went back by the same route [that we had taken a year before]\footnote{Lit. came back} by the same route [that we had taken a year before]'
\z 
 
\ea \label{ZQ.53}
\textit{kuřmā berārdū berdmā čenū wēmāw wešīyē ʕejāyibū diɫwešīyē ʕejāyib} \\ 
\gll kuř=mā ber-ārd-∅=ū berd-∅=mā čenū wē=mā=ū wešī-ē ʕejāyib=ū diɫwešī-ē ʕejāyib \\ 
 son/boy\textsc{.m}\textsc{=\textsc{1pl}} out- bring\textsc{.pst}\textsc{-3sg}\textsc{.m}=and take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1pl}} with\textsc{.ez.gen}=and happiness\textsc{.m}\textsc{-indf} unbelievable=and joy\textsc{.m}\textsc{-indf} unbelievable \\ 
\glt `We took out the son, and took him with us. [We were full of] unbelievable happiness; [full of] an unbelievable sense of joy. '
\z 
 
\ea \label{HB.32}
\textit{ēme milmē Šešìk} \\ 
\gll ēme mi-l-mē Šešìk \\ 
 \textsc{1pl} \textsc{ind-}go\textsc{.prs}\textsc{-1pl} \textsc{pn} \\ 
\glt `We will go to Shashk.'
\z 
 
\ea \label{HB.33}
\textit{milmē gošew šārī, āwāyekey} \\ 
\gll mi-l-mē goše=w šār-ī āwāī-ekey \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-1pl} corner\textsc{=ez}\textsc{.gen} city\textsc{.m}\textsc{-obl}\textsc{.m} village\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `We will go to the city suburb, to the village. '
\z 
 
\ea \label{HB.34}
\textit{čowe neferēwe mē, pēwāymāwe} \\ 
\gll č=o=we nefer-ēwe- ē pēwāy=mā=we \\ 
 in\textsc{=dem}.dist\textsc{=\textsc{post}} person\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} welcoming\textsc{.m}\textsc{=1pl}\textsc{=\textsc{post}} \\ 
\glt `There, a person will welcome us.'
\z 
 
\ea \label{HB.35}
\textit{xūɫkemā kero, beromā yānew wēšā} \\ 
\gll xūɫke=mā ∅-ker-o ∅-ber-o=mā yāne=ū wē=šā \\ 
 invitation\textsc{=1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=1pl} house=and \textsc{reflx}\textsc{=3pl} \\ 
\glt `He will invite us; he will take us to his\footnote{Lit. their.} house.'
\z 
 
\ea \label{HB.42}
\textit{pekemā megino ētir} \\ 
\gll peke=mā me-gin-o ētir \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} no\_longer \\ 
\glt `we should not be worried anymore'
\z 
 
\ea \label{HB.43}
\textit{pekemā megino, to goš min dāre!’} \\ 
\gll peke=mā me-gin-o to goš min ∅-dār-e \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \textsc{2sg} ear\textsc{.m} 1sg \textsc{imp-}hold\textsc{.prs}-\textsc{2sg} \\ 
\glt `We should not be worried; listen to me!’'
\z 
 
\ea \label{PM.5}
\textit{fermāndew merekuřekey, reʕīsekešā māčo, ēme bē ā šexsī metāwmē, e čē menmēwe zerer deymē čī mentēqey’} \\ 
\gll fermānde=ū merekuř-ekey reʕīs-eke=šā m-āč-o ēme bē ā šexs-ī me-tāw-mē e=čē ∅-men-mē=we zerer ∅-de-ymē č=ī mentēqe-ī \\ 
 leader\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} leader\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} in=here \textsc{sbjv-}stay\textsc{.prs}\textsc{-1pl}\textsc{=compl} harm\textsc{.m} \textsc{sbjv-}give\textsc{.prs}-\textsc{1pl} in=\textsc{dem.prox} region\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The head of the grasshoppers said, ‘Without [permission from] this person we cannot stay here and do harm to this region.'
\z 
 
\ea \label{PM.9}
\textit{ījāze bizāne āđ ījāze miđo ēme e čē nehār kermē yām ne} \\ 
\gll ījāze bi-zān-e āđ ījāze mi-đe-o ēme e=čē nehār ∅-ker-mē yām ne \\ 
 permission \textsc{imp-}know\textsc{.prs}-\textsc{2sg} 3sg\textsc{.dir}\textsc{.m} permission \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \textsc{1pl} in=here lunch\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} or no \\ 
\glt `See if he lets us stay here for lunch or not'
\z 
 
\ea \label{PM.11}
\textit{māčo, qurbān ēme tāyfēnmē āmēnmēre,} \\ 
\gll m-āč-o qurbān ēme tāyfe-ē=nmē āmē=nmē=re \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{1pl} tirbe\textsc{.m}\textsc{-indf}\textsc{=cop}\textsc{.1pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{=\textsc{povb}} \\ 
\glt `[and] said, ‘Sir, we are a family [who] have come here.'
\z 
 
\ea \label{PM.12}
\textit{īnāymē fiɫāne yāgē} \\ 
\gll īnā-ymē fiɫān=e yāgē \\ 
 \textsc{deic-1pl} such\_and\_such\textsc{=ez}\textsc{.cmpd} place.f \\ 
\glt `We are at such-and-such a place.'
\z 
 
\ea \label{PM.14}
\textit{eger to ījāze miđey ēme e čē nehārē kermē ījāze be desū ton} \\ 
\gll eger to ījāze mi-đe-ī ēme e=čē nehār-ē ∅-ker-mē ījāze be des=ū to=n \\ 
 if \textsc{2sg} permission \textsc{ind-}give\textsc{.prs}-\textsc{2sg} \textsc{1pl} in=here lunch\textsc{.m}\textsc{-indf} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} permission to hand\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `It is in your hands whether to let us stay here for lunch.'
\z 
 
\ea \label{PM.15}
\textit{āmēnmē xizmetū to ījāzemā biđey} \\ 
\gll āmē=nmē xizmet=ū to ījāze=mā bi-đe-ī \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} permission\textsc{=1pl} \textsc{sbjv-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `I\footnote{Lit. We.} have come to your service so that you might permit us.'
\z 
 
\ea \label{PM.17}
\textit{ānā ēme wēmā mehelēmā hen tūl} \\ 
\gll āne=ā ēme wē=mā mehele-ē=mā hen-∅ tūl \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{1pl} \textsc{reflx}\textsc{=1pl} neighbourhood\textsc{.m}\textsc{-indf}\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} long \\ 
\glt `There, I, myself, own a big neighbourhood.’\footnote{Lit. We, ourselves, have a big neighbourhood.}'
\z 
 
\ea \label{PM.20}
\textit{fermāwebē, fermāwo lodē e čā muɫkū ēme wēmā, herčēmā hen čāne ā firmānde wēš zāno heqū ēme kāmā} \\ 
\gll fermāwe b-ē ∅-fermāw-o ∅-lo-dē e=čā muɫk=ū ēme wē=mā herčē=mā hen-∅ č=ā=ne ā firmānde wē=š ∅-zān-o heq=ū ēme kām=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} in=there property\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{reflx}\textsc{=1pl} whatever\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} leader\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} right\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} which\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go! In that property of mine, whatever we possess there, your boss knows which portion [of land] is mine.'
\z 
 
\ea \label{PM.36}
\textit{min pewkī ījāzem dān hinekew ēme wero, bilo zerer je kesī negino} \\ 
\gll min pewkī ījāze=m dā=n hin-eke=ū ēme ∅-wer-o bi-l-o zerer je kes-ī ne-gin-o \\ 
 \textsc{1sg} that\_is\_why permission\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} ez.pron\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{sbjv-}eat\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} harm\textsc{.m} from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `I let them [eat our crop] so that they would eat our crop and go away, [but] not do harm to others.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.38}
\textit{girđū šārezūrī ēsāɫ merāʕetū ēme kero je pārī zīyāter pey toyčū pey ēmeyč,} \\ 
\gll girđ=ū šārezūr-ī ēsāɫ merāʕet=ū ēme ∅-ker-o je pār-ī zīyā-ter pey to=yč=ū pey ēme=yč \\ 
 all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} this\_year care\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} from last\_year\textsc{-obl}\textsc{.m} increased\textsc{-cmpr} for \textsc{2sg}\textsc{=add}=and for \textsc{1pl}\textsc{=add} \\ 
\glt `This year, all the people in Sharazour care for us— both you and I—, more than the previous year.'
\z 
 
\ea \label{PM.39}
\textit{hīčkāmmā mihtājē mebīmē pīney} \\ 
\gll hīčkām=mā mihtāj-ē me-b-īmē p=īney \\ 
 no\_one\textsc{=1pl} needy\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs} \textsc{1pl} by=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \\ 
\glt `None of us would be in need of these [products] '
\z 
 
\ea \label{PM.43}
\textit{girđčē benjānī bo xīyārī bo bermē bā pey verzīyerīū pey īđīū pey ewīū pey ewīū pey ewī} \\ 
\gll girđčē benjānī ∅-b-o xīyārī ∅-b-o ∅-ber-mē bā pey verzīyer-ī=ū pey īđī=ū pey ewī=ū pey ewī=ū pey ewī \\ 
 everything tomato.f \textsc{ind-}be\textsc{.prs}\textsc{-3sg} cucumber \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}take\textsc{.prs}\textsc{-1pl} \textsc{hort} for farm\_labourer\textsc{.m}\textsc{-obl}\textsc{.m}=and for 3sg\textsc{.obl}\textsc{.m}=and for 3sg\textsc{.obl}\textsc{.m}=and for 3sg\textsc{.obl}\textsc{.m}=and for 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `Everything [was there]; There were cucumbers; there were tomatoes. [People said], ‘Let’s take [these products] to the farm-labourer; to this [person]; to that [person]; to that [person] and so on.'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.5}
\textit{eger newešīēmā bīyebo duktur nebīyenū} \\ 
\gll eger newešī-ē=mā bīye b-o duktur ne-bīye=n=ū \\ 
 if illness\textsc{.m}\textsc{-indf}\textsc{=1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} physician\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \\ 
\glt `If there might have been an illness, there was no doctor. '
\z 
 
\ea \label{DG.6}
\textit{ā wextīekey lūēnmē e wāge duʕāw xeyrīš kerdēnew} \\ 
\gll ā wext-ī-ekey lūē=nmē e=wāge duʕā=ū xeyr-ī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} in=there prayer\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} goodness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `We would go there [i.e. to them and] they would bless us'
\z 
 
\ea \label{DG.7}
\textit{xwāy derdemā deʕfeš kerdēne} \\ 
\gll xwā-ī derde=mā deʕfe=š kerdē=ne \\ 
 God\textsc{.m}\textsc{-obl}\textsc{.m} illness\textsc{\textsc{.f}}\textsc{=1pl} exclusion\textsc{\textsc{.f}}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `[and] God would protect us from illness.'
\z 
 
\ea \label{DG.35}
\textit{nimāzā bilmē āweyānī} \\ 
\gll ni-m-āz-ā bi-l-mē āweyānī \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} habitat\textsc{\textsc{.f}} \\ 
\glt `we are not allowed to go to the village.'
\z 
 
\ea \label{DG.36}
\textit{be čāwel ħeywāndārē bīyēnmēū} \\ 
\gll be čāwel ħeywāndār-ē bīyē=nmē=ū \\ 
 by in\_the\_past animal\_husbandry\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}=and \\ 
\glt `In the past, our job was animal husbandry;'
\z 
 
\ea \label{DG.37}
\textit{bāxdārē bīyēnmēū} \\ 
\gll bāxdār-ē bīyē=nmē=ū \\ 
 gardener\textsc{.m}\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}=and \\ 
\glt `we were gardeners.'
\z 
 
\ea \label{DG.42}
\textit{nizīkū āweyānīyē nimāzā ginmē} \\ 
\gll nizīk=ū āweyānī-ē ni-m-āz-ā ∅-gin-mē \\ 
 close\textsc{=ez}\textsc{.gen} habitat\textsc{\textsc{.m}}\textsc{-indf}\textsc{\textsc{.f}} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} sbjv\textsc{-f}all\textsc{.prs}\textsc{-1pl} \\ 
\glt `They don’t let us anywhere near a village.'
\z 
 
\ea \label{DG.64}
\textit{ā wextīekey beɫē, min bilo ā memleketū wēšā, kiseɫīš čenē, ēme zāmdārē nekero ēme kečē nekero} \\ 
\gll ā wext-ī-ekey beɫē min bi-l-o ā memleket=ū wē=šā kiseɫī=š čenē ēme zāmdār-ē ne-ker-o ēme keč-ē ne-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yes \textsc{1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3pl} tortoise\textsc{\textsc{.f}}\textsc{=3sg} in \textsc{1pl} wounded\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `[The Sheikh said], ‘Yes!' '[The tortoise said] I [say] he [should] go to his region; there are tortoises there. He shall not injure us; he shall not make us crooked. '
\z 
 
\ea \label{DG.66}
\textit{ē.. dewāyekeš sāwe pore dā ēme kečē nebīmē} \\ 
\gll ē.. dewāī-eke=š ∅-sāw-e pore dā ēme keč-ē ne-b-īmē \\ 
 eh medication\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} at \textsc{hort} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs}-\textsc{1pl} \\ 
\glt `He shall rub the medicine on him; he shall not make us crooked.'
\z 
 
\ea \label{DG.68}
\textit{berešo memliketū wētā welē sermā deydē} \\ 
\gll ∅-ber-e=š=o memliket=ū wē=tā welē ser=mā ∅-de-ydē \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=\textsc{compl}} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=2pl} but head\textsc{.m}\textsc{=1pl} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} \\ 
\glt `Take him to your region, but pay us a visit.'
\z 
 
\ea \label{DG.73}
\textit{īnē dāstānū ī šēxāū ēme bīyen gewregerīšā} \\ 
\gll īnē dāstāne=ū ī šēx-ā=ū ēme bīye=n gewregerī=šā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{\textsc{.f}}\textsc{.3sg} story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} sheikh\textsc{.m}\textsc{-obl}\textsc{.pl}=\textsc{ez.gen} \textsc{1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} greatness\textsc{=3pl} \\ 
\glt `This is the story of our Sheikhs and their greatness.'
\z 
 

\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.21}
\textit{ī bāxū bāxātū hewrāmānī, memliketū hewrāmānī, xeɫkū hewrāmānī ā wextīyekey ēme hukimřewiš bīyēnmē} \\ 
\gll ī bāx=ū bāx-āt=ū hewrāmān-ī memliket=ū hewrāmān-ī xeɫk=ū hewrāmān ī ā wext-ī-ekey ēme hukimřew=iš bīyē=nmē \\ 
 \textsc{dem.prox} garden\textsc{.m}\textsc{=ez}\textsc{.gen} garden\textsc{.m}\textsc{-pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} people\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \textsc{dem.prox} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} governor\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `We have ruled over\footnote{Lit. We have been rulers.} these gardens of Hawraman, this region of Hawraman, [and] this people of Hawraman.'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.39}
\textit{bē bā nīšmēre seḷāhū wēmā kermē} \\ 
\gll bē bā ∅-nīš-mē=re seḷāh=ū wē=mā ∅-ker-mē \\ 
 come\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{hort} \textsc{sbjv-}sit\textsc{.prs}\textsc{-1pl}\textsc{=\textsc{povb}} solution\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `[He said to Sibhan Agha], ‘Let us sit together and come to a solution.’'
\z 
 
\ea \label{DP.44}
\textit{wāt, bē hem mēmān bo nīmeřo hem bā qisē kermē} \\ 
\gll wāt b-ē hem mēmān ∅-b-o nīmeřo hem b-ā qisē-ē ∅-ker-mē \\ 
 say\textsc{.pst} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} both guest\textsc{.m} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} noon both be\textsc{.prs}\textsc{-3pl} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `He said, ‘Come and be our guest for lunch, and we will discuss the matter.’'
\z 
 
\ea \label{ZP.36}
\textit{īne māčmēš pene šālīyāre sīyāw lālow kināčekēn} \\ 
\gll īne m-āč-mē=š pene šālīyār=e sīyāw lālo=ū kināčē-ekē=n \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn}\textsc{=dem} black maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This [person], whom we call Shaliyar Siya, is the maternal uncle of the [king’s] daughter.'
\z 
 
\ea \label{ZP.51}
\textit{ētir kināče keřū lāɫe bo māčmēš pene tewenū sirewē} \\ 
\gll ētir kināčē=e keře=ū lāɫ-e ∅-b-o m-āč-mē=š pene tewenī=ū sirewē \\ 
 \textsc{disc.ptcl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} deaf\textsc{.f}=and mute\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} \\ 
\glt `Since the girl was deaf-mute, we call it the healing stone.'
\z 
 
\ea \label{ZP.63}
\textit{mizyo, mizyā werāwerū māčmēš pene dūtifɫe} \\ 
\gll mi-zye-o mi-zye-ā werāwer=ū m-āč-mē=š pene dūtifɫe \\ 
 \textsc{ind-}go\_out\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn} \\ 
\glt `[Then] they left [Hawraman village and went to a place which] we call Dutifla'
\z 
 
\ea \label{ZP.73}
\textit{īse māčmēš pene tenūrū dēwā} \\ 
\gll īse m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 now \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Now we call that place Tanūrū dēwā ‘the demons’ oven’'
\z 
 
\ea \label{ZP.74}
\textit{ā degākē ehā čew deso bīyēne duē kutē māčmēš pene tenūrū dēwā} \\ 
\gll ā degā-ekē ehā č=ew des=o bīyē=ne duē kut-ē m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 \textsc{dem.dist} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{intj} in=\textsc{dem.dist} hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} two half\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Do you see that the village is divided in half on the other side? We call that place Tanūrū dēwā.'
\z 
 
\ea \label{ZP.84}
\textit{persā, ēme yānew fiɫāne kesīmā gerekā} \\ 
\gll ∅-pers-ā ēme yāne=ū fiɫān=e kes-ī=mā gerek=ā \\ 
 \textsc{ind-}ask\textsc{.prs}\textsc{-3pl} \textsc{1pl} house\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} necessary\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[and] asked [people], ‘We are looking for\footnote{Lit. we want.} the house of such-and-such a person.'
\z 
 
\ea \label{ZP.88}
\textit{ā wextīekey ʕerz kerā, dey ēme ī kināčē jeryāniš e čīnew} \\ 
\gll ā wext-ī-ekey ʕerz ∅-ker-ā dey ēme ī kināčē jeryān=iš e=čīne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} petition\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{1pl} \textsc{dem.prox} girl\textsc{.f} story\textsc{.m}\textsc{=3sg} in=like=and \\ 
\glt `Then they (her uncle) said, ‘This girl’s story is like this.'
\z 
 
\ea \label{ZP.94}
\textit{ā wextī ēme xeplezeřetīnē bēnē tā pāyīz} \\ 
\gll ā wext-ī ēme xeplezeřetīnē b-ēn-ē tā pāyīz \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{1pl} \textsc{pn}\textsc{.dir}\textsc{.pl} be\textsc{.prs}\textsc{-pstc}\textsc{-3pl} until autumn \\ 
\glt `Back then, there were farms of black corn\footnote{A type of corn. Also a type of bread made from corn flour.} until autumn.'
\z 
 
\ea \label{ZP.122}
\textit{sereš pey sereš biřmēw čāštē kermēw} \\ 
\gll sere=š pey sere=š ∅-biř-mē=ū čāštī-ē ∅-ker-mē=ū \\ 
 head\textsc{.m}\textsc{=3sg} for head\textsc{.m}\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl}=and meal\textsc{.f}\textsc{-indf} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl}=and \\ 
\glt `We shall kill it and cook food.'
\z 
 
\ea \label{ZP.123}
\textit{merāsēmekey řās kermēwew} \\ 
\gll merāsēm-ekey řās ∅-ker-mē=we=ū \\ 
 ceremony\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} true \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl}\textsc{=compl}\textsc{\textsc{=ez.gen}} \\ 
\glt `And carry out the ceremony.'
\z 
 
\ea \label{ZP.129}
\textit{šūnū ēmeyčre her meno} \\ 
\gll šūn=ū ēme=yč=re her ∅-men-o \\ 
 after\textsc{\textsc{=ez.gen}} \textsc{1pl}\textsc{=add}\textsc{=\textsc{post}} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `It will continue to exist after us as well.'
\z 
 
\ea \label{ŽP.4}
\textit{pewkīmā leqemšā miđā māčāš pīr šelīyār} \\ 
\gll pewkī=mā leqem=šā mi-đe-ā m-āč-ā=š pīr šelīyār \\ 
 that\_is\_why\textsc{=1pl} title\textsc{.m}\textsc{=3pl} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{pn} \textsc{pn} \\ 
\glt `That is why he’s been given this title [and] is called Pir Shaliyar.'
\z 
 
\ea \label{ŽP.62}
\textit{wēregā pīr šelīyār mēwe sey misefā wextē mēwe, ʕerzū ženekēš kero māčo, bā ēšew īmti̩hānēš kermē lāw wēmāwe newsinmēš} \\ 
\gll wēregā pīr šelīyār m-ē=we sey misefā wext-ē m-ē=we ʕerz=ū ženī-ekē=š ∅-ker-o m-āč-o bā ēšew īmti̩hān-ē=š ∅-ker-mē lā=ū wē=mā=we ne-ūsin-mē=š \\ 
 evening \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{pn} \textsc{pn} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} tonight test\textsc{-indf}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=1pl}\textsc{=\textsc{post}} \textsc{neg-}put\_to\_sleep\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \\ 
\glt `Pir Shaliyar returned in the evening; when Say Mustafa returned, he (Hama Yoso) said to his wife, ‘Let’s test him tonight; let’s not put him to sleep next to us.’'
\z 
 
\ea \label{ŽP.63}
\textit{yāgē jīyekāre būsinmēš} \\ 
\gll yāgē-ē jīyekār-e b-ūsin-mē=š \\ 
 place\textsc{.f}\textsc{-indf} separate\textsc{-f} \textsc{sbjv-}put\_to\_sleep\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \\ 
\glt `We will put him to sleep in a separate place.'
\z 
 
\ea \label{ŽP.64}
\textit{īmtīhānē teriš kermē bizānmē řāsā diron} \\ 
\gll īmtīhān-ē ter=iš ∅-ker-mē bi-zān-mē řās=ā diro=n \\ 
 test\textsc{-indf} another\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} true\textsc{=cop}\textsc{.3sg}\textsc{.m} lie\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We will test him again, see if it [his supernatural power] is right or it is wrong.'
\z 
 
\ea \label{ŽP.72}
\textit{ʕerzū heremēš kero, fermāwo be heremēš weɫāhī ētir ēme metāwmē, sey misefā metāwmē hīč kārēš pene kermē īlā kīyānmēš bilo weno, bilo kīyānmēš weno} \\ 
\gll ʕerz=ū hereme-ē=š ∅-ker-o ∅-fermāw-o be hereme-ē=š weɫāhī ētir ēme me-tāw-mē sey misefā me-tāw-mē hīč kār-ē=š pene ∅-ker-mē īlā ∅-kīyān-mē=š bi-l-o ∅-wen-o bi-l-o ∅-kīyān-mē=š ∅-wen-o \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} indeed \textsc{disc.ptcl} \textsc{1pl} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} \textsc{pn} \textsc{pn} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} except \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}read\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}read\textsc{-3sg} \\ 
\glt `He (Pir Hama Yoso) said to his wife, ‘Indeed, we can no longer be helpful to Say Mustafa,\footnote{Lit. we can do nothing to him.} except that we send him to study; that we send him, to go [and] study.'
\z 
 
\ea \label{ŽP.75}
\textit{ētir mebo ēme kārēš pene kermē} \\ 
\gll ētir me-b-o ēme kār-ē=š pene ∅-ker-mē \\ 
 no\_longer \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{1pl} task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `We can no longer do anything for him.'
\z 
 
\ea \label{ŽP.78}
\textit{meger kīyānmēš bilo weno be feqīgerī} \\ 
\gll meger ∅-kīyān-mē=š bi-l-o ∅-wen-o be feqīgerī \\ 
 if\_only \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}read\textsc{-3sg} by Islamic\_jurisprudence \\ 
\glt `If only we could send him to study Islamic jurisprudence.'
\z 
 
\ea \label{ŽP.186}
\textit{mēnē, mēnē pēse māčmē biɫbeřū hurwezā milegāw} \\ 
\gll m-ē-nē m-ē-nē pēse m-āč-mē biɫbeř=ū hur-wez-ā milegā=ū \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} as\_if \textsc{ind-}say\textsc{.prs}\textsc{-1pl} \textsc{pn}=and \textsc{pvb-}climb\textsc{.prs}\textsc{-3pl} top\_of\_mountain=and \\ 
\glt `They (the girl and her uncle) came [to the direction of Hawraman]; they climbed a peak of the mountain'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.228}
\textit{yomā āne tišit miđo} \\ 
\gll yo=mā āne tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give [pomegranate] molasses.'
\z 
 
\ea \label{ŽP.229}
\textit{yoīčmā kučē pīyāzīt miđo} \\ 
\gll yo=īč=mā kuč-ē pīyāz-ī=t mi-đe-o \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} little\textsc{-indf} onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give you some onions.'
\z 
 
\ea \label{ŽP.230}
\textit{yoīč mā yoīč māčo, kāke gīyān! minīč šimšū yerū heywānēt mido pene} \\ 
\gll yo=īč=mā yo=īč m-āč-o kāke gīyān min=īč šimšī=ū yeher=ū heywān-ē=t mi-de-o pene \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother.vob dear.\textsc{voc} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `One [said], One person said, ‘Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{ŽP.231}
\textit{čāštekē kermē} \\ 
\gll čāštī-ekē ∅-ker-mē \\ 
 meal\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `Let us make a meal.'
\z 
 
\ea \label{ŽP.232}
\textit{čāštekē kermē kermēš be zemāwine peyt} \\ 
\gll čāštī-ekē ∅-ker-mē ∅-ker-mē=š be zemāwine pey=t \\ 
 meal\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl}\textsc{=3sg} by wedding for\textsc{=\textsc{2sg}} \\ 
\glt `Let’s cook a meal and hold a wedding for you. '
\z 
 
\ea \label{ŽP.233}
\textit{nīkāhiš kermē peyt} \\ 
\gll nīkāh=iš ∅-ker-mē pey=t \\ 
 marriage\textsc{.m}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} for\textsc{=\textsc{2sg}} \\ 
\glt `Let us marry her to you. '
\z 
 
\ea \label{ŽP.234}
\textit{qey mekero nimāymē peket gino} \\ 
\gll qey me-ker-o ni-m-āz-mē peke=t ∅-gin-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-1pl} strenght\textsc{.f}\textsc{=\textsc{2sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `It is all right. We will not let you worry about it.'
\z 
 
\ea \label{ŽP.235}
\textit{nimāymē peket gino} \\ 
\gll ni-m-āz-mē peke=t ∅-gin-o \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-1pl} strenght\textsc{.f}\textsc{=\textsc{2sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `We will not let you worry about it.'
\z 
 
\ea \label{ŽP.241}
\textit{čāštekē hiɫošīnekē kermē} \\ 
\gll čāštī-ekē hiɫošīne-ekē ∅-ker-mē \\ 
 meal\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{pn}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `Let us make the meal called Hiloshina.'
\z 
 
\ea \label{ŽP.264}
\textit{lālo mēwe wesāyil mārowe pey kināčē pēse īsey māčmē teleš pey berā} \\ 
\gll lālo m-ē=we wesāyil m-ār-o=we pey kināčē pēse īse-ī m-āč-mē tele=š pey ∅-ber-ā \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} appliance\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=compl} to girl\textsc{.dir}\textsc{.f} as\_if now\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-1pl} wedding\_gift\textsc{=3sg} to \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \\ 
\glt `The uncle went back. He brought household equipment for the girl, just like nowadays when we say, ‘They [the bride’s family] take gifts for her [the bride].’'
\z 
 
\ea \label{BP.9}
\textit{ā zemān bābā xwāđā bīyen, ā zemān ī mexlūqū hewrāmānātū ēme reʕīsšā je ʕērāqew čā kurdewārīū ʕērāqīne bīyen} \\ 
\gll ā zemān bābā xwāđā bīye=n ā zemān ī mexlūq=ū hewrāmān-āt=ū ēme reʕīs=šā je ʕērāq=ew čā kurdewārī=ū ʕērāq-ī=ne bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} time\textsc{.m} \textsc{dem.prox} people\textsc{=ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{=ez.gen} \textsc{1pl} leader\textsc{.m}\textsc{=3pl} from \textsc{pn}\textsc{=\textsc{post}} there kurdish\_region\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Babab Khwada, the people of Hawraman were under Iraq’s rule, under the control of Kurdish regions in Iraq.'
\z 
 
\ea \label{BP.10}
\textit{meselen āđīšā hukim kerdēnmē} \\ 
\gll meselen āđīšā hukim kerdē=nmē \\ 
 for\_example 3pl\textsc{.obl} rule\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `For instance, they ruled over us.'
\z 
 
\ea \label{BP.11}
\textit{zemānū bābāw pīr šelīyārī bābā xwāđāy āā je ʕērāqne ēme hukmū hewrāmānīšā kerden, biřew bārešā sānān} \\ 
\gll zemān=ū bābā=ū pīr šelīyār-ī bābā xwāđā-ī āā je ʕērāq=ne ēme hukm=ū hewrāmān-ī=šā kerde=n biře=ū bāre=šā sānā=n \\ 
 time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{intj} from \textsc{pn}\textsc{=\textsc{post}} \textsc{1pl} rule\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Pir Shaliyar’s grandfather, we were ruled by Iraq, [and] they would collect tax.'
\z 
 
\ea \label{BP.12}
\textit{biřew bārešā sānān, āmēnē pēse čāweɫī begzāyāw wēmā} \\ 
\gll biře=ū bāre=šā sānā=n āmē=nē pēse č=ā eweɫ-ī begzā-yā=ū wē=mā \\ 
 portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} nobleman\textsc{-obl}\textsc{.pl}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1pl}} \\ 
\glt `They would collect tax. They would come (here) [and] like our noblemen..'
\z 
 
\ea \label{BP.17}
\textit{ā pāse hetā hečkesē ženīš dē šū, bešepīyākew ēme begzāyē ēme} \\ 
\gll ā pāse hetā hečkes-ē ženī=š de-ē šū bešepīyā-eke=ū ēme begzāye-ē ēme \\ 
 \textsc{dem.dist} like even whoever\textsc{-indf} woman.f\textsc{=3sg} give.prs\textsc{-pstc}\textsc{.3sg} husband\textsc{.m} quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{1pl} nobleman\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} \\ 
\glt `[Things were] like this. Even when someone wanted to get his daughter married, he had to pay a quota  to our noblemen.'
\z 
 
\ea \label{BP.18}
\textit{eger ženīmā dē be bešepīyākew yo terī čīwēš zereriš sānē čene wāčēnēš sūrāne} \\ 
\gll eger ženī=mā de-ē be bešepīyā-eke=ū yo ter-ī čīw-ē=š zerer=iš sān-ē čene wāč-ēn-ē=š sūrāne \\ 
 if woman.f\textsc{=\textsc{1pl}} give.prs\textsc{-cond} to quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} thing\textsc{.m}\textsc{-indf}\textsc{=3sg} harm\textsc{.m}\textsc{=3sg} buy.prs\textsc{-pstc}\textsc{.3sg} from say.prs.\textsc{sbjv-}pstc\textsc{-3pl}\textsc{=3sg} \textsc{pn} \\ 
\glt `If one\footnote{Lit. we.} had married women from the quota of another [lord], then they [the lord of the bride's village] would charge him [the bride’s father]. This money is called Surana.'
\z 
 
\ea \label{BP.24}
\textit{e čē biřew bārešā sānān je hewrāmānī pēse begzāyekāw ī beynū ēme} \\ 
\gll e=čē biře=ū bāre=šā sānā=n je hewrāmān-ī pēse begzāye-ekā=ū ī beyn=ū ēme \\ 
 in=here portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{-obl}\textsc{.m} like nobleman\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=ez.gen} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `[who] would collect taxes from Hawraman, [just] like our noblemen [in recent] times.'
\z 
 
\ea \label{BP.42}
\textit{yānew ēme neferēšā kīyāsenū} \\ 
\gll yāne=ū ēme nefer-ē=šā kīyāse=n=ū \\ 
 house\textsc{.m}\textsc{=ez.gen} \textsc{1pl} person\textsc{.m}\textsc{-indf}\textsc{=3pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `One [tax collector] would be sent to our house;'
\z 
 
\ea \label{BP.55}
\textit{girđtā beydē mizgī bā bīmēwe yo} \\ 
\gll girđ=tā b-e-ydē mizgī bā ∅-b-īmē=we yo \\ 
 all=\textsc{2pl} \textsc{sbjv-}come.prs-\textsc{2pl} mosque\textsc{.m} \textsc{hort} \textsc{sbjv-}be.prs-\textsc{1pl}\textsc{=compl} one\textsc{.m} \\ 
\glt `You all come to the mosque; let us be one.'
\z 
 
\ea \label{BP.56}
\textit{jā girđmā bilmē pey ʕērāqī pey lāw ī āġeyšā bilmērewo,} \\ 
\gll jā girđ=mā bi-l-mē pey ʕērāq-ī pey lā=ū ī āġe-ī=šā bi-l-mē=re=o \\ 
 afterwards all\textsc{=\textsc{1pl}} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} for at\_the\_place\_of\textsc{=ez.gen} \textsc{dem.prox} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}\textsc{=compl} \\ 
\glt `Then, we all go to Iraq, to their agha. We go down there.'
\z 
 
\ea \label{BP.60}
\textit{ēmeyč māčmē, ī meʕmūrē šime kīyāsēnē bā biřew bāremā dēw nāmēwe bārāšo} \\ 
\gll ēme=yč m-āč-mē ī meʕmūr-ē šime kīyāsē=nē ∅-b-ā biře=ū bāre=mā dā-ē=ū ne-āmā-ē=we b-ār-ā=š=o \\ 
 \textsc{1pl}\textsc{=add} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{ind-}be.prs\textsc{-3pl} portion\textsc{.m}=and profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}bring.prs\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `We will say, ‘We paid tax to the officers whom you had sent, [however,] they didn’t return to bring it back.'
\z 
 
\ea \label{BP.63}
\textit{ēmeyč muxteserū kelāmī ā īney wāčmē,} \\ 
\gll ēme=yč muxteser=ū kelām-ī ā īney wāč-mē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} \textsc{dem.dist} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} say.prs.sbjv\textsc{-\textsc{1pl}} \\ 
\glt `As for us, to cut a long story short, we say this.'
\z 
 
\ea \label{BP.68}
\textit{ēme milmē īney māčmē} \\ 
\gll ēme mi-l-mē īney m-āč-mē \\ 
 \textsc{1pl} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \\ 
\glt `Then, we will go [to Iraq] and say this (i.e. what we agreed on)'
\z 
 
\ea \label{BP.69}
\textit{īney māčmē.’} \\ 
\gll īney m-āč-mē \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \\ 
\glt `We will say this.’'
\z 
 
\ea \label{BP.78}
\textit{wāče, bē dērā ēme milāy milmē pey ʕērāqī be peyē} \\ 
\gll wāč-e b-ē dēr=ā ēme mi-lāy mi-l-mē pey ʕērāq-ī be peyē \\ 
 say.prs-\textsc{imp-}\textsc{2sg} be.prs\textsc{-pstc}\textsc{.3sg} late\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} \textsc{ind-}go\textsc{.adv} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} to foot\textsc{.obl} \\ 
\glt `Tell him, “Come, it’s getting late. We are going to Iraq on foot.'
\z 
 
\ea \label{BP.79}
\textit{milāymē pey ʕērāqī milāymē ī xeberī bermēwe} \\ 
\gll mi-lā-ymē pey ʕērāq-ī mi-lā-ymē ī xeber-ī ∅-ber-mē=we \\ 
 \textsc{ind-}go.prs \textsc{1pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs \textsc{1pl} \textsc{dem.prox} news\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}take.prs\textsc{-\textsc{1pl}}\textsc{=compl} \\ 
\glt `We are going to Iraq. We are going [there] to pass this news [to the agha in Iraq]”'
\z 
 
\ea \label{BP.87}
\textit{jārēne jimdēre mexɫūq hečē henē mizgīne bā bilimdē pey ʕērāqī} \\ 
\gll jār-ē=ne ∅-jim-dē=re mexɫūq hečē hen-ē mizgī=ne ∅-b-ā bi-l-imdē pey ʕērāq-ī \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} people\textsc{.m} no\_matter\_how\_many exist\textsc{-3pl} mosque\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}/\textsc{2pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They all set off; every person who was in the mosque [said], ‘Let’s go to Iraq.'
\z 
 
\ea \label{BP.88}
\textit{bilmēre wā ʕērāqī} \\ 
\gll bi-l-mē=re wā ʕērāq-ī \\ 
 \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} there \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Let’s go there, to Iraq. '
\z 
 
\ea \label{BP.89}
\textit{ētir ēme xo diruē meyeymē} \\ 
\gll ētir ēme xo diruē me-de-ymē \\ 
 \textsc{disc.ptcl} \textsc{1pl} dis.ptc lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `We are not going to lie.'
\z 
 
\ea \label{BP.90}
\textit{māčmē ī meʕmūrū šime āmē lāw ēme} \\ 
\gll m-āč-mē ī meʕmūr=ū šime āmā-ē lā=ū ēme \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} come\textsc{.pst}\textsc{-3pl} at\_the\_place\_of\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `We will say [to the agha], “your officers came to our region".'
\z 
 
\ea \label{BP.91}
\textit{biřew bāremā dēw} \\ 
\gll biře=ū bāre=mā dā-ē=ū \\ 
 portion\textsc{.m}\textsc{=ez.gen} profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \\ 
\glt `We paid them tax.'
\z 
 
\ea \label{BP.95}
\textit{ēmeyč muxteserū kelāmī, xo ēme diruē meyeymē} \\ 
\gll ēme=yč muxteser=ū kelām-ī xo ēme diruē me-de-ymē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} dis.ptc \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `As for us, to cut a long story short, we are not lying.'
\z 
 
\ea \label{BP.96}
\textit{ī meʕmūrū šimemā sere biřyēnē} \\ 
\gll ī meʕmūr=ū šime=mā sere biřyē=nē \\ 
 \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl}\textsc{=\textsc{1pl}} head\textsc{.m} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We have decapitated your officers.'
\z 
 
\ea \label{BP.97}
\textit{seremā biřyē, īne sereū ēmew īne šimšērū šime} \\ 
\gll sere=mā biřī-ē īne sere=ū ēme=ū īne šimšēr=ū šime \\ 
 head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} \textsc{2pl} \\ 
\glt `We have decapitated them. Here are their heads and the agha’s swords.\footnote{Lit. These are our heads and these are the agha’s swords.}'
\z 
 
\ea \label{BP.101}
\textit{māčā ēme čī bilmēre ʕērāq wēmā kermē zīnānū} \\ 
\gll m-āč-ā ēme čī bi-l-mē=re ʕērāq wē=mā ∅-ker-mē zīnān=ū \\ 
 \textsc{ind-}say.prs\textsc{-3pl} \textsc{1pl} why \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} \textsc{pn} \textsc{reflx}\textsc{=\textsc{1pl}} \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}} prison\textsc{.m}=and \\ 
\glt `They (the people) said, ‘Why should we go to Iraq and put ourselves in   prison? '
\z 
 
\ea \label{BP.102}
\textit{pādšā ezyetmā keronew misībetmā biđe ver} \\ 
\gll pādšā ezyet=mā ∅-ker-one=ū misībet=mā bi-đe-o ver \\ 
 king\textsc{.m} irritation\textsc{.m}\textsc{=\textsc{1pl}} \textsc{sbjv-}do.prs\textsc{-3sg}\textsc{=ez.gen} calamity\textsc{.m}\textsc{=\textsc{1pl}} \textsc{sbjv-}give.prs\textsc{-3sg} front \\ 
\glt `The king is going to upset us and give us a hard time.'
\z 
 
\ea \label{BP.103}
\textit{be xwā gēɫmēwe} \\ 
\gll be xwā ∅-gēɫ-mē=we \\ 
 to God\textsc{.m} \textsc{sbjv-}turn\textsc{-\textsc{1pl}}\textsc{=compl} \\ 
\glt `Indeed, we shall return [to the village]'
\z 
 
\ea \label{BP.109}
\textit{fermāwo, īne konē xeɫkeke čenū ēme bē?} \\ 
\gll ∅-fermāw-o īne ko=nē xeɫk-eke čenū ēme b-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} people\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} with\textsc{.ez.gen} \textsc{1pl} be.prs\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Baba Khwada) said, ‘Where are the people who were with us?'
\z 
 
\ea \label{BP.110}
\textit{ēme firē bēnmē. ṣew neferē bēnmē āmāymē řāre.} \\ 
\gll ēme fire-ē b-ēn-mē ṣew nefer-ē b-ēn-mē āmā-īmē řā=re \\ 
 \textsc{1pl} a\_lot\textsc{-pl} be.\textsc{prs}\textsc{-pstc}\textsc{-\textsc{1pl}} hundred person\textsc{.m}\textsc{-dir.pl} be.\textsc{prs}\textsc{-pstc}\textsc{-\textsc{1pl}} come\textsc{.pst}-\textsc{1pl} road.f\textsc{=\textsc{post}} \\ 
\glt `We were a large number. We were one hundred  people who came here.'
\z 
 
\ea \label{BP.117}
\textit{ētir āđē wātenšā, ēme diruē meyeymē} \\ 
\gll ētir āđē wāte=n=šā ēme diruē me-de-ymē \\ 
 \textsc{disc.ptcl} 3pl\textsc{.dir} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `They said, ‘We are not going to lie.'
\z 
 
\ea \label{BP.118}
\textit{řūē memenmērew je kesī metersmē īlā je xwāyne bo} \\ 
\gll řūē me-men-mē=re=ū je kes-ī me-ters-mē īlā je xwā-ī=ne ∅-b-o \\ 
 face\textsc{.obl} \textsc{neg-}remain.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}=and from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\_afraid.prs\textsc{-\textsc{1pl}} except from God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{sbjv-}be.prs\textsc{-3sg} \\ 
\glt `We are not going to implore [the king]. We are not afraid of anyone except God.'
\z 
 
\ea \label{BP.119}
\textit{milmē īney qomīyān,} \\ 
\gll mi-l-mē īney qomyā=n \\ 
 \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We are going [there since] this has happened. '
\z 
 
\ea \label{BP.120}
\textit{jereyāniš kermē pey pādšāw ʕērāqī} \\ 
\gll jereyān=iš ∅-ker-mē pey pādšā=w ʕērāq-ī \\ 
 story\textsc{.m}\textsc{=3sg} \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}} for king\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We will tell the king the story.'
\z 
 
\ea \label{BP.121}
\textit{māčmē ēme serū īney ī pīyēmā kuštēnē”’} \\ 
\gll m-āč-mē ēme ser=ū īney ī pīyē=mā kuštē=nē \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{1pl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox} man\textsc{.dir}\textsc{.pl}\textsc{=\textsc{1pl}} kill\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We say, ‘We have killed these men because of this.’'
\z 
 
\ea \label{BP.122}
\textit{ā wextī reʕīsū milāre ʕērāq lāw āġekeyšā māčo seremā biřyēnē bābāw pīr šelīyārī bo, ā wextīyekey ħemey ġeybīū yoyč bo nāmē ħemey wirdīkɫe,} \\ 
\gll ā wext-ī reʕīs=ū mi-l-ā=re ʕērāq lā=ū āġe-ekey=šā m-āč-o sere=mā biřyē=nē bābā=ū pīr šelīyār-ī ∅-b-o ā wext-ī-ekey ħeme=ī ġeybī=ū yo=yč ∅-b-o nāmē ħeme=ī wirdīkɫe \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} leader\textsc{.m}\textsc{=ez.gen} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{pn} at\_the\_place\_of\textsc{=ez.gen} agha\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}say.prs\textsc{-3sg} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}=ez.\textsc{attr} invisible=and one\textsc{.m}\textsc{=add} \textsc{ind-}be.prs\textsc{-3sg} name.f \textsc{pn}=ez.\textsc{attr} small \\ 
\glt `The head of those who went to Iraq to the agha, saying, ‘We have decapitated [your men], was Pir Shaliyar’s grandfather, Baba Khwada. In addition [there were] Hama the invisible and Little Hama. '
\z 
 
\ea \label{BP.123}
\textit{ħemey wirdīkɫe ā wextī, pēse īsey māčā heɫāy neyāwān sinʕe zīnāniš nekermē yā čīw yā čāne} \\ 
\gll ħeme=ī wirdīkɫe ā wext-ī pēse īse-ī m-āč-ā heɫāy ne-yāwā=n sinʕe zīnān=iš ne-ker-mē yā čīw yā č=ā=ne \\ 
 \textsc{pn}=ez.\textsc{attr} small \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} like now\textsc{-obl}\textsc{.m} \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} age.f prison\textsc{.m}\textsc{=3sg} \textsc{neg-}do.prs\textsc{-\textsc{1pl}} either thing\textsc{.m} either in=\textsc{dem.dist}\textsc{=\textsc{post}} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) will say, ‘Hama Wirdikla has not reached adulthood; we won’t imprison him, and so on.’'
\z 
 
\ea \label{BP.124}
\textit{ā wextī milārewe māčā, wiɫā ī meʕmūrū šime kīyāsēbēnētā pey īne, īnemā āne heke šime gerekbē dēmā penew} \\ 
\gll ā wext-ī mi-l-ā=re=we m-āč-ā wiɫā ī meʕmūr=ū šime kīyāsē b-ēn-ē=tā pey īne īne=mā āne heke šime gerek b-ē dā-ē=mā pene=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.\textsc{prs}\textsc{-3pl}\textsc{=\textsc{povb}}\textsc{=compl} \textsc{ind-}say.prs\textsc{-3pl} by\_God \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl}=\textsc{2pl} for \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1pl}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} if \textsc{2pl} necessary be.prs\textsc{-pstc}\textsc{.3sg} give\textsc{.pst}\textsc{-3pl}\textsc{=\textsc{1pl}} to=and \\ 
\glt `Then, they went [to the agha and] said, ‘Indeed, the officers whom you had sent to us, whatever [ taxation] you had asked for, we gave them,'
\z 
 
\ea \label{BP.127}
\textit{ēmeyč muxteserū kelāmī seremā biřyēnē} \\ 
\gll ēme=yč muxteser=ū kelām-ī sere=mā biřyē=nē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `So, to cut a long story short, we decapitated them.'
\z 
 
\ea \label{BP.128}
\textit{īne serew ēmew īne šimšērū āġey} \\ 
\gll īne sere=ū ēme=ū īne šimšēr=ū āġe-ī \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Here are their heads and the agha’s swords.'
\z 
 
\ea \label{BP.156}
\textit{ā dey hešpišā dēndēne īse ēme tekbīrmā čēš bo čēš kermēū} \\ 
\gll =ā dey hešpišī-ā dē=ndē=ne īse ēme tekbīr=mā čēš ∅-b-o čēš ∅-ker-mē=ū \\ 
=\textsc{ptcl} \textsc{disc.ptcl} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}.\textsc{2pl}\textsc{=\textsc{post}} now \textsc{1pl} advice\textsc{=\textsc{1pl}} what \textsc{sbjv-}be.prs\textsc{-3sg} what \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}}=and \\ 
\glt `Look, you have gotten lice. What is the plan now for us? What should we do?'
\z 
 
\ea \label{BP.157}
\textit{ēme mezānmē tekbīrmā čēš bo īnīšā} \\ 
\gll ēme me-zān-mē tekbīr=mā čēš ∅-b-o īnīšā \\ 
 \textsc{1pl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} advice\textsc{=\textsc{1pl}} what \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `We don’t know what we should do, and so on.'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.161}
\textit{bizāne yānew kē pī činnē hāy wexte nānē čīwēšā dānī ārdenit pey ēme, ārdenit pey ēmew} \\ 
\gll bi-zān-e yāne=ū kē p=ī činnē hāy wext=e nān-ē čīw-ē=šā dā=nī ārde=n=it pey ēme ārde=n=it pey ēme=ū \\ 
 \textsc{sbjv-}know.prs-\textsc{2sg} house\textsc{.m}\textsc{=ez.gen} who at=\textsc{dem.prox} some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{=dem} meal\textsc{.m}\textsc{-indf} thing\textsc{.m}\textsc{-indf}\textsc{=3pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl}=and \\ 
\glt `see which household has given you food, which you brought to us, during this period!'
\z 
 
\ea \label{BP.181}
\textit{mēnē e čā ħemey ġeybī fermāwo, fermāwo, min mānīyēnmē e čē bā kučē nīšmēre} \\ 
\gll m-ē-nē e=čā ħeme=ī ġeybī ∅-fermāw-o ∅-fermāw-o min mānīyē=nmē e=čē bā kuč-ē ∅-nīš-mē=re \\ 
 \textsc{ind-}come.prs\textsc{-3pl} in=there \textsc{pn}=ez.\textsc{attr} invisible \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{1sg} tired\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} in=here \textsc{hort} little\textsc{-indf} \textsc{sbjv-}sit.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} \\ 
\glt `They went [to the region between Zawar and Tawela]. Hama Ghaybi said, ‘I, we are tired. Let us rest\footnote{Lit. sit.} a little bit here.'
\z 
 
\ea \label{BP.182}
\textit{seyēmēw min čurtēč kerū kučē būsū} \\ 
\gll sey-ēmē=ū min čurt-ē=č ∅-ker-ū kuč-ē b-ūs-ū \\ 
 rest.prs-\textsc{1pl}=and \textsc{1sg} nap\textsc{-indf}\textsc{=add} \textsc{ind-}do.prs\textsc{-\textsc{1sg}} little\textsc{-indf} \textsc{sbjv-}sleep.prs\textsc{-\textsc{1sg}} \\ 
\glt ` I shall rest,\footnote{Lit. My rest comes.} take a nap [and] sleep a little bit. They were a bit tired.'
\z 
 
\ea \label{BP.186}
\textit{kuřēmā bo nāmēš binyeydē miħemeđ} \\ 
\gll kuř-ē=mā ∅-b-o nāmē=š bi-nye-ydē miħemeđ \\ 
 boy\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{ind-}be.prs\textsc{-3sg} name.f\textsc{=3sg} \textsc{imp-}put.prs-\textsc{2pl} \textsc{pn} \\ 
\glt `We will have a son. Name him Mohammad.'
\z 
 
\ea \label{BP.191}
\textit{dērāw ēme milāy milmēwew řāmā dūrene, milāy milmēwew} \\ 
\gll dēr=ā=ū ēme mi-lāy mi-l-mē=we=ū řā=mā dūr-e=ne mi-lāy mi-l-mē=we=ū \\ 
 late\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{1pl} \textsc{ind-}go\textsc{.adv} \textsc{ind-}go.prs\textsc{-\textsc{1pl}}\textsc{=compl}=and road.f\textsc{=\textsc{1pl}} far\textsc{-f}\textsc{=cop}\textsc{.3sg}.f \textsc{ind-}go\textsc{.adv} \textsc{ind-}go.prs\textsc{-\textsc{1pl}}\textsc{=compl}=and \\ 
\glt `[Baba Khwada said], ‘It’s getting late. We shall be going. We have a long way [to go]. We shall be going.’'
\z 
 
\ea \label{BP.202}
\textit{mēnēwe māčā, hāɫ jeryān ēme e pēsemā ser qomīyān} \\ 
\gll m-ē-nē=we m-āč-ā hāɫ jeryān ēme e=pēse=mā ser qomyā=n \\ 
 \textsc{ind-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say.prs\textsc{-3pl} state\textsc{.m} story\textsc{.m} \textsc{1pl} in=like\textsc{=\textsc{1pl}} on happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went back [and] said, ‘We have been through such and such.’'
\z 
 
\ea \label{ŠJ.31}
\textit{ēme xo mātiɫū tonīmē} \\ 
\gll ēme xo mātiɫ=ū to=nīmē \\ 
 \textsc{1pl} dis.ptc suspended\textsc{=ez.gen} \textsc{2sg}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `We have been waiting for you.'
\z 
 
\ea \label{ŠJ.32}
\textit{pey lāw to āmēnmēw} \\ 
\gll pey lā=ū to āmē=nmē=ū \\ 
 to at\_the\_place\_of\textsc{=ez.gen} \textsc{2sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}=and \\ 
\glt `We have come to you.’'
\z 
 
\ea \label{ŠJ.34}
\textit{māčā pīyew jāfir sānīnmē} \\ 
\gll m-āč-ā pīye=w jāfir sān-ī=nmē \\ 
 \textsc{ind-}say.prs\textsc{-3pl} man\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `They said, ‘We are Jafir Sān’s men.’'
\z 
 
\ea \label{ŠJ.36}
\textit{āmēnmē jafir sānī fermāwān lodē} \\ 
\gll āmē=nmē jafir sān-ī fermāwā=n ∅-lo-dē \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \\ 
\glt `We have come here. Jafir San said, ‘Go [and bring him].’'
\z 
 
\ea \label{ŠJ.44}
\textit{māčo qurbān! be xwā xuřyeymēš venī} \\ 
\gll m-āč-o qurbān be xwā xuřye-ymē=š venī \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} to God\textsc{.m} shout\textsc{.pst} \textsc{1pl}\textsc{=3sg} at \\ 
\glt `They said, ‘Sir, indeed, [as] he shouted at us.'
\z 
 
\ea \label{ŠJ.45}
\textit{weɫē bīyēnmē āro duē řoē ētir mezānmē koge milmē} \\ 
\gll weɫē bīyē=nmē āro duē řo-ē ētir me-zān-mē koge mi-l-mē \\ 
 front be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} today two day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{disc.ptcl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} where \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \\ 
\glt `We have spread out for two days; we didn’t know where we were going.’'
\z 
 
\ea \label{ŠJ.88}
\textit{kārū ēme mānā īse īnā desū xwāyne} \\ 
\gll kār=ū ēme mānā īse īnā-∅ des=ū xwā-ī=ne \\ 
 task\textsc{.m}\textsc{=ez.gen} \textsc{1pl} that\_is now deic\textsc{-3sg}\textsc{.m} hand\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Our task now rests within God’s hands.'
\z 
 
\ea \label{ŠJ.89}
\textit{herčē ārezūmā bo herčē xwāy meyliš bo tāwo bispāro pēme} \\ 
\gll herčē ārezū=mā ∅-b-o herčē xwā-ī meyl=iš ∅-b-o ∅-tāw-o bi-spār-o p=ēme \\ 
 however\_much wish\textsc{=\textsc{1pl}} \textsc{sbjv-}be.prs\textsc{-3sg} however\_much God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg} \textsc{sbjv-}be.prs\textsc{-3sg} \textsc{ind-}can.prs\textsc{-3sg} \textsc{sbjv-}hand\_over.prs\textsc{-3sg} to=\textsc{1pl} \\ 
\glt `Whatever wishes we might have depend on God’s willingness to grant [them] to us.'
\z 
 
\ea \label{KŠ.14}
\textit{māčmēš pene suɫtān mehmūđ bī} \\ 
\gll m-āč-mē=š pene suɫtān mehmūđ bī-∅ \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-\textsc{1pl}}\textsc{=3sg} to sultan \textsc{pn} be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `[Let’s say] his (the king's) name was Sultan Mahmud.'
\z 
 
\ea \label{KŠ.40}
\textit{kināčēw pāđšāyč māčo, kuřū šuwāneyč bārdē ēme tāqetmā pey āney mē} \\ 
\gll kināčē=ū pāđšā-ī=č m-āč-o kuř=ū šuwāne-ī=č b-ār-dē ēme tāqet=mā pey āney m-ē \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} \textsc{1pl} persistence\textsc{.m}\textsc{=\textsc{1pl}} to \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `The king’s daughter said, ‘Fetch the shepherd’s son. He will entertain us.\footnote{Lit. Our patience comes by him.}’'
\z 
 
\ea \label{KŠ.43}
\textit{māčo be eyāšū tāteyšū āđīšā, bā īne menowe pey nokerū ēme epēne pey werdesīmā pey hermānberū yāneymā} \\ 
\gll m-āč-o be eyā=š=ū tāte-ī=š=ū āđīšā bā īne ∅-men-o=we pey noker=ū ēme epēne pey werdes-ī=mā pey hermānber=ū yāne-ī=mā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by mother\textsc{=3sg}=and father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and 3pl\textsc{.obl} \textsc{hort} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} to servant\textsc{.m}\textsc{=ez.gen} \textsc{1pl} here to attendant\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} to worker\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `She (the girl) said to her parents and others , ‘Let him stay here as our servant here, as our attendant, as our domestic worker.'
\z 
 
\ea \label{ŽH.6}
\textit{wiɫātēwe termā hen ew dīmew ēme, pēsew zāwerīū tewēɫey} \\ 
\gll wiɫāt-ēwe ter=mā hen-∅ ew dīm=ew ēme pēse=ū zāwer-ī=ū tewēɫe-ī \\ 
 country\textsc{.m}\textsc{-indf} other\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} \textsc{dem.dist} side\textsc{.m}\textsc{=\textsc{post}} \textsc{1pl} like\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `There is another country on the other side of us; something like Zawar (in Iran) and Tawela (in Iraq) village.'
\z 
 
\ea \label{ŽH.11}
\textit{māčo, dey bilmē minīč mewt čene} \\ 
\gll m-āč-o dey bi-l-mē min=īč m-e-ū=t čene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} 1sg\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{2sg}} with \\ 
\glt `He (Hayas) said, ‘Let’s go, I am also coming with you.’'
\z 
 
\ea \label{ŽH.12}
\textit{milā řāhur māčo, heyās māčoš pene, kāke gīyān pilekānēwe binye bā bilmē!’} \\ 
\gll mi-l-ā řā=hur m-āč-o heyās m-āč-o=š pene kāke gīyān pilekānī-ēwe bi-nye-e bā bi-l-mē \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to brother.vob dear.\textsc{voc} stair\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}put\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \\ 
\glt `They set off. He, said, [i.e.] Hayas said to him, ‘Dear brother, set up a ladder for us to move !’'\footnote{A cryptic message meaning ‘tell a story’.}
\z 
 
\ea \label{ŽH.23}
\textit{māčo, ne wiɫā řoɫe, āmānē pāynū serekoē wāčmēš pene be zuwānū hewrāmīyānew wēmā derey} \\ 
\gll m-āč-o ne wiɫā řoɫe āmā-ānē pāyn=ū sereko-ē ∅-wāč-mē=š pene be zuwān=ū hewrāmī-yāne=ū wē=mā dere-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no by\_God child.\textsc{voc} come\textsc{.pst}\textsc{-1sg} bottom\textsc{=ez}\textsc{.gen} uphill\textsc{.m}\textsc{-indf} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-1pl}\textsc{=3sg} to by language\textsc{.m}\textsc{=ez}\textsc{.gen} Hawrami-vernacular\_of\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} valley\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘No, child, I went to the bottom of the slope —we call it ‘daray’ in our Hawrami language—'
\z 
 
\ea \label{ŽH.25}
\textit{wātiš, pilekānē čine bā bilmē!”} \\ 
\gll wāt=iš pilekānī-ē ∅-čin-e bā bi-l-mē \\ 
 say\textsc{.pst}\textsc{=3sg} stair\textsc{\textsc{.f}}\textsc{-indf} \textsc{ind-}set\_up\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \\ 
\glt `He said, “Set up a ladder for us to move!”’'
\z 
 
\ea \label{ŽH.27}
\textit{wātiš, tāte āne wāteniš, hekāyetē weš gīsnene, bā ī řē kučē keme kermē be hekāyetekeyt rāz biřmē} \\ 
\gll wāt=iš tāte āne wāte=n=iš hekāyet-ē weš-∅ ∅-gīsn-e=ne bā ī řē kuč-ē kem-e ∅-ker-mē be hekāyet-ekey=t rāz ∅-biř-mē \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} tale\textsc{.m}\textsc{-indf} well-m \textsc{imp-}start\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{hort} \textsc{dem.prox} road\textsc{.obl}\textsc{\textsc{.f}}.sg little\textsc{-indf} s\textsc{hort}\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} by tale\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{2sg}} tale\textsc{.m} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl} \\ 
\glt `She said, ‘Father, he had said, “Start [telling] a good tale, so that we [are entertained] and our way seems shorter. Let us finish your tale.”'
\z 
 
\ea \label{ŽH.30}
\textit{wātiš, qisēmā nekerdē āmāymē nizīkū dege,} \\ 
\gll wāt=iš qisē=mā ne-kerd-ē āmā-īmē nizīk=ū dege \\ 
 say\textsc{.pst}\textsc{=3sg} talk\textsc{.dir}\textsc{.pl}\textsc{=1pl} \textsc{neg-}do\textsc{.pst}\textsc{-3pl} come\textsc{.pst} \textsc{1pl} close\textsc{=ez}\textsc{.gen} village\textsc{.obl}\textsc{\textsc{.f}} \\ 
\glt `He said, ‘We didn’t talk [until] we got close to the village.'
\z 
 
\ea \label{ŽH.50}
\textit{ā wextīyekey, māčo be tāteyš, dā ā mēmānī bāre pey yāneymā!’} \\ 
\gll ā wext-ī-ekey m-āč-o be tāte-ī=š dā ā mēmān-ī b-ār-e pey yāne-ī=mā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{hort} \textsc{dem.dist} guest\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \\ 
\glt `The, she (the older sister) said to her father, ‘Bring that guest home!’'
\z 
 
\ea \label{ŽH.55}
\textit{māčo, wārđmā xizmetīū tāšāmā sere} \\ 
\gll m-āč-o wārđ=mā xizmet-ī=ū tāšā=mā sere \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} eat\textsc{.pst}\textsc{=1pl} service\textsc{.m}\textsc{-obl}\textsc{.m}=and shave/cut\textsc{.pst}\textsc{=1pl} head\textsc{.m} \\ 
\glt `He (Hayas) said [in a poem], ‘We ate at your sevice, and shaved our head,'
\z 
 
\ea \label{ŽH.56}
\textit{kināčēčmā gerekene binyeymē were’} \\ 
\gll kināčē=č=mā gerek-e=ne bi-nye-ymē were \\ 
 girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{=add}\textsc{=1pl} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \textsc{sbjv-}put\textsc{.prs} \textsc{1pl} front \\ 
\glt `We want the daughter; we would like to put her in front [for marriage]'
\z 
 
\ea \label{ŽH.57}
\textit{ā wextīyekey āyeč māčo, weɫāhī tāte, dāwāw ēme yoymā kero’} \\ 
\gll ā wext-ī-ekey āye=č m-āč-o weɫāhī tāte dāwā=w ēme yo-ī=mā ∅-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{\textsc{.f}}\textsc{.dir}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed father\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{1pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then she (the eldest daughter) said, ‘Indeed, father, he is asking for the hand of one of us.’'
\z 
 
\ea \label{ŽH.87}
\textit{māčo, īnē māčmēšā hēɫe’} \\ 
\gll m-āč-o īnē m-āč-mē=šā hēɫe \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.3pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3pl} egg\textsc{.m} \\ 
\glt `He (the sultan) said, ‘We call these eggs.’'
\z 
 
\ea \label{ŽH.110}
\textit{šūnū duē řoeyre, kināčekē kīyāno lāw suɫtānī māčo, ēme pīyākemā kīyāstiš meʕmūrīyet nīyā, nāmāno’} \\ 
\gll šūn=ū duē řo-e-ī=re kināčē-ekē ∅-kīyān-o lā=ū suɫtān-ī m-āč-o ēme pīyā-eke=mā kīyāst-∅=iš meʕmūrīyet nīye=ā ne-āmā=n=o \\ 
 after\textsc{=ez}\textsc{.gen} two\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} sul\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} send\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=3sg} mission\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `After the second day, the girl (his wife) sent (someone) to the Sultan to tell [him], ‘He (the sultan) sent my husband on a mission, [however] he is not around; He hasn’t come back [yet].’'
\z 
 
\ea \label{RE.2}
\textit{ēme eger zemānē yā qeymīēmā yā zemānū wēmā ženīmā ārdēbo} \\ 
\gll ēme eger zemān-ē yā qeymī-ē=mā yā zemān=ū wē=mā ženī=mā ārdē b-o \\ 
 \textsc{1pl} if time\textsc{.m}\textsc{-indf} either the\_elderly\textsc{-indf}\textsc{=\textsc{1pl}} either time\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1pl}} woman\textsc{.dir}\textsc{.f}\textsc{=\textsc{1pl}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f} be\textsc{.prs}\textsc{-3sg} \\ 
\glt ` Once, in the time of our elders or in our time, if one of us married,'
\z 
 
\ea \label{RE.3}
\textit{āmēnē lūēnmē werū berew kābrāy, hāwmāɫū wēmā} \\ 
\gll āmē=nē lūē=nmē wer=ū bere=ū kābrā-ī hāwmāɫ=ū wē=mā \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} front\textsc{.m}\textsc{\textsc{=ez.gen}} door\textsc{.m}\textsc{\textsc{=ez.gen}} man\textsc{-obl}\textsc{.m} neighbour\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1pl}} \\ 
\glt `they (our elderly family members) would go to the door of the fellow (the girl’s father), [who was] the neighbour.'
\z 
 
\ea \label{RE.4}
\textit{bābe gīyān ī kināčēte yā tātemā lūān yā māmomā yā kesūkārmā} \\ 
\gll bābe gīyān ī kināčē=t=e yā tāte=mā lūā=n yā māmo=mā yā kesūkār=mā \\ 
 father.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} either father\textsc{.m}\textsc{=\textsc{1pl}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} either paternal\_uncle\textsc{=\textsc{1pl}} either relative\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `They would say, ‘Dear brother,\footnote{Lit. father.} this daughter of yours \—either our father would go [to the girl’s father], or our uncle, or a relative of ours—'
\z 
 
\ea \label{RE.41}
\textit{ī memliketū ēme ne welē wiɫātāre ētir tāte bīyen čūēw desū šuwāney} \\ 
\gll ī memliket=ū ēme ne welē wiɫāt-ā=re ētir tāte bīye=n čū-ē=w des=ū šuwāne-ī \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} no but region\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{disc.ptcl} father\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} stick\textsc{.m}\textsc{-indf}\textsc{=ez.gen} hand\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[It is not common] in our region, but elsewhere, the father has become like the stick in the shepherd’s hand.\footnote{Meaning that he has no power over whom his children will marry.}'
\z 
 
\ea \label{RE.42}
\textit{ēme zemānmā řisūmātū ewsāymā ečīne bīyen} \\ 
\gll ēme zemān=mā řisūmāt=ū ewsā-ī=mā e=čīne bīye=n \\ 
 \textsc{1pl} time\textsc{.m}\textsc{=\textsc{1pl}} traditions\textsc{.m}\textsc{=ez.gen} past\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} in=like be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the old times, our customs were like this. '
\z 
 
\ea \label{RE.43}
\textit{e čē duē begēmā bīyēnē lūēnē} \\ 
\gll e=čē duē beg-ē=mā bīyē=nē lūē=nē \\ 
 in=here two Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Then, even if you might not have given [your daughter in marriage], We had two begs ‘lords’ to whom they (the girl's family) would go.'
\z 
 
\ea \label{RE.51}
\textit{ewīč wāten bā bizānmē} \\ 
\gll ew=īč wāte=n bā bi-zān-mē \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{hort} \textsc{sbjv-}know\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `He (the lord) would say, ‘Let me see.’\footnote{Lit. Let us know.}'
\z 
 
\ea \label{RE.56}
\textit{zemānē ēme ī begēmā bīyēnē} \\ 
\gll zemān-ē ēme ī beg-ē=mā bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} \textsc{1pl} \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Once we had these noblemen [in our region]'
\z 
 
\ea \label{RE.62}
\textit{ī memliketū ēme řisūmātū āwe zemānīš e čīne bīyen} \\ 
\gll ī memliket=ū ēme řisūmāt=ū āwe zemān-ī=š e=čīne bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} traditions\textsc{.m}\textsc{=ez.gen} that time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} in=like be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The traditions of earlier times were like this in our region.'
\z 
 
\ea \label{RE.63}
\textit{ī begēmā bīyēnē, begē} \\ 
\gll ī beg-ē=mā bīyē=nē beg-ē \\ 
 \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} Beg\textsc{.m}\textsc{-indf} \\ 
\glt `We had these noblemen.'
\z 
 
\ea \label{RE.64}
\textit{bē ā begā ēme netāwānmā ħīč kermē} \\ 
\gll bē ā beg-ā ēme ne-tāwā=n=mā ħīč ∅-ker-mē \\ 
 without \textsc{dem.dist} Beg\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{1pl} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1pl}} nothing \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `Without them, we were not able to do anything.'
\z 
 
\ea \label{RE.65}
\textit{berdēnmēšā lo keney bē heqū} \\ 
\gll berdē=nmē=šā lo keney bē heq=ū \\ 
 take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=3pl} fodder\_grass mow.in without salary\textsc{.m}=and \\ 
\glt `They would take us to mow fodder grass for free [i.e. without wages].'
\z 
 
\ea \label{RE.66}
\textit{berdēnmēšā diřē keney bē heqū} \\ 
\gll berdē=nmē=šā diřē keney bē heq=ū \\ 
 take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=3pl} prickle mow.in without salary\textsc{.m}=and \\ 
\glt `They would take us to mow prickles for free [i.e. without wages].'
\z 
 
\ea \label{RE.67}
\textit{lūēnmē bāyē māsēmā berdēnē bē ewū} \\ 
\gll lūē=nmē bāye-ē mās-ē=mā berdē=nē bē ew=ū \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} pot\textsc{.m}\textsc{-dir}\textsc{.pl} yoghurt\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1pl}} take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} without 3sg\textsc{.dir}\textsc{.m}=and \\ 
\glt `We would go [to the noblemen and] take them bowls of yoghurt free [of charge]'
\z 
 
\ea \label{RE.68}
\textit{desmā gērten sīnemāwe siɫām kerdey} \\ 
\gll des=mā gērte=n sīne=mā=we siɫām kerdey \\ 
 hand\textsc{.m}\textsc{=\textsc{1pl}} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} chest\textsc{.m}\textsc{=\textsc{1pl}}\textsc{=\textsc{post}} hello\textsc{.m} do\textsc{.inf} \\ 
\glt `We would fold our arms across our chest [as a sign of respect and] greet them.'
\z 
 
\ea \label{RE.69}
\textit{tersēnmē weršāw} \\ 
\gll tersē=nmē wer=šā=ū \\ 
 be\_afraid\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} in\_front\_of\textsc{=3pl}=and \\ 
\glt `We were scared by them.'
\z 
 
\ea \label{RE.70}
\textit{zemānēč pāyīz āmāno kewčē xelemā bīyebo teġārē} \\ 
\gll zemān-ē=č pāyīz āmā=n=o kewč-ē xele=mā bīye b-o teġār-ē \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=add} autumn come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} six\_kilos\textsc{-indf} wheat\textsc{.m}\textsc{=\textsc{1pl}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} 120\_kilos\textsc{-indf} \\ 
\glt `We might have had a taghar\footnote{A unit of weight equivalent to 120 kilograms.} of wheat when it was autumn.'
\z 
 
\ea \label{ŽE.50}
\textit{ī doīnete dēnmēū ī šelemīnēte dēnmē} \\ 
\gll ī doīne=t=e dē=nmē=ū ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{\textsc{=ez.gen}} \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this dwina. \footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{ŽE.51}
\textit{ī šelemīnēte dēnmē} \\ 
\gll ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `you gave us this Shalamina. \footnote{Dish made of cracked wheat and turnip.}'
\z 
 
\ea \label{ŽE.52}
\textit{ī xeplēte dēnmē} \\ 
\gll ī xeplē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this xapla '
\z 
 
\ea \label{ŽE.53}
\textit{īnēmā wārdēne} \\ 
\gll īnē=mā wārdē=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=1pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We have eaten this [meal]. '
\z 
 
\ea \label{ŽE.76}
\textit{wāčē, weɫā bā kināčekē donmē’} \\ 
\gll wāč-ē weɫā bā kināčē-ekē ∅-don-mē \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} indeed \textsc{hort} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}talk\_to\textsc{-1pl} \\ 
\glt `She (the mother) would say, ‘Let us talk to the girl.’'
\z 
 
\ea \label{ŽE.83}
\textit{tā qewmekāmā wiš kermē qewmekā wiš kermē bizānmē ī qewmīmā řezāšāne yā řezāšā nīyene} \\ 
\gll tā qewm-ekā=mā wiš ∅-ker-mē qewm-ekā wiš ∅-ker-mē bi-zān-mē ī qewm-ī=mā řezā=šā=ne yā řezā=šā nīye=ne \\ 
 until relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=1pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} agreement\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} or agreement\textsc{.f}\textsc{=3pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We will inform our relatives. We will inform our relatives to see if they agree [to the  marriage] or not.'
\z 
 
\ea \label{ŽM.19}
\textit{ī menteqew ēme kemūkuřīn} \\ 
\gll ī menteqe=ū ēme kemūkořī=n \\ 
 \textsc{dem.prox} region\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Our region is in poverty.'
\z 
 
\ea \label{ŽM.21}
\textit{ī mentēqew ēme nebīyeyš hen nebīyey} \\ 
\gll ī mentēqe=ū ēme nebīyey=š hen-∅ nebīyey \\ 
 \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} poverty\textsc{.m} \\ 
\glt `Our region is in poverty, [and] shortage.'
\z 
 
\ea \label{ŽM.22}
\textit{ā wextī mišyē bilmē ġārijne čīw bārmēw čīw bermē} \\ 
\gll ā wext-ī mišyē bi-l-mē ġārij=ne čīw b-ār-mē=ū čīw ∅-ber-mē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} aux\textsc{.pst} \textsc{sbjv-}go\textsc{.prs}\textsc{-\textsc{1pl}} abroad\textsc{=\textsc{post}} thing\textsc{.m} \textsc{sbjv-}bring\textsc{.prs}\textsc{-\textsc{1pl}}=and thing\textsc{.m} \textsc{sbjv-}take\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt ` We had to go out [of Hawraman] to fetch and take stuff [to sell].'
\z 
 
\ea \label{ŽM.23}
\textit{luēnmē ferz tārān kārmā kerđen šiš māngē} \\ 
\gll luē=nmē ferz tārān kār=mā kerđe=n šiš mānge-ē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} guess\textsc{.m} \textsc{pn} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go, say, to Tehran and work for six months.'
\z 
 
\ea \label{ŽM.24}
\textit{luēnmē kirmān kārmā kerđen šiš māngē} \\ 
\gll luē=nmē kirmān kār=mā kerđe=n šiš mānge-ē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kerman and work for six months.'
\z 
 
\ea \label{ŽM.25}
\textit{jezīrew kīšīne kārmā kerđen šiš māngē} \\ 
\gll jezīre=w kīš-ī=ne kār=mā kerđe=n šiš mānge-ē \\ 
 island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kish Island and work for six months.'
\z 
 
\ea \label{ŽM.31}
\textit{dilē bedbextīne bīyēnmē} \\ 
\gll dilē bedbextī=ne bīyē=nmē \\ 
 inside hardship\textsc{.m}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `We experienced hardship.'
\z 
 
\ea \label{ŽM.32}
\textit{ā wextīyekey ečāge hīč xeberēmā nebīyenū telefūnīčū žimārew telefūnī nebēw} \\ 
\gll ā wext-ī-ekey e=čāge hīč xeber-ē=mā ne-bīye=n=ū telefūn=īč=ū žimāre=ū telefūn-ī ne-b-ē=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} in=there nothing news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and telephone\textsc{.m}\textsc{=add}=and number\textsc{.m}\textsc{=ez.gen} telephone\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and \\ 
\glt `I\footnote{Lit. We.}  had no news [of my family] from there. There was no phone or phone number at the time.'
\z 
 
\ea \label{ŽM.33}
\textit{šesū šiš sāʕetē ēme be māšīn jā luēnmē binderʕebāsū} \\ 
\gll šes=ū šiš sāʕet-ē ēme be māšīn jā luē=nmē binderʕebās=ū \\ 
 sixty\textsc{=ez.gen} six hour\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} by car\textsc{.m} afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and \\ 
\glt `It took us sixty-three hours to reach Bandar Abbas by car.'
\z 
 
\ea \label{ŽM.34}
\textit{luēnmē čārekū kerdenmā ew des} \\ 
\gll luē=nmē čārek=ū kerde=n=mā ew des \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{dem.dist} hand\textsc{.m} \\ 
\glt `We went to Charak and went to the other side.'
\z 
 
\ea \label{ŽM.35}
\textit{šiš māngē wēmā čā kārmā kerđen} \\ 
\gll šiš mānge-ē wē=mā čā kār=mā kerđe=n \\ 
 six month\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{reflx}\textsc{=\textsc{1pl}} there work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I worked there for [periods of] six months.'
\z 
 
\ea \label{ŽM.36}
\textit{ī šiš mānge xeberū hīčīmā nezānān} \\ 
\gll ī šiš mānge=e xeber=ū hīč-ī=mā ne-zānā=n \\ 
 \textsc{dem.prox} six month\textsc{.f}\textsc{=dem} news\textsc{.m}\textsc{=ez.gen} nothing\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I didn’t have any updates [from my family] during those periods of six months.'
\z 
 
\ea \label{ŽM.38}
\textit{extīyārmā mištū ā fārsāne bīyen jezīrew kīšī} \\ 
\gll extīyār=mā mišt=ū ā fārs-ā=ne bīye=n jezīre=ū kīš-ī \\ 
 volition\textsc{.m}\textsc{=\textsc{1pl}} fist\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} Persian\textsc{-3pl}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We were under the rule of Persians at Kish island.'
\z 
 
\ea \label{ŽM.39}
\textit{imrīkāī āmēnē ečā kārmā pey kerdēnē} \\ 
\gll imrīkāī āmē=nē e=čā kār=mā pey kerdē=nē \\ 
 \textsc{pn} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} in=there work\textsc{.m}\textsc{=\textsc{1pl}} to do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Americans came [there]. We had them work [for us].'
\z 
 
\ea \label{ŽM.40}
\textit{ēme čāgene sāxtmānē sāznēnmē} \\ 
\gll ēme čāgene sāxtmān-ē sāznē=nmē \\ 
 \textsc{1pl} there building\textsc{.m}\textsc{-dir}\textsc{.pl} build\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `We built buildings there.'
\z 
 
\ea \label{ŽM.45}
\textit{ēme jezīre čenū ā jezīrey čeneš bēnmē duē sāʕetē beyniš bē} \\ 
\gll ēme jezīre čenū ā jezīre-ī čene=š b-ēn-mē duē sāʕet-ē beyn=iš b-ē \\ 
 \textsc{1pl} island with\textsc{\textsc{.ez}.gen} \textsc{dem.dist} island\textsc{-obl}\textsc{.m} at\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1pl}} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} distance\textsc{.m}\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `It (Kuwait) was two hours away from the island we were in.'
\z 
 
\ea \label{ŽM.46}
\textit{řoē sī penj timenē kār kerēnmē} \\ 
\gll řo-ē sī penj timen-ē kār ker-ēn-mē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} task\textsc{.m} do\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1pl}} \\ 
\glt `I worked for a daily salary of thirty-five tomans. '
\z 
 
\ea \label{ŽM.50}
\textit{ēme čāne kārmā kerđenū} \\ 
\gll ēme č=ā=ne kār=mā kerđe=n=ū \\ 
 \textsc{1pl} in=\textsc{dem.dist}\textsc{=\textsc{post}} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `We worked there.'
\z 
 
\ea \label{ŽM.51}
\textit{wextē āmēnmēwe ī mentēqe ētir čē musteqerē bīyēnmē} \\ 
\gll wext-ē āmē=nmē=we ī mentēqe ētir čē musteqer-ē bīyē=nmē \\ 
 time\textsc{.m}\textsc{-indf} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \textsc{disc}.\textsc{ptcl} here settled\textsc{\textsc{-pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `When I returned to this region, I settled down here.'
\z 
 
\ea \label{ŽM.52}
\textit{luēnmē benenū āmēnmēwew} \\ 
\gll luē=nmē benen=ū āmē=nmē=we=ū \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl}=and \\ 
\glt `We would go to Banan (the summer habitat) and return (home).'
\z 
 
\ea \label{ŽM.53}
\textit{ʕerzū birāw wēm kerū āmēnmēwe ī mentēqe} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū āmē=nmē=we ī mentēqe \\ 
 petition\textsc{.m}\textsc{=ez.gen} brother\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \\ 
\glt `Let me tell you. I returned to this region.'
\z 
 
\ea \label{ŽM.54}
\textit{ētir īsey āmēnmē tā wē yāwēnmē be kemūkuřīū hewrāmānī} \\ 
\gll ētir īse-ī āmē=nmē tā wē yāwē=nmē be kemūkořī=ū hewrāmān-ī \\ 
 \textsc{disc}.\textsc{ptcl} now\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} until there arrive\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} by poverty\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have made it so far, [despite] poverty and destitution.'
\z 
 
\ea \label{ŽM.57}
\textit{ā wextīyekey ēme īne zinđigīš pene kermē be nebīyey} \\ 
\gll ā wext-ī-ekey ēme īne zinđigī=š pene ∅-ker-mē be nebīyey \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} life\textsc{.m}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1pl}} with poverty\textsc{.m} \\ 
\glt `Well, we live in poverty.'
\z 
 
\ea \label{ŽM.58}
\textit{be kemūkořī nebīyey zinđigīš pene kermē} \\ 
\gll be kemūkořī nebīyey zinđigī=š pene ∅-ker-mē \\ 
 by poverty\textsc{.m} poverty\textsc{.m} life\textsc{.m}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `We live in poverty and destitution,'
\z 
 
\ea \label{ŽM.59}
\textit{enāne ēme wārīyātēmā nīyā} \\ 
\gll enā=ne ēme wārīyāt-ē=mā nīye=ā \\ 
 otherwise\textsc{=\textsc{post}} \textsc{1pl} income\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `because we have no income.'
\z 
 
\ea \label{ŽM.60}
\textit{wārīyātmā nīyā kāsbīmā nīyā kārmā nīyā} \\ 
\gll wārīyāt=mā nīye=ā kāsbī=mā nīye=ā kār=mā nīye=ā \\ 
 income\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} business\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} work\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We have no income, no business, no jobs.'
\z 
 
