\chapter{All 254 examples of 1sg in these texts are found in the following sentences:}

\ea \label{ZB.37}
\textit{īne min īnem žinīyen lāw, fermāwištū šēxīn, šā ʕelādīnīwe} \\ 
\gll īne min īne=m žinīye=n lā=ū fermāwišt=ū šēx-ī=n šā ʕelādīn-ī=we \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{1sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} speech\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have heard this tale from —it is the Sheikh’s words— Shah Aladin.'
\z 
 
\ea \label{ZQ.1}
\textit{ʕerzū xizmetīt kerū} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū \\ 
 petition\textsc{.m}\textsc{=ez.gen} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `Let me tell you!\footnote{Lit. May a petition be for your service.}'
\z 
 
\ea \label{ZQ.2}
\textit{hetā min žinīyenim pīyewe āmā xizmetū šā ʕelādīnī} \\ 
\gll hetā min žinīye=n=im pīye-we āmā xizmet=ū šā ʕelādīn-ī \\ 
 even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} man\textsc{.m} come\textsc{.pst}\textsc{.3sg} service\textsc{.m}\textsc{=ez.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have even heard that a man came to the sevice of Shah Aladin. '
\z 
 
\ea \label{ZQ.7}
\textit{ʕerziš kerđ, qurbān bā dāstānēt pey gēɫnūwe, dāstāne} \\ 
\gll ʕerz=iš kerđ qurbān ∅-b-ā dāstāne-ē=t pey ∅-gēɫn-ū=we dāstāne \\ 
 petition\textsc{.m}\textsc{=3sg} do\textsc{.pst} sir.\textsc{voc} \textsc{ind-}be.prs\textsc{-3pl} story.f\textsc{-indf}\textsc{=\textsc{2sg}} to \textsc{sbjv-}lead.prs\textsc{-\textsc{1sg}}\textsc{=compl} story.f \\ 
\glt `He said, ‘Sir, Let me tell you a story, the story!’'
\z 
 
\ea \label{ZQ.13}
\textit{ženīem bē ženekēmīč bārū hemɫīe bē} \\ 
\gll ženīe=m b-ē ženī-ekē=m=īč bār=ū hemɫī-e b-ē \\ 
 woman.f\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs\textsc{.3sg} woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}}\textsc{=add} load\textsc{.m}\textsc{=ez.gen} pregnant\textsc{-f} \textsc{sbjv-}come.prs\textsc{.3sg} \\ 
\glt `I had a wife. My wife was pregnant.'
\z 
 
\ea \label{ZQ.14}
\textit{ženekēm qomyāš venī kelekewe} \\ 
\gll ženī-ekē=m qomyā=š venī kel-eke=we \\ 
 woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}} happen\textsc{.pst}\textsc{.3sg}\textsc{=3sg} at mountain\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `My wife was about to deliver a baby. '
\z 
 
\ea \label{ZQ.23}
\textit{yoīčšā her gīyānēš čenī bē gēɫānē qiřoɫe dārēm yoso} \\ 
\gll yo=īč=šā her gīyān-ē=š čenī b-ē gēɫā-ānē qiřoɫ=e dār-ē=m yos=o \\ 
 one\textsc{.m}\textsc{=add}\textsc{=3pl} just soul\textsc{.m}\textsc{-indf}\textsc{=3sg} in \textsc{sbjv-}come.prs\textsc{.3sg} turn\textsc{.pst}\textsc{-\textsc{1sg}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} find\textsc{.pst}\textsc{=compl} \\ 
\glt `As for the one which was half-alive, I wandered around [and] found a tree hollow.'
\z 
 
\ea \label{ZQ.24}
\textit{berdim nīyāmne qiřoɫe dāreke} \\ 
\gll berd-∅=im nīyā=m=ne qiřoɫ=e dār-eke \\ 
 take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1sg}} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `I took [him] and put him in the tree hollow.'
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{ZQ.26}
\textit{yošā berd nīyāmne qiřoɫe dāreke qurbān!} \\ 
\gll yo=šā berd-∅ nīyā=m=ne qiřoɫ=e dār-eke qurbān \\ 
 one\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{-3sg}\textsc{.m} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} sir.\textsc{voc} \\ 
\glt `Sir, I took one [and] put him in the tree hollow. '
\z 
 
\ea \label{ZQ.35}
\textit{luānē yāgekē vātim, be xwā īse serē midew bizānū} \\ 
\gll luā-ānē yāgē-ekē vāt=im be xwā īse ser-ē mi-de-ū bi-zān-ū \\ 
 go\textsc{.pst}\textsc{-\textsc{1sg}} place.f\textsc{-def}\textsc{.obl}.f say\textsc{.pst}\textsc{=\textsc{1sg}} by God\textsc{.m} now \textsc{clf}\textsc{-indf} \textsc{ind-}give.prs\textsc{-\textsc{1sg}} \textsc{sbjv-}know.prs\textsc{-\textsc{1sg}} \\ 
\glt `[When] we went to the place [of the other child], I said [to myself], ‘By God, I will have a look [and] see [what happens].’'
\z 
 
\ea \label{ZQ.36}
\textit{luānē nizīkū qiřoɫekey bīyāwe} \\ 
\gll luā-ānē nizīk=ū qiřoɫ-ekey bī-ā=we \\ 
 go\textsc{.pst}\textsc{-\textsc{1sg}} close=\textsc{ez.gen} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} be\textsc{.pst}\textsc{-3pl}\textsc{=compl} \\ 
\glt `I went [and] got closer to the tree hollow.'
\z 
 
\ea \label{ZQ.37}
\textit{miđyew sīwīnīš mē qiřoɫekey} \\ 
\gll mi-đye-ū sīwīnī=š m-ē qiřoɫ-ekey \\ 
 \textsc{ind-}look.prs\textsc{-\textsc{1sg}} voice\textsc{=3sg} \textsc{ind-}come.prs\textsc{.3sg} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I noticed a voice came from the hollow.'
\z 
 
\ea \label{ZQ.38}
\textit{nizīk bīyāwe temāšem kerd wextē miđyew kuřeke bīyen peřū qiřoɫēwe, qiřoɫekey} \\ 
\gll nizīk bī-ā=we temāše=m kerd wext-ē mi-đye-ū kuř-eke bīye=n peř=ū qiřoɫ-ēwe qiřoɫ-ekey \\ 
 close be\textsc{.pst}\textsc{-3pl}\textsc{=compl} looking\textsc{=\textsc{1sg}} do\textsc{.pst} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{=ez.gen} hollow\textsc{.m}\textsc{-indf} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I got closer and looked inside. I noticed that the boy [had grown up so much that he] had filled a tree hollow, i.e. the tree hollow. '
\z 
 
\ea \label{ZQ.39}
\textit{wextē luānē werew temāšem kerd miđyew be ʕemrew xwāy e čī qiřolī| ēnnew čīčew ženē pīfē peydā bīyenē| ʕeyn pīfē} \\ 
\gll wext-ē luā-ānē wer=ew temāše=m kerd mi-đye-ū be ʕemre=ū xwā-ī ečī qiřoɫ-ī ēnne=w čīče=ū ženī-ē pīfē-ē peydā bīyē=nē ʕeyn pīfē \\ 
 time\textsc{.m}\textsc{-indf} go\textsc{.pst}\textsc{-\textsc{1sg}} front\textsc{.m}\textsc{=\textsc{post}} looking\textsc{=\textsc{1sg}} do\textsc{.pst} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} by order.f\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} in.\textsc{dem.prox} hollow\textsc{.m}\textsc{-obl}\textsc{.m} as\_much\_as\textsc{=ez.gen} nipple\textsc{=ez.gen} woman.f\textsc{-obl}.f burr.f\textsc{-dir}\textsc{.pl} visible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like balloon.f \\ 
\glt `When I got closer and looked inside, I noticed that, by the order of God, some burrs the size of a woman's nipple had appeared in the tree hollow, like balloons.'
\z 
 
\ea \label{ZQ.45}
\textit{wextē miđyew kuř bīyen peřū qiřoɫē} \\ 
\gll wext-ē mi-đye-ū kuř bīye=n peř=ū qiřoɫ-ē \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} son/boy\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{=ez.gen} hollow\textsc{.m}\textsc{-indf} \\ 
\glt `When I looked, the boy had [grown up so much that he had] filled a tree hollow.'
\z 
 
\ea \label{ZQ.46}
\textit{wātim, beydē temāšew lutfū xwāy čenīnā’} \\ 
\gll wāt=im b-e-ydē temāše=ū lutf=ū xwā-ī čenīn=ā \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs-\textsc{2pl} looking\textsc{=ez.gen} grace\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} how\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I said [to the people], ‘Come and see, God’s grace is like this!'
\z 
 
\ea \label{ZQ.48}
\textit{wātim, beydē dilū āweđānīekey} \\ 
\gll wāt=im b-e-ydē dil=ū āweđānī-ekey \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs-\textsc{2pl} inside\textsc{=ez.gen} oasis\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I said, ‘Come to the meadow’'
\z 
 
\ea \label{ZQ.50}
\textit{wātim, beydē dilū āweđānīekey} \\ 
\gll wāt=im b-e-ydē dil=ū āweđānī-ekey \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs-\textsc{2pl} inside\textsc{=ez.gen} oasis\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I said, ‘By God, the son has grown up to the size of the tree hollow.’'
\z 
 
\ea \label{ZQ.52}
\textit{vātim, weɫā!} \\ 
\gll vāt=im weɫā \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} indeed \\ 
\glt `I said, ‘Indeed!’'
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{HB.10}
\textit{min bilā bēnēve, bilā bāršā pey bāro bāršā pey bero} \\ 
\gll min bi-l-ā b-ē-nē=ve bi-l-ā bār=šā pey b-ār-o bār=šā pey ∅-ber-o \\ 
 1sg \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} load\textsc{.m}\textsc{=3pl} for \textsc{sbjv-}bring\textsc{.prs}\textsc{-3sg} load\textsc{.m}\textsc{=3pl} for \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `[They had donkeys] to go [for loads] and return [home with them]; they would go [to the destination], put the loads on them, and take the loads with them.'
\z 
 
\ea \label{HB.19}
\textit{māčo, āy dāxekey dāxim bē min čen bē ʕeqiɫ bīyā!} \\ 
\gll m-āč-o āy dāx-eke=y dāx=im b-ē min čen bē-ʕeqiɫ bī-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} sorrow\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}.\textsc{attr} sorrow\textsc{=1sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{1sg} how\_much without-wisdom be\textsc{.pst}\textsc{-1sg} \\ 
\glt `He (the donkey breeder) said, ‘Oh, I wish I was dead! Look how stupid I was!’'
\z 
 
\ea \label{HB.20}
\textit{newātim qurbān min bilū yānew kē hārđī bārū} \\ 
\gll ne-wāt=im qurbān min bi-l-w yāne=ū kē hārđī b-ār-w \\ 
 \textsc{neg-}say\textsc{.pst}\textsc{=1sg} sir.\textsc{voc} 1sg \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} house=and who flour\textsc{\textsc{.f}}\textsc{.obl} \textsc{sbjv-}bring\textsc{.prs}\textsc{-1sg} \\ 
\glt `I didn’t say, ‘Sir, to whose house should I go [and] bring flour!'
\z 
 
\ea \label{HB.21}
\textit{min bilū āsāwū kē hārdī bārū} \\ 
\gll min bi-l-w āsāw=ū kē hārdī b-ār=ū \\ 
 1sg \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} mill\textsc{.m}\textsc{=ez}\textsc{.gen} who flour\textsc{\textsc{.f}} \textsc{sbjv-}bring\textsc{.prs}\textsc{=ez}\textsc{.gen} \\ 
\glt `To whose mill I should go and bring flour?'
\z 
 
\ea \label{HB.22}
\textit{zeřīčiš neyānā bilū yāgē bisānū} \\ 
\gll zeř=yč=iš ne-yā=nā bi-l-w yāgē-ē bi-sān-w \\ 
 money\textsc{.m}\textsc{=add}\textsc{=3sg} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} place\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}buy\textsc{.prs}\textsc{-1sg} \\ 
\glt `He hasn’t given me money either so that I could go to a place [and] buy [flour]'
\z 
 
\ea \label{HB.23}
\textit{ī gīre čēš bī min wārd?’} \\ 
\gll ī gīr=e čēš bī-∅ min wārd \\ 
 \textsc{dem.prox} hook\textsc{.m}\textsc{=dem} what be\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{1sg} eat\textsc{.pst} \\ 
\glt `What is this situation that I am caught in?’'
\z 
 
\ea \label{HB.36}
\textit{simerī kerā tūrekekew min} \\ 
\gll simerī ∅-ker-ā tūreke-eke=w min \\ 
 straw\textsc{\textsc{.f}} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} sack\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg \\ 
\glt `They (the family) will put [some] straw into my sack;'
\z 
 
\ea \label{HB.43}
\textit{pekemā megino, to goš min dāre!’} \\ 
\gll peke=mā me-gin-o to goš min ∅-dār-e \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \textsc{2sg} ear\textsc{.m} 1sg \textsc{imp-}hold\textsc{.prs}-\textsc{2sg} \\ 
\glt `We should not be worried; listen to me!’'
\z 
 
\ea \label{HB.51}
\textit{wātiš, fermāwdē bēydē mēmānū minindē} \\ 
\gll wāt=iš ∅-fermāw-dē b-ē-ydē mēmān=ū min=indē \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{imp-}say\textsc{.prs}\textsc{-2pl} \textsc{imp-}come\textsc{.prs}\textsc{-2pl} guest\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg\textsc{=cop}\textsc{.2pl} \\ 
\glt `He said, ‘Please come! You are my guests.'
\z 
 
\ea \label{HB.52}
\textit{mēmānū minindē'} \\ 
\gll mēmān=ū min=indē \\ 
 guest\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg\textsc{=cop}\textsc{.2pl} \\ 
\glt `You are my guests.’'
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{HB.68}
\textit{milūwe, min činnē sāɫēn īnānē xizmetū šēxīne} \\ 
\gll mi-l-w=we min činnē sāɫe-ē=n īnā-ānē xizmet=ū šēx-ī=ne \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{compl}} 1sg some\textsc{.pl} year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-1sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I will go back! It’s been several years that I am at the service of the sheikh;'
\z 
 
\ea \label{HB.69}
\textit{herbeneš pey kerūnew renjberiš bīyenā} \\ 
\gll herbene=š pey ∅-ker-ūne=ū renjber=iš bīye=nā \\ 
 donkey breeder\textsc{.m}\textsc{=3sg} for \textsc{ind-}do\textsc{.prs}\textsc{-1sg}\textsc{=ez}\textsc{.gen} worker\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \\ 
\glt `I take care of the donkeys for him; I have been his worker.'
\z 
 
\ea \label{HB.70}
\textit{ēnnew herēm nezānān} \\ 
\gll ēnne=w her-ē=m ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Wasn’t I as much of a confidante as a donkey?' \footnote{Lit. didn’t I know as much as a donkey!}
\z 
 
\ea \label{HB.71}
\textit{ī herī zūwāniš zānā min hīčim nezānān!} \\ 
\gll ī her-ī zūwān=iš zānā min hīč=im ne-zānā=n \\ 
 \textsc{dem.prox} donkey\textsc{.m}\textsc{-obl}\textsc{.m} language\textsc{.m}\textsc{=3sg} know\textsc{.pst}\textsc{.ptcp}\textsc{.m} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The donkey knew the [Sheikh’s] language; I didn’t know a thing!'
\z 
 
\ea \label{HB.80}
\textit{čiřoš hezretū šēxī fermāwo, erē řoɫem čī toryānī?} \\ 
\gll ∅-čiř-o=š hezret=ū šēx=ī ∅-fermāw-o erē řoɫe=m čī toryā=nī? \\ 
 \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} child\textsc{.m}\textsc{=1sg} why get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt ` His highness the Sheikh summoned him, i.e. the donkey breeder, [and] said, ‘Oh son! why were you offended? '
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{HB.83}
\textit{ēnnew herē qābilīm nebē hāɫīm kerī!} \\ 
\gll ēnne=w her-ē qābilī=m ne-b-ē hāɫī=m ∅-ker-ī \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} deserved\textsc{=1sg} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} understood\textsc{=1sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-obl}\textsc{.m} \\ 
\glt `Wasn’t I as trustworthy as a donkey that you didn’t make me understand!'
\z 
 
\ea \label{HB.84}
\textit{ēnnew herē, īse her be herī wēš qisē kero min hīčim nezānān!’} \\ 
\gll ēnne=w her-ē īse her be her-ī wē=š qisē-ē ∅-ker-o min hīč=im ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} now both by donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As much as a donkey! now the donkey could speak even though it was a donkey, but I didn’t know a thing.'
\z 
 
\ea \label{HB.92}
\textit{čīm hāɫī nekerdenī!} \\ 
\gll čī=m hāɫī ne-kerde=nī \\ 
 why\textsc{=1sg} understood \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Why haven’t I made you understand?'
\z 
 
\ea \label{HB.93}
\textit{hāɫīm nekerdenī řās, ānīšā} \\ 
\gll hāɫī=m ne-kerde=nī řās ānīšā \\ 
 understood\textsc{=1sg} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} really 3pl.dem\textsc{.obl} \\ 
\glt `Haven’t I really made you understand? and so on.'
\z 


\ea \label{PM.1}
\textit{ʕerzū xizmetīt kerū hetā min žinīyenim pāse mešhūr bīyen šewē šā ʕelādīnī, šewē peɫē merekuř peɫē merekuř mē šārezūr} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū hetā min žinīye=n=im pāse mešhūr bīye=n šewe-ē šā ʕelādīn-ī šewe-ē peɫ-ē merekuř peɫ-ē merekuř m-ē šārezūr \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} like famous be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} night.f\textsc{-obl}.f king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} night.f\textsc{-obl}.f \textsc{clf}\textsc{-indf} grasshopper \textsc{clf}\textsc{-indf} grasshopper \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{pn} \\ 
\glt `Let me tell you .\footnote{Lit. May a petition be for your service.} I have even heard (this). It is well known that a swarm of grasshoppers came to Sharazour one night.'
\z 
 
\ea \label{PM.13}
\textit{kīyāsenāšā xizmetū to} \\ 
\gll kīyāse=nā=šā xizmet=ū to \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=3pl} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} \\ 
\glt `I have been sent to your service.'
\z 
 
\ea \label{PM.36}
\textit{min pewkī ījāzem dān hinekew ēme wero, bilo zerer je kesī negino} \\ 
\gll min pewkī ījāze=m dā=n hin-eke=ū ēme ∅-wer-o bi-l-o zerer je kes-ī ne-gin-o \\ 
 \textsc{1sg} that\_is\_why permission\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} ez.pron\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{sbjv-}eat\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} harm\textsc{.m} from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `I let them [eat our crop] so that they would eat our crop and go away, [but] not do harm to others.'
\z 
 
\ea \label{PM.45}
\textit{min īse ī nīštenāre īsrāhetīč kerū} \\ 
\gll min īse ī nīšte=nā=re īsrāhet=īč ∅-ker-ū \\ 
 \textsc{1sg} now \textsc{dem.prox} sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=\textsc{povb}} rest\textsc{.m}\textsc{=add} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Now I am seated and I am resting.'
\z 
 
\ea \label{PM.46}
\textit{čowe ew mē peym māro} \\ 
\gll č=o=we ew m-ē pey=m m-ār-o \\ 
 in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From that direction, he  comes [to me] and brings [food] for me'
\z 
 
\ea \label{PM.47}
\textit{ēđ čowe mē peym māro} \\ 
\gll ēđ č=o=we m-ē pey=m m-ār-o \\ 
 3sg.prox\textsc{.dir} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From this direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.48}
\textit{ew čowe mē peym māro} \\ 
\gll ew č=o=we m-ē pey=m m-ār-o \\ 
 \textsc{dem.dist} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From that direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.49}
\textit{fermāwo dey ētir min čī nāřihet bīyā!} \\ 
\gll ∅-fermāw-o dey ētir min čī nāřihet bī-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{disc.ptcl} \textsc{1sg} why sad become\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `So, why did I get worried at all?'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DG.38}
\textit{ā wextīyekey, ētir, īse, min yerē sāɫē ī birāme tūšū ī derdēe āmānū} \\ 
\gll ā wext-ī-ekey ētir īse min yerē sāɫe-ē ī birā=m=e tūš=ū ī derde-ē=e āmā=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} encounter\textsc{=ez}\textsc{.gen} \textsc{dem.prox} illness\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=dem} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Now, as for me, it’s been three years that my brother is suffering from this illness. '
\z 
 
\ea \label{DG.39}
\textit{nimāzā bilū āweyānī} \\ 
\gll ni-m-āz-ā bi-l-ū āweyānī \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} habitat\textsc{\textsc{.f}} \\ 
\glt `I am not allowed to go to the village.'
\z 
 
\ea \label{DG.40}
\textit{šewēne, řone milū nān peyā kerūnew} \\ 
\gll šewe-ē=ne řo=ne mi-l-ū nān peyā ∅-ker-ūne=ū \\ 
 night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} day\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} bread\textsc{.m} visible \textsc{ind-}do\textsc{.prs}\textsc{-1sg}=and \\ 
\glt `I go [and] find bread during the day.'
\z 
 
\ea \label{DG.41}
\textit{šewēne milūwe lāš} \\ 
\gll šewe-ē=ne mi-l-ū=we lā=š \\ 
 night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}go\textsc{.prs}\textsc{-1sg}\textsc{=compl} at\_the\_place\_of\textsc{=3sg} \\ 
\glt `I go back to him at night.'
\z 
 
\ea \label{DG.43}
\textit{ī birāme derdeš, derdēš īnā po derdū gulīš māčā dāđē!} \\ 
\gll ī birā=m=e derde=š derde-ē=š īnā-∅ p==oderde=ū gulī=š m-āč-ā dāđ-ē \\ 
 \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} illness\textsc{\textsc{.f}}\textsc{=3sg} illness\textsc{\textsc{.f}}\textsc{-indf}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at=\textsc{=\textsc{post}}illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} yell\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `This brother of mine is ill; it is called leprosy; Please help!'
\z 
 
\ea \label{DG.45}
\textit{lue xāneqā tā xeberit dewwe pene’} \\ 
\gll ∅-lu-e xāneqā tā xeber=it ∅-de-ū=we pene \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} monastery\textsc{.m} until news\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}give\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{compl}} to \\ 
\glt `Go to the monastery until I let you know.'
\z 
 
\ea \label{DG.59}
\textit{gāhes pī qisē mine xwā řehm kero, weš bo} \\ 
\gll gāhes p=ī qisē-ē min=e xwā řehm ∅-ker-o weš ∅-b-o \\ 
 maybe by\textsc{=dem}.prox talk\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{1sg}\textsc{=dem} God\textsc{.m} mercy\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} well \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Through my intervention may God have mercy on him so that he gets well.'
\z 
 
\ea \label{DG.64}
\textit{ā wextīekey beɫē, min bilo ā memleketū wēšā, kiseɫīš čenē, ēme zāmdārē nekero ēme kečē nekero} \\ 
\gll ā wext-ī-ekey beɫē min bi-l-o ā memleket=ū wē=šā kiseɫī=š čenē ēme zāmdār-ē ne-ker-o ēme keč-ē ne-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yes \textsc{1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3pl} tortoise\textsc{\textsc{.f}}\textsc{=3sg} in \textsc{1pl} wounded\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `[The Sheikh said], ‘Yes!' '[The tortoise said] I [say] he [should] go to his region; there are tortoises there. He shall not injure us; he shall not make us crooked. '
\z 
 
\ea \label{DG.67}
\textit{šā ʕelādīnīč fermāwo birākem wextē īnē zānā min gunākē māy mē milimre} \\ 
\gll šā ʕelādīn-ī=č ∅-fermāw-o birā-eke=m wext-ē īnē ∅-zān-ā min gunā-ekē m-āy m-ē mil=im=re \\ 
 king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} time\textsc{.m}\textsc{-indf} \textsc{dem.prox}\textsc{.3pl} \textsc{ind-}know\textsc{.prs}\textsc{-3pl} \textsc{1sg} sin\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}come\textsc{.pst}\textsc{.adv} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} shoulder\textsc{.m}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `Shah Aladin said, ‘Brother!—as they (i.e. the tortoises) know [about the medicine] I won’t shoulder the burden of injuring them.'
\z 
 
\ea \label{DP.20}
\textit{wāteniš, min zīndigīū bābiɫ bābīm e čē bīyen} \\ 
\gll wāte=n=iš min zīndigī=ū bāb-iɫ bāb-ī=m e=čē bīye=n \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} life\textsc{.m}\textsc{=ez}\textsc{.gen} ancestor\textsc{.m}-of ancestor\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg} in=here be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` saying, ‘I, my ancestors, have all lived here.'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.23}
\textit{min īney dew deso towe ey wēm kowe bilū} \\ 
\gll min īney ∅-de-ū des=o to=we ey wē=m ko=we bi-l-w \\ 
 \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}give\textsc{.prs}\textsc{-1sg} hand\textsc{.m}\textsc{=\textsc{post}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{intj} \textsc{reflx}\textsc{=1sg} where\textsc{=\textsc{post}} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} \\ 
\glt `If I hand over these to you, where should I go myself?'
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.40}
\textit{ewīč nimē māčo, nexeyr nimewnew mitmānim nīyā penet.} \\ 
\gll ew=yč ni-mi-ē mi-āč-o nexeyr ni-mi-e-wne=ū mitmān=im nīye=ā pene=t \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-1sg}=and trust\textsc{.m}\textsc{=1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} to\textsc{=\textsc{2sg}} \\ 
\glt ` Sibhan Agha\footnote{Lit. He.} did not go [to Jamsher Shah], saying, ‘No, I won’t come [to you]; I don’t trust you.’'
\z 
 
\ea \label{DP.41}
\textit{desgīrim kero āđ nūrīš firen} \\ 
\gll desgīr=im ∅-ker-o āđ nūrī=š fire=n \\ 
 hand.cufffed\textsc{=1sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m} force\textsc{.m}\textsc{=3sg} many\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is going to arrest me. He has many men.'
\z 
 
\ea \label{DP.45}
\textit{āđ kārū wēšū min hinū wēm} \\ 
\gll āđ kār=ū wē=š=ū min hin=ū wē=m \\ 
 3sg\textsc{.dir}\textsc{.m} task\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3sg}\textsc{=and} \textsc{1sg} \textsc{z.pron}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} \\ 
\glt `Then, he goes his way, and I go mine.'
\z 
\ea \label{DP.48}
\textit{māčo, dey min īney pīneym wāten} \\ 
\gll mi-āč-o dey min īney p=īney=m wāte=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=1sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` Jamsher Shah said [to Sibhan Agha], ‘When I swore an oath, I swore on this sparrow;'
\z 
 
\ea \label{DP.49}
\textit{ī gīyānū qefesew sīney minne īnen āđe lūā} \\ 
\gll ī gīyān=ū qefese=ū sīne=y min=ne īne=n āđe lūā \\ 
 \textsc{dem.prox} soul\textsc{.m}\textsc{=ez}\textsc{.gen} cage\textsc{.m}=and chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{\textsc{.f}} go\textsc{.pst} \\ 
\glt ` The soul in my rib cage [that I swore an oath on] was this [sparrow]. It [just] flew away.'
\z 
 
\ea \label{DP.50}
\textit{ānen mirīčēɫēm wer dān, to xeyrū min siɫāmet} \\ 
\gll āne=n mirīčēɫe-ē=m wer dā=n to xeyr=ū min siɫāmet \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} small\_bird\textsc{.m}\textsc{-indf}\textsc{=1sg} out give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `’I have let the bird go away. Now, you follow your life, and I will follow mine!\footnote{Lit. Well-being for you and health for me.}’'
\z 
 
\ea \label{DP.51}
\textit{dā dey dewršā gērdē bā deġāɫet kerū} \\ 
\gll dā dey dewr=šā ∅-gēr-dē bā deġāɫet ∅-ker-w \\ 
 \textsc{hort} \textsc{disc.ptcl} round\textsc{=3pl} \textsc{imp-}take\textsc{.prs}\textsc{-2pl} \textsc{hort} intervention\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt ` Jamsher Shah\footnote{Lit. He.} said [to his men], ‘Encircle them so that I can defeat [Sibhan Agha].'
\z 
 
\ea \label{DP.55}
\textit{to xeyrū min siɫāmet} \\ 
\gll to xeyr=ū min siɫāmet \\ 
 \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `Follow your own life, and I will follow mine.'
\z 
 
\ea \label{ZP.19}
\textit{yānē yāwān pene tāwū hem šerīʕetū hem terīqet tāwū girđ kārēwe kero} \\ 
\gll yānē yāwā=n pene ∅-tāw-ū hem šerīʕet=ū hem terīqet ∅-tāw-ū girđ kār-ē=we ∅-ker-o \\ 
 that\_means arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} both sharia\_law\textsc{\textsc{=ez.gen}} both denomination \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} every task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{post}} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This meant that he had become experienced (in his field). He could practise Sharia law, lead an order (of dervishes), and carry out all tasks [appointed to him]'
\z 
 
\ea \label{ZP.32}
\textit{šewē āmān werm be min wāteniš,} \\ 
\gll šewe-ē āmā=n werm be min wāte=n=iš \\ 
 night\textsc{.f}\textsc{-obl}\textsc{.f} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} sleep to \textsc{1sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \\ 
\glt `[The king said], ‘At night, he came to my dream [and] told me,'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.48}
\textit{ī kināčē nārīwe ētir pey min her ēnne weše bowe’} \\ 
\gll ī kināčē ne-ār-ī=we ētir pey min her ēnne weš-e b-o=we \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} any\_more for \textsc{1sg} emph as\_much\_as well\textsc{-f} be\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `Please do not bring this daughter [of mine] back to me as long as she is cured.'
\z 
 
\ea \label{ZP.52}
\textit{pāse māčāš pene tewenū sirewē kināčekē mārū serū tewenekē} \\ 
\gll pāse m-āč-ā=š pene tewenī=ū sirewē kināčē-ekē m-ār-ū ser=ū tewenī-ekē \\ 
 like \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-\textsc{1sg}} on\textsc{\textsc{=ez.gen}} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `It is called the healing stone. He (the uncle) took the girl to the top of the stone.'
\z 
 
\ea \label{ZP.89}
\textit{kināčēw pādšāw fiɫān henānew miniš kerdenā be wekēɫ} \\ 
\gll kināčē=ū pādšā=w fiɫān hen-ā=ne=ū min=iš kerde=nā be wekēɫ \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such place\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \textsc{1sg}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to ad\textsc{voc}ate\textsc{.m} \\ 
\glt `She is the daughter of … the king of such-and-such place has put me in charge.\footnote{Lit. made me advocate.}'
\z 
 
\ea \label{ZP.90}
\textit{mišo māreš biřū pey ī šexsīw ī šexse eger ī kināčēše weše kerde,} \\ 
\gll mišo māre=š ∅-biř-ū pey ī šexs-ī=ū ī šexs=e eger ī kināčē=š=e weš-e kerd-e \\ 
 aux marriage\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for dem.prox saint\textsc{-obl}\textsc{.m}=and dem.prox saint\textsc{=dem} if dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f} \\ 
\glt `I shall marry her to this person [on the condition that] this person has healed this girl.'
\z 
 
\ea \label{ZP.98}
\textit{māčo, min īneyā jeryānim e čīnenū} \\ 
\gll m-āč-o min īne=ā jeryān=im e=čīne=n=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} story\textsc{.m}\textsc{=\textsc{1sg}} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He said, ‘I— look, my story is like this.'
\z 
 
\ea \label{ZP.99}
\textit{pādšāy kerdenā wekēɫū} \\ 
\gll pādšā-ī kerde=nā wekēɫ=ū \\ 
 king\textsc{.m}\textsc{-obl}\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m}=and \\ 
\glt `The king has put me in charge.'
\z 
 
\ea \label{ZP.100}
\textit{ī kināčē eger weše bīyowe mišyo māreš biřū pey to be mezebū wētāw} \\ 
\gll ī kināčē eger weš-e bīyē-o=we mišyo māre=š ∅-biř-ū pey to be mezeb=ū wē=tā=ū \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} if well\textsc{-f} be\textsc{.prs}.\textsc{cond}\textsc{.f}?\textsc{-3sg}\textsc{=compl} aux marraige\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for \textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `If the girl is healed, I shall marry her to you, according to your religion.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.102}
\textit{īne pādšāzāyēnew derdū min meweronew} \\ 
\gll īne pādšāzāyē=ne=ū derd=ū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}=and illness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She is no use to me.'
\z 
 
\ea \label{ZP.103}
\textit{min metāwū īne nigebānīye kerū} \\ 
\gll min me-tāw-ū īne nigebānī-e ∅-ker-ū \\ 
 \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} guardian\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt ` I cannot take care of her.'
\z 
 
\ea \label{ZP.104}
\textit{īne pādšāzāyēne čenū min meweronew} \\ 
\gll īne pādšāzāyē=ne čenū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}} with\textsc{.ez.gen} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She will not eat with me [under the same roof].'
\z 
 
\ea \label{ZP.105}
\textit{min čīwēwim nīyārē lāyiqū ī pādšāzāye bo} \\ 
\gll min čīw-ēw=im nīye=ā=rē lāyiq=ū ī pādšāzāye ∅-b-o \\ 
 \textsc{1sg} thing\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} deserved\textsc{\textsc{=ez.gen}} dem.prox princess\textsc{.obl}\textsc{.f} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `I have nothing which would be up to the standards of this princess.'
\z 
 
\ea \label{ZP.107}
\textit{ewīč māčo, nexeyre lāloekeš min tāze pādšā kerdenā wekēɫū} \\ 
\gll ew=īč m-āč-o nexeyre lālo-eke=š min tāze pādšā kerde=nā wekēɫ=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{1sg} anyway king\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m}=and \\ 
\glt ` The uncle said, ‘No! I—the king has put me in charge.'
\z 
 
\ea \label{ZP.108}
\textit{wātenīčiš, mišyo neberūšo”} \\ 
\gll wāte=n=īč=iš mišyo ne-ber-ū=š=o \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=add}\textsc{=3sg} aux \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `He has said, “I shall not take her back.”'
\z 
 
\ea \label{ZP.109}
\textit{ētir min metāwū berūšo, hergīz meberūšo’} \\ 
\gll ētir min me-tāw-ū ∅-ber-ū=š=o hergīz me-ber-ū=š=o \\ 
 \textsc{disc.ptcl} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} never \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `I cannot take her back. I will never take her back.’'
\z 
 
\ea \label{ZP.116}
\textit{māčo, meberūšo mišyo to māreš kerīnew”} \\ 
\gll m-āč-o me-ber-ū=š=o mišyo to māre=š ∅-ker-īne=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} aux \textsc{2sg} marraige\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs} \textsc{2sg}=and \\ 
\glt `He says, “I will not take her back; you shall marry her.”'
\z 
 
\ea \label{ZP.117}
\textit{īđīč hīč minīč hīčim nīyārēw} \\ 
\gll īđ=īč hīč min=īč hīč=im nīye=ā=rē=ū \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} nothing \textsc{1sg}\textsc{=add} nothing\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `She [has] nothing, nor do I.'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.119}
\textit{yoīč māčo, dey minīč beʕzē pīyāzīt midonew’} \\ 
\gll yo=īč m-āč-o dey min=īč beʕzē pīyāz-ī=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg}\textsc{=add} some onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Another said, ‘I will give you some onions.’'
\z 
 
\ea \label{ZP.120}
\textit{yoīč māčo, minīč beʕzē tiršū henārāt midonew’} \\ 
\gll yo=īč m-āč-o min=īč beʕzē tirš=ū henār-ā=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} some molasses\textsc{\textsc{=ez.gen}} pomegranate\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt ` Another said, ‘I will give you some pomegranate molasses.’'
\z 
 
\ea \label{ZP.121}
\textit{kābrāyēč māčo, dey birā, minīč serē heywānit midew} \\ 
\gll kābrā-ē=č m-āč-o dey birā min=īč ser-ē heywān=it mi-de-ū \\ 
 man\textsc{-indf}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{1sg}\textsc{=add} \textsc{clf}\textsc{-indf} animal\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `A man said, ‘Brother, I will give you an animal.’'
\z 
 
\ea \label{ZP.128}
\textit{weɫāhī īne šime māčdēšū īnā vistāwe řā pey min čūn šime be xātirū xwāytā kerd hetā xwā dinyāš weše kerdēne ī merāsēme her bo her meno’} \\ 
\gll weɫāhī īne šime m-āč-dē=š=ū īnā-∅ vist=tā=we řā pey min čūn šime be xātir=ū xwā-ī=tā kerd hetā xwā dinyā=š weš-e kerdē=ne ī merāsēm=e her ∅-b-o her ∅-men-o \\ 
 indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} \textsc{ind-}say\textsc{.prs}-\textsc{2pl}\textsc{=3sg}=and deic\textsc{-3sg}\textsc{.m} throw\textsc{.pst}=\textsc{2pl}\textsc{=compl} road\textsc{.f} for \textsc{1sg} because \textsc{2pl} to sake\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}be\textsc{.prs}\textsc{-3sg} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `Indeed, what you are saying and the way you organised [this ceremony] for me, since you did it in good faith, so long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ŽP.28}
\textit{ētir xo āđ meselen hetīm bo, lāw māmoyšā bo řūeš mebo wāčo, yā metāwū yā tāwū} \\ 
\gll ētir xo āđ meselen hetīm ∅-b-o lā=ū māmo-ī=šā ∅-b-o řūe=š me-b-o ∅-wāč-o yā me-tāw-ū yā ∅-tāw-ū \\ 
 \textsc{disc.ptcl} dis.ptc 3sg\textsc{.dir}\textsc{.m} for\_example orphan \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} face\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-3sg} either \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} either \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Well, he was an orphan. He was at his uncle’s. He was ashamed to say, ‘I cannot, or I can.’'
\z 
 
\ea \label{ŽP.51}
\textit{milo werāweriš miđyo e pāse min wāt— tomekeš šānā zemīnekeyneū} \\ 
\gll mi-l-o werāwer=iš mi-đye-o e=pāse min wāt tom-eke=š šānā zemīn-ekey=ne=ū \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} in\_front\_of\textsc{=3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} in=like \textsc{1sg} say\textsc{.pst} seed\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} scatter\textsc{.pst} land\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt ` He went to him [and] saw that —as I said— he (Pir Shaliyar) scattered the seeds in the field'
\z 
 
\ea \label{ŽP.57}
\textit{pīyāke mēwe ʕerzū heme yosoy kero māčo, weɫāhī qurbān min tūšū jeryānē e čīney āmānē} \\ 
\gll pīyā-eke m-ē=we ʕerz=ū heme yoso-ī ∅-ker-o m-āč-o weɫāhī qurbān min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sir.\textsc{voc} \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `The man returned [and said] to Hama Yoso, ‘Indeed, Sir! I have witnessed an event like this.'
\z 
 
\ea \label{ŽP.58}
\textit{min tūšū jeryānē e čīney āmānē} \\ 
\gll min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `I have witnessed an event like this.’'
\z 
 
\ea \label{ŽP.74}
\textit{māčo min īnem dī be sey misefāywe} \\ 
\gll m-āč-o min īne=m dī-∅ be sey misefā-ī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} see\textsc{.pst}\textsc{-3sg}\textsc{.m} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘I witnessed this about Say Mustafa.’'
\z 
 
\ea \label{ŽP.76}
\textit{čemim pene kotū řāsāw} \\ 
\gll čem=im pene kot=ū řās=ā=ū \\ 
 eye\textsc{=\textsc{1sg}} to fall\textsc{.pst}\textsc{.3sg}\textsc{.m}=and true\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `I caught sight of him. It’s true [that he has supernatural powers]'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.104}
\textit{zeřekey bere min, qey mekero} \\ 
\gll zeř-ekey ∅-ber-e min qey me-ker-o \\ 
 money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{1sg} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take the money. I will be fine.’'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.136}
\textit{fermāwo, min her noʕē āmānā āmānāw īnānē čēgew’} \\ 
\gll ∅-fermāw-o min her noʕ-ē āmā=nā āmā=nā=ū īnā-ānē čēge=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} emph kind/way\textsc{-indf} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}=and deic\textsc{-\textsc{1sg}} here=and \\ 
\glt `He said, ‘Well, I have come here anyway. I am here.’'
\z 
 
\ea \label{ŽP.138}
\textit{māčoš lāwe māčo min pey īney āmānā āmānāw} \\ 
\gll m-āč-o=š lā=we m-āč-o min pey īney āmā=nā āmā=nā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} to dem.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}=and \\ 
\glt `He said to him, ‘I have come for this [reason]’'
\z 
 
\ea \label{ŽP.160}
\textit{min wēm pādšānā} \\ 
\gll min wē=m pādšā=nā \\ 
 \textsc{1sg} \textsc{reflx}\textsc{=\textsc{1sg}} king\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I am the king.'
\z 
 
\ea \label{ŽP.161}
\textit{metāwū serū textīwe wezūrew bilū ginū gelū kināčēw be peyū luāy řāreū degā be degā} \\ 
\gll me-tāw-ū ser=ū text-ī=we wez-ū=re=ū bi-l-ū ∅-gin-ū gel=ū kināčē=ū be pe-ī=ū luāy řā=re=ū degā be degā \\ 
 \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} on\textsc{\textsc{=ez.gen}} throne\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} climb\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{povb}}=and \textsc{sbjv-}go\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-\textsc{1sg}} with\textsc{\textsc{=ez.gen}} daughter\textsc{.dir}\textsc{.pl}=and by foot\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} go.inf road\textsc{.f}\textsc{=\textsc{post}}=and village\textsc{.obl}\textsc{.f} by village\textsc{.obl}\textsc{.f} \\ 
\glt `I cannot leave the throne [and] go [and] accompany the girl, [while] walking on foot on the roads [that go] village by village.'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.198}
\textit{fermāwo, weɫā čūn ānā bēheyānī kerūt penū ʕāɫemī} \\ 
\gll ∅-fermāw-o weɫā čūn āne=ā bēheyā=nī ∅-ker-ū=t pen=ū ʕāɫem-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed as \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} shameless\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{2sg}} trouble\textsc{.m}\textsc{\textsc{=ez.gen}} world\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) said, ‘Indeed, since you are so indecent, I will disgrace you.’'
\z 
 
\ea \label{ŽP.199}
\textit{čūn bēheyānī āškirāt kerū} \\ 
\gll čūn bēheyā=nī āškirā=t ∅-ker-ū \\ 
 because shameless\textsc{=cop}\textsc{.\textsc{2sg}} visible\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Since you are indecent, I will reveal you.'
\z 
 
\ea \label{ŽP.206}
\textit{ā wextī lālo kināčē māčo, min, tāze pādšāy kerdenā wekēɫ} \\ 
\gll ā wext-ī lālo kināčē m-āč-o min tāze pādšā-ī kerde=nā wekēɫ \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} anyway king\textsc{.m}\textsc{-obl}\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m} \\ 
\glt `Then, the girl’s uncle said, ‘As for me, well, the king has put me in charge.\footnote{Lit. the king has made me advocate.}'
\z 
 
\ea \label{ŽP.207}
\textit{mišyo ī kināčē nīkāh kerū pey toū be mezebū wētāū} \\ 
\gll mišyo ī kināčē nīkāh ∅-ker-ū pey to=ū be mezeb=ū wē=tā=ū \\ 
 aux dem.prox girl\textsc{.dir}\textsc{.f} marriage\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} to \textsc{2sg}=and by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{ŽP.208}
\textit{xo min meberūšo} \\ 
\gll xo min me-ber-ū=š=o \\ 
 dis.ptc \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `I will not take her back.'
\z 
 
\ea \label{ŽP.209}
\textit{pāđšāy desūriš dān be minū} \\ 
\gll pāđšā-ī desūr=iš dā=n be min=ū \\ 
 king\textsc{.m}\textsc{-obl}\textsc{.m} task/order\textsc{.m}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by \textsc{1sg}=and \\ 
\glt `The king has ordered me.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.213}
\textit{īnē be min mewero kināčē pādšāy,} \\ 
\gll īnē be min me-wer-o kināčē pādšā-ī \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} with \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg} girl\textsc{.dir}\textsc{.f} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `She will not eat with me [under the same roof], the king’s daughter.'
\z 
 
\ea \label{ŽP.214}
\textit{min āne nānmā xeplezeřātīnen} \\ 
\gll min āne nān=m=ā xeplezeřātīne=n \\ 
 \textsc{1sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Look, my meal is xaplazaratine.’'
\z 
 
\ea \label{ŽP.217}
\textit{fermāwo, min īne nānmārēū} \\ 
\gll ∅-fermāw-o min īne nān=m=ā=rē=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}}=and \\ 
\glt `He said, ‘This is my meal.’'
\z 
 
\ea \label{ŽP.218}
\textit{ečīne nānē ečīne werū’} \\ 
\gll e=čīne nān-ē e=čīne ∅-wer-ū \\ 
 in=like meal\textsc{.m}\textsc{-indf} in=like \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `I eat such food.'
\z 
 
\ea \label{ŽP.225}
\textit{her lāloeke māčo, mišyo ī kināčē min meberūšo māreš biřūnew’} \\ 
\gll her lālo-eke m-āč-o mišyo ī kināčē min me-ber-ū=š=o māre=š ∅-biř-ūne=ū \\ 
 emph maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} aux dem.prox girl\textsc{.dir}\textsc{.f} \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} marraige\textsc{=3sg} \textsc{ind-}cut\textsc{.prs}\textsc{-\textsc{1sg}}=and \\ 
\glt ` The uncle kept saying, ‘I shall not take this girl back. I shall marry her [to Pir Shaliyar].’'
\z 
 
\ea \label{ŽP.226}
\textit{āđīč māčo, min metāwū ī kināčē wey kerū īne nānmā} \\ 
\gll āđ=īč m-āč-o min me-tāw-ū ī kināčē wey ∅-ker-ū īne nān=m=ā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox girl\textsc{.dir}\textsc{.f} raising \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `People … he (Pir Shaliyar) said, ‘I cannot take care of this girl. This is my food.’'
\z 
 
\ea \label{ŽP.230}
\textit{yoīč mā yoīč māčo, kāke gīyān! minīč šimšū yerū heywānēt mido pene} \\ 
\gll yo=īč=mā yo=īč m-āč-o kāke gīyān min=īč šimšī=ū yeher=ū heywān-ē=t mi-de-o pene \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother.vob dear.\textsc{voc} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `One [said], One person said, ‘Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{ŽP.236}
\textit{min hiɫošet miđo’} \\ 
\gll min hiɫoše=t mi-đe-o \\ 
 \textsc{1sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `[One of them said], ‘I will give you cracked wheat.’'
\z 
 
\ea \label{ŽP.237}
\textit{yoīč māčo, minīč tišit miđo’} \\ 
\gll yo=īč m-āč-o min=īč tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you [pomegranate] molasses.’'
\z 
 
\ea \label{ŽP.238}
\textit{yoīč māčo, minīč šimšū yeherit miđo} \\ 
\gll yo=īč m-āč-o min=īč šimšī=ū yeher=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you spleen and liver.’'
\z 
 
\ea \label{ŽP.239}
\textit{yoīč māčo minīč mekī mārū} \\ 
\gll yo=īč m-āč-o min=īč mekī m-ār-ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} salt\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `One said, ‘I will give you salt.’'
\z 
 
\ea \label{ŽP.240}
\textit{yoīč māčo minīč pīyāzī mārū} \\ 
\gll yo=īč m-āč-o min=īč pīyāz-ī m-ār-ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} onion\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `One said, ‘I will give you onions.’'
\z 
 
\ea \label{BP.19}
\textit{ā zemāne māčū} \\ 
\gll ā zemān=e m-āč-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{ind-}say.prs\textsc{-\textsc{1sg}} \\ 
\glt `I am talking about earlier times. '
\z 
 
\ea \label{BP.57}
\textit{āđ māčo, konē ī meʕmūrē mine činnē wextēn? je hewrāmān bīyēnē āmēnē pey biřew bāre} \\ 
\gll āđ m-āč-o ko=nē ī meʕmūr-ē min=e činnē wext-ē=n je hewrāmān bīyē=nē āmē=nē pey biře=ū bāre \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} where\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1sg}\textsc{=dem} some\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} for portion\textsc{.m}=and profit\textsc{.m} \\ 
\glt `He (the agha) says, ‘where have my officers been during this time? They were in Hawraman; they had gone to collect tax.'
\z 
 
\ea \label{BP.82}
\textit{kure māčo wiɫā min mišārle bonā} \\ 
\gll kure m-āč-o wiɫā min mišār-le bo=nā \\ 
 \textsc{fill} \textsc{ind-}say.prs\textsc{-3sg} by\_God \textsc{1sg} hand\_saw\textsc{.m}\textsc{-dim} become.prs\textsc{=cop}\textsc{.3pl} \\ 
\glt `He said, ‘I —there was only a small hand saw.'
\z 
 
\ea \label{BP.83}
\textit{dēr nejātim bonā serem biřēnā} \\ 
\gll dēr nejāt=im bo=nā sere=m biřē=nā \\ 
 late salvation\textsc{.m}\textsc{=\textsc{1sg}} become.prs\textsc{=cop}\textsc{.3pl} head\textsc{.m}\textsc{=\textsc{1sg}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `I finished decapitating late. I beheaded them.'
\z 
 
\ea \label{BP.84}
\textit{be mišārle serem biřēnā,} \\ 
\gll be mišār-le sere=m biřē=nā \\ 
 to hand\_saw\textsc{.m}\textsc{-dim} head\textsc{.m}\textsc{=\textsc{1sg}} a\_while\textsc{=cop}\textsc{.3pl} \\ 
\glt `I decapitated them using a small hand saw.'
\z 
 
\ea \label{BP.139}
\textit{berā kerāšāne zīnāneke, wextē miđyā mezā nū če noʕē bīyen ħejījīyāyčšā girten xuɫkū hejījī} \\ 
\gll ∅-ber-ā ∅-ker-ā=šā=ne zīnān-eke wext-ē mi-đye-ā me-zān-ū če noʕ-ē bīye=n ħejījī-yā=yčšā girte=n xuɫk=ū hejīj-ī \\ 
 \textsc{ind-}take.prs\textsc{-3pl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=3pl}\textsc{=\textsc{povb}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-\textsc{1sg}} what kind/way\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} people\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They took [them] and put them in the jail, they realised that —they didn’t know for what reason— [but] a person from Hajij had also been arrested.'
\z 
 
\ea \label{BP.155}
\textit{fermāwo āđīč, weɫā min pey šime girewū} \\ 
\gll ∅-fermāw-o āđ=īč weɫā min pey šime ∅-girew-ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} indeed \textsc{1sg} for \textsc{2pl} \textsc{ind-}cry.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘I’m crying for you.’'
\z 
 
\ea \label{BP.172}
\textit{hejījīyeke, fermāwo hejījīyeke, be qurbānit bim sūre gīyān yekē tirī lē de leqē teriš de wene’} \\ 
\gll hejījī-eke ∅-fermāw-o hejījī-eke be qurbān=it ∅-b-im sūr=e gīyān yek-ē tir=ī lē ∅-de-e leq-ē ter=iš ∅-de-e wene \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} to sacrifice\textsc{=\textsc{2sg}} \textsc{sbjv-}be.prs\textsc{-\textsc{1sg}} red=ez\textsc{.cmpd} dear.\textsc{voc} one\textsc{-indf} other\textsc{=3sg} at \textsc{imp-}give.prs-\textsc{2sg} kick\textsc{-indf} another\textsc{=3sg} \textsc{imp-}give.prs-\textsc{2sg} at \\ 
\glt `The [prisoner from] Hajij said [to Baba Khwada], ‘May I be your sacrifice, you red-skin, kick it once more! Kick it once more!”'
\z 
 
\ea \label{BP.181}
\textit{mēnē e čā ħemey ġeybī fermāwo, fermāwo, min mānīyēnmē e čē bā kučē nīšmēre} \\ 
\gll m-ē-nē e=čā ħeme=ī ġeybī ∅-fermāw-o ∅-fermāw-o min mānīyē=nmē e=čē bā kuč-ē ∅-nīš-mē=re \\ 
 \textsc{ind-}come.prs\textsc{-3pl} in=there \textsc{pn}=ez.\textsc{attr} invisible \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{1sg} tired\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} in=here \textsc{hort} little\textsc{-indf} \textsc{sbjv-}sit.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} \\ 
\glt `They went [to the region between Zawar and Tawela]. Hama Ghaybi said, ‘I, we are tired. Let us rest\footnote{Lit. sit.} a little bit here.'
\z 
 
\ea \label{BP.182}
\textit{seyēmēw min čurtēč kerū kučē būsū} \\ 
\gll sey-ēmē=ū min čurt-ē=č ∅-ker-ū kuč-ē b-ūs-ū \\ 
 rest.prs-\textsc{1pl}=and \textsc{1sg} nap\textsc{-indf}\textsc{=add} \textsc{ind-}do.prs\textsc{-\textsc{1sg}} little\textsc{-indf} \textsc{sbjv-}sleep.prs\textsc{-\textsc{1sg}} \\ 
\glt ` I shall rest,\footnote{Lit. My rest comes.} take a nap [and] sleep a little bit. They were a bit tired.'
\z 
 
\ea \label{BP.183}
\textit{kučē būsū wermīčim mēw kučēš mānīyānā} \\ 
\gll kuč-ē b-ūs-ū werm=īč=im m-ē=ū kuč-ē=š mānīyā=nā \\ 
 little\textsc{-indf} \textsc{sbjv-}sleep.prs\textsc{-\textsc{1sg}} sleep\textsc{=add}\textsc{=\textsc{1sg}} \textsc{ind-}come.prs\textsc{.3sg}=and little\textsc{-indf}\textsc{=3sg} tired\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I shall sleep a little bit. I feel sleepy.\footnote{Lit. my sleep comes.} I am a bit tired.'
\z 
 
\ea \label{BP.184}
\textit{fermāwo, min kučē mūsū welē wesīyetēm hen kerū:} \\ 
\gll ∅-fermāw-o min kuč-ē m-ūs-ū welē wesīyet-ē=m hen-∅ ∅-ker-ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{1sg} little\textsc{-indf} \textsc{ind-}sleep.prs\textsc{-\textsc{1sg}} but will.\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} exist\textsc{-3sg}\textsc{.m} \textsc{ind-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, I will sleep a bit. However, I have a [last] will to give:'
\z 
 
\ea \label{BP.185}
\textit{šime luāydēwe min kuřēm je lemū ženekēmne, hāmīlēne} \\ 
\gll šime luā-ydē=we min kuř-ē=m je leme=ū ženī-ekē=m=ne hāmīlē=ne \\ 
 \textsc{2pl} go\textsc{.pst}-\textsc{2pl}\textsc{=compl} \textsc{1sg} son/boy\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} from stomach.f\textsc{=ez.gen} woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} pregnant.f\textsc{=\textsc{post}} \\ 
\glt `You go back. My wife is pregnant with my son.\footnote{Lit. there is a son in my wife’s belly.} She is pregnant.'
\z 
 
\ea \label{ŠJ.21}
\textit{lodē bārdēš peym šē ʕumerī} \\ 
\gll ∅-lo-dē b-ār-dē=š pey=m šē ʕumer-ī \\ 
 \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}bring.prs-\textsc{2pl}\textsc{=3sg} for\textsc{=\textsc{1sg}} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Go [and] bring Sheikh Umer to me.’'
\z 
 
\ea \label{ŠJ.37}
\textit{fermāwo hurbēzdē gumē bīdē gumē bīdē werū čemīmo!} \\ 
\gll ∅-fermāw-o hur- b-ēz-dē gum-ē ∅-b-īdē gum-ē ∅-b-īdē wer=ū čem-ī=m=o \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{pvb-}\textsc{imp-}rise.prs-\textsc{2pl} lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} out\_of\textsc{=ez.gen} eye\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} \\ 
\glt ` He said, ‘Get up and go away! Get out of my sight!'
\z 
 
\ea \label{ŠJ.52}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `‘Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.53}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.61}
\textit{māčo erē min āmānā zānū} \\ 
\gll m-āč-o erē min āmā=nā ∅-zān-ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} yes \textsc{1sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{ind-}know.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Yes, I have come [here]; I know [where it is].’'
\z 
 
\ea \label{ŠJ.62}
\textit{māčo berem pištū berew wiɫāxekā} \\ 
\gll m-āč-o ∅-ber-e=m pišt=ū bere=ū wiɫāx-ekā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}take.prs-\textsc{2sg}\textsc{=\textsc{1sg}} back\textsc{=ez.gen} door\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} \\ 
\glt `He (Jafir Sān) said, ‘Take me behind the door of the horses’ stable!’'
\z 
 
\ea \label{ŠJ.74}
\textit{fermāwo lodē lāš wāčdē kāriš be min nīyā} \\ 
\gll ∅-fermāw-o ∅-lo-dē lā=š wāč-dē kār=iš be min nīye=ā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} at\_the\_place\_of\textsc{=3sg} say.prs-\textsc{imp-}\textsc{2pl} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go to him [and] say, ‘His business is not with me.'
\z 
 
\ea \label{ŠJ.76}
\textit{hīč kārēš be min nīyā} \\ 
\gll hīč kār-ē=š be min nīye=ā \\ 
 nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He has no business with me.’'
\z 
 
\ea \label{ŠJ.77}
\textit{milā māčā qurbān šē ʕusmānī fermāwā kāriš be min nīyā} \\ 
\gll mi-l-ā m-āč-ā qurbān šē ʕusmān-ī fermāwā kār=iš be min nīye=ā \\ 
 \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] said [to Jafir Sān], ‘Sir, the sheikh said, “He has no business with me.'
\z 
 
\ea \label{ŠJ.86}
\textit{fermāwo qurbān hermāne be desū min nīyenew} \\ 
\gll ∅-fermāw-o qurbān hermāne be des=ū min nīye=ne=ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} work\textsc{.f} to hand\textsc{.m}\textsc{=ez.gen} \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `He (the sheikh) said, ‘Sir, the task [of ruling] is not in my hands;'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{KŠ.1}
\textit{ew wextī īnē her girđ řāzū eyemnē lāšo bīyenā eyām} \\ 
\gll ew wext-ī īnē her girđ řāz=ū eye=m=nē lā=š=o bīye=nā eyā=m \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox}\textsc{.3pl} just all tale\textsc{.m}\textsc{=ez.gen} mother\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3pl} at\_the\_place\_of\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} mother\textsc{=\textsc{1sg}} \\ 
\glt `Well, these are all my mother’s tales. I was with my mother.'
\z 
 
\ea \label{KŠ.2}
\textit{dūr gošdārā neweše bē lāšo min bēnē} \\ 
\gll dūr gošdār-ā neweš-e b-ē lā=š=o min b-ēn-ē \\ 
 far listener\textsc{.m}\textsc{-obl}\textsc{.pl} ill\textsc{-f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} at\_the\_place\_of\textsc{=3sg}\textsc{=\textsc{post}} \textsc{1sg} be\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1sg}} \\ 
\glt `May it be far from the audience, she was sick. I was [staying] with her.'
\z 
 
\ea \label{KŠ.9}
\textit{māčo, nāmerdit pene kerdā, čī pāset pene kerdā?’} \\ 
\gll m-āč-o nāmerd=it pene kerd-ā čī pāse=t pene kerd-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} just\textsc{=\textsc{2sg}} with do\textsc{.pst}\textsc{-\textsc{1sg}} why this\textsc{=\textsc{2sg}} with do\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘You didn’t act justly towards me! why did you do this to me?'
\z 
 
\ea \label{KŠ.10}
\textit{māčo, to kēnī min werew bāɫdārīm teqnān’} \\ 
\gll m-āč-o to kē=nī min wer=ew bāɫdār-ī=m teqnā=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{2sg} who\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{1sg} at\textsc{=\textsc{post}} bird\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} shoot\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (the king) said, ‘Who are you? I shot the birds.'
\z 
 
\ea \label{KŠ.11}
\textit{māčo, min čārenwīsnā, čārenwīs!’} \\ 
\gll m-āč-o min čārenwīs=nā čārenwīs \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} fortune\_teller\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} fortune\_teller\textsc{.m} \\ 
\glt `He said, ‘I am a fortune teller\footnote{Lit. destiny writer.}, a fortune teller.'
\z 
 
\ea \label{KŠ.13}
\textit{māčo, niwīsem pey kuřū šuwānī’} \\ 
\gll m-āč-o niwīs-e=m pey kuř=ū šuwāne-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} write\textsc{.pst}\textsc{-3sg}\textsc{.f}\textsc{=\textsc{1sg}} to son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the fortune teller) said, ‘I have chosen her for a shepherd’s son.’'
\z 
 
\ea \label{KŠ.28}
\textit{wātiš, qurbān herkey gele berī epāne min ī bizeme čekyo} \\ 
\gll wāt=iš qurbān herkey gele ∅-ber-ī epāne min ī bize=m=e ∅-čekye-o \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} each\_time herd\textsc{.m} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} there \textsc{1sg} \textsc{dem.prox} goat\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{ind-}drain\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `[and] said, ‘Sir, whenever you take the herd there, my goat[‘s milk] is drained.'
\z 
 
\ea \label{KŠ.29}
\textit{āđīč wāt, dey xāsā min gino šūnī[šre]} \\ 
\gll āđ=īč wāt dey xās=ā min ∅-gin-o šūnī=š=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} track\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `He (the shepherd) said, ‘All right, I will look after her (the goat).’'
\z 
 
\ea \label{KŠ.51}
\textit{māčo, be xwā pāđšām seɫāmet kuřū šuwāney heke to kuštēū ā wextīyekey hukmit dāū īnīšā īse īnā yānetā īnīšā} \\ 
\gll m-āč-o be xwā pāđšā=m seɫāmet kuř=ū šuwāne-ī heke to kušt-ē=ū ā wext-ī-ekey hukm=it dā=ū īnīšā īse īnā-∅ yāne=tā īnīšā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} king\textsc{.m}\textsc{=\textsc{1sg}} healthy son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} if \textsc{2sg} kill\textsc{.pst}\textsc{-3pl}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} rule\textsc{.m}\textsc{=\textsc{2sg}} give\textsc{.pst}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} now \textsc{deic}\textsc{-3sg}\textsc{.m} house\textsc{.m}=\textsc{2pl} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `They said [in the letter], ‘Indeed, may my king be healthy, the shepherd’s son whom you ordered to be killed is now in your house and so on’'
\z 
 
\ea \label{KŠ.52}
\textit{āđīč māčo, wāčdē bē melāqātīm} \\ 
\gll āđ=īč m-āč-o wāč-dē b-ē melāqātī=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} say\textsc{.prs}-\textsc{imp-}\textsc{2pl} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} visitor\textsc{.m}\textsc{=\textsc{1sg}} \\ 
\glt `He (the king) said, ‘Tell him to come to visit me.’'
\z 
 
\ea \label{KŠ.56}
\textit{yewāšē čāge, čogene šewē minwīso, māčo, ey kesūkārū min, ey henū nīyenū min, ey īnīšā ke mintā gereknā, tāyfew min heɫbetene ī kuře biřiš dāwe, be her jor ī kuřī beynne berdē.’} \\ 
\gll yewāšē čāge č=oge=ne šewe-ē mi-nwīs-o m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey īnīšā ke min=tā gerek=nā tāyfe=ū min heɫbete=ne ī kuř=e biř=iš dā=we be her jor ī kuř-ī beyn=ne ∅-ber-dē \\ 
 then there in=there\textsc{=\textsc{post}} night\textsc{.f}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} rel \textsc{1sg}=\textsc{2pl} necessary\textsc{=cop}\textsc{.\textsc{1sg}} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} by any means \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} between\textsc{=\textsc{post}} \textsc{ind-}take\textsc{.prs}-\textsc{2pl} \\ 
\glt `Then, one night he (the king) wrote [a letter], saying, ‘O my relatives, O my close family members , O those who want me, [are] from my tribe, [when you read this letter] it is obvious that this boy has arrived. By any means possible, kill him.'
\z 
 
\ea \label{KŠ.72}
\textit{niwīsebēš, ey kesūkārū min, ey henū nīyenū min, heɫbetene ī kuřī biřiš dāwe, šime be wesīlēwe kuždēš, beynne berdēš!} \\ 
\gll niwīse b-ē=š ey kesūkār=ū min ey hen=ū nīyen=ū min heɫbete=ne ī kuř-ī biř=iš dā=we šime be wesīle-ēwe ∅-kuž-dē=š beyn=ne ∅-ber-dē=š \\ 
 write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} \textsc{2pl} by tool\textsc{.m}\textsc{-indf} \textsc{imp-}kill\textsc{.prs}-\textsc{2pl}\textsc{=3sg} between\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2pl}\textsc{=3sg} \\ 
\glt `He had written, ‘O my relatives, O my close family members, [when you read this letter] it is evident that this boy has arrived. Kill him with a thing, destroy him!'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.81}
\textit{yewāšē seʕbē hurmēzāwe māčo, xeyrāmāyš kero merhebāyš kero māčo, tāteym hāɫū ēhwāɫtā?’} \\ 
\gll yewāšē seʕbe-ē hur- mi-ēz-ā=we m-āč-o xeyrāmāy=š ∅-ker-o merhebā-ī=š ∅-ker-o m-āč-o tāte-ī=m hāɫ=ū ēhwāɫ=tā \\ 
 then morning\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} welcoming\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} hello\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} state\textsc{.m}\textsc{\textsc{=ez.gen}} demeanour=\textsc{2pl} \\ 
\glt `The following day, they woke up. She said … she welcomed him (the shepherd’s son), greeted him [and] said, ‘How is my father doing?’'
\z 
 
\ea \label{KŠ.83}
\textit{xulāsew kelāmī qisēš čenū min kerdēw} \\ 
\gll xulāse=ū kelām-ī qisē=š čenū min kerd-ē=ū \\ 
 summary=and speech\textsc{-obl}\textsc{.m} talk\textsc{.dir}\textsc{.pl}\textsc{=3sg} with\textsc{\textsc{.ez}.gen} \textsc{1sg} do\textsc{.pst}\textsc{-3pl}=and \\ 
\glt `In short, he (the king) talked to me.'
\z 
 
\ea \label{KŠ.84}
\textit{īnā qāqezēčiš dānā pey} \\ 
\gll īne=ā qāqez-ē=č=iš dā=nā pey \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} letter\textsc{.m}\textsc{-indf}\textsc{=add}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to \\ 
\glt `Look, he has given me a letter.'
\z 
 
\ea \label{KŠ.85}
\textit{pey lāw šime ārdenmo} \\ 
\gll pey lā=ū šime ārde=n=m=o \\ 
 to at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=compl} \\ 
\glt `I have brought it for you.'
\z 
 
\ea \label{KŠ.86}
\textit{wātiš dey lue āney fiɫānīū fiɫānī čiře ī qewmāme bā bēnēwe qāqezekey wenā} \\ 
\gll wāt=iš dey ∅-lu-e āney fiɫānī=ū fiɫānī ∅-čiř-e ī qewm-ā=m=e bā b-ē-nē=we qāqez-ekey ∅-wen-ā \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} such\_and\_such\_person=and such\_and\_such\_person \textsc{imp-}call\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{hort} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}read\textsc{.prs}\textsc{-3pl} \\ 
\glt `She said, ‘Then go, and summon such-and-such, our relatives, so that they come and read the letter.’'
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{ŽH.1}
\textit{min zāroɫe bēnē, lāw eyemo} \\ 
\gll min zāroɫe b-ēn-ē lā=ū eye=m=o \\ 
 \textsc{1sg} child be\textsc{.prs}\textsc{-pstc}\textsc{-1sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} mother\textsc{.obl}\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `I was a child. [I was] with my mother.'
\z 
 
\ea \label{ŽH.2}
\textit{eyām qisē kerdē wāčē, zemānū qedīmī pādšā hāřūne(l) řešīd bīyenū nā siɫtān mehmūw heyāsī žīr} \\ 
\gll eyā=m qisē-ē kerdē-ē wāč-ē zemān=ū qedīm-ī pādšā hāřūn-e(l) řešīd bīye=n=ū nā siɫtān mehmū=ū heyās=ī žīr \\ 
 mother\textsc{=1sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-3pl} say\textsc{.prs-}pstc\textsc{.3sg} time\textsc{.m}\textsc{=ez}\textsc{.gen} old\_time\textsc{-obl}\textsc{.m} king\textsc{.m} \textsc{pn}-of \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and no sultan\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez}.\textsc{attr} shrewd \\ 
\glt `My mother would narrate [tales]. She said, ‘In the olden days, there was Harun Rashid, the king. No, [there was] Sultan Mahmud and Hayas, the shrewd.'
\z 
 
\ea \label{ŽH.3}
\textit{wāčē řoē je řoā qāɫešā bo, serū mezān ū čēšī} \\ 
\gll wāč-ē řo-ē je řo-ā qāɫe=šā ∅-b-o sere=ū me-zān-ū čēš-ī \\ 
 say\textsc{.prs-}pstc\textsc{.3sg} day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} quarrel\textsc{\textsc{.f}}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} head\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what\textsc{-obl}\textsc{.m} \\ 
\glt `According to her (my mother),\footnote{Lit. She would say.}  Once they (Sultan Mahmud and Hayas) got into a quarrel with each other, about something I don’t know.'
\z 
 
\ea \label{ŽH.10}
\textit{māčo, milū} \\ 
\gll m-āč-o mi-l-ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} \\ 
\glt `He (the man) said, ‘I’m going [to my village]’'
\z 
 
\ea \label{ŽH.11}
\textit{māčo, dey bilmē minīč mewt čene} \\ 
\gll m-āč-o dey bi-l-mē min=īč m-e-ū=t čene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} 1sg\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{2sg}} with \\ 
\glt `He (Hayas) said, ‘Let’s go, I am also coming with you.’'
\z 
 
\ea \label{ŽH.13}
\textit{ewīč māčo, min pilekānī mezānū binyewnew mezānū čēš māčī!’} \\ 
\gll ew=yč m-āč-o min pilekānī me-zān-ū bi-nye-ūne=ū me-zān-ū čēš m-āč-ī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg stair\textsc{\textsc{.f}} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} \textsc{sbjv-}put\textsc{.prs}\textsc{-1sg}=and \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what \textsc{ind-}say\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘I can’t set up a ladder and I don’t understand\footnote{Lit. know.}  what you are saying!’'
\z 
 
\ea \label{ŽH.23}
\textit{māčo, ne wiɫā řoɫe, āmānē pāynū serekoē wāčmēš pene be zuwānū hewrāmīyānew wēmā derey} \\ 
\gll m-āč-o ne wiɫā řoɫe āmā-ānē pāyn=ū sereko-ē ∅-wāč-mē=š pene be zuwān=ū hewrāmī-yāne=ū wē=mā dere-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no by\_God child.\textsc{voc} come\textsc{.pst}\textsc{-1sg} bottom\textsc{=ez}\textsc{.gen} uphill\textsc{.m}\textsc{-indf} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-1pl}\textsc{=3sg} to by language\textsc{.m}\textsc{=ez}\textsc{.gen} Hawrami-vernacular\_of\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} valley\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘No, child, I went to the bottom of the slope —we call it ‘daray’ in our Hawrami language—'
\z 
 
\ea \label{ŽH.24}
\textit{neferē bī řefēqim,} \\ 
\gll nefer-ē bī-∅ řefēq=im \\ 
 person\textsc{.m}\textsc{-indf} be\textsc{.pst}\textsc{-3sg}\textsc{.m} friend\textsc{=1sg} \\ 
\glt `I made friends with a person. '
\z 
 
\ea \label{ŽH.26}
\textit{māčo, nezānām} \\ 
\gll m-āč-o ne-zānā=m \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}know\textsc{.pst}\textsc{=1sg} \\ 
\glt `He (the man) said, ‘I didn’t understand [his point].’'
\z 
 
\ea \label{ŽH.31}
\textit{wātim peneš to milī ko?”} \\ 
\gll wāt=im pene=š to mi-l-ī ko \\ 
 say\textsc{.pst}\textsc{=1sg} to\textsc{=3sg} \textsc{2sg} \textsc{ind-}go\textsc{.prs}-\textsc{2sg} where \\ 
\glt `I said to him, “Where are you going?”’'
\z 
 
\ea \label{ŽH.33}
\textit{wātiš, tāte, āne wāteniš, eger to xuɫkem mekerī, meberīm yāne milū mizgī’} \\ 
\gll wāt=iš tāte āne wāte=n=iš eger to xuɫke=m me-ker-ī me-ber-ī=m yāne mi-l-ū mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} if \textsc{2sg} invitation\textsc{=1sg} \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \textsc{neg-}take\textsc{.prs}-\textsc{2sg}\textsc{=1sg} house\textsc{.m}\textsc{.dir.sg} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} mosque\textsc{.m} \\ 
\glt `She said, ‘Father, he meant,\footnote{Lit. said.} “If you do not invite me and won’t take me home, I will go to the mosque.”'
\z 
 
\ea \label{ŽH.37}
\textit{řo yerē āmāwe wātiš be kināčekē, řoɫe řefēqekem her mizgīne īnā mizgīne} \\ 
\gll řo yerē āmā=we wāt=iš be kināčē-ekē řoɫe řefēq-eke=m her mizgī=ne īnā-∅ mizgī=ne \\ 
 day\textsc{.m} three come\textsc{.pst}\textsc{.3sg}\textsc{=compl} say\textsc{.pst}\textsc{=3sg} to daughter\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} child.\textsc{voc} friend\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} just mosque\textsc{.m}\textsc{=\textsc{post}} deic\textsc{-3sg}\textsc{.m} mosque\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `On the third day, he (the father) returned [home and] said to the daughter, ‘Child, my friend is still in the mosque; he is at the mosque.’'
\z 
 
\ea \label{ŽH.41}
\textit{ā wextīyekey kināčekē beroš řāre māčo, be xwā īse kučēš čene werū.} \\ 
\gll ā wext-ī-ekey kināčē-ekē ∅-ber-o=š řā=re m-āč-o be xwā īse kuč-ē=š čene ∅-wer-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} now little\textsc{-indf}\textsc{=3sg} from \textsc{ind-}eat\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then, the girl set off on the way. She said, ‘Well, I shall eat some of it now.’'
\z 
 
\ea \label{ŽH.42}
\textit{kučēš čene werū ī māsī’} \\ 
\gll kuč-ē=š čene ∅-wer-ū ī mās-ī \\ 
 little\textsc{-indf}\textsc{=3sg} with \textsc{ind-}eat\textsc{.prs}\textsc{-1sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `I shall eat some of it.'
\z 
 
\ea \label{ŽH.49}
\textit{māčo, beɫē kečim kerd!’} \\ 
\gll m-āč-o beɫē keč=im kerd \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes crooked\textsc{=1sg} do\textsc{.pst} \\ 
\glt `She (the younger sister), ‘Indeed, I made it less.’'
\z 
 
\ea \label{ŽH.53}
\textit{māčo, ī serū řīšīme kučē tāše’} \\ 
\gll m-āč-o ī sere=ū řīš-ī=m=e kuč-ē ∅-tāš-e \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox} head\textsc{.m}=and beard\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} little\textsc{-indf} \textsc{imp-}shave\textsc{.prs}-\textsc{2sg} \\ 
\glt `He (Hayas) said [to the girl], ‘shave my face a bit.’'
\z 
 
\ea \label{ŽH.59}
\textit{māčo, dey min šūyš kerū pene’} \\ 
\gll m-āč-o dey min šū-ī=š ∅-ker-ū pene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} 1sg husband\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} to \\ 
\glt `She said, ‘I will marry him.'
\z 
 
\ea \label{ŽH.60}
\textit{šūyš kerū pene} \\ 
\gll šū-ī=š ∅-ker-ū pene \\ 
 husband\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} to \\ 
\glt `I will marry him.’'
\z 
 
\ea \label{ŽH.67}
\textit{āđīč māčo, čiřdēš peym’} \\ 
\gll āđ=īč m-āč-o ∅-čiř-dē=š pey=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}call\textsc{.prs}\textsc{-2pl}\textsc{=3sg} for\textsc{=1sg} \\ 
\glt `He (the Sultan) said, ‘Call him for me.’'
\z 
 
\ea \label{ŽH.78}
\textit{āđīč māčo, min desūrim pēsnen pey wēt āmānā} \\ 
\gll āđ=īč m-āč-o min desūr=im pēsne=n pey wē=t āmā=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg task/order\textsc{.m}\textsc{=1sg} like\_this\textsc{=cop}\textsc{.3sg}\textsc{.m} for \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \\ 
\glt `He said, ‘My task is like this! I have come for you.’'
\z 
 
\ea \label{ŽH.79}
\textit{ā wextī yewāšē māčo, dey tehemuɫ kere bā hēɫeřūenit pey kerū, hāɫīt kerū'} \\ 
\gll ā wext-ī yewāšē m-āč-o dey tehemuɫ ∅-ker-e bā hēɫeřūen=it pey ∅-ker-ū hāɫī=t ∅-ker-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} then \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} patience \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{hort} fried\_egg\textsc{=\textsc{2sg}} for \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} understood\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then she said, ‘All right, wait a little bit, let me cook you fried eggs, [and] make you understand.'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.104}
\textit{ānā dersē ʕālit dānē} \\ 
\gll āne=ā ders-ē ʕāl=it dā-ānē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} lesson\textsc{.m}\textsc{-indf} good\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{-1sg} \\ 
\glt `You gave me a good lesson.'
\z 
 
\ea \label{ŽH.105}
\textit{memnūnitnā} \\ 
\gll memnūn=it=nā \\ 
 grateful\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.1sg} \\ 
\glt `I am grateful to you.'
\z 
 
\ea \label{ŽH.113}
\textit{māčo, qurbān čī bilūwe! min činnē hāy wextē bāxēm bāxēm ārd ber} \\ 
\gll m-āč-o qurbān čī bi-l-ū=we min činnē hāy wext-ē bāx-ē=m bāx-ē=m ārd-∅ ber \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} why \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg}\textsc{=compl} 1sg some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{-indf} garden\textsc{.m}\textsc{-indf}\textsc{=1sg} garden\textsc{.m}\textsc{-indf}\textsc{=1sg} bring\textsc{.pst}\textsc{-3sg}\textsc{.m} product \\ 
\glt `He (Hayas) said, ‘Sir, why should I go back (home)? After a long time my garden has produced [fruit]'
\z 
 
\ea \label{ŽH.114}
\textit{pāse zānū šērē lūān dilēš!} \\ 
\gll pāse ∅-zān-ū šēr-ē lūā=n dilē=š \\ 
 like \textsc{ind-}know\textsc{.prs}\textsc{-1sg} lion\textsc{.m}\textsc{-indf} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} inside\textsc{=3sg} \\ 
\glt `Now to me it is as if a lion has trespassed in it.'
\z 
 
\ea \label{ŽH.116}
\textit{āđīč wāt, ā šēre minnā} \\ 
\gll āđ=īč wāt ā šēr=e min=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{dem.dist} lion\textsc{.m}\textsc{=dem} 1sg\textsc{=cop}\textsc{.1sg} \\ 
\glt `He (the sultan) said, ‘I am that lion.'
\z 
 
\ea \label{ŽH.117}
\textit{kārēš pene kerdenā be mehremū wēm qubuɫim kerden.} \\ 
\gll kār-ē=š pene kerde=nā be mehrem=ū wē=m qubuɫ=im kerde=n \\ 
 task\textsc{.m}\textsc{-indf}\textsc{=3sg} to do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `She made  me such that I have accepted her as my own kindred. '
\z 
 
\ea \label{RE.5}
\textit{āmān lūān wāteniš, birā gīyān! min xū wēt zānī hāwmāɫitnā} \\ 
\gll āmā=n lūā=n wāte=n=iš birā gīyān min xū wē=t ∅-zān-ī hāwmāɫ=it=nā \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} brother\textsc{.m} dear.\textsc{voc} \textsc{1sg} \textsc{disc.ptcl} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} neighbour\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `They would go [and] say, ‘Dear brother, you know [well] that I am your neighbour.'
\z 
 
\ea \label{RE.6}
\textit{be bīyew nebīyeym zānī} \\ 
\gll be bīyey=w nebīyey=m ∅-zān-ī \\ 
 to resources\textsc{.m}\textsc{\textsc{=ez.gen}} s\textsc{hort}age\textsc{.m}\textsc{=\textsc{1sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} \\ 
\glt `You know about my resources and shortages. '
\z 
 
\ea \label{RE.7}
\textit{ʕerzū birāw wēm kerū ī kināčēte de pī kuřīme} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū ī kināčē=t=e ∅-de-e p=ī kuř-ī=m=e \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} brother\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} at=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \\ 
\glt `Let me tell you,\footnote{Lit. May a petition be at the service of my brother.} ‘Give your daughter to my son [in marriage]'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.20}
\textit{lue kārekem pey kere} \\ 
\gll ∅-lu-e kār-eke=m pey ∅-ker-e \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} task\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} for \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Go do [this] task for me.'
\z 
 
\ea \label{RE.29}
\textit{kābrā melo wāčo, ženīm de} \\ 
\gll kābrā me-l-o ∅-wāč-o ženī=m ∅-de-e \\ 
 man \textsc{neg-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} woman\textsc{.f}\textsc{=\textsc{1sg}} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `The fellow wouldn’t go [to his parents and] say, ‘Find a girl for me!’\footnote{Lit. Give me a woman.}'
\z 
 
\ea \label{RE.36}
\textit{min, qewɫim dān pey fiɫāne kesī} \\ 
\gll min qewɫ=im dā=n pey fiɫān=e kes-ī \\ 
 \textsc{1sg} arrangement\textsc{.m}\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `I have promised such-and-such a person.'
\z 
 
\ea \label{RE.37}
\textit{miđey mehr kerū,} \\ 
\gll mi-đe-ī mehr ∅-ker-ū \\ 
 \textsc{ind-}give\textsc{.prs}-\textsc{2sg} mariage\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `If you accept,\footnote{Lit. give.} I will marry [him]'
\z 
 
\ea \label{RE.38}
\textit{međey mehr kerū,} \\ 
\gll me-đe-ī mehr ∅-ker-ū \\ 
 \textsc{neg-}give\textsc{.prs}-\textsc{2sg} mariage\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `If you don’t accept, I will marry [him]'
\z 
 
\ea \label{RE.49}
\textit{min kināčēt dēne kuřū fiɫāne kesī} \\ 
\gll min kināčē=t dē=ne kuř=ū fiɫān=e kes-ī \\ 
 \textsc{1sg} daughter\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}} son/boy\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and say to the lord], ‘I have given my\footnote{Lit. your.} daughter to the son of such-and-such person.'
\z 
 
\ea \label{RE.60}
\textit{sūrāne yānē mišo bārī to zeř dey jā min māčū erē} \\ 
\gll sūrāne yānē mišo ∅- b-ār-ī to zeř ∅-de-ī jā min m-āč-ū erē \\ 
 \textsc{pn} that\_means aux \textsc{ind-}\textsc{sbjv-}bring\textsc{.prs}-\textsc{2sg} \textsc{2sg} money\textsc{.m} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} afterwards \textsc{1sg} \textsc{ind-}say\textsc{.prs}\textsc{-\textsc{1sg}} yes \\ 
\glt `Surana means that you should give me money, then I will say, ‘Yes [to the marriage]'
\z 
 
\ea \label{ŽE.43}
\textit{eđāw tātew min her beđbext bīyen} \\ 
\gll eđā=ū tāte=ū min her beđbext bīye=n \\ 
 mother\textsc{.f}=and father\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} emph poor be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `My parents were poor.'
\z 
 
\ea \label{ŽE.59}
\textit{seʕbē tā wēregā min duē nānē bē} \\ 
\gll seʕbe-ē tā wēregā min duē nān-ē b-ē \\ 
 morning\textsc{.f}\textsc{-obl}\textsc{.f} until evening \textsc{1sg} two meal\textsc{.m}\textsc{-indf} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `I, there were a few\footnote{Lit. two} [loads of] bread [for me to cook] from morning to evening.'
\z 
 
\ea \label{ŽE.60}
\textit{nānim peten} \\ 
\gll nān=im pete=n \\ 
 bread\textsc{.m}\textsc{=\textsc{1sg}} bake\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I would bake bread. '
\z 
 
\ea \label{ŽE.61}
\textit{āwīm ārdēne} \\ 
\gll āwī=m ārdē=ne \\ 
 water\textsc{.f}\textsc{=\textsc{1sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `I brought water.'
\z 
 
\ea \label{ŽE.62}
\textit{luēnā ħeywānim bīyen} \\ 
\gll luē=nā ħeywān=im bīye=n \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.\textsc{1sg}} animal\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I went … I had animals.'
\z 
 
\ea \label{ŽE.63}
\textit{ħeywāndārīm kerden} \\ 
\gll ħeywāndārī=m kerde=n \\ 
 animal\_husbandry\textsc{=\textsc{1sg}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I tended animals.'
\z 
 
\ea \label{ŽE.64}
\textit{herkām girewēbā bīnīenim koɫīmre} \\ 
\gll herkām girewē b-ā bīnīe=n=im koɫī=m=re \\ 
 whoever cry\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be\textsc{.prs}\textsc{-3pl} tie\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} shoulder\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} \\ 
\glt `Each [of my kids] who might have cried, I would put on my shoulders.'
\z 
 
\ea \label{ŽE.65}
\textit{bēškēm, bēškēne bīyēnē} \\ 
\gll bēške-ē=m bēške-ē=ne bīyē=nē \\ 
 cot\textsc{.f}\textsc{-indf}\textsc{=\textsc{1sg}} cot\textsc{.f}\textsc{-indf}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They were in a cot.'
\z 
 
\ea \label{ŽE.66}
\textit{hurim gērtēnē} \\ 
\gll hur=im gērtē=nē \\ 
 \textsc{pvb}\textsc{=\textsc{1sg}} take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `I lifted them.'
\z 
 
\ea \label{ŽE.67}
\textit{āwū nānim dēnē nīyēnēm nām bēškekē} \\ 
\gll āwī=ū nān=im dē=nē nīyē=nē=m nām bēške-ekē \\ 
 water\textsc{.f}\textsc{\textsc{=ez.gen}} bread\textsc{.m}\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} inside cot\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `I gave them food [and] put them back in the cot.'
\z 
 
\ea \label{ŽE.73}
\textit{ā wextī to mīsāl wātēt, ā fiɫāne kesem gerekene’} \\ 
\gll ā wext-ī to mīsāl wāt-ē=t ā fiɫān=e kese=m gerek-e=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg} for\_example say\textsc{.pst}\textsc{-cond}\textsc{=\textsc{2sg}} \textsc{dem.dist} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.f}\textsc{=\textsc{1sg}} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `In earlier times, let us say\footnote{Lit. for instance.} you said, ‘I want that certain person (i.e., girl).''
\z 
 
\ea \label{ŽE.74}
\textit{luēnī lāw ī kuřī wāčēnī, dā lue wāče eđēš bizāne ī kināčēše miđom pene yā ne’} \\ 
\gll lu-ēn-ī lā=ū ī kuř-ī wāč-ēn-ī dā ∅-lu-e ∅-wāč-e eđē=š bi-zān-e ī kināčē=š=e mi-đe-o=m pene yā ne \\ 
 go\textsc{.prs}\textsc{-pstc}-\textsc{2sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.prs}\textsc{-pstc}-\textsc{2sg} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}say\textsc{.prs-}.\textsc{2sg} mother\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=3sg}\textsc{=dem} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{1sg}} to either no \\ 
\glt ` You would go to this boy [seated next to the narrator] and would say, ‘Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!’'
\z 
 
\ea \label{ŽE.77}
\textit{eger kināčekēšā donē wātēš, erē kerū’} \\ 
\gll eger kināčē-ekē=šā don-ē wāt-ē=š erē ∅-ker-ū \\ 
 if girl\textsc{.obl}\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3pl} talk\_to\textsc{-cond} say\textsc{.pst}\textsc{-cond}\textsc{=3sg} yes \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `After they talk to the girl [and] she said, ‘Yes, I do’, '
\z 
 
\ea \label{ŽE.81}
\textit{wāčē, erē miđewš’} \\ 
\gll wāč-ē erē mi-đe-ū=š \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} yes \textsc{ind-}give\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg} \\ 
\glt `They would say, ‘Yes, I will give her.’'
\z 
 
\ea \label{ŽM.1}
\textit{heštā sāɫē ʕemirmā} \\ 
\gll heštā sāɫe-ē ʕemr=m=ā \\ 
 eighty year\textsc{.f}\textsc{-dir}\textsc{.pl} age\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I am eighty years old.'
\z 
 
\ea \label{ŽM.2}
\textit{tātem duē ženīš ārdēnē} \\ 
\gll tāte=m duē ženī=š ārdē=nē \\ 
 father\textsc{.m}\textsc{=\textsc{1sg}} two woman\textsc{.dir}\textsc{.pl}\textsc{=3sg} bring\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `My father married two women.'
\z 
 
\ea \label{ŽM.3}
\textit{min kuřū duēnā} \\ 
\gll min kuř=ū duē=nā \\ 
 \textsc{1sg} son/boy\textsc{.m}\textsc{=ez.gen} second\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I am from the second one.'
\z 
 
\ea \label{ŽM.4}
\textit{šūnū āneyre, šūnū řāy eweɫ je kārīgerīwe serem beršīyen} \\ 
\gll šūn=ū āney=re šūn=ū řā=ī eweɫ je kārīgerī=we sere=m ber-šīye=n \\ 
 after\textsc{=ez.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} after\textsc{=ez.gen} road\textsc{.f}=ez.\textsc{attr} first from labourer\_job\textsc{=compl} head\textsc{.m}\textsc{=\textsc{1sg}} out- go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Well, I first learned\footnote{Lit. my head went out.}  to work as a labourer.'
\z 
 
\ea \label{ŽM.5}
\textit{je herbeneywe serem beršīyen} \\ 
\gll je herbene-ī=we sere=m ber-šīye=n \\ 
 from donkey breeder\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} head\textsc{.m}\textsc{=\textsc{1sg}} out- go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I learned to work as a donkey breeder.'
\z 
 
\ea \label{ŽM.6}
\textit{šūnīre āmānā desim kerđen kārēz biřyey} \\ 
\gll šūnī=re āmā=nā des=im kerđe=n kārēz biřyey \\ 
 afterwards\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} hand\textsc{.m}\textsc{=\textsc{1sg}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} subterranean\_canal\textsc{.m} cut\textsc{.inf} \\ 
\glt `Then, I started\footnote{Lit. put hand on.} to dig subterranean canals.'
\z 
 
\ea \label{ŽM.8}
\textit{īse ħewt sāɫēč āmānāwe serū īney dūkāndārnā} \\ 
\gll īse ħewt sāɫe-ē=č āmā=nā=we ser=ū īney dūkāndār=nā \\ 
 now seven year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} shopkeeper\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{ŽM.9}
\textit{qāčim meřyānū kewtenā yānew} \\ 
\gll qāč=im meřyā=n=ū kewte=nā yāne=ū \\ 
 leg\textsc{.m}\textsc{=\textsc{1sg}} break\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} house\textsc{.m}=and \\ 
\glt `My leg is broken. I am confined to home.\footnote{Lit. I have fallen at home.}'
\z 
 
\ea \label{ŽM.10}
\textit{ā wextīyekey pīr bīyenā} \\ 
\gll ā wext-ī-ekey pīr bīye=nā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} old be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I have aged.'
\z 
 
\ea \label{ŽM.11}
\textit{čemēm korē bīyēnēw} \\ 
\gll čem-ē=m kor-ē bīyē=nē=ū \\ 
 eye\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} blind\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `My eyesight has deteriorated.\footnote{My eyes have gotten blind.} '
\z 
 
\ea \label{ŽM.12}
\textit{īne dāstānū minā} \\ 
\gll īne dāstāne=ū min=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{=ez.gen} \textsc{1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is my story.'
\z 
 
\ea \label{ŽM.13}
\textit{eger tūlū ʕemrīm temāmū dinyēm wiškinān pey žīwāyū dinyē} \\ 
\gll eger tūl=ū ʕemr-ī=m temām=ū dinyē=m wiškinā=n pey žīwāy=ū dinyē \\ 
 if lenght\textsc{.m}\textsc{=ez.gen} age\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} all\textsc{=ez.gen} world\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}} search\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to live\textsc{.inf}\textsc{=ez.gen} world\textsc{.obl}\textsc{.f} \\ 
\glt `Throughout my life, I have searched all around the world to make ends meet.'
\z 
 
\ea \label{ŽM.14}
\textit{dizīm nekerđenū hīzīm nekerdenū girđkār nebīyenā} \\ 
\gll dizī=m ne-kerđe=n=ū hīzī=m ne-kerde=n=ū girđkār ne-bīye=nā \\ 
 theft\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and adultery\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and know\_all \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I have not committed theft or adultery. I was not a know-all person.'
\z 
 
\ea \label{ŽM.15}
\textit{ā wextīyekey ečī newʕe řū be werū wēmre řū be xwā řisqēm peyā kerđen} \\ 
\gll ā wext-ī-ekey ečī newʕ=e řū be wer=ū wē=m=re řū be xwā řisq-ē=m peyā kerđe=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} like\_this kind/way\textsc{=dem} face by front\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} face by God\textsc{.m} living\textsc{-indf}\textsc{=\textsc{1sg}} visible do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I have taken refuge in God and found living.'
\z 
 
\ea \label{ŽM.16}
\textit{dārāw īse elʕān hāɫēhāzirī dārāw penj kuřānāw yerē kināčā, heštū wēmū ženē īne de neferē} \\ 
\gll dārā=w īse elʕān hāɫēhāzir-ī dārā=w penj kuř-ā=nā=ū yerē kināčē-ā hešt=ū wē=m=ū ženī-ē īne de nefer-ē \\ 
 owner\textsc{.m}\textsc{=ez.gen} now now at\_the\_moment\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{=ez.gen} five son/boy\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1sg}}=and three daughter\textsc{.f}\textsc{-obl}\textsc{.pl} eight=and \textsc{reflx}\textsc{=\textsc{1sg}}=and woman\textsc{.f}\textsc{-indf} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} ten person\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `I have, at the moment, five sons and three daughters, [which makes] eight, [plus] myself and a wife, [making us] ten people.'
\z 
 
\ea \label{ŽM.17}
\textit{tā īse yāwānā be řehmū xwāywe} \\ 
\gll tā īse yāwā=nā be řehm=ū xwā-ī=we \\ 
 until now arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} by mercy\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have made\footnote{Lit. arrived.} it so far, thanks to God's blessing.'
\z 
 
\ea \label{ŽM.18}
\textit{tā wē yāwānā} \\ 
\gll tā wē yāwā=nā \\ 
 until there arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I have made it so far.'
\z 
 
\ea \label{ŽM.20}
\textit{min zemānē zemāniš bīyen ženīm ārdēne girđū kārēw nebīyey bīyen} \\ 
\gll min zemān-ē zemān=iš bīye=n ženī=m ārdē=ne girđ=ū kār-ē=ū nebīyey bīye=n \\ 
 \textsc{1sg} time\textsc{.m}\textsc{-indf} time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} woman\textsc{.f}\textsc{=\textsc{1sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} all\textsc{=ez.gen} task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=ez.gen} poverty\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As for me, once I got married, there was poverty all around.'
\z 
 
\ea \label{ŽM.26}
\textit{ā wextīyekey ħewt no, yānze zāroɫēm bīyēnē} \\ 
\gll ā wext-ī-ekey ħewt no yānze zāroɫe-ē=m bīyē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} seven nine eleven child\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `I had seven or nine or eleven children.'
\z 
 
\ea \label{ŽM.27}
\textit{ī yānze zāroɫē ečē āsēnēm jīyā milū ī ženīme milū nebīyeywe} \\ 
\gll ī yānze zāroɫe-ē e=čē āsē=nē=m jīyā mil=ū ī ženī=m=e mil=ū nebīyey=we \\ 
 \textsc{dem.prox} eleven child\textsc{-dir}\textsc{.pl} in=here leave\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} separate on\textsc{=ez.gen} \textsc{dem.prox} woman\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} on\textsc{=ez.gen} poverty\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Because of destitution, [I had to] leave these eleven children to my wife.'
\z 
 
\ea \label{ŽM.28}
\textit{min ečāgene kārīger bīyenā šiš māngē xeberēm nezānān} \\ 
\gll min e=čāgene kārīger bīye=nā šiš mānge-ē xeber-ē=m ne-zānā=n \\ 
 \textsc{1sg} in=there labourer\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} six month\textsc{.f}\textsc{-dir}\textsc{.pl} news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` I was a worker there. I was unaware of them\footnote{Lit. I didn’t know their news.}  (my children) for six months,'
\z 
 
\ea \label{ŽM.29}
\textit{bizānū kewtī kewtēnē wārđīšā wārden dizīš dizīyen} \\ 
\gll bi-zān-ū kewtī kewtē=nē wārđī=šā wārde=n dizī=š dizīye=n \\ 
 \textsc{sbjv-}know\textsc{.prs}\textsc{-\textsc{1sg}} fall\textsc{.adv} fall\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} eat\textsc{.adv}\textsc{=3pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} steal\textsc{.adv}\textsc{=3sg} steal\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[I was not around to] witness if they (the children) had fallen, if they had eaten, or stolen [something].'
\z 
 
\ea \label{ŽM.30}
\textit{ā wextīyekey řūtē bīyēnē qūtē bīyēnē bīyenšā nebīyenšā xeberēm nezānēnē} \\ 
\gll ā wext-ī-ekey řūt-ē bīyē=nē qūt-ē bīyē=nē bīye=n=šā ne-bīye=n=šā xeber-ē=m ne-zānē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} naked\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} bare\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} news\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `[I was unaware if] they were without clothes; if they had [enough]; if they didn’t have [enough]. I didn't know their news.'
\z 
 
\ea \label{ŽM.47}
\textit{penj řoē čāne min řemezānim gērten} \\ 
\gll penj řo-ē č=ā=ne min řemezān=im gērte=n \\ 
 five day\textsc{.m}\textsc{-indf} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{1sg} \textsc{pn}\textsc{=\textsc{1sg}} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[while there] I fasted five days during Ramadan.'
\z 
 
\ea \label{ŽM.49}
\textit{řoē sī penj timenē šesū yerē sāɫē čēweɫī kerū ētir meyānū sāɫū činē bē} \\ 
\gll řo-ē sī penj timen-ē šes=ū yerē sāɫe-ē čēweɫī ∅-ker-ū ētir me-zān-ū sāɫe-ū činē b-ē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} sixty\textsc{=ez.gen} three year\textsc{.f}\textsc{-dir}\textsc{.pl} ago \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc}.\textsc{ptcl} \textsc{neg-}know\textsc{.prs}\textsc{-\textsc{1sg}} year\textsc{.f}\textsc{-\textsc{1sg}} how\_many\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `A daily salary of thirty-five tomans, it was sixty-three years ago. I don’t know which year it was/'
\z 
 
\ea \label{ŽM.53}
\textit{ʕerzū birāw wēm kerū āmēnmēwe ī mentēqe} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū āmē=nmē=we ī mentēqe \\ 
 petition\textsc{.m}\textsc{=ez.gen} brother\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \\ 
\glt `Let me tell you. I returned to this region.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
