\chapter{All 244 examples of 2sg in these texts are found in the following sentences:}

\ea \label{ZB.55}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{ZB.60}
\textit{welē īse be qurʕān ne šēx menen, ne meɫā menen, ne kesē hen be weš dāđiš berī lā} \\ 
\gll welē īse be qurʕān ne šēx mene=n ne meɫā mene=n ne kes-ē hen-∅ be weš dāđ=iš ∅-ber-ī lā \\ 
 but now to \textsc{pn} neither sheikh\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} nor mullah\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} nor person\textsc{.m}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} to good yell\textsc{.m}\textsc{=3sg} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} at\_the\_place\_of \\ 
\glt `However, nowadays, I swear on the Quran, neither a [true] sheikh nor a [true] mullah remains, nor is there anyone one can take counsel with.'
\z 
 
\ea \label{ZB.61}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 



\ea \label{ZQ.1}
\textit{ʕerzū xizmetīt kerū} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū \\ 
 petition\textsc{.m}\textsc{=ez.gen} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `Let me tell you!\footnote{Lit. May a petition be for your service.}'
\z 
 
\ea \label{ZQ.7}
\textit{ʕerziš kerđ, qurbān bā dāstānēt pey gēɫnūwe, dāstāne} \\ 
\gll ʕerz=iš kerđ qurbān ∅-b-ā dāstāne-ē=t pey ∅-gēɫn-ū=we dāstāne \\ 
 petition\textsc{.m}\textsc{=3sg} do\textsc{.pst} sir.\textsc{voc} \textsc{ind-}be.prs\textsc{-3pl} story.f\textsc{-indf}\textsc{=\textsc{2sg}} to \textsc{sbjv-}lead.prs\textsc{-\textsc{1sg}}\textsc{=compl} story.f \\ 
\glt `He said, ‘Sir, Let me tell you a story, the story!’'
\z 
 
\ea \label{ZQ.8}
\textit{fermāwāš, dā dey řoɫe čēš māčī’} \\ 
\gll fermāwā=š dā dey řoɫe čēš m-āč-ī \\ 
 say\textsc{.pst}\textsc{=3sg} give\textsc{.pst} \textsc{disc.ptcl} child.\textsc{voc} what \textsc{ind-}say.prs-\textsc{2sg} \\ 
\glt `The Shah said, ‘Go on son, What is it ?\footnote{Lit. What are you saying?}’'
\z 
 
\ea \label{ZQ.16}
\textit{duē, be gīrmā wārd čenū ženīū čenū heywānīū dewɫetīū bē āwīū nāřihet e pāwe} \\ 
\gll duē be gīr=mā wārd čenū ženī=ū čenū heywān-ī=ū dewɫet-ī=ū bē āwī=ū nāřihet e=pāwe \\ 
 two by hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} woman.f=and with\textsc{.ez.gen} animal\textsc{.m}\textsc{-obl}\textsc{.m}=and animal\textsc{-obl}\textsc{.m}=and come\textsc{.imp}\textsc{.\textsc{2sg}} water.f=and sad in=there \\ 
\glt `I was stuck there with my wife, livestock, a lack of water, and a lack of comfort. '
\z 
 
\ea \label{ZQ.20}
\textit{her wāčēnī bečkēwen gīyāniš čenīn her āne!} \\ 
\gll her wāč-ēn-ī bečke-ēwe=n gīyān=iš čenī=n her āne \\ 
 just say.prs\textsc{-pstc}-\textsc{2sg} baby\textsc{-indf}\textsc{=cop}\textsc{.3sg}\textsc{.m} soul\textsc{.m}\textsc{=3sg} in\textsc{=cop}\textsc{.3sg}\textsc{.m} just that\_much \\ 
\glt `One would say it’s a child who is just alive, just that!'
\z 
 
\ea \label{ZQ.51}
\textit{vātšā, čēš māčī?’} \\ 
\gll vāt=šā čēš m-āč-ī \\ 
 say\textsc{.pst}\textsc{=3pl} what \textsc{ind-}say.prs-\textsc{2sg} \\ 
\glt `They said, ‘Are you serious?\footnote{Lit. What are you saying?}’ '
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{HB.6}
\textit{hezretū šēxīč fermāwo, dey konī?} \\ 
\gll hezret=ū šēx=yč ∅-fermāw-o dey ko=nī \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} where\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `His highness the Sheikh said, ‘Where are you?’'
\z 
 
\ea \label{HB.15}
\textit{kābrāyč herbenew hereyn, milo herey gurj kerowew gino řā,} \\ 
\gll kābrā=īč herbene=ū her-e-ī=n mi-l-o her-e-ī gurj ∅-ker-o=we=ū ∅-gin-o řā \\ 
 man\textsc{=add} donkey\_breeder\textsc{.m}=and donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} alert \textsc{ind-}do\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}}\textsc{=ez}\textsc{.gen} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} road\textsc{\textsc{.f}} \\ 
\glt `The fellow who was a donkey breeder, went and prepared the donkey, [and] set off.'
\z 
 
\ea \label{HB.17}
\textit{miđyo herzretū šēxī nefermāwā bilue yānew āney bāre’} \\ 
\gll mi-đye-o herzret=ū šēx-ī ne-fermāwā bi-lue yāne=ū āney b-ār-e \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}say\textsc{.pst} \textsc{sbjv-}go\textsc{.imp}\textsc{.\textsc{2sg}} house=and \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `He noticed that his highness the Sheikh hadn’t told [him], ‘Go to the house of that (i.e. such-and-such) person and bring [back the flour]’'
\z 
 
\ea \label{HB.18}
\textit{yā zeřiš neyā pene, lue čiko bisāne bāre!’} \\ 
\gll yā zeř=iš ne-dā pene ∅-lu-e či=ko bi-sān-e b-ār-e \\ 
 or money\textsc{.m}\textsc{=3sg} \textsc{neg-}give\textsc{.pst} to \textsc{imp-}go.\textsc{imp-}\textsc{2sg} from=where \textsc{imp-}buy\textsc{.prs}-\textsc{2sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `Nor had he given him money, [saying], ‘Go [and] buy from a place\footnote{Lit. from where.} [and] bring [back the flour]’'
\z 
 
\ea \label{HB.31}
\textit{hereke āmāwe zuwāb wātiš, xefetit nebo!} \\ 
\gll her-eke āmā=we zuwāb wāt=iš xefet=it ne-b-o \\ 
 donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} come\textsc{.pst}\textsc{.3sg}\textsc{=compl} reply\textsc{.m} say\textsc{.pst}\textsc{=3sg} sorrow\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The donkey started to speak [and] said, ‘Don’t worry!'
\z 
 
\ea \label{HB.38}
\textit{toyč berā yāne} \\ 
\gll to=īč ∅-ber-ā yāne \\ 
 \textsc{2sg}\textsc{=add} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} house \\ 
\glt `they will take you to [their] house;'
\z 
 
\ea \label{HB.39}
\textit{hēɫeřuānīt pey kerā} \\ 
\gll hēɫeřuānī=t pey ∅-ker-ā \\ 
 fried\_egg\textsc{\textsc{.f}}\textsc{=\textsc{2sg}} for \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \\ 
\glt `they will cook fried eggs for you;'
\z 
 
\ea \label{HB.40}
\textit{nānīč miđā to} \\ 
\gll nān=yč mi-đe-ā to \\ 
 bread\textsc{.m}\textsc{=add} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{2sg} \\ 
\glt `they will give you a meal;'
\z 
 
\ea \label{HB.41}
\textit{toīč nān werī} \\ 
\gll to=yč nān ∅-wer-ī \\ 
 \textsc{2sg}\textsc{=add} bread\textsc{.m} \textsc{ind-}out\textsc{-obl}\textsc{.m} \\ 
\glt `[and] you will eat your meal.'
\z 
 
\ea \label{HB.43}
\textit{pekemā megino, to goš min dāre!’} \\ 
\gll peke=mā me-gin-o to goš min ∅-dār-e \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \textsc{2sg} ear\textsc{.m} 1sg \textsc{imp-}hold\textsc{.prs}-\textsc{2sg} \\ 
\glt `We should not be worried; listen to me!’'
\z 
 
\ea \label{HB.54}
\textit{pāse herey wāt, simerīš kerdē torekekew āđī} \\ 
\gll pāse her-e-ī wāt simerī=š kerdē-ē toreke-eke=ū āđī \\ 
 like donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} say\textsc{.pst} straw\textsc{.dir}\textsc{.pl}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-dir}\textsc{.pl} sack\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}=and 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `As the donkey said, he (the man) put straw into his (i.e. donkey’s) sack.'
\z 
 
\ea \label{HB.62}
\textit{vātiš, dey xuře vene bere} \\ 
\gll vāt=iš dey ∅-xuř-e vene ∅-ber-e \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}shout\textsc{.prs}-\textsc{2sg} \textsc{post} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `[and] said, ‘Command the donkey\footnote{Lit. shout at him.}; move him ; take him;'
\z 
 
\ea \label{HB.63}
\textit{bere’} \\ 
\gll ∅-ber-e \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt ` take him. '
\z 
 
\ea \label{HB.72}
\textit{milowew bārū herey wizorew} \\ 
\gll mi-l-o=we=ū bār=ū her-e-ī ∅-wiz-o=re=ū \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}}=and load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} \textsc{ind-}throw\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}}=and \\ 
\glt `He went back and put down the donkey’s load.'
\z 
 
\ea \label{HB.80}
\textit{čiřoš hezretū šēxī fermāwo, erē řoɫem čī toryānī?} \\ 
\gll ∅-čiř-o=š hezret=ū šēx=ī ∅-fermāw-o erē řoɫe=m čī toryā=nī? \\ 
 \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} child\textsc{.m}\textsc{=1sg} why get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt ` His highness the Sheikh summoned him, i.e. the donkey breeder, [and] said, ‘Oh son! why were you offended? '
\z 
 
\ea \label{HB.81}
\textit{čī ʕājiz bīyenī?’} \\ 
\gll čī ʕājiz bīye=nī \\ 
 why upset be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Why did you get annoyed?’'
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{HB.86}
\textit{īse kē be zūwānū herī zāno bējge to?} \\ 
\gll īse kē be zūwān=ū her-ī ∅-zān-o bējge to \\ 
 now who by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `Now who can understand the language of donkeys except you?'
\z 
 
\ea \label{HB.87}
\textit{īse bējgem to kē be zuwānū herī zāno?} \\ 
\gll īse bējgem to kē be zuwān=ū her-ī ∅-zān-o \\ 
 now except\_for \textsc{2sg} who by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `Now, except for you, who can understand the language of donkeys?'
\z 
 
\ea \label{HB.88}
\textit{kē be zuwānū herī zāno bejgem to} \\ 
\gll kē be zuwān=ū her-ī ∅-zān-o bejgem to \\ 
 who by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `Who can understand the language of donkeys except you?'
\z 
 
\ea \label{HB.89}
\textit{ētir to čī ādiz bīyenī?} \\ 
\gll ētir to čī ādiz bīye=nī \\ 
 \textsc{disc.ptcl} \textsc{2sg} why upset be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Then why have you become upset?'
\z 
 
\ea \label{HB.90}
\textit{dā dey hurbēze ī mexlūqī girdiš gēɫeš pone} \\ 
\gll dā dey hur-b-ēz-e ī mexlūq-ī gird=iš gēɫ-e=š pone \\ 
 \textsc{hort} \textsc{disc.ptcl} \textsc{pvb-}\textsc{sbjv-}rise\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} people\textsc{-obl}\textsc{.m} all\textsc{=3sg} wander\textsc{.prs}-\textsc{2sg}\textsc{=3sg} at \\ 
\glt `Come on, get up and search among all these people,'
\z 
 
\ea \label{HB.91}
\textit{bizāne neferē ter hen be zūwānū herī hāɫīš bo bejge to!} \\ 
\gll bi-zān-e nefer-ē ter hen-∅ be zūwān=ū her-ī hāɫī=š ∅-b-o bejge to \\ 
 \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} person\textsc{.m}\textsc{-dir}\textsc{.pl} other exist\textsc{-3sg}\textsc{.m} by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} understood\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `see if there is a person, apart from you, who can understand the language of donkeys!'
\z 
 
\ea \label{HB.92}
\textit{čīm hāɫī nekerdenī!} \\ 
\gll čī=m hāɫī ne-kerde=nī \\ 
 why\textsc{=1sg} understood \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Why haven’t I made you understand?'
\z 
 
\ea \label{HB.93}
\textit{hāɫīm nekerdenī řās, ānīšā} \\ 
\gll hāɫī=m ne-kerde=nī řās ānīšā \\ 
 understood\textsc{=1sg} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} really 3pl.dem\textsc{.obl} \\ 
\glt `Haven’t I really made you understand? and so on.'
\z 


\ea \label{PM.1}
\textit{ʕerzū xizmetīt kerū hetā min žinīyenim pāse mešhūr bīyen šewē šā ʕelādīnī, šewē peɫē merekuř peɫē merekuř mē šārezūr} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū hetā min žinīye=n=im pāse mešhūr bīye=n šewe-ē šā ʕelādīn-ī šewe-ē peɫ-ē merekuř peɫ-ē merekuř m-ē šārezūr \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} like famous be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} night.f\textsc{-obl}.f king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} night.f\textsc{-obl}.f \textsc{clf}\textsc{-indf} grasshopper \textsc{clf}\textsc{-indf} grasshopper \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{pn} \\ 
\glt `Let me tell you .\footnote{Lit. May a petition be for your service.} I have even heard (this). It is well known that a swarm of grasshoppers came to Sharazour one night.'
\z 
 
\ea \label{PM.8}
\textit{fermāwo, lue xizmetū āđī!} \\ 
\gll ∅-fermāw-o ∅-lu-e xizmet=ū āđī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `[and] said, ‘Go into his presence!'
\z 
 
\ea \label{PM.9}
\textit{ījāze bizāne āđ ījāze miđo ēme e čē nehār kermē yām ne} \\ 
\gll ījāze bi-zān-e āđ ījāze mi-đe-o ēme e=čē nehār ∅-ker-mē yām ne \\ 
 permission \textsc{imp-}know\textsc{.prs}-\textsc{2sg} 3sg\textsc{.dir}\textsc{.m} permission \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \textsc{1pl} in=here lunch\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} or no \\ 
\glt `See if he lets us stay here for lunch or not'
\z 
 
\ea \label{PM.13}
\textit{kīyāsenāšā xizmetū to} \\ 
\gll kīyāse=nā=šā xizmet=ū to \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=3pl} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} \\ 
\glt `I have been sent to your service.'
\z 
 
\ea \label{PM.14}
\textit{eger to ījāze miđey ēme e čē nehārē kermē ījāze be desū ton} \\ 
\gll eger to ījāze mi-đe-ī ēme e=čē nehār-ē ∅-ker-mē ījāze be des=ū to=n \\ 
 if \textsc{2sg} permission \textsc{ind-}give\textsc{.prs}-\textsc{2sg} \textsc{1pl} in=here lunch\textsc{.m}\textsc{-indf} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} permission to hand\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `It is in your hands whether to let us stay here for lunch.'
\z 
 
\ea \label{PM.15}
\textit{āmēnmē xizmetū to ījāzemā biđey} \\ 
\gll āmē=nmē xizmet=ū to ījāze=mā bi-đe-ī \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} permission\textsc{=1pl} \textsc{sbjv-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `I\footnote{Lit. We.} have come to your service so that you might permit us.'
\z 
 
\ea \label{PM.29}
\textit{māčo qurbān īne čēšit kerdū?} \\ 
\gll m-āč-o qurbān īne čēš=it kerd=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=\textsc{2sg}} do\textsc{.pst}=and \\ 
\glt `They said, ‘Sir, what have you done?'
\z 
 
\ea \label{PM.30}
\textit{īne bē ījāzew to īne nekiryān} \\ 
\gll īne b-ē ījāze=ū to īne ne-kir-yā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} permission\textsc{\textsc{=ez.gen}} \textsc{2sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{neg-}do.pass-past.pass\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This couldn’t have been done without your permission.'
\z 
 
\ea \label{PM.31}
\textit{tenhā hinū ton, geɫēwew kesī zereriš neginān je šārezūrne} \\ 
\gll tenhā hin=ū to=n geɫā-ēwe=ū kes-ī zerer=iš ne-ginā=n je šārezūr=ne \\ 
 not\_even ez.pron\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} leaf\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} person\textsc{.m}\textsc{-obl}\textsc{.m} harm\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} \\ 
\glt `It’s all yours [otherwise] no one’s leaves have been harmed in Sharazour. '
\z 
 
\ea \label{PM.33}
\textit{fermāwo, nāřihet nebī!} \\ 
\gll ∅-fermāw-o nāřihet ne-b-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘May you not be worried!''
\z 
 
\ea \label{PM.34}
\textit{nāřihet nebī āne meʕmūre bē terefū xwāywe āmābē} \\ 
\gll nāřihet ne-b-ī āne meʕmūre b-ē teref=ū xwā-ī=we āmā b-ē \\ 
 sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} agent be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Don’t worry! It was an agent [who] was sent  by God.'
\z 
 
\ea \label{PM.38}
\textit{girđū šārezūrī ēsāɫ merāʕetū ēme kero je pārī zīyāter pey toyčū pey ēmeyč,} \\ 
\gll girđ=ū šārezūr-ī ēsāɫ merāʕet=ū ēme ∅-ker-o je pār-ī zīyā-ter pey to=yč=ū pey ēme=yč \\ 
 all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} this\_year care\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} from last\_year\textsc{-obl}\textsc{.m} increased\textsc{-cmpr} for \textsc{2sg}\textsc{=add}=and for \textsc{1pl}\textsc{=add} \\ 
\glt `This year, all the people in Sharazour care for us— both you and I—, more than the previous year.'
\z 
 
\ea \label{PM.40}
\textit{bilowe xemit nebo} \\ 
\gll bi-l-o=we xem=it ne-b-o \\ 
 \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} sorrow\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Go back; don’t worry’'
\z 
 
\ea \label{PM.53}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 

 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.20}
\textit{hāɫīt bo?} \\ 
\gll hāɫī=t b-o \\ 
 understood\textsc{=\textsc{2sg}} be\textsc{.prs}\textsc{-3sg} \\ 
\glt `You see?'
\z 
 
\ea \label{DG.26}
\textit{to mišo ī birāyte šewēne bere} \\ 
\gll to mišo ī birā-ī=t=e šewe-ē=ne ∅-ber-e \\ 
 \textsc{2sg} aux \textsc{dem.prox} brother\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `You should take this brother of yours [there] at night.'
\z 
 
\ea \label{DG.27}
\textit{lue wēt wize milū pāɫāyekā šā ʕelādīnīre} \\ 
\gll ∅-lu-e wē=t ∅-wiz-e mil=ū pāɫā-ekā šā ʕelādīn-ī=re \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}throw\textsc{.prs}-\textsc{2sg} on\textsc{=ez}\textsc{.gen} shoe\textsc{-def}\textsc{.obl}\textsc{.pl} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Go, throw yourself at the feet\footnote{Lit. shoes.} of Shah Aladin'
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DG.44}
\textit{īđīč fermāwo, lue čā nīšere,} \\ 
\gll īđ=īč ∅-fermāw-o ∅-lu-e čā ∅-nīš-e=re \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} there \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \\ 
\glt `Shah Aladin  said, ‘Go [and] sit there;'
\z 
 
\ea \label{DG.45}
\textit{lue xāneqā tā xeberit dewwe pene’} \\ 
\gll ∅-lu-e xāneqā tā xeber=it ∅-de-ū=we pene \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} monastery\textsc{.m} until news\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}give\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{compl}} to \\ 
\glt `Go to the monastery until I let you know.'
\z 
 
\ea \label{DG.47}
\textit{winū kīseɫē sāwī be} \\ 
\gll winī=ū kīseɫī-ē ∅-sāw-ī be \\ 
 blood\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} tortoise\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{sbjv-}rub\textsc{.prs}-\textsc{2sg} to \\ 
\glt `[The Sheikh said to the man], ‘You may rub tortoise’s blood on …'
\z 
 
\ea \label{DG.49}
\textit{winū kīseɫē zāmdāre kere} \\ 
\gll winī=ū kīseɫī-ē zāmdār-e ∅-ker-e \\ 
 blood\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} tortoise\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} wounded\textsc{-f} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Make a tortoise wounded!'
\z 
 
\ea \label{DG.50}
\textit{winekēš sāwe pā qelbū pīyāyre} \\ 
\gll winī-ekē=š ∅-sāw-e p=ā qelb=ū pīyā-ī=re \\ 
 blood\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} by\textsc{=dem}.dist heart\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `[then] rub its blood on the heart of the [sick] man'
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.57}
\textit{lue xāneqā nān bere, bideš penew} \\ 
\gll ∅-lu-e xāneqā nān ∅-ber-e bi-de=š pene=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} monastery\textsc{.m} bread\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}give\textsc{.prs}\textsc{=3sg} to=and \\ 
\glt `Go to the monastery; take food [and] give it to him.'
\z 
 
\ea \label{DG.58}
\textit{ā wextīyekey, gēɫdē lodē penā bere be xwāy} \\ 
\gll ā wext-ī-ekey ∅-gēɫ-dē ∅-lo-dē penā ∅-ber-e be xwā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}wander\textsc{.prs}\textsc{-2pl} \textsc{imp-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} refuge \textsc{imp-}take\textsc{.prs}-\textsc{2sg} by God\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, go back [to your place]; Take refuge with God.'
\z 
 
\ea \label{DG.66}
\textit{ē.. dewāyekeš sāwe pore dā ēme kečē nebīmē} \\ 
\gll ē.. dewāī-eke=š ∅-sāw-e pore dā ēme keč-ē ne-b-īmē \\ 
 eh medication\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} at \textsc{hort} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs}-\textsc{1pl} \\ 
\glt `He shall rub the medicine on him; he shall not make us crooked.'
\z 
 
\ea \label{DG.68}
\textit{berešo memliketū wētā welē sermā deydē} \\ 
\gll ∅-ber-e=š=o memliket=ū wē=tā welē ser=mā ∅-de-ydē \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=\textsc{compl}} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=2pl} but head\textsc{.m}\textsc{=1pl} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} \\ 
\glt `Take him to your region, but pay us a visit.'
\z 
 
\ea \label{DG.72}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{DP.6}
\textit{hāḷītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.23}
\textit{min īney dew deso towe ey wēm kowe bilū} \\ 
\gll min īney ∅-de-ū des=o to=we ey wē=m ko=we bi-l-w \\ 
 \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}give\textsc{.prs}\textsc{-1sg} hand\textsc{.m}\textsc{=\textsc{post}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{intj} \textsc{reflx}\textsc{=1sg} where\textsc{=\textsc{post}} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} \\ 
\glt `If I hand over these to you, where should I go myself?'
\z 
 
\ea \label{DP.24}
\textit{to luew čiko āmānī luewewe} \\ 
\gll to lue=ū či=ko āmā=nī lue=wewe \\ 
 \textsc{2sg} go\textsc{.imp}\textsc{.\textsc{2sg}}=and from=where come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} go\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `Go back to where you came from, go back!'
\z 
 
\ea \label{DP.39}
\textit{bē bā nīšmēre seḷāhū wēmā kermē} \\ 
\gll bē bā ∅-nīš-mē=re seḷāh=ū wē=mā ∅-ker-mē \\ 
 come\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{hort} \textsc{sbjv-}sit\textsc{.prs}\textsc{-1pl}\textsc{=\textsc{povb}} solution\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `[He said to Sibhan Agha], ‘Let us sit together and come to a solution.’'
\z 
 
\ea \label{DP.40}
\textit{ewīč nimē māčo, nexeyr nimewnew mitmānim nīyā penet.} \\ 
\gll ew=yč ni-mi-ē mi-āč-o nexeyr ni-mi-e-wne=ū mitmān=im nīye=ā pene=t \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-1sg}=and trust\textsc{.m}\textsc{=1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} to\textsc{=\textsc{2sg}} \\ 
\glt ` Sibhan Agha\footnote{Lit. He.} did not go [to Jamsher Shah], saying, ‘No, I won’t come [to you]; I don’t trust you.’'
\z 
 
\ea \label{DP.42}
\textit{māčo, ne bē!’} \\ 
\gll mi-āč-o ne ∅-bē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{imp-}come\textsc{.imp}\textsc{.\textsc{2sg}} \\ 
\glt `Jamsher Shah\footnote{Lit. He} said, ‘No [I won’t arrest you]. Come!’'
\z 
 
\ea \label{DP.43}
\textit{lue wāče bē!’} \\ 
\gll ∅-lue ∅-wāč-e ∅-bē \\ 
 \textsc{imp-}go\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}say\textsc{.prs-}\textsc{2sg} \textsc{imp-}come\textsc{.imp}\textsc{.\textsc{2sg}} \\ 
\glt `[He said to his men], ‘Go [and] tell him to come!’'
\z 
 
\ea \label{DP.50}
\textit{ānen mirīčēɫēm wer dān, to xeyrū min siɫāmet} \\ 
\gll āne=n mirīčēɫe-ē=m wer dā=n to xeyr=ū min siɫāmet \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} small\_bird\textsc{.m}\textsc{-indf}\textsc{=1sg} out give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `’I have let the bird go away. Now, you follow your life, and I will follow mine!\footnote{Lit. Well-being for you and health for me.}’'
\z 
 
\ea \label{DP.55}
\textit{to xeyrū min siɫāmet} \\ 
\gll to xeyr=ū min siɫāmet \\ 
 \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `Follow your own life, and I will follow mine.'
\z 
 
\ea \label{ZP.33}
\textit{ī kināčēw to dermānū derdīš īnā lā ī pīyāywe’} \\ 
\gll ī kināčē=ū to dermān=ū derd-ī=š īnā-∅ lā ī pīyā-ī=we \\ 
 dem.prox daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} \textsc{2sg} treatment\textsc{.m}\textsc{\textsc{=ez.gen}} illness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \\ 
\glt `‘This daughter of yours, the treatment for her illness lies with this man (i.e., Pir Shaliyar)'
\z 
 
\ea \label{ZP.34}
\textit{meger berī kīyānīš pey hewrāmānī} \\ 
\gll meger ∅-ber-ī ∅-kīyān-ī=š pey hewrāmān-ī \\ 
 if\_only \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}send\textsc{.prs}-\textsc{2sg}\textsc{=3sg} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Maybe if you take her to him.'
\z 
 
\ea \label{ZP.39}
\textit{māčo, ī kināčē bere degā be degāw} \\ 
\gll m-āč-o ī kināčē ∅-ber-e degā be degā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.obl}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `[and] said, ‘Take this girl village by village,'
\z 
 
\ea \label{ZP.44}
\textit{to eger ī kināčē berdet xizmetū ī pīyāy} \\ 
\gll to eger ī kināčē berd-e=t xizmet=ū ī pīyā-ī \\ 
 \textsc{2sg} if dem.prox daughter\textsc{.dir}\textsc{.f} take\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=\textsc{2sg}} service\textsc{.m}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `‘When you have taken this girl into the presence of this man'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.100}
\textit{ī kināčē eger weše bīyowe mišyo māreš biřū pey to be mezebū wētāw} \\ 
\gll ī kināčē eger weš-e bīyē-o=we mišyo māre=š ∅-biř-ū pey to be mezeb=ū wē=tā=ū \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} if well\textsc{-f} be\textsc{.prs}.\textsc{cond}\textsc{.f}?\textsc{-3sg}\textsc{=compl} aux marraige\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for \textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `If the girl is healed, I shall marry her to you, according to your religion.'
\z 
 
\ea \label{ZP.106}
\textit{berešo. mubārekū sāhēbīš bo!’} \\ 
\gll ∅-ber-e=š=o mubārek=ū sāhēb-ī=š ∅-b-o \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=compl} blessed\textsc{\textsc{=ez.gen}} owner\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take her back. May she be happy with her owner (i.e., father)!'
\z 
 
\ea \label{ZP.116}
\textit{māčo, meberūšo mišyo to māreš kerīnew”} \\ 
\gll m-āč-o me-ber-ū=š=o mišyo to māre=š ∅-ker-īne=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} aux \textsc{2sg} marraige\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs} \textsc{2sg}=and \\ 
\glt `He says, “I will not take her back; you shall marry her.”'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.119}
\textit{yoīč māčo, dey minīč beʕzē pīyāzīt midonew’} \\ 
\gll yo=īč m-āč-o dey min=īč beʕzē pīyāz-ī=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg}\textsc{=add} some onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Another said, ‘I will give you some onions.’'
\z 
 
\ea \label{ZP.120}
\textit{yoīč māčo, minīč beʕzē tiršū henārāt midonew’} \\ 
\gll yo=īč m-āč-o min=īč beʕzē tirš=ū henār-ā=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} some molasses\textsc{\textsc{=ez.gen}} pomegranate\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt ` Another said, ‘I will give you some pomegranate molasses.’'
\z 
 
\ea \label{ZP.121}
\textit{kābrāyēč māčo, dey birā, minīč serē heywānit midew} \\ 
\gll kābrā-ē=č m-āč-o dey birā min=īč ser-ē heywān=it mi-de-ū \\ 
 man\textsc{-indf}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{1sg}\textsc{=add} \textsc{clf}\textsc{-indf} animal\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `A man said, ‘Brother, I will give you an animal.’'
\z 
 
\ea \label{ŽP.18}
\textit{kīyānāš māčā, heywānekey bere} \\ 
\gll ∅-kīyān-ā=š m-āč-ā heywān-ekey ∅-ber-e \\ 
 \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} animal\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `They (his uncle’s family) sent him [shepherding]. They said, ‘Take the animals ….'
\z 
 
\ea \label{ŽP.21}
\textit{heywānekey bere pey ewā} \\ 
\gll heywān-ekey ∅-ber-e pey ewā \\ 
 animal\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to there \\ 
\glt `Take the animals there [the valley of Pir Shaliyar] '
\z 
 
\ea \label{ŽP.24}
\textit{milo mēweū māčā, hizmēč bāre’} \\ 
\gll mi-l-o m-ē=we=ū m-āč-ā hizmē=č b-ār-e \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl}=and \textsc{ind-}say\textsc{.prs}\textsc{-3pl} firewood\textsc{.f}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt ` He went [shepherding and] came back [home]. They said, ‘Bring firewood as well.’'
\z 
 
\ea \label{ŽP.27}
\textit{māčā, dey to metāwī -sey misefāw- kučēš yewē berī nehāɫū pīr šelīyārā šānīšā pižgīnīšā bā bizyā’} \\ 
\gll m-āč-ā dey to me-tāw-ī sey misefā=ū kuč-ē=š yewe-ē ∅-ber-ī nehāɫ=ū pīr šelīyār-ā ∅-šān-ī=šā ∅-pižgīn-ī=šā bā bi-zye-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{2sg} \textsc{neg-}can\textsc{.prs}-\textsc{2sg} \textsc{pn} \textsc{pn}=and little\textsc{-indf}\textsc{=3sg} barley\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \textsc{sbjv-}throw\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}scatter\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \textsc{hort} \textsc{sbjv-}grow\textsc{.prs}\textsc{-3pl} \\ 
\glt `They said, ‘You are not able, Say Mustafa, to take some barley seeds to the valley of Pir Shaliyar, [and] scatter them [on the field], [and] plant them, so that they grow.'
\z 
 
\ea \label{ŽP.39}
\textit{milo mēwe māčā, šānāt tomeke?’} \\ 
\gll mi-l-o m-ē=we m-āč-ā šānā=t tom eke \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} scatter\textsc{.pst}\textsc{=\textsc{2sg}} \\ 
\glt `He would go [to the field] and come back. They would say, ‘Did you plant the seeds?’'
\z 
 
\ea \label{ŽP.43}
\textit{māčo bē dengē māmo yoso, dā lue šonū sey misefāyre bizāne řās māčo!} \\ 
\gll m-āč-o bē- deng-ē māmo yoso dā ∅-lu-e šūn=ū sey misefā-ī=re bi-zān-e řās m-āč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}without noise\textsc{.m}\textsc{-indf} paternal\_uncle \textsc{pn} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} after\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} truth \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \\ 
\glt `Uncle Yoso secretly said [to the servant], ‘Go after Say Mustafa, see if he is telling the truth!'
\z 
 
\ea \label{ŽP.60}
\textit{māčo, erē řās māčī.’} \\ 
\gll m-āč-o erē řās m-āč-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes truth \textsc{ind-}say\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘Yes, you are telling the truth.’'
\z 
 
\ea \label{ŽP.69}
\textit{āne je zemīnī pēse wāčī xwāy ketē pey kīyāsen} \\ 
\gll āne je zemīn-ī pēse ∅-wāč-ī xwā-ī ket-ē pey kīyāse=n \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} from earth\textsc{.m}\textsc{-obl}\textsc{.m} as\_if \textsc{sbjv-}say\textsc{.prs-}\textsc{2sg} God\textsc{.m}\textsc{-obl}\textsc{.m} bed\textsc{-indf} to send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[He had levitated] this much from the ground, as if God had sent him a bed.'
\z 
 
\ea \label{ŽP.91}
\textit{fermāwo hezretū xos, dey to mišyo tešrīf berī bilīwe, pey menteqew wētāw} \\ 
\gll ∅-fermāw-o hezret=ū xos dey to mišyo tešrīf ∅-ber-ī bi-l-ī=we pey menteqe=ū wē=tā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{disc.ptcl} \textsc{2sg} aux honouring \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} to region=and \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `His Highness Khos said, ‘You should leave [here and] return to your region.'
\z 
 
\ea \label{ŽP.92}
\textit{bilī čāy xizmetū mirīđū mensūb weš kerīw} \\ 
\gll bi-l-ī čā-ī xizmet=ū mirīđ=ū mensūb weš ∅-ker-ī=ū \\ 
 \textsc{sbjv-}go\textsc{.prs}-\textsc{2sg} there\textsc{-obl}\textsc{.m} service\textsc{.m}\textsc{\textsc{=ez.gen}} adherent=and appointees well \textsc{sbjv-}do\textsc{.prs}\textsc{-obl}\textsc{.m}=and \\ 
\glt `You should go there and be of good service to adherents and appointees, '
\z 
 
\ea \label{ŽP.93}
\textit{bilīwe pey ā menteqew wētāw, šerʕī ders wāčīw} \\ 
\gll bi-l-ī=we pey ā menteqe=ū wē=tā=ū šerʕī ders ∅-wāč-ī=ū \\ 
 \textsc{sbjv-}\textsc{-obl}\textsc{.m}\textsc{=compl} to \textsc{dem.dist} region=and \textsc{reflx}=\textsc{2pl}=and sharia lesson\textsc{.m} \textsc{sbjv-}say\textsc{.prs-}\textsc{2sg}=and \\ 
\glt `Go to your region [and] teach Sharia law'
\z 
 
\ea \label{ŽP.94}
\textit{hem šerēʕetū hem terēqet to āmādenī tekmēɫnī} \\ 
\gll hem šerēʕet=ū hem terēqet to āmāde=nī tekmēɫ=nī \\ 
 both sharia\_law=and and denomination \textsc{2sg} ready\textsc{=cop}\textsc{.\textsc{2sg}} complete\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `You are ready both [to lead] an order (of dervishes) and [to practise] Sharia law. You are complete.'
\z 
 
\ea \label{ŽP.95}
\textit{luewe} \\ 
\gll ∅-lu-e=we \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `Go back!’'
\z 
 
\ea \label{ŽP.103}
\textit{to lue zeřekey bere čenū wēt pey xerjīt lodē} \\ 
\gll to ∅-lu-e zeř-ekey ∅-ber-e čenū wē=t pey xerjī=t ∅-lo-dē \\ 
 \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} with\textsc{.ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} to consumption\textsc{=\textsc{2sg}} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} \\ 
\glt `You go, take the money with you for your expenses.'
\z 
 
\ea \label{ŽP.104}
\textit{zeřekey bere min, qey mekero} \\ 
\gll zeř-ekey ∅-ber-e min qey me-ker-o \\ 
 money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{1sg} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take the money. I will be fine.’'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.106}
\textit{fermāwo, nā nā meger berīš’} \\ 
\gll ∅-fermāw-o nā nā meger ∅-ber-ī=š \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no no must \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg} \\ 
\glt `He (the man from Hajij) said, ‘No, you must take it.’'
\z 
 
\ea \label{ŽP.116}
\textit{mēwe ētir ā zemāne pīr šelīyār mēwe menteqe des kero, jārē āne minyā māčāš meřū āsinī} \\ 
\gll m-ē=we ētir ā zemān=e pīr šelīyār m-ē=we menteqe des=t ∅-ker-o jārē āne mi-nye-ā m-āč-ā=š meře=ū āsin-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} once \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) returned [to Hawraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ‘Iron cave.’'
\z 
 
\ea \label{ŽP.126}
\textit{māčā, qurbān to desūrēt dā be sey misefāy bilowe šerēʕetū terēqet ders wāčo} \\ 
\gll m-āč-ā qurbān to desūr-ē=t dā be sey misefā-ī bi-l-o=we šerēʕet=ū terēqet ders ∅-wāč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} \textsc{2sg} task/order\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{hort} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} sharia\_law=and denomination lesson\textsc{.m} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} \\ 
\glt `They (the gossips) said, ‘Sir, you ordered Say Mustafa to go back [to Hawraman and] teach Sharia law and [the ways of] orders (of dervishes).'
\z 
 
\ea \label{ŽP.135}
\textit{suāleš čene kero, māčo, čī āmānīwe pey īnīšā?’} \\ 
\gll suāle=š čene ∅-ker-o m-āč-o čī āmā=nī=we pey īnīšā \\ 
 question\textsc{.f}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} to dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He asked him [and] said, ‘Why have you come here?\footnote{Lit. for these.}’'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.156}
\textit{māčā, ī kināčē to tāwo weše kerowe’} \\ 
\gll m-āč-ā ī kināčē to ∅-tāw-o weš-e ∅-ker-o=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `It was said [to him in the dream], ‘He can cure this daughter of yours.’'
\z 
 
\ea \label{ŽP.159}
\textit{to bo lue! ī kināčē bere} \\ 
\gll to b-o ∅-lu-e ī kināčē ∅-ber-e \\ 
 \textsc{2sg} \textsc{imp-}come\textsc{.prs}\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `‘Come [here and] go! Take the girl.'
\z 
 
\ea \label{ŽP.162}
\textit{dēhāt be dēhāt ī kināčē bere} \\ 
\gll dēhāt be dēhāt ī kināčē ∅-ber-e \\ 
 village\textsc{.f} by village\textsc{.f} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `Take the girl village by village.'
\z 
 
\ea \label{ŽP.163}
\textit{eger to ī kināčēte berdeū ī pīyā her kes hen} \\ 
\gll eger to ī kināčē=t=e berd-e=ū ī pīyā her kes hen-∅ \\ 
 if \textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} take\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m} emph person\textsc{.m} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `If you take the girl, [no matter] who this man (the doctor) is,'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.198}
\textit{fermāwo, weɫā čūn ānā bēheyānī kerūt penū ʕāɫemī} \\ 
\gll ∅-fermāw-o weɫā čūn āne=ā bēheyā=nī ∅-ker-ū=t pen=ū ʕāɫem-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed as \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} shameless\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{2sg}} trouble\textsc{.m}\textsc{\textsc{=ez.gen}} world\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) said, ‘Indeed, since you are so indecent, I will disgrace you.’'
\z 
 
\ea \label{ŽP.199}
\textit{čūn bēheyānī āškirāt kerū} \\ 
\gll čūn bēheyā=nī āškirā=t ∅-ker-ū \\ 
 because shameless\textsc{=cop}\textsc{.\textsc{2sg}} visible\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Since you are indecent, I will reveal you.'
\z 
 
\ea \label{ŽP.207}
\textit{mišyo ī kināčē nīkāh kerū pey toū be mezebū wētāū} \\ 
\gll mišyo ī kināčē nīkāh ∅-ker-ū pey to=ū be mezeb=ū wē=tā=ū \\ 
 aux dem.prox girl\textsc{.dir}\textsc{.f} marriage\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} to \textsc{2sg}=and by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{ŽP.212}
\textit{berešo qey mekeronew} \\ 
\gll ∅-ber-e=š=o qey me-ker-one=ū \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Take her back. It is all right!'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.228}
\textit{yomā āne tišit miđo} \\ 
\gll yo=mā āne tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give [pomegranate] molasses.'
\z 
 
\ea \label{ŽP.229}
\textit{yoīčmā kučē pīyāzīt miđo} \\ 
\gll yo=īč=mā kuč-ē pīyāz-ī=t mi-đe-o \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} little\textsc{-indf} onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give you some onions.'
\z 
 
\ea \label{ŽP.230}
\textit{yoīč mā yoīč māčo, kāke gīyān! minīč šimšū yerū heywānēt mido pene} \\ 
\gll yo=īč=mā yo=īč m-āč-o kāke gīyān min=īč šimšī=ū yeher=ū heywān-ē=t mi-de-o pene \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother.vob dear.\textsc{voc} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `One [said], One person said, ‘Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{ŽP.232}
\textit{čāštekē kermē kermēš be zemāwine peyt} \\ 
\gll čāštī-ekē ∅-ker-mē ∅-ker-mē=š be zemāwine pey=t \\ 
 meal\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl}\textsc{=3sg} by wedding for\textsc{=\textsc{2sg}} \\ 
\glt `Let’s cook a meal and hold a wedding for you. '
\z 
 
\ea \label{ŽP.233}
\textit{nīkāhiš kermē peyt} \\ 
\gll nīkāh=iš ∅-ker-mē pey=t \\ 
 marriage\textsc{.m}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} for\textsc{=\textsc{2sg}} \\ 
\glt `Let us marry her to you. '
\z 
 
\ea \label{ŽP.234}
\textit{qey mekero nimāymē peket gino} \\ 
\gll qey me-ker-o ni-m-āz-mē peke=t ∅-gin-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-1pl} strenght\textsc{.f}\textsc{=\textsc{2sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `It is all right. We will not let you worry about it.'
\z 
 
\ea \label{ŽP.235}
\textit{nimāymē peket gino} \\ 
\gll ni-m-āz-mē peke=t ∅-gin-o \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-1pl} strenght\textsc{.f}\textsc{=\textsc{2sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `We will not let you worry about it.'
\z 
 
\ea \label{ŽP.236}
\textit{min hiɫošet miđo’} \\ 
\gll min hiɫoše=t mi-đe-o \\ 
 \textsc{1sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `[One of them said], ‘I will give you cracked wheat.’'
\z 
 
\ea \label{ŽP.237}
\textit{yoīč māčo, minīč tišit miđo’} \\ 
\gll yo=īč m-āč-o min=īč tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you [pomegranate] molasses.’'
\z 
 
\ea \label{ŽP.238}
\textit{yoīč māčo, minīč šimšū yeherit miđo} \\ 
\gll yo=īč m-āč-o min=īč šimšī=ū yeher=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you spleen and liver.’'
\z 
 
\ea \label{ŽP.262}
\textit{lue! to luewe hem wesāylīč bere pey ī pīyāyū ī kināčēw} \\ 
\gll ∅-lu-e to ∅-lu-e=we hem wesāyl=īč ∅-ber-e pey ī pīyā-ī=ū ī kināčē=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} both appliance\textsc{.m}\textsc{=add} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}=and dem.prox daughter\textsc{.dir}\textsc{.pl}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter'
\z 
 
\ea \label{ŽP.263}
\textit{ētir to lue meylū wētā’} \\ 
\gll ētir to ∅-lu-e meyl=ū wē=t=ā \\ 
 \textsc{disc.ptcl} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} willingness\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then it’s up to you [to stay in Hawraman]'
\z 
 
\ea \label{BP.13}
\textit{to hošit nīyā, tāteyt miqdārēš hošā—} \\ 
\gll to hoš=it nīye=ā tāte-ī=t miqdārē=š hoš=ā \\ 
 \textsc{2sg} memory\textsc{.m}\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} a\_little\textsc{=3sg} memory\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `—you are not aware [of that period], [but] your father remembers it a bit—'
\z 
 
\ea \label{BP.34}
\textit{tumez mekere, pewkī melāwe, epēgene gerekšān genekārī kerā} \\ 
\gll tumez me-ker-e pewkī me-l-ā=we epēgene gerek=šā=n genekārī ∅-ker-ā \\ 
 surprisingly \textsc{neg-}do.prs-\textsc{2sg} that\_is\_why \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} here necessary\textsc{=3pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} debauchery\textsc{.m} \textsc{sbjv-}do.prs\textsc{-3pl} \\ 
\glt `Ostensibly, they hadn’t gone back because they wished to engage in debauchery. '
\z 
 
\ea \label{BP.78}
\textit{wāče, bē dērā ēme milāy milmē pey ʕērāqī be peyē} \\ 
\gll wāč-e b-ē dēr=ā ēme mi-lāy mi-l-mē pey ʕērāq-ī be peyē \\ 
 say.prs-\textsc{imp-}\textsc{2sg} be.prs\textsc{-pstc}\textsc{.3sg} late\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} \textsc{ind-}go\textsc{.adv} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} to foot\textsc{.obl} \\ 
\glt `Tell him, “Come, it’s getting late. We are going to Iraq on foot.'
\z 
 
\ea \label{BP.81}
\textit{māčo, čī dēr āmāy?} \\ 
\gll m-āč-o čī dēr āmā-ī \\ 
 \textsc{ind-}say.prs\textsc{-3sg} why late come\textsc{.pst}-\textsc{2sg} \\ 
\glt `He (Baba Khwada) said, ‘Why have you come late?’'
\z 
 
\ea \label{BP.154}
\textit{māčā, čī girewī āne miheme?’} \\ 
\gll m-āč-ā čī ∅-girew-ī āne miheme \\ 
 \textsc{ind-}say.prs\textsc{-3pl} why \textsc{ind-}cry.prs-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{pn} \\ 
\glt `They said, ‘Mihamad, why are you crying?’'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.161}
\textit{bizāne yānew kē pī činnē hāy wexte nānē čīwēšā dānī ārdenit pey ēme, ārdenit pey ēmew} \\ 
\gll bi-zān-e yāne=ū kē p=ī činnē hāy wext=e nān-ē čīw-ē=šā dā=nī ārde=n=it pey ēme ārde=n=it pey ēme=ū \\ 
 \textsc{sbjv-}know.prs-\textsc{2sg} house\textsc{.m}\textsc{=ez.gen} who at=\textsc{dem.prox} some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{=dem} meal\textsc{.m}\textsc{-indf} thing\textsc{.m}\textsc{-indf}\textsc{=3pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl}=and \\ 
\glt `see which household has given you food, which you brought to us, during this period!'
\z 
 
\ea \label{BP.162}
\textit{wāče, kučē dūr ginowe”} \\ 
\gll wāč-e kuč-ē dūr ∅-gin-o=we \\ 
 say.prs-\textsc{imp-}\textsc{2sg} little\textsc{-indf} far \textsc{ind-}fall.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `Tell [them], “Go to a remote place”'
\z 
 
\ea \label{BP.163}
\textit{eger gošiš dārāyne dūr dūr kewto,} \\ 
\gll eger goš=iš dārā-ī=ne dūr dūr kewt-∅=o \\ 
 if ear\textsc{.m}\textsc{=3sg} hold\textsc{.pst}-\textsc{2sg}\textsc{=\textsc{povb}} far far fall\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `If they listen to you, they will go away.'
\z 
 
\ea \label{BP.164}
\textit{gošiš nedārāyne ētir jā bowe xeber bārewe} \\ 
\gll goš=iš ne-dārā-ī=ne ētir jā b-o=we xeber b-ār-e=we \\ 
 ear\textsc{.m}\textsc{=3sg} \textsc{neg-}hold\textsc{.pst}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{disc.ptcl} afterwards \textsc{imp-}come.prs\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{imp-}bring.prs-\textsc{2sg}\textsc{=compl} \\ 
\glt `If they don’t listen to you, then, come back and let us know.'
\z 
 
\ea \label{BP.165}
\textit{wāče her kesī nān dānī, wāče, ēšew be xātirū xwāy toū īđū ī kiɫfetū yāneyte dūr ginewe čī šārī’} \\ 
\gll wāč-e her kes-ī nān dā=nī wāč-e ēšew be xātir=ū xwā-ī to=ū īđ=ū ī kiɫfet=ū yāne-ī=t=e dūr ∅-gin-e=we čī šār-ī \\ 
 say.prs-\textsc{imp-}\textsc{2sg} just person\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} say.prs-\textsc{imp-}\textsc{2sg} tonight to sake\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}=and 3sg.prox\textsc{.dir}\textsc{.m}=and \textsc{dem.prox} child\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} far \textsc{imp-}fall.prs-\textsc{2sg}\textsc{=compl} why city\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Tell whoever gave you food, say, “Tonight, for the sake of God, you and your household get away from this town.'
\z 
 
\ea \label{BP.166}
\textit{lue bīrūnā”} \\ 
\gll ∅-lu-e bīrūnā \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} outside \\ 
\glt `Go away!'
\z 
 
\ea \label{BP.172}
\textit{hejījīyeke, fermāwo hejījīyeke, be qurbānit bim sūre gīyān yekē tirī lē de leqē teriš de wene’} \\ 
\gll hejījī-eke ∅-fermāw-o hejījī-eke be qurbān=it ∅-b-im sūr=e gīyān yek-ē tir=ī lē ∅-de-e leq-ē ter=iš ∅-de-e wene \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} to sacrifice\textsc{=\textsc{2sg}} \textsc{sbjv-}be.prs\textsc{-\textsc{1sg}} red=ez\textsc{.cmpd} dear.\textsc{voc} one\textsc{-indf} other\textsc{=3sg} at \textsc{imp-}give.prs-\textsc{2sg} kick\textsc{-indf} another\textsc{=3sg} \textsc{imp-}give.prs-\textsc{2sg} at \\ 
\glt `The [prisoner from] Hajij said [to Baba Khwada], ‘May I be your sacrifice, you red-skin, kick it once more! Kick it once more!”'
\z 
 
\ea \label{BP.199}
\textit{qetārleyē tewenīš hesār pey kēšānī} \\ 
\gll qetārle-ē tewenī=š hesār pey kēšā=nī \\ 
 small\_wall\textsc{-indf} stone.f\textsc{=3sg} pond\textsc{.m} for pull\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `They made a small wall made of stone for him.'
\z 
 
\ea \label{ŠJ.14}
\textit{māčā qurbān to čī bīyenī pīyā, čī bīyenī jāfir sān?} \\ 
\gll m-āč-ā qurbān to čī bīye=nī pīyā čī bīye=nī jāfir sān \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{2sg} why be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} man\textsc{.m} why be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{pn} \textsc{pn} \\ 
\glt `[and] said, ‘Sir, how have you become such a man? How have you turned into the Jafir Sān (that you are)?'
\z 
 
\ea \label{ŠJ.30}
\textit{māčā qurbān īne čiko bīyenī} \\ 
\gll m-āč-ā qurbān īne či=ko bīye=nī \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=where be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `They said, ‘Sir, where have you been?'
\z 
 
\ea \label{ŠJ.31}
\textit{ēme xo mātiɫū tonīmē} \\ 
\gll ēme xo mātiɫ=ū to=nīmē \\ 
 \textsc{1pl} dis.ptc suspended\textsc{=ez.gen} \textsc{2sg}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `We have been waiting for you.'
\z 
 
\ea \label{ŠJ.32}
\textit{pey lāw to āmēnmēw} \\ 
\gll pey lā=ū to āmē=nmē=ū \\ 
 to at\_the\_place\_of\textsc{=ez.gen} \textsc{2sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}=and \\ 
\glt `We have come to you.’'
\z 
 
\ea \label{ŠJ.40}
\textit{ħāɫīt bo?} \\ 
\gll ħāɫī=t ∅-b-o \\ 
 understood\textsc{=\textsc{2sg}} \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `[Addressing the researcher], do you get it?'
\z 
 
\ea \label{ŠJ.52}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `‘Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.53}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.60}
\textit{zānī bārbenū wiɫāxāš čikowen?} \\ 
\gll ∅-zān-ī bārben=ū wiɫāx-ā=š či=ko=we=n \\ 
 \textsc{ind-}know.prs-\textsc{2sg} stableman\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=3sg} in=where\textsc{=\textsc{post}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `‘Do you know where the horses’ stable is?’'
\z 
 
\ea \label{ŠJ.62}
\textit{māčo berem pištū berew wiɫāxekā} \\ 
\gll m-āč-o ∅-ber-e=m pišt=ū bere=ū wiɫāx-ekā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}take.prs-\textsc{2sg}\textsc{=\textsc{1sg}} back\textsc{=ez.gen} door\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} \\ 
\glt `He (Jafir Sān) said, ‘Take me behind the door of the horses’ stable!’'
\z 
 
\ea \label{ŠJ.87}
\textit{be desū to nīyenew} \\ 
\gll be des=ū to nīye=ne=ū \\ 
 to hand\textsc{.m}\textsc{=ez.gen} \textsc{2sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `It is not in your hands either.'
\z 
 
\ea \label{ŠJ.92}
\textit{mišo řū we yek neferī kerī ke derwārdenū meʕlūmā; meʕlūmārī} \\ 
\gll mišo řū=we yek nefer-ī ∅-ker-ī ke derđwārde=n=ū meʕlūm=ā meʕlūm=ā=rī \\ 
 aux face=to one person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}do.prs-\textsc{2sg} rel qualified\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and obvious\textsc{=cop}\textsc{.3sg}\textsc{.m} obvious\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `One should plead with one [ruler] who is qualified and distinct.'
\z 
 
\ea \label{ŠJ.101}
\textit{dey āro be kē wāčī derđwārde bo je dinyāne kes nemenen} \\ 
\gll dey āro be kē ∅-wāč-ī derđwārde ∅-b-o je dinyā=ne kes ne-mene=n \\ 
 \textsc{disc.ptcl} today to who \textsc{sbjv-}say.prs.\textsc{sbjv-}\textsc{2sg} qualified\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} from world\textsc{.dir}.f\textsc{=\textsc{post}} person\textsc{.m} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Nowadays, who can you refer to in the world who would be qualified? No one has remained.'
\z 
 
\ea \label{ŠJ.102}
\textit{enāne hetā fermāwī ī mentēqew ʕērāqū ērān desū ānīšāwe bīyen} \\ 
\gll enā=ne hetā ∅-fermāw-ī ī mentēqe=ū ʕērāq=ū ērān des=ū ānīšā=we bīye=n \\ 
 otherwise\textsc{=\textsc{post}} even \textsc{ind-}say.prs-\textsc{2sg} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{=ez.gen} \textsc{pn} hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All these regions in Iraq and Iran were under their rule.'
\z 
 
\ea \label{KŠ.9}
\textit{māčo, nāmerdit pene kerdā, čī pāset pene kerdā?’} \\ 
\gll m-āč-o nāmerd=it pene kerd-ā čī pāse=t pene kerd-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} just\textsc{=\textsc{2sg}} with do\textsc{.pst}\textsc{-\textsc{1sg}} why this\textsc{=\textsc{2sg}} with do\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘You didn’t act justly towards me! why did you do this to me?'
\z 
 
\ea \label{KŠ.10}
\textit{māčo, to kēnī min werew bāɫdārīm teqnān’} \\ 
\gll m-āč-o to kē=nī min wer=ew bāɫdār-ī=m teqnā=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{2sg} who\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{1sg} at\textsc{=\textsc{post}} bird\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} shoot\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (the king) said, ‘Who are you? I shot the birds.'
\z 
 
\ea \label{KŠ.12}
\textit{māčo, xās to čārenwīsnī, kināčēw pādšāyt īse pey kē niwīse? nāmēš pey kē?} \\ 
\gll m-āč-o xās to čārenwīs=nī kināčē=ū pādšā-ī=t īse pey kē niwīs-e nāmē=š pey kē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} fine \textsc{2sg} fortune\_teller\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} now to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} name\textsc{.f}\textsc{=3sg} to who \\ 
\glt `He (the king) said, ‘All right, [assuming that] you’re a fortune teller, whom have you considered [as the future husband] for the king’s daughter?\footnote{Lit. Whom did you write the king’s daughter for?} Who is she chosen for?'
\z 
 
\ea \label{KŠ.15}
\textit{ā wextīyekey, yewāšē wātiš pey, kināčēw pādšāyt pey kē niwīse?} \\ 
\gll ā wext-ī-ekey yewāšē wāt=iš pey kināčē=ū pādšā-ī=t pey kē niwīs-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} then say\textsc{.pst}\textsc{=3sg} to daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Then, he (the king) said, ‘Whom have you considered [as the future husband] for the king’s daughter?'
\z 
 
\ea \label{KŠ.28}
\textit{wātiš, qurbān herkey gele berī epāne min ī bizeme čekyo} \\ 
\gll wāt=iš qurbān herkey gele ∅-ber-ī epāne min ī bize=m=e ∅-čekye-o \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} each\_time herd\textsc{.m} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} there \textsc{1sg} \textsc{dem.prox} goat\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{ind-}drain\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `[and] said, ‘Sir, whenever you take the herd there, my goat[‘s milk] is drained.'
\z 
 
\ea \label{KŠ.51}
\textit{māčo, be xwā pāđšām seɫāmet kuřū šuwāney heke to kuštēū ā wextīyekey hukmit dāū īnīšā īse īnā yānetā īnīšā} \\ 
\gll m-āč-o be xwā pāđšā=m seɫāmet kuř=ū šuwāne-ī heke to kušt-ē=ū ā wext-ī-ekey hukm=it dā=ū īnīšā īse īnā-∅ yāne=tā īnīšā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} king\textsc{.m}\textsc{=\textsc{1sg}} healthy son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} if \textsc{2sg} kill\textsc{.pst}\textsc{-3pl}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} rule\textsc{.m}\textsc{=\textsc{2sg}} give\textsc{.pst}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} now \textsc{deic}\textsc{-3sg}\textsc{.m} house\textsc{.m}=\textsc{2pl} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `They said [in the letter], ‘Indeed, may my king be healthy, the shepherd’s son whom you ordered to be killed is now in your house and so on’'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.82}
\textit{māčo, tātet weš bēw hāɫū ēhwāɫiš xās bē} \\ 
\gll m-āč-o tāte=t weš b-ē=ū hāɫ=ū ēhwāɫ=iš xās b-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m}\textsc{=\textsc{2sg}} good be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and state\textsc{.m}\textsc{\textsc{=ez.gen}} demeanour\textsc{=3sg} fine be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (the shepherd’s son) said, ‘Your father was all right; he was feeling well.'
\z 
 
\ea \label{KŠ.86}
\textit{wātiš dey lue āney fiɫānīū fiɫānī čiře ī qewmāme bā bēnēwe qāqezekey wenā} \\ 
\gll wāt=iš dey ∅-lu-e āney fiɫānī=ū fiɫānī ∅-čiř-e ī qewm-ā=m=e bā b-ē-nē=we qāqez-ekey ∅-wen-ā \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} such\_and\_such\_person=and such\_and\_such\_person \textsc{imp-}call\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{hort} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}read\textsc{.prs}\textsc{-3pl} \\ 
\glt `She said, ‘Then go, and summon such-and-such, our relatives, so that they come and read the letter.’'
\z 
 
\ea \label{KŠ.87}
\textit{lue čiře wextē miđyo wāteniš, luāydē ā mejlistā girđ nīštīdēre hurneēzdēwe her čā māreš biřdē} \\ 
\gll ∅-lu-e ∅-čiř-e wext-ē mi-đye-o wāte=n=iš luā-ydē ā mejlis=tā girđ nīšt-īdē=re hur-ne-ēz-dē=we her čā māre=š ∅-biř-dē \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}call\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} go\textsc{.pst}-\textsc{2pl} \textsc{dem.dist} gathering\textsc{.m}=\textsc{2pl} all sit\textsc{.pst}-\textsc{2pl}\textsc{=\textsc{povb}} \textsc{pvb-}\textsc{neg-}rise\textsc{.prs}-\textsc{2pl}\textsc{=compl} just there marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2pl} \\ 
\glt `They (the relatives) went [to the palace]. They learned that he (the king) had said [in the letter], ‘Go and sit in that gathering. Then do not rise [from your seats] until you have married her to him.’'
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{KŠ.97}
\textit{yewāšē neferēwe řemo weɫēwe māčo be xānī, be xwā āne to kārū xwāyt tefre dā,} \\ 
\gll yewāšē nefer-ēwe ∅-řem-o weɫē=we m-āč-o be xān-ī be xwā āne to kār=ū xwā-ī=t tefre dā \\ 
 then person\textsc{.m}\textsc{-indf} \textsc{ind-}run\textsc{.prs}\textsc{-3sg} front\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by chief\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2sg} task\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} evasion give\textsc{.pst} \\ 
\glt `Then, a person went forward and said to the Khan, ‘Indeed, you [tried to] evade God’s destiny.'
\z 
 
\ea \label{KŠ.98}
\textit{yārū tifengeket teqnā, ženū šuwāney kušī,} \\ 
\gll yārū tifeng-eke=t teqnā ženī=ū šuwāne-ī ∅-kuš-ī \\ 
 \textsc{fill} gun\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} shoot\textsc{.pst} woman\textsc{.f}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}kill\textsc{.prs}-\textsc{2sg} \\ 
\glt `You fired a bullet from your gun; [and] you [intended] to kill the shepherd’s wife.'
\z 
 
\ea \label{KŠ.99}
\textit{ā wextīyekey čārenwīsī niwīsebēš ā kināčē to pey kuřū šuwāney niwīsyene} \\ 
\gll ā wext-ī-ekey čārenwīs-ī niwīse b-ē=š ā kināčē to pey kuř=ū šuwāne-ī niwīsye=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} fortune\_teller\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{dem.dist} daughter\textsc{.dir}\textsc{.f} \textsc{2sg} to son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}.pass\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `Then, the fortune teller had allocated\footnote{Lit. written.} your daughter for the shepherd’s son. '
\z 
 
\ea \label{KŠ.100}
\textit{āne kuřū šuwāneyn serū esbekey to} \\ 
\gll āne kuř=ū šuwāne-ī=n ser=ū esb-ekey to \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} on\textsc{\textsc{=ez.gen}} horse\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{2sg} \\ 
\glt `Look, the shepherd’s son [is] on your horse.'
\z 
 
\ea \label{KŠ.101}
\textit{āne kināčekētne ženīšne} \\ 
\gll āne kināčē-ekē=t=ne ženī=š=ne \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.f} woman\textsc{.f}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `That is your daughter. She is his wife.'
\z 
 
\ea \label{KŠ.104}
\textit{ā wextīyekey īmāniš ārd čā, wātiš xwā to heqnī} \\ 
\gll ā wext-ī-ekey īmān=iš ārd-∅ čā wāt=iš xwā to heq=nī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} belief\textsc{=3sg} bring\textsc{.pst}\textsc{-3sg}\textsc{.m} there say\textsc{.pst}\textsc{=3sg} God\textsc{.m} \textsc{2sg} right\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Then, he (the king) believed in God [and] said, ‘God, you are right!'
\z 
 
\ea \label{KŠ.105}
\textit{kārū to řāsā, kes metāwoš tefre biđo} \\ 
\gll kār=ū to řās=ā kes me-tāw-o=š tefre bi-đe-o \\ 
 task\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} true\textsc{=cop}\textsc{.3sg}\textsc{.m} person\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-3sg}\textsc{=3sg} evasion \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `Your deed is right! No one can avoid destiny.'
\z 
 
\ea \label{ŽH.9}
\textit{māčo, ko milī?’} \\ 
\gll m-āč-o ko mi-l-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} where \textsc{ind-}go\textsc{.prs}-\textsc{2sg} \\ 
\glt `He (Hayas) said, ‘Where are you going?’'
\z 
 
\ea \label{ŽH.11}
\textit{māčo, dey bilmē minīč mewt čene} \\ 
\gll m-āč-o dey bi-l-mē min=īč m-e-ū=t čene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} 1sg\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{2sg}} with \\ 
\glt `He (Hayas) said, ‘Let’s go, I am also coming with you.’'
\z 
 
\ea \label{ŽH.12}
\textit{milā řāhur māčo, heyās māčoš pene, kāke gīyān pilekānēwe binye bā bilmē!’} \\ 
\gll mi-l-ā řā=hur m-āč-o heyās m-āč-o=š pene kāke gīyān pilekānī-ēwe bi-nye-e bā bi-l-mē \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to brother.vob dear.\textsc{voc} stair\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}put\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \\ 
\glt `They set off. He, said, [i.e.] Hayas said to him, ‘Dear brother, set up a ladder for us to move !’'\footnote{A cryptic message meaning ‘tell a story’.}
\z 
 
\ea \label{ŽH.13}
\textit{ewīč māčo, min pilekānī mezānū binyewnew mezānū čēš māčī!’} \\ 
\gll ew=yč m-āč-o min pilekānī me-zān-ū bi-nye-ūne=ū me-zān-ū čēš m-āč-ī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg stair\textsc{\textsc{.f}} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} \textsc{sbjv-}put\textsc{.prs}\textsc{-1sg}=and \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what \textsc{ind-}say\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘I can’t set up a ladder and I don’t understand\footnote{Lit. know.}  what you are saying!’'
\z 
 
\ea \label{ŽH.17}
\textit{īse koge bilī yānew kē milī?’} \\ 
\gll īse koge bi-l-ī yāne=ū kē mi-l-ī \\ 
 now where \textsc{sbjv-}go\textsc{.prs}-\textsc{2sg} house\textsc{.m}=and who \textsc{ind-}go\textsc{.prs}-\textsc{2sg} \\ 
\glt `Where shall you be going? Whose house are you going to?’'
\z 
 
\ea \label{ŽH.18}
\textit{māčo, eger to nebī, mizgī’} \\ 
\gll m-āč-o eger to ne-b-ī mizgī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} if \textsc{2sg} \textsc{neg-}be\textsc{.prs}-\textsc{2sg} mosque\textsc{.m} \\ 
\glt `He (Hayas) said, ‘If it is not you[r house, I will go to] the mosque.’'
\z 
 
\ea \label{ŽH.22}
\textit{ā wextīyekey, kināče gewrekēš māčo, tāte, her wēt āmānīwe?’} \\ 
\gll ā wext-ī-ekey kināčē=e gewre-ekē=š m-āč-o tāte her wē=t āmā=nī=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} big\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m} just \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `Then, his older daughter said, ‘Father, have you come back alone?’'
\z 
 
\ea \label{ŽH.25}
\textit{wātiš, pilekānē čine bā bilmē!”} \\ 
\gll wāt=iš pilekānī-ē ∅-čin-e bā bi-l-mē \\ 
 say\textsc{.pst}\textsc{=3sg} stair\textsc{\textsc{.f}}\textsc{-indf} \textsc{ind-}set\_up\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \\ 
\glt `He said, “Set up a ladder for us to move!”’'
\z 
 
\ea \label{ŽH.27}
\textit{wātiš, tāte āne wāteniš, hekāyetē weš gīsnene, bā ī řē kučē keme kermē be hekāyetekeyt rāz biřmē} \\ 
\gll wāt=iš tāte āne wāte=n=iš hekāyet-ē weš-∅ ∅-gīsn-e=ne bā ī řē kuč-ē kem-e ∅-ker-mē be hekāyet-ekey=t rāz ∅-biř-mē \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} tale\textsc{.m}\textsc{-indf} well-m \textsc{imp-}start\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{hort} \textsc{dem.prox} road\textsc{.obl}\textsc{\textsc{.f}}.sg little\textsc{-indf} s\textsc{hort}\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} by tale\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{2sg}} tale\textsc{.m} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl} \\ 
\glt `She said, ‘Father, he had said, “Start [telling] a good tale, so that we [are entertained] and our way seems shorter. Let us finish your tale.”'
\z 
 
\ea \label{ŽH.28}
\textit{to nezānānt!’} \\ 
\gll to ne-zānā=n=t \\ 
 \textsc{2sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `[But] you didn’t understand [his point].’'
\z 
 
\ea \label{ŽH.31}
\textit{wātim peneš to milī ko?”} \\ 
\gll wāt=im pene=š to mi-l-ī ko \\ 
 say\textsc{.pst}\textsc{=1sg} to\textsc{=3sg} \textsc{2sg} \textsc{ind-}go\textsc{.prs}-\textsc{2sg} where \\ 
\glt `I said to him, “Where are you going?”’'
\z 
 
\ea \label{ŽH.32}
\textit{wātiš, eger to nebī mizgī’} \\ 
\gll wāt=iš eger to ne-b-ī mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} if \textsc{2sg} \textsc{neg-}be\textsc{.prs}-\textsc{2sg} mosque\textsc{.m} \\ 
\glt `He (Hayas) said, ‘If it is not you[r house, I will go to] the mosque.’'
\z 
 
\ea \label{ŽH.33}
\textit{wātiš, tāte, āne wāteniš, eger to xuɫkem mekerī, meberīm yāne milū mizgī’} \\ 
\gll wāt=iš tāte āne wāte=n=iš eger to xuɫke=m me-ker-ī me-ber-ī=m yāne mi-l-ū mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} if \textsc{2sg} invitation\textsc{=1sg} \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \textsc{neg-}take\textsc{.prs}-\textsc{2sg}\textsc{=1sg} house\textsc{.m}\textsc{.dir.sg} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} mosque\textsc{.m} \\ 
\glt `She said, ‘Father, he meant,\footnote{Lit. said.} “If you do not invite me and won’t take me home, I will go to the mosque.”'
\z 
 
\ea \label{ŽH.34}
\textit{hāɫīt bo?} \\ 
\gll hāɫī=t ∅-b-o \\ 
 understood\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `You see?'
\z 
 
\ea \label{ŽH.38}
\textit{vātiš, dey tāte dey bā čīwēweš pey bere!’} \\ 
\gll vāt=iš dey tāte dey bā čīw-ēwe=š pey ∅-ber-e \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} father\textsc{.m} \textsc{disc.ptcl} \textsc{hort} thing\textsc{-indf}\textsc{=3sg} for \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `She said, ‘Father, get him something.’'
\z 
 
\ea \label{ŽH.40}
\textit{wātiš be wāɫe wičkilekēš ī māsī bere deš pā mēmān mizgī čenīw nānī’} \\ 
\gll wāt=iš be wāɫē=e wičkile-ekē=š ī mās-ī ∅-ber-e ∅-de-e=š p=ā mēmān mizgī čenī=ū nān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} by sister\textsc{\textsc{.f}}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} to\textsc{=dem}.dist guest\textsc{.m} mosque\textsc{.m} with\textsc{=ez}\textsc{.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said to her little daughter, ‘Take this yoghurt [and] together with [some] bread give it to the the guest in the mosque.’ '
\z 
 
\ea \label{ŽH.44}
\textit{māčo, jā heweřekey bāro} \\ 
\gll m-āč-o jā heweř-ekey b-ār-e=o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} afterwards pot\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg}\textsc{=compl} \\ 
\glt `She said, ‘Bring the pot back afterwards.’'
\z 
 
\ea \label{ŽH.45}
\textit{māčo, wāče be dēdēt māngeš keče bē’} \\ 
\gll m-āč-o ∅-wāč-e be dēdē=t mānge=š keč-e b-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}say\textsc{.prs-}\textsc{2sg} to older\_sister\textsc{\textsc{.f}}\textsc{=\textsc{2sg}} month\textsc{\textsc{.f}}\textsc{=3sg} crooked\textsc{-f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Hayas) said, ‘Tell your sister\footnote{Lit. maternal aunt.} that her moon was incomplete .’'\footnote{A cryptic message which in this context means that the yoghurt has been fingered by the girl’s daughter}
\z 
 
\ea \label{ŽH.47}
\textit{māčo, mēmānekey wāt, wāče dēdēt} \\ 
\gll m-āč-o mēmān-ekey wāt ∅-wāč-e dēdē=t \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} guest\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} say\textsc{.pst} \textsc{imp-}say\textsc{.prs-}\textsc{2sg} older\_sister\textsc{\textsc{.f}}\textsc{=\textsc{2sg}} \\ 
\glt `She said, ‘The guest said, “Tell your sister [that her moon is incomplete]”'
\z 
 
\ea \label{ŽH.48}
\textit{māčo, je māseket keč nekerđebo řāře wāɫē!’} \\ 
\gll m-āč-o je mās-eke=t keč ne-kerđe b-o řā=ře wāɫē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} from yoghurt\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} crooked \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} sister\textsc{\textsc{.f}} \\ 
\glt `She (the older sister) had said [to her younger sister], ‘Sister, could the yoghurt be reduced’'
\z 
 
\ea \label{ŽH.50}
\textit{ā wextīyekey, māčo be tāteyš, dā ā mēmānī bāre pey yāneymā!’} \\ 
\gll ā wext-ī-ekey m-āč-o be tāte-ī=š dā ā mēmān-ī b-ār-e pey yāne-ī=mā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{hort} \textsc{dem.dist} guest\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \\ 
\glt `The, she (the older sister) said to her father, ‘Bring that guest home!’'
\z 
 
\ea \label{ŽH.53}
\textit{māčo, ī serū řīšīme kučē tāše’} \\ 
\gll m-āč-o ī sere=ū řīš-ī=m=e kuč-ē ∅-tāš-e \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox} head\textsc{.m}=and beard\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} little\textsc{-indf} \textsc{imp-}shave\textsc{.prs}-\textsc{2sg} \\ 
\glt `He (Hayas) said [to the girl], ‘shave my face a bit.’'
\z 
 
\ea \label{ŽH.69}
\textit{māčo, bereš pey kurdistānī’} \\ 
\gll m-āč-o ∅-ber-e=š pey kurdistān-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said [to Hayas], ‘Take this to Kurdistan. '
\z 
 
\ea \label{ŽH.78}
\textit{āđīč māčo, min desūrim pēsnen pey wēt āmānā} \\ 
\gll āđ=īč m-āč-o min desūr=im pēsne=n pey wē=t āmā=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg task/order\textsc{.m}\textsc{=1sg} like\_this\textsc{=cop}\textsc{.3sg}\textsc{.m} for \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \\ 
\glt `He said, ‘My task is like this! I have come for you.’'
\z 
 
\ea \label{ŽH.79}
\textit{ā wextī yewāšē māčo, dey tehemuɫ kere bā hēɫeřūenit pey kerū, hāɫīt kerū'} \\ 
\gll ā wext-ī yewāšē m-āč-o dey tehemuɫ ∅-ker-e bā hēɫeřūen=it pey ∅-ker-ū hāɫī=t ∅-ker-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} then \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} patience \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{hort} fried\_egg\textsc{=\textsc{2sg}} for \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} understood\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then she said, ‘All right, wait a little bit, let me cook you fried eggs, [and] make you understand.'
\z 
 
\ea \label{ŽH.82}
\textit{xo hāɫīt bo hēɫē yānē kāmēnē} \\ 
\gll xo hāɫī=t ∅-b-o hēɫe-ē yānē kām-ē=nē \\ 
 dis.ptc understood\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} egg\textsc{.m}\textsc{-dir}\textsc{.pl} that\_means which\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Do you understand what eggs are?'
\z 
 
\ea \label{ŽH.94}
\textit{māčo, werešā} \\ 
\gll m-āč-o ∅-wer-e=šā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}eat\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \\ 
\glt `[and] said [to the sultan], ‘Eat them [one by one]!’'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.104}
\textit{ānā dersē ʕālit dānē} \\ 
\gll āne=ā ders-ē ʕāl=it dā-ānē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} lesson\textsc{.m}\textsc{-indf} good\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{-1sg} \\ 
\glt `You gave me a good lesson.'
\z 
 
\ea \label{ŽH.105}
\textit{memnūnitnā} \\ 
\gll memnūn=it=nā \\ 
 grateful\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.1sg} \\ 
\glt `I am grateful to you.'
\z 
 
\ea \label{ŽH.112}
\textit{māčo, āđīč čiřoš māčo, čī nelānīwe?’} \\ 
\gll m-āč-o āđ=īč ∅-čiř-o=š m-āč-o čī ne-lā=nī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `He (the sultan) called him (Hayas) [and] said, ‘Why don’t you go back (to your house)?’'
\z 
 
\ea \label{ŽH.118}
\textit{luewe pey yāney!’} \\ 
\gll ∅-lu-e=we pey yāne-ī \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} for house\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[Now] go back home!'
\z 
 
\ea \label{ŽH.119}
\textit{dey duʕāw seɫāmetīt} \\ 
\gll dey duʕā=ū seɫāmetī=t \\ 
 \textsc{disc.ptcl} prayer\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} health\textsc{=\textsc{2sg}} \\ 
\glt `That’s it. May good prayers  be for you.'
\z 



\ea \label{RE.4}
\textit{bābe gīyān ī kināčēte yā tātemā lūān yā māmomā yā kesūkārmā} \\ 
\gll bābe gīyān ī kināčē=t=e yā tāte=mā lūā=n yā māmo=mā yā kesūkār=mā \\ 
 father.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} either father\textsc{.m}\textsc{=\textsc{1pl}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} either paternal\_uncle\textsc{=\textsc{1pl}} either relative\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `They would say, ‘Dear brother,\footnote{Lit. father.} this daughter of yours \—either our father would go [to the girl’s father], or our uncle, or a relative of ours—'
\z 
 
\ea \label{RE.5}
\textit{āmān lūān wāteniš, birā gīyān! min xū wēt zānī hāwmāɫitnā} \\ 
\gll āmā=n lūā=n wāte=n=iš birā gīyān min xū wē=t ∅-zān-ī hāwmāɫ=it=nā \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} brother\textsc{.m} dear.\textsc{voc} \textsc{1sg} \textsc{disc.ptcl} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} neighbour\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `They would go [and] say, ‘Dear brother, you know [well] that I am your neighbour.'
\z 
 
\ea \label{RE.6}
\textit{be bīyew nebīyeym zānī} \\ 
\gll be bīyey=w nebīyey=m ∅-zān-ī \\ 
 to resources\textsc{.m}\textsc{\textsc{=ez.gen}} s\textsc{hort}age\textsc{.m}\textsc{=\textsc{1sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} \\ 
\glt `You know about my resources and shortages. '
\z 
 
\ea \label{RE.7}
\textit{ʕerzū birāw wēm kerū ī kināčēte de pī kuřīme} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū ī kināčē=t=e ∅-de-e p=ī kuř-ī=m=e \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} brother\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} at=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \\ 
\glt `Let me tell you,\footnote{Lit. May a petition be at the service of my brother.} ‘Give your daughter to my son [in marriage]'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.11}
\textit{ētir kābrā fers kere zemānū ā ānē hītē pāɫē bīyēnēū čwār gezē pārče bīyenū} \\ 
\gll ētir kābrā fers ∅-ker-e zemān=ū ā ānē hīt-ē pāɫā-ē bīyē=nē=ū čwār gez-ē pārče bīye=n=ū \\ 
 \textsc{disc.ptcl} man guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} \textsc{dem.dist}\textsc{.3pl} pair\textsc{-indf} shoe\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and four ell\textsc{.m}\textsc{-indf} textile\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Then, let us say,\footnote{Lit. imagine.} at the time, [the wedding gift] consisted of a pair of shoes [and] four gaz\footnote{A unit of measure equivalent to 72 cm.} of textile.'
\z 
 
\ea \label{RE.17}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{RE.19}
\textit{lue hārdekā bārewe} \\ 
\gll ∅-lu-e hārdī-ekā b-ār-e=we \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} flour\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.pl} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg}\textsc{=compl} \\ 
\glt `Go bring the flour.'
\z 
 
\ea \label{RE.20}
\textit{lue kārekem pey kere} \\ 
\gll ∅-lu-e kār-eke=m pey ∅-ker-e \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} task\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} for \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Go do [this] task for me.'
\z 
 
\ea \label{RE.27}
\textit{ētir īse zemān fāřyān hāɫīt bo?} \\ 
\gll ētir īse zemān fāřyā=n hāɫī=t ∅-b-o \\ 
 \textsc{disc.ptcl} now time\textsc{.m} change\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} understood\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Now, things have changed. You see?'
\z 
 
\ea \label{RE.29}
\textit{kābrā melo wāčo, ženīm de} \\ 
\gll kābrā me-l-o ∅-wāč-o ženī=m ∅-de-e \\ 
 man \textsc{neg-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} woman\textsc{.f}\textsc{=\textsc{1sg}} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `The fellow wouldn’t go [to his parents and] say, ‘Find a girl for me!’\footnote{Lit. Give me a woman.}'
\z 
 
\ea \label{RE.37}
\textit{miđey mehr kerū,} \\ 
\gll mi-đe-ī mehr ∅-ker-ū \\ 
 \textsc{ind-}give\textsc{.prs}-\textsc{2sg} mariage\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `If you accept,\footnote{Lit. give.} I will marry [him]'
\z 
 
\ea \label{RE.38}
\textit{međey mehr kerū,} \\ 
\gll me-đe-ī mehr ∅-ker-ū \\ 
 \textsc{neg-}give\textsc{.prs}-\textsc{2sg} mariage\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `If you don’t accept, I will marry [him]'
\z 
 
\ea \label{RE.39}
\textit{to čikārenī} \\ 
\gll to čikāre=nī \\ 
 \textsc{2sg} what\_job\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `It is not your business.\footnote{Lit. What is your business.}’'
\z 
 
\ea \label{RE.40}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{RE.49}
\textit{min kināčēt dēne kuřū fiɫāne kesī} \\ 
\gll min kināčē=t dē=ne kuř=ū fiɫān=e kes-ī \\ 
 \textsc{1sg} daughter\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}} son/boy\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and say to the lord], ‘I have given my\footnote{Lit. your.} daughter to the son of such-and-such person.'
\z 
 
\ea \label{RE.50}
\textit{to qibūɫ kerī yā mekerī} \\ 
\gll to qibūɫ ∅-ker-ī yā me-ker-ī \\ 
 \textsc{2sg} accepted \textsc{ind-}do\textsc{.prs}-\textsc{2sg} either \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Will you consent to it or not?'
\z 
 
\ea \label{RE.53}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{RE.55}
\textit{māčo, dey lue deš} \\ 
\gll m-āč-o dey ∅-lu-e ∅-de-e=š \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} \\ 
\glt `He (the nobleman) would say [to the girl’s father], ‘Go [and] give her’'
\z 
 
\ea \label{RE.60}
\textit{sūrāne yānē mišo bārī to zeř dey jā min māčū erē} \\ 
\gll sūrāne yānē mišo ∅- b-ār-ī to zeř ∅-de-ī jā min m-āč-ū erē \\ 
 \textsc{pn} that\_means aux \textsc{ind-}\textsc{sbjv-}bring\textsc{.prs}-\textsc{2sg} \textsc{2sg} money\textsc{.m} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} afterwards \textsc{1sg} \textsc{ind-}say\textsc{.prs}\textsc{-\textsc{1sg}} yes \\ 
\glt `Surana means that you should give me money, then I will say, ‘Yes [to the marriage]'
\z 
 
\ea \label{ŽE.4}
\textit{ī degā toš vīnī čoɫe bīyēne} \\ 
\gll ī degā to=š ∅-vīn-ī čoɫ-e bīyē=ne \\ 
 \textsc{dem.prox} village\textsc{.obl}\textsc{\textsc{.f}} \textsc{2sg}\textsc{=3sg} \textsc{ind-}see\textsc{.prs}-\textsc{2sg} deserted-\textsc{f} be\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `This village, which you see, was deserted.'
\z 
 
\ea \label{ŽE.49}
\textit{wātenšā, xwā gīyān hemdū senā pey to’} \\ 
\gll wāte=n=šā xwā gīyān hemd=ū senā pey to \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} God\textsc{.m} dear.\textsc{voc} praise\textsc{\textsc{=ez.gen}} admiration for \textsc{2sg} \\ 
\glt `[They] the would say, ‘Dear God, may our praises be for you!’'
\z 
 
\ea \label{ŽE.50}
\textit{ī doīnete dēnmēū ī šelemīnēte dēnmē} \\ 
\gll ī doīne=t=e dē=nmē=ū ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{\textsc{=ez.gen}} \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this dwina. \footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{ŽE.51}
\textit{ī šelemīnēte dēnmē} \\ 
\gll ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `you gave us this Shalamina. \footnote{Dish made of cracked wheat and turnip.}'
\z 
 
\ea \label{ŽE.52}
\textit{ī xeplēte dēnmē} \\ 
\gll ī xeplē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this xapla '
\z 
 
\ea \label{ŽE.54}
\textit{xwā hemdū senā pey to!’} \\ 
\gll xwā hemd=ū senā pey to \\ 
 God\textsc{.m} praise\textsc{\textsc{=ez.gen}} admiration for \textsc{2sg} \\ 
\glt `God, may our praises be for you!’'
\z 
 
\ea \label{ŽE.73}
\textit{ā wextī to mīsāl wātēt, ā fiɫāne kesem gerekene’} \\ 
\gll ā wext-ī to mīsāl wāt-ē=t ā fiɫān=e kese=m gerek-e=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg} for\_example say\textsc{.pst}\textsc{-cond}\textsc{=\textsc{2sg}} \textsc{dem.dist} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.f}\textsc{=\textsc{1sg}} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `In earlier times, let us say\footnote{Lit. for instance.} you said, ‘I want that certain person (i.e., girl).''
\z 
 
\ea \label{ŽE.74}
\textit{luēnī lāw ī kuřī wāčēnī, dā lue wāče eđēš bizāne ī kināčēše miđom pene yā ne’} \\ 
\gll lu-ēn-ī lā=ū ī kuř-ī wāč-ēn-ī dā ∅-lu-e ∅-wāč-e eđē=š bi-zān-e ī kināčē=š=e mi-đe-o=m pene yā ne \\ 
 go\textsc{.prs}\textsc{-pstc}-\textsc{2sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.prs}\textsc{-pstc}-\textsc{2sg} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}say\textsc{.prs-}.\textsc{2sg} mother\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=3sg}\textsc{=dem} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{1sg}} to either no \\ 
\glt ` You would go to this boy [seated next to the narrator] and would say, ‘Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!’'
\z 
 
\ea \label{ŽE.75}
\textit{kuřekeyč ē māčī, kināčekēt miđey pī kuřī?’} \\ 
\gll kuř-ekey=č ē m-āč-ī kināčē-ekē=t mi-đe-ī p=ī kuř-ī \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} come\textsc{.prs}\textsc{.pst}c\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The boy comes. [Addressing her grand son] you would say [to her mother], ‘Will you give your daughter to this boy [in marriage]?’'
\z 
 
\ea \label{ŽE.78}
\textit{wāčē, dey lue duē yerē pīyē bēnē’} \\ 
\gll wāč-ē dey ∅-lu-e duē yerē pīyē b-ē-nē \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc.}\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} two three man\textsc{.dir}\textsc{.pl} \textsc{sbjv-}come\textsc{.prs}\textsc{-3pl} \\ 
\glt `She (the mother) would say, ‘Go! Let two, three men come.'
\z 
 
\ea \label{ŽE.80}
\textit{ēnē hījbīš kerē wāčē, ī kināčē miđey pī kuřī?’} \\ 
\gll ēnē hījbī=š ker-ē wāč-ē ī kināčē mi-đe-ī p=ī kuř-ī \\ 
 come\textsc{.prs}\textsc{.pstc}\textsc{.3pl} betrothal\textsc{=3sg} do\textsc{.prs}\textsc{-pstc} say\textsc{.prs-pstc}\textsc{.3sg} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `They would come [and] act as intermediary to ask for the hand of the girl. They would say, ‘Will you give this girl to this boy?’'
\z 
 
\ea \label{ŽE.82}
\textit{wāčē, dey lue yānew wēt nīšere tā duē māngē ter yerē māngē ter} \\ 
\gll wāč-ē dey ∅-lu-e yāne=ū wē=t ∅-nīš-e=re tā duē mānge-ē ter yerē mānge-ē ter \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc}.\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} until two month\textsc{.f}\textsc{-dir}\textsc{.pl} other three month\textsc{.f}\textsc{-dir}\textsc{.pl} other \\ 
\glt `They (the family of the girl) would say [to the boy], ‘Go back home [and] wait\footnote{Lit. sit.} for two, three months.'
\z 
 
\ea \label{ŽM.37}
\textit{ħāɫītā} \\ 
\gll ħāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
\ea \label{ŽM.48}
\textit{hāɫītā?} \\ 
\gll hāɫī=t=ā \\ 
 understood\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `You see?'
\z 
 
