\chapter{All 133 examples of add in these texts are found in the following sentences:}

\ea \label{ZB.1}
\textit{řoē je řoā yoe mē yoyč mē xizmetū šā ʕelādīnī, her ā šā ʕelādīne} \\ 
\gll řo-ē je řo-ā yo-e m-ē yo=yč m-ē xizmet=ū šā ʕelādīn-ī her ā šā ʕelādīn=e \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m}\textsc{-f} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} one\textsc{.m}\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} emph \textsc{dem.dist} king\textsc{.m} \textsc{pn}\textsc{=dem} \\ 
\glt `Once,\footnote{Lit. One day of days.} a woman and a man\footnote{Lit. One (f.) came, one (m.) came.} came to the service of šā ʕalādīn, the same šā ʕalādīn.'
\z 
 
\ea \label{ZB.8}
\textit{mē šewē je šewā ženēweš bo wēšū kināčē ezemēč lāšo bo} \\ 
\gll m-ē šewe-ē je šewe-ā ženī-ēwe=š ∅-b-o wē=š=ū kināčē ezem-e-ē=č lā=š=o ∅-b-o \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} night\textsc{.f}\textsc{-indf} from night\textsc{.f}\textsc{-obl}\textsc{.pl} woman\textsc{.f}\textsc{-indf}\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{reflx}\textsc{=3sg}=and girl\textsc{.obl}\textsc{.f} adolescent\textsc{-f}\textsc{-indf}\textsc{=add} at\_the\_place\_of\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `There came a night when —he had a wife. She and an adolescent daughter [of his] were with him.'
\z 
 
\ea \label{ZB.23}
\textit{beʕzēčiš meno milwā xizānā meřekē} \\ 
\gll beʕzē=č=iš ∅-men-o mi-lw-ā ∅-xiz-ānē meře-ekē \\ 
 some\textsc{=add}\textsc{=3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{ind-}creep\textsc{-3pl} cave\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `Some of [the animals] which had survived went and squeezed themselves into the cave.'
\z 
 
\ea \label{ZB.24}
\textit{šewē be ʕemreū xwāy ženēč nīšore duē zāroɫē wīno her ā šewē} \\ 
\gll šewe-ē be ʕemre=ū xwā-ī ženī-ē=č ∅-nīš-o=re duē zāroɫe-ē ∅-wīn-o her ā šewe-ē \\ 
 night\textsc{.f}\textsc{-obl}\textsc{.f} to order\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m} woman\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=add} \textsc{ind-}sit\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} two child\textsc{-dir}\textsc{.pl} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} emph \textsc{dem.dist} night\textsc{.f}\textsc{-obl}\textsc{.f} \\ 
\glt `During that very night, by the order of God, the wife gave birth to two babies.\footnote{Lit. She sat down [and] saw two babies.} '
\z 
 
\ea \label{ZB.35}
\textit{mekiryoyč ī kuřī xinīkno} \\ 
\gll me-kir-īye-o=yč ī kuř-ī ∅-xinīkn-o \\ 
 \textsc{neg-}do.pass-pass\textsc{-3sg}\textsc{=add} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv}-strangle\textsc{.prs}\textsc{-3sg} \\ 
\glt `We should not strangle the [baby] boys.\footnote{Lit. It can't be done that we strangle the baby boys.}'
\z 
 
\ea \label{ZB.50}
\textit{ā wextīekey īđīčšā ārdenū} \\ 
\gll ā wext-ī-ekey īđī=č=šā ārde=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{.obl}\textsc{.m}\textsc{=add}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They (the family) brought him back.'
\z 
 
\ea \label{ZB.54}
\textit{ī bizēče powe āwēzāne bo mē e čē ēđ čeknošū milo lā yāgekēš} \\ 
\gll ī bizē=č=e p=o=we āwēzān-e ∅-b-o m-ē e=čē ēđ čekn-o=š=ū mi-l-o lā yāgē-ekē=š \\ 
 \textsc{dem.prox} goat\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=dem} at=\textsc{dem.dist}\textsc{=compl} hanging\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} in=here 3sg.prox\textsc{.dir} suckle\textsc{.prs}\textsc{-3sg}\textsc{=3sg}=and \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of place\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \\ 
\glt `The goat would stay in a hanging position, the child would feed from its udder, and it would go back to its place.'
\z 
 
\ea \label{ZQ.13}
\textit{ženīem bē ženekēmīč bārū hemɫīe bē} \\ 
\gll ženīe=m b-ē ženī-ekē=m=īč bār=ū hemɫī-e b-ē \\ 
 woman.f\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs\textsc{.3sg} woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}}\textsc{=add} load\textsc{.m}\textsc{=ez.gen} pregnant\textsc{-f} \textsc{sbjv-}come.prs\textsc{.3sg} \\ 
\glt `I had a wife. My wife was pregnant.'
\z 
 
\ea \label{ZQ.19}
\textit{yoyčšā her gīyāne gūwīnēš čenī bē melāqāt} \\ 
\gll yo=yč=šā her gīyān=e gūwīne-ē=š čenī b-ē melāqāt \\ 
 one\textsc{.m}\textsc{=add}\textsc{=3pl} just soul\textsc{.m}=ez\textsc{.cmpd} half\textsc{-indf}\textsc{=3sg} in \textsc{sbjv-}come.prs\textsc{.3sg} visit \\ 
\glt `[The other] one of them was only half alive.\footnote{Lit. had little soul left in him.}    '
\z 
 
\ea \label{ZQ.21}
\textit{ēmeyč gīrmā wārd čenū her wī} \\ 
\gll ēme=yč gīr=mā wārd čenū her wē-ī \\ 
 \textsc{1pl}\textsc{=add} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} just two\textsc{-obl}\textsc{.m} \\ 
\glt `We were stuck with both babies.'
\z 
 
\ea \label{ZQ.22}
\textit{yošā kerdmā bāboɫē ārdmā} \\ 
\gll yo=šā kerd-∅=mā bāboɫē ārd-∅=mā \\ 
 one\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1pl}} swaddling bring\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `We rolled one in clothing [and] took him.'
\z 
 
\ea \label{ZQ.23}
\textit{yoīčšā her gīyānēš čenī bē gēɫānē qiřoɫe dārēm yoso} \\ 
\gll yo=īč=šā her gīyān-ē=š čenī b-ē gēɫā-ānē qiřoɫ=e dār-ē=m yos=o \\ 
 one\textsc{.m}\textsc{=add}\textsc{=3pl} just soul\textsc{.m}\textsc{-indf}\textsc{=3sg} in \textsc{sbjv-}come.prs\textsc{.3sg} turn\textsc{.pst}\textsc{-\textsc{1sg}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} find\textsc{.pst}\textsc{=compl} \\ 
\glt `As for the one which was half-alive, I wandered around [and] found a tree hollow.'
\z 
 
\ea \label{ZQ.32}
\textit{āy bāwke řo, āy tāte řo īđīč luāw ewīč luā} \\ 
\gll āy bāwke řo āy tāte řo īđ=īč luā=ū ew=īč luā \\ 
 \textsc{intj} fahter.\textsc{voc} onom \textsc{intj} father\textsc{.m} onom 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg}=and 3sg\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg} \\ 
\glt `Oh dear God\footnote{The expressions bāwka řo and tāta řo are used to mourn someone’s death, or to express distress.}, he (the child whom I took with me) died; he (the child whom I left behind) died too!\footnote{Lit. This one went; that one, too, went.} '
\z 
 
\ea \label{HB.6}
\textit{hezretū šēxīč fermāwo, dey konī?} \\ 
\gll hezret=ū šēx=yč ∅-fermāw-o dey ko=nī \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} where\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `His highness the Sheikh said, ‘Where are you?’'
\z 
 
\ea \label{HB.15}
\textit{kābrāyč herbenew hereyn, milo herey gurj kerowew gino řā,} \\ 
\gll kābrā=īč herbene=ū her-e-ī=n mi-l-o her-e-ī gurj ∅-ker-o=we=ū ∅-gin-o řā \\ 
 man\textsc{=add} donkey\_breeder\textsc{.m}=and donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} alert \textsc{ind-}do\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}}\textsc{=ez}\textsc{.gen} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} road\textsc{\textsc{.f}} \\ 
\glt `The fellow who was a donkey breeder, went and prepared the donkey, [and] set off.'
\z 
 
\ea \label{HB.22}
\textit{zeřīčiš neyānā bilū yāgē bisānū} \\ 
\gll zeř=yč=iš ne-yā=nā bi-l-w yāgē-ē bi-sān-w \\ 
 money\textsc{.m}\textsc{=add}\textsc{=3sg} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} place\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}buy\textsc{.prs}\textsc{-1sg} \\ 
\glt `He hasn’t given me money either so that I could go to a place [and] buy [flour]'
\z 
 
\ea \label{HB.38}
\textit{toyč berā yāne} \\ 
\gll to=īč ∅-ber-ā yāne \\ 
 \textsc{2sg}\textsc{=add} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} house \\ 
\glt `they will take you to [their] house;'
\z 
 
\ea \label{HB.40}
\textit{nānīč miđā to} \\ 
\gll nān=yč mi-đe-ā to \\ 
 bread\textsc{.m}\textsc{=add} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{2sg} \\ 
\glt `they will give you a meal;'
\z 
 
\ea \label{HB.41}
\textit{toīč nān werī} \\ 
\gll to=yč nān ∅-wer-ī \\ 
 \textsc{2sg}\textsc{=add} bread\textsc{.m} \textsc{ind-}out\textsc{-obl}\textsc{.m} \\ 
\glt `[and] you will eat your meal.'
\z 
 
\ea \label{HB.45}
\textit{ewīč goš miđo pene} \\ 
\gll ew=yč goš mi-đe-o pene \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} ear\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `He listened to it [i.e. the donkey].'
\z 
 
\ea \label{HB.65}
\textit{xeylēč menebē bilowe Bīyārē} \\ 
\gll xeylē=č mene b-ē bi-l-o=we Bīyārē \\ 
 so much\textsc{=add} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{pn} \\ 
\glt `He had a long way to go back to Biyare.'
\z 
 
\ea \label{HB.66}
\textit{xeylēč menebē bilowe Bīyārēw} \\ 
\gll xeylē=č mene b-ē bi-l-o=we Bīyārē=ū \\ 
 so much\textsc{=add} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{pn}=and \\ 
\glt `He had a long way to go back to Biyare.'
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{PM.38}
\textit{girđū šārezūrī ēsāɫ merāʕetū ēme kero je pārī zīyāter pey toyčū pey ēmeyč,} \\ 
\gll girđ=ū šārezūr-ī ēsāɫ merāʕet=ū ēme ∅-ker-o je pār-ī zīyā-ter pey to=yč=ū pey ēme=yč \\ 
 all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} this\_year care\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} from last\_year\textsc{-obl}\textsc{.m} increased\textsc{-cmpr} for \textsc{2sg}\textsc{=add}=and for \textsc{1pl}\textsc{=add} \\ 
\glt `This year, all the people in Sharazour care for us— both you and I—, more than the previous year.'
\z 
 
\ea \label{PM.45}
\textit{min īse ī nīštenāre īsrāhetīč kerū} \\ 
\gll min īse ī nīšte=nā=re īsrāhet=īč ∅-ker-ū \\ 
 \textsc{1sg} now \textsc{dem.prox} sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=\textsc{povb}} rest\textsc{.m}\textsc{=add} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Now I am seated and I am resting.'
\z 
 
\ea \label{DG.16}
\textit{kābrāyč werū řehmū wēšū milo pāɫ midowe birākeyšo} \\ 
\gll kābrā=īč wer=ū řehm=ū wē=š=ū mi-l-o pāɫ mi-de-o=we birā-ekey=š=o \\ 
 man\textsc{=add} out\_of\textsc{=ez}\textsc{.gen} mercy\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3sg}=and \textsc{ind-}go\textsc{.prs}\textsc{-3sg} side/flank\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `The fellow, out of mercy, went and supported his brother.'
\z 
 
\ea \label{DG.44}
\textit{īđīč fermāwo, lue čā nīšere,} \\ 
\gll īđ=īč ∅-fermāw-o ∅-lu-e čā ∅-nīš-e=re \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} there \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \\ 
\glt `Shah Aladin  said, ‘Go [and] sit there;'
\z 
 
\ea \label{DG.67}
\textit{šā ʕelādīnīč fermāwo birākem wextē īnē zānā min gunākē māy mē milimre} \\ 
\gll šā ʕelādīn-ī=č ∅-fermāw-o birā-eke=m wext-ē īnē ∅-zān-ā min gunā-ekē m-āy m-ē mil=im=re \\ 
 king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} time\textsc{.m}\textsc{-indf} \textsc{dem.prox}\textsc{.3pl} \textsc{ind-}know\textsc{.prs}\textsc{-3pl} \textsc{1sg} sin\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}come\textsc{.pst}\textsc{.adv} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} shoulder\textsc{.m}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `Shah Aladin said, ‘Brother!—as they (i.e. the tortoises) know [about the medicine] I won’t shoulder the burden of injuring them.'
\z 
 
\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.17}
\textit{ā wextīyekey īđīč āmān dāwāw pādšāgerīš kerdēnew} \\ 
\gll ā wext-ī-ekey īđ=yč āmā=n dāwā=w pādšāgerī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} kingship\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `He (i.e., Jamsher Shah) came [and] claimed kingship.'
\z 
 
\ea \label{DP.35}
\textit{īzāfeyš wizo pā sextre māčāš qeɫāw hewrāmānī} \\ 
\gll īzāfe-ī=š wiz-o p=ā sext=re mi-āč-ā=š qeɫā=w hewrāmān-ī \\ 
 addition\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg} by\textsc{=dem}.dist mountain\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He would throw whatever was extra [of the snow] at that hard-to-access mountain called the citadel of Hawraman.'
\z 
 
\ea \label{DP.40}
\textit{ewīč nimē māčo, nexeyr nimewnew mitmānim nīyā penet.} \\ 
\gll ew=yč ni-mi-ē mi-āč-o nexeyr ni-mi-e-wne=ū mitmān=im nīye=ā pene=t \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-1sg}=and trust\textsc{.m}\textsc{=1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} to\textsc{=\textsc{2sg}} \\ 
\glt ` Sibhan Agha\footnote{Lit. He.} did not go [to Jamsher Shah], saying, ‘No, I won’t come [to you]; I don’t trust you.’'
\z 
 
\ea \label{ZP.86}
\textit{māčā xeɫkīč, ī pīyā šimeš pey persdē xo yānēw e čāneš nīyāw} \\ 
\gll m-āč-ā xeɫk=īč ī pīyā šime=š pey ∅-pers-dē xo yāne-ē=ū e=čāne=š nīye=ā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} people\textsc{.m}\textsc{=add} dem.prox man\textsc{.m} \textsc{2pl}\textsc{=3sg} for \textsc{ind-}ask\textsc{.prs}-\textsc{2pl} dis.ptc house\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} in=like\_that\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The people said, ‘This man whom you ask for, he does not have a proper house.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.107}
\textit{ewīč māčo, nexeyre lāloekeš min tāze pādšā kerdenā wekēɫū} \\ 
\gll ew=īč m-āč-o nexeyre lālo-eke=š min tāze pādšā kerde=nā wekēɫ=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{1sg} anyway king\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m}=and \\ 
\glt ` The uncle said, ‘No! I—the king has put me in charge.'
\z 
 
\ea \label{ZP.108}
\textit{wātenīčiš, mišyo neberūšo”} \\ 
\gll wāte=n=īč=iš mišyo ne-ber-ū=š=o \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=add}\textsc{=3sg} aux \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `He has said, “I shall not take her back.”'
\z 
 
\ea \label{ZP.115}
\textit{māčā, dey īne ī pīyā ī ženīče ārdēne} \\ 
\gll m-āč-ā dey īne ī pīyā ī ženī=č=e ārdē=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox man\textsc{.m} dem.prox woman\textsc{.dir}f\textsc{=add}\textsc{=dem} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `[and] said, ‘Well, this man has brought this woman [here]'
\z 
 
\ea \label{ZP.117}
\textit{īđīč hīč minīč hīčim nīyārēw} \\ 
\gll īđ=īč hīč min=īč hīč=im nīye=ā=rē=ū \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} nothing \textsc{1sg}\textsc{=add} nothing\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `She [has] nothing, nor do I.'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.119}
\textit{yoīč māčo, dey minīč beʕzē pīyāzīt midonew’} \\ 
\gll yo=īč m-āč-o dey min=īč beʕzē pīyāz-ī=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg}\textsc{=add} some onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Another said, ‘I will give you some onions.’'
\z 
 
\ea \label{ZP.120}
\textit{yoīč māčo, minīč beʕzē tiršū henārāt midonew’} \\ 
\gll yo=īč m-āč-o min=īč beʕzē tirš=ū henār-ā=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} some molasses\textsc{\textsc{=ez.gen}} pomegranate\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt ` Another said, ‘I will give you some pomegranate molasses.’'
\z 
 
\ea \label{ZP.121}
\textit{kābrāyēč māčo, dey birā, minīč serē heywānit midew} \\ 
\gll kābrā-ē=č m-āč-o dey birā min=īč ser-ē heywān=it mi-de-ū \\ 
 man\textsc{-indf}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{1sg}\textsc{=add} \textsc{clf}\textsc{-indf} animal\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `A man said, ‘Brother, I will give you an animal.’'
\z 
 
\ea \label{ZP.129}
\textit{šūnū ēmeyčre her meno} \\ 
\gll šūn=ū ēme=yč=re her ∅-men-o \\ 
 after\textsc{\textsc{=ez.gen}} \textsc{1pl}\textsc{=add}\textsc{=\textsc{post}} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `It will continue to exist after us as well.'
\z 
 
\ea \label{ŽP.24}
\textit{milo mēweū māčā, hizmēč bāre’} \\ 
\gll mi-l-o m-ē=we=ū m-āč-ā hizmē=č b-ār-e \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl}=and \textsc{ind-}say\textsc{.prs}\textsc{-3pl} firewood\textsc{.f}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt ` He went [shepherding and] came back [home]. They said, ‘Bring firewood as well.’'
\z 
 
\ea \label{ŽP.25}
\textit{derdeyčiš hēzmē māro} \\ 
\gll derdey=č=iš hēzmī-ē m-ār-o \\ 
 load\textsc{.m}\textsc{=add}\textsc{=3sg} firewood\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt ` He brought [home] a load of firewood on his shoulders.'
\z 
 
\ea \label{ŽP.29}
\textit{werū mejbūrī yewāyč bero} \\ 
\gll wer=ū mejbūrī yewe-ā=yč ∅-ber-o \\ 
 out\_of\textsc{\textsc{=ez.gen}} obligation barley\textsc{.f}\textsc{-3pl}\textsc{=add} \textsc{ind-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt ` Out of obligation, he took the barley seeds, too.'
\z 
 
\ea \label{ŽP.37}
\textit{bizekēčiš be lūlekē hurpiřā jīyā leweřyāyšā} \\ 
\gll bize-ekē=č=iš be lūle-ekē hur-∅-piř-ā jīyā leweřyāy=šā \\ 
 goat\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=3sg} by flute\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{pvb-}\textsc{ind-}jump\textsc{.prs}\textsc{-3pl} instead\_of graze.inf\textsc{=3pl} \\ 
\glt `With the sound of the flute, the goats danced instead of grazing.'
\z 
 
\ea \label{ŽP.48}
\textit{šuwānew bizāyč bo} \\ 
\gll šuwāne=ū bize-ā=yč ∅-b-o \\ 
 shepherd\textsc{.m}\textsc{\textsc{=ez.gen}} goat\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=add} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `And he is shepherding the goats.'
\z 
 
\ea \label{ŽP.77}
\textit{ū pīyāyče řāsiš wāt ūneš gēɫnāwew} \\ 
\gll ū pīyā=yč=e řās=iš wāt ūne=š gēɫnā=we=ū \\ 
 \textsc{dem.dist}\textsc{.3sg} man\textsc{.m}\textsc{=add}\textsc{=dem} truth\textsc{=3sg} say\textsc{.pst} \textsc{dem.dist}\textsc{=3sg} narrate\textsc{.pst}\textsc{=\textsc{post}}=and \\ 
\glt `And the man (the servant) told the truth when he said that [those words]'
\z 
 
\ea \label{ŽP.84}
\textit{misefā yāwān pene tāwo je šerīʕetīč īstīfāde kero je terēqetīč} \\ 
\gll misefā yāwā=n pene ∅-tāw-o je šerīʕet-ī=č īstīfāde ∅-ker-o je terēqet-ī=č \\ 
 \textsc{pn} arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \textsc{ind-}can\textsc{.prs}\textsc{-3sg} from sharia\_law\textsc{-obl}\textsc{.m}\textsc{=add} use \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} from denomination\textsc{-obl}\textsc{.m}\textsc{=add} \\ 
\glt `Mustafa has got [knowledgeable] enough that he can both practise Sharia law and lead an order (of dervishes).'
\z 
 
\ea \label{ŽP.88}
\textit{čūn je šerēʕetīčnew je terēqetīčne yāwān pene} \\ 
\gll čūn je šerēʕet-ī=č=ne=ū je terēqet-ī=č=ne yāwā=n pene \\ 
 because from sharia\_law\textsc{-obl}\textsc{.m}\textsc{=add}\textsc{=\textsc{post}}=and from denomination\textsc{-obl}\textsc{.m}\textsc{=add}\textsc{=\textsc{post}} arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \\ 
\glt `Because he has both mastered Sharia law and [is capable of leading an] order (of dervishes).'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.125}
\textit{māčā kābrāyē pēse īse ā wextīč qisekerē her bīyēnē} \\ 
\gll m-āč-ā kābrā-ē pēse īse ā wext-ī=č qiseker-ē her bīyē=nē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} man\textsc{-indf} as\_if now \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} gossiper\textsc{-dir}\textsc{.pl} emph be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They said — Like nowadays, back then, there were also gossips.'
\z 
 
\ea \label{ŽP.157}
\textit{pādšāyč māčo be lālo kināčekē be šālyāre sīyāwīš} \\ 
\gll pādšā=yč m-āč-o be lālo kināčē-ekē be šālyār=e sīyāw-ī=š \\ 
 king\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} by \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `The king said to [his] daughter’s uncle, to Shaliyar Syaw'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.219}
\textit{xo ā wextī wēčiš yāneš her mebo} \\ 
\gll xo ā wext-ī wē=č=iš yāne=š her me-b-o \\ 
 dis.ptc \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=add}\textsc{=3sg} house\textsc{.m}\textsc{=3sg} emph \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He had no house at that time.'
\z 
 
\ea \label{ŽP.226}
\textit{āđīč māčo, min metāwū ī kināčē wey kerū īne nānmā} \\ 
\gll āđ=īč m-āč-o min me-tāw-ū ī kināčē wey ∅-ker-ū īne nān=m=ā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox girl\textsc{.dir}\textsc{.f} raising \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `People … he (Pir Shaliyar) said, ‘I cannot take care of this girl. This is my food.’'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.229}
\textit{yoīčmā kučē pīyāzīt miđo} \\ 
\gll yo=īč=mā kuč-ē pīyāz-ī=t mi-đe-o \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} little\textsc{-indf} onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give you some onions.'
\z 
 
\ea \label{ŽP.230}
\textit{yoīč mā yoīč māčo, kāke gīyān! minīč šimšū yerū heywānēt mido pene} \\ 
\gll yo=īč=mā yo=īč m-āč-o kāke gīyān min=īč šimšī=ū yeher=ū heywān-ē=t mi-de-o pene \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother.vob dear.\textsc{voc} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `One [said], One person said, ‘Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{ŽP.237}
\textit{yoīč māčo, minīč tišit miđo’} \\ 
\gll yo=īč m-āč-o min=īč tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you [pomegranate] molasses.’'
\z 
 
\ea \label{ŽP.238}
\textit{yoīč māčo, minīč šimšū yeherit miđo} \\ 
\gll yo=īč m-āč-o min=īč šimšī=ū yeher=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you spleen and liver.’'
\z 
 
\ea \label{ŽP.239}
\textit{yoīč māčo minīč mekī mārū} \\ 
\gll yo=īč m-āč-o min=īč mekī m-ār-ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} salt\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `One said, ‘I will give you salt.’'
\z 
 
\ea \label{ŽP.240}
\textit{yoīč māčo minīč pīyāzī mārū} \\ 
\gll yo=īč m-āč-o min=īč pīyāz-ī m-ār-ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} onion\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `One said, ‘I will give you onions.’'
\z 
 
\ea \label{ŽP.243}
\textit{pīr šelīyār fermāwo, tehāā weɫāhī īne šime desūrtā dāw wāttā čūnke āđē be diɫtā wāt řāstā wāt hetā xwāy dinyāš weše kerdēne her menowe kemīč mekero her zīyā kero} \\ 
\gll pīr šelīyār ∅-fermāw-o tehāā weɫāhī īne šime desūr=tā dā=ū wāt=tā čūnke āđē be diɫ=tā wāt řās=tā wāt hetā xwā-ī dinyā=š weš-e kerdē=ne her ∅-men-o=we kem=īč me-ker-o her zīyā ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} task/order\textsc{.m}=\textsc{2pl} give\textsc{.pst}=and say\textsc{.pst}=\textsc{2pl} since 3pl\textsc{.dir} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} truth=\textsc{2pl} say\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} small\textsc{=add} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar said, ‘Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] continues to exist; it will not be diminished; it will grow.'
\z 
 
\ea \label{ŽP.245}
\textit{ī merāsēme her menowe hīčkeyč kem mekero her zīyāy kero īlāy žīyāyne bo zīyāyiš pene kero} \\ 
\gll ī merāsēm=e her ∅-men-o=we hīčke=yč kem me-ker-o her zīyāy ∅-ker-o īlāy žīyāy=ne ∅-b-o zīyāy=iš pene ∅-ker-o \\ 
 dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} nothing\textsc{=add} little \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increase \textsc{ind-}do\textsc{.prs}\textsc{-3sg} for\_sure increased\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} increased\textsc{=3sg} to \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This ceremony will be repeated, it won’t diminish. It will continue to grow. It will grow [into bigger ceremonies]'
\z 
 
\ea \label{ŽP.262}
\textit{lue! to luewe hem wesāylīč bere pey ī pīyāyū ī kināčēw} \\ 
\gll ∅-lu-e to ∅-lu-e=we hem wesāyl=īč ∅-ber-e pey ī pīyā-ī=ū ī kināčē=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} both appliance\textsc{.m}\textsc{=add} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}=and dem.prox daughter\textsc{.dir}\textsc{.pl}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter'
\z 
 
\ea \label{ŽP.266}
\textit{ēti melowe āđīč melowe} \\ 
\gll ēti me-l-o=we āđ=īč me-l-o=we \\ 
 \textsc{disc.ptcl} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `After all, he didn’t go back [to Egypt]. He did not go back either.'
\z 
 
\ea \label{BP.38}
\textit{ī kābrēče je ʕērāqo āmēnē, wēregānew nīmeřonew seʕbne bexšnāyšā bexšnēnēwe} \\ 
\gll ī kābrā-ē=č=e je ʕērāq=o āmē=nē wēregā=ne=ū nīmeřo=ne=ū seʕbe=ne bexšnāy=šā bexšn-ēn-ē=we \\ 
 \textsc{dem.prox} man\textsc{-dir}\textsc{.pl}\textsc{=add}\textsc{=dem} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} evening\textsc{=\textsc{post}}=and noon\textsc{=\textsc{post}}=and morning.f\textsc{=\textsc{post}} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-indf}\textsc{=compl} \\ 
\glt `They (people) would donate [food] to the fellows (the tax collectors) who had come from Iraq, in the evenings, mornings, and at noon. '
\z 
 
\ea \label{BP.60}
\textit{ēmeyč māčmē, ī meʕmūrē šime kīyāsēnē bā biřew bāremā dēw nāmēwe bārāšo} \\ 
\gll ēme=yč m-āč-mē ī meʕmūr-ē šime kīyāsē=nē ∅-b-ā biře=ū bāre=mā dā-ē=ū ne-āmā-ē=we b-ār-ā=š=o \\ 
 \textsc{1pl}\textsc{=add} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{ind-}be.prs\textsc{-3pl} portion\textsc{.m}=and profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}bring.prs\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `We will say, ‘We paid tax to the officers whom you had sent, [however,] they didn’t return to bring it back.'
\z 
 
\ea \label{BP.63}
\textit{ēmeyč muxteserū kelāmī ā īney wāčmē,} \\ 
\gll ēme=yč muxteser=ū kelām-ī ā īney wāč-mē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} \textsc{dem.dist} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} say.prs.sbjv\textsc{-\textsc{1pl}} \\ 
\glt `As for us, to cut a long story short, we say this.'
\z 
 
\ea \label{BP.75}
\textit{mēnē mizgī jemʕē bā} \\ 
\gll m-ē-nē mizgī jemʕ-ē ∅-b-ā \\ 
 \textsc{ind-}come.prs\textsc{-3pl} mosque\textsc{.m} added\textsc{-pstc}\textsc{.3sg} \textsc{ind-}be.prs\textsc{-3pl} \\ 
\glt `They (people) went to the mosque [and] gathered there.'
\z 
 
\ea \label{BP.86}
\textit{āđīč mē werew māčo, fermāwo, dā dey dey jārēne jimdēre’} \\ 
\gll āđ=īč m-ē wer=ew m-āč-o ∅-fermāw-o dā dey dey jār-ē=ne ∅-jim-dē=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}come.prs\textsc{.3sg} front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{disc.ptcl} \textsc{disc.ptcl} time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} \\ 
\glt `He (Baba Khwada) came forward [and] said, ‘Please, set off the way all of you.’'
\z 
 
\ea \label{BP.95}
\textit{ēmeyč muxteserū kelāmī, xo ēme diruē meyeymē} \\ 
\gll ēme=yč muxteser=ū kelām-ī xo ēme diruē me-de-ymē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} dis.ptc \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `As for us, to cut a long story short, we are not lying.'
\z 
 
\ea \label{BP.98}
\textit{ā wextī ēti seʕbe milā milāne mizgī, mizyāre beynē gināre řā beynē bilā pey ʕērāqī jemʕē bā ħemey ġeybīū bābā xuāđā ginā weɫēwe} \\ 
\gll ā wext-ī ēti seʕbe mi-l-ā mi-l-ān=e mizgī mi- zye-ā=re beynē ∅-gin-ā=re řā beynē bi-l-ā pey ʕērāq-ī jemʕ-ē ∅-b-ā ħeme=ī ġeybī=ū bābā xuāđā ∅-gin-ā weɫē=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} morning.f \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}go.prs\textsc{-3pl}=drct mosque\textsc{.m} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} a\_while \textsc{ind-}fall.prs\textsc{-3pl}\textsc{=\textsc{povb}} road.f a\_while \textsc{sbjv-}go.prs\textsc{-3pl} for \textsc{pn}\textsc{-obl}\textsc{.m} added\textsc{-pstc}\textsc{.3sg} \textsc{ind-}be.prs\textsc{-3pl} \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m} \textsc{pn} \textsc{ind-}fall.prs\textsc{-3pl} front\textsc{=\textsc{post}} \\ 
\glt `So, in the morning they went to the mosque. They went a little way towards Iraq. They all gathered. Hama Ghaybi and Baba Khwada went in front of them.'
\z 
 
\ea \label{BP.100}
\textit{āwetū xuɫkī īse ā wextīč xuɫk her tersezāɫ bo,} \\ 
\gll āđet=ū xuɫk-ī īse ā wext-ī=č xuɫk her tersezāɫ ∅-b-o \\ 
 habit\textsc{=ez.gen} people\textsc{.m}\textsc{-obl}\textsc{.m} now \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} people\textsc{.m} just coward \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `Just like people nowdays, in earlier times people were intimidated [by kings]. '
\z 
 
\ea \label{BP.104}
\textit{ewēč her serešā kumnānre} \\ 
\gll ewē=č her sere=šā kumnā=n=re \\ 
 3sg\textsc{.dir}.f\textsc{=add} just head\textsc{.m}\textsc{=3pl} bend\_down\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} \\ 
\glt `They (the people) nodded [in agreement]'
\z 
 
\ea \label{BP.122}
\textit{ā wextī reʕīsū milāre ʕērāq lāw āġekeyšā māčo seremā biřyēnē bābāw pīr šelīyārī bo, ā wextīyekey ħemey ġeybīū yoyč bo nāmē ħemey wirdīkɫe,} \\ 
\gll ā wext-ī reʕīs=ū mi-l-ā=re ʕērāq lā=ū āġe-ekey=šā m-āč-o sere=mā biřyē=nē bābā=ū pīr šelīyār-ī ∅-b-o ā wext-ī-ekey ħeme=ī ġeybī=ū yo=yč ∅-b-o nāmē ħeme=ī wirdīkɫe \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} leader\textsc{.m}\textsc{=ez.gen} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{pn} at\_the\_place\_of\textsc{=ez.gen} agha\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}say.prs\textsc{-3sg} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}=ez.\textsc{attr} invisible=and one\textsc{.m}\textsc{=add} \textsc{ind-}be.prs\textsc{-3sg} name.f \textsc{pn}=ez.\textsc{attr} small \\ 
\glt `The head of those who went to Iraq to the agha, saying, ‘We have decapitated [your men], was Pir Shaliyar’s grandfather, Baba Khwada. In addition [there were] Hama the invisible and Little Hama. '
\z 
 
\ea \label{BP.127}
\textit{ēmeyč muxteserū kelāmī seremā biřyēnē} \\ 
\gll ēme=yč muxteser=ū kelām-ī sere=mā biřyē=nē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `So, to cut a long story short, we decapitated them.'
\z 
 
\ea \label{BP.132}
\textit{ħemey wirdīkɫeyč pēse īse māčā, heɫāy neyāwān bilūqū sinʕēš} \\ 
\gll ħeme=ī wirdīkɫe=yč pēse īse m-āč-ā heɫāy ne-yāwā=n bilūq=ū sinʕe-ē=š \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} like now \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} maturity\textsc{.m}\textsc{=ez.gen} age.f\textsc{-obl}.f\textsc{=3sg} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) said, ‘he hasn’t hit puberty yet.'
\z 
 
\ea \label{BP.140}
\textit{āđīč her īnā je zīnānekene} \\ 
\gll āđ=īč her īnā-∅ je zīnān-eke=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} just deic\textsc{-3sg}\textsc{.m} from prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man from Hajij) was also in the prison. '
\z 
 
\ea \label{BP.143}
\textit{ħemey wirdīkɫeyč mičkile bo, wirdīkɫe bo pāne girđ seferē milo pā šārne gēɫo} \\ 
\gll ħeme=ī wirdīkɫe=yč mičkile ∅-b-o wirdīkɫe ∅-b-o p=ā=ne girđ sefer-ē mi-l-o p=ā šār=ne ∅-gēɫ-o \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} small \textsc{ind-}be.prs\textsc{-3sg} small \textsc{ind-}be.prs\textsc{-3sg} at=\textsc{dem.dist}\textsc{=\textsc{post}} all trip\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} at=\textsc{dem.dist} city\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}wander.prs\textsc{-3sg} \\ 
\glt `Hama Wirdikla was small and young. He would constantly go around in the city.'
\z 
 
\ea \label{BP.155}
\textit{fermāwo āđīč, weɫā min pey šime girewū} \\ 
\gll ∅-fermāw-o āđ=īč weɫā min pey šime ∅-girew-ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} indeed \textsc{1sg} for \textsc{2pl} \textsc{ind-}cry.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘I’m crying for you.’'
\z 
 
\ea \label{BP.182}
\textit{seyēmēw min čurtēč kerū kučē būsū} \\ 
\gll sey-ēmē=ū min čurt-ē=č ∅-ker-ū kuč-ē b-ūs-ū \\ 
 rest.prs-\textsc{1pl}=and \textsc{1sg} nap\textsc{-indf}\textsc{=add} \textsc{ind-}do.prs\textsc{-\textsc{1sg}} little\textsc{-indf} \textsc{sbjv-}sleep.prs\textsc{-\textsc{1sg}} \\ 
\glt ` I shall rest,\footnote{Lit. My rest comes.} take a nap [and] sleep a little bit. They were a bit tired.'
\z 
 
\ea \label{BP.183}
\textit{kučē būsū wermīčim mēw kučēš mānīyānā} \\ 
\gll kuč-ē b-ūs-ū werm=īč=im m-ē=ū kuč-ē=š mānīyā=nā \\ 
 little\textsc{-indf} \textsc{sbjv-}sleep.prs\textsc{-\textsc{1sg}} sleep\textsc{=add}\textsc{=\textsc{1sg}} \textsc{ind-}come.prs\textsc{.3sg}=and little\textsc{-indf}\textsc{=3sg} tired\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I shall sleep a little bit. I feel sleepy.\footnote{Lit. my sleep comes.} I am a bit tired.'
\z 
 
\ea \label{BP.189}
\textit{āđēč miqdārē be enāzew seʕātē nīšāre dīyāršo} \\ 
\gll āđē=č miqdārē be enāze=ū seʕāt-ē ∅-nīš-ā=re dīyār=š=o \\ 
 3pl\textsc{.dir}\textsc{=add} a\_little to proportion\textsc{=ez.gen} hour\textsc{.m}\textsc{-indf} \textsc{ind-}sit.prs\textsc{-3pl}\textsc{=\textsc{povb}} region\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `They (Baba Khwada and Hama Wirdikla) waited for him for about an hour.'
\z 
 
\ea \label{ŠJ.20}
\textit{āđīč seʕbē pīyā kīyāno māčo lodē bārdēš!} \\ 
\gll āđ=īč seʕbe-ē pīyā ∅-kīyān-o m-āč-o ∅-lo-dē b-ār-dē=š \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} morning.f\textsc{-obl}.f man\textsc{.m} \textsc{ind-}send.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}bring.prs-\textsc{2pl}\textsc{=3sg} \\ 
\glt `The next day, he (Jafir Sān) sent some men [and] said, ‘Go [and] bring him!'
\z 
 
\ea \label{ŠJ.52}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `‘Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.53}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.54}
\textit{esb zīnī kero peyšū} \\ 
\gll esb zīnī ∅-ker-o pey=š=ū \\ 
 horse saddle \textsc{ind-}do.prs\textsc{-3sg} for\textsc{=3sg}=and \\ 
\glt `He saddled the horse for him. At night, he set off quickly,'
\z 
 
\ea \label{ŠJ.82}
\textit{ʕeybekeyčiš memeno} \\ 
\gll ʕeyb-eke=yč=iš me-men-o \\ 
 fault\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=add}\textsc{=3sg} \textsc{neg-}remain.prs\textsc{-3sg} \\ 
\glt `The defect had disappeared.'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{KŠ.17}
\textit{āđīč hukmiš dā wātiš, ženē leme peře mebo bo} \\ 
\gll āđ=īč hukm=iš dā wāt=iš ženī-ē leme peř-e me-b-o ∅-b-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} rule\textsc{.m}\textsc{=3sg} give\textsc{.pst} say\textsc{.pst}\textsc{=3sg} woman\textsc{.f}\textsc{-indf} stomach\textsc{.f} full\textsc{-f} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the king) issued an order [and] said, ‘No pregnant women should stay alive!’\footnote{Lit. No pregnant women should be.}'
\z 
 
\ea \label{KŠ.18}
\textit{benāxātirē īney āyeč bilo, hāmɫē bē} \\ 
\gll benāxātirē īney āye=č bi-l-o hāmɫē b-ē \\ 
 for\_the\_reason \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} 3sg\textsc{.f}\textsc{.dir}\textsc{=add} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} pregnant\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `The reason [for issuing this order] was that she (the shepherd’s wife), who was pregnant, would disappear.'
\z 
 
\ea \label{KŠ.29}
\textit{āđīč wāt, dey xāsā min gino šūnī[šre]} \\ 
\gll āđ=īč wāt dey xās=ā min ∅-gin-o šūnī=š=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} track\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `He (the shepherd) said, ‘All right, I will look after her (the goat).’'
\z 
 
\ea \label{KŠ.40}
\textit{kināčēw pāđšāyč māčo, kuřū šuwāneyč bārdē ēme tāqetmā pey āney mē} \\ 
\gll kināčē=ū pāđšā-ī=č m-āč-o kuř=ū šuwāne-ī=č b-ār-dē ēme tāqet=mā pey āney m-ē \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} \textsc{1pl} persistence\textsc{.m}\textsc{=\textsc{1pl}} to \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `The king’s daughter said, ‘Fetch the shepherd’s son. He will entertain us.\footnote{Lit. Our patience comes by him.}’'
\z 
 
\ea \label{KŠ.52}
\textit{āđīč māčo, wāčdē bē melāqātīm} \\ 
\gll āđ=īč m-āč-o wāč-dē b-ē melāqātī=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} say\textsc{.prs}-\textsc{imp-}\textsc{2pl} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} visitor\textsc{.m}\textsc{=\textsc{1sg}} \\ 
\glt `He (the king) said, ‘Tell him to come to visit me.’'
\z 
 
\ea \label{KŠ.54}
\textit{āđīč milo čenīū duē yerē milo} \\ 
\gll āđ=īč mi-l-o čenī=ū duē yerē mi-l-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} with\textsc{=ez.gen} two three \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the shepherd’s son) too went, together with two, three [persons].'
\z 
 
\ea \label{KŠ.63}
\textit{āđīč gēɫo āšpezxānekene kučē čīw menebo čāštū čīw wero} \\ 
\gll āđ=īč ∅-gēɫ-o āšpezxāne-eke=ne kuč-ē čīw mene b-o čāštī=ū čīw ∅-wer-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}wander\textsc{.prs}\textsc{-3sg} kitchen\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} little\textsc{-indf} thing\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} meal\textsc{.f}\textsc{\textsc{=ez.gen}} thing\textsc{.m} \textsc{ind-}eat\textsc{.prs}\textsc{-3sg} \\ 
\glt `He looked into the kitchen. There was some leftovers. He ate them. '
\z 
 
\ea \label{KŠ.65}
\textit{kināčekē hāyš bowe wextē miđyo āmāno lāw tāteyčšo mēw} \\ 
\gll kināčē-ekē hāy=š ∅-b-o=we wext-ē mi-đye-o āmā=n=o lā=ū tāte-ī=č=š=o m-ē=ū \\ 
 girl\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} awareness\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}=and \\ 
\glt `The [king’s] daughter became aware [of the coughing], and noticed that he (the shepherd’s son) had returned; he had returned from the place of her father.'
\z 
 
\ea \label{KŠ.73}
\textit{īnēč dīyāš pore bezeyīš āmā pore} \\ 
\gll īnē=č dīyā=š pore bezeyī=š āmā pore \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=add} look\textsc{.pst}\textsc{=3sg} at pity\textsc{.m}\textsc{=3sg} come\textsc{.pst}\textsc{.3sg} at \\ 
\glt `She looked at him. She took pity on him.'
\z 
 
\ea \label{KŠ.84}
\textit{īnā qāqezēčiš dānā pey} \\ 
\gll īne=ā qāqez-ē=č=iš dā=nā pey \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} letter\textsc{.m}\textsc{-indf}\textsc{=add}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to \\ 
\glt `Look, he has given me a letter.'
\z 
 
\ea \label{KŠ.94}
\textit{yewāšē ewēč milā wero, xeɫk milo pāyšāwe} \\ 
\gll yewāšē ewē=č mi-l-ā ∅-wer-o xeɫk mi-l-o pā-ī=šā=we \\ 
 then 3pl\textsc{.dir}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{ind-}out\textsc{-3sg} people\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} foot\textsc{-obl}\textsc{.m}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `Then, they (the shepherd’s son’s family) went to the front [of the palace]; people welcomed them.'
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{ŽH.11}
\textit{māčo, dey bilmē minīč mewt čene} \\ 
\gll m-āč-o dey bi-l-mē min=īč m-e-ū=t čene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} 1sg\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{2sg}} with \\ 
\glt `He (Hayas) said, ‘Let’s go, I am also coming with you.’'
\z 
 
\ea \label{ŽH.13}
\textit{ewīč māčo, min pilekānī mezānū binyewnew mezānū čēš māčī!’} \\ 
\gll ew=yč m-āč-o min pilekānī me-zān-ū bi-nye-ūne=ū me-zān-ū čēš m-āč-ī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg stair\textsc{\textsc{.f}} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} \textsc{sbjv-}put\textsc{.prs}\textsc{-1sg}=and \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what \textsc{ind-}say\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘I can’t set up a ladder and I don’t understand\footnote{Lit. know.}  what you are saying!’'
\z 
 
\ea \label{ŽH.19}
\textit{āđīč milo mizgī} \\ 
\gll āđ=īč mi-l-o mizgī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} mosque\textsc{.m} \\ 
\glt `He (Hayas) went to the mosque.'
\z 
 
\ea \label{ŽH.46}
\textit{ewīč milowe gēɫnowe kināče wičkilekēšā} \\ 
\gll ew=īč mi-l-o=we ∅-gēɫn-o=we kināčē=e wičkile-ekē=šā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}lead\textsc{.prs}\textsc{-3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3pl} \\ 
\glt `She, i.e. the girl, went back [and] narrated the story [to her sister];'
\z 
 
\ea \label{ŽH.56}
\textit{kināčēčmā gerekene binyeymē were’} \\ 
\gll kināčē=č=mā gerek-e=ne bi-nye-ymē were \\ 
 girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{=add}\textsc{=1pl} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \textsc{sbjv-}put\textsc{.prs} \textsc{1pl} front \\ 
\glt `We want the daughter; we would like to put her in front [for marriage]'
\z 
 
\ea \label{ŽH.57}
\textit{ā wextīyekey āyeč māčo, weɫāhī tāte, dāwāw ēme yoymā kero’} \\ 
\gll ā wext-ī-ekey āye=č m-āč-o weɫāhī tāte dāwā=w ēme yo-ī=mā ∅-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{\textsc{.f}}\textsc{.dir}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed father\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{1pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then she (the eldest daughter) said, ‘Indeed, father, he is asking for the hand of one of us.’'
\z 
 
\ea \label{ŽH.65}
\textit{yewāšē āđīč mārošo mēwe kināčēw luhonī bo teřū čermew īnīšā} \\ 
\gll yewāšē āđ=īč m-ār-o=š=o m-ē=we kināčē=ū luhon-ī ∅-b-o teř=ū čerme=ū īnīšā \\ 
 then 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} fresh=and white=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `He (Hayas) brought her (i.e. his wife). She was from Luhon; she was young, white, and so on.'
\z 
 
\ea \label{ŽH.67}
\textit{āđīč māčo, čiřdēš peym’} \\ 
\gll āđ=īč m-āč-o ∅-čiř-dē=š pey=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}call\textsc{.prs}\textsc{-2pl}\textsc{=3sg} for\textsc{=1sg} \\ 
\glt `He (the Sultan) said, ‘Call him for me.’'
\z 
 
\ea \label{ŽH.72}
\textit{yewāšē ēđīč wēš wezore milo pey yāneyšā} \\ 
\gll yewāšē ēđ=īč wē=š wez-o=re mi-l-o pey yāne-ī=šā \\ 
 then 3sg.prox\textsc{.dir}\textsc{=add} \textsc{reflx}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \\ 
\glt `Then, he (the sultan) prepared himself [and] went towards their (i.e. Hayas’s) house.'
\z 
 
\ea \label{ŽH.78}
\textit{āđīč māčo, min desūrim pēsnen pey wēt āmānā} \\ 
\gll āđ=īč m-āč-o min desūr=im pēsne=n pey wē=t āmā=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg task/order\textsc{.m}\textsc{=1sg} like\_this\textsc{=cop}\textsc{.3sg}\textsc{.m} for \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \\ 
\glt `He said, ‘My task is like this! I have come for you.’'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.112}
\textit{māčo, āđīč čiřoš māčo, čī nelānīwe?’} \\ 
\gll m-āč-o āđ=īč ∅-čiř-o=š m-āč-o čī ne-lā=nī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `He (the sultan) called him (Hayas) [and] said, ‘Why don’t you go back (to your house)?’'
\z 
 
\ea \label{ŽH.116}
\textit{āđīč wāt, ā šēre minnā} \\ 
\gll āđ=īč wāt ā šēr=e min=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{dem.dist} lion\textsc{.m}\textsc{=dem} 1sg\textsc{=cop}\textsc{.1sg} \\ 
\glt `He (the sultan) said, ‘I am that lion.'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.22}
\textit{šūnū ženekēčre ā wextīekey ētir hemāhengīšā weš bīyen} \\ 
\gll šūn=ū ženī-ekē=č=re ā wext-ī-ekey ētir hemāhengī=šā weš bīye=n \\ 
 after\textsc{=ez.gen} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} cooperation\textsc{.m}\textsc{=3pl} good be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `After the marriage, they (the couple's families) were on good terms.'
\z 
 
\ea \label{RE.26}
\textit{zemānēč ženekēš berdēne ētir čo dumā bīyēnē tēkeɫ} \\ 
\gll zemān-ē=č ženī-ekē=š berdē=ne ētir č=o dumā bīyē=nē tēkeɫ \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=add} woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} take\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} \textsc{disc.ptcl} in=\textsc{dem.dist} afterwards be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} mixed \\ 
\glt `When the brides were taken, they (the families) would become one.'
\z 
 
\ea \label{RE.32}
\textit{kuřekeyč qisē kero} \\ 
\gll kuř-ekey=č qisē-ē ∅-ker-o \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} talk\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The boy, too, talks [to the girl]'
\z 
 
\ea \label{RE.51}
\textit{ewīč wāten bā bizānmē} \\ 
\gll ew=īč wāte=n bā bi-zān-mē \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{hort} \textsc{sbjv-}know\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `He (the lord) would say, ‘Let me see.’\footnote{Lit. Let us know.}'
\z 
 
\ea \label{RE.52}
\textit{pīney bāyē māsīč bero} \\ 
\gll p=īney bāye-ē mās=īč ∅-ber-o \\ 
 for=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} pot\textsc{.m}\textsc{-dir}\textsc{.pl} yoghurt\textsc{.m}\textsc{=add} \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `For the reason that he (the girl’s father) would take a bowl of yoghurt [to him] too.'
\z 
 
\ea \label{RE.70}
\textit{zemānēč pāyīz āmāno kewčē xelemā bīyebo teġārē} \\ 
\gll zemān-ē=č pāyīz āmā=n=o kewč-ē xele=mā bīye b-o teġār-ē \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=add} autumn come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} six\_kilos\textsc{-indf} wheat\textsc{.m}\textsc{=\textsc{1pl}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} 120\_kilos\textsc{-indf} \\ 
\glt `We might have had a taghar\footnote{A unit of weight equivalent to 120 kilograms.} of wheat when it was autumn.'
\z 
 
\ea \label{ŽE.1}
\textit{dey čēgeyč qeyīm ʕāɫifšā kennenū} \\ 
\gll dey čēge=īč qeyīm ʕāɫif=šā kenne=n=ū \\ 
 \textsc{disc.ptcl} here\textsc{=add} old\_time grass\textsc{.dir}\textsc{.m}\textsc{=3pl} mow\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `In the past, they (people) mowed grass.'
\z 
 
\ea \label{ŽE.15}
\textit{zimsānīč āmān tā luēnē hāne teqqānū} \\ 
\gll zimsān=īč āmā=n tā luē=nē hāne teqqā=n=ū \\ 
 winter\textsc{.m}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} water\_spring\textsc{.m} dig\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The winter came. They would go and dig [to find] water springs.'
\z 
 
\ea \label{ŽE.23}
\textit{čēgeyč āwī her nebīyēne bīnder} \\ 
\gll čēge=īč āwī her ne-bīyē=ne bīnder \\ 
 here\textsc{=add} water\textsc{.f} emph \textsc{neg-}be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} at\_all \\ 
\glt `There was no water here.'
\z 
 
\ea \label{ŽE.75}
\textit{kuřekeyč ē māčī, kināčekēt miđey pī kuřī?’} \\ 
\gll kuř-ekey=č ē m-āč-ī kināčē-ekē=t mi-đe-ī p=ī kuř-ī \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} come\textsc{.prs}\textsc{.pst}c\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The boy comes. [Addressing her grand son] you would say [to her mother], ‘Will you give your daughter to this boy [in marriage]?’'
\z 
 
\ea \label{ŽM.8}
\textit{īse ħewt sāɫēč āmānāwe serū īney dūkāndārnā} \\ 
\gll īse ħewt sāɫe-ē=č āmā=nā=we ser=ū īney dūkāndār=nā \\ 
 now seven year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} shopkeeper\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{ŽM.32}
\textit{ā wextīyekey ečāge hīč xeberēmā nebīyenū telefūnīčū žimārew telefūnī nebēw} \\ 
\gll ā wext-ī-ekey e=čāge hīč xeber-ē=mā ne-bīye=n=ū telefūn=īč=ū žimāre=ū telefūn-ī ne-b-ē=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} in=there nothing news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and telephone\textsc{.m}\textsc{=add}=and number\textsc{.m}\textsc{=ez.gen} telephone\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and \\ 
\glt `I\footnote{Lit. We.}  had no news [of my family] from there. There was no phone or phone number at the time.'
\z 
 
\ea \label{ŽM.42}
\textit{kuweyt, āđīč wēš kēšwerēwā kuweyt} \\ 
\gll kuweyt āđ=īč wē=š kēšwer-ēw=ā kuweyt \\ 
 \textsc{pn} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{reflx}\textsc{=3sg} country\textsc{-indf}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \\ 
\glt `Kuwait is a country on its own.'
\z 
 
\ea \label{ŽM.56}
\textit{čēweɫ nebīyey bē ētir īseyč xū nīyā} \\ 
\gll čēweɫ nebīyey b-ē ētir īse=yč xū nīye=ā \\ 
 previously poverty\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{disc.ptcl} now\textsc{=add} \textsc{disc.ptcl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the past, there was poverty. But even nowadays, there is poverty.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
