\chapter{All 35 examples of attr in these texts are found in the following sentences:}

\ea \label{ZQ.9}
\textit{vātiš, qurbān luāymē pār yā činnē sāɫē čēweɫter luāymē pey sārāɫū ērānī, koy sān} \\ 
\gll vāt=iš qurbān luā-ymē pār yā činnē sāɫe-ē čēweɫter luā-ymē pey sārāɫ=ū ērān-ī ko=y sān \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} go\textsc{.pst} \textsc{1pl} last\_year either some\textsc{.pl} year.f\textsc{-indf} ago go\textsc{.pst} \textsc{1pl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} where=ez.\textsc{attr} \textsc{pn} \\ 
\glt `He said, ‘Sir, we went, say, last year or a few years ago, we went to the Saral region in Iran, to the San mountain.'
\z 
 
\ea \label{ZQ.42}
\textit{xwāy deseɫātē čāneš bīyen ī pīfēše weše kerdēne} \\ 
\gll xwā-ī deseɫāt=ē čāne=š bīye=n ī pīfē=š=e weš-e kerdē=ne \\ 
 God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m}=ez.\textsc{attr} like\_that\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox} burr.f\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `God’s work was such: he had made this burr; '
\z 
 
\ea \label{HB.19}
\textit{māčo, āy dāxekey dāxim bē min čen bē ʕeqiɫ bīyā!} \\ 
\gll m-āč-o āy dāx-eke=y dāx=im b-ē min čen bē-ʕeqiɫ bī-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} sorrow\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}.\textsc{attr} sorrow\textsc{=1sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{1sg} how\_much without-wisdom be\textsc{.pst}\textsc{-1sg} \\ 
\glt `He (the donkey breeder) said, ‘Oh, I wish I was dead! Look how stupid I was!’'
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{HB.75}
\textit{xeber midā hezretū šēxī} \\ 
\gll xeber mi-de-ā hezret=ū šēx=ī \\ 
 news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \\ 
\glt `They (the sheikh’s servants) passed on the news to his highness the Sheikh.'
\z 
 
\ea \label{HB.78}
\textit{hurmēzo hezretū šēxī pīyāy kīyāno} \\ 
\gll hur-m-ēz-o hezret=ū šēx=ī pīyā-ī ∅-kīyān-o \\ 
 \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} man\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then\footnote{Lit. He rose.}, his highness the Sheikh sent men [to the donkey breeder]'
\z 
 
\ea \label{HB.80}
\textit{čiřoš hezretū šēxī fermāwo, erē řoɫem čī toryānī?} \\ 
\gll ∅-čiř-o=š hezret=ū šēx=ī ∅-fermāw-o erē řoɫe=m čī toryā=nī? \\ 
 \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} child\textsc{.m}\textsc{=1sg} why get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt ` His highness the Sheikh summoned him, i.e. the donkey breeder, [and] said, ‘Oh son! why were you offended? '
\z 
 
\ea \label{HB.85}
\textit{hezretū šēxī fermāwo, dey xāsā bābā herā gā!} \\ 
\gll hezret=ū šēx=ī ∅-fermāw-o dey xās=ā bābā her-ā gā \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} grandfather\textsc{.m} donkey\textsc{.m}\textsc{-obl}\textsc{.pl} fornicate\textsc{.pst} \\ 
\glt `His highness the Sheikh said, ‘That is fine! God damn you!\footnote{Lit. May donkeys fornicate with your ancestors!}'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.49}
\textit{ī gīyānū qefesew sīney minne īnen āđe lūā} \\ 
\gll ī gīyān=ū qefese=ū sīne=y min=ne īne=n āđe lūā \\ 
 \textsc{dem.prox} soul\textsc{.m}\textsc{=ez}\textsc{.gen} cage\textsc{.m}=and chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{\textsc{.f}} go\textsc{.pst} \\ 
\glt ` The soul in my rib cage [that I swore an oath on] was this [sparrow]. It [just] flew away.'
\z 
 
\ea \label{ŽP.1}
\textit{nāmē heqetīneš sey misefān} \\ 
\gll nāmē=ī heqetīne=š sey misefā=n \\ 
 name\textsc{.f}\textsc{=ez}.\textsc{attr} real\textsc{=3sg} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `His real name is Say Mustafa'
\z 
 
\ea \label{ŽP.2}
\textit{nāmē wēš pīr šelīyāre sey misefān} \\ 
\gll nāmē=ī wē=š pīr šelīyār-e sey misefā=n \\ 
 name\textsc{.f}\textsc{=ez}.\textsc{attr} \textsc{reflx}\textsc{=3sg} \textsc{pn} \textsc{pn}\textsc{-def} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s name is [initially] Say Mustafa.'
\z 
 
\ea \label{ŽP.107}
\textit{be zor miđoš milū sey misefāyre zerey} \\ 
\gll be zor mi-đe-o=š mil=ū sey misefā-ī=re zer-e=y \\ 
 by force \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=3sg} on\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} money\textsc{.m}\textsc{-def}\textsc{=ez}.\textsc{attr} \\ 
\glt `He (the man from Hajij) did not take no for an answer\footnote{Lit. with force.} [and] gave him (Pir Shaliyar) the money.'
\z 
 
\ea \label{ŽP.177}
\textit{ī dēwēnē ī kināčēšā bestēne} \\ 
\gll ī dēw=ē=nē ī kināčē=šā bestē=ne \\ 
 dem.prox demon\textsc{.m}\textsc{=ez}.\textsc{attr}\textsc{=cop}\textsc{.3pl} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3pl} tie\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `It was the demons who had muted the girl.'
\z 
 
\ea \label{BP.46}
\textit{xeber yāwo be bābā xwāđāw ħemey ġeybī} \\ 
\gll xeber ∅-yāw-o be bābā xwāđā=ū ħeme=ī ġeybī \\ 
 news\textsc{.dir}\textsc{.m} \textsc{ind-}arrive.prs\textsc{-3sg} to grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible \\ 
\glt `The news reached Baba Khwada and Hama the invisible.'
\z 
 
\ea \label{BP.51}
\textit{ħemey ġeybīw bābāw pīr šelīyārī, bābā xwāđā māčā be xeɫkekey,} \\ 
\gll ħeme=ī ġeybī=ū bābā=ū pīr šelīyār-ī bābā xwāđā m-āč-ā be xeɫk-ekey \\ 
 \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3pl} to people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Hama Ghaybi and Pir Shaliyar’s grandfather, i.e., Baba Khwada, said to the people, '
\z 
 
\ea \label{BP.62}
\textit{je hewrāmānne kārī nāšerʕī kerēnē.’} \\ 
\gll je hewrāmān=ne kār=ī nāšerʕī ker-ēn-ē \\ 
 from \textsc{pn}\textsc{=\textsc{post}} task\textsc{.m}=ez.\textsc{attr} unlawful do.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `They committed unlawful acts in Hawraman.’'
\z 
 
\ea \label{BP.94}
\textit{nāmēwew xerīkū genekārīw kārī nāšerʕīšā kerdenū} \\ 
\gll ne-āmā-ē=we=ū xerīk=ū genekārī=ū kār=ī nāšerʕī=šā kerde=n=ū \\ 
 \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-indf}\textsc{=compl}=and busy\textsc{=ez.gen} debauchery\textsc{.m}=and task\textsc{.m}=ez.\textsc{attr} unlawful\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They didn’t go back [to Iraq]. They commited unlawful acts.'
\z 
 
\ea \label{BP.98}
\textit{ā wextī ēti seʕbe milā milāne mizgī, mizyāre beynē gināre řā beynē bilā pey ʕērāqī jemʕē bā ħemey ġeybīū bābā xuāđā ginā weɫēwe} \\ 
\gll ā wext-ī ēti seʕbe mi-l-ā mi-l-ān=e mizgī mi- zye-ā=re beynē ∅-gin-ā=re řā beynē bi-l-ā pey ʕērāq-ī jemʕ-ē ∅-b-ā ħeme=ī ġeybī=ū bābā xuāđā ∅-gin-ā weɫē=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} morning.f \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}go.prs\textsc{-3pl}=drct mosque\textsc{.m} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} a\_while \textsc{ind-}fall.prs\textsc{-3pl}\textsc{=\textsc{povb}} road.f a\_while \textsc{sbjv-}go.prs\textsc{-3pl} for \textsc{pn}\textsc{-obl}\textsc{.m} added\textsc{-pstc}\textsc{.3sg} \textsc{ind-}be.prs\textsc{-3pl} \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m} \textsc{pn} \textsc{ind-}fall.prs\textsc{-3pl} front\textsc{=\textsc{post}} \\ 
\glt `So, in the morning they went to the mosque. They went a little way towards Iraq. They all gathered. Hama Ghaybi and Baba Khwada went in front of them.'
\z 
 
\ea \label{BP.105}
\textit{bābā xwāđāw ħemey ġeybīū ħemey wirdīkɫe her ginēnē weɫē īnē weɫēne ejošā girđ īnē šūnīšāwe} \\ 
\gll bābā xwāđā=ū ħeme=ī ġeybī=ū ħeme=ī wirdīkɫe her gin-ēn-ē weɫē īnē weɫē=ne ejo=šā girđ īnē šūnī=šā=we \\ 
 grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible=and \textsc{pn}=ez.\textsc{attr} small just fall.prs\textsc{-pstc}\textsc{-3pl} front \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} front\textsc{=\textsc{post}} guess.prs\textsc{=3pl} all \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} track.f\textsc{=3pl}\textsc{=compl} \\ 
\glt `Baba Khwada, Hama the invisible, and little Hama were in front, assuming that everyone was following them.'
\z 
 
\ea \label{BP.113}
\textit{fermāwo ħemey ġeybī, āy wiɫāhī hewrāmī tā meřeber} \\ 
\gll ∅-fermāw-o ħeme=ī ġeybī āy wiɫāhī hewrāmī tā meřeber \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{pn}=ez.\textsc{attr} invisible \textsc{intj} indeed Hawrami until \textsc{pn} \\ 
\glt `Hama Ghaybi said, ‘Well, By God, hawrāmī tā mařabar'
\z 
 
\ea \label{BP.122}
\textit{ā wextī reʕīsū milāre ʕērāq lāw āġekeyšā māčo seremā biřyēnē bābāw pīr šelīyārī bo, ā wextīyekey ħemey ġeybīū yoyč bo nāmē ħemey wirdīkɫe,} \\ 
\gll ā wext-ī reʕīs=ū mi-l-ā=re ʕērāq lā=ū āġe-ekey=šā m-āč-o sere=mā biřyē=nē bābā=ū pīr šelīyār-ī ∅-b-o ā wext-ī-ekey ħeme=ī ġeybī=ū yo=yč ∅-b-o nāmē ħeme=ī wirdīkɫe \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} leader\textsc{.m}\textsc{=ez.gen} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{pn} at\_the\_place\_of\textsc{=ez.gen} agha\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}say.prs\textsc{-3sg} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}=ez.\textsc{attr} invisible=and one\textsc{.m}\textsc{=add} \textsc{ind-}be.prs\textsc{-3sg} name.f \textsc{pn}=ez.\textsc{attr} small \\ 
\glt `The head of those who went to Iraq to the agha, saying, ‘We have decapitated [your men], was Pir Shaliyar’s grandfather, Baba Khwada. In addition [there were] Hama the invisible and Little Hama. '
\z 
 
\ea \label{BP.123}
\textit{ħemey wirdīkɫe ā wextī, pēse īsey māčā heɫāy neyāwān sinʕe zīnāniš nekermē yā čīw yā čāne} \\ 
\gll ħeme=ī wirdīkɫe ā wext-ī pēse īse-ī m-āč-ā heɫāy ne-yāwā=n sinʕe zīnān=iš ne-ker-mē yā čīw yā č=ā=ne \\ 
 \textsc{pn}=ez.\textsc{attr} small \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} like now\textsc{-obl}\textsc{.m} \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} age.f prison\textsc{.m}\textsc{=3sg} \textsc{neg-}do.prs\textsc{-\textsc{1pl}} either thing\textsc{.m} either in=\textsc{dem.dist}\textsc{=\textsc{post}} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) will say, ‘Hama Wirdikla has not reached adulthood; we won’t imprison him, and so on.’'
\z 
 
\ea \label{BP.132}
\textit{ħemey wirdīkɫeyč pēse īse māčā, heɫāy neyāwān bilūqū sinʕēš} \\ 
\gll ħeme=ī wirdīkɫe=yč pēse īse m-āč-ā heɫāy ne-yāwā=n bilūq=ū sinʕe-ē=š \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} like now \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} maturity\textsc{.m}\textsc{=ez.gen} age.f\textsc{-obl}.f\textsc{=3sg} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) said, ‘he hasn’t hit puberty yet.'
\z 
 
\ea \label{BP.138}
\textit{hemey wirdīkɫey mekerā zīnān ewīšā kerā zīnān} \\ 
\gll heme=ī wirdīkɫe-ī me-ker-ā zīnān ewīšā ∅-ker-ā zīnān \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{-obl}\textsc{.m} \textsc{neg-}do.prs\textsc{-3pl} prison\textsc{.m} 3pl\textsc{.obl} \textsc{ind-}do.prs\textsc{-3pl} prison\textsc{.m} \\ 
\glt `They didn’t put Hama Wirdikla in jail. They put them (Baba Khwada and Hama Ghaybi) in jail.'
\z 
 
\ea \label{BP.143}
\textit{ħemey wirdīkɫeyč mičkile bo, wirdīkɫe bo pāne girđ seferē milo pā šārne gēɫo} \\ 
\gll ħeme=ī wirdīkɫe=yč mičkile ∅-b-o wirdīkɫe ∅-b-o p=ā=ne girđ sefer-ē mi-l-o p=ā šār=ne ∅-gēɫ-o \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} small \textsc{ind-}be.prs\textsc{-3sg} small \textsc{ind-}be.prs\textsc{-3sg} at=\textsc{dem.dist}\textsc{=\textsc{post}} all trip\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} at=\textsc{dem.dist} city\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}wander.prs\textsc{-3sg} \\ 
\glt `Hama Wirdikla was small and young. He would constantly go around in the city.'
\z 
 
\ea \label{BP.178}
\textit{āhā māčā īnā čū dīmne lāw beynū zāwerīnew tewēɫēne ħemey ġeybī hen} \\ 
\gll āhā m-āč-ā īnā-∅ č=ū dīm=ne lā=ū beyn=ū zāwer-ī=ne=ū tewēɫē=ne ħeme=ī ġeybī hen-∅ \\ 
 \textsc{intj} \textsc{ind-}say.prs\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{.3sg} side\textsc{.m}\textsc{=\textsc{post}} at\_the\_place\_of\textsc{=ez.gen} among\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{=\textsc{post}} \textsc{pn}=ez.\textsc{attr} invisible exist\textsc{-3sg}\textsc{.m} \\ 
\glt `It is said that, on this side of the border, there is the shrine of Hama the invisible between the villages of Zawar and Tawela.'
\z 
 
\ea \label{BP.181}
\textit{mēnē e čā ħemey ġeybī fermāwo, fermāwo, min mānīyēnmē e čē bā kučē nīšmēre} \\ 
\gll m-ē-nē e=čā ħeme=ī ġeybī ∅-fermāw-o ∅-fermāw-o min mānīyē=nmē e=čē bā kuč-ē ∅-nīš-mē=re \\ 
 \textsc{ind-}come.prs\textsc{-3pl} in=there \textsc{pn}=ez.\textsc{attr} invisible \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{1sg} tired\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} in=here \textsc{hort} little\textsc{-indf} \textsc{sbjv-}sit.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} \\ 
\glt `They went [to the region between Zawar and Tawela]. Hama Ghaybi said, ‘I, we are tired. Let us rest\footnote{Lit. sit.} a little bit here.'
\z 
 
\ea \label{BP.203}
\textit{ā wextī penj šiš neferē je hewrāmāno milā hesārē muweqetīyū ħemey ġeybī kēšā} \\ 
\gll ā wext-ī penj šiš nefer-ē je hewrāmān=o mi-l-ā hesār-ē muweqetī=ū ħeme=ī ġeybī kēšā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} five six person\textsc{.m}\textsc{-indf} from \textsc{pn}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} wall\textsc{-indf} temporary\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible pull\textsc{.pst} \\ 
\glt `Then five, six people went from Hawraman [and] raised a temporary wall around Hama Ghaybi’s shrine.'
\z 
 
\ea \label{BP.205}
\textit{mēnēwe čā dimāy, ženū ħemey ġeybī hāmīlē bo kuřē wīno,} \\ 
\gll m-ē-nē=we čā dimā-ī ženī=ū ħeme=ī ġeybī hāmīlē ∅-b-o kuř-ē ∅-wīn-o \\ 
 \textsc{ind-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} there afterwards\textsc{-obl}\textsc{.m} woman.f=and \textsc{pn}=ez.\textsc{attr} invisible pregnant.f \textsc{ind-}be.prs\textsc{-3sg} son/boy\textsc{.m}\textsc{-indf} \textsc{ind-}see.prs\textsc{-3sg} \\ 
\glt `After they came back [to Hawraman], Hama Ghaybi’s wife, who was pregnant, gave birth to a boy [Lit. She sees a boy].'
\z 
 
\ea \label{BP.206}
\textit{kuřey nāmē minyā mihemeđ} \\ 
\gll kuř-e-ī nāmē=ī mi-nye-ā mihemeđ \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m} name.f=ez.\textsc{attr} \textsc{ind-}put.prs\textsc{-3pl} \textsc{pn} \\ 
\glt `They named the son Muhammad.'
\z 
 
\ea \label{BP.208}
\textit{īne dāstānū īney, īnē dāstānū bābāw pīr šelīyārīū ħemey ġeybī bīyēnē} \\ 
\gll īne dāstāne=ū īney īnē dāstāne=ū bābā=ū pīr šelīyār-ī=ū ħeme=ī ġeybī bīyē=nē \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story.f\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \textsc{pn}=ez.\textsc{attr} invisible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `This was the story of Pir Shaliyar’s grandfather and Hama Ghaybi.'
\z 
 
\ea \label{ŽH.2}
\textit{eyām qisē kerdē wāčē, zemānū qedīmī pādšā hāřūne(l) řešīd bīyenū nā siɫtān mehmūw heyāsī žīr} \\ 
\gll eyā=m qisē-ē kerdē-ē wāč-ē zemān=ū qedīm-ī pādšā hāřūn-e(l) řešīd bīye=n=ū nā siɫtān mehmū=ū heyās=ī žīr \\ 
 mother\textsc{=1sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-3pl} say\textsc{.prs-}pstc\textsc{.3sg} time\textsc{.m}\textsc{=ez}\textsc{.gen} old\_time\textsc{-obl}\textsc{.m} king\textsc{.m} \textsc{pn}-of \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and no sultan\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez}.\textsc{attr} shrewd \\ 
\glt `My mother would narrate [tales]. She said, ‘In the olden days, there was Harun Rashid, the king. No, [there was] Sultan Mahmud and Hayas, the shrewd.'
\z 
 
\ea \label{ŽM.4}
\textit{šūnū āneyre, šūnū řāy eweɫ je kārīgerīwe serem beršīyen} \\ 
\gll šūn=ū āney=re šūn=ū řā=ī eweɫ je kārīgerī=we sere=m ber-šīye=n \\ 
 after\textsc{=ez.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} after\textsc{=ez.gen} road\textsc{.f}=ez.\textsc{attr} first from labourer\_job\textsc{=compl} head\textsc{.m}\textsc{=\textsc{1sg}} out- go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Well, I first learned\footnote{Lit. my head went out.}  to work as a labourer.'
\z 
 
