\chapter{All 30 examples of cmpd in these texts are found in the following sentences:}

\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.19}
\textit{yoyčšā her gīyāne gūwīnēš čenī bē melāqāt} \\ 
\gll yo=yč=šā her gīyān=e gūwīne-ē=š čenī b-ē melāqāt \\ 
 one\textsc{.m}\textsc{=add}\textsc{=3pl} just soul\textsc{.m}=ez\textsc{.cmpd} half\textsc{-indf}\textsc{=3sg} in \textsc{sbjv-}come.prs\textsc{.3sg} visit \\ 
\glt `[The other] one of them was only half alive.\footnote{Lit. had little soul left in him.}    '
\z 
 
\ea \label{ZQ.23}
\textit{yoīčšā her gīyānēš čenī bē gēɫānē qiřoɫe dārēm yoso} \\ 
\gll yo=īč=šā her gīyān-ē=š čenī b-ē gēɫā-ānē qiřoɫ=e dār-ē=m yos=o \\ 
 one\textsc{.m}\textsc{=add}\textsc{=3pl} just soul\textsc{.m}\textsc{-indf}\textsc{=3sg} in \textsc{sbjv-}come.prs\textsc{.3sg} turn\textsc{.pst}\textsc{-\textsc{1sg}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} find\textsc{.pst}\textsc{=compl} \\ 
\glt `As for the one which was half-alive, I wandered around [and] found a tree hollow.'
\z 
 
\ea \label{ZQ.24}
\textit{berdim nīyāmne qiřoɫe dāreke} \\ 
\gll berd-∅=im nīyā=m=ne qiřoɫ=e dār-eke \\ 
 take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1sg}} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `I took [him] and put him in the tree hollow.'
\z 
 
\ea \label{ZQ.26}
\textit{yošā berd nīyāmne qiřoɫe dāreke qurbān!} \\ 
\gll yo=šā berd-∅ nīyā=m=ne qiřoɫ=e dār-eke qurbān \\ 
 one\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{-3sg}\textsc{.m} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} sir.\textsc{voc} \\ 
\glt `Sir, I took one [and] put him in the tree hollow. '
\z 
 
\ea \label{HB.11}
\textit{fermāwo, dā lodē wāčdē be fiɫāne kesī bilo herekey bero} \\ 
\gll ∅-fermāw-o dā ∅-lo-dē wāč-dē be fiɫān=e kes-ī bi-l-o her-ekey ∅-ber-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} \textsc{imp-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} say\textsc{.prs}.sbjv\textsc{-2pl} by such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the Sheikh) said [to the people around him], ‘Go and say to such-and-such a person that [he should] go [and] bring his donkey; '
\z 
 
\ea \label{HB.76}
\textit{māčā, qurbān wiɫā fiɫāne kes toryānū ānā} \\ 
\gll m-āč-ā qurbān wiɫā fiɫān=e kes toryā=n=ū āne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} by\_God such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m} get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \\ 
\glt `They said, ‘Sir, indeed, the fellow has got offended; there he is’'
\z 
 
\ea \label{PM.12}
\textit{īnāymē fiɫāne yāgē} \\ 
\gll īnā-ymē fiɫān=e yāgē \\ 
 \textsc{deic-1pl} such\_and\_such\textsc{=ez}\textsc{.cmpd} place.f \\ 
\glt `We are at such-and-such a place.'
\z 
 
\ea \label{ZP.31}
\textit{māčo, māčā neferē īnārē je menteqew hewrāmānātī be nāmēw fiɫāne kesī} \\ 
\gll m-āč-o m-āč-ā nefer-ē īnā-∅=rē je menteqe=ū hewrāmān-āt-ī be nāmē=ū fiɫān=e kes-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} person\textsc{.m}\textsc{-indf} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} from region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} to name\textsc{.f}\textsc{\textsc{=ez.gen}} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `It was said [to him in the dream] that there is a person in the Hawram region called such-and-such.'
\z 
 
\ea \label{ZP.51}
\textit{ētir kināče keřū lāɫe bo māčmēš pene tewenū sirewē} \\ 
\gll ētir kināčē=e keře=ū lāɫ-e ∅-b-o m-āč-mē=š pene tewenī=ū sirewē \\ 
 \textsc{disc.ptcl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} deaf\textsc{.f}=and mute\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} \\ 
\glt `Since the girl was deaf-mute, we call it the healing stone.'
\z 
 
\ea \label{ZP.84}
\textit{persā, ēme yānew fiɫāne kesīmā gerekā} \\ 
\gll ∅-pers-ā ēme yāne=ū fiɫān=e kes-ī=mā gerek=ā \\ 
 \textsc{ind-}ask\textsc{.prs}\textsc{-3pl} \textsc{1pl} house\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} necessary\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[and] asked [people], ‘We are looking for\footnote{Lit. we want.} the house of such-and-such a person.'
\z 
 
\ea \label{ŽP.30}
\textit{zengene wišɫālē bero} \\ 
\gll zengene=e wišɫāle-ē ∅-ber-o \\ 
 hoe\textsc{.m}\textsc{=ez}\textsc{.cmpd} small\textsc{-indf} \textsc{ind-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `He took a small hoe, too.'
\z 
 
\ea \label{ŽP.157}
\textit{pādšāyč māčo be lālo kināčekē be šālyāre sīyāwīš} \\ 
\gll pādšā=yč m-āč-o be lālo kināčē-ekē be šālyār=e sīyāw-ī=š \\ 
 king\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} by \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `The king said to [his] daughter’s uncle, to Shaliyar Syaw'
\z 
 
\ea \label{ŽP.215}
\textit{āne zerātiš šānī miđo, zerāte sūrī} \\ 
\gll āne zerāt=iš šānī mi-đe-o zerāt=e sūrī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} corn\textsc{=3sg} show \textsc{ind-}give\textsc{.prs}\textsc{-3sg} corn\textsc{=ez}\textsc{.cmpd} Syrian \\ 
\glt `He (Pir Shaliyar) showed him some corn, [some] red corn.'
\z 
 
\ea \label{ŽP.249}
\textit{e wextīyekey yo tiše henār mārow} \\ 
\gll e=wext-ī-ekey yo tiš=e henār m-ār-o=ū \\ 
 in=time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} one\textsc{.m} molasses\textsc{=ez}\textsc{.cmpd} pomegranate\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `One fetched pomegranate molasses.'
\z 
 
\ea \label{ŽP.267}
\textit{ānen īse šelīyāre sīyāwā} \\ 
\gll āne=n īse šelīyār=e sīyāw=ā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[His shrine] is over there now; [he is called] Shaliyar Siyaw.'
\z 
 
\ea \label{BP.77}
\textit{fermāwo, dā lodē bizāndē fiɫāne kes čī nāmā?} \\ 
\gll ∅-fermāw-o dā ∅-lo-dē bi-zān-dē fiɫān=e kes čī ne-āmā-∅ \\ 
 \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{sbjv-}know.prs-\textsc{2pl} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m} why \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `He (Baba Khwada) said, ‘Go [and] see\footnote{Lit. know} why such-and-such person hasn’t come [to the mosque]?'
\z 
 
\ea \label{BP.172}
\textit{hejījīyeke, fermāwo hejījīyeke, be qurbānit bim sūre gīyān yekē tirī lē de leqē teriš de wene’} \\ 
\gll hejījī-eke ∅-fermāw-o hejījī-eke be qurbān=it ∅-b-im sūr=e gīyān yek-ē tir=ī lē ∅-de-e leq-ē ter=iš ∅-de-e wene \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} to sacrifice\textsc{=\textsc{2sg}} \textsc{sbjv-}be.prs\textsc{-\textsc{1sg}} red=ez\textsc{.cmpd} dear.\textsc{voc} one\textsc{-indf} other\textsc{=3sg} at \textsc{imp-}give.prs-\textsc{2sg} kick\textsc{-indf} another\textsc{=3sg} \textsc{imp-}give.prs-\textsc{2sg} at \\ 
\glt `The [prisoner from] Hajij said [to Baba Khwada], ‘May I be your sacrifice, you red-skin, kick it once more! Kick it once more!”'
\z 
 
\ea \label{BP.207}
\textit{īse čiřāš pene mihemeđe wirdīkɫe} \\ 
\gll īse ∅-čiř-ā=š pene mihemeđ=e wirdīkɫe \\ 
 now \textsc{ind-}call.prs\textsc{-3pl}\textsc{=3sg} to \textsc{pn}=ez\textsc{.cmpd} small \\ 
\glt `Nowadays, he is called Little  Muhammad.'
\z 
 
\ea \label{ŽH.22}
\textit{ā wextīyekey, kināče gewrekēš māčo, tāte, her wēt āmānīwe?’} \\ 
\gll ā wext-ī-ekey kināčē=e gewre-ekē=š m-āč-o tāte her wē=t āmā=nī=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} big\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m} just \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `Then, his older daughter said, ‘Father, have you come back alone?’'
\z 
 
\ea \label{ŽH.40}
\textit{wātiš be wāɫe wičkilekēš ī māsī bere deš pā mēmān mizgī čenīw nānī’} \\ 
\gll wāt=iš be wāɫē=e wičkile-ekē=š ī mās-ī ∅-ber-e ∅-de-e=š p=ā mēmān mizgī čenī=ū nān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} by sister\textsc{\textsc{.f}}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} to\textsc{=dem}.dist guest\textsc{.m} mosque\textsc{.m} with\textsc{=ez}\textsc{.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said to her little daughter, ‘Take this yoghurt [and] together with [some] bread give it to the the guest in the mosque.’ '
\z 
 
\ea \label{ŽH.46}
\textit{ewīč milowe gēɫnowe kināče wičkilekēšā} \\ 
\gll ew=īč mi-l-o=we ∅-gēɫn-o=we kināčē=e wičkile-ekē=šā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}lead\textsc{.prs}\textsc{-3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3pl} \\ 
\glt `She, i.e. the girl, went back [and] narrated the story [to her sister];'
\z 
 
\ea \label{RE.18}
\textit{firmānbirđār bīyen, tā sāhībū ī kināčē wāčo, xeber bido wāčo, fiɫāne kār} \\ 
\gll firmānbirđār bīye=n tā sāhīb=ū ī kināčē ∅-wāč-o xeber bi-de-o ∅-wāč-o fiɫān=e kār \\ 
 obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until owner\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} news\textsc{.dir}\textsc{.m} \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} such\_and\_such=ez\textsc{.cmpd} task\textsc{.m} \\ 
\glt `He was obedient. The girl’s father\footnote{Lit. owner.} would say [to him] and ask him, ‘[Do] such-and-such tasks;'
\z 
 
\ea \label{RE.36}
\textit{min, qewɫim dān pey fiɫāne kesī} \\ 
\gll min qewɫ=im dā=n pey fiɫān=e kes-ī \\ 
 \textsc{1sg} arrangement\textsc{.m}\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `I have promised such-and-such a person.'
\z 
 
\ea \label{RE.49}
\textit{min kināčēt dēne kuřū fiɫāne kesī} \\ 
\gll min kināčē=t dē=ne kuř=ū fiɫān=e kes-ī \\ 
 \textsc{1sg} daughter\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}} son/boy\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and say to the lord], ‘I have given my\footnote{Lit. your.} daughter to the son of such-and-such person.'
\z 
 
\ea \label{ŽE.25}
\textit{yānew ētir ne yāne bīyenū} \\ 
\gll yāne=ū ētir ne yāne=e bīye=n=\\ 
 house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{disc.ptcl} no house\textsc{.m}\textsc{=ez}\textsc{.cmpd} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `As for houses, there was no [proper] house.'
\z 
 
\ea \label{ŽE.26}
\textit{denīle girđ yo dilne yāne xākīne bīyen} \\ 
\gll denīle girđ yo dil=ne yāne=e xākī=ne bīye=n \\ 
 family\_of\_ten? all one\textsc{.m} inside\textsc{=\textsc{post}} house\textsc{.m}\textsc{=ez}\textsc{.cmpd} earthen\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[In] a single family, all [the members] lived\footnote{Lit. were.} in an earthen house.'
\z 
 
\ea \label{ŽE.42}
\textit{čirāwe desīšā, čirāwekešā weš kerden bilā pey yānew pīr šelīyārī pey čīwī bārānšo} \\ 
\gll čirāw=e desī=šā čirāw-eke=šā weš kerde=n bi-l-ā pey yāne=ū pīr šelīyār-ī pey čīw-ī b-ār-ān=š=o \\ 
 lamp\textsc{.m}\textsc{=ez}\textsc{.cmpd} hand\textsc{=3pl} lamp\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} for house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} for thing\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `They lit the lamps to go the Pir Shaliyar’s shrine\footnote{Lit. house.} [and] to fetch things.'
\z 
 
\ea \label{ŽE.44}
\textit{denīwlē yāne bīyen ħewt kiɫfetēš bīyēnē} \\ 
\gll denīwlē yāne=e bīye=n ħewt kiɫfet-ē=š bīyē=nē \\ 
 small house\textsc{.m}\textsc{=ez}\textsc{.cmpd} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} seven child\textsc{-indf}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `There was a small house [for us to live]. We were seven children.'
\z 
 
\ea \label{ŽE.73}
\textit{ā wextī to mīsāl wātēt, ā fiɫāne kesem gerekene’} \\ 
\gll ā wext-ī to mīsāl wāt-ē=t ā fiɫān=e kese=m gerek-e=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg} for\_example say\textsc{.pst}\textsc{-cond}\textsc{=\textsc{2sg}} \textsc{dem.dist} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.f}\textsc{=\textsc{1sg}} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `In earlier times, let us say\footnote{Lit. for instance.} you said, ‘I want that certain person (i.e., girl).''
\z 
 
