\chapter{All 441 examples of dem in these texts are found in the following sentences:}

\ea \label{ZB.1}
\textit{řoē je řoā yoe mē yoyč mē xizmetū šā ʕelādīnī, her ā šā ʕelādīne} \\ 
\gll řo-ē je řo-ā yo-e m-ē yo=yč m-ē xizmet=ū šā ʕelādīn-ī her ā šā ʕelādīn=e \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m}\textsc{-f} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} one\textsc{.m}\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} emph \textsc{dem.dist} king\textsc{.m} \textsc{pn}\textsc{=dem} \\ 
\glt `Once,\footnote{Lit. One day of days.} a woman and a man\footnote{Lit. One (f.) came, one (m.) came.} came to the service of šā ʕalādīn, the same šā ʕalādīn.'
\z 
 
\ea \label{ZB.6}
\textit{ā wextīekey čā nīšorew tā hāmin bo} \\ 
\gll ā wext-ī-ekey čā ∅-nīš-o=re=o tā hāmin ∅-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} there \textsc{ind-}sit\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}}\textsc{=compl} until summer\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then, they settled there until the summer.'
\z 
 
\ea \label{ZB.12}
\textit{tejrebeš bo ʕemriš dewrū ṣewū wīs sāɫāne bo wīs sāɫē sī sāɫē pā dewrne} \\ 
\gll tejrebe=š ∅-b-o ʕemr=iš dewr=ū ṣew=ū wīs sāɫe-ā=ne ∅-b-o wīs sāɫe-ē sī sāɫe-ē p=ā dewr=ne \\ 
 experience\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} age\textsc{.m}\textsc{=3sg} around\textsc{\textsc{=ez.gen}} hundred=and twenty year\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} twenty year\textsc{.f}\textsc{-dir}\textsc{.pl} thirty year\textsc{.f}\textsc{-dir}\textsc{.pl} at=\textsc{dem.dist} around\textsc{=\textsc{post}} \\ 
\glt `He was experienced. He was around 120 years old, about 120 or 130,'
\z 
 
\ea \label{ZB.17}
\textit{māčo, be xwā wāro! ānā hewrē sīyāw pise četrē kewt āsmān’} \\ 
\gll m-āč-o be xwā ∅-wār-o āne=ā hewr-ē sīyāw pise četre-ē kewt-∅ āsmān \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to God\textsc{.m} \textsc{ind-}rain\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} cloud\textsc{-indf} black like umbrella\textsc{-obl}\textsc{.f} fall\textsc{.pst}\textsc{-3sg}\textsc{.m} sky\textsc{.m} \\ 
\glt ` He said, ‘By God, it will rain. There, a black cloud has covered the sky like an umbrella!’'
\z 
 
\ea \label{ZB.19}
\textit{jā ā wextī čāwire bo, čāwrekēne fire gināwe kem} \\ 
\gll jā ā wext-ī čāwire ∅-b-o čāwre-ekē=ne fire ∅-gin-ā=we kem \\ 
 afterwards \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} tent\textsc{.f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} tent\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} a\_lot \textsc{ind-}fall\textsc{.prs}\textsc{-3pl}\textsc{=compl} little \\ 
\glt `Back then, there were tents. They tossed around [in bed] in the tent for a while.'
\z 
 
\ea \label{ZB.24}
\textit{šewē be ʕemreū xwāy ženēč nīšore duē zāroɫē wīno her ā šewē} \\ 
\gll šewe-ē be ʕemre=ū xwā-ī ženī-ē=č ∅-nīš-o=re duē zāroɫe-ē ∅-wīn-o her ā šewe-ē \\ 
 night\textsc{.f}\textsc{-obl}\textsc{.f} to order\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m} woman\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=add} \textsc{ind-}sit\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} two child\textsc{-dir}\textsc{.pl} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} emph \textsc{dem.dist} night\textsc{.f}\textsc{-obl}\textsc{.f} \\ 
\glt `During that very night, by the order of God, the wife gave birth to two babies.\footnote{Lit. She sat down [and] saw two babies.} '
\z 
 
\ea \label{ZB.28}
\textit{āwī luēnērew serū āne hārdīšā bīyēnēū} \\ 
\gll āwī luē=nē=re=ū ser=ū āne hārdī=šā bīyē=nē=ū \\ 
 water\textsc{.f} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}}=and on\textsc{\textsc{=ez.gen}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} flour\textsc{.f}\textsc{=3pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `The water had gone up to the level of the flour [sacks] that they had.'
\z 
 
\ea \label{ZB.29}
\textit{kučē qenū čāyū ā wextīekey, ne nān hen, ne čāy hen, ne čīw hen, ħīč} \\ 
\gll kuč-ē qen=ū čāy=ū ā wext-ī-ekey ne nān hen-∅ ne čāy hen-∅ ne čīw hen-∅ ħīč \\ 
 little\textsc{-indf} sugar\_cube=and tea\textsc{.m}=and \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m} nor tea\textsc{.m} exist\textsc{-3sg}\textsc{.m} nor thing\textsc{.m} exist\textsc{-3sg}\textsc{.m} nothing \\ 
\glt `[There was] some sugar and tea, well then, there was neither bread, nor water, nor anything else, [there was] nothing [left]'
\z 
 
\ea \label{ZB.32}
\textit{ā wextīyekey zāroɫē duē bāwšīšāne} \\ 
\gll ā wext-ī-ekey zāroɫe-ē duē bāwšī=šā=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} child\textsc{-dir}\textsc{.pl} two embrace\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `And they held the babies in their arms.'
\z 
 
\ea \label{ZB.35}
\textit{mekiryoyč ī kuřī xinīkno} \\ 
\gll me-kir-īye-o=yč ī kuř-ī ∅-xinīkn-o \\ 
 \textsc{neg-}do.pass-pass\textsc{-3sg}\textsc{=add} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv}-strangle\textsc{.prs}\textsc{-3sg} \\ 
\glt `We should not strangle the [baby] boys.\footnote{Lit. It can't be done that we strangle the baby boys.}'
\z 
 
\ea \label{ZB.37}
\textit{īne min īnem žinīyen lāw, fermāwištū šēxīn, šā ʕelādīnīwe} \\ 
\gll īne min īne=m žinīye=n lā=ū fermāwišt=ū šēx-ī=n šā ʕelādīn-ī=we \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{1sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} speech\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have heard this tale from —it is the Sheikh’s words— Shah Aladin.'
\z 
 
\ea \label{ZB.38}
\textit{ā wextīekey minyoš qiřoɫū beřū dārēweū} \\ 
\gll ā wext-ī-ekey mi-nye-o=š qiřoɫ=ū beřū dār-ēwe=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}put\textsc{.prs}\textsc{-3sg}\textsc{=3sg} hollow\textsc{.m}\textsc{\textsc{=ez.gen}} oak\textsc{.m} tree\textsc{.m}\textsc{-indf}=and \\ 
\glt `Anyhow, he put him in a tree hollow, in an oak tree.'
\z 
 
\ea \label{ZB.40}
\textit{milā ew kuřekey yoyšā berā ānešā zilterū ʕālter bo āđī berā} \\ 
\gll mi-l-ā ew kuř-ekey yo-ī=šā ∅-ber-ā āne=šā zil-ter=ū ʕāl-ter ∅-b-o āđī ∅-ber-ā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3pl} big\textsc{-cmpr}\textsc{\textsc{=ez.gen}} good\textsc{-cmpr} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} 3sg\textsc{.obl}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \\ 
\glt `They went away [and took] that son. They took one of them (i.e., the boys), the one who was bigger and healthier; they took him.'
\z 
 
\ea \label{ZB.42}
\textit{čī mede xwāwinū ʕālemīyānī pāse īrāyešā bizēw menēne jīāy sāyqekewe} \\ 
\gll č=ī mede xwāwin=ū ʕālemīyān-ī pāse īrāye=š=ā bize-ēw menē=ne jīāy sāyqe-eke=we \\ 
 at=\textsc{dem.prox} time owner\textsc{\textsc{=ez.gen}} worlds\textsc{.m}\textsc{-obl}\textsc{.m} like volition\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} goat\textsc{.f}\textsc{-indf} remain\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} behind lightning\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `The will of the lord of all worlds was such that a goat was accidentally left behind [healthy] from the lightning [that caused the flood]'
\z 
 
\ea \label{ZB.50}
\textit{ā wextīekey īđīčšā ārdenū} \\ 
\gll ā wext-ī-ekey īđī=č=šā ārde=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{.obl}\textsc{.m}\textsc{=add}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They (the family) brought him back.'
\z 
 
\ea \label{ZB.54}
\textit{ī bizēče powe āwēzāne bo mē e čē ēđ čeknošū milo lā yāgekēš} \\ 
\gll ī bizē=č=e p=o=we āwēzān-e ∅-b-o m-ē e=čē ēđ čekn-o=š=ū mi-l-o lā yāgē-ekē=š \\ 
 \textsc{dem.prox} goat\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=dem} at=\textsc{dem.dist}\textsc{=compl} hanging\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} in=here 3sg.prox\textsc{.dir} suckle\textsc{.prs}\textsc{-3sg}\textsc{=3sg}=and \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of place\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \\ 
\glt `The goat would stay in a hanging position, the child would feed from its udder, and it would go back to its place.'
\z 
 
\ea \label{ZB.58}
\textit{īne dāstānū gewregerīū ā mešeyxāne} \\ 
\gll īne dāstāne=ū gewregerī=ū ā mešeyxā=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{\textsc{=ez.gen}} greatness\textsc{\textsc{=ez.gen}} \textsc{dem.dist} sheikh\textsc{.obl}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `This is the story of the greatness of the sheikhs.'
\z 
 
\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.10}
\textit{čowe sārāɫū ērānīwe gēɫāymēwe} \\ 
\gll č=o=we sārāɫ=ū ērān-ī=we gēɫā-ymē=we \\ 
 in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=compl} turn\textsc{.pst} \textsc{1pl}\textsc{=compl} \\ 
\glt `On the way back, we returned from the Saral region of Iran.'
\z 
 
\ea \label{ZQ.11}
\textit{wextū ī jimāymārā āmāymēwe āmāymē kelēwe} \\ 
\gll wext=ū ī jimāy=mā=rā āmā-ymē=we āmā-īmē kel-ēwe \\ 
 time\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} move\textsc{.inf}\textsc{=\textsc{1pl}}\textsc{=\textsc{povb}} come\textsc{.pst} \textsc{1pl}\textsc{=compl} come\textsc{.pst} \textsc{1pl} mountain\textsc{-indf} \\ 
\glt `On our return (journey), we went\footnote{Lit. came.}  to a mountain. '
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{ZQ.29}
\textit{āne ke berdmā šiš māngē menn} \\ 
\gll āne ke berd-∅=mā šiš mānge-ē menn \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{rel} take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{1pl} six month.f\textsc{-dir}\textsc{.pl} remain\textsc{.pst} \\ 
\glt `The one whom we took [with us] lived for six months.'
\z 
 
\ea \label{ZQ.34}
\textit{her pā henāhur luāymēwe} \\ 
\gll her p=ā hen-ā=hur luā-ymē=we \\ 
 just at=\textsc{dem.dist} side\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} go\textsc{.pst} \textsc{1pl}\textsc{=compl} \\ 
\glt `We went back by the same route [that we had taken a year before]\footnote{Lit. came back} by the same route [that we had taken a year before]'
\z 
 
\ea \label{ZQ.39}
\textit{wextē luānē werew temāšem kerd miđyew be ʕemrew xwāy e čī qiřolī| ēnnew čīčew ženē pīfē peydā bīyenē| ʕeyn pīfē} \\ 
\gll wext-ē luā-ānē wer=ew temāše=m kerd mi-đye-ū be ʕemre=ū xwā-ī ečī qiřoɫ-ī ēnne=w čīče=ū ženī-ē pīfē-ē peydā bīyē=nē ʕeyn pīfē \\ 
 time\textsc{.m}\textsc{-indf} go\textsc{.pst}\textsc{-\textsc{1sg}} front\textsc{.m}\textsc{=\textsc{post}} looking\textsc{=\textsc{1sg}} do\textsc{.pst} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} by order.f\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} in.\textsc{dem.prox} hollow\textsc{.m}\textsc{-obl}\textsc{.m} as\_much\_as\textsc{=ez.gen} nipple\textsc{=ez.gen} woman.f\textsc{-obl}.f burr.f\textsc{-dir}\textsc{.pl} visible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like balloon.f \\ 
\glt `When I got closer and looked inside, I noticed that, by the order of God, some burrs the size of a woman's nipple had appeared in the tree hollow, like balloons.'
\z 
 
\ea \label{ZQ.41}
\textit{āne memkeš īnā je pīfekēwe jīyā bīyeno} \\ 
\gll āne memke=š īnā-∅ je pīfē-ekē=we jīyā bīye=n=o \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} breast\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} from burr.f\textsc{-def}\textsc{.obl}.f\textsc{=compl} separate be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `The breast was clearly distinguished from the burr. '
\z 
 
\ea \label{ZQ.42}
\textit{xwāy deseɫātē čāneš bīyen ī pīfēše weše kerdēne} \\ 
\gll xwā-ī deseɫāt=ē čāne=š bīye=n ī pīfē=š=e weš-e kerdē=ne \\ 
 God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m}=ez.\textsc{attr} like\_that\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox} burr.f\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `God’s work was such: he had made this burr; '
\z 
 
\ea \label{ZQ.44}
\textit{ī kuře her memeš wārden her memes wārden} \\ 
\gll ī kuř=e her meme=š wārde=n her meme=š wārde=n \\ 
 \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} just breast\textsc{.dir}\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} just breast\textsc{.dir}\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The boy had kept breast-feeding [from the tree]; he had kept breast-feeding .'
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{ZQ.55}
\textit{fermāwo, dey āne ētir deseɫāt xwāy, pey her čē meyliš bē xwāy deseɫātē čānēš henē} \\ 
\gll ∅-fermāw-o dey āne ētir deseɫāt xwā-ī pey herčē meyl=iš b-ē xwā-ī deseɫāt-ē čāne-ē=š hen-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} power\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} to whatever willingness\textsc{=3sg} \textsc{sbjv-}come.prs\textsc{.3sg} God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m-dir.pl} such\textsc{-pl}\textsc{=3sg} \textsc{exist}\textsc{-3pl} \\ 
\glt ` The Shah said, ‘Well, then, this is in God’s hands. Whatever God wishes, he has the power [to exert]'
\z 
 
\ea \label{HB.2}
\textit{nānewēkēw šēh ʕelādīnī seʕbē hurmēzāwe xeber miđā māčā, xeber deydē be hezretū šēxī, be xwā xāneqāne ne nān hen, ne hārdī henē} \\ 
\gll nānewē-ekē=ū šēh ʕelādīn-ī seʕbe-ē hur-mi-ēz ā=we xeber mi-đe-ā mi-āč-ā xeber ∅-de-ydē be hezret=ū šēx-ī be xwā xāneqā=ne=ne nān hen-∅=ne hārdī hen-ē \\ 
 baker\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{.pl}=and sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} morning\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs} \textsc{dem.dist}\textsc{=compl} news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} news\textsc{.m} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} to his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}}=neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m}=nor flour\textsc{\textsc{.f}} exist\textsc{-3pl} \\ 
\glt `Sheikh Aladin’s bakers got up in the morning and passed on the news, saying, ‘Pass on the news to his highness the Sheikh. By God, there is neither bread nor flour in the monastery. '
\z 
 
\ea \label{HB.8}
\textit{bāɫānisbet ā wextīye, be weɫāx be} \\ 
\gll bāɫānisbet ā wext-ī=e be weɫāx be \\ 
 God\_forbid \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=dem} by horse/ donkey\textsc{.m} by \\ 
\glt `May it be far [from you]; back then, they had a donkey for carrying loads.'
\z 
 
\ea \label{HB.17}
\textit{miđyo herzretū šēxī nefermāwā bilue yānew āney bāre’} \\ 
\gll mi-đye-o herzret=ū šēx-ī ne-fermāwā bi-lue yāne=ū āney b-ār-e \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}say\textsc{.pst} \textsc{sbjv-}go\textsc{.imp}\textsc{.\textsc{2sg}} house=and \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `He noticed that his highness the Sheikh hadn’t told [him], ‘Go to the house of that (i.e. such-and-such) person and bring [back the flour]’'
\z 
 
\ea \label{HB.23}
\textit{ī gīre čēš bī min wārd?’} \\ 
\gll ī gīr=e čēš bī-∅ min wārd \\ 
 \textsc{dem.prox} hook\textsc{.m}\textsc{=dem} what be\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{1sg} eat\textsc{.pst} \\ 
\glt `What is this situation that I am caught in?’'
\z 
 
\ea \label{HB.28}
\textit{xeylē rāziɫū diɫīš kerdēš} \\ 
\gll xeylē rāziɫ=ū diɫ-ī=š kerd-ē=š \\ 
 so much request\textsc{=ez}\textsc{.gen} heart\textsc{.m} \textsc{dem.prox}\textsc{=3sg} do\textsc{.pst} 3pl\textsc{=3sg} \\ 
\glt `He struggled a lot.'
\z 
 
\ea \label{HB.34}
\textit{čowe neferēwe mē, pēwāymāwe} \\ 
\gll č=o=we nefer-ēwe- ē pēwāy=mā=we \\ 
 in\textsc{=dem}.dist\textsc{=\textsc{post}} person\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} welcoming\textsc{.m}\textsc{=1pl}\textsc{=\textsc{post}} \\ 
\glt `There, a person will welcome us.'
\z 
 
\ea \label{HB.44}
\textit{hereke mēwe juwāb ī qisā kero} \\ 
\gll her-eke m-ē=we juwāb ī qisē-ā ∅-ker-o \\ 
 donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} reply\textsc{.m} \textsc{dem.prox} talk\textsc{\textsc{.f}}\textsc{-obl}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The donkey started to speak; it spoke.'
\z 
 
\ea \label{HB.49}
\textit{yo kābrāēwe čowewe āmān řākēre} \\ 
\gll yo kābrā-ēwe č=owe=we āmā=n řā-ekē=re \\ 
 one\textsc{.m} man\textsc{-indf} in\textsc{=dem}.dist\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} road\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \\ 
\glt `A man came from the other direction onto their road'
\z 
 
\ea \label{HB.59}
\textit{ā wextīyekey jimēre kābrā kot gelšā} \\ 
\gll ā wext-ī-ekey ∅-jimā-ē=re kābrā kot gel=šā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}move\textsc{.pst}\textsc{-3pl}\textsc{=\textsc{povb}} man fall\textsc{.pst}\textsc{.3sg}\textsc{.m} with\textsc{=3pl} \\ 
\glt `Then, when they returned, the fellow accompanied them.'
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{HB.71}
\textit{ī herī zūwāniš zānā min hīčim nezānān!} \\ 
\gll ī her-ī zūwān=iš zānā min hīč=im ne-zānā=n \\ 
 \textsc{dem.prox} donkey\textsc{.m}\textsc{-obl}\textsc{.m} language\textsc{.m}\textsc{=3sg} know\textsc{.pst}\textsc{.ptcp}\textsc{.m} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The donkey knew the [Sheikh’s] language; I didn’t know a thing!'
\z 
 
\ea \label{HB.76}
\textit{māčā, qurbān wiɫā fiɫāne kes toryānū ānā} \\ 
\gll m-āč-ā qurbān wiɫā fiɫān=e kes toryā=n=ū āne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} by\_God such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m} get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \\ 
\glt `They said, ‘Sir, indeed, the fellow has got offended; there he is’'
\z 
 
\ea \label{HB.90}
\textit{dā dey hurbēze ī mexlūqī girdiš gēɫeš pone} \\ 
\gll dā dey hur-b-ēz-e ī mexlūq-ī gird=iš gēɫ-e=š pone \\ 
 \textsc{hort} \textsc{disc.ptcl} \textsc{pvb-}\textsc{sbjv-}rise\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} people\textsc{-obl}\textsc{.m} all\textsc{=3sg} wander\textsc{.prs}-\textsc{2sg}\textsc{=3sg} at \\ 
\glt `Come on, get up and search among all these people,'
\z 
 
\ea \label{HB.93}
\textit{hāɫīm nekerdenī řās, ānīšā} \\ 
\gll hāɫī=m ne-kerde=nī řās ānīšā \\ 
 understood\textsc{=1sg} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} really 3pl.dem\textsc{.obl} \\ 
\glt `Haven’t I really made you understand? and so on.'
\z 


\ea \label{PM.5}
\textit{fermāndew merekuřekey, reʕīsekešā māčo, ēme bē ā šexsī metāwmē, e čē menmēwe zerer deymē čī mentēqey’} \\ 
\gll fermānde=ū merekuř-ekey reʕīs-eke=šā m-āč-o ēme bē ā šexs-ī me-tāw-mē e=čē ∅-men-mē=we zerer ∅-de-ymē č=ī mentēqe-ī \\ 
 leader\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} leader\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} in=here \textsc{sbjv-}stay\textsc{.prs}\textsc{-1pl}\textsc{=compl} harm\textsc{.m} \textsc{sbjv-}give\textsc{.prs}-\textsc{1pl} in=\textsc{dem.prox} region\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The head of the grasshoppers said, ‘Without [permission from] this person we cannot stay here and do harm to this region.'
\z 
 
\ea \label{PM.17}
\textit{ānā ēme wēmā mehelēmā hen tūl} \\ 
\gll āne=ā ēme wē=mā mehele-ē=mā hen-∅ tūl \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{1pl} \textsc{reflx}\textsc{=1pl} neighbourhood\textsc{.m}\textsc{-indf}\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} long \\ 
\glt `There, I, myself, own a big neighbourhood.’\footnote{Lit. We, ourselves, have a big neighbourhood.}'
\z 
 
\ea \label{PM.18}
\textit{muɫkēšā bīyebo čāne temāmū muɫkekey kuliš je zerāʕetne bīyebo āmābonew} \\ 
\gll muɫk-ē=šā bīye b-o č=ā=ne temām=ū muɫk-ekey kul=iš je zerāʕet=ne bīye b-o āmā b-one=ū \\ 
 property\textsc{.m}\textsc{-indf}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} in=\textsc{dem.dist}\textsc{=\textsc{post}} all\textsc{\textsc{=ez.gen}} property\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} all\textsc{=3sg} from agriculture\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `—īt appears that he\footnote{they} had a big property. The property might have been filled with crops; and they might have been ripe.'
\z 
 
\ea \label{PM.20}
\textit{fermāwebē, fermāwo lodē e čā muɫkū ēme wēmā, herčēmā hen čāne ā firmānde wēš zāno heqū ēme kāmā} \\ 
\gll fermāwe b-ē ∅-fermāw-o ∅-lo-dē e=čā muɫk=ū ēme wē=mā herčē=mā hen-∅ č=ā=ne ā firmānde wē=š ∅-zān-o heq=ū ēme kām=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} in=there property\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{reflx}\textsc{=1pl} whatever\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} leader\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} right\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} which\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go! In that property of mine, whatever we possess there, your boss knows which portion [of land] is mine.'
\z 
 
\ea \label{PM.29}
\textit{māčo qurbān īne čēšit kerdū?} \\ 
\gll m-āč-o qurbān īne čēš=it kerd=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=\textsc{2sg}} do\textsc{.pst}=and \\ 
\glt `They said, ‘Sir, what have you done?'
\z 
 
\ea \label{PM.30}
\textit{īne bē ījāzew to īne nekiryān} \\ 
\gll īne b-ē ījāze=ū to īne ne-kir-yā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} permission\textsc{\textsc{=ez.gen}} \textsc{2sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{neg-}do.pass-past.pass\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This couldn’t have been done without your permission.'
\z 
 
\ea \label{PM.32}
\textit{īne čī e pēse?} \\ 
\gll īne čī e=pēse \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} why in=like \\ 
\glt `Why is this situation like this?'
\z 
 
\ea \label{PM.34}
\textit{nāřihet nebī āne meʕmūre bē terefū xwāywe āmābē} \\ 
\gll nāřihet ne-b-ī āne meʕmūre b-ē teref=ū xwā-ī=we āmā b-ē \\ 
 sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} agent be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Don’t worry! It was an agent [who] was sent  by God.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.39}
\textit{hīčkāmmā mihtājē mebīmē pīney} \\ 
\gll hīčkām=mā mihtāj-ē me-b-īmē p=īney \\ 
 no\_one\textsc{=1pl} needy\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs} \textsc{1pl} by=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \\ 
\glt `None of us would be in need of these [products] '
\z 
 
\ea \label{PM.44}
\textit{werzīyereke māčo, weɫāhī šēxī ʕelādīne řāsiš fermāwān} \\ 
\gll werzīyer-eke m-āč-o weɫāhī šēx-ī ʕelādīn=e řās=iš fermāwā=n \\ 
 farm\_labourer\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{=dem} truth\textsc{=3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The farm-labourer said; ‘Indeed, Sheikh Aladin was right.'
\z 
 
\ea \label{PM.45}
\textit{min īse ī nīštenāre īsrāhetīč kerū} \\ 
\gll min īse ī nīšte=nā=re īsrāhet=īč ∅-ker-ū \\ 
 \textsc{1sg} now \textsc{dem.prox} sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=\textsc{povb}} rest\textsc{.m}\textsc{=add} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Now I am seated and I am resting.'
\z 
 
\ea \label{PM.46}
\textit{čowe ew mē peym māro} \\ 
\gll č=o=we ew m-ē pey=m m-ār-o \\ 
 in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From that direction, he  comes [to me] and brings [food] for me'
\z 
 
\ea \label{PM.47}
\textit{ēđ čowe mē peym māro} \\ 
\gll ēđ č=o=we m-ē pey=m m-ār-o \\ 
 3sg.prox\textsc{.dir} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From this direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.48}
\textit{ew čowe mē peym māro} \\ 
\gll ew č=o=we m-ē pey=m m-ār-o \\ 
 \textsc{dem.dist} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From that direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.51}
\textit{īne heɫbet xwāy meylišā īne} \\ 
\gll īne heɫbet xwā-ī meyl=š=ā īne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} obviously God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \\ 
\glt `This is indeed God’s will'
\z 
 
\ea \label{PM.52}
\textit{īne heɫbet xwāy meylšā} \\ 
\gll īne heɫbet xwā-ī meyl=š=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} obviously God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is indeed God’s will’'
\z 
 
\ea \label{DG.3}
\textit{ā wextīekey yerē kuřēš bīyēnē} \\ 
\gll ā wext-ī-ekey yerē kuř-ē=š bīyē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `[Shah Aladin] had three sons.'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.6}
\textit{ā wextīekey lūēnmē e wāge duʕāw xeyrīš kerdēnew} \\ 
\gll ā wext-ī-ekey lūē=nmē e=wāge duʕā=ū xeyr-ī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} in=there prayer\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} goodness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `We would go there [i.e. to them and] they would bless us'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.13}
\textit{derdekēš, ā wextīekey, derdū gulī bo, dūrū gošdārā} \\ 
\gll derde-ekē=š ā wext-ī-ekey derde=ū gulī ∅-b-o dūr=ū gošdār-ā \\ 
 illness\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy \textsc{ind-}be\textsc{.prs}\textsc{-3sg} far\textsc{=ez}\textsc{.gen} listener\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `May it be far from the audience, his illness was leprosy.'
\z 
 
\ea \label{DG.15}
\textit{ā wextīyekey, birākeyš ber kerā} \\ 
\gll ā wext-ī-ekey birā-ekey=š ber ∅-ker-ā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} out \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \\ 
\glt `His brother was expelled [from the village]'
\z 
 
\ea \label{DG.21}
\textit{ā wextīyekey tā yerē sāɫē e čā kešo āđ guzerān kero lā birākeyšo} \\ 
\gll ā wext-ī-ekey tā yerē sāɫe-ē e=čā keš=o āđ guzerān ∅-ker-o lā birā-ekey=š=o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} until three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} in=there mountain\textsc{.m}\textsc{=\textsc{post}} 3sg\textsc{.dir}\textsc{.m} subsistence\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} at\_the\_place\_of brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `For three years, he stayed with his brothers in those mountains '
\z 
 
\ea \label{DG.23}
\textit{ā wextīyekey, nimāzā bilo āweyānī} \\ 
\gll ā wext-ī-ekey ni-m-āz-ā bi-l-o āweyānī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} habitat\textsc{\textsc{.m}} \\ 
\glt `People wouldn’t let him go to the village.'
\z 
 
\ea \label{DG.26}
\textit{to mišo ī birāyte šewēne bere} \\ 
\gll to mišo ī birā-ī=t=e šewe-ē=ne ∅-ber-e \\ 
 \textsc{2sg} aux \textsc{dem.prox} brother\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `You should take this brother of yours [there] at night.'
\z 
 
\ea \label{DG.33}
\textit{xāneqāw bīyārēne, ā wextīyekey dāđ kero} \\ 
\gll xāneqā=ū bīyārē=ne ā wext-ī-ekey dāđ ∅-ker-o \\ 
 monastery\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yell\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He gave a yell of despair at the monastery of Biyare.'
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DG.38}
\textit{ā wextīyekey, ētir, īse, min yerē sāɫē ī birāme tūšū ī derdēe āmānū} \\ 
\gll ā wext-ī-ekey ētir īse min yerē sāɫe-ē ī birā=m=e tūš=ū ī derde-ē=e āmā=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} encounter\textsc{=ez}\textsc{.gen} \textsc{dem.prox} illness\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=dem} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Now, as for me, it’s been three years that my brother is suffering from this illness. '
\z 
 
\ea \label{DG.43}
\textit{ī birāme derdeš, derdēš īnā po derdū gulīš māčā dāđē!} \\ 
\gll ī birā=m=e derde=š derde-ē=š īnā-∅ p==oderde=ū gulī=š m-āč-ā dāđ-ē \\ 
 \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} illness\textsc{\textsc{.f}}\textsc{=3sg} illness\textsc{\textsc{.f}}\textsc{-indf}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at=\textsc{=\textsc{post}}illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} yell\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `This brother of mine is ill; it is called leprosy; Please help!'
\z 
 
\ea \label{DG.46}
\textit{šēx mutewej bo, wextē miđyo derdū ī pīyāy lā xwāyū xwāwinū ʕāɫemīyānī mutewej bo, řānimonīš kero} \\ 
\gll šēx mutewej ∅-b-o wext-ē mi-đye-o derde=ū ī pīyā-ī lā xwā-ī=ū xwāwin=ū ʕāɫemīyān-ī mutewej ∅-b-o řānimonī=š ∅-ker-o \\ 
 sheikh\textsc{.m} aware \textsc{ind-}be\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} man\textsc{.m}\textsc{-obl}\textsc{.m} at\_the\_place\_of God\textsc{.m}\textsc{-obl}\textsc{.m}=and owner\textsc{=ez}\textsc{.gen} worlds\textsc{.m}\textsc{-obl}\textsc{.m} aware \textsc{ind-}be\textsc{.prs}\textsc{-3sg} guidance\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The Sheikh understood. When he looked at the man’s illness, the lord of all universes realised [what] his illness [was and] guided him (i.e. the Sheikh).'
\z 
 
\ea \label{DG.50}
\textit{winekēš sāwe pā qelbū pīyāyre} \\ 
\gll winī-ekē=š ∅-sāw-e p=ā qelb=ū pīyā-ī=re \\ 
 blood\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} by\textsc{=dem}.dist heart\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `[then] rub its blood on the heart of the [sick] man'
\z 
 
\ea \label{DG.51}
\textit{ī pose āne gino, pos māro nāwe} \\ 
\gll ī pos=e āne ∅-gin-o pos m-ār-o nāwe \\ 
 \textsc{dem.prox} skin\textsc{.m}\textsc{=dem} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} ind\textsc{-f}all\textsc{.prs}\textsc{-3sg} skin\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} inside \\ 
\glt `his skin will fall off, and [a new] skin will replace it'
\z 
 
\ea \label{DG.53}
\textit{šā ʕelādīn īney īney māčo be, fermāwo be pīyā, hīnī birāw pīyākey} \\ 
\gll šā ʕelādīn īney īney m-āč-o be ∅-fermāw-o be pīyā hīn-ī birā=ū pīyā-ekey \\ 
 king\textsc{.m} \textsc{pn} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by man\textsc{.m} \textsc{fill}\textsc{-obl}\textsc{.m} brother\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Shah Aladin said this to the man, to the brother of the [sick] man.'
\z 
 
\ea \label{DG.55}
\textit{ā wextīyekey īne vīyero, qerār bo} \\ 
\gll ā wext-ī-ekey īne vīyer-o qerār ∅-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} pass\textsc{.prs}\textsc{-3sg} arrangement \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then, this was settled. The deal was ...'
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.58}
\textit{ā wextīyekey, gēɫdē lodē penā bere be xwāy} \\ 
\gll ā wext-ī-ekey ∅-gēɫ-dē ∅-lo-dē penā ∅-ber-e be xwā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}wander\textsc{.prs}\textsc{-2pl} \textsc{imp-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} refuge \textsc{imp-}take\textsc{.prs}-\textsc{2sg} by God\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, go back [to your place]; Take refuge with God.'
\z 
 
\ea \label{DG.59}
\textit{gāhes pī qisē mine xwā řehm kero, weš bo} \\ 
\gll gāhes p=ī qisē-ē min=e xwā řehm ∅-ker-o weš ∅-b-o \\ 
 maybe by\textsc{=dem}.prox talk\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{1sg}\textsc{=dem} God\textsc{.m} mercy\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} well \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Through my intervention may God have mercy on him so that he gets well.'
\z 
 
\ea \label{DG.60}
\textit{šūnū īneyre āhā lā xwāwe inne duʕāš meqboɫe bo} \\ 
\gll šūn=ū īney=re āhā lā xwā=we inne duʕā=š meqboɫ-e ∅-b-o \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} \textsc{intj} at\_the\_place\_of God\textsc{.m}\textsc{=\textsc{post}} so\_much prayer\textsc{\textsc{.f}}\textsc{=3sg} acceptable\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well then, his prayers were accepted by God.'
\z 
 
\ea \label{DG.64}
\textit{ā wextīekey beɫē, min bilo ā memleketū wēšā, kiseɫīš čenē, ēme zāmdārē nekero ēme kečē nekero} \\ 
\gll ā wext-ī-ekey beɫē min bi-l-o ā memleket=ū wē=šā kiseɫī=š čenē ēme zāmdār-ē ne-ker-o ēme keč-ē ne-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yes \textsc{1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3pl} tortoise\textsc{\textsc{.f}}\textsc{=3sg} in \textsc{1pl} wounded\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `[The Sheikh said], ‘Yes!' '[The tortoise said] I [say] he [should] go to his region; there are tortoises there. He shall not injure us; he shall not make us crooked. '
\z 
 
\ea \label{DG.67}
\textit{šā ʕelādīnīč fermāwo birākem wextē īnē zānā min gunākē māy mē milimre} \\ 
\gll šā ʕelādīn-ī=č ∅-fermāw-o birā-eke=m wext-ē īnē ∅-zān-ā min gunā-ekē m-āy m-ē mil=im=re \\ 
 king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} time\textsc{.m}\textsc{-indf} \textsc{dem.prox}\textsc{.3pl} \textsc{ind-}know\textsc{.prs}\textsc{-3pl} \textsc{1sg} sin\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}come\textsc{.pst}\textsc{.adv} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} shoulder\textsc{.m}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `Shah Aladin said, ‘Brother!—as they (i.e. the tortoises) know [about the medicine] I won’t shoulder the burden of injuring them.'
\z 
 
\ea \label{DG.70}
\textit{kuře wextē miđyo ānā čā šāzāyēwe} \\ 
\gll kuř-e wext-ē mi-đye-o āne=ā čā šāzā-ēwe \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.dir} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} there prince\textsc{.m}\textsc{-indf} \\ 
\glt `The [sick] boy has become like a prince.'
\z 
 
\ea \label{DG.73}
\textit{īnē dāstānū ī šēxāū ēme bīyen gewregerīšā} \\ 
\gll īnē dāstāne=ū ī šēx-ā=ū ēme bīye=n gewregerī=šā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{\textsc{.f}}\textsc{.3sg} story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} sheikh\textsc{.m}\textsc{-obl}\textsc{.pl}=\textsc{ez.gen} \textsc{1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} greatness\textsc{=3pl} \\ 
\glt `This is the story of our Sheikhs and their greatness.'
\z 
 

\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.5}
\textit{āmān gerekiš bīyen ī memliketī dāgīr kero pey wēš} \\ 
\gll āmā=n gerek=iš bīye=n ī memliket-ī dāgīr ∅-ker-o pey wē=š \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} necessary\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} occupied \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} for \textsc{reflx}\textsc{=3sg} \\ 
\glt `He wanted to occupy this region [and control it] for his own interest.'
\z 
 
\ea \label{DP.9}
\textit{āmān lūān ā qeɫāw hewrāmānī e pāgewe nīštenre} \\ 
\gll āmā=n lūā=n ā qeɫā=w hewrāmān-ī e=pāgewe nīšte=n=re \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} in=there sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} \\ 
\glt `He came [here and] went to the citadel of Hawraman over there and settled there.'
\z 
 
\ea \label{DP.12}
\textit{ā wextī ī jemšēr šā, sibhān āġey čew deso bīyen} \\ 
\gll ā wext-ī ī jemšēr šā sibhān āġe-ī č=ew des=o bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} \textsc{pn} king\textsc{.m} \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} in\textsc{=dem}.dist hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Jamsher Shah, eh … Sibhan Agha was on the other side [of the village].'
\z 
 
\ea \label{DP.13}
\textit{dāwāw āġeyīš kerdēne} \\ 
\gll dāwā=w āġe-ī=š kerdē=ne \\ 
 demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} agha\textsc{.m}\textsc{-nmlz}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `He (i.e., Jamsher Shah) claimed governorship [over] Hawraman. '
\z 
 
\ea \label{DP.14}
\textit{ī memlikete desū āđīwe bīyen} \\ 
\gll ī memliket=e des=ū āđī=we bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=dem} hand\textsc{.m}\textsc{=ez}\textsc{.gen} 3sg\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This region was under Sibhan Agha’s control:'\footnote{Lit. This region was in his hands.}
\z 
 
\ea \label{DP.17}
\textit{ā wextīyekey īđīč āmān dāwāw pādšāgerīš kerdēnew} \\ 
\gll ā wext-ī-ekey īđ=yč āmā=n dāwā=w pādšāgerī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} kingship\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `He (i.e., Jamsher Shah) came [and] claimed kingship.'
\z 
 
\ea \label{DP.18}
\textit{gerekiš bīyen hewrāmānī dāwā kero, dāgīr kero} \\ 
\gll gerek=iš bīye=n hewrāmān-ī dāwā ∅-ker-o dāgīr ∅-ker-o \\ 
 necessary\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} demand\textsc{\textsc{.f}} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} occupied \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wanted to occupy Hawraman.'
\z 
 
\ea \label{DP.21}
\textit{ī bāxū bāxātū hewrāmānī, memliketū hewrāmānī, xeɫkū hewrāmānī ā wextīyekey ēme hukimřewiš bīyēnmē} \\ 
\gll ī bāx=ū bāx-āt=ū hewrāmān-ī memliket=ū hewrāmān-ī xeɫk=ū hewrāmān ī ā wext-ī-ekey ēme hukimřew=iš bīyē=nmē \\ 
 \textsc{dem.prox} garden\textsc{.m}\textsc{=ez}\textsc{.gen} garden\textsc{.m}\textsc{-pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} people\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \textsc{dem.prox} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} governor\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `We have ruled over\footnote{Lit. We have been rulers.} these gardens of Hawraman, this region of Hawraman, [and] this people of Hawraman.'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.23}
\textit{min īney dew deso towe ey wēm kowe bilū} \\ 
\gll min īney ∅-de-ū des=o to=we ey wē=m ko=we bi-l-w \\ 
 \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}give\textsc{.prs}\textsc{-1sg} hand\textsc{.m}\textsc{=\textsc{post}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{intj} \textsc{reflx}\textsc{=1sg} where\textsc{=\textsc{post}} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} \\ 
\glt `If I hand over these to you, where should I go myself?'
\z 
 
\ea \label{DP.25}
\textit{šūnū āneyre beynē e čā bīyenū} \\ 
\gll šūn=ū āney=re beynē e=čā bīye=n=w \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} a\_while in=there be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Jamsher Shah was still there (i.e., in the citadel) afterwards.'
\z 
 
\ea \label{DP.26}
\textit{sibhān āġey ā wextīyekey āġe bīyen ħīn bīyen} \\ 
\gll sibhān āġe-ī ā wext-ī-ekey āġe bīye=n ħīn bīye=n \\ 
 \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} agha\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{fill} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Sibhan Agha was thingummy, the ruler.'
\z 
 
\ea \label{DP.28}
\textit{ī māɫū miɫke hinū āđī bīyenū desšo bīyenū} \\ 
\gll ī māɫ=w miɫk=e hin=ū āđī bīye=n=w des=š=o bīye=n=ū \\ 
 \textsc{dem.prox} property\textsc{.m}=and property\textsc{.m}\textsc{=dem} ez.pron\textsc{=ez}\textsc{.gen} 3sg\textsc{.obl}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and hand\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `These lands and properties here were his. They were under his control.'
\z 
 
\ea \label{DP.30}
\textit{šūnū āneyre wextē zāno hāmin āmāw gīriš kerd sibhān āġe milo āwekēū jemšēr āġey biřo, hesāryewe} \\ 
\gll šūn=ū āney=re wext-ē ∅-zān-o hāmin āmā=ū gīr=iš kerd sibhān āġe mi-l-o āwī-ekē=w jemšēr āġe-ī ∅- bi-ř-o hesār-ī=ewe \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} time\textsc{.m}\textsc{-indf} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} summer\textsc{.m} come\textsc{.pst}\textsc{.3sg}=and hook\textsc{.m}\textsc{=3sg} do\textsc{.pst} \textsc{pn} agha\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} water\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}=and \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}\textsc{sbjv-}cut\textsc{.prs}\textsc{-3sg} pond\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Then, as Sibhan Agha\footnote{Lit. he.} knew summer was coming [and] Jamsher Agha\footnote{Lit. he.} was stuck, Sibhan Agha went and cut off Jamsher Shah’s water supply [and led it to a] pond.'
\z 
 
\ea \label{DP.32}
\textit{āhekū gunjēw ārdēnešā ew dīmū qeɫē} \\ 
\gll āhek=ū gunj-ē=ū ārdē=ne=šā ew dīm=ū qeɫā-ē \\ 
 lime\textsc{.m}\textsc{=ez}\textsc{.gen} earthenware\_ pipe\textsc{.m}\textsc{-dir}\textsc{.pl}=and bring\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}\textsc{=3pl} \textsc{dem.dist} side\textsc{.m}\textsc{=ez}\textsc{.gen} castle\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \\ 
\glt `They took the water to the other side of the citadel.'
\z 
 
\ea \label{DP.35}
\textit{īzāfeyš wizo pā sextre māčāš qeɫāw hewrāmānī} \\ 
\gll īzāfe-ī=š wiz-o p=ā sext=re mi-āč-ā=š qeɫā=w hewrāmān-ī \\ 
 addition\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg} by\textsc{=dem}.dist mountain\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He would throw whatever was extra [of the snow] at that hard-to-access mountain called the citadel of Hawraman.'
\z 
 
\ea \label{DP.36}
\textit{wextē werwekē kem pā kotāy māro pey kemī milo hurgēro hurmēzo hur čolekēwe gēro minyo bāxeɫēš} \\ 
\gll wext-ē werwe-ekē kem p=ā kotā-ī mi-ār-o pey kem-ī mi-l-o hur-gēr-o hur-m-ēz-o hur čoleke-ē-ewe ∅-gēr-o mi-nye-o bāxeɫē=š \\ 
 time\textsc{.m}\textsc{-indf} snow\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} little by\textsc{=dem}.dist s\textsc{hort}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} for little\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pvb-}take\textsc{.prs}\textsc{-3sg} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg} \textsc{povb} sparrow\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{-indf} \textsc{ind-}take\textsc{.prs}\textsc{-3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3sg} embrace\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \\ 
\glt `When the little snow (that was left) was about to disappear [from the mountains], Jamsher Shah\footnote{Lit. he.} went [and] grabbed a sparrow [and] put it on his chest [under his clothing]'
\z 
 
\ea \label{DP.37}
\textit{be xātirū īney qesem wero qesemiš negino} \\ 
\gll be xātir=ū īney qesem ∅-wer-o qesem=iš ne-gin-o \\ 
 by sake\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} oath\textsc{.m} \textsc{sbjv-}out\textsc{-3sg} oath\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wanted to swear an oath using the sparrow. He did not want to swear falsely on himself.'
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.46}
\textit{ā wextīyekey wextē mēnē ā wextī herčē qisē kerā, herčē behsē kerā herčē hesābē kerā, qēxāw hewrāmānī herčē bā ā wextī isɫāhšā mebo} \\ 
\gll ā wext-ī-ekey wext-ē mi-ē-nē ā wext-ī herčē qisē-ē ∅-ker-ā herčē behse-ē ∅-ker-ā herčē hesāb-ē ∅-ker-ā qēxā=w hewrāmān-ī herčē ∅-b-ā ā wext-ī isɫāh=šā me-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} however\_much talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much discussion\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much account\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} chieftain\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} however\_much \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} amendment\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then Sibhan Agha and his men came [to Jamsher Shah]. No matter how much the headmen of Hawraman talked [to both of them], discussed the issue, and tried to mediate, they (i.e., Jamsher Shah and Sibhan Agha) disagreed.'
\z 
 
\ea \label{DP.48}
\textit{māčo, dey min īney pīneym wāten} \\ 
\gll mi-āč-o dey min īney p=īney=m wāte=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=1sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` Jamsher Shah said [to Sibhan Agha], ‘When I swore an oath, I swore on this sparrow;'
\z 
 
\ea \label{DP.49}
\textit{ī gīyānū qefesew sīney minne īnen āđe lūā} \\ 
\gll ī gīyān=ū qefese=ū sīne=y min=ne īne=n āđe lūā \\ 
 \textsc{dem.prox} soul\textsc{.m}\textsc{=ez}\textsc{.gen} cage\textsc{.m}=and chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{\textsc{.f}} go\textsc{.pst} \\ 
\glt ` The soul in my rib cage [that I swore an oath on] was this [sparrow]. It [just] flew away.'
\z 
 
\ea \label{DP.50}
\textit{ānen mirīčēɫēm wer dān, to xeyrū min siɫāmet} \\ 
\gll āne=n mirīčēɫe-ē=m wer dā=n to xeyr=ū min siɫāmet \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} small\_bird\textsc{.m}\textsc{-indf}\textsc{=1sg} out give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `’I have let the bird go away. Now, you follow your life, and I will follow mine!\footnote{Lit. Well-being for you and health for me.}’'
\z 
 
\ea \label{ZP.3}
\textit{čī degāw hewrāmānīne serāwinū degākēwe je eyē bīyen} \\ 
\gll č=ī degā=ū hewrāmān-ī=ne serāwin=ū degā-ekē=we je eyē bīye=n \\ 
 in=dem.prox village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} top\_of\_village\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} from mother\textsc{.obl}\textsc{.f} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was born in the village of Hawraman, at the top point of the village.'
\z 
 
\ea \label{ZP.6}
\textit{ā šexse īnā ew desū degāw hewrāmānīne māčāš pene pīr ħeme yoso māmoš bīyen} \\ 
\gll ā šexs=e īnā-∅ ew des=ū degā=ū hewrāmān-ī=ne m-āč-ā=š pene pīr ħeme yoso māmo=š bīye=n \\ 
 \textsc{dem.dist} saint\textsc{=dem} deic\textsc{-3sg}\textsc{.m} \textsc{dem.dist} hand\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn} \textsc{pn} \textsc{pn} paternal\_uncle\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The saint [whose grave is] on the other side of Hawraman is called Pir Hama Yoso. He is his (Pir Shaliyar’s) paternal uncle.'
\z 
 
\ea \label{ZP.7}
\textit{āne māmoš bīyen} \\ 
\gll āne māmo=š bīye=n \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} paternal\_uncle\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` He was his uncle.'
\z 
 
\ea \label{ZP.23}
\textit{e pā menteqew hewrāmānī ētir feqīyēwe bīyenū wenānšū} \\ 
\gll e=p=ā menteqe=ū hewrāmān-ī ētir feqī-ēwe bīye=n=ū wenā=n=š=ū \\ 
 in=by=\textsc{dem.dist} region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} theologian\textsc{.m}\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} read\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}=and \\ 
\glt `In this Hawraman region, well, he was a theologian; he was literate.'
\z 
 
\ea \label{ZP.26}
\textit{neweše gino herčīw duktirū ī lāū ew lā berāš duktir hīč duktirē ʕelēješ mekero} \\ 
\gll neweš-e ∅-gin-o herčīw duktir=ū ī lā=ū ew lā ∅-ber-ā=š duktir hīč duktir-ē ʕelēje=š me-ker-o \\ 
 ill\textsc{-f} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} how\_often physician\textsc{.m}=and dem.prox at\_the\_place\_of=and \textsc{dem.dist} at\_the\_place\_of \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3sg} physician\textsc{.m} nothing physician\textsc{.m}\textsc{-indf} treatment\textsc{.f}\textsc{=3sg} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt ` No matter how often she was taken to doctors here and there, no doctor could cure her.'
\z 
 
\ea \label{ZP.33}
\textit{ī kināčēw to dermānū derdīš īnā lā ī pīyāywe’} \\ 
\gll ī kināčē=ū to dermān=ū derd-ī=š īnā-∅ lā ī pīyā-ī=we \\ 
 dem.prox daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} \textsc{2sg} treatment\textsc{.m}\textsc{\textsc{=ez.gen}} illness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \\ 
\glt `‘This daughter of yours, the treatment for her illness lies with this man (i.e., Pir Shaliyar)'
\z 
 
\ea \label{ZP.36}
\textit{īne māčmēš pene šālīyāre sīyāw lālow kināčekēn} \\ 
\gll īne m-āč-mē=š pene šālīyār=e sīyāw lālo=ū kināčē-ekē=n \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn}\textsc{=dem} black maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This [person], whom we call Shaliyar Siya, is the maternal uncle of the [king’s] daughter.'
\z 
 
\ea \label{ZP.39}
\textit{māčo, ī kināčē bere degā be degāw} \\ 
\gll m-āč-o ī kināčē ∅-ber-e degā be degā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.obl}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `[and] said, ‘Take this girl village by village,'
\z 
 
\ea \label{ZP.40}
\textit{ētir xo ā zemāne eɫbet wesīle nebīyen yā māšīn yā čīwē} \\ 
\gll ētir xo ā zemān=e eɫbet wesīle ne-bīye=n yā māšīn yā čīw-ē \\ 
 \textsc{disc.ptcl} dis.ptc \textsc{dem.dist} time\textsc{.m}\textsc{=dem} of\_course tool\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} or car\textsc{.m} or thing\textsc{-indf} \\ 
\glt `—Well, back then, indeed, there was no means of transport, be it a car or anything [else.]'
\z 
 
\ea \label{ZP.43}
\textit{jā pādšākey wāten be lāloe ew lāloe kināčekē,} \\ 
\gll jā pādšā-eke-ī wāte=n be lālo-e ew lālo=e kināčē-ekē \\ 
 afterwards king\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to maternal\_uncle\textsc{.m}\textsc{-def} \textsc{dem.dist} maternal\_uncle\textsc{.m}\textsc{=dem} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `Oh, the king had said to the uncle, to [his] daughter’s uncle,'
\z 
 
\ea \label{ZP.44}
\textit{to eger ī kināčē berdet xizmetū ī pīyāy} \\ 
\gll to eger ī kināčē berd-e=t xizmet=ū ī pīyā-ī \\ 
 \textsc{2sg} if dem.prox daughter\textsc{.dir}\textsc{.f} take\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=\textsc{2sg}} service\textsc{.m}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `‘When you have taken this girl into the presence of this man'
\z 
 
\ea \label{ZP.45}
\textit{ī pīyāy tāwāš ī kināčēše kerde be qerārē wešeš kerdewe} \\ 
\gll ī pīyā-ī tāwā=š ī kināčē=š=e kerd-e be qerār-ē weš-e=š kerd-e=we \\ 
 dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} can\textsc{.pst}\textsc{=3sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} do\textsc{.pst}\textsc{-f}\textsc{.3sg} to arrangement\textsc{-indf} well\textsc{-f}\textsc{=3sg} do\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=compl} \\ 
\glt `[and if] the man has been able to cure the girl'
\z 
 
\ea \label{ZP.48}
\textit{ī kināčē nārīwe ētir pey min her ēnne weše bowe’} \\ 
\gll ī kināčē ne-ār-ī=we ētir pey min her ēnne weš-e b-o=we \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} any\_more for \textsc{1sg} emph as\_much\_as well\textsc{-f} be\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `Please do not bring this daughter [of mine] back to me as long as she is cured.'
\z 
 
\ea \label{ZP.49}
\textit{ā wextīekey lālo kināčekē māro} \\ 
\gll ā wext-ī-ekey lālo kināčē-ekē m-ār-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} maternal\_uncle\textsc{.m} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then, the girl’s uncle brought her (to Hawraman)'
\z 
 
\ea \label{ZP.54}
\textit{čāweɫ tā jāwe nelābē pāne pēwīyē tewenekē tewenē zile bē} \\ 
\gll čāweɫ tā jāwe ne-lā b-ē p=āne pēwīye-ē tewenī-ekē tewenī-ē zil-e b-ē \\ 
 in\_the\_past until road \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-indf} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\_visible\textsc{.prs}\textsc{-pstc}\textsc{.3sg} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} stone\textsc{.f}\textsc{-indf} big\textsc{-f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `In the past, when there was no road, the stone was visible from afar. It was a big stone.'
\z 
 
\ea \label{ZP.61}
\textit{her ā dāđe kero, ētir māčo, erē weɫā sirewā’ hurmēzowe miđo wene māroš hesārū činārārū} \\ 
\gll her ā dāđ=e ∅-ker-o ētir m-āč-o erē weɫā sirewā hur-m-ēz-o=we mi-đe-o wene m-ār-o=š hesār=ū činār-ā=re=ū \\ 
 emph \textsc{dem.dist} yell\textsc{.m}\textsc{=dem} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes indeed be cured\textsc{.pst}\textsc{.3sg} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} at \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} pond\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \\ 
\glt `After he shouted [and] said, ‘Indeed she is healed’, then he rose, [and] set off, and took the girl to Hasar Chinara.'
\z 
 
\ea \label{ZP.65}
\textit{pīr šelīyār ā vextī yānekew yānew tāteyšne bo, pū ser degāwe} \\ 
\gll pīr šelīyār ā vext-ī yāne-eke=ū yāne=ū tāte-ī=š=ne ∅-b-o p=ū ser degā=we \\ 
 \textsc{pn} \textsc{pn} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} house\textsc{.m}\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at=\textsc{dem.dist}\textsc{.3sg} on village\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar, well, he was [living] in his father’s house, [which was located] at the top of the village'
\z 
 
\ea \label{ZP.67}
\textit{tewejo kero wextē miđyo ī ženīū ī pīyā mēnē řāře duē dēwē īnē dewršāne} \\ 
\gll tewejo ∅-ker-o wext-ē mi-đye-o ī ženī=ū ī pīyā m-ē-nē řā=ře duē dēw-ē īnē dewr=šā=ne \\ 
 caution \textsc{ind-}do\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} dem.prox woman\textsc{.f}=and dem.prox man\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} road\textsc{.f}\textsc{=\textsc{post}} two demon\textsc{.m}\textsc{-dir}\textsc{.pl} dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} around\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `He noticed that when these two persons were on their way, they were surrounded by two demons.'
\z 
 
\ea \label{ZP.68}
\textit{ā wextīekey čā xuřo yošā gēɫowe} \\ 
\gll ā wext-ī-ekey čā ∅-xuř-o yo-ī=šā ∅-gēɫ-o=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} there \textsc{ind-}shout\textsc{.prs}\textsc{-3sg} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}wander\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `Then, he shouted at them [and] one of them returned.'
\z 
 
\ea \label{ZP.70}
\textit{fermāwo, gēɫdēwe bēheyāyā wāz bārdē čā zāyfē!’} \\ 
\gll ∅-fermāw-o ∅-gēɫ-dē=we bēheyā-yā wāz b-ār-dē č=ā zāyfe-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}wander\textsc{.prs}-\textsc{2pl}\textsc{=compl} shameless\textsc{-obl}\textsc{.pl} request \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} from=\textsc{dem.dist} woman\textsc{-obl}\textsc{.f} \\ 
\glt `He said, ‘Go back, you shameless [demons]! Let go of that poor girl!'
\z 
 
\ea \label{ZP.73}
\textit{īse māčmēš pene tenūrū dēwā} \\ 
\gll īse m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 now \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Now we call that place Tanūrū dēwā ‘the demons’ oven’'
\z 
 
\ea \label{ZP.74}
\textit{ā degākē ehā čew deso bīyēne duē kutē māčmēš pene tenūrū dēwā} \\ 
\gll ā degā-ekē ehā č=ew des=o bīyē=ne duē kut-ē m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 \textsc{dem.dist} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{intj} in=\textsc{dem.dist} hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} two half\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Do you see that the village is divided in half on the other side? We call that place Tanūrū dēwā.'
\z 
 
\ea \label{ZP.77}
\textit{mē tenūrū dēwā} \\ 
\gll m-ē tenūr=ū dēw-ā \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `They arrived at Tanūrū dēwā.'
\z 
 
\ea \label{ZP.80}
\textit{werū čemāw girdū mexeleqeɫāyre til bowe ginore ā dere qoɫe} \\ 
\gll wer=ū čem-ā=ū gird=ū mexeleqeɫā-ī=re til ∅-b-o=we gin-o=re ā dere qoɫ=e \\ 
 out\_of\textsc{\textsc{=ez.gen}} eye\textsc{-3pl}\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl} fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{dem.dist} valley\textsc{.m} deep\textsc{=dem} \\ 
\glt `in front of people’s eyes, he (the demon) rolled down and fell into a deep valley.'
\z 
 
\ea \label{ZP.81}
\textit{čā dēweke čāgene fotīyo} \\ 
\gll čā dēw-eke čāgene fotīye-o \\ 
 there demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} there die\textsc{.prs}\textsc{-3sg} \\ 
\glt `The demon died there.'
\z 
 
\ea \label{ZP.86}
\textit{māčā xeɫkīč, ī pīyā šimeš pey persdē xo yānēw e čāneš nīyāw} \\ 
\gll m-āč-ā xeɫk=īč ī pīyā šime=š pey ∅-pers-dē xo yāne-ē=ū e=čāne=š nīye=ā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} people\textsc{.m}\textsc{=add} dem.prox man\textsc{.m} \textsc{2pl}\textsc{=3sg} for \textsc{ind-}ask\textsc{.prs}-\textsc{2pl} dis.ptc house\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} in=like\_that\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The people said, ‘This man whom you ask for, he does not have a proper house.'
\z 
 
\ea \label{ZP.87}
\textit{ženīū yāneš nīyārē īnīšā} \\ 
\gll ženī=ū yāne=š nīye=ā=rē īnīšā \\ 
 woman\textsc{.f}=and house\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He has no wife, nor a house, and so on.'
\z 
 
\ea \label{ZP.88}
\textit{ā wextīekey ʕerz kerā, dey ēme ī kināčē jeryāniš e čīnew} \\ 
\gll ā wext-ī-ekey ʕerz ∅-ker-ā dey ēme ī kināčē jeryān=iš e=čīne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} petition\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{1pl} \textsc{dem.prox} girl\textsc{.f} story\textsc{.m}\textsc{=3sg} in=like=and \\ 
\glt `Then they (her uncle) said, ‘This girl’s story is like this.'
\z 
 
\ea \label{ZP.90}
\textit{mišo māreš biřū pey ī šexsīw ī šexse eger ī kināčēše weše kerde,} \\ 
\gll mišo māre=š ∅-biř-ū pey ī šexs-ī=ū ī šexs=e eger ī kināčē=š=e weš-e kerd-e \\ 
 aux marriage\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for dem.prox saint\textsc{-obl}\textsc{.m}=and dem.prox saint\textsc{=dem} if dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f} \\ 
\glt `I shall marry her to this person [on the condition that] this person has healed this girl.'
\z 
 
\ea \label{ZP.91}
\textit{ī kināčē weše bīyēnew īnīšā’} \\ 
\gll ī kināčē weš-e bīyē=ne=ū īnīšā \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} well\textsc{-f} be\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}}=and dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `The girl is now healed, and so on.'
\z 
 
\ea \label{ZP.94}
\textit{ā wextī ēme xeplezeřetīnē bēnē tā pāyīz} \\ 
\gll ā wext-ī ēme xeplezeřetīnē b-ēn-ē tā pāyīz \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{1pl} \textsc{pn}\textsc{.dir}\textsc{.pl} be\textsc{.prs}\textsc{-pstc}\textsc{-3pl} until autumn \\ 
\glt `Back then, there were farms of black corn\footnote{A type of corn. Also a type of bread made from corn flour.} until autumn.'
\z 
 
\ea \label{ZP.95}
\textit{šāmne ā tome pīr šelīyārī wēš ārden je šāmne bē} \\ 
\gll šām=ne ā tom=e pīr šelīyār-ī wē=š ārde=n je šām=ne b-ē \\ 
 \textsc{pn}\textsc{=\textsc{post}} \textsc{dem.dist} seed\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `From Damascus, Pir Shaliyar had brought its seeds. He was in Damascus (for a while).'
\z 
 
\ea \label{ZP.97}
\textit{ā wextī mārāš mē ʕerzū pīr šelīyārī kero lāloekeš} \\ 
\gll ā wext-ī m-ār-ā=š m-ē ʕerz=ū pīr šelīyār-ī ∅-ker-o lālo-eke=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \\ 
\glt `They (the people) brought her. They came (to Pir Shaliyar). The uncle said to Pir Shaliyar,'
\z 
 
\ea \label{ZP.98}
\textit{māčo, min īneyā jeryānim e čīnenū} \\ 
\gll m-āč-o min īne=ā jeryān=im e=čīne=n=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} story\textsc{.m}\textsc{=\textsc{1sg}} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He said, ‘I— look, my story is like this.'
\z 
 
\ea \label{ZP.100}
\textit{ī kināčē eger weše bīyowe mišyo māreš biřū pey to be mezebū wētāw} \\ 
\gll ī kināčē eger weš-e bīyē-o=we mišyo māre=š ∅-biř-ū pey to be mezeb=ū wē=tā=ū \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} if well\textsc{-f} be\textsc{.prs}.\textsc{cond}\textsc{.f}?\textsc{-3sg}\textsc{=compl} aux marraige\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for \textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `If the girl is healed, I shall marry her to you, according to your religion.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.102}
\textit{īne pādšāzāyēnew derdū min meweronew} \\ 
\gll īne pādšāzāyē=ne=ū derd=ū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}=and illness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She is no use to me.'
\z 
 
\ea \label{ZP.103}
\textit{min metāwū īne nigebānīye kerū} \\ 
\gll min me-tāw-ū īne nigebānī-e ∅-ker-ū \\ 
 \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} guardian\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt ` I cannot take care of her.'
\z 
 
\ea \label{ZP.104}
\textit{īne pādšāzāyēne čenū min meweronew} \\ 
\gll īne pādšāzāyē=ne čenū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}} with\textsc{.ez.gen} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She will not eat with me [under the same roof].'
\z 
 
\ea \label{ZP.105}
\textit{min čīwēwim nīyārē lāyiqū ī pādšāzāye bo} \\ 
\gll min čīw-ēw=im nīye=ā=rē lāyiq=ū ī pādšāzāye ∅-b-o \\ 
 \textsc{1sg} thing\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} deserved\textsc{\textsc{=ez.gen}} dem.prox princess\textsc{.obl}\textsc{.f} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `I have nothing which would be up to the standards of this princess.'
\z 
 
\ea \label{ZP.113}
\textit{xeɫk mē dewreš mido māčā, āne čēšā? čēšā āne?’} \\ 
\gll xeɫk m-ē dewre=š mi-de-o m-āč-ā āne čēš=ā čēš=ā āne \\ 
 people\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} rotation\textsc{=3sg} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=cop}\textsc{.3sg}\textsc{.m} what\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \\ 
\glt `People came, encircled him, and asked,\footnote{Lit. said.} ‘What is going on? What is going on?’'
\z 
 
\ea \label{ZP.115}
\textit{māčā, dey īne ī pīyā ī ženīče ārdēne} \\ 
\gll m-āč-ā dey īne ī pīyā ī ženī=č=e ārdē=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox man\textsc{.m} dem.prox woman\textsc{.dir}f\textsc{=add}\textsc{=dem} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `[and] said, ‘Well, this man has brought this woman [here]'
\z 
 
\ea \label{ZP.126}
\textit{ā wextī pīr šelīyār fermāwo, šime īne māčdē řāstā wāt} \\ 
\gll ā wext-ī pīr šelīyār ∅-fermāw-o šime īne m-āč-dē řās=tā wāt \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{2pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} \\ 
\glt `Then Pir Shaliyar said, ‘What you are saying is right.’\footnote{Lit. You said this; you told the truth.}'
\z 
 
\ea \label{ZP.128}
\textit{weɫāhī īne šime māčdēšū īnā vistāwe řā pey min čūn šime be xātirū xwāytā kerd hetā xwā dinyāš weše kerdēne ī merāsēme her bo her meno’} \\ 
\gll weɫāhī īne šime m-āč-dē=š=ū īnā-∅ vist=tā=we řā pey min čūn šime be xātir=ū xwā-ī=tā kerd hetā xwā dinyā=š weš-e kerdē=ne ī merāsēm=e her ∅-b-o her ∅-men-o \\ 
 indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} \textsc{ind-}say\textsc{.prs}-\textsc{2pl}\textsc{=3sg}=and deic\textsc{-3sg}\textsc{.m} throw\textsc{.pst}=\textsc{2pl}\textsc{=compl} road\textsc{.f} for \textsc{1sg} because \textsc{2pl} to sake\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}be\textsc{.prs}\textsc{-3sg} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `Indeed, what you are saying and the way you organised [this ceremony] for me, since you did it in good faith, so long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ZP.130}
\textit{hetā xwāy dinyāš weše kerdēne ī merāsēme kem mekero} \\ 
\gll hetā xwā-ī dinyā=š weš-e kerdē=ne ī merāsēm=e kem me-ker-o \\ 
 as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} small \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `So long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ŽP.6}
\textit{ānen īse īnā lāw hānew degē} \\ 
\gll āne=n īse īnā-∅ lā=ū hāne=ū degē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} water\_spring\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f} \\ 
\glt `That (his shrine) is on the other side of the village.'
\z 
 
\ea \label{ŽP.7}
\textit{āne tātešā} \\ 
\gll āne tāte=š=ā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} father\textsc{.m}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is his father.'
\z 
 
\ea \label{ŽP.12}
\textit{māmo yoso ā pīr heme yoso māmošā} \\ 
\gll māmo yoso ā pīr heme yoso māmo=š=ā \\ 
 paternal\_uncle \textsc{pn} \textsc{dem.dist} \textsc{pn} \textsc{pn} \textsc{pn} paternal\_uncle\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Uncle Yoso, [otherwise called] Pir Hama Yoso, was his uncle.'
\z 
 
\ea \label{ŽP.19}
\textit{ānā ew dīmū qeɫāw kitāy} \\ 
\gll āne=ā ew dīm=ū qeɫā=ū kitā-ī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `—Look, it is [located] on the other side of the Kita castle'
\z 
 
\ea \label{ŽP.31}
\textit{milo ew dīmū qeɫāw kitā} \\ 
\gll mi-l-o ew dīm=ū qeɫā=w kitā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `He went to the other side of the Kita castle.'
\z 
 
\ea \label{ŽP.44}
\textit{ī yewēše šānēnē pī zemīnne} \\ 
\gll ī yewe-ē=š=e šānē=nē p=ī zemīn=ne \\ 
 dem.prox barley\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3sg}\textsc{=dem} scatter\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} at=dem.prox land\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Has he planted the seeds in the field?'
\z 
 
\ea \label{ŽP.45}
\textit{yā ī yewēše zāxēre kerdēnē?} \\ 
\gll yā ī yewe-ē=š=e zāxēre kerdē=nē \\ 
 either dem.prox barley\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3sg}\textsc{=dem} sotrage do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Or has he [instead] stored the seeds [for himself]'
\z 
 
\ea \label{ŽP.49}
\textit{īneš netāwān kero} \\ 
\gll īne=š ne-tāwā=n ∅-ker-o \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He cannot do these [tasks].’'
\z 
 
\ea \label{ŽP.52}
\textit{zengenekeš wistrew pāne zengene zemīnekey wēš} \\ 
\gll zengene-eke=š wist=re=we p=āne zengene zemīn-ekey wē=š \\ 
 hoe\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} throw\textsc{.pst}\textsc{=\textsc{povb}}\textsc{=compl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} hoe\textsc{.m} land\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} \\ 
\glt `He put the hoe there, in his field.'
\z 
 
\ea \label{ŽP.67}
\textit{wurmēzāwe bilā seriš bilā zānā jeryān čēšā} \\ 
\gll wur-m-ēz-ā=we bi-l-ā ser=iš bi-l-ā ∅-zān-ā jeryān čēš=ā \\ 
 \textsc{pvb-}\textsc{ind-}\textsc{dem.dist}\textsc{=compl} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} on\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-3pl} story\textsc{.m} what\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the uncle and his wife) woke up to go to his bed and went to see the story (i.e. what was happening).'
\z 
 
\ea \label{ŽP.68}
\textit{wextē miđyā īnā āne je zemīnī hewāwe wuten} \\ 
\gll wext-ē mi-đye-ā īnā-∅ āne je zemīn-ī hewā=we wute=n \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} from earth\textsc{.m}\textsc{-obl}\textsc{.m} air\textsc{=\textsc{post}} sleep\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They saw that he (Pir Shaliyar) had levitated and slept aloft in the air at this height\footnote{Lit. this much.} (informant indicates the height of the bed with his hand).'
\z 
 
\ea \label{ŽP.69}
\textit{āne je zemīnī pēse wāčī xwāy ketē pey kīyāsen} \\ 
\gll āne je zemīn-ī pēse ∅-wāč-ī xwā-ī ket-ē pey kīyāse=n \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} from earth\textsc{.m}\textsc{-obl}\textsc{.m} as\_if \textsc{sbjv-}say\textsc{.prs-}\textsc{2sg} God\textsc{.m}\textsc{-obl}\textsc{.m} bed\textsc{-indf} to send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[He had levitated] this much from the ground, as if God had sent him a bed.'
\z 
 
\ea \label{ŽP.70}
\textit{āne īnā hewāwe} \\ 
\gll āne īnā-∅ hewā=we \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \\ 
\glt `He (his body) was aloft in the air [up to this height]'
\z 
 
\ea \label{ŽP.71}
\textit{pīr heme yoso wēš milo temāšeš kero miđyo āne īnā hewāwe zāno īne ētir kārī xwāyīn} \\ 
\gll pīr heme yoso wē=š mi-l-o temāše=š ∅-ker-o mi-đye-o āne īnā-∅ hewā=we ∅-zān-o īne ētir kār-ī xwāyī=n \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{reflx}\textsc{=3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} looking\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} task\textsc{.m}\textsc{-obl}\textsc{.m} divine\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Hama Yoso went [and] looked at him; he noticed that he (Pir Shaliyar) was aloft in the air [up until a certain height] and came to realise that that was God’s work.'
\z 
 
\ea \label{ŽP.74}
\textit{māčo min īnem dī be sey misefāywe} \\ 
\gll m-āč-o min īne=m dī-∅ be sey misefā-ī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} see\textsc{.pst}\textsc{-3sg}\textsc{.m} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘I witnessed this about Say Mustafa.’'
\z 
 
\ea \label{ŽP.77}
\textit{ū pīyāyče řāsiš wāt ūneš gēɫnāwew} \\ 
\gll ū pīyā=yč=e řās=iš wāt ūne=š gēɫnā=we=ū \\ 
 \textsc{dem.dist}\textsc{.3sg} man\textsc{.m}\textsc{=add}\textsc{=dem} truth\textsc{=3sg} say\textsc{.pst} \textsc{dem.dist}\textsc{=3sg} narrate\textsc{.pst}\textsc{=\textsc{post}}=and \\ 
\glt `And the man (the servant) told the truth when he said that [those words]'
\z 
 
\ea \label{ŽP.80}
\textit{je her yāgē weno, čā dimāyē wextē miđyā ēnneš wenā qābilyetiš peyā kerd kīyānāš pey beġāy pey xizmetū, kīyānāš pey beġāy wenāy} \\ 
\gll je her yāgē-ē ∅-wen-o č=ā dimā-ē wext-ē mi-đye-ā ēnne=š wenā qābilyet=iš peyā kerd ∅-kīyān-ā=š pey beġā-ī pey xizmet=ū ∅-kīyān-ā=š pey beġā-ī wenāy \\ 
 from emph place\textsc{.f}\textsc{-indf} \textsc{ind-}read\textsc{-3sg} in=\textsc{dem.dist} afterwards\textsc{-obl}\textsc{.f} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} so\_much\textsc{=3sg} read\textsc{.pst} capability\textsc{.m}\textsc{=3sg} visible do\textsc{.pst} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn}\textsc{-obl}\textsc{.m} to service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn}\textsc{-obl}\textsc{.m} read.inf \\ 
\glt `Wherever he studied, after a while, they realised that he studied to the extent that he deserved to be sent to Baghdad into the presence of his highness [Khos]'
\z 
 
\ea \label{ŽP.93}
\textit{bilīwe pey ā menteqew wētāw, šerʕī ders wāčīw} \\ 
\gll bi-l-ī=we pey ā menteqe=ū wē=tā=ū šerʕī ders ∅-wāč-ī=ū \\ 
 \textsc{sbjv-}\textsc{-obl}\textsc{.m}\textsc{=compl} to \textsc{dem.dist} region=and \textsc{reflx}=\textsc{2pl}=and sharia lesson\textsc{.m} \textsc{sbjv-}say\textsc{.prs-}\textsc{2sg}=and \\ 
\glt `Go to your region [and] teach Sharia law'
\z 
 
\ea \label{ŽP.97}
\textit{řāne mēnē pāne yā guzerēwene mesīrēwene čenū hejījīyē ginā pēwere} \\ 
\gll řā=ne m-ē-nē p=āne yā guzer-ēwe=ne mesīr-ēwe=ne čenū hejījī-ē ∅-gin-ā pēwere \\ 
 road\textsc{.f}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} either path\textsc{-indf}\textsc{=\textsc{post}} path\textsc{-indf}\textsc{=\textsc{post}} with\textsc{.ez.gen} from\_Hajij\textsc{-indf} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl} together \\ 
\glt `On the road, on a path, he ran into a person from Hajij.'
\z 
 
\ea \label{ŽP.98}
\textit{ētir herčī mēnē ā zemāne wesīlew māšīnū čīwī mebo} \\ 
\gll ētir herčī m-ē-nē ā zemān=e wesīle=ū māšīn=ū čīw-ī me-b-o \\ 
 \textsc{disc.ptcl} what\_ever \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} tool\textsc{.m}=and car\textsc{.m}=and thing\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well, in that period, there were no means of transportation, cars, etc.'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.116}
\textit{mēwe ētir ā zemāne pīr šelīyār mēwe menteqe des kero, jārē āne minyā māčāš meřū āsinī} \\ 
\gll m-ē=we ētir ā zemān=e pīr šelīyār m-ē=we menteqe des=t ∅-ker-o jārē āne mi-nye-ā m-āč-ā=š meře=ū āsin-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} once \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) returned [to Hawraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ‘Iron cave.’'
\z 
 
\ea \label{ŽP.119}
\textit{čāne weɫēne luānre wā čilēš kēštēne} \\ 
\gll č=ā=ne weɫēne luā=n=re wā čilē=š kēštē=ne \\ 
 from=\textsc{dem.dist}\textsc{=\textsc{post}} in\_the\_begining go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} there forty\_days\_fasting\textsc{.dir}\textsc{.f}\textsc{=3sg} pull\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `From long ago, he would go there and fast for forty days.'
\z 
 
\ea \label{ŽP.123}
\textit{ā čilexāne čilexāne lemē} \\ 
\gll ā čilexāne čilexāne lemē \\ 
 \textsc{dem.dist} \textsc{pn} \textsc{pn} \textsc{pn}\textsc{.obl}\textsc{.f} \\ 
\glt `[he would fast] in that place of fasting [called] Chilaxana Lame.'
\z 
 
\ea \label{ŽP.124}
\textit{ā wextī xeberē yāwno hezretū xosī} \\ 
\gll ā wext-ī xeber-ē ∅-yāwn-o hezret=ū xos-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} news\textsc{.m}\textsc{-indf} \textsc{ind-}cause\_to\_arrive\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, the news reached His Highness Khos.'
\z 
 
\ea \label{ŽP.125}
\textit{māčā kābrāyē pēse īse ā wextīč qisekerē her bīyēnē} \\ 
\gll m-āč-ā kābrā-ē pēse īse ā wext-ī=č qiseker-ē her bīyē=nē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} man\textsc{-indf} as\_if now \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} gossiper\textsc{-dir}\textsc{.pl} emph be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They said — Like nowadays, back then, there were also gossips.'
\z 
 
\ea \label{ŽP.135}
\textit{suāleš čene kero, māčo, čī āmānīwe pey īnīšā?’} \\ 
\gll suāle=š čene ∅-ker-o m-āč-o čī āmā=nī=we pey īnīšā \\ 
 question\textsc{.f}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} to dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He asked him [and] said, ‘Why have you come here?\footnote{Lit. for these.}’'
\z 
 
\ea \label{ŽP.137}
\textit{ā wextīyekey her noʕē māčoš lāwew} \\ 
\gll ā wext-ī-ekey her noʕ-ē m-āč-o=š lā=w=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} emph kind/way\textsc{-indf} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to=and \\ 
\glt `Then, he (the man) implicitly told him (Pir Shaliyar).'
\z 
 
\ea \label{ŽP.138}
\textit{māčoš lāwe māčo min pey īney āmānā āmānāw} \\ 
\gll m-āč-o=š lā=we m-āč-o min pey īney āmā=nā āmā=nā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} to dem.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}=and \\ 
\glt `He said to him, ‘I have come for this [reason]’'
\z 
 
\ea \label{ŽP.143}
\textit{milowe yānew ētir pāne nāmēwe peyā kero} \\ 
\gll mi-l-o=we yāne=ū ētir p=āne nāmē-ewe peyā ∅-ker-o \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m}=and \textsc{disc.ptcl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} name\textsc{.f}\textsc{-indf} visible \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He went there (to Hawraman) and got a bit famous.\footnote{Lit. He found a name.}'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.156}
\textit{māčā, ī kināčē to tāwo weše kerowe’} \\ 
\gll m-āč-ā ī kināčē to ∅-tāw-o weš-e ∅-ker-o=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `It was said [to him in the dream], ‘He can cure this daughter of yours.’'
\z 
 
\ea \label{ŽP.159}
\textit{to bo lue! ī kināčē bere} \\ 
\gll to b-o ∅-lu-e ī kināčē ∅-ber-e \\ 
 \textsc{2sg} \textsc{imp-}come\textsc{.prs}\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `‘Come [here and] go! Take the girl.'
\z 
 
\ea \label{ŽP.162}
\textit{dēhāt be dēhāt ī kināčē bere} \\ 
\gll dēhāt be dēhāt ī kināčē ∅-ber-e \\ 
 village\textsc{.f} by village\textsc{.f} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `Take the girl village by village.'
\z 
 
\ea \label{ŽP.163}
\textit{eger to ī kināčēte berdeū ī pīyā her kes hen} \\ 
\gll eger to ī kināčē=t=e berd-e=ū ī pīyā her kes hen-∅ \\ 
 if \textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} take\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m} emph person\textsc{.m} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `If you take the girl, [no matter] who this man (the doctor) is,'
\z 
 
\ea \label{ŽP.164}
\textit{tāwāš ī duktire ī kināčēše weše kerdew} \\ 
\gll tāwā=š ī duktir=e ī kināčē=š=e weš-e kerd-e=ū \\ 
 can\textsc{.pst}\textsc{=3sg} dem.prox phy\textsc{=dem} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `If the doctor can cure the girl,'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.167}
\textit{silēne řeđē bā tewenē čā beynne bē beynū silēnē biɫberīne bē} \\ 
\gll silēne řeđ-ē bā tewenī-ē č=ā beyn=ne b-ē beyn=ū silēnē biɫber-ī=ne b-ē \\ 
 \textsc{pn} crossing\textsc{-dir}\textsc{.pl} \textsc{hort} stone\textsc{.f}\textsc{-indf} at=\textsc{dem.dist} between\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} between\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `They passed Slen. There was a stone in the road between Slen and Bilbar.'
\z 
 
\ea \label{ŽP.171}
\textit{tewenē pāne bē nīšāre seyāwe} \\ 
\gll tewenī-ē p=āne b-ē ∅-nīš-ā=re sey-ā=we \\ 
 stone\textsc{.f}\textsc{-indf} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}sit\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} rest\textsc{.prs}\textsc{-3pl}\textsc{=compl} \\ 
\glt `It was a wide stone. There was a stone there. They sat [on the stone] to rest.'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.176}
\textit{tewejoēšā kero wextē miđyo duē dānē dēwē yošā īnā peštīšne mē řāre yošā werišne} \\ 
\gll tewejo-ē=šā ∅-ker-o wext-ē mi-đye-o duē dāne-ē dēw-ē yo=šā īnā-∅ peštī=š=ne m-ē řā=re yo=šā wer=iš=ne \\ 
 caution\textsc{-indf}\textsc{=3pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} two \textsc{clf}\textsc{-dir}\textsc{.pl} demon\textsc{.m}\textsc{-dir}\textsc{.pl} one\textsc{.m}\textsc{=3pl} deic\textsc{-3sg}\textsc{.m} back\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} road\textsc{.f}\textsc{=\textsc{post}} one\textsc{.m}\textsc{=3pl} front\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `He noticed [their presence]. He realised that two demons [were with them], one behind her [the girl], one in front of her.'
\z 
 
\ea \label{ŽP.177}
\textit{ī dēwēnē ī kināčēšā bestēne} \\ 
\gll ī dēw=ē=nē ī kināčē=šā bestē=ne \\ 
 dem.prox demon\textsc{.m}\textsc{=ez}.\textsc{attr}\textsc{=cop}\textsc{.3pl} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3pl} tie\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `It was the demons who had muted the girl.'
\z 
 
\ea \label{ŽP.178}
\textit{nitqšā girtēne ī due dēwene} \\ 
\gll nitq=šā girtē=ne ī due dēwe=ne \\ 
 speech\textsc{=3pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox two\textsc{.f} demon\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `They had muted her speech, these two demons.'
\z 
 
\ea \label{ŽP.187}
\textit{milegā ā wextī mizyā werāwerū dūtifɫey} \\ 
\gll milegā ā wext-ī mi-zye-ā werāwer=ū dūtifɫey \\ 
 top\_of\_mountain \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `Then they came out in front of  Dutifla.'
\z 
 
\ea \label{ŽP.192}
\textit{dēwē keřetē milo peštīšāne keřetē mē weršāne} \\ 
\gll dēw-ē keřet-ē mi-l-o peštī=šā=ne keřet-ē m-ē wer=šā=ne \\ 
 demon\textsc{.m}\textsc{-dir}\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} back\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} front\textsc{.m}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt ` A demon would sometimes follow them, [and] sometimes go ahead of them.'
\z 
 
\ea \label{ŽP.196}
\textit{tenūrū māčāš pene tenūrū dēwā’} \\ 
\gll tenūr=ū m-āč-ā=š pene tenūr=ū dēw-ā \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-3pl} \\ 
\glt `They call [that place next to the tree] Tanuru Dewa ‘the demons’ oven.’'
\z 
 
\ea \label{ŽP.198}
\textit{fermāwo, weɫā čūn ānā bēheyānī kerūt penū ʕāɫemī} \\ 
\gll ∅-fermāw-o weɫā čūn āne=ā bēheyā=nī ∅-ker-ū=t pen=ū ʕāɫem-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed as \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} shameless\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{2sg}} trouble\textsc{.m}\textsc{\textsc{=ez.gen}} world\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) said, ‘Indeed, since you are so indecent, I will disgrace you.’'
\z 
 
\ea \label{ŽP.200}
\textit{tewejoēš wene kero pēse gāw til bore e pā tenūrū dēwāre til bowew ginorewe. ānen māčāš pene tenūrū dēwā} \\ 
\gll tewejo-ē=š wene ∅-ker-o pēse gāw til b-o=re e=pā tenūr=ū dēw-ā=re til b-o=we=ū gin-o=re=we āne=n m-āč-ā=š pene tenūr=ū dēw-ā \\ 
 caution\textsc{-indf}\textsc{=3sg} \textsc{post} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} as\_if bull\textsc{.m} rolling be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=that oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} rolling be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}}=and fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}}\textsc{=compl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He (Pir Shaliyar) paid attention to him. He (the demon) rolled down like a cow. It is there. It is called Tanuru Dewa.'
\z 
 
\ea \label{ŽP.201}
\textit{tenūrū dēwā ētir pēse gāw til bowe e čā werāwerū girdū mexeleqeɫāw hewrāmānīwe dēweke til bowe ginore soɫāwe} \\ 
\gll tenūr=ū dēw-ā ētir pēse gāw til ∅-b-o=we e=čā werāwer=ū gird=ū mexeleqeɫā=ū hewrāmān-ī=we dēw-eke til ∅-b-o=we ∅-gin-o=re soɫāwe \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{disc.ptcl} as\_if bull\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=there in\_front\_of\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt ` Then it was as if a cow had rolled down; the [demon] rolled down and fell [into a place called] Solawa before all the people’s eyes.'
\z 
 
\ea \label{ŽP.202}
\textit{dēwe čāne topo miro} \\ 
\gll dēwe č=ā=ne ∅-top-o ∅-mir-o \\ 
 demon\textsc{.f} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}die\textsc{.prs}\textsc{-3sg} \textsc{ind-}die\textsc{.prs}\textsc{-3sg} \\ 
\glt `The demon died there.'
\z 
 
\ea \label{ŽP.204}
\textit{perso, jeryān čikonū? ī yānew kābrāy kēnū īne kēnū?} \\ 
\gll ∅-pers-o jeryān čiko=n=ū ī yāne=ū kābrā-ī kē=n=ū īne kē=n=ū \\ 
 \textsc{ind-}ask\textsc{.prs}\textsc{-3sg} story\textsc{.m} where\textsc{=cop}\textsc{.3sg}\textsc{.m}=and dem.prox house\textsc{.m}=and man\textsc{-obl}\textsc{.m} who\textsc{=cop}\textsc{.3sg}\textsc{.m}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} who\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He came to the village. He asked, ‘[what is the story]? Where is the house of the fellow? Who is he? Who is he?'
\z 
 
\ea \label{ŽP.206}
\textit{ā wextī lālo kināčē māčo, min, tāze pādšāy kerdenā wekēɫ} \\ 
\gll ā wext-ī lālo kināčē m-āč-o min tāze pādšā-ī kerde=nā wekēɫ \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} anyway king\textsc{.m}\textsc{-obl}\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m} \\ 
\glt `Then, the girl’s uncle said, ‘As for me, well, the king has put me in charge.\footnote{Lit. the king has made me advocate.}'
\z 
 
\ea \label{ŽP.207}
\textit{mišyo ī kināčē nīkāh kerū pey toū be mezebū wētāū} \\ 
\gll mišyo ī kināčē nīkāh ∅-ker-ū pey to=ū be mezeb=ū wē=tā=ū \\ 
 aux dem.prox girl\textsc{.dir}\textsc{.f} marriage\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} to \textsc{2sg}=and by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.213}
\textit{īnē be min mewero kināčē pādšāy,} \\ 
\gll īnē be min me-wer-o kināčē pādšā-ī \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} with \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg} girl\textsc{.dir}\textsc{.f} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `She will not eat with me [under the same roof], the king’s daughter.'
\z 
 
\ea \label{ŽP.214}
\textit{min āne nānmā xeplezeřātīnen} \\ 
\gll min āne nān=m=ā xeplezeřātīne=n \\ 
 \textsc{1sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Look, my meal is xaplazaratine.’'
\z 
 
\ea \label{ŽP.215}
\textit{āne zerātiš šānī miđo, zerāte sūrī} \\ 
\gll āne zerāt=iš šānī mi-đe-o zerāt=e sūrī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} corn\textsc{=3sg} show \textsc{ind-}give\textsc{.prs}\textsc{-3sg} corn\textsc{=ez}\textsc{.cmpd} Syrian \\ 
\glt `He (Pir Shaliyar) showed him some corn, [some] red corn.'
\z 
 
\ea \label{ŽP.216}
\textit{āne šānī mido,} \\ 
\gll āne šānī mi-de-o \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} show \textsc{ind-}give\textsc{.prs} \\ 
\glt `He showed him some [corn and]'
\z 
 
\ea \label{ŽP.217}
\textit{fermāwo, min īne nānmārēū} \\ 
\gll ∅-fermāw-o min īne nān=m=ā=rē=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}}=and \\ 
\glt `He said, ‘This is my meal.’'
\z 
 
\ea \label{ŽP.219}
\textit{xo ā wextī wēčiš yāneš her mebo} \\ 
\gll xo ā wext-ī wē=č=iš yāne=š her me-b-o \\ 
 dis.ptc \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=add}\textsc{=3sg} house\textsc{.m}\textsc{=3sg} emph \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He had no house at that time.'
\z 
 
\ea \label{ŽP.220}
\textit{ā wextī māčā, dey qurbān} \\ 
\gll ā wext-ī m-āč-ā dey qurbān \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} sir.\textsc{voc} \\ 
\glt `Then, they (people) said, Oh, Sir!’'
\z 
 
\ea \label{ŽP.222}
\textit{māčā, īne čēšā?} \\ 
\gll m-āč-ā īne čēš=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the people) said, ‘What is this?’'
\z 
 
\ea \label{ŽP.223}
\textit{īnē ī qisē čēšene? čēš māčdē? māčā čēš?} \\ 
\gll īnē ī qisē čēše=ne čēš m-āč-dē m-āč-ā čēš \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} dem.prox talk\textsc{.dir}\textsc{.f} what\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} what \textsc{ind-}say\textsc{.prs}-\textsc{2pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} what \\ 
\glt `What is this talk? What are you saying? What are they saying?'
\z 
 
\ea \label{ŽP.225}
\textit{her lāloeke māčo, mišyo ī kināčē min meberūšo māreš biřūnew’} \\ 
\gll her lālo-eke m-āč-o mišyo ī kināčē min me-ber-ū=š=o māre=š ∅-biř-ūne=ū \\ 
 emph maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} aux dem.prox girl\textsc{.dir}\textsc{.f} \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} marraige\textsc{=3sg} \textsc{ind-}cut\textsc{.prs}\textsc{-\textsc{1sg}}=and \\ 
\glt ` The uncle kept saying, ‘I shall not take this girl back. I shall marry her [to Pir Shaliyar].’'
\z 
 
\ea \label{ŽP.226}
\textit{āđīč māčo, min metāwū ī kināčē wey kerū īne nānmā} \\ 
\gll āđ=īč m-āč-o min me-tāw-ū ī kināčē wey ∅-ker-ū īne nān=m=ā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox girl\textsc{.dir}\textsc{.f} raising \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `People … he (Pir Shaliyar) said, ‘I cannot take care of this girl. This is my food.’'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.228}
\textit{yomā āne tišit miđo} \\ 
\gll yo=mā āne tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give [pomegranate] molasses.'
\z 
 
\ea \label{ŽP.242}
\textit{qey mekero, mādām desūr e pēsen pādšā řāzīe nīā ī ženī bilowe ī kināčē bilowe qey mekero!’} \\ 
\gll qey me-ker-o mādām desūr e=pēse=n pādšā řāzīe nīye=ā ī ženī bi-l-o=we ī kināčē bi-l-o=we qey me-ker-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} as\_long\_as task/order\textsc{.m} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} satisfied \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} dem.prox woman\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `There is no problem. Now that things have developed this way, [and] the king is not happy for the girl to return, it’s all right! [She can stay here]’'
\z 
 
\ea \label{ŽP.243}
\textit{pīr šelīyār fermāwo, tehāā weɫāhī īne šime desūrtā dāw wāttā čūnke āđē be diɫtā wāt řāstā wāt hetā xwāy dinyāš weše kerdēne her menowe kemīč mekero her zīyā kero} \\ 
\gll pīr šelīyār ∅-fermāw-o tehāā weɫāhī īne šime desūr=tā dā=ū wāt=tā čūnke āđē be diɫ=tā wāt řās=tā wāt hetā xwā-ī dinyā=š weš-e kerdē=ne her ∅-men-o=we kem=īč me-ker-o her zīyā ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} task/order\textsc{.m}=\textsc{2pl} give\textsc{.pst}=and say\textsc{.pst}=\textsc{2pl} since 3pl\textsc{.dir} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} truth=\textsc{2pl} say\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} small\textsc{=add} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar said, ‘Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] continues to exist; it will not be diminished; it will grow.'
\z 
 
\ea \label{ŽP.244}
\textit{ī bezme šime wāttāw ī čāštīū īne ī desūre šime ījrātā kerd hetā xwāy dinyāš weše kerdēne šime řāstā wāt be diɫtā wāt} \\ 
\gll ī bezm=e šime wāt=tā=ū ī čāštī=ū īne ī desūr=e šime ījrā=tā kerd hetā xwā-ī dinyā=š weš-e kerdē=ne šime řās=tā wāt be diɫ=tā wāt \\ 
 dem.prox jollification\textsc{.dir}\textsc{.m}\textsc{=dem} \textsc{2pl} say\textsc{.pst}=\textsc{2pl}=and dem.prox meal\textsc{.f}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox task/order\textsc{.m}\textsc{=dem} \textsc{2pl} performance=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} \\ 
\glt `[After] the things you said, and the meals which you prepared, so long as the world continues to exist, you are right [in so doing]; you did these [things] in good faith.'
\z 
 
\ea \label{ŽP.245}
\textit{ī merāsēme her menowe hīčkeyč kem mekero her zīyāy kero īlāy žīyāyne bo zīyāyiš pene kero} \\ 
\gll ī merāsēm=e her ∅-men-o=we hīčke=yč kem me-ker-o her zīyāy ∅-ker-o īlāy žīyāy=ne ∅-b-o zīyāy=iš pene ∅-ker-o \\ 
 dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} nothing\textsc{=add} little \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increase \textsc{ind-}do\textsc{.prs}\textsc{-3sg} for\_sure increased\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} increased\textsc{=3sg} to \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This ceremony will be repeated, it won’t diminish. It will continue to grow. It will grow [into bigger ceremonies]'
\z 
 
\ea \label{ŽP.258}
\textit{ā wextīyekey lālo pīr šelīyār hīn lālo heremū pīr šelīyārī milowe xeber berowe pey pādšāy} \\ 
\gll ā wext-ī-ekey lālo pīr šelīyār hīn lālo hereme=ū pīr šelīyār-ī mi-l-o=we xeber ∅-ber-o=we pey pādšā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} maternal\_uncle\textsc{.m} \textsc{pn} \textsc{pn} \textsc{fill} maternal\_uncle\textsc{.m} wife\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=compl} to king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, Pir Shaliyar’s uncle, eh, Pir Shaliyar’s wife’s uncle, went back [to Egypt] and passed the news to the king.'
\z 
 
\ea \label{ŽP.262}
\textit{lue! to luewe hem wesāylīč bere pey ī pīyāyū ī kināčēw} \\ 
\gll ∅-lu-e to ∅-lu-e=we hem wesāyl=īč ∅-ber-e pey ī pīyā-ī=ū ī kināčē=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} both appliance\textsc{.m}\textsc{=add} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}=and dem.prox daughter\textsc{.dir}\textsc{.pl}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter'
\z 
 
\ea \label{ŽP.267}
\textit{ānen īse šelīyāre sīyāwā} \\ 
\gll āne=n īse šelīyār=e sīyāw=ā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[His shrine] is over there now; [he is called] Shaliyar Siyaw.'
\z 
 
\ea \label{BP.3}
\textit{yānē yānē meseleyš serū īneyn yānē bābā xwāy dā, kuřēš peyā bī} \\ 
\gll yānē yānē mesele-ī=š ser=ū īne-ī=n yānē bābā xwā-ī dā kuř-ē=š peyā bī-∅ \\ 
 that\_means that\_means issue\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} that\_means grandfather\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} give\textsc{.pst} son\textsc{.m}\textsc{-indf}\textsc{=3sg} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `The point with his name is that God granted the grandfather [a son]. A boy  was born to him .\footnote{Lit. A son of him appeared.}'
\z 
 
\ea \label{BP.9}
\textit{ā zemān bābā xwāđā bīyen, ā zemān ī mexlūqū hewrāmānātū ēme reʕīsšā je ʕērāqew čā kurdewārīū ʕērāqīne bīyen} \\ 
\gll ā zemān bābā xwāđā bīye=n ā zemān ī mexlūq=ū hewrāmān-āt=ū ēme reʕīs=šā je ʕērāq=ew čā kurdewārī=ū ʕērāq-ī=ne bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} time\textsc{.m} \textsc{dem.prox} people\textsc{=ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{=ez.gen} \textsc{1pl} leader\textsc{.m}\textsc{=3pl} from \textsc{pn}\textsc{=\textsc{post}} there kurdish\_region\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Babab Khwada, the people of Hawraman were under Iraq’s rule, under the control of Kurdish regions in Iraq.'
\z 
 
\ea \label{BP.12}
\textit{biřew bārešā sānān, āmēnē pēse čāweɫī begzāyāw wēmā} \\ 
\gll biře=ū bāre=šā sānā=n āmē=nē pēse č=ā eweɫ-ī begzā-yā=ū wē=mā \\ 
 portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} nobleman\textsc{-obl}\textsc{.pl}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1pl}} \\ 
\glt `They would collect tax. They would come (here) [and] like our noblemen..'
\z 
 
\ea \label{BP.15}
\textit{hečkes xelēš bē, xelēš čāne sānēnē} \\ 
\gll hečkes xele-ē=š b-ē xele-ē=š č=ā=ne sānē=nē \\ 
 whoever wheat\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be.prs\textsc{-pstc}\textsc{.3sg} wheat\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} from=\textsc{dem.dist}\textsc{=\textsc{post}} take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Whoever had wheat, they would collect the wheat from him as tax.'
\z 
 
\ea \label{BP.17}
\textit{ā pāse hetā hečkesē ženīš dē šū, bešepīyākew ēme begzāyē ēme} \\ 
\gll ā pāse hetā hečkes-ē ženī=š de-ē šū bešepīyā-eke=ū ēme begzāye-ē ēme \\ 
 \textsc{dem.dist} like even whoever\textsc{-indf} woman.f\textsc{=3sg} give.prs\textsc{-pstc}\textsc{.3sg} husband\textsc{.m} quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{1pl} nobleman\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} \\ 
\glt `[Things were] like this. Even when someone wanted to get his daughter married, he had to pay a quota  to our noblemen.'
\z 
 
\ea \label{BP.19}
\textit{ā zemāne māčū} \\ 
\gll ā zemān=e m-āč-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{ind-}say.prs\textsc{-\textsc{1sg}} \\ 
\glt `I am talking about earlier times. '
\z 
 
\ea \label{BP.20}
\textit{jā ī dāstānē hinū bābāw pīr šelīyārīne} \\ 
\gll jā ī dāstāne-ē hin=ū bābā=ū pīr šelīyār-ī=ne \\ 
 afterwards \textsc{dem.prox} story.f\textsc{-obl}.f ez.pron\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `Anyhow, this story is about the grandfather of Pir Shaliyar.'
\z 
 
\ea \label{BP.21}
\textit{bābāw pīr šelīyārī hukmū bābāw pīr šelīyārī ā zemāne hewrāmān bīyen je ʕērāqo āmēnē hukimšā kerden} \\ 
\gll bābā=ū pīr šelīyār-ī hukm=ū bābā=ū pīr šelīyār-ī ā zemān=e hewrāmān bīye=n je ʕērāq=o āmē=nē hukim=šā kerde=n \\ 
 grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} rule\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} rule\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s grandfather, at the time of his religious rule, Hawraman was under the rule of Iraq.'
\z 
 
\ea \label{BP.24}
\textit{e čē biřew bārešā sānān je hewrāmānī pēse begzāyekāw ī beynū ēme} \\ 
\gll e=čē biře=ū bāre=šā sānā=n je hewrāmān-ī pēse begzāye-ekā=ū ī beyn=ū ēme \\ 
 in=here portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{-obl}\textsc{.m} like nobleman\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=ez.gen} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `[who] would collect taxes from Hawraman, [just] like our noblemen [in recent] times.'
\z 
 
\ea \label{BP.36}
\textit{je terefū ženū menū čīwe genekārī kerā nelāwe} \\ 
\gll je teref=ū ženī=ū men=ū čīw=e genekārī ∅-ker-ā ne-l-ā=we \\ 
 from side\textsc{=ez.gen} woman.f=and \textsc{rdp}=and thing\textsc{.m}\textsc{=dem} debauchery\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They wished to engage in debauchery with women and so on. They didn’t go back.'
\z 
 
\ea \label{BP.38}
\textit{ī kābrēče je ʕērāqo āmēnē, wēregānew nīmeřonew seʕbne bexšnāyšā bexšnēnēwe} \\ 
\gll ī kābrā-ē=č=e je ʕērāq=o āmē=nē wēregā=ne=ū nīmeřo=ne=ū seʕbe=ne bexšnāy=šā bexšn-ēn-ē=we \\ 
 \textsc{dem.prox} man\textsc{-dir}\textsc{.pl}\textsc{=add}\textsc{=dem} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} evening\textsc{=\textsc{post}}=and noon\textsc{=\textsc{post}}=and morning.f\textsc{=\textsc{post}} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-indf}\textsc{=compl} \\ 
\glt `They (people) would donate [food] to the fellows (the tax collectors) who had come from Iraq, in the evenings, mornings, and at noon. '
\z 
 
\ea \label{BP.40}
\textit{her wēregāne pēse čāweɫ sempāšē bēnē, yoīšā dēnē duīšā dēnē yānew yoy ī meʕmūrīye ke ʕērāqo āmēnē, ā zemānū pīr šelīyārī bexšnāyšā bexšnēnē} \\ 
\gll her wēregā=ne pēse čāweɫ sempāš-ē b-ēn-ē yo-ī=šā de-ēn-ē duē-ī=šā de-ēn-ē yāne=ū yo-ī ī meʕmūr-ī=e ke ʕērāq=o āmē=nē ā zemān=ū pīr šelīyār-ī bexšnāy=šā bexšn-ēn-ē \\ 
 just evening\textsc{=\textsc{post}} like in\_the\_past poison\_sprayer\textsc{-dir}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} two\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} house\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} officer\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=dem} rel \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `As in the old days when there were people whose job was spraying pesticide on trees, [and] each would be hosted at the house of one [host], the officers who had come from Iraq would be fed in the time of Pir Shaliyar. '
\z 
 
\ea \label{BP.43}
\textit{yānew yo terī, her šewē her wēregāne ī mēmānē mēmānū yoy bīyēne} \\ 
\gll yāne=ū yo ter-ī her šewe-ē her wēregā=ne ī mēmān-ē mēmān=ū yo-ī bīyē=ne \\ 
 house\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} just night.f\textsc{-obl}.f just evening\textsc{=\textsc{post}} \textsc{dem.prox} guest\textsc{.m}\textsc{-dir}\textsc{.pl} guest\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `[likewise, one tax collector] would be the guest in the house of one [fellow from Hawraman]  each evening, each night.'
\z 
 
\ea \label{BP.45}
\textit{milā xeɫk pāne milo miđyo īne ā xerīkū genekārīnē ī kābrāy je āmēnē, ī meʕmūrē xerīkū genekārīnēw} \\ 
\gll mi-l-ā xeɫk p=ā=ne mi-l-o mi-đye-o īne=ā xerīk=ū genekārī=nē ī kābrā-ī je āmē=nē ī meʕmūr-ē xerīk=ū genekārī=nē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3pl} people\textsc{.m} at=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}look.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} man\textsc{-obl}\textsc{.m} from come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `People realised that these men were busy engaging in debauchery; these officers were busy engaing in debauchery.'
\z 
 
\ea \label{BP.47}
\textit{māčā, ī īnē āmēnē melāwe’} \\ 
\gll m-āč-ā ī īnē āmē=nē me-l-ā=we \\ 
 \textsc{ind-}say.prs\textsc{-3pl} \textsc{dem.prox} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They (people) said, ‘These (tax collectors) have come here; they have not gone back yet.’'
\z 
 
\ea \label{BP.53}
\textit{mēmānekeyš sere biřo māđām īne kāršā bīyen} \\ 
\gll mēmān-ekey=š sere ∅-biř-o māđām īne kār=šā bīye=n \\ 
 guest\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} head\textsc{.m} \textsc{sbjv-}cut.prs\textsc{-3sg} as\_long\_as \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Since they (the tax collectors) are involved in this (debauchery), everyone should decapitate his guest.'
\z 
 
\ea \label{BP.54}
\textit{seʕbē girđtā beydē mizgī, ā mizgīyew hewrāmānī,} \\ 
\gll seʕbe-ē girđ=tā b-e-ydē mizgī ā mizgī=e=ū hewrāmān-ī \\ 
 morning.f\textsc{-obl}.f all=\textsc{2pl} \textsc{sbjv-}come.prs-\textsc{2pl} mosque\textsc{.m} \textsc{dem.dist} mosque\textsc{.m}\textsc{=dem}\textsc{=ez.gen}\textsc{-obl}\textsc{.m} \\ 
\glt `Tomorrow, all of you attend the mosque, that mosque of Hawraman.'
\z 
 
\ea \label{BP.56}
\textit{jā girđmā bilmē pey ʕērāqī pey lāw ī āġeyšā bilmērewo,} \\ 
\gll jā girđ=mā bi-l-mē pey ʕērāq-ī pey lā=ū ī āġe-ī=šā bi-l-mē=re=o \\ 
 afterwards all\textsc{=\textsc{1pl}} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} for at\_the\_place\_of\textsc{=ez.gen} \textsc{dem.prox} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}\textsc{=compl} \\ 
\glt `Then, we all go to Iraq, to their agha. We go down there.'
\z 
 
\ea \label{BP.57}
\textit{āđ māčo, konē ī meʕmūrē mine činnē wextēn? je hewrāmān bīyēnē āmēnē pey biřew bāre} \\ 
\gll āđ m-āč-o ko=nē ī meʕmūr-ē min=e činnē wext-ē=n je hewrāmān bīyē=nē āmē=nē pey biře=ū bāre \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} where\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1sg}\textsc{=dem} some\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} for portion\textsc{.m}=and profit\textsc{.m} \\ 
\glt `He (the agha) says, ‘where have my officers been during this time? They were in Hawraman; they had gone to collect tax.'
\z 
 
\ea \label{BP.60}
\textit{ēmeyč māčmē, ī meʕmūrē šime kīyāsēnē bā biřew bāremā dēw nāmēwe bārāšo} \\ 
\gll ēme=yč m-āč-mē ī meʕmūr-ē šime kīyāsē=nē ∅-b-ā biře=ū bāre=mā dā-ē=ū ne-āmā-ē=we b-ār-ā=š=o \\ 
 \textsc{1pl}\textsc{=add} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{ind-}be.prs\textsc{-3pl} portion\textsc{.m}=and profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}bring.prs\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `We will say, ‘We paid tax to the officers whom you had sent, [however,] they didn’t return to bring it back.'
\z 
 
\ea \label{BP.63}
\textit{ēmeyč muxteserū kelāmī ā īney wāčmē,} \\ 
\gll ēme=yč muxteser=ū kelām-ī ā īney wāč-mē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} \textsc{dem.dist} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} say.prs.sbjv\textsc{-\textsc{1pl}} \\ 
\glt `As for us, to cut a long story short, we say this.'
\z 
 
\ea \label{BP.64}
\textit{jā īne qewɫ kerā yo} \\ 
\gll jā īne qewɫ ∅-ker-ā yo \\ 
 afterwards \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} arrangement\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} one\textsc{.m} \\ 
\glt `Then, they (the people of Hawraman) all agreed [about the plan]\footnote{Lit. they made their oaths one.}'
\z 
 
\ea \label{BP.66}
\textit{īney māčā beynū wēšāne} \\ 
\gll īney m-āč-ā beyn=ū wē=šā=ne \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-3pl} among\textsc{=ez.gen} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `They said this to each other,'
\z 
 
\ea \label{BP.68}
\textit{ēme milmē īney māčmē} \\ 
\gll ēme mi-l-mē īney m-āč-mē \\ 
 \textsc{1pl} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \\ 
\glt `Then, we will go [to Iraq] and say this (i.e. what we agreed on)'
\z 
 
\ea \label{BP.69}
\textit{īney māčmē.’} \\ 
\gll īney m-āč-mē \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \\ 
\glt `We will say this.’'
\z 
 
\ea \label{BP.71}
\textit{neferē yāneš pī ser degāwe bo, tirēštew kārdū čīwiš mebo} \\ 
\gll nefer-ē yāne=š p=ī ser degā=we ∅-b-o tirēšte=ū kārdī=ū čīw=iš me-b-o \\ 
 person\textsc{.m}\textsc{-indf} house\textsc{.m}\textsc{=3sg} at=\textsc{dem.prox} on village.f\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3sg} ax\textsc{.m}=and knife.f\textsc{=ez.gen} thing\textsc{.m}\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg} \\ 
\glt `The house of one of the locals was at the top of the village. He had neither axe, nor knife, nor anything else. '
\z 
 
\ea \label{BP.76}
\textit{miđyā īne be mišārī sereš biřyen nāmā nāmā nīyā} \\ 
\gll mi-đye-ā īne be mišār-ī sere=š biřye=n ne-āmā-∅ ne-āmā-∅ nīye=ā \\ 
 \textsc{ind-}look.prs\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} to hand\_saw\textsc{.m}\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=3sg} cut\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They noticed that the one who had decapitated his guests with a hand saw hadn’t come to the mosque. He hadn’t arrived. He wasn’t there.'
\z 
 
\ea \label{BP.79}
\textit{milāymē pey ʕērāqī milāymē ī xeberī bermēwe} \\ 
\gll mi-lā-ymē pey ʕērāq-ī mi-lā-ymē ī xeber-ī ∅-ber-mē=we \\ 
 \textsc{ind-}go.prs \textsc{1pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs \textsc{1pl} \textsc{dem.prox} news\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}take.prs\textsc{-\textsc{1pl}}\textsc{=compl} \\ 
\glt `We are going to Iraq. We are going [there] to pass this news [to the agha in Iraq]”'
\z 
 
\ea \label{BP.80}
\textit{milo, milā māčā miđyā čowe āmā} \\ 
\gll mi-l-o mi-l-ā m-āč-ā mi-đye-ā č=o=we āmā-∅ \\ 
 \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} \textsc{ind-}look.prs\textsc{-3pl} in=\textsc{dem.dist}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `They were about to go and collect him. They noticed that he came from the other side.'
\z 
 
\ea \label{BP.90}
\textit{māčmē ī meʕmūrū šime āmē lāw ēme} \\ 
\gll m-āč-mē ī meʕmūr=ū šime āmā-ē lā=ū ēme \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} come\textsc{.pst}\textsc{-3pl} at\_the\_place\_of\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `We will say [to the agha], “your officers came to our region".'
\z 
 
\ea \label{BP.96}
\textit{ī meʕmūrū šimemā sere biřyēnē} \\ 
\gll ī meʕmūr=ū šime=mā sere biřyē=nē \\ 
 \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl}\textsc{=\textsc{1pl}} head\textsc{.m} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We have decapitated your officers.'
\z 
 
\ea \label{BP.97}
\textit{seremā biřyē, īne sereū ēmew īne šimšērū šime} \\ 
\gll sere=mā biřī-ē īne sere=ū ēme=ū īne šimšēr=ū šime \\ 
 head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} \textsc{2pl} \\ 
\glt `We have decapitated them. Here are their heads and the agha’s swords.\footnote{Lit. These are our heads and these are the agha’s swords.}'
\z 
 
\ea \label{BP.98}
\textit{ā wextī ēti seʕbe milā milāne mizgī, mizyāre beynē gināre řā beynē bilā pey ʕērāqī jemʕē bā ħemey ġeybīū bābā xuāđā ginā weɫēwe} \\ 
\gll ā wext-ī ēti seʕbe mi-l-ā mi-l-ān=e mizgī mi- zye-ā=re beynē ∅-gin-ā=re řā beynē bi-l-ā pey ʕērāq-ī jemʕ-ē ∅-b-ā ħeme=ī ġeybī=ū bābā xuāđā ∅-gin-ā weɫē=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} morning.f \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}go.prs\textsc{-3pl}=drct mosque\textsc{.m} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} a\_while \textsc{ind-}fall.prs\textsc{-3pl}\textsc{=\textsc{povb}} road.f a\_while \textsc{sbjv-}go.prs\textsc{-3pl} for \textsc{pn}\textsc{-obl}\textsc{.m} added\textsc{-pstc}\textsc{.3sg} \textsc{ind-}be.prs\textsc{-3pl} \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m} \textsc{pn} \textsc{ind-}fall.prs\textsc{-3pl} front\textsc{=\textsc{post}} \\ 
\glt `So, in the morning they went to the mosque. They went a little way towards Iraq. They all gathered. Hama Ghaybi and Baba Khwada went in front of them.'
\z 
 
\ea \label{BP.100}
\textit{āwetū xuɫkī īse ā wextīč xuɫk her tersezāɫ bo,} \\ 
\gll āđet=ū xuɫk-ī īse ā wext-ī=č xuɫk her tersezāɫ ∅-b-o \\ 
 habit\textsc{=ez.gen} people\textsc{.m}\textsc{-obl}\textsc{.m} now \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} people\textsc{.m} just coward \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `Just like people nowdays, in earlier times people were intimidated [by kings]. '
\z 
 
\ea \label{BP.105}
\textit{bābā xwāđāw ħemey ġeybīū ħemey wirdīkɫe her ginēnē weɫē īnē weɫēne ejošā girđ īnē šūnīšāwe} \\ 
\gll bābā xwāđā=ū ħeme=ī ġeybī=ū ħeme=ī wirdīkɫe her gin-ēn-ē weɫē īnē weɫē=ne ejo=šā girđ īnē šūnī=šā=we \\ 
 grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible=and \textsc{pn}=ez.\textsc{attr} small just fall.prs\textsc{-pstc}\textsc{-3pl} front \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} front\textsc{=\textsc{post}} guess.prs\textsc{=3pl} all \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} track.f\textsc{=3pl}\textsc{=compl} \\ 
\glt `Baba Khwada, Hama the invisible, and little Hama were in front, assuming that everyone was following them.'
\z 
 
\ea \label{BP.107}
\textit{wextē miđyo her ā yerē neferēnē ewēšā gird gēɫēnēwe} \\ 
\gll wext-ē mi-đye-o her ā yerē nefer-ē=nē ewēšā gird gēɫē=nē=we \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} just \textsc{dem.dist} three person\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} \\ 
\glt `They realised that there were only the three of them. All the rest had returned [to Hawraman]'
\z 
 
\ea \label{BP.108}
\textit{wextē miđyo her ā yerē neferēnē ewēšā gird gēɫēnēwe} \\ 
\gll wext-ē mi-đye-o her ā yerē nefer-ē=nē ewēšā gird gēɫē=nē=we \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} just \textsc{dem.dist} three person\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} \\ 
\glt ` They realised that there were only the three of them. All the rest had returned [to Hawraman].'
\z 
 
\ea \label{BP.109}
\textit{fermāwo, īne konē xeɫkeke čenū ēme bē?} \\ 
\gll ∅-fermāw-o īne ko=nē xeɫk-eke čenū ēme b-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} people\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} with\textsc{.ez.gen} \textsc{1pl} be.prs\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Baba Khwada) said, ‘Where are the people who were with us?'
\z 
 
\ea \label{BP.111}
\textit{īne konē ewēšā? čēššā pene āmā?} \\ 
\gll īne ko=nē ewēšā čēš=šā pene āmā-∅ \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} what\textsc{=3pl} to come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `Where are they? What happened to them?'
\z 
 
\ea \label{BP.112}
\textit{īne konē ewē? čēššā pene āmā?} \\ 
\gll īne ko=nē ewē čēš=šā pene āmā-∅ \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} what\textsc{=3pl} to come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `Where are they? What happened to them?'
\z 
 
\ea \label{BP.119}
\textit{milmē īney qomīyān,} \\ 
\gll mi-l-mē īney qomyā=n \\ 
 \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We are going [there since] this has happened. '
\z 
 
\ea \label{BP.121}
\textit{māčmē ēme serū īney ī pīyēmā kuštēnē”’} \\ 
\gll m-āč-mē ēme ser=ū īney ī pīyē=mā kuštē=nē \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{1pl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox} man\textsc{.dir}\textsc{.pl}\textsc{=\textsc{1pl}} kill\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We say, ‘We have killed these men because of this.’'
\z 
 
\ea \label{BP.122}
\textit{ā wextī reʕīsū milāre ʕērāq lāw āġekeyšā māčo seremā biřyēnē bābāw pīr šelīyārī bo, ā wextīyekey ħemey ġeybīū yoyč bo nāmē ħemey wirdīkɫe,} \\ 
\gll ā wext-ī reʕīs=ū mi-l-ā=re ʕērāq lā=ū āġe-ekey=šā m-āč-o sere=mā biřyē=nē bābā=ū pīr šelīyār-ī ∅-b-o ā wext-ī-ekey ħeme=ī ġeybī=ū yo=yč ∅-b-o nāmē ħeme=ī wirdīkɫe \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} leader\textsc{.m}\textsc{=ez.gen} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{pn} at\_the\_place\_of\textsc{=ez.gen} agha\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}say.prs\textsc{-3sg} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}=ez.\textsc{attr} invisible=and one\textsc{.m}\textsc{=add} \textsc{ind-}be.prs\textsc{-3sg} name.f \textsc{pn}=ez.\textsc{attr} small \\ 
\glt `The head of those who went to Iraq to the agha, saying, ‘We have decapitated [your men], was Pir Shaliyar’s grandfather, Baba Khwada. In addition [there were] Hama the invisible and Little Hama. '
\z 
 
\ea \label{BP.123}
\textit{ħemey wirdīkɫe ā wextī, pēse īsey māčā heɫāy neyāwān sinʕe zīnāniš nekermē yā čīw yā čāne} \\ 
\gll ħeme=ī wirdīkɫe ā wext-ī pēse īse-ī m-āč-ā heɫāy ne-yāwā=n sinʕe zīnān=iš ne-ker-mē yā čīw yā č=ā=ne \\ 
 \textsc{pn}=ez.\textsc{attr} small \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} like now\textsc{-obl}\textsc{.m} \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} age.f prison\textsc{.m}\textsc{=3sg} \textsc{neg-}do.prs\textsc{-\textsc{1pl}} either thing\textsc{.m} either in=\textsc{dem.dist}\textsc{=\textsc{post}} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) will say, ‘Hama Wirdikla has not reached adulthood; we won’t imprison him, and so on.’'
\z 
 
\ea \label{BP.124}
\textit{ā wextī milārewe māčā, wiɫā ī meʕmūrū šime kīyāsēbēnētā pey īne, īnemā āne heke šime gerekbē dēmā penew} \\ 
\gll ā wext-ī mi-l-ā=re=we m-āč-ā wiɫā ī meʕmūr=ū šime kīyāsē b-ēn-ē=tā pey īne īne=mā āne heke šime gerek b-ē dā-ē=mā pene=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.\textsc{prs}\textsc{-3pl}\textsc{=\textsc{povb}}\textsc{=compl} \textsc{ind-}say.prs\textsc{-3pl} by\_God \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl}=\textsc{2pl} for \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1pl}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} if \textsc{2pl} necessary be.prs\textsc{-pstc}\textsc{.3sg} give\textsc{.pst}\textsc{-3pl}\textsc{=\textsc{1pl}} to=and \\ 
\glt `Then, they went [to the agha and] said, ‘Indeed, the officers whom you had sent to us, whatever [ taxation] you had asked for, we gave them,'
\z 
 
\ea \label{BP.128}
\textit{īne serew ēmew īne šimšērū āġey} \\ 
\gll īne sere=ū ēme=ū īne šimšēr=ū āġe-ī \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Here are their heads and the agha’s swords.'
\z 
 
\ea \label{BP.133}
\textit{īneyšā mekerdē zīnān} \\ 
\gll īney=šā me-ker-dē zīnān \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=3pl} \textsc{neg-}do.prs-\textsc{2pl} prison\textsc{.m} \\ 
\glt `Do not put this one of them in prison.'
\z 
 
\ea \label{BP.143}
\textit{ħemey wirdīkɫeyč mičkile bo, wirdīkɫe bo pāne girđ seferē milo pā šārne gēɫo} \\ 
\gll ħeme=ī wirdīkɫe=yč mičkile ∅-b-o wirdīkɫe ∅-b-o p=ā=ne girđ sefer-ē mi-l-o p=ā šār=ne ∅-gēɫ-o \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} small \textsc{ind-}be.prs\textsc{-3sg} small \textsc{ind-}be.prs\textsc{-3sg} at=\textsc{dem.dist}\textsc{=\textsc{post}} all trip\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} at=\textsc{dem.dist} city\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}wander.prs\textsc{-3sg} \\ 
\glt `Hama Wirdikla was small and young. He would constantly go around in the city.'
\z 
 
\ea \label{BP.144}
\textit{milo pāne swāɫū nānī kero,} \\ 
\gll mi-l-o p=ā=ne swāɫ=ū nān-ī ∅-ker-o \\ 
 \textsc{ind-}go.prs\textsc{-3sg} at=\textsc{dem.dist}\textsc{=\textsc{post}} begging\textsc{=ez.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}do.prs\textsc{-3sg} \\ 
\glt `He would go there and beg for some food.'
\z 
 
\ea \label{BP.150}
\textit{ī due nefere hešpišā dēnēne, gīršā wārden} \\ 
\gll ī due nefer=e hešpišī-ā dē=nē=ne gīr=šā wārde=n \\ 
 \textsc{dem.prox} two.f person\textsc{.m}\textsc{=dem} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} hook\textsc{.m}\textsc{=3pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `These two persons (Baba Khwada and Hama Ghaybi) had gotten lice. They had gotten stuck.'
\z 
 
\ea \label{BP.154}
\textit{māčā, čī girewī āne miheme?’} \\ 
\gll m-āč-ā čī ∅-girew-ī āne miheme \\ 
 \textsc{ind-}say.prs\textsc{-3pl} why \textsc{ind-}cry.prs-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{pn} \\ 
\glt `They said, ‘Mihamad, why are you crying?’'
\z 
 
\ea \label{BP.157}
\textit{ēme mezānmē tekbīrmā čēš bo īnīšā} \\ 
\gll ēme me-zān-mē tekbīr=mā čēš ∅-b-o īnīšā \\ 
 \textsc{1pl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} advice\textsc{=\textsc{1pl}} what \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `We don’t know what we should do, and so on.'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.161}
\textit{bizāne yānew kē pī činnē hāy wexte nānē čīwēšā dānī ārdenit pey ēme, ārdenit pey ēmew} \\ 
\gll bi-zān-e yāne=ū kē p=ī činnē hāy wext=e nān-ē čīw-ē=šā dā=nī ārde=n=it pey ēme ārde=n=it pey ēme=ū \\ 
 \textsc{sbjv-}know.prs-\textsc{2sg} house\textsc{.m}\textsc{=ez.gen} who at=\textsc{dem.prox} some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{=dem} meal\textsc{.m}\textsc{-indf} thing\textsc{.m}\textsc{-indf}\textsc{=3pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl}=and \\ 
\glt `see which household has given you food, which you brought to us, during this period!'
\z 
 
\ea \label{BP.165}
\textit{wāče her kesī nān dānī, wāče, ēšew be xātirū xwāy toū īđū ī kiɫfetū yāneyte dūr ginewe čī šārī’} \\ 
\gll wāč-e her kes-ī nān dā=nī wāč-e ēšew be xātir=ū xwā-ī to=ū īđ=ū ī kiɫfet=ū yāne-ī=t=e dūr ∅-gin-e=we čī šār-ī \\ 
 say.prs-\textsc{imp-}\textsc{2sg} just person\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} say.prs-\textsc{imp-}\textsc{2sg} tonight to sake\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}=and 3sg.prox\textsc{.dir}\textsc{.m}=and \textsc{dem.prox} child\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} far \textsc{imp-}fall.prs-\textsc{2sg}\textsc{=compl} why city\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Tell whoever gave you food, say, “Tonight, for the sake of God, you and your household get away from this town.'
\z 
 
\ea \label{BP.167}
\textit{ā wextī milo mēwe herkesī goš dārābone, dūr kewtenwe je, kučē je šāreke meselen e čē luān tā milemārfā wēšū xānewādeš} \\ 
\gll ā wext-ī mi-l-o m-ē=we herkesī goš dārā b-one dūr kewte=n=we je kuč-ē je šār-eke meselen e=čē luā=n tā milemārfā wē=š=ū xānewāde=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}come.prs\textsc{.3sg}\textsc{=compl} whoever ear\textsc{.m} hold\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3sg} far fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} from little\textsc{-indf} from city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} for\_example in=here go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until \textsc{pn} \textsc{reflx}\textsc{=3sg}=and family\textsc{.m}\textsc{=3sg} \\ 
\glt `Then, he went [to town] and came back. Whoever might have listened to him had gone away from the city. For instance, they had gone together with their family as far as Milamarfa\footnote{name of a summer habitat.}.'
\z 
 
\ea \label{BP.168}
\textit{ā wextī bābāw pīr šelīyārī milowe dūrē leqē māɫo zīnānekeyre} \\ 
\gll ā wext-ī bābā=ū pīr šelīyār-ī mi-l-o=we dūr-ē leq-ē ∅-māɫ-o zīnān-ekey=re \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs\textsc{-3sg}\textsc{=compl} far\textsc{-indf} kick\textsc{-indf} \textsc{ind-}smash.prs\textsc{-3sg} prison\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Then, Pir Shaliyar’s grandfather went backwards in his cell and kicked the wall of the prison.'
\z 
 
\ea \label{BP.173}
\textit{bāzem dūbāre ī sefer wextē milowe dimā leqē māɫo zīnānekeyre zīnāneke yekser veɫādīm bo} \\ 
\gll bāzem dūbāre ī sefer wext-ē mi-l-o=we dimā leq-ē ∅-māɫ-o zīnān-ekey=re zīnān-eke yekser veɫādīm ∅-b-o \\ 
 again again \textsc{dem.prox} trip time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg}\textsc{=compl} afterwards kick\textsc{-indf} \textsc{ind-}smash.prs\textsc{-3sg} prison\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} totally collapsed \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `This time, when he went backward and kicked the prison [wall], the prison totally collapsed.'
\z 
 
\ea \label{BP.178}
\textit{āhā māčā īnā čū dīmne lāw beynū zāwerīnew tewēɫēne ħemey ġeybī hen} \\ 
\gll āhā m-āč-ā īnā-∅ č=ū dīm=ne lā=ū beyn=ū zāwer-ī=ne=ū tewēɫē=ne ħeme=ī ġeybī hen-∅ \\ 
 \textsc{intj} \textsc{ind-}say.prs\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{.3sg} side\textsc{.m}\textsc{=\textsc{post}} at\_the\_place\_of\textsc{=ez.gen} among\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{=\textsc{post}} \textsc{pn}=ez.\textsc{attr} invisible exist\textsc{-3sg}\textsc{.m} \\ 
\glt `It is said that, on this side of the border, there is the shrine of Hama the invisible between the villages of Zawar and Tawela.'
\z 
 
\ea \label{BP.197}
\textit{ā wextī ētir āđē zānā ġeyb bīyen zemīnegere luān} \\ 
\gll ā wext-ī ētir āđē ∅-zān-ā ġeyb bīye=n zemīn-eke=re luā=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} 3pl\textsc{.dir} \textsc{ind-}know.prs\textsc{-3pl} disappearance be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} land\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, they realised that he had disappeared into the ground. '
\z 
 
\ea \label{BP.203}
\textit{ā wextī penj šiš neferē je hewrāmāno milā hesārē muweqetīyū ħemey ġeybī kēšā} \\ 
\gll ā wext-ī penj šiš nefer-ē je hewrāmān=o mi-l-ā hesār-ē muweqetī=ū ħeme=ī ġeybī kēšā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} five six person\textsc{.m}\textsc{-indf} from \textsc{pn}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} wall\textsc{-indf} temporary\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible pull\textsc{.pst} \\ 
\glt `Then five, six people went from Hawraman [and] raised a temporary wall around Hama Ghaybi’s shrine.'
\z 
 
\ea \label{BP.204}
\textit{īse ā hesāre kēštenšā kerdenšā hesārē qābilīyetdārū} \\ 
\gll īse ā hesār=e kēšte=n=šā kerde=n=šā hesār-ē qābilīyetdār=ū \\ 
 now \textsc{dem.dist} wall\textsc{=dem} weigh\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} wall\textsc{-indf} decent=and \\ 
\glt `Now, that [temporary] wall has become a decent wall.'
\z 
 
\ea \label{BP.208}
\textit{īne dāstānū īney, īnē dāstānū bābāw pīr šelīyārīū ħemey ġeybī bīyēnē} \\ 
\gll īne dāstāne=ū īney īnē dāstāne=ū bābā=ū pīr šelīyār-ī=ū ħeme=ī ġeybī bīyē=nē \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story.f\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \textsc{pn}=ez.\textsc{attr} invisible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `This was the story of Pir Shaliyar’s grandfather and Hama Ghaybi.'
\z 
 
\ea \label{BP.209}
\textit{luēnē īnešā wene qomīyān} \\ 
\gll luē=nē īne=šā wene qomyā=n \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3pl} at happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] these things happened to them.'
\z 
 
\ea \label{BP.210}
\textit{ī kāršā āmān milo} \\ 
\gll ī kār=šā āmā=n mil=o \\ 
 \textsc{dem.prox} task\textsc{.m}\textsc{=3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} on\textsc{=\textsc{post}} \\ 
\glt `These things happened to them.'
\z 
 
\ea \label{BP.211}
\textit{werū īney diruē neđāw werū īney je lāw xwāywe negināwe diro ħīčē suālē diruēšā neđēne} \\ 
\gll wer=ū īney diruē ne-đā=ū wer=ū īney je lā=ū xwā-ī=we ne-gin-ā=we diro ħīč-ē suāl-ē diruē=šā ne-đē=ne \\ 
 out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} lie.f \textsc{neg-}give\textsc{.pst}=and out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} from at\_the\_place\_of\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \textsc{neg-}fall.prs\textsc{-3pl}\textsc{=compl} lie nothing\textsc{-indf} question\textsc{-indf} lie.f\textsc{=3pl} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `In order not to lie, in order not to be liars in the presence of God, they didn’t tell a lie.'
\z 
 
\ea \label{BP.212}
\textit{īnē dāstānū āđī tā ewā} \\ 
\gll īnē dāstāne=ū āđī tā ewā \\ 
 \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} until there \\ 
\glt `This was his tale. [It was] until there.'
\z 
 

\ea \label{ŠJ.7}
\textit{be beqēwew tifengē hīnē tifengū ā wextī herčē bīyen pā kešāne gēɫān} \\ 
\gll be beq-ēwe=ū tifeng-ē hīn-ē tifeng=ū ā wext-ī herčē bīye=n p=ā keš-ā=ne gēɫā=n \\ 
 to partridge\textsc{.m}\textsc{-indf}=and gun\textsc{-indf} \textsc{fill}\textsc{-indf} gun\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} whatever be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at=\textsc{dem.dist} mountain\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} wander\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[He had] a partridge and a gun—of whatever type it was back then—[and] would wander in those mountains.'
\z 
 
\ea \label{ŠJ.8}
\textit{pā kešāne gēɫānū wēregā luāno} \\ 
\gll p=ā keš-ā=ne gēɫā=n=ū wēregā luā=n=o \\ 
 at=\textsc{dem.dist} mountain\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} wander\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and evening go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `He would wander in those mountains and would come back in the evening. '
\z 
 
\ea \label{ŠJ.9}
\textit{duē yerē mēnē pāne xeɫkū mentīqew hīnī jāfir sānū luhonī} \\ 
\gll duē yerē m-ē-nē p=ā=ne xeɫk=ū mentīqe=ū hīn-ī jāfir sān=ū luhon-ī \\ 
 two three \textsc{ind-}come.prs\textsc{-3pl} at=\textsc{dem.dist}\textsc{=\textsc{post}} people\textsc{.m}\textsc{=ez.gen} region\textsc{.m}\textsc{=ez.gen} \textsc{fill}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Two three men went\footnote{Lit. came} there [to Biyare]; they were from Jafir Sān’s region, from Luhon.'
\z 
 
\ea \label{ŠJ.12}
\textit{tifengiš henū īneš henū āneš henū!} \\ 
\gll tifeng=iš hen-∅=ū īne=š hen-∅=ū āne=š hen-∅=ū \\ 
 gun\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist-3sg.m}=and \\ 
\glt `He has a gun, he has this, and he has that!'
\z 
 
\ea \label{ŠJ.19}
\textit{ānen pīyā!} \\ 
\gll āne=n pīyā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} man\textsc{.m} \\ 
\glt ` He is the man!’'
\z 
 
\ea \label{ŠJ.30}
\textit{māčā qurbān īne čiko bīyenī} \\ 
\gll m-āč-ā qurbān īne či=ko bīye=nī \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=where be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `They said, ‘Sir, where have you been?'
\z 
 
\ea \label{ŠJ.48}
\textit{perčī nāqāyme miđyo bīyen ženī ēhwāɫiš fāřyān} \\ 
\gll per- č=ī nāqāym=e mi-đye-o bīye=n ženī ēhwāɫ=iš fāřyā=n \\ 
 \textsc{ptcl}- at=\textsc{dem.prox} sudden\textsc{=dem} \textsc{ind-}look.prs\textsc{-3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} woman.f demeanour\textsc{=3sg} change\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All of a sudden, he realised that he had turned into a woman; his demeanour had changed;'
\z 
 
\ea \label{ŠJ.49}
\textit{ēhwāɫiš fāřyān} \\ 
\gll ēhwāɫ=iš fāřyā=n \\ 
 demeanour\textsc{=3sg} change\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `His demeanour had changed.'
\z 
 
\ea \label{ŠJ.51}
\textit{kirž māčo be wiɫā[xdār] ī pīyāw yāneyš} \\ 
\gll kirž m-āč-o be wiɫāxdār ī pīyā=ū yāne-ī=š \\ 
 quick \textsc{ind-}say.prs\textsc{-3sg} to stableman \textsc{dem.prox} man\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `Quickly, he said to the stableman, to the servant in his house,'
\z 
 
\ea \label{ŠJ.52}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `‘Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.53}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.59}
\textit{ā wextī mē milo, jāfir sān māčo be pīyākeyš} \\ 
\gll ā wext-ī m-ē mi-l-o jāfir sān m-āč-o be pīyā-ekey=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{.3sg} \textsc{ind-}go.prs\textsc{-3sg} \textsc{pn} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} to man\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `Jafir Sān went [to Tawela]. He said to his servant,'
\z 
 
\ea \label{ŠJ.64}
\textit{wiɫāxekā, berekey kero milone ew dīm wēš wizone tewēɫeke} \\ 
\gll wiɫāx-ekā bere-ekey ∅-ker-o mi-l-one ew dīm wē=š ∅-wiz-one tewēɫe-eke \\ 
 horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} door\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}do.prs\textsc{-3sg} \textsc{ind-}go.prs\textsc{-3sg} \textsc{dem.dist} side\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}throw.prs\textsc{-3sg} stable\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `He opened the door and went inside. He forced himself into the stable.'
\z 
 
\ea \label{ŠJ.90}
\textit{xo řīk megino ī kāre pey girđ kesī bo} \\ 
\gll xo řīk me-gin-o ī kār=e pey girđ kes-ī ∅-b-o \\ 
 dis.ptc arranged \textsc{neg-}fall.prs\textsc{-3sg} \textsc{dem.prox} task\textsc{.m}\textsc{=dem} to all person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} \\ 
\glt `These (tasks) cannot be done by everyone.'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{ŠJ.99}
\textit{kāke gīyān īnī ečānē bīyēnē} \\ 
\gll kāke gīyān īnī e=čāne-ē bīyē=nē \\ 
 brother.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.pl} in=like\_that\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Dear brother, they were like this.'
\z 
 
\ea \label{ŠJ.102}
\textit{enāne hetā fermāwī ī mentēqew ʕērāqū ērān desū ānīšāwe bīyen} \\ 
\gll enā=ne hetā ∅-fermāw-ī ī mentēqe=ū ʕērāq=ū ērān des=ū ānīšā=we bīye=n \\ 
 otherwise\textsc{=\textsc{post}} even \textsc{ind-}say.prs-\textsc{2sg} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{=ez.gen} \textsc{pn} hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All these regions in Iraq and Iran were under their rule.'
\z 
 
\ea \label{ŠJ.103}
\textit{desū ānīšāwe bīyen} \\ 
\gll des=ū ānīšā=we bīye=n \\ 
 hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `It was under their rule.'
\z 
 
\ea \label{KŠ.1}
\textit{ew wextī īnē her girđ řāzū eyemnē lāšo bīyenā eyām} \\ 
\gll ew wext-ī īnē her girđ řāz=ū eye=m=nē lā=š=o bīye=nā eyā=m \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox}\textsc{.3pl} just all tale\textsc{.m}\textsc{=ez.gen} mother\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3pl} at\_the\_place\_of\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} mother\textsc{=\textsc{1sg}} \\ 
\glt `Well, these are all my mother’s tales. I was with my mother.'
\z 
 
\ea \label{KŠ.3}
\textit{wāčē ħukmū ā wextī milo kešene} \\ 
\gll wāč-ē ħukm=ū ā wext-ī mi-l-o keš=ene \\ 
 say\textsc{.prs}\textsc{-pstc}\textsc{.3sg} ruler\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} mountain\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `According to her,\footnote{Lit. She said.} the king of time went to a mountain.'
\z 
 
\ea \label{KŠ.15}
\textit{ā wextīyekey, yewāšē wātiš pey, kināčēw pādšāyt pey kē niwīse?} \\ 
\gll ā wext-ī-ekey yewāšē wāt=iš pey kināčē=ū pādšā-ī=t pey kē niwīs-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} then say\textsc{.pst}\textsc{=3sg} to daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Then, he (the king) said, ‘Whom have you considered [as the future husband] for the king’s daughter?'
\z 
 
\ea \label{KŠ.18}
\textit{benāxātirē īney āyeč bilo, hāmɫē bē} \\ 
\gll benāxātirē īney āye=č bi-l-o hāmɫē b-ē \\ 
 for\_the\_reason \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} 3sg\textsc{.f}\textsc{.dir}\textsc{=add} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} pregnant\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `The reason [for issuing this order] was that she (the shepherd’s wife), who was pregnant, would disappear.'
\z 
 
\ea \label{KŠ.23}
\textit{ā wextī zāwɫeke peyā bī šūnū peyā bīyeyšre eyākēš merđe} \\ 
\gll ā wext-ī zāwɫe-eke peyā bī-∅ šūn=ū peyā bīyey=š=re eđā-ekē=š merđ-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} child\textsc{-def}\textsc{.dir}\textsc{.m} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} after\textsc{=ez.gen} visible be\textsc{.inf}\textsc{=3sg}\textsc{=\textsc{post}} mother\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} die\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Well, the child was born. His mother died after his birth.'
\z 
 
\ea \label{KŠ.28}
\textit{wātiš, qurbān herkey gele berī epāne min ī bizeme čekyo} \\ 
\gll wāt=iš qurbān herkey gele ∅-ber-ī epāne min ī bize=m=e ∅-čekye-o \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} each\_time herd\textsc{.m} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} there \textsc{1sg} \textsc{dem.prox} goat\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{ind-}drain\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `[and] said, ‘Sir, whenever you take the herd there, my goat[‘s milk] is drained.'
\z 
 
\ea \label{KŠ.30}
\textit{luā kewt šūnī wextē miđyo kuřēwe āmā čowe, wešɫe} \\ 
\gll luā kewt-∅ šūnī wext-ē mi-đye-o kuř-ēwe āmā č=o=we wešɫe \\ 
 go\textsc{.pst}\textsc{.3sg} fall\textsc{.pst}\textsc{-3sg}\textsc{.m} track\textsc{.f} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} boy\textsc{.m}\textsc{-indf} come\textsc{.pst}\textsc{.3sg} from=\textsc{dem.dist}\textsc{=compl} adorable \\ 
\glt `He (the shepherd) went [and] looked after her. He saw that a baby boy approached [the goat]'
\z 
 
\ea \label{KŠ.38}
\textit{ew wextīekey yewāšē tā bo be tāqetē xeɫkī, xeɫk beroš yānew wēš} \\ 
\gll ew wext-ī-ekey yewāšē tā ∅-b-o be tāqet-ē xeɫk-ī xeɫk ∅-ber-o=š yāne=ū wē=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} then until \textsc{ind-}be\textsc{.prs}\textsc{-3sg} by support\textsc{.m}\textsc{-indf} people\textsc{.m}\textsc{-obl}\textsc{.m} people\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} house\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg} \\ 
\glt `Well then, as he became an entertainer for people, people would take him into their own houses.'
\z 
 
\ea \label{KŠ.40}
\textit{kināčēw pāđšāyč māčo, kuřū šuwāneyč bārdē ēme tāqetmā pey āney mē} \\ 
\gll kināčē=ū pāđšā-ī=č m-āč-o kuř=ū šuwāne-ī=č b-ār-dē ēme tāqet=mā pey āney m-ē \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} \textsc{1pl} persistence\textsc{.m}\textsc{=\textsc{1pl}} to \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `The king’s daughter said, ‘Fetch the shepherd’s son. He will entertain us.\footnote{Lit. Our patience comes by him.}’'
\z 
 
\ea \label{KŠ.43}
\textit{māčo be eyāšū tāteyšū āđīšā, bā īne menowe pey nokerū ēme epēne pey werdesīmā pey hermānberū yāneymā} \\ 
\gll m-āč-o be eyā=š=ū tāte-ī=š=ū āđīšā bā īne ∅-men-o=we pey noker=ū ēme epēne pey werdes-ī=mā pey hermānber=ū yāne-ī=mā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by mother\textsc{=3sg}=and father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and 3pl\textsc{.obl} \textsc{hort} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} to servant\textsc{.m}\textsc{=ez.gen} \textsc{1pl} here to attendant\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} to worker\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `She (the girl) said to her parents and others , ‘Let him stay here as our servant here, as our attendant, as our domestic worker.'
\z 
 
\ea \label{KŠ.46}
\textit{ā wextīyekey xeɫk milo māčo be bābeyšū īnīšā māčā, be xwā pāse dīyārā} \\ 
\gll ā wext-ī-ekey xeɫk mi-l-o m-āč-o be bābe-ī=š=ū īnīšā m-āč-ā be xwā pāse dīyār=ā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} people\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} by God\textsc{.m} like apparent\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then people went [and] said to his father and so on, ‘Indeed, it seems that [he is well-integrated into the king’s family]'
\z 
 
\ea \label{KŠ.51}
\textit{māčo, be xwā pāđšām seɫāmet kuřū šuwāney heke to kuštēū ā wextīyekey hukmit dāū īnīšā īse īnā yānetā īnīšā} \\ 
\gll m-āč-o be xwā pāđšā=m seɫāmet kuř=ū šuwāne-ī heke to kušt-ē=ū ā wext-ī-ekey hukm=it dā=ū īnīšā īse īnā-∅ yāne=tā īnīšā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} king\textsc{.m}\textsc{=\textsc{1sg}} healthy son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} if \textsc{2sg} kill\textsc{.pst}\textsc{-3pl}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} rule\textsc{.m}\textsc{=\textsc{2sg}} give\textsc{.pst}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} now \textsc{deic}\textsc{-3sg}\textsc{.m} house\textsc{.m}=\textsc{2pl} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `They said [in the letter], ‘Indeed, may my king be healthy, the shepherd’s son whom you ordered to be killed is now in your house and so on’'
\z 
 
\ea \label{KŠ.56}
\textit{yewāšē čāge, čogene šewē minwīso, māčo, ey kesūkārū min, ey henū nīyenū min, ey īnīšā ke mintā gereknā, tāyfew min heɫbetene ī kuře biřiš dāwe, be her jor ī kuřī beynne berdē.’} \\ 
\gll yewāšē čāge č=oge=ne šewe-ē mi-nwīs-o m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey īnīšā ke min=tā gerek=nā tāyfe=ū min heɫbete=ne ī kuř=e biř=iš dā=we be her jor ī kuř-ī beyn=ne ∅-ber-dē \\ 
 then there in=there\textsc{=\textsc{post}} night\textsc{.f}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} rel \textsc{1sg}=\textsc{2pl} necessary\textsc{=cop}\textsc{.\textsc{1sg}} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} by any means \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} between\textsc{=\textsc{post}} \textsc{ind-}take\textsc{.prs}-\textsc{2pl} \\ 
\glt `Then, one night he (the king) wrote [a letter], saying, ‘O my relatives, O my close family members , O those who want me, [are] from my tribe, [when you read this letter] it is obvious that this boy has arrived. By any means possible, kill him.'
\z 
 
\ea \label{KŠ.57}
\textit{īnē dilē nāmekeyne minvīso} \\ 
\gll īnē dilē nāme-ekey=ne mi-nvīs-o \\ 
 \textsc{dem.prox}\textsc{.3pl} inside letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wrote this inside the letter [and gave it to the shepherd’s son].'
\z 
 
\ea \label{KŠ.59}
\textit{yewāšē dewrū dimāw ʕīšāy kero, šewe kero wārāy wāro mē ew peřū degē} \\ 
\gll yewāšē dewr=ū dimā=ū ʕīšā-ī ∅-ker-o šewe ∅-ker-o wārāy ∅-wār-o m-ē ew peř=ū degē \\ 
 then around\textsc{\textsc{=ez.gen}} afterwards\textsc{\textsc{=ez.gen}} evening\_prayers\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} night\textsc{.f} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} rain\textsc{.adv} \textsc{ind-}rain\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f} \\ 
\glt `Then, it was either during the evening prayers or during the night that he arrived at the other side of the village while it was raining. '
\z 
 
\ea \label{KŠ.72}
\textit{niwīsebēš, ey kesūkārū min, ey henū nīyenū min, heɫbetene ī kuřī biřiš dāwe, šime be wesīlēwe kuždēš, beynne berdēš!} \\ 
\gll niwīse b-ē=š ey kesūkār=ū min ey hen=ū nīyen=ū min heɫbete=ne ī kuř-ī biř=iš dā=we šime be wesīle-ēwe ∅-kuž-dē=š beyn=ne ∅-ber-dē=š \\ 
 write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} \textsc{2pl} by tool\textsc{.m}\textsc{-indf} \textsc{imp-}kill\textsc{.prs}-\textsc{2pl}\textsc{=3sg} between\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2pl}\textsc{=3sg} \\ 
\glt `He had written, ‘O my relatives, O my close family members, [when you read this letter] it is evident that this boy has arrived. Kill him with a thing, destroy him!'
\z 
 
\ea \label{KŠ.73}
\textit{īnēč dīyāš pore bezeyīš āmā pore} \\ 
\gll īnē=č dīyā=š pore bezeyī=š āmā pore \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=add} look\textsc{.pst}\textsc{=3sg} at pity\textsc{.m}\textsc{=3sg} come\textsc{.pst}\textsc{.3sg} at \\ 
\glt `She looked at him. She took pity on him.'
\z 
 
\ea \label{KŠ.76}
\textit{īne wēš xeberiš nīyāw sewāđiš nīyā} \\ 
\gll īne wē=š xeber=iš nīye=ā=ū sewāđ=iš nīye=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{reflx}\textsc{=3sg} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and literacy\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He wasn’t aware [of the death plot] that \_\_ he was illiterate\_\_'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.81}
\textit{yewāšē seʕbē hurmēzāwe māčo, xeyrāmāyš kero merhebāyš kero māčo, tāteym hāɫū ēhwāɫtā?’} \\ 
\gll yewāšē seʕbe-ē hur- mi-ēz-ā=we m-āč-o xeyrāmāy=š ∅-ker-o merhebā-ī=š ∅-ker-o m-āč-o tāte-ī=m hāɫ=ū ēhwāɫ=tā \\ 
 then morning\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} welcoming\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} hello\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} state\textsc{.m}\textsc{\textsc{=ez.gen}} demeanour=\textsc{2pl} \\ 
\glt `The following day, they woke up. She said … she welcomed him (the shepherd’s son), greeted him [and] said, ‘How is my father doing?’'
\z 
 
\ea \label{KŠ.82}
\textit{māčo, tātet weš bēw hāɫū ēhwāɫiš xās bē} \\ 
\gll m-āč-o tāte=t weš b-ē=ū hāɫ=ū ēhwāɫ=iš xās b-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m}\textsc{=\textsc{2sg}} good be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and state\textsc{.m}\textsc{\textsc{=ez.gen}} demeanour\textsc{=3sg} fine be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (the shepherd’s son) said, ‘Your father was all right; he was feeling well.'
\z 
 
\ea \label{KŠ.84}
\textit{īnā qāqezēčiš dānā pey} \\ 
\gll īne=ā qāqez-ē=č=iš dā=nā pey \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} letter\textsc{.m}\textsc{-indf}\textsc{=add}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to \\ 
\glt `Look, he has given me a letter.'
\z 
 
\ea \label{KŠ.86}
\textit{wātiš dey lue āney fiɫānīū fiɫānī čiře ī qewmāme bā bēnēwe qāqezekey wenā} \\ 
\gll wāt=iš dey ∅-lu-e āney fiɫānī=ū fiɫānī ∅-čiř-e ī qewm-ā=m=e bā b-ē-nē=we qāqez-ekey ∅-wen-ā \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} such\_and\_such\_person=and such\_and\_such\_person \textsc{imp-}call\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{hort} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}read\textsc{.prs}\textsc{-3pl} \\ 
\glt `She said, ‘Then go, and summon such-and-such, our relatives, so that they come and read the letter.’'
\z 
 
\ea \label{KŠ.87}
\textit{lue čiře wextē miđyo wāteniš, luāydē ā mejlistā girđ nīštīdēre hurneēzdēwe her čā māreš biřdē} \\ 
\gll ∅-lu-e ∅-čiř-e wext-ē mi-đye-o wāte=n=iš luā-ydē ā mejlis=tā girđ nīšt-īdē=re hur-ne-ēz-dē=we her čā māre=š ∅-biř-dē \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}call\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} go\textsc{.pst}-\textsc{2pl} \textsc{dem.dist} gathering\textsc{.m}=\textsc{2pl} all sit\textsc{.pst}-\textsc{2pl}\textsc{=\textsc{povb}} \textsc{pvb-}\textsc{neg-}rise\textsc{.prs}-\textsc{2pl}\textsc{=compl} just there marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2pl} \\ 
\glt `They (the relatives) went [to the palace]. They learned that he (the king) had said [in the letter], ‘Go and sit in that gathering. Then do not rise [from your seats] until you have married her to him.’'
\z 
 
\ea \label{KŠ.89}
\textit{ā wextīyekey māreš biřā peyšū} \\ 
\gll ā wext-ī-ekey māre=š ∅-biř-ā pey=š=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} marraige\textsc{=3sg} \textsc{ind-}cut\textsc{.prs}\textsc{-3pl} for\textsc{=3sg}=and \\ 
\glt `They married her to him.'
\z 
 
\ea \label{KŠ.91}
\textit{ētir ew čone āyā xeber zāno yā mezāno,} \\ 
\gll ētir ew č=o=ne āyā xeber ∅-zān-o yā me-zān-o \\ 
 \textsc{disc.ptcl} \textsc{dem.dist} in=\textsc{dem.dist}\textsc{=\textsc{post}} whether news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} or \textsc{neg-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `[It is not clear] whether he (the king) was aware of the wedding or not. '
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{KŠ.97}
\textit{yewāšē neferēwe řemo weɫēwe māčo be xānī, be xwā āne to kārū xwāyt tefre dā,} \\ 
\gll yewāšē nefer-ēwe ∅-řem-o weɫē=we m-āč-o be xān-ī be xwā āne to kār=ū xwā-ī=t tefre dā \\ 
 then person\textsc{.m}\textsc{-indf} \textsc{ind-}run\textsc{.prs}\textsc{-3sg} front\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by chief\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2sg} task\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} evasion give\textsc{.pst} \\ 
\glt `Then, a person went forward and said to the Khan, ‘Indeed, you [tried to] evade God’s destiny.'
\z 
 
\ea \label{KŠ.99}
\textit{ā wextīyekey čārenwīsī niwīsebēš ā kināčē to pey kuřū šuwāney niwīsyene} \\ 
\gll ā wext-ī-ekey čārenwīs-ī niwīse b-ē=š ā kināčē to pey kuř=ū šuwāne-ī niwīsye=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} fortune\_teller\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{dem.dist} daughter\textsc{.dir}\textsc{.f} \textsc{2sg} to son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}.pass\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `Then, the fortune teller had allocated\footnote{Lit. written.} your daughter for the shepherd’s son. '
\z 
 
\ea \label{KŠ.100}
\textit{āne kuřū šuwāneyn serū esbekey to} \\ 
\gll āne kuř=ū šuwāne-ī=n ser=ū esb-ekey to \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} on\textsc{\textsc{=ez.gen}} horse\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{2sg} \\ 
\glt `Look, the shepherd’s son [is] on your horse.'
\z 
 
\ea \label{KŠ.101}
\textit{āne kināčekētne ženīšne} \\ 
\gll āne kināčē-ekē=t=ne ženī=š=ne \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.f} woman\textsc{.f}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `That is your daughter. She is his wife.'
\z 
 
\ea \label{KŠ.102}
\textit{āney zāroɫešān} \\ 
\gll āney zāroɫe=šā=n \\ 
 \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} child\textsc{=3pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `That is their child.'
\z 
 
\ea \label{KŠ.104}
\textit{ā wextīyekey īmāniš ārd čā, wātiš xwā to heqnī} \\ 
\gll ā wext-ī-ekey īmān=iš ārd-∅ čā wāt=iš xwā to heq=nī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} belief\textsc{=3sg} bring\textsc{.pst}\textsc{-3sg}\textsc{.m} there say\textsc{.pst}\textsc{=3sg} God\textsc{.m} \textsc{2sg} right\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Then, he (the king) believed in God [and] said, ‘God, you are right!'
\z 
 
\ea \label{ŽH.5}
\textit{e čē milo bilo pey ew dīmū kelī} \\ 
\gll e=čē mi-l-o bi-l-o pey ew dīm=ū kel-ī \\ 
 in=here \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} for \textsc{dem.dist} side\textsc{.m}\textsc{=ez}\textsc{.gen} mountain\textsc{-obl}\textsc{.m} \\ 
\glt `He went from here to the other side of the mountain.'
\z 
 
\ea \label{ŽH.6}
\textit{wiɫātēwe termā hen ew dīmew ēme, pēsew zāwerīū tewēɫey} \\ 
\gll wiɫāt-ēwe ter=mā hen-∅ ew dīm=ew ēme pēse=ū zāwer-ī=ū tewēɫe-ī \\ 
 country\textsc{.m}\textsc{-indf} other\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} \textsc{dem.dist} side\textsc{.m}\textsc{=\textsc{post}} \textsc{1pl} like\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `There is another country on the other side of us; something like Zawar (in Iran) and Tawela (in Iraq) village.'
\z 
 
\ea \label{ŽH.8}
\textit{ā wextīyekey yewāšē milo derey, tūšū neferēwe mē} \\ 
\gll ā wext-ī-ekey yewāšē mi-l-o dere-ī tūš=ū nefer-ēwe m-ē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{fill} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} valley\textsc{.m}\textsc{-obl}\textsc{.m} encounter\textsc{=ez}\textsc{.gen} person\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `Then, he went to a slope [and] ran into a person.'
\z 
 
\ea \label{ŽH.22}
\textit{ā wextīyekey, kināče gewrekēš māčo, tāte, her wēt āmānīwe?’} \\ 
\gll ā wext-ī-ekey kināčē=e gewre-ekē=š m-āč-o tāte her wē=t āmā=nī=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} big\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m} just \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `Then, his older daughter said, ‘Father, have you come back alone?’'
\z 
 
\ea \label{ŽH.27}
\textit{wātiš, tāte āne wāteniš, hekāyetē weš gīsnene, bā ī řē kučē keme kermē be hekāyetekeyt rāz biřmē} \\ 
\gll wāt=iš tāte āne wāte=n=iš hekāyet-ē weš-∅ ∅-gīsn-e=ne bā ī řē kuč-ē kem-e ∅-ker-mē be hekāyet-ekey=t rāz ∅-biř-mē \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} tale\textsc{.m}\textsc{-indf} well-m \textsc{imp-}start\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{hort} \textsc{dem.prox} road\textsc{.obl}\textsc{\textsc{.f}}.sg little\textsc{-indf} s\textsc{hort}\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} by tale\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{2sg}} tale\textsc{.m} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl} \\ 
\glt `She said, ‘Father, he had said, “Start [telling] a good tale, so that we [are entertained] and our way seems shorter. Let us finish your tale.”'
\z 
 
\ea \label{ŽH.33}
\textit{wātiš, tāte, āne wāteniš, eger to xuɫkem mekerī, meberīm yāne milū mizgī’} \\ 
\gll wāt=iš tāte āne wāte=n=iš eger to xuɫke=m me-ker-ī me-ber-ī=m yāne mi-l-ū mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} if \textsc{2sg} invitation\textsc{=1sg} \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \textsc{neg-}take\textsc{.prs}-\textsc{2sg}\textsc{=1sg} house\textsc{.m}\textsc{.dir.sg} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} mosque\textsc{.m} \\ 
\glt `She said, ‘Father, he meant,\footnote{Lit. said.} “If you do not invite me and won’t take me home, I will go to the mosque.”'
\z 
 
\ea \label{ŽH.35}
\textit{ā wextī yewāšē berdiš luā mizgī} \\ 
\gll ā wext-ī yewāšē berd-∅=iš luā mizgī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} then take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=3sg} go\textsc{.pst}\textsc{.3sg} mosque\textsc{.m} \\ 
\glt `Well, he (the man) had taken him (Hayas) to the mosque. '
\z 
 
\ea \label{ŽH.40}
\textit{wātiš be wāɫe wičkilekēš ī māsī bere deš pā mēmān mizgī čenīw nānī’} \\ 
\gll wāt=iš be wāɫē=e wičkile-ekē=š ī mās-ī ∅-ber-e ∅-de-e=š p=ā mēmān mizgī čenī=ū nān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} by sister\textsc{\textsc{.f}}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} to\textsc{=dem}.dist guest\textsc{.m} mosque\textsc{.m} with\textsc{=ez}\textsc{.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said to her little daughter, ‘Take this yoghurt [and] together with [some] bread give it to the the guest in the mosque.’ '
\z 
 
\ea \label{ŽH.41}
\textit{ā wextīyekey kināčekē beroš řāre māčo, be xwā īse kučēš čene werū.} \\ 
\gll ā wext-ī-ekey kināčē-ekē ∅-ber-o=š řā=re m-āč-o be xwā īse kuč-ē=š čene ∅-wer-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} now little\textsc{-indf}\textsc{=3sg} from \textsc{ind-}eat\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then, the girl set off on the way. She said, ‘Well, I shall eat some of it now.’'
\z 
 
\ea \label{ŽH.42}
\textit{kučēš čene werū ī māsī’} \\ 
\gll kuč-ē=š čene ∅-wer-ū ī mās-ī \\ 
 little\textsc{-indf}\textsc{=3sg} with \textsc{ind-}eat\textsc{.prs}\textsc{-1sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `I shall eat some of it.'
\z 
 
\ea \label{ŽH.50}
\textit{ā wextīyekey, māčo be tāteyš, dā ā mēmānī bāre pey yāneymā!’} \\ 
\gll ā wext-ī-ekey m-āč-o be tāte-ī=š dā ā mēmān-ī b-ār-e pey yāne-ī=mā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{hort} \textsc{dem.dist} guest\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \\ 
\glt `The, she (the older sister) said to her father, ‘Bring that guest home!’'
\z 
 
\ea \label{ŽH.53}
\textit{māčo, ī serū řīšīme kučē tāše’} \\ 
\gll m-āč-o ī sere=ū řīš-ī=m=e kuč-ē ∅-tāš-e \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox} head\textsc{.m}=and beard\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} little\textsc{-indf} \textsc{imp-}shave\textsc{.prs}-\textsc{2sg} \\ 
\glt `He (Hayas) said [to the girl], ‘shave my face a bit.’'
\z 
 
\ea \label{ŽH.57}
\textit{ā wextīyekey āyeč māčo, weɫāhī tāte, dāwāw ēme yoymā kero’} \\ 
\gll ā wext-ī-ekey āye=č m-āč-o weɫāhī tāte dāwā=w ēme yo-ī=mā ∅-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{\textsc{.f}}\textsc{.dir}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed father\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{1pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then she (the eldest daughter) said, ‘Indeed, father, he is asking for the hand of one of us.’'
\z 
 
\ea \label{ŽH.62}
\textit{īne e čē čenū āġeyš jengšā bīyen, be ādizī lūān} \\ 
\gll īne e=čē čenū āġe-ī=š jeng=šā bīye=n be ādizī lūā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=here with.ez\textsc{.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} quarrel\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} with desperation go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (Hayas) got into a quarrel with his master. He had left [town] in desperation'
\z 
 
\ea \label{ŽH.65}
\textit{yewāšē āđīč mārošo mēwe kināčēw luhonī bo teřū čermew īnīšā} \\ 
\gll yewāšē āđ=īč m-ār-o=š=o m-ē=we kināčē=ū luhon-ī ∅-b-o teř=ū čerme=ū īnīšā \\ 
 then 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} fresh=and white=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `He (Hayas) brought her (i.e. his wife). She was from Luhon; she was young, white, and so on.'
\z 
 
\ea \label{ŽH.70}
\textit{pīney bilo šewē neēwe} \\ 
\gll p=īney bi-l-o šewe-ē ne-ē=we \\ 
 for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{neg-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `The reason [was] so that he (Hayas) would go and not come back for the night. '
\z 
 
\ea \label{ŽH.71}
\textit{řākē dūrew inne dūre bo, pīney šewē biř nekero bēwe} \\ 
\gll řā-ekē dūr=ew inne dūr-e ∅-b-o p=īney šewe-ē biř ne-ker-o b-ē=we \\ 
 road\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} far\textsc{=\textsc{post}} so\_much far\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} piece \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `The distance was so long that he (Hayas) wouldn’t be able to return for the night.'
\z 
 
\ea \label{ŽH.75}
\textit{ā wextīyekey hurmēzo bereš kerowe pey} \\ 
\gll ā wext-ī-ekey hur-m-ēz-o bere=š ∅-ker-o=we pey \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg} door\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} for \\ 
\glt `Then, she rose and opened the door to him.'
\z 
 
\ea \label{ŽH.79}
\textit{ā wextī yewāšē māčo, dey tehemuɫ kere bā hēɫeřūenit pey kerū, hāɫīt kerū'} \\ 
\gll ā wext-ī yewāšē m-āč-o dey tehemuɫ ∅-ker-e bā hēɫeřūen=it pey ∅-ker-ū hāɫī=t ∅-ker-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} then \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} patience \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{hort} fried\_egg\textsc{=\textsc{2sg}} for \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} understood\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then she said, ‘All right, wait a little bit, let me cook you fried eggs, [and] make you understand.'
\z 
 
\ea \label{ŽH.83}
\textit{ā wextī dānēšā sīyāw kero} \\ 
\gll ā wext-ī dāne-ē=šā sīyāw ∅-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} one\textsc{-indf}\textsc{=3pl} black \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The girl coloured one of them black;'
\z 
 
\ea \label{ŽH.86}
\textit{māčo be tūwāɫekeyšā, īne čēšen?’} \\ 
\gll m-āč-o be tūwāɫe-ekey=šā īne čēše=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by shell\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what \\ 
\glt `She pointed to the (egg)shells [and] said, ‘what are these? \footnote{Lit. What is this}''
\z 
 
\ea \label{ŽH.87}
\textit{māčo, īnē māčmēšā hēɫe’} \\ 
\gll m-āč-o īnē m-āč-mē=šā hēɫe \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.3pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3pl} egg\textsc{.m} \\ 
\glt `He (the sultan) said, ‘We call these eggs.’'
\z 
 
\ea \label{ŽH.88}
\textit{māčo, īne bīye čin řengī?’} \\ 
\gll m-āč-o īne bī-e čin řengī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.pst} 3sg\textsc{\textsc{.f}} how\_many colour \\ 
\glt `She said, ‘How many colours have they become?’'
\z 
 
\ea \label{ŽH.89}
\textit{māčo, āne bīyē yerē řengī} \\ 
\gll m-āč-o āne bī-ē yerē řengī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.pst} 3pl three colour \\ 
\glt `He said, ‘They have turned into three colours’'
\z 
 
\ea \label{ŽH.95}
\textit{māčo, īne her hēɫe bī’} \\ 
\gll m-āč-o īne her hēɫe bī-∅ \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} just egg\textsc{.m} be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `He said, ‘This one was just an egg.’'
\z 
 
\ea \label{ŽH.100}
\textit{āne řengw ji berīš ferq kero} \\ 
\gll āne řengī=ū ji ber-ī=š ferq ∅-ker-o \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} colour=and from outside\textsc{-obl}\textsc{.m}\textsc{=3sg} difference \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `It [is] just the colour. They differ in appearance.'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.104}
\textit{ānā dersē ʕālit dānē} \\ 
\gll āne=ā ders-ē ʕāl=it dā-ānē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} lesson\textsc{.m}\textsc{-indf} good\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{-1sg} \\ 
\glt `You gave me a good lesson.'
\z 
 
\ea \label{ŽH.106}
\textit{ā wextī mizyore mizyore milo, milowe} \\ 
\gll ā wext-ī mi-zye-o=re mi-zye-o=re mi-l-o mi-l-o=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `The he left [the house] and returned [home]'
\z 
 
\ea \label{ŽH.107}
\textit{ew mēwe sebāyš, tā pērēš} \\ 
\gll ew m-ē=we sebā-ī=š tā pērē=š \\ 
 \textsc{dem.dist} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} tomorrow\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} until day\_after\_tomorrow\textsc{=3sg} \\ 
\glt `He (Hayas) returned (home) the next morning, or in two days.'
\z 
 
\ea \label{ŽH.109}
\textit{melowe yāne, duē řoē pā degāne bo} \\ 
\gll me-l-o=we yāne duē řo-ē p=ā degā=ne ∅-b-o \\ 
 \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m} two day\textsc{.m}\textsc{-dir}\textsc{.pl} at\textsc{=dem}.dist village\textsc{.obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He didn’t go back home. He was [wandering] in the village for two days. '
\z 
 
\ea \label{ŽH.111}
\textit{māčo, āmāno īnā’} \\ 
\gll m-āč-o āmā=n=o īne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (the sultan) said, ‘He has returned; there he is.’'
\z 
 
\ea \label{ŽH.116}
\textit{āđīč wāt, ā šēre minnā} \\ 
\gll āđ=īč wāt ā šēr=e min=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{dem.dist} lion\textsc{.m}\textsc{=dem} 1sg\textsc{=cop}\textsc{.1sg} \\ 
\glt `He (the sultan) said, ‘I am that lion.'
\z 
 
\ea \label{RE.4}
\textit{bābe gīyān ī kināčēte yā tātemā lūān yā māmomā yā kesūkārmā} \\ 
\gll bābe gīyān ī kināčē=t=e yā tāte=mā lūā=n yā māmo=mā yā kesūkār=mā \\ 
 father.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} either father\textsc{.m}\textsc{=\textsc{1pl}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} either paternal\_uncle\textsc{=\textsc{1pl}} either relative\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `They would say, ‘Dear brother,\footnote{Lit. father.} this daughter of yours \—either our father would go [to the girl’s father], or our uncle, or a relative of ours—'
\z 
 
\ea \label{RE.7}
\textit{ʕerzū birāw wēm kerū ī kināčēte de pī kuřīme} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū ī kināčē=t=e ∅-de-e p=ī kuř-ī=m=e \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} brother\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} at=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \\ 
\glt `Let me tell you,\footnote{Lit. May a petition be at the service of my brother.} ‘Give your daughter to my son [in marriage]'
\z 
 
\ea \label{RE.8}
\textit{ā wextī eger xwāy īrāđeš bīyebo serū qisekē} \\ 
\gll ā wext-ī eger xwā-ī īrāđe=š bīye b-o ser=ū qisē-ekē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} if God\textsc{.m}\textsc{-obl}\textsc{.m} volition\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} on\textsc{\textsc{=ez.gen}} talk\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `Then, if God was willing over the subject,'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.11}
\textit{ētir kābrā fers kere zemānū ā ānē hītē pāɫē bīyēnēū čwār gezē pārče bīyenū} \\ 
\gll ētir kābrā fers ∅-ker-e zemān=ū ā ānē hīt-ē pāɫā-ē bīyē=nē=ū čwār gez-ē pārče bīye=n=ū \\ 
 \textsc{disc.ptcl} man guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} \textsc{dem.dist}\textsc{.3pl} pair\textsc{-indf} shoe\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and four ell\textsc{.m}\textsc{-indf} textile\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Then, let us say,\footnote{Lit. imagine.} at the time, [the wedding gift] consisted of a pair of shoes [and] four gaz\footnote{A unit of measure equivalent to 72 cm.} of textile.'
\z 
 
\ea \label{RE.14}
\textit{ā wextīekey hurēsēnē hemāhengīšā bīyen} \\ 
\gll ā wext-ī-ekey hur-ēsē=nē hemāhengī=šā bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pvb-}rise\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} cooperation\textsc{.m}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, they would coordinate (the wedding)'
\z 
 
\ea \label{RE.16}
\textit{ey pīyā ētir emīnū yāne bīyenū xizmetkār bīyen} \\ 
\gll ey pīyā ētir emīn=ū yāne bīye=n=ū xizmetkār bīye=n \\ 
 \textsc{dem.prox} man\textsc{.m} \textsc{disc.ptcl} confidant\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{.dir}.sg be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez.gen} servant\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was this man who was the family’s confidant. He was a servant.'
\z 
 
\ea \label{RE.18}
\textit{firmānbirđār bīyen, tā sāhībū ī kināčē wāčo, xeber bido wāčo, fiɫāne kār} \\ 
\gll firmānbirđār bīye=n tā sāhīb=ū ī kināčē ∅-wāč-o xeber bi-de-o ∅-wāč-o fiɫān=e kār \\ 
 obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until owner\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} news\textsc{.dir}\textsc{.m} \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} such\_and\_such=ez\textsc{.cmpd} task\textsc{.m} \\ 
\glt `He was obedient. The girl’s father\footnote{Lit. owner.} would say [to him] and ask him, ‘[Do] such-and-such tasks;'
\z 
 
\ea \label{RE.21}
\textit{ā wextīekey firmānbirđār bīyen tā ženekēš berdēne} \\ 
\gll ā wext-ī-ekey firmānbirđār bīye=n tā ženī-ekē=š berdē=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} take\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `He was at their service until the girl married.'
\z 
 
\ea \label{RE.22}
\textit{šūnū ženekēčre ā wextīekey ētir hemāhengīšā weš bīyen} \\ 
\gll šūn=ū ženī-ekē=č=re ā wext-ī-ekey ētir hemāhengī=šā weš bīye=n \\ 
 after\textsc{=ez.gen} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} cooperation\textsc{.m}\textsc{=3pl} good be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `After the marriage, they (the couple's families) were on good terms.'
\z 
 
\ea \label{RE.25}
\textit{ew āmān xeyr āmāyē weš āmāyē beynne bīyenū} \\ 
\gll ew āmā=n xeyr āmāy-ē weš āmāy-ē beyn=ne bīye=n=ū \\ 
 \textsc{dem.dist} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} goodness\textsc{.m} come\textsc{.inf}\textsc{-indf} good come\textsc{.inf}\textsc{-indf} between\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They\footnote{Lit. He.} would come and greet each other and welcome each other.'
\z 
 
\ea \label{RE.26}
\textit{zemānēč ženekēš berdēne ētir čo dumā bīyēnē tēkeɫ} \\ 
\gll zemān-ē=č ženī-ekē=š berdē=ne ētir č=o dumā bīyē=nē tēkeɫ \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=add} woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} take\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} \textsc{disc.ptcl} in=\textsc{dem.dist} afterwards be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} mixed \\ 
\glt `When the brides were taken, they (the families) would become one.'
\z 
 
\ea \label{RE.41}
\textit{ī memliketū ēme ne welē wiɫātāre ētir tāte bīyen čūēw desū šuwāney} \\ 
\gll ī memliket=ū ēme ne welē wiɫāt-ā=re ētir tāte bīye=n čū-ē=w des=ū šuwāne-ī \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} no but region\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{disc.ptcl} father\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} stick\textsc{.m}\textsc{-indf}\textsc{=ez.gen} hand\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[It is not common] in our region, but elsewhere, the father has become like the stick in the shepherd’s hand.\footnote{Meaning that he has no power over whom his children will marry.}'
\z 
 
\ea \label{RE.52}
\textit{pīney bāyē māsīč bero} \\ 
\gll p=īney bāye-ē mās=īč ∅-ber-o \\ 
 for=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} pot\textsc{.m}\textsc{-dir}\textsc{.pl} yoghurt\textsc{.m}\textsc{=add} \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `For the reason that he (the girl’s father) would take a bowl of yoghurt [to him] too.'
\z 
 
\ea \label{RE.56}
\textit{zemānē ēme ī begēmā bīyēnē} \\ 
\gll zemān-ē ēme ī beg-ē=mā bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} \textsc{1pl} \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Once we had these noblemen [in our region]'
\z 
 
\ea \label{RE.57}
\textit{ī begē qibūɫ kerā} \\ 
\gll ī beg-ē qibūɫ ∅-ker-ā \\ 
 \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf} accepted \textsc{sbjv-}do\textsc{.prs}\textsc{-3pl} \\ 
\glt `[After] the noblemen should consent [to the marriage]'
\z 
 
\ea \label{RE.58}
\textit{ā wextīyekey āwyānešā sānān} \\ 
\gll ā wext-ī-ekey āwyāne=šā sānā=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, they (the noblemen) would ask for awyana (i.e. the money in exchange for giving consent to marriage)'
\z 
 
\ea \label{RE.61}
\textit{sūrāne ānen} \\ 
\gll sūrāne āne=n \\ 
 \textsc{pn} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Surana was that.'
\z 
 
\ea \label{RE.62}
\textit{ī memliketū ēme řisūmātū āwe zemānīš e čīne bīyen} \\ 
\gll ī memliket=ū ēme řisūmāt=ū āwe zemān-ī=š e=čīne bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} traditions\textsc{.m}\textsc{=ez.gen} that time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} in=like be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The traditions of earlier times were like this in our region.'
\z 
 
\ea \label{RE.63}
\textit{ī begēmā bīyēnē, begē} \\ 
\gll ī beg-ē=mā bīyē=nē beg-ē \\ 
 \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} Beg\textsc{.m}\textsc{-indf} \\ 
\glt `We had these noblemen.'
\z 
 
\ea \label{RE.64}
\textit{bē ā begā ēme netāwānmā ħīč kermē} \\ 
\gll bē ā beg-ā ēme ne-tāwā=n=mā ħīč ∅-ker-mē \\ 
 without \textsc{dem.dist} Beg\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{1pl} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1pl}} nothing \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `Without them, we were not able to do anything.'
\z 
 
\ea \label{ŽE.4}
\textit{ī degā toš vīnī čoɫe bīyēne} \\ 
\gll ī degā to=š ∅-vīn-ī čoɫ-e bīyē=ne \\ 
 \textsc{dem.prox} village\textsc{.obl}\textsc{\textsc{.f}} \textsc{2sg}\textsc{=3sg} \textsc{ind-}see\textsc{.prs}-\textsc{2sg} deserted-\textsc{f} be\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `This village, which you see, was deserted.'
\z 
 
\ea \label{ŽE.37}
\textit{hēzmī ētir zimsān āmān hurēsēnē yerē āwīr, āwīre kučēšā nīyēnēre bī kipē} \\ 
\gll hēzmī ētir zimsān āmā=n hur-ēsē=nē yerē āwīr āwīr=e kuč-ē=šā nīyē=nē=re bī-∅ kipē \\ 
 firewood\textsc{.f} \textsc{disc}.\textsc{ptcl} winter\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pvb-}rise\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} three fire\textsc{.m} fire\textsc{.m}\textsc{=dem} little\textsc{-dir}\textsc{.pl}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} be\textsc{.pst}\textsc{-3sg}\textsc{.m} sealed \\ 
\glt `Firewood …  Then when it was winter they would rise and start a small fire [and it would] damp down.'
\z 
 
\ea \label{ŽE.50}
\textit{ī doīnete dēnmēū ī šelemīnēte dēnmē} \\ 
\gll ī doīne=t=e dē=nmē=ū ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{\textsc{=ez.gen}} \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this dwina. \footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{ŽE.51}
\textit{ī šelemīnēte dēnmē} \\ 
\gll ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `you gave us this Shalamina. \footnote{Dish made of cracked wheat and turnip.}'
\z 
 
\ea \label{ŽE.52}
\textit{ī xeplēte dēnmē} \\ 
\gll ī xeplē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this xapla '
\z 
 
\ea \label{ŽE.53}
\textit{īnēmā wārdēne} \\ 
\gll īnē=mā wārdē=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=1pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We have eaten this [meal]. '
\z 
 
\ea \label{ŽE.56}
\textit{čāweɫī sextī nebīyen} \\ 
\gll č=ā eweɫ-ī sextī ne-bīye=n \\ 
 in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} difficulty\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the beginning there was no hardship.'
\z 
 
\ea \label{ŽE.57}
\textit{čāweɫī hīč nebīyen} \\ 
\gll č=ā eweɫ-ī hīč ne-bīye=n \\ 
 in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} nothing \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the beginning there was nothing.'
\z 
 
\ea \label{ŽE.71}
\textit{pāne wešɫe wešɫeyī yānekoɫēšā kerden} \\ 
\gll p=āne wešɫe wešɫeyī yānekoɫē=šā kerde=n \\ 
 at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} well happy\_manner moving\_house\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They left the house in a happy manner.'
\z 
 
\ea \label{ŽE.73}
\textit{ā wextī to mīsāl wātēt, ā fiɫāne kesem gerekene’} \\ 
\gll ā wext-ī to mīsāl wāt-ē=t ā fiɫān=e kese=m gerek-e=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg} for\_example say\textsc{.pst}\textsc{-cond}\textsc{=\textsc{2sg}} \textsc{dem.dist} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.f}\textsc{=\textsc{1sg}} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `In earlier times, let us say\footnote{Lit. for instance.} you said, ‘I want that certain person (i.e., girl).''
\z 
 
\ea \label{ŽE.74}
\textit{luēnī lāw ī kuřī wāčēnī, dā lue wāče eđēš bizāne ī kināčēše miđom pene yā ne’} \\ 
\gll lu-ēn-ī lā=ū ī kuř-ī wāč-ēn-ī dā ∅-lu-e ∅-wāč-e eđē=š bi-zān-e ī kināčē=š=e mi-đe-o=m pene yā ne \\ 
 go\textsc{.prs}\textsc{-pstc}-\textsc{2sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.prs}\textsc{-pstc}-\textsc{2sg} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}say\textsc{.prs-}.\textsc{2sg} mother\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=3sg}\textsc{=dem} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{1sg}} to either no \\ 
\glt ` You would go to this boy [seated next to the narrator] and would say, ‘Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!’'
\z 
 
\ea \label{ŽE.75}
\textit{kuřekeyč ē māčī, kināčekēt miđey pī kuřī?’} \\ 
\gll kuř-ekey=č ē m-āč-ī kināčē-ekē=t mi-đe-ī p=ī kuř-ī \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} come\textsc{.prs}\textsc{.pst}c\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The boy comes. [Addressing her grand son] you would say [to her mother], ‘Will you give your daughter to this boy [in marriage]?’'
\z 
 
\ea \label{ŽE.80}
\textit{ēnē hījbīš kerē wāčē, ī kināčē miđey pī kuřī?’} \\ 
\gll ēnē hījbī=š ker-ē wāč-ē ī kināčē mi-đe-ī p=ī kuř-ī \\ 
 come\textsc{.prs}\textsc{.pstc}\textsc{.3pl} betrothal\textsc{=3sg} do\textsc{.prs}\textsc{-pstc} say\textsc{.prs-pstc}\textsc{.3sg} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `They would come [and] act as intermediary to ask for the hand of the girl. They would say, ‘Will you give this girl to this boy?’'
\z 
 
\ea \label{ŽE.83}
\textit{tā qewmekāmā wiš kermē qewmekā wiš kermē bizānmē ī qewmīmā řezāšāne yā řezāšā nīyene} \\ 
\gll tā qewm-ekā=mā wiš ∅-ker-mē qewm-ekā wiš ∅-ker-mē bi-zān-mē ī qewm-ī=mā řezā=šā=ne yā řezā=šā nīye=ne \\ 
 until relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=1pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} agreement\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} or agreement\textsc{.f}\textsc{=3pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We will inform our relatives. We will inform our relatives to see if they agree [to the  marriage] or not.'
\z 
 
\ea \label{ŽM.4}
\textit{šūnū āneyre, šūnū řāy eweɫ je kārīgerīwe serem beršīyen} \\ 
\gll šūn=ū āney=re šūn=ū řā=ī eweɫ je kārīgerī=we sere=m ber-šīye=n \\ 
 after\textsc{=ez.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} after\textsc{=ez.gen} road\textsc{.f}=ez.\textsc{attr} first from labourer\_job\textsc{=compl} head\textsc{.m}\textsc{=\textsc{1sg}} out- go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Well, I first learned\footnote{Lit. my head went out.}  to work as a labourer.'
\z 
 
\ea \label{ŽM.7}
\textit{je kārēz biřyeywe pī heštā sāɫe, īse ħewt sāɫē} \\ 
\gll je kārēz biřyey=we p=ī heštā sāɫe=e īse ħewt sāɫe-ē \\ 
 from subterranean\_canal\textsc{.m} cut\textsc{.inf}\textsc{=compl} at=\textsc{dem.prox} eighty year\textsc{.f}\textsc{=dem} now seven year\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `Ever since [the time I was] digging canals until now that I am eighty years old [I have been working]'
\z 
 
\ea \label{ŽM.8}
\textit{īse ħewt sāɫēč āmānāwe serū īney dūkāndārnā} \\ 
\gll īse ħewt sāɫe-ē=č āmā=nā=we ser=ū īney dūkāndār=nā \\ 
 now seven year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} shopkeeper\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{ŽM.10}
\textit{ā wextīyekey pīr bīyenā} \\ 
\gll ā wext-ī-ekey pīr bīye=nā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} old be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I have aged.'
\z 
 
\ea \label{ŽM.12}
\textit{īne dāstānū minā} \\ 
\gll īne dāstāne=ū min=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{=ez.gen} \textsc{1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is my story.'
\z 
 
\ea \label{ŽM.15}
\textit{ā wextīyekey ečī newʕe řū be werū wēmre řū be xwā řisqēm peyā kerđen} \\ 
\gll ā wext-ī-ekey ečī newʕ=e řū be wer=ū wē=m=re řū be xwā řisq-ē=m peyā kerđe=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} like\_this kind/way\textsc{=dem} face by front\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} face by God\textsc{.m} living\textsc{-indf}\textsc{=\textsc{1sg}} visible do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I have taken refuge in God and found living.'
\z 
 
\ea \label{ŽM.16}
\textit{dārāw īse elʕān hāɫēhāzirī dārāw penj kuřānāw yerē kināčā, heštū wēmū ženē īne de neferē} \\ 
\gll dārā=w īse elʕān hāɫēhāzir-ī dārā=w penj kuř-ā=nā=ū yerē kināčē-ā hešt=ū wē=m=ū ženī-ē īne de nefer-ē \\ 
 owner\textsc{.m}\textsc{=ez.gen} now now at\_the\_moment\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{=ez.gen} five son/boy\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1sg}}=and three daughter\textsc{.f}\textsc{-obl}\textsc{.pl} eight=and \textsc{reflx}\textsc{=\textsc{1sg}}=and woman\textsc{.f}\textsc{-indf} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} ten person\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `I have, at the moment, five sons and three daughters, [which makes] eight, [plus] myself and a wife, [making us] ten people.'
\z 
 
\ea \label{ŽM.19}
\textit{ī menteqew ēme kemūkuřīn} \\ 
\gll ī menteqe=ū ēme kemūkořī=n \\ 
 \textsc{dem.prox} region\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Our region is in poverty.'
\z 
 
\ea \label{ŽM.21}
\textit{ī mentēqew ēme nebīyeyš hen nebīyey} \\ 
\gll ī mentēqe=ū ēme nebīyey=š hen-∅ nebīyey \\ 
 \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} poverty\textsc{.m} \\ 
\glt `Our region is in poverty, [and] shortage.'
\z 
 
\ea \label{ŽM.22}
\textit{ā wextī mišyē bilmē ġārijne čīw bārmēw čīw bermē} \\ 
\gll ā wext-ī mišyē bi-l-mē ġārij=ne čīw b-ār-mē=ū čīw ∅-ber-mē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} aux\textsc{.pst} \textsc{sbjv-}go\textsc{.prs}\textsc{-\textsc{1pl}} abroad\textsc{=\textsc{post}} thing\textsc{.m} \textsc{sbjv-}bring\textsc{.prs}\textsc{-\textsc{1pl}}=and thing\textsc{.m} \textsc{sbjv-}take\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt ` We had to go out [of Hawraman] to fetch and take stuff [to sell].'
\z 
 
\ea \label{ŽM.26}
\textit{ā wextīyekey ħewt no, yānze zāroɫēm bīyēnē} \\ 
\gll ā wext-ī-ekey ħewt no yānze zāroɫe-ē=m bīyē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} seven nine eleven child\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `I had seven or nine or eleven children.'
\z 
 
\ea \label{ŽM.27}
\textit{ī yānze zāroɫē ečē āsēnēm jīyā milū ī ženīme milū nebīyeywe} \\ 
\gll ī yānze zāroɫe-ē e=čē āsē=nē=m jīyā mil=ū ī ženī=m=e mil=ū nebīyey=we \\ 
 \textsc{dem.prox} eleven child\textsc{-dir}\textsc{.pl} in=here leave\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} separate on\textsc{=ez.gen} \textsc{dem.prox} woman\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} on\textsc{=ez.gen} poverty\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Because of destitution, [I had to] leave these eleven children to my wife.'
\z 
 
\ea \label{ŽM.30}
\textit{ā wextīyekey řūtē bīyēnē qūtē bīyēnē bīyenšā nebīyenšā xeberēm nezānēnē} \\ 
\gll ā wext-ī-ekey řūt-ē bīyē=nē qūt-ē bīyē=nē bīye=n=šā ne-bīye=n=šā xeber-ē=m ne-zānē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} naked\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} bare\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} news\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `[I was unaware if] they were without clothes; if they had [enough]; if they didn’t have [enough]. I didn't know their news.'
\z 
 
\ea \label{ŽM.32}
\textit{ā wextīyekey ečāge hīč xeberēmā nebīyenū telefūnīčū žimārew telefūnī nebēw} \\ 
\gll ā wext-ī-ekey e=čāge hīč xeber-ē=mā ne-bīye=n=ū telefūn=īč=ū žimāre=ū telefūn-ī ne-b-ē=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} in=there nothing news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and telephone\textsc{.m}\textsc{=add}=and number\textsc{.m}\textsc{=ez.gen} telephone\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and \\ 
\glt `I\footnote{Lit. We.}  had no news [of my family] from there. There was no phone or phone number at the time.'
\z 
 
\ea \label{ŽM.34}
\textit{luēnmē čārekū kerdenmā ew des} \\ 
\gll luē=nmē čārek=ū kerde=n=mā ew des \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{dem.dist} hand\textsc{.m} \\ 
\glt `We went to Charak and went to the other side.'
\z 
 
\ea \label{ŽM.36}
\textit{ī šiš mānge xeberū hīčīmā nezānān} \\ 
\gll ī šiš mānge=e xeber=ū hīč-ī=mā ne-zānā=n \\ 
 \textsc{dem.prox} six month\textsc{.f}\textsc{=dem} news\textsc{.m}\textsc{=ez.gen} nothing\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I didn’t have any updates [from my family] during those periods of six months.'
\z 
 
\ea \label{ŽM.38}
\textit{extīyārmā mištū ā fārsāne bīyen jezīrew kīšī} \\ 
\gll extīyār=mā mišt=ū ā fārs-ā=ne bīye=n jezīre=ū kīš-ī \\ 
 volition\textsc{.m}\textsc{=\textsc{1pl}} fist\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} Persian\textsc{-3pl}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We were under the rule of Persians at Kish island.'
\z 
 
\ea \label{ŽM.43}
\textit{ā zemāne kuweyt je girđū kēšwerekā dewɫetmenter bē} \\ 
\gll ā zemān=e kuweyt je girđ=ū kēšwer-ekā dewɫetmen-ter b-ē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} from all\textsc{=ez.gen} country\textsc{-def}\textsc{.obl}\textsc{.pl} rich\textsc{-cmpr} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Back then, Kuwait was the richest of all countries.'
\z 
 
\ea \label{ŽM.45}
\textit{ēme jezīre čenū ā jezīrey čeneš bēnmē duē sāʕetē beyniš bē} \\ 
\gll ēme jezīre čenū ā jezīre-ī čene=š b-ēn-mē duē sāʕet-ē beyn=iš b-ē \\ 
 \textsc{1pl} island with\textsc{\textsc{.ez}.gen} \textsc{dem.dist} island\textsc{-obl}\textsc{.m} at\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1pl}} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} distance\textsc{.m}\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `It (Kuwait) was two hours away from the island we were in.'
\z 
 
\ea \label{ŽM.47}
\textit{penj řoē čāne min řemezānim gērten} \\ 
\gll penj řo-ē č=ā=ne min řemezān=im gērte=n \\ 
 five day\textsc{.m}\textsc{-indf} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{1sg} \textsc{pn}\textsc{=\textsc{1sg}} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[while there] I fasted five days during Ramadan.'
\z 
 
\ea \label{ŽM.50}
\textit{ēme čāne kārmā kerđenū} \\ 
\gll ēme č=ā=ne kār=mā kerđe=n=ū \\ 
 \textsc{1pl} in=\textsc{dem.dist}\textsc{=\textsc{post}} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `We worked there.'
\z 
 
\ea \label{ŽM.51}
\textit{wextē āmēnmēwe ī mentēqe ētir čē musteqerē bīyēnmē} \\ 
\gll wext-ē āmē=nmē=we ī mentēqe ētir čē musteqer-ē bīyē=nmē \\ 
 time\textsc{.m}\textsc{-indf} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \textsc{disc}.\textsc{ptcl} here settled\textsc{\textsc{-pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `When I returned to this region, I settled down here.'
\z 
 
\ea \label{ŽM.53}
\textit{ʕerzū birāw wēm kerū āmēnmēwe ī mentēqe} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū āmē=nmē=we ī mentēqe \\ 
 petition\textsc{.m}\textsc{=ez.gen} brother\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \\ 
\glt `Let me tell you. I returned to this region.'
\z 
 
\ea \label{ŽM.57}
\textit{ā wextīyekey ēme īne zinđigīš pene kermē be nebīyey} \\ 
\gll ā wext-ī-ekey ēme īne zinđigī=š pene ∅-ker-mē be nebīyey \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} life\textsc{.m}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1pl}} with poverty\textsc{.m} \\ 
\glt `Well, we live in poverty.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
