\chapter{All 480 examples of dir in these texts are found in the following sentences:}

\ea \label{ZB.1}
\textit{řoē je řoā yoe mē yoyč mē xizmetū šā ʕelādīnī, her ā šā ʕelādīne} \\ 
\gll řo-ē je řo-ā yo-e m-ē yo=yč m-ē xizmet=ū šā ʕelādīn-ī her ā šā ʕelādīn=e \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m}\textsc{-f} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} one\textsc{.m}\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} emph \textsc{dem.dist} king\textsc{.m} \textsc{pn}\textsc{=dem} \\ 
\glt `Once,\footnote{Lit. One day of days.} a woman and a man\footnote{Lit. One (f.) came, one (m.) came.} came to the service of šā ʕalādīn, the same šā ʕalādīn.'
\z 
 
\ea \label{ZB.9}
\textit{kuřēš bā, kuřekēš germīyānne nimēniš čene} \\ 
\gll kuř-ē=š b-ā kuř-ekē=š germīyān=ne ni-m-ē-nē=iš čene \\ 
 son\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.prs}\textsc{-3pl} son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.pl}\textsc{=3sg} \textsc{pn}\textsc{=\textsc{post}} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-3pl}\textsc{=3sg} with \\ 
\glt `He had sons. His sons did not accompany him [;they stayed in] Garmiyan. '
\z 
 
\ea \label{ZB.12}
\textit{tejrebeš bo ʕemriš dewrū ṣewū wīs sāɫāne bo wīs sāɫē sī sāɫē pā dewrne} \\ 
\gll tejrebe=š ∅-b-o ʕemr=iš dewr=ū ṣew=ū wīs sāɫe-ā=ne ∅-b-o wīs sāɫe-ē sī sāɫe-ē p=ā dewr=ne \\ 
 experience\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} age\textsc{.m}\textsc{=3sg} around\textsc{\textsc{=ez.gen}} hundred=and twenty year\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} twenty year\textsc{.f}\textsc{-dir}\textsc{.pl} thirty year\textsc{.f}\textsc{-dir}\textsc{.pl} at=\textsc{dem.dist} around\textsc{=\textsc{post}} \\ 
\glt `He was experienced. He was around 120 years old, about 120 or 130,'
\z 
 
\ea \label{ZB.13}
\textit{māčo, ženekē} \\ 
\gll m-āč-o ženī-ekē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} \\ 
\glt `He said, ‘Wife!’'
\z 
 
\ea \label{ZB.16}
\textit{ewēč māčo, wārānū čēšī} \\ 
\gll ewē=č m-āč-o wārān=ū čēš-ī \\ 
 3sg\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} rain\textsc{\textsc{=ez.gen}} what\textsc{-obl}\textsc{.m} \\ 
\glt `She said, ‘What rain?!’'
\z 
 
\ea \label{ZB.17}
\textit{māčo, be xwā wāro! ānā hewrē sīyāw pise četrē kewt āsmān’} \\ 
\gll m-āč-o be xwā ∅-wār-o āne=ā hewr-ē sīyāw pise četre-ē kewt-∅ āsmān \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to God\textsc{.m} \textsc{ind-}rain\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} cloud\textsc{-indf} black like umbrella\textsc{-obl}\textsc{.f} fall\textsc{.pst}\textsc{-3sg}\textsc{.m} sky\textsc{.m} \\ 
\glt ` He said, ‘By God, it will rain. There, a black cloud has covered the sky like an umbrella!’'
\z 
 
\ea \label{ZB.21}
\textit{heywāne āwē berde} \\ 
\gll heywāne āwē berd-e \\ 
 animal\textsc{.dir}\textsc{.f} water\textsc{.obl}\textsc{.f} take\textsc{.pst} 3sg\textsc{.f} \\ 
\glt `The flood\ref{Lit. water} took away the animals.'
\z 
 
\ea \label{ZB.24}
\textit{šewē be ʕemreū xwāy ženēč nīšore duē zāroɫē wīno her ā šewē} \\ 
\gll šewe-ē be ʕemre=ū xwā-ī ženī-ē=č ∅-nīš-o=re duē zāroɫe-ē ∅-wīn-o her ā šewe-ē \\ 
 night\textsc{.f}\textsc{-obl}\textsc{.f} to order\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m} woman\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=add} \textsc{ind-}sit\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} two child\textsc{-dir}\textsc{.pl} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} emph \textsc{dem.dist} night\textsc{.f}\textsc{-obl}\textsc{.f} \\ 
\glt `During that very night, by the order of God, the wife gave birth to two babies.\footnote{Lit. She sat down [and] saw two babies.} '
\z 
 
\ea \label{ZB.27}
\textit{seʕbē milā wextī miđyā penj bizē menēnēw} \\ 
\gll seʕbe-ē mi-l-ā wext-ī mi-đye-ā penj bize-ē menē=nē=ū \\ 
 morning\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} five goat\textsc{.f}\textsc{-dir}\textsc{.pl} remain\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `The next day, when they looked around, they saw that five goats had survived [from the flood]'
\z 
 
\ea \label{ZB.28}
\textit{āwī luēnērew serū āne hārdīšā bīyēnēū} \\ 
\gll āwī luē=nē=re=ū ser=ū āne hārdī=šā bīyē=nē=ū \\ 
 water\textsc{.f} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}}=and on\textsc{\textsc{=ez.gen}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} flour\textsc{.f}\textsc{=3pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `The water had gone up to the level of the flour [sacks] that they had.'
\z 
 
\ea \label{ZB.31}
\textit{ħeywānē miđā wer. girđ wizāš weɫē.} \\ 
\gll ħeywān-ē mi-đe-ā wer girđ ∅-wiz-ā=š weɫē \\ 
 animal\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} out all \textsc{ind-}throw\textsc{.prs}\textsc{-3pl}\textsc{=3sg} front \\ 
\glt `They herded the animals out [and] they drove them forth.'
\z 
 
\ea \label{ZB.32}
\textit{ā wextīyekey zāroɫē duē bāwšīšāne} \\ 
\gll ā wext-ī-ekey zāroɫe-ē duē bāwšī=šā=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} child\textsc{-dir}\textsc{.pl} two embrace\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `And they held the babies in their arms.'
\z 
 
\ea \label{ZB.33}
\textit{yeženekē biřē pīyāke māčo, melāmā pey’} \\ 
\gll yeženī-ekē biřē pīyā-eke m-āč-o me-l-ā=mā pey \\ 
 woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} a\_while man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}go\textsc{.prs}\textsc{-3pl}\textsc{=1pl} for \\ 
\glt `The woman (or) the man said several times, ‘It's not possible for us [to take the babies]. \footnote{they don't go for us}’'
\z 
 
\ea \label{ZB.37}
\textit{īne min īnem žinīyen lāw, fermāwištū šēxīn, šā ʕelādīnīwe} \\ 
\gll īne min īne=m žinīye=n lā=ū fermāwišt=ū šēx-ī=n šā ʕelādīn-ī=we \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{1sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} speech\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have heard this tale from —it is the Sheikh’s words— Shah Aladin.'
\z 
 
\ea \label{ZB.40}
\textit{milā ew kuřekey yoyšā berā ānešā zilterū ʕālter bo āđī berā} \\ 
\gll mi-l-ā ew kuř-ekey yo-ī=šā ∅-ber-ā āne=šā zil-ter=ū ʕāl-ter ∅-b-o āđī ∅-ber-ā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3pl} big\textsc{-cmpr}\textsc{\textsc{=ez.gen}} good\textsc{-cmpr} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} 3sg\textsc{.obl}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \\ 
\glt `They went away [and took] that son. They took one of them (i.e., the boys), the one who was bigger and healthier; they took him.'
\z 
 
\ea \label{ZB.42}
\textit{čī mede xwāwinū ʕālemīyānī pāse īrāyešā bizēw menēne jīāy sāyqekewe} \\ 
\gll č=ī mede xwāwin=ū ʕālemīyān-ī pāse īrāye=š=ā bize-ēw menē=ne jīāy sāyqe-eke=we \\ 
 at=\textsc{dem.prox} time owner\textsc{\textsc{=ez.gen}} worlds\textsc{.m}\textsc{-obl}\textsc{.m} like volition\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} goat\textsc{.f}\textsc{-indf} remain\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} behind lightning\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `The will of the lord of all worlds was such that a goat was accidentally left behind [healthy] from the lightning [that caused the flood]'
\z 
 
\ea \label{ZB.43}
\textit{xwāy pāse īrāyešā āđē milā řeđē bāw} \\ 
\gll xwā-ī pāse īrāye=š=ā āđē mi-l-ā řeđ-ē b-ā=ū \\ 
 God\textsc{.m}\textsc{-obl}\textsc{.m} like volition\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} crossing\textsc{-dir}\textsc{.pl} be\textsc{.prs}\textsc{-3pl}=and \\ 
\glt `God’s will was such that they went away, [and] crossed [that region],'
\z 
 
\ea \label{ZB.44}
\textit{bizē her werew mē} \\ 
\gll bize-ē her wer=ew m-ē \\ 
 goat\textsc{.f}\textsc{-dir}\textsc{.pl} emph front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `and the goat kept going\footnote{Lit. coming.} forward [to the tree]'
\z 
 
\ea \label{ZB.45}
\textit{pey minyo serū nerekeyū, čērhur zāroɫeke šot wero} \\ 
\gll pey mi-nye-o ser=ū nere-ekey=ū čēr=hur zāroɫe-eke šot ∅-wer-o \\ 
 for \textsc{ind-}put\textsc{.prs}\textsc{-3sg} on\textsc{\textsc{=ez.gen}} oak\_tree\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{\textsc{=ez.gen}} under\textsc{=\textsc{post}} child\textsc{-def}\textsc{.dir}\textsc{.m} milk\textsc{.m} \textsc{ind-}eat\textsc{.prs}\textsc{-3sg} \\ 
\glt `[and] put its feet on the trunk of the oak tree [and] the baby drank [its] milk from below.'
\z 
 
\ea \label{ZB.46}
\textit{sāɫē ter mēnēwe, māčo, bilmē bizānmē tekbīrī ētir řāwēžmā čēš bo’ āđē, wiškesāɫīn?} \\ 
\gll sāɫe-ē ter m-ē-nē=we m-āč-o bi-l-mē bi-zān-mē tekbīr-ī ētir řāwēž=mā čēš ∅-b-o āđē wiškesāɫī=n \\ 
 year\textsc{.f}\textsc{-dir}\textsc{.pl} other \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} advice\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} conjecture\textsc{=1pl} what \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} 3pl\textsc{.dir} drought\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The following year, they went back. He (the father) said, ‘We shall go [and] see what is ahead of us with this period of drought.’'
\z 
 
\ea \label{ZB.49}
\textit{milo wextē miđyo zāroleke bīyen peřū qiřoɫekey} \\ 
\gll mi-l-o wext-ē mi-đye-o zāroɫe-eke bīye=n peř=ū qiřoɫ-ekey \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} child\textsc{-def}\textsc{.dir}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{\textsc{=ez.gen}} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `The father went [closer and] noticed that the baby boy [had grown up so much that] he had filled the tree hollow.'
\z 
 
\ea \label{ZB.54}
\textit{ī bizēče powe āwēzāne bo mē e čē ēđ čeknošū milo lā yāgekēš} \\ 
\gll ī bizē=č=e p=o=we āwēzān-e ∅-b-o m-ē e=čē ēđ čekn-o=š=ū mi-l-o lā yāgē-ekē=š \\ 
 \textsc{dem.prox} goat\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=dem} at=\textsc{dem.dist}\textsc{=compl} hanging\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} in=here 3sg.prox\textsc{.dir} suckle\textsc{.prs}\textsc{-3sg}\textsc{=3sg}=and \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of place\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \\ 
\glt `The goat would stay in a hanging position, the child would feed from its udder, and it would go back to its place.'
\z 
 
\ea \label{ZB.57}
\textit{hāɫbete bero gēroš kēštebēš šiš kīloē āđ čewī zīyāter bīyebē} \\ 
\gll hāɫbete ∅-ber-o ∅-gēr-o=š kēšte b-ē=š šiš kīlo-ē āđ č=ewī zīyā-ter bīye b-ē \\ 
 of\_course \textsc{ind-}take\textsc{.prs}\textsc{-3sg} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} weigh\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} six kilo\textsc{.m}\textsc{-indf} 3sg\textsc{.dir}\textsc{.m} from=3sg\textsc{.obl}\textsc{.m} increased\textsc{-cmpr} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Indeed, he (the father) took him; he grabbed him. When he weighed him, he weighed six kilos more than him (the other son)'
\z 
 
\ea \label{ZB.58}
\textit{īne dāstānū gewregerīū ā mešeyxāne} \\ 
\gll īne dāstāne=ū gewregerī=ū ā mešeyxā=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{\textsc{=ez.gen}} greatness\textsc{\textsc{=ez.gen}} \textsc{dem.dist} sheikh\textsc{.obl}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `This is the story of the greatness of the sheikhs.'
\z 
 
\ea \label{ZQ.12}
\textit{kelēwe dīmāndīmē bē fire tūl bē, dāresān bē āwīš nebē fire nāřihet} \\ 
\gll kel-ēwe dīmāndīm-ē b-ē fire tūl b-ē dāresān b-ē āwī=š ne-b-ē fire nāřihet \\ 
 mountain\textsc{-indf} wide\textsc{-indf} \textsc{sbjv-}come.prs\textsc{.3sg} a\_lot long be.prs\textsc{-pstc}\textsc{.3sg} woodland\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} water\textsc{.dir}.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-pstc}\textsc{.3sg} a\_lot sad \\ 
\glt `It was a big mountain; it was very high; it was a forest; [it was] very uncomfortable.'
\z 
 
\ea \label{ZQ.14}
\textit{ženekēm qomyāš venī kelekewe} \\ 
\gll ženī-ekē=m qomyā=š venī kel-eke=we \\ 
 woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}} happen\textsc{.pst}\textsc{.3sg}\textsc{=3sg} at mountain\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `My wife was about to deliver a baby. '
\z 
 
\ea \label{ZQ.15}
\textit{duē bečkēš dīyē zāroē} \\ 
\gll duē bečke-ē=š dī-ē zāro-ē \\ 
 two baby\textsc{-dir}\textsc{.pl}\textsc{=3sg} see\textsc{.pst}\textsc{-3pl} child\textsc{-dir}\textsc{.pl} \\ 
\glt `She gave birth to two babies.'
\z 
 
\ea \label{ZQ.17}
\textit{kuřeke yošā weš kewtebē kuřū} \\ 
\gll kuř-eke yo=šā weš kewte b-ē kuř=ū \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} one\textsc{.m}\textsc{=3pl} good fall\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-pstc}\textsc{.3sg} son/boy\textsc{.m}=and \\ 
\glt `One of the newborn boys was healthy. '
\z 
 
\ea \label{ZQ.18}
\textit{āđmā tehwēɫ gērtū} \\ 
\gll āđ=mā tehwēɫ gērt=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=\textsc{1pl}} delivery\textsc{.m} grab\textsc{.pst}\textsc{=ez.gen} \\ 
\glt `We got hold of him.'
\z 
 
\ea \label{ZQ.24}
\textit{berdim nīyāmne qiřoɫe dāreke} \\ 
\gll berd-∅=im nīyā=m=ne qiřoɫ=e dār-eke \\ 
 take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1sg}} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `I took [him] and put him in the tree hollow.'
\z 
 
\ea \label{ZQ.26}
\textit{yošā berd nīyāmne qiřoɫe dāreke qurbān!} \\ 
\gll yo=šā berd-∅ nīyā=m=ne qiřoɫ=e dār-eke qurbān \\ 
 one\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{-3sg}\textsc{.m} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} sir.\textsc{voc} \\ 
\glt `Sir, I took one [and] put him in the tree hollow. '
\z 
 
\ea \label{ZQ.29}
\textit{āne ke berdmā šiš māngē menn} \\ 
\gll āne ke berd-∅=mā šiš mānge-ē menn \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{rel} take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{1pl} six month.f\textsc{-dir}\textsc{.pl} remain\textsc{.pst} \\ 
\glt `The one whom we took [with us] lived for six months.'
\z 
 
\ea \label{ZQ.30}
\textit{dimāw šiš māngē fewtiš kerd lūā} \\ 
\gll dimā=ū šiš mānge-ē fewt=iš kerd-∅ lūā \\ 
 afterwards\textsc{=ez.gen} six month.f\textsc{-dir}\textsc{.pl} death\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{-3sg}\textsc{.m} go\textsc{.pst} \\ 
\glt `After six months, he passed away.'
\z 
 
\ea \label{ZQ.32}
\textit{āy bāwke řo, āy tāte řo īđīč luāw ewīč luā} \\ 
\gll āy bāwke řo āy tāte řo īđ=īč luā=ū ew=īč luā \\ 
 \textsc{intj} fahter.\textsc{voc} onom \textsc{intj} father\textsc{.m} onom 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg}=and 3sg\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg} \\ 
\glt `Oh dear God\footnote{The expressions bāwka řo and tāta řo are used to mourn someone’s death, or to express distress.}, he (the child whom I took with me) died; he (the child whom I left behind) died too!\footnote{Lit. This one went; that one, too, went.} '
\z 
 
\ea \label{ZQ.38}
\textit{nizīk bīyāwe temāšem kerd wextē miđyew kuřeke bīyen peřū qiřoɫēwe, qiřoɫekey} \\ 
\gll nizīk bī-ā=we temāše=m kerd wext-ē mi-đye-ū kuř-eke bīye=n peř=ū qiřoɫ-ēwe qiřoɫ-ekey \\ 
 close be\textsc{.pst}\textsc{-3pl}\textsc{=compl} looking\textsc{=\textsc{1sg}} do\textsc{.pst} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{=ez.gen} hollow\textsc{.m}\textsc{-indf} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I got closer and looked inside. I noticed that the boy [had grown up so much that he] had filled a tree hollow, i.e. the tree hollow. '
\z 
 
\ea \label{ZQ.39}
\textit{wextē luānē werew temāšem kerd miđyew be ʕemrew xwāy e čī qiřolī| ēnnew čīčew ženē pīfē peydā bīyenē| ʕeyn pīfē} \\ 
\gll wext-ē luā-ānē wer=ew temāše=m kerd mi-đye-ū be ʕemre=ū xwā-ī ečī qiřoɫ-ī ēnne=w čīče=ū ženī-ē pīfē-ē peydā bīyē=nē ʕeyn pīfē \\ 
 time\textsc{.m}\textsc{-indf} go\textsc{.pst}\textsc{-\textsc{1sg}} front\textsc{.m}\textsc{=\textsc{post}} looking\textsc{=\textsc{1sg}} do\textsc{.pst} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} by order.f\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} in.\textsc{dem.prox} hollow\textsc{.m}\textsc{-obl}\textsc{.m} as\_much\_as\textsc{=ez.gen} nipple\textsc{=ez.gen} woman.f\textsc{-obl}.f burr.f\textsc{-dir}\textsc{.pl} visible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like balloon.f \\ 
\glt `When I got closer and looked inside, I noticed that, by the order of God, some burrs the size of a woman's nipple had appeared in the tree hollow, like balloons.'
\z 
 
\ea \label{ZQ.41}
\textit{āne memkeš īnā je pīfekēwe jīyā bīyeno} \\ 
\gll āne memke=š īnā-∅ je pīfē-ekē=we jīyā bīye=n=o \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} breast\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} from burr.f\textsc{-def}\textsc{.obl}.f\textsc{=compl} separate be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `The breast was clearly distinguished from the burr. '
\z 
 
\ea \label{ZQ.44}
\textit{ī kuře her memeš wārden her memes wārden} \\ 
\gll ī kuř=e her meme=š wārde=n her meme=š wārde=n \\ 
 \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} just breast\textsc{.dir}\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} just breast\textsc{.dir}\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The boy had kept breast-feeding [from the tree]; he had kept breast-feeding .'
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{ZQ.55}
\textit{fermāwo, dey āne ētir deseɫāt xwāy, pey her čē meyliš bē xwāy deseɫātē čānēš henē} \\ 
\gll ∅-fermāw-o dey āne ētir deseɫāt xwā-ī pey herčē meyl=iš b-ē xwā-ī deseɫāt-ē čāne-ē=š hen-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} power\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} to whatever willingness\textsc{=3sg} \textsc{sbjv-}come.prs\textsc{.3sg} God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m-dir.pl} such\textsc{-pl}\textsc{=3sg} \textsc{exist}\textsc{-3pl} \\ 
\glt ` The Shah said, ‘Well, then, this is in God’s hands. Whatever God wishes, he has the power [to exert]'
\z 
 
\ea \label{HB.2}
\textit{nānewēkēw šēh ʕelādīnī seʕbē hurmēzāwe xeber miđā māčā, xeber deydē be hezretū šēxī, be xwā xāneqāne ne nān hen, ne hārdī henē} \\ 
\gll nānewē-ekē=ū šēh ʕelādīn-ī seʕbe-ē hur-mi-ēz ā=we xeber mi-đe-ā mi-āč-ā xeber ∅-de-ydē be hezret=ū šēx-ī be xwā xāneqā=ne=ne nān hen-∅=ne hārdī hen-ē \\ 
 baker\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{.pl}=and sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} morning\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs} \textsc{dem.dist}\textsc{=compl} news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} news\textsc{.m} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} to his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}}=neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m}=nor flour\textsc{\textsc{.f}} exist\textsc{-3pl} \\ 
\glt `Sheikh Aladin’s bakers got up in the morning and passed on the news, saying, ‘Pass on the news to his highness the Sheikh. By God, there is neither bread nor flour in the monastery. '
\z 
 
\ea \label{HB.12}
\textit{bilo pey šeškū Bešāretī, āsāwekew šeškī} \\ 
\gll bi-l-o pey šešk=ū Bešāret-ī āsāw-eke=w šešk-ī \\ 
 \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} for \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} mill\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] go to Shashk Basharat\footnote{A village in Iraqi Kurdistan.}, to the mill of Shashk'
\z 
 
\ea \label{HB.19}
\textit{māčo, āy dāxekey dāxim bē min čen bē ʕeqiɫ bīyā!} \\ 
\gll m-āč-o āy dāx-eke=y dāx=im b-ē min čen bē-ʕeqiɫ bī-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} sorrow\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}.\textsc{attr} sorrow\textsc{=1sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{1sg} how\_much without-wisdom be\textsc{.pst}\textsc{-1sg} \\ 
\glt `He (the donkey breeder) said, ‘Oh, I wish I was dead! Look how stupid I was!’'
\z 
 
\ea \label{HB.31}
\textit{hereke āmāwe zuwāb wātiš, xefetit nebo!} \\ 
\gll her-eke āmā=we zuwāb wāt=iš xefet=it ne-b-o \\ 
 donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} come\textsc{.pst}\textsc{.3sg}\textsc{=compl} reply\textsc{.m} say\textsc{.pst}\textsc{=3sg} sorrow\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The donkey started to speak [and] said, ‘Don’t worry!'
\z 
 
\ea \label{HB.36}
\textit{simerī kerā tūrekekew min} \\ 
\gll simerī ∅-ker-ā tūreke-eke=w min \\ 
 straw\textsc{\textsc{.f}} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} sack\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg \\ 
\glt `They (the family) will put [some] straw into my sack;'
\z 
 
\ea \label{HB.44}
\textit{hereke mēwe juwāb ī qisā kero} \\ 
\gll her-eke m-ē=we juwāb ī qisē-ā ∅-ker-o \\ 
 donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} reply\textsc{.m} \textsc{dem.prox} talk\textsc{\textsc{.f}}\textsc{-obl}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The donkey started to speak; it spoke.'
\z 
 
\ea \label{HB.45}
\textit{ewīč goš miđo pene} \\ 
\gll ew=yč goš mi-đe-o pene \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} ear\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `He listened to it [i.e. the donkey].'
\z 
 
\ea \label{HB.48}
\textit{dāš wene hereš berd gošew šārū šeškī} \\ 
\gll dā=š wene her-e=š berd-∅ goše=w šār=ū šešk-ī \\ 
 give\textsc{.pst}\textsc{=3sg} \textsc{post} donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{=3sg} take\textsc{.pst}\textsc{-3sg}\textsc{.m} corner\textsc{=ez}\textsc{.gen} city\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He set off. He took the donkey to the suburb of Shashk city.'
\z 
 
\ea \label{HB.54}
\textit{pāse herey wāt, simerīš kerdē torekekew āđī} \\ 
\gll pāse her-e-ī wāt simerī=š kerdē-ē toreke-eke=ū āđī \\ 
 like donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} say\textsc{.pst} straw\textsc{.dir}\textsc{.pl}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-dir}\textsc{.pl} sack\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}=and 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `As the donkey said, he (the man) put straw into his (i.e. donkey’s) sack.'
\z 
 
\ea \label{HB.56}
\textit{hēɫeřuenīšā kerde pey sāhēbū hereyū pey herbenew hereyū} \\ 
\gll hēɫeřuenī=šā kerd-e pey sāhēb=ū her-e-ī=ū pey herbene=ū her-e-ī=ū \\ 
 fried\_egg\textsc{.dir}\textsc{\textsc{.f}}\textsc{=3pl} do\textsc{.pst}-f\textsc{.3sg} for owner\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} for donkey breeder\textsc{.m}=and donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \\ 
\glt `They (the host’s family) cooked fried eggs for the owner of the donkey, for the donkey breeder.'
\z 
 
\ea \label{HB.58}
\textit{girđ kārēšā kerdū} \\ 
\gll girđ kār-ē=šā kerd=w \\ 
 all of task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3pl} do\textsc{.pst}=and \\ 
\glt `They carried out all their tasks.'
\z 
 
\ea \label{HB.60}
\textit{āmā āsāwekew Šeškī, kābrāke} \\ 
\gll āmā āsāw-eke=w Šešk-ī kābrā-eke \\ 
 come\textsc{.pst}\textsc{.3sg} mill\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} man\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `The fellow came to the Shashk mill.'
\z 
 
\ea \label{HB.61}
\textit{duē ferdē hārdīš dē qewū herekeywe peyšū} \\ 
\gll duē ferdē hārdī=š dā-ē qew=ū her-ekey=we pey=š=w \\ 
 two sack\textsc{.dir}\textsc{.pl} flour\textsc{.dir}\textsc{\textsc{.f}}\textsc{.pl}\textsc{=3sg} give\textsc{.pst}\textsc{-dir}\textsc{.pl} on\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} for\textsc{=3sg}=and \\ 
\glt `He put two sacks of flour on the back of the donkey'
\z 
 
\ea \label{HB.68}
\textit{milūwe, min činnē sāɫēn īnānē xizmetū šēxīne} \\ 
\gll mi-l-w=we min činnē sāɫe-ē=n īnā-ānē xizmet=ū šēx-ī=ne \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{compl}} 1sg some\textsc{.pl} year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-1sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I will go back! It’s been several years that I am at the service of the sheikh;'
\z 
 
\ea \label{HB.76}
\textit{māčā, qurbān wiɫā fiɫāne kes toryānū ānā} \\ 
\gll m-āč-ā qurbān wiɫā fiɫān=e kes toryā=n=ū āne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} by\_God such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m} get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \\ 
\glt `They said, ‘Sir, indeed, the fellow has got offended; there he is’'
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{HB.84}
\textit{ēnnew herē, īse her be herī wēš qisē kero min hīčim nezānān!’} \\ 
\gll ēnne=w her-ē īse her be her-ī wē=š qisē-ē ∅-ker-o min hīč=im ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} now both by donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As much as a donkey! now the donkey could speak even though it was a donkey, but I didn’t know a thing.'
\z 
 
\ea \label{HB.91}
\textit{bizāne neferē ter hen be zūwānū herī hāɫīš bo bejge to!} \\ 
\gll bi-zān-e nefer-ē ter hen-∅ be zūwān=ū her-ī hāɫī=š ∅-b-o bejge to \\ 
 \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} person\textsc{.m}\textsc{-dir}\textsc{.pl} other exist\textsc{-3sg}\textsc{.m} by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} understood\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `see if there is a person, apart from you, who can understand the language of donkeys!'
\z 
 
\ea \label{PM.5}
\textit{fermāndew merekuřekey, reʕīsekešā māčo, ēme bē ā šexsī metāwmē, e čē menmēwe zerer deymē čī mentēqey’} \\ 
\gll fermānde=ū merekuř-ekey reʕīs-eke=šā m-āč-o ēme bē ā šexs-ī me-tāw-mē e=čē ∅-men-mē=we zerer ∅-de-ymē č=ī mentēqe-ī \\ 
 leader\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} leader\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} in=here \textsc{sbjv-}stay\textsc{.prs}\textsc{-1pl}\textsc{=compl} harm\textsc{.m} \textsc{sbjv-}give\textsc{.prs}-\textsc{1pl} in=\textsc{dem.prox} region\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The head of the grasshoppers said, ‘Without [permission from] this person we cannot stay here and do harm to this region.'
\z 
 
\ea \label{PM.9}
\textit{ījāze bizāne āđ ījāze miđo ēme e čē nehār kermē yām ne} \\ 
\gll ījāze bi-zān-e āđ ījāze mi-đe-o ēme e=čē nehār ∅-ker-mē yām ne \\ 
 permission \textsc{imp-}know\textsc{.prs}-\textsc{2sg} 3sg\textsc{.dir}\textsc{.m} permission \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \textsc{1pl} in=here lunch\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} or no \\ 
\glt `See if he lets us stay here for lunch or not'
\z 
 
\ea \label{PM.10}
\textit{nimāyindeke mē xizmetū šēxī} \\ 
\gll nimāyinde-eke m-ē xizmet=ū šēx-ī \\ 
 representative\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The messenger came into the Sheikh’s presence'
\z 
 
\ea \label{PM.17}
\textit{ānā ēme wēmā mehelēmā hen tūl} \\ 
\gll āne=ā ēme wē=mā mehele-ē=mā hen-∅ tūl \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{1pl} \textsc{reflx}\textsc{=1pl} neighbourhood\textsc{.m}\textsc{-indf}\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} long \\ 
\glt `There, I, myself, own a big neighbourhood.’\footnote{Lit. We, ourselves, have a big neighbourhood.}'
\z 
 
\ea \label{PM.23}
\textit{šewē merekuř nīšo milo mehelekew šēxī ʕelādīnī} \\ 
\gll šewe-ē merekuř ∅-nīš-o mi-l-o mehele-eke=ū šēx-ī ʕelādīn-ī \\ 
 night.f\textsc{-obl}.f grasshopper \textsc{ind-}sit\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} neighbourhood\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `The grasshoppers stayed\footnote{Lit. sat.} [there]. They attacked\footnote{Lit. sit on.} Sheikh Aladin’s neighbourhood.'
\z 
 
\ea \label{PM.27}
\textit{her šewēw meheleke fire bīyenū kem wārdenšū lūān} \\ 
\gll her šewe-ē=ū mehele-eke fire bīye=n=ū kem wārde=n=š=ū lūā=n \\ 
 just night.f\textsc{-obl}.f=and neighbourhood\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} a\_lot be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and little eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}=and go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Whatever had been there in the neighbourhood, whether large or small, they (i.e., the grasshoppers) had eaten during the night [and] gone away.'
\z 
 
\ea \label{PM.29}
\textit{māčo qurbān īne čēšit kerdū?} \\ 
\gll m-āč-o qurbān īne čēš=it kerd=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=\textsc{2sg}} do\textsc{.pst}=and \\ 
\glt `They said, ‘Sir, what have you done?'
\z 
 
\ea \label{PM.30}
\textit{īne bē ījāzew to īne nekiryān} \\ 
\gll īne b-ē ījāze=ū to īne ne-kir-yā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} permission\textsc{\textsc{=ez.gen}} \textsc{2sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{neg-}do.pass-past.pass\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This couldn’t have been done without your permission.'
\z 
 
\ea \label{PM.32}
\textit{īne čī e pēse?} \\ 
\gll īne čī e=pēse \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} why in=like \\ 
\glt `Why is this situation like this?'
\z 
 
\ea \label{PM.34}
\textit{nāřihet nebī āne meʕmūre bē terefū xwāywe āmābē} \\ 
\gll nāřihet ne-b-ī āne meʕmūre b-ē teref=ū xwā-ī=we āmā b-ē \\ 
 sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} agent be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Don’t worry! It was an agent [who] was sent  by God.'
\z 
 
\ea \label{PM.35}
\textit{eger yerē řoē šārezūrne menēwe geɫāyē sewz šārezūrne nāzē} \\ 
\gll eger yerē řo-ē šārezūr=ne men-ē=we geɫā-ē sewz šārezūr=ne ne-āz-ē \\ 
 if three day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{pn}\textsc{=\textsc{post}} remain\textsc{.prs}\textsc{-cond}\textsc{.3pl}\textsc{=compl} leaf\textsc{.m}\textsc{-indf} green \textsc{pn}\textsc{=\textsc{post}} \textsc{neg-}let\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `If the grasshoppers had stayed in Sharazour for three days, they wouldn’t have left behind a single green leaf in Sharazour.'
\z 
 
\ea \label{PM.36}
\textit{min pewkī ījāzem dān hinekew ēme wero, bilo zerer je kesī negino} \\ 
\gll min pewkī ījāze=m dā=n hin-eke=ū ēme ∅-wer-o bi-l-o zerer je kes-ī ne-gin-o \\ 
 \textsc{1sg} that\_is\_why permission\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} ez.pron\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{sbjv-}eat\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} harm\textsc{.m} from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `I let them [eat our crop] so that they would eat our crop and go away, [but] not do harm to others.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.44}
\textit{werzīyereke māčo, weɫāhī šēxī ʕelādīne řāsiš fermāwān} \\ 
\gll werzīyer-eke m-āč-o weɫāhī šēx-ī ʕelādīn=e řās=iš fermāwā=n \\ 
 farm\_labourer\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{=dem} truth\textsc{=3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The farm-labourer said; ‘Indeed, Sheikh Aladin was right.'
\z 
 
\ea \label{PM.47}
\textit{ēđ čowe mē peym māro} \\ 
\gll ēđ č=o=we m-ē pey=m m-ār-o \\ 
 3sg.prox\textsc{.dir} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From this direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.51}
\textit{īne heɫbet xwāy meylišā īne} \\ 
\gll īne heɫbet xwā-ī meyl=š=ā īne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} obviously God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \\ 
\glt `This is indeed God’s will'
\z 
 
\ea \label{PM.52}
\textit{īne heɫbet xwāy meylšā} \\ 
\gll īne heɫbet xwā-ī meyl=š=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} obviously God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is indeed God’s will’'
\z 
 
\ea \label{DG.3}
\textit{ā wextīekey yerē kuřēš bīyēnē} \\ 
\gll ā wext-ī-ekey yerē kuř-ē=š bīyē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `[Shah Aladin] had three sons.'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.11}
\textit{řoē je řoā yo mē xizmetiš, xizmetū šā ʕelādīnī} \\ 
\gll řo-ē je řo-ā yo m-ē xizmet=iš xizmet=ū šā ʕelādīn-ī \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{=3sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Once\footnote{Lit. One day of days.}, a man\footnote{Lit. one.}  came into his presence, to Shah Aladin'
\z 
 
\ea \label{DG.21}
\textit{ā wextīyekey tā yerē sāɫē e čā kešo āđ guzerān kero lā birākeyšo} \\ 
\gll ā wext-ī-ekey tā yerē sāɫe-ē e=čā keš=o āđ guzerān ∅-ker-o lā birā-ekey=š=o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} until three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} in=there mountain\textsc{.m}\textsc{=\textsc{post}} 3sg\textsc{.dir}\textsc{.m} subsistence\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} at\_the\_place\_of brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `For three years, he stayed with his brothers in those mountains '
\z 
 
\ea \label{DG.24}
\textit{řoē je řoā māčā pīyāē hen nāmēš šā ʕelādīnārēū} \\ 
\gll řo-ē je řo-ā m-āč-ā pīyā-ē hen-∅ nāmē=š šā ʕelādīn=ā=rē=ū \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} man\textsc{.m}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} name\textsc{\textsc{.f}}\textsc{=3sg} king\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `Once, people said, ‘There is a man called Aladin. '
\z 
 
\ea \label{DG.28}
\textit{gā hen lāɫēwe xwāy xwā derdeš deʕfe kero} \\ 
\gll gāhen ∅-lāɫye-ē=we xwā-ī xwā derde=š deʕfe ∅-ker-o \\ 
 maybe \textsc{sbjv-}beseech\textsc{-dir}\textsc{.pl}\textsc{=\textsc{compl}} God\textsc{.m}\textsc{-obl}\textsc{.m} God\textsc{.m} illness\textsc{\textsc{.f}}\textsc{=3sg} exclusion\textsc{\textsc{.f}} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `and beg to God; may God cure him.’'
\z 
 
\ea \label{DG.29}
\textit{šewē, de, de řoē bo, yā pāngze řoē bo yā wīs řoē bo, je qeredāxo māroš tā heɫebje} \\ 
\gll šewe-ē de de řo-ē ∅-b-o yā pāngze řo-ē ∅-b-o yā wīs řo-ē ∅-b-o je qeredāx=o m-ār-o=š tā heɫebje \\ 
 night\textsc{\textsc{.f}}\textsc{-indf} ten ten day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} or fifteeen day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} or twenty day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} from \textsc{pn}\textsc{=\textsc{post}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} until \textsc{pn} \\ 
\glt `Whether it was one night, ten days, fifteen days, or twenty days, he took his brother from Qaradax to Halabja.'
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DG.36}
\textit{be čāwel ħeywāndārē bīyēnmēū} \\ 
\gll be čāwel ħeywāndār-ē bīyē=nmē=ū \\ 
 by in\_the\_past animal\_husbandry\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}=and \\ 
\glt `In the past, our job was animal husbandry;'
\z 
 
\ea \label{DG.37}
\textit{bāxdārē bīyēnmēū} \\ 
\gll bāxdār-ē bīyē=nmē=ū \\ 
 gardener\textsc{.m}\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}=and \\ 
\glt `we were gardeners.'
\z 
 
\ea \label{DG.38}
\textit{ā wextīyekey, ētir, īse, min yerē sāɫē ī birāme tūšū ī derdēe āmānū} \\ 
\gll ā wext-ī-ekey ētir īse min yerē sāɫe-ē ī birā=m=e tūš=ū ī derde-ē=e āmā=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} encounter\textsc{=ez}\textsc{.gen} \textsc{dem.prox} illness\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=dem} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Now, as for me, it’s been three years that my brother is suffering from this illness. '
\z 
 
\ea \label{DG.43}
\textit{ī birāme derdeš, derdēš īnā po derdū gulīš māčā dāđē!} \\ 
\gll ī birā=m=e derde=š derde-ē=š īnā-∅ p==oderde=ū gulī=š m-āč-ā dāđ-ē \\ 
 \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} illness\textsc{\textsc{.f}}\textsc{=3sg} illness\textsc{\textsc{.f}}\textsc{-indf}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at=\textsc{=\textsc{post}}illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} yell\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `This brother of mine is ill; it is called leprosy; Please help!'
\z 
 
\ea \label{DG.44}
\textit{īđīč fermāwo, lue čā nīšere,} \\ 
\gll īđ=īč ∅-fermāw-o ∅-lu-e čā ∅-nīš-e=re \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} there \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \\ 
\glt `Shah Aladin  said, ‘Go [and] sit there;'
\z 
 
\ea \label{DG.51}
\textit{ī pose āne gino, pos māro nāwe} \\ 
\gll ī pos=e āne ∅-gin-o pos m-ār-o nāwe \\ 
 \textsc{dem.prox} skin\textsc{.m}\textsc{=dem} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} ind\textsc{-f}all\textsc{.prs}\textsc{-3sg} skin\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} inside \\ 
\glt `his skin will fall off, and [a new] skin will replace it'
\z 
 
\ea \label{DG.55}
\textit{ā wextīyekey īne vīyero, qerār bo} \\ 
\gll ā wext-ī-ekey īne vīyer-o qerār ∅-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} pass\textsc{.prs}\textsc{-3sg} arrangement \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then, this was settled. The deal was ...'
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.66}
\textit{ē.. dewāyekeš sāwe pore dā ēme kečē nebīmē} \\ 
\gll ē.. dewāī-eke=š ∅-sāw-e pore dā ēme keč-ē ne-b-īmē \\ 
 eh medication\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} at \textsc{hort} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs}-\textsc{1pl} \\ 
\glt `He shall rub the medicine on him; he shall not make us crooked.'
\z 
 
\ea \label{DG.67}
\textit{šā ʕelādīnīč fermāwo birākem wextē īnē zānā min gunākē māy mē milimre} \\ 
\gll šā ʕelādīn-ī=č ∅-fermāw-o birā-eke=m wext-ē īnē ∅-zān-ā min gunā-ekē m-āy m-ē mil=im=re \\ 
 king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} time\textsc{.m}\textsc{-indf} \textsc{dem.prox}\textsc{.3pl} \textsc{ind-}know\textsc{.prs}\textsc{-3pl} \textsc{1sg} sin\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}come\textsc{.pst}\textsc{.adv} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} shoulder\textsc{.m}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `Shah Aladin said, ‘Brother!—as they (i.e. the tortoises) know [about the medicine] I won’t shoulder the burden of injuring them.'
\z 
 
\ea \label{DG.69}
\textit{milowe māngēš pene mešo mēwe} \\ 
\gll mi-l-o=we mānge-ē=š pene me-š-o m-ē=we \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} month\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=3sg} to \textsc{neg-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `He went back. A month passed [and] he came back.'
\z 
 
\ea \label{DG.70}
\textit{kuře wextē miđyo ānā čā šāzāyēwe} \\ 
\gll kuř-e wext-ē mi-đye-o āne=ā čā šāzā-ēwe \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.dir} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} there prince\textsc{.m}\textsc{-indf} \\ 
\glt `The [sick] boy has become like a prince.'
\z 
 
\ea \label{DG.73}
\textit{īnē dāstānū ī šēxāū ēme bīyen gewregerīšā} \\ 
\gll īnē dāstāne=ū ī šēx-ā=ū ēme bīye=n gewregerī=šā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{\textsc{.f}}\textsc{.3sg} story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} sheikh\textsc{.m}\textsc{-obl}\textsc{.pl}=\textsc{ez.gen} \textsc{1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} greatness\textsc{=3pl} \\ 
\glt `This is the story of our Sheikhs and their greatness.'
\z 
 

\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.8}
\textit{penjā šes neferē šūnīšo bīyēnēw} \\ 
\gll penjā šes nefer-ē šūnī=š=o bīyē=nē=ū \\ 
 fifty sixty person\textsc{.m}\textsc{-dir}\textsc{.pl} track\textsc{\textsc{.f}}\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `Some fifty, sixty persons backed him.'
\z 
 
\ea \label{DP.11}
\textit{ħewt sāɫē pāwe bīyen} \\ 
\gll ħewt sāɫe-ē pāwe bīye=n \\ 
 seven year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} there be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was up there for seven years.'
\z 
 
\ea \label{DP.15}
\textit{āđ pādšāš bīyen} \\ 
\gll āđ pādšā=š bīye=n \\ 
 3sg\textsc{.dir}\textsc{.m} king\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was the king.'
\z 
 
\ea \label{DP.16}
\textit{tā pādšā nebīyen āđ pādšāš bīyen} \\ 
\gll tā pādšā ne-bīye=n āđ pādšā=š bīye=n \\ 
 until king\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{.m} king\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was its king until there were no kings.'
\z 
 
\ea \label{DP.17}
\textit{ā wextīyekey īđīč āmān dāwāw pādšāgerīš kerdēnew} \\ 
\gll ā wext-ī-ekey īđ=yč āmā=n dāwā=w pādšāgerī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} kingship\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `He (i.e., Jamsher Shah) came [and] claimed kingship.'
\z 
 
\ea \label{DP.31}
\textit{be gunjēšā ārdēne} \\ 
\gll be gunj-ē=šā ārdē=ne \\ 
 by earthenware\_ pipe\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `They (i.e., Jamsher Shah’s’s men) carried [Jamsher Shah’s] water supply through earthenware pipes.'
\z 
 
\ea \label{DP.32}
\textit{āhekū gunjēw ārdēnešā ew dīmū qeɫē} \\ 
\gll āhek=ū gunj-ē=ū ārdē=ne=šā ew dīm=ū qeɫā-ē \\ 
 lime\textsc{.m}\textsc{=ez}\textsc{.gen} earthenware\_ pipe\textsc{.m}\textsc{-dir}\textsc{.pl}=and bring\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}\textsc{=3pl} \textsc{dem.dist} side\textsc{.m}\textsc{=ez}\textsc{.gen} castle\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \\ 
\glt `They took the water to the other side of the citadel.'
\z 
 
\ea \label{DP.34}
\textit{āwekē biřo milonū wer werwe māro be nīrūekeyš de koɫē penj koɫē gi řoē} \\ 
\gll āwī-ekē ∅-biř-o mi-l-on=w wer werwe mi-ār-o be nīrū-ekey=š de koɫ-ē penj koɫ-ē gi řo-ē \\ 
 water\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} \textsc{ind-}cut\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}=and out snow\textsc{\textsc{.f}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} by force\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} ten load/shoulder\textsc{.m}\textsc{-dir}\textsc{.pl} five load/shoulder\textsc{.m}\textsc{-dir}\textsc{.pl} each day\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `He cut off the water supply. He (i.e., Jamsher Shah) fetched snow, five or ten loads daily, using his men.'
\z 
 
\ea \label{DP.40}
\textit{ewīč nimē māčo, nexeyr nimewnew mitmānim nīyā penet.} \\ 
\gll ew=yč ni-mi-ē mi-āč-o nexeyr ni-mi-e-wne=ū mitmān=im nīye=ā pene=t \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-1sg}=and trust\textsc{.m}\textsc{=1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} to\textsc{=\textsc{2sg}} \\ 
\glt ` Sibhan Agha\footnote{Lit. He.} did not go [to Jamsher Shah], saying, ‘No, I won’t come [to you]; I don’t trust you.’'
\z 
 
\ea \label{DP.41}
\textit{desgīrim kero āđ nūrīš firen} \\ 
\gll desgīr=im ∅-ker-o āđ nūrī=š fire=n \\ 
 hand.cufffed\textsc{=1sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m} force\textsc{.m}\textsc{=3sg} many\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is going to arrest me. He has many men.'
\z 
 
\ea \label{DP.44}
\textit{wāt, bē hem mēmān bo nīmeřo hem bā qisē kermē} \\ 
\gll wāt b-ē hem mēmān ∅-b-o nīmeřo hem b-ā qisē-ē ∅-ker-mē \\ 
 say\textsc{.pst} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} both guest\textsc{.m} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} noon both be\textsc{.prs}\textsc{-3pl} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `He said, ‘Come and be our guest for lunch, and we will discuss the matter.’'
\z 
 
\ea \label{DP.45}
\textit{āđ kārū wēšū min hinū wēm} \\ 
\gll āđ kār=ū wē=š=ū min hin=ū wē=m \\ 
 3sg\textsc{.dir}\textsc{.m} task\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3sg}\textsc{=and} \textsc{1sg} \textsc{z.pron}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} \\ 
\glt `Then, he goes his way, and I go mine.'
\z 
\ea \label{DP.46}
\textit{ā wextīyekey wextē mēnē ā wextī herčē qisē kerā, herčē behsē kerā herčē hesābē kerā, qēxāw hewrāmānī herčē bā ā wextī isɫāhšā mebo} \\ 
\gll ā wext-ī-ekey wext-ē mi-ē-nē ā wext-ī herčē qisē-ē ∅-ker-ā herčē behse-ē ∅-ker-ā herčē hesāb-ē ∅-ker-ā qēxā=w hewrāmān-ī herčē ∅-b-ā ā wext-ī isɫāh=šā me-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} however\_much talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much discussion\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much account\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} chieftain\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} however\_much \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} amendment\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then Sibhan Agha and his men came [to Jamsher Shah]. No matter how much the headmen of Hawraman talked [to both of them], discussed the issue, and tried to mediate, they (i.e., Jamsher Shah and Sibhan Agha) disagreed.'
\z 
 
\ea \label{DP.49}
\textit{ī gīyānū qefesew sīney minne īnen āđe lūā} \\ 
\gll ī gīyān=ū qefese=ū sīne=y min=ne īne=n āđe lūā \\ 
 \textsc{dem.prox} soul\textsc{.m}\textsc{=ez}\textsc{.gen} cage\textsc{.m}=and chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{\textsc{.f}} go\textsc{.pst} \\ 
\glt ` The soul in my rib cage [that I swore an oath on] was this [sparrow]. It [just] flew away.'
\z 
 
\ea \label{DP.50}
\textit{ānen mirīčēɫēm wer dān, to xeyrū min siɫāmet} \\ 
\gll āne=n mirīčēɫe-ē=m wer dā=n to xeyr=ū min siɫāmet \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} small\_bird\textsc{.m}\textsc{-indf}\textsc{=1sg} out give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `’I have let the bird go away. Now, you follow your life, and I will follow mine!\footnote{Lit. Well-being for you and health for me.}’'
\z 
 
\ea \label{DP.53}
\textit{āđ bowe pādšā tā zemānēwe nāđir šā mē serū dewrī} \\ 
\gll āđ ∅-b-o=we pādšā tā zemān-ē-ewe nāđir šā m-ē ser=ū dewr-ī \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} king\textsc{.m} until time\textsc{.m}\textsc{-indf}\textsc{-indf} \textsc{pn} king\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} on\textsc{=ez}\textsc{.gen} period\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (i.e., Jamsher Shah) became the king up until the time of Nader Shah.'
\z 
 
\ea \label{DP.54}
\textit{nāđir šā mē serū dewrī her wī berkenārē kero} \\ 
\gll nāđir šā m-ē ser=ū dewr-ī her wē-ī berkenār-ē ∅-ker-o \\ 
 \textsc{pn} king\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} on\textsc{=ez}\textsc{.gen} period\textsc{.m}\textsc{-obl}\textsc{.m} both two\textsc{-obl}\textsc{.m} ousted\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Nader Shah came to power [and] expelled both [Jamsher Shah and Sibhan Agha]'
\z 
 
\ea \label{ZP.7}
\textit{āne māmoš bīyen} \\ 
\gll āne māmo=š bīye=n \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} paternal\_uncle\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` He was his uncle.'
\z 
 
\ea \label{ZP.19}
\textit{yānē yāwān pene tāwū hem šerīʕetū hem terīqet tāwū girđ kārēwe kero} \\ 
\gll yānē yāwā=n pene ∅-tāw-ū hem šerīʕet=ū hem terīqet ∅-tāw-ū girđ kār-ē=we ∅-ker-o \\ 
 that\_means arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} both sharia\_law\textsc{\textsc{=ez.gen}} both denomination \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} every task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{post}} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This meant that he had become experienced (in his field). He could practise Sharia law, lead an order (of dervishes), and carry out all tasks [appointed to him]'
\z 
 
\ea \label{ZP.25}
\textit{čā dimāy, kināčēw pādšāw misrī neweše gino} \\ 
\gll čā dimā-ī kināčē=ū pādšā=w misr-ī neweš-e ∅-gin-o \\ 
 there afterwards\textsc{-obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} ill\textsc{-f} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `After a while, the king of Egypt’s daughter fell sick.'
\z 
 
\ea \label{ZP.33}
\textit{ī kināčēw to dermānū derdīš īnā lā ī pīyāywe’} \\ 
\gll ī kināčē=ū to dermān=ū derd-ī=š īnā-∅ lā ī pīyā-ī=we \\ 
 dem.prox daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} \textsc{2sg} treatment\textsc{.m}\textsc{\textsc{=ez.gen}} illness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \\ 
\glt `‘This daughter of yours, the treatment for her illness lies with this man (i.e., Pir Shaliyar)'
\z 
 
\ea \label{ZP.35}
\textit{ētir ewīč wēš pādšā bo} \\ 
\gll ētir ew=yč wē=š pādšā ∅-b-o \\ 
 \textsc{disc.ptcl} 3sg\textsc{.dir}\textsc{.m} \textsc{reflx}\textsc{=3sg} king\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well, he was a king.'
\z 
 
\ea \label{ZP.36}
\textit{īne māčmēš pene šālīyāre sīyāw lālow kināčekēn} \\ 
\gll īne m-āč-mē=š pene šālīyār=e sīyāw lālo=ū kināčē-ekē=n \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn}\textsc{=dem} black maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This [person], whom we call Shaliyar Siya, is the maternal uncle of the [king’s] daughter.'
\z 
 
\ea \label{ZP.43}
\textit{jā pādšākey wāten be lāloe ew lāloe kināčekē,} \\ 
\gll jā pādšā-eke-ī wāte=n be lālo-e ew lālo=e kināčē-ekē \\ 
 afterwards king\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to maternal\_uncle\textsc{.m}\textsc{-def} \textsc{dem.dist} maternal\_uncle\textsc{.m}\textsc{=dem} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `Oh, the king had said to the uncle, to [his] daughter’s uncle,'
\z 
 
\ea \label{ZP.44}
\textit{to eger ī kināčē berdet xizmetū ī pīyāy} \\ 
\gll to eger ī kināčē berd-e=t xizmet=ū ī pīyā-ī \\ 
 \textsc{2sg} if dem.prox daughter\textsc{.dir}\textsc{.f} take\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=\textsc{2sg}} service\textsc{.m}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `‘When you have taken this girl into the presence of this man'
\z 
 
\ea \label{ZP.45}
\textit{ī pīyāy tāwāš ī kināčēše kerde be qerārē wešeš kerdewe} \\ 
\gll ī pīyā-ī tāwā=š ī kināčē=š=e kerd-e be qerār-ē weš-e=š kerd-e=we \\ 
 dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} can\textsc{.pst}\textsc{=3sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} do\textsc{.pst}\textsc{-f}\textsc{.3sg} to arrangement\textsc{-indf} well\textsc{-f}\textsc{=3sg} do\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=compl} \\ 
\glt `[and if] the man has been able to cure the girl'
\z 
 
\ea \label{ZP.51}
\textit{ētir kināče keřū lāɫe bo māčmēš pene tewenū sirewē} \\ 
\gll ētir kināčē=e keře=ū lāɫ-e ∅-b-o m-āč-mē=š pene tewenī=ū sirewē \\ 
 \textsc{disc.ptcl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} deaf\textsc{.f}=and mute\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} \\ 
\glt `Since the girl was deaf-mute, we call it the healing stone.'
\z 
 
\ea \label{ZP.54}
\textit{čāweɫ tā jāwe nelābē pāne pēwīyē tewenekē tewenē zile bē} \\ 
\gll čāweɫ tā jāwe ne-lā b-ē p=āne pēwīye-ē tewenī-ekē tewenī-ē zil-e b-ē \\ 
 in\_the\_past until road \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-indf} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\_visible\textsc{.prs}\textsc{-pstc}\textsc{.3sg} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} stone\textsc{.f}\textsc{-indf} big\textsc{-f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `In the past, when there was no road, the stone was visible from afar. It was a big stone.'
\z 
 
\ea \label{ZP.58}
\textit{lāloekeš māčo, erē sirewā, erē weɫā sirewā’} \\ 
\gll lālo-eke=š m-āč-o erē sirewā erē weɫā sirewā \\ 
 maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes be cured\textsc{.pst}\textsc{.3sg} yes indeed be cured\textsc{.pst}\textsc{.3sg} \\ 
\glt `Her uncle said, ‘Yes, she is healed; indeed, she is healed.'
\z 
 
\ea \label{ZP.60}
\textit{e čā dengē kero māčo, erē erē weɫā sirewā’} \\ 
\gll e=čā deng-ē ∅-ker-o m-āč-o erē erē weɫā sirewā \\ 
 in=there noise\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes yes indeed be cured\textsc{.pst}\textsc{.3sg} \\ 
\glt `He shouted there [and] said, ‘Indeed, she is healed.’'
\z 
 
\ea \label{ZP.65}
\textit{pīr šelīyār ā vextī yānekew yānew tāteyšne bo, pū ser degāwe} \\ 
\gll pīr šelīyār ā vext-ī yāne-eke=ū yāne=ū tāte-ī=š=ne ∅-b-o p=ū ser degā=we \\ 
 \textsc{pn} \textsc{pn} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} house\textsc{.m}\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at=\textsc{dem.dist}\textsc{.3sg} on village\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar, well, he was [living] in his father’s house, [which was located] at the top of the village'
\z 
 
\ea \label{ZP.67}
\textit{tewejo kero wextē miđyo ī ženīū ī pīyā mēnē řāře duē dēwē īnē dewršāne} \\ 
\gll tewejo ∅-ker-o wext-ē mi-đye-o ī ženī=ū ī pīyā m-ē-nē řā=ře duē dēw-ē īnē dewr=šā=ne \\ 
 caution \textsc{ind-}do\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} dem.prox woman\textsc{.f}=and dem.prox man\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} road\textsc{.f}\textsc{=\textsc{post}} two demon\textsc{.m}\textsc{-dir}\textsc{.pl} dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} around\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `He noticed that when these two persons were on their way, they were surrounded by two demons.'
\z 
 
\ea \label{ZP.72}
\textit{ewšā her mēš čene bēheyā bo mēš čene} \\ 
\gll ew=šā her m-ē=š čene bē- heyā ∅-b-o m-ē=š čene \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=3pl} emph \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=3sg} withwithout decency \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=3sg} with \\ 
\glt `The other one was still coming with her; he was shameless; he kept coming with her.'
\z 
 
\ea \label{ZP.74}
\textit{ā degākē ehā čew deso bīyēne duē kutē māčmēš pene tenūrū dēwā} \\ 
\gll ā degā-ekē ehā č=ew des=o bīyē=ne duē kut-ē m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 \textsc{dem.dist} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{intj} in=\textsc{dem.dist} hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} two half\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Do you see that the village is divided in half on the other side? We call that place Tanūrū dēwā.'
\z 
 
\ea \label{ZP.81}
\textit{čā dēweke čāgene fotīyo} \\ 
\gll čā dēw-eke čāgene fotīye-o \\ 
 there demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} there die\textsc{.prs}\textsc{-3sg} \\ 
\glt `The demon died there.'
\z 
 
\ea \label{ZP.89}
\textit{kināčēw pādšāw fiɫān henānew miniš kerdenā be wekēɫ} \\ 
\gll kināčē=ū pādšā=w fiɫān hen-ā=ne=ū min=iš kerde=nā be wekēɫ \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such place\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \textsc{1sg}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to ad\textsc{voc}ate\textsc{.m} \\ 
\glt `She is the daughter of … the king of such-and-such place has put me in charge.\footnote{Lit. made me advocate.}'
\z 
 
\ea \label{ZP.90}
\textit{mišo māreš biřū pey ī šexsīw ī šexse eger ī kināčēše weše kerde,} \\ 
\gll mišo māre=š ∅-biř-ū pey ī šexs-ī=ū ī šexs=e eger ī kināčē=š=e weš-e kerd-e \\ 
 aux marriage\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for dem.prox saint\textsc{-obl}\textsc{.m}=and dem.prox saint\textsc{=dem} if dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f} \\ 
\glt `I shall marry her to this person [on the condition that] this person has healed this girl.'
\z 
 
\ea \label{ZP.93}
\textit{kināčēw pādšāy mārā lā pīr šelīyārī} \\ 
\gll kināčē=ū pādšā-ī m-ār-ā lā pīr šelīyār-ī \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl} at\_the\_place\_of \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They brought the king’s daughter to Pir Shaliyar.'
\z 
 
\ea \label{ZP.94}
\textit{ā wextī ēme xeplezeřetīnē bēnē tā pāyīz} \\ 
\gll ā wext-ī ēme xeplezeřetīnē b-ēn-ē tā pāyīz \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{1pl} \textsc{pn}\textsc{.dir}\textsc{.pl} be\textsc{.prs}\textsc{-pstc}\textsc{-3pl} until autumn \\ 
\glt `Back then, there were farms of black corn\footnote{A type of corn. Also a type of bread made from corn flour.} until autumn.'
\z 
 
\ea \label{ZP.97}
\textit{ā wextī mārāš mē ʕerzū pīr šelīyārī kero lāloekeš} \\ 
\gll ā wext-ī m-ār-ā=š m-ē ʕerz=ū pīr šelīyār-ī ∅-ker-o lālo-eke=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \\ 
\glt `They (the people) brought her. They came (to Pir Shaliyar). The uncle said to Pir Shaliyar,'
\z 
 
\ea \label{ZP.98}
\textit{māčo, min īneyā jeryānim e čīnenū} \\ 
\gll m-āč-o min īne=ā jeryān=im e=čīne=n=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} story\textsc{.m}\textsc{=\textsc{1sg}} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He said, ‘I— look, my story is like this.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.102}
\textit{īne pādšāzāyēnew derdū min meweronew} \\ 
\gll īne pādšāzāyē=ne=ū derd=ū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}=and illness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She is no use to me.'
\z 
 
\ea \label{ZP.103}
\textit{min metāwū īne nigebānīye kerū} \\ 
\gll min me-tāw-ū īne nigebānī-e ∅-ker-ū \\ 
 \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} guardian\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt ` I cannot take care of her.'
\z 
 
\ea \label{ZP.104}
\textit{īne pādšāzāyēne čenū min meweronew} \\ 
\gll īne pādšāzāyē=ne čenū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}} with\textsc{.ez.gen} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She will not eat with me [under the same roof].'
\z 
 
\ea \label{ZP.107}
\textit{ewīč māčo, nexeyre lāloekeš min tāze pādšā kerdenā wekēɫū} \\ 
\gll ew=īč m-āč-o nexeyre lālo-eke=š min tāze pādšā kerde=nā wekēɫ=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{1sg} anyway king\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m}=and \\ 
\glt ` The uncle said, ‘No! I—the king has put me in charge.'
\z 
 
\ea \label{ZP.113}
\textit{xeɫk mē dewreš mido māčā, āne čēšā? čēšā āne?’} \\ 
\gll xeɫk m-ē dewre=š mi-de-o m-āč-ā āne čēš=ā čēš=ā āne \\ 
 people\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} rotation\textsc{=3sg} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=cop}\textsc{.3sg}\textsc{.m} what\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \\ 
\glt `People came, encircled him, and asked,\footnote{Lit. said.} ‘What is going on? What is going on?’'
\z 
 
\ea \label{ZP.114}
\textit{qisē kerā beynū wēšāne} \\ 
\gll qisē ∅-ker-ā beyn=ū wē=šā=ne \\ 
 talk\textsc{.dir}\textsc{.f} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} among\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `They talked among themselves.'
\z 
 
\ea \label{ZP.115}
\textit{māčā, dey īne ī pīyā ī ženīče ārdēne} \\ 
\gll m-āč-ā dey īne ī pīyā ī ženī=č=e ārdē=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox man\textsc{.m} dem.prox woman\textsc{.dir}f\textsc{=add}\textsc{=dem} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `[and] said, ‘Well, this man has brought this woman [here]'
\z 
 
\ea \label{ZP.117}
\textit{īđīč hīč minīč hīčim nīyārēw} \\ 
\gll īđ=īč hīč min=īč hīč=im nīye=ā=rē=ū \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} nothing \textsc{1sg}\textsc{=add} nothing\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `She [has] nothing, nor do I.'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.126}
\textit{ā wextī pīr šelīyār fermāwo, šime īne māčdē řāstā wāt} \\ 
\gll ā wext-ī pīr šelīyār ∅-fermāw-o šime īne m-āč-dē řās=tā wāt \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{2pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} \\ 
\glt `Then Pir Shaliyar said, ‘What you are saying is right.’\footnote{Lit. You said this; you told the truth.}'
\z 
 
\ea \label{ZP.128}
\textit{weɫāhī īne šime māčdēšū īnā vistāwe řā pey min čūn šime be xātirū xwāytā kerd hetā xwā dinyāš weše kerdēne ī merāsēme her bo her meno’} \\ 
\gll weɫāhī īne šime m-āč-dē=š=ū īnā-∅ vist=tā=we řā pey min čūn šime be xātir=ū xwā-ī=tā kerd hetā xwā dinyā=š weš-e kerdē=ne ī merāsēm=e her ∅-b-o her ∅-men-o \\ 
 indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} \textsc{ind-}say\textsc{.prs}-\textsc{2pl}\textsc{=3sg}=and deic\textsc{-3sg}\textsc{.m} throw\textsc{.pst}=\textsc{2pl}\textsc{=compl} road\textsc{.f} for \textsc{1sg} because \textsc{2pl} to sake\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}be\textsc{.prs}\textsc{-3sg} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `Indeed, what you are saying and the way you organised [this ceremony] for me, since you did it in good faith, so long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ZP.130}
\textit{hetā xwāy dinyāš weše kerdēne ī merāsēme kem mekero} \\ 
\gll hetā xwā-ī dinyā=š weš-e kerdē=ne ī merāsēm=e kem me-ker-o \\ 
 as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} small \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `So long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ŽP.6}
\textit{ānen īse īnā lāw hānew degē} \\ 
\gll āne=n īse īnā-∅ lā=ū hāne=ū degē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} water\_spring\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f} \\ 
\glt `That (his shrine) is on the other side of the village.'
\z 
 
\ea \label{ŽP.7}
\textit{āne tātešā} \\ 
\gll āne tāte=š=ā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} father\textsc{.m}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is his father.'
\z 
 
\ea \label{ŽP.19}
\textit{ānā ew dīmū qeɫāw kitāy} \\ 
\gll āne=ā ew dīm=ū qeɫā=ū kitā-ī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `—Look, it is [located] on the other side of the Kita castle'
\z 
 
\ea \label{ŽP.28}
\textit{ētir xo āđ meselen hetīm bo, lāw māmoyšā bo řūeš mebo wāčo, yā metāwū yā tāwū} \\ 
\gll ētir xo āđ meselen hetīm ∅-b-o lā=ū māmo-ī=šā ∅-b-o řūe=š me-b-o ∅-wāč-o yā me-tāw-ū yā ∅-tāw-ū \\ 
 \textsc{disc.ptcl} dis.ptc 3sg\textsc{.dir}\textsc{.m} for\_example orphan \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} face\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-3sg} either \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} either \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Well, he was an orphan. He was at his uncle’s. He was ashamed to say, ‘I cannot, or I can.’'
\z 
 
\ea \label{ŽP.35}
\textit{zengeneke zengenāre kero} \\ 
\gll zengene-eke zengenāre ∅-ker-o \\ 
 hoe\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} ploughing \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The hoe [started to] plough.'
\z 
 
\ea \label{ŽP.44}
\textit{ī yewēše šānēnē pī zemīnne} \\ 
\gll ī yewe-ē=š=e šānē=nē p=ī zemīn=ne \\ 
 dem.prox barley\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3sg}\textsc{=dem} scatter\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} at=dem.prox land\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Has he planted the seeds in the field?'
\z 
 
\ea \label{ŽP.45}
\textit{yā ī yewēše zāxēre kerdēnē?} \\ 
\gll yā ī yewe-ē=š=e zāxēre kerdē=nē \\ 
 either dem.prox barley\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3sg}\textsc{=dem} sotrage do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Or has he [instead] stored the seeds [for himself]'
\z 
 
\ea \label{ŽP.46}
\textit{āđ čenī tāwo kewčē yewē šāno zemīnnew wēregā bēwe} \\ 
\gll āđ čenī ∅-tāw-o kewč-ē yewe-ē ∅-šān-o zemīn=ne=ū wēregā b-ē=we \\ 
 3sg\textsc{.dir}\textsc{.m} how \textsc{ind-}can\textsc{.prs}\textsc{-3sg} six\_kilos\textsc{-indf} barley\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{sbjv-}scatter\textsc{.prs}\textsc{-3sg} land\textsc{.m}\textsc{=\textsc{post}}=and evening \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `How can he plant six kilos [of barley] in the field [and] return in the evening?'
\z 
 
\ea \label{ŽP.49}
\textit{īneš netāwān kero} \\ 
\gll īne=š ne-tāwā=n ∅-ker-o \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He cannot do these [tasks].’'
\z 
 
\ea \label{ŽP.51}
\textit{milo werāweriš miđyo e pāse min wāt— tomekeš šānā zemīnekeyneū} \\ 
\gll mi-l-o werāwer=iš mi-đye-o e=pāse min wāt tom-eke=š šānā zemīn-ekey=ne=ū \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} in\_front\_of\textsc{=3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} in=like \textsc{1sg} say\textsc{.pst} seed\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} scatter\textsc{.pst} land\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt ` He went to him [and] saw that —as I said— he (Pir Shaliyar) scattered the seeds in the field'
\z 
 
\ea \label{ŽP.52}
\textit{zengenekeš wistrew pāne zengene zemīnekey wēš} \\ 
\gll zengene-eke=š wist=re=we p=āne zengene zemīn-ekey wē=š \\ 
 hoe\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} throw\textsc{.pst}\textsc{=\textsc{povb}}\textsc{=compl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} hoe\textsc{.m} land\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} \\ 
\glt `He put the hoe there, in his field.'
\z 
 
\ea \label{ŽP.54}
\textit{bizekē wēš lūlē ženo} \\ 
\gll bize-ekē wē=š lūle-ē ∅-žen-o \\ 
 goat\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.pl} \textsc{reflx}\textsc{=3sg} flute\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{ind-}play\textsc{.prs}\textsc{-3sg} \\ 
\glt `The goats ... he (Pir Shaliyar) played the flute.'
\z 
 
\ea \label{ŽP.55}
\textit{bizekē jīyā leweřyāyšā hurpiřā} \\ 
\gll bize-ekē jīyā leweřyāy=šā hur-∅-piř-ā \\ 
 goat\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.pl} instead\_of graze.inf\textsc{=3pl} \textsc{pvb-}\textsc{ind-}jump\textsc{.prs}\textsc{-3pl} \\ 
\glt `The goats were dancing instead of grazing.'
\z 
 
\ea \label{ŽP.56}
\textit{bizekē semāy kerā hurpiřā} \\ 
\gll bize-ekē semā-ī ∅-ker-ā hur-∅-piř-ā \\ 
 goat\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.pl} dance\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{pvb-}\textsc{ind-}jump\textsc{.prs}\textsc{-3pl} \\ 
\glt `The goats were dancing.'
\z 
 
\ea \label{ŽP.57}
\textit{pīyāke mēwe ʕerzū heme yosoy kero māčo, weɫāhī qurbān min tūšū jeryānē e čīney āmānē} \\ 
\gll pīyā-eke m-ē=we ʕerz=ū heme yoso-ī ∅-ker-o m-āč-o weɫāhī qurbān min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sir.\textsc{voc} \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `The man returned [and said] to Hama Yoso, ‘Indeed, Sir! I have witnessed an event like this.'
\z 
 
\ea \label{ŽP.68}
\textit{wextē miđyā īnā āne je zemīnī hewāwe wuten} \\ 
\gll wext-ē mi-đye-ā īnā-∅ āne je zemīn-ī hewā=we wute=n \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} from earth\textsc{.m}\textsc{-obl}\textsc{.m} air\textsc{=\textsc{post}} sleep\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They saw that he (Pir Shaliyar) had levitated and slept aloft in the air at this height\footnote{Lit. this much.} (informant indicates the height of the bed with his hand).'
\z 
 
\ea \label{ŽP.69}
\textit{āne je zemīnī pēse wāčī xwāy ketē pey kīyāsen} \\ 
\gll āne je zemīn-ī pēse ∅-wāč-ī xwā-ī ket-ē pey kīyāse=n \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} from earth\textsc{.m}\textsc{-obl}\textsc{.m} as\_if \textsc{sbjv-}say\textsc{.prs-}\textsc{2sg} God\textsc{.m}\textsc{-obl}\textsc{.m} bed\textsc{-indf} to send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[He had levitated] this much from the ground, as if God had sent him a bed.'
\z 
 
\ea \label{ŽP.70}
\textit{āne īnā hewāwe} \\ 
\gll āne īnā-∅ hewā=we \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \\ 
\glt `He (his body) was aloft in the air [up to this height]'
\z 
 
\ea \label{ŽP.71}
\textit{pīr heme yoso wēš milo temāšeš kero miđyo āne īnā hewāwe zāno īne ētir kārī xwāyīn} \\ 
\gll pīr heme yoso wē=š mi-l-o temāše=š ∅-ker-o mi-đye-o āne īnā-∅ hewā=we ∅-zān-o īne ētir kār-ī xwāyī=n \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{reflx}\textsc{=3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} looking\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} task\textsc{.m}\textsc{-obl}\textsc{.m} divine\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Hama Yoso went [and] looked at him; he noticed that he (Pir Shaliyar) was aloft in the air [up until a certain height] and came to realise that that was God’s work.'
\z 
 
\ea \label{ŽP.74}
\textit{māčo min īnem dī be sey misefāywe} \\ 
\gll m-āč-o min īne=m dī-∅ be sey misefā-ī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} see\textsc{.pst}\textsc{-3sg}\textsc{.m} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘I witnessed this about Say Mustafa.’'
\z 
 
\ea \label{ŽP.82}
\textit{šūnū miqdārēre hezretū xosī fermāwo duē yerē īmtihānēš kero} \\ 
\gll šūn=ū miqdārē=re hezret=ū xos-ī ∅-fermāw-o duē yerē īmtihān-ē=š ∅-ker-o \\ 
 after\textsc{\textsc{=ez.gen}} a\_little\textsc{=\textsc{post}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} two three test\textsc{-dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `After a while, his Highness Khos said … He tested him two, three times.'
\z 
 
\ea \label{ŽP.83}
\textit{īmtihānēš kero, fermāwo, exī mistefā rejilun kāmilun yānē be ʕerebī wēšā pāseš fermāwān hezretū xosī} \\ 
\gll īmtihān-ē=š ∅-ker-o ∅-fermāw-o exī mistefā rejilun kāmilun yānē be ʕerebī wē=šā pāse=š fermāwā=n hezret=ū xos-ī \\ 
 test\textsc{-dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} my\_brother \textsc{pn} man complete that\_means by Arabic \textsc{reflx}\textsc{=3pl} this\textsc{=3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He tested him [and] said, ‘My brother Mustafa is a well-rounded man, which in their Arabic means, his Highness Khos said,'
\z 
 
\ea \label{ŽP.97}
\textit{řāne mēnē pāne yā guzerēwene mesīrēwene čenū hejījīyē ginā pēwere} \\ 
\gll řā=ne m-ē-nē p=āne yā guzer-ēwe=ne mesīr-ēwe=ne čenū hejījī-ē ∅-gin-ā pēwere \\ 
 road\textsc{.f}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} either path\textsc{-indf}\textsc{=\textsc{post}} path\textsc{-indf}\textsc{=\textsc{post}} with\textsc{.ez.gen} from\_Hajij\textsc{-indf} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl} together \\ 
\glt `On the road, on a path, he ran into a person from Hajij.'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.112}
\textit{hejījīyeke čāy dimā wālīore miđyo īnā zeřeš penen} \\ 
\gll hejījī-eke čā-ī dimā wālīye-o=re mi-đye-o īnā-∅ zeř-e=š pene=n \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} there\textsc{-obl}\textsc{.m} afterwards be\_astonished\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} deic\textsc{-3sg}\textsc{.m} money\textsc{.m}\textsc{-def}\textsc{=3sg} with\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The man from Hajij noticed later. He realised that he had money with him.'
\z 
 
\ea \label{ŽP.116}
\textit{mēwe ētir ā zemāne pīr šelīyār mēwe menteqe des kero, jārē āne minyā māčāš meřū āsinī} \\ 
\gll m-ē=we ētir ā zemān=e pīr šelīyār m-ē=we menteqe des=t ∅-ker-o jārē āne mi-nye-ā m-āč-ā=š meře=ū āsin-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} once \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) returned [to Hawraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ‘Iron cave.’'
\z 
 
\ea \label{ŽP.119}
\textit{čāne weɫēne luānre wā čilēš kēštēne} \\ 
\gll č=ā=ne weɫēne luā=n=re wā čilē=š kēštē=ne \\ 
 from=\textsc{dem.dist}\textsc{=\textsc{post}} in\_the\_begining go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} there forty\_days\_fasting\textsc{.dir}\textsc{.f}\textsc{=3sg} pull\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `From long ago, he would go there and fast for forty days.'
\z 
 
\ea \label{ŽP.125}
\textit{māčā kābrāyē pēse īse ā wextīč qisekerē her bīyēnē} \\ 
\gll m-āč-ā kābrā-ē pēse īse ā wext-ī=č qiseker-ē her bīyē=nē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} man\textsc{-indf} as\_if now \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} gossiper\textsc{-dir}\textsc{.pl} emph be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They said — Like nowadays, back then, there were also gossips.'
\z 
 
\ea \label{ŽP.132}
\textit{pīyākew hezretū xosī milo meřū āsinī tūšiš mē} \\ 
\gll pīyā-eke=ū hezret=ū xos-ī mi-l-o meře=ū āsin-ī tūš=iš m-ē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} encounter\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `His Highness Khos’s man went [and] ran into him (Pir Shaliyar) in the Iron cave.'
\z 
 
\ea \label{ŽP.134}
\textit{xo āđ mezāno hezretū xosī kīyāseniš} \\ 
\gll xo āđ me-zān-o hezret=ū xos-ī kīyāse=n=iš \\ 
 dis.ptc 3sg\textsc{.dir}\textsc{.m} \textsc{neg-}know\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \\ 
\glt `Well, he (Pir Shaliyar) did not realise that His Highness Khos had sent him (the man)'
\z 
 
\ea \label{ŽP.140}
\textit{čil dānē čilē kēšo} \\ 
\gll čil dāne-ē čilē ∅-kēš-o \\ 
 forty \textsc{clf}\textsc{-dir}\textsc{.pl} forty\_days\_fasting\textsc{.f} \textsc{ind-}pull\textsc{.prs}\textsc{-3sg} \\ 
\glt `He fasted for forty rounds of forty days of fasting.'
\z 
 
\ea \label{ŽP.143}
\textit{milowe yānew ētir pāne nāmēwe peyā kero} \\ 
\gll mi-l-o=we yāne=ū ētir p=āne nāmē-ewe peyā ∅-ker-o \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m}=and \textsc{disc.ptcl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} name\textsc{.f}\textsc{-indf} visible \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He went there (to Hawraman) and got a bit famous.\footnote{Lit. He found a name.}'
\z 
 
\ea \label{ŽP.148}
\textit{hīč dukturē, hečī dukturē, hečī kīšwerē beroš} \\ 
\gll hīč duktur-ē hečī duktur-ē hečī kīšwer-ē ∅-ber-o=š \\ 
 nothing physician\textsc{.m}\textsc{-indf} every physician\textsc{.m}\textsc{-dir}\textsc{.pl} every country\textsc{-dir}\textsc{.pl} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \\ 
\glt `No doctor, no doctor [could cure her no matter] how many countries he took her to.'
\z 
 
\ea \label{ŽP.149}
\textit{kināčekē feleje bo felejīyekeyš je diɫišne bo} \\ 
\gll kināčē-ekē felej-e ∅-b-o felejī-ekey=š je diɫ=iš=ne ∅-b-o \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} disabled\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} disability\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} from heart\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The girl was disabled. Her disability came from her tongue\footnote{Lit. heart.}'
\z 
 
\ea \label{ŽP.150}
\textit{qisē metāwo kero nitqiš bīnīyo} \\ 
\gll qisē-ē me-tāw-o ∅-ker-o nitq=iš bīnīye-o \\ 
 talk\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} speech\textsc{=3sg} close\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `She was not able to speak. She had lost the ability to speak.\footnote{Lit. her speech was closed.}'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.155}
\textit{wermekeš hāɫīš kerā puxte} \\ 
\gll werm-eke=š hāɫī=š ∅-ker-ā puxte \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} understood\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} properly \\ 
\glt `The dream made him understand well.'
\z 
 
\ea \label{ŽP.156}
\textit{māčā, ī kināčē to tāwo weše kerowe’} \\ 
\gll m-āč-ā ī kināčē to ∅-tāw-o weš-e ∅-ker-o=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `It was said [to him in the dream], ‘He can cure this daughter of yours.’'
\z 
 
\ea \label{ŽP.157}
\textit{pādšāyč māčo be lālo kināčekē be šālyāre sīyāwīš} \\ 
\gll pādšā=yč m-āč-o be lālo kināčē-ekē be šālyār=e sīyāw-ī=š \\ 
 king\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} by \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `The king said to [his] daughter’s uncle, to Shaliyar Syaw'
\z 
 
\ea \label{ŽP.159}
\textit{to bo lue! ī kināčē bere} \\ 
\gll to b-o ∅-lu-e ī kināčē ∅-ber-e \\ 
 \textsc{2sg} \textsc{imp-}come\textsc{.prs}\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `‘Come [here and] go! Take the girl.'
\z 
 
\ea \label{ŽP.161}
\textit{metāwū serū textīwe wezūrew bilū ginū gelū kināčēw be peyū luāy řāreū degā be degā} \\ 
\gll me-tāw-ū ser=ū text-ī=we wez-ū=re=ū bi-l-ū ∅-gin-ū gel=ū kināčē=ū be pe-ī=ū luāy řā=re=ū degā be degā \\ 
 \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} on\textsc{\textsc{=ez.gen}} throne\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} climb\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{povb}}=and \textsc{sbjv-}go\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-\textsc{1sg}} with\textsc{\textsc{=ez.gen}} daughter\textsc{.dir}\textsc{.pl}=and by foot\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} go.inf road\textsc{.f}\textsc{=\textsc{post}}=and village\textsc{.obl}\textsc{.f} by village\textsc{.obl}\textsc{.f} \\ 
\glt `I cannot leave the throne [and] go [and] accompany the girl, [while] walking on foot on the roads [that go] village by village.'
\z 
 
\ea \label{ŽP.162}
\textit{dēhāt be dēhāt ī kināčē bere} \\ 
\gll dēhāt be dēhāt ī kināčē ∅-ber-e \\ 
 village\textsc{.f} by village\textsc{.f} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `Take the girl village by village.'
\z 
 
\ea \label{ŽP.163}
\textit{eger to ī kināčēte berdeū ī pīyā her kes hen} \\ 
\gll eger to ī kināčē=t=e berd-e=ū ī pīyā her kes hen-∅ \\ 
 if \textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} take\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m} emph person\textsc{.m} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `If you take the girl, [no matter] who this man (the doctor) is,'
\z 
 
\ea \label{ŽP.164}
\textit{tāwāš ī duktire ī kināčēše weše kerdew} \\ 
\gll tāwā=š ī duktir=e ī kināčē=š=e weš-e kerd-e=ū \\ 
 can\textsc{.pst}\textsc{=3sg} dem.prox phy\textsc{=dem} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `If the doctor can cure the girl,'
\z 
 
\ea \label{ŽP.166}
\textit{lālo gino gelū kināčē māroš dey degā be degāw kēšwer be kēšwerū māroš silēne} \\ 
\gll lālo ∅-gin-o gel=ū kināčē m-ār-o=š dey degā be degā=ū kēšwer be kēšwer=ū m-ār-o=š silēne \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} with\textsc{\textsc{=ez.gen}} girl\textsc{.dir}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{disc.ptcl} village\textsc{.obl}\textsc{.f} by village\textsc{.f}=and country by country=and \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{pn} \\ 
\glt `The uncle set off on the road with the girl. He took her. He took her village by village and country by country until [they arrived] at Slen.'
\z 
 
\ea \label{ŽP.167}
\textit{silēne řeđē bā tewenē čā beynne bē beynū silēnē biɫberīne bē} \\ 
\gll silēne řeđ-ē bā tewenī-ē č=ā beyn=ne b-ē beyn=ū silēnē biɫber-ī=ne b-ē \\ 
 \textsc{pn} crossing\textsc{-dir}\textsc{.pl} \textsc{hort} stone\textsc{.f}\textsc{-indf} at=\textsc{dem.dist} between\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} between\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `They passed Slen. There was a stone in the road between Slen and Bilbar.'
\z 
 
\ea \label{ŽP.169}
\textit{mēnē čenū kināčekē serū tewenekēwe nīšāre} \\ 
\gll m-ē-nē čenū kināčē-ekē ser=ū tewenī-ekē=we ∅-nīš-ā=re \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} with\textsc{.ez.gen} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} on\textsc{\textsc{=ez.gen}} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} \textsc{ind-}sit\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \\ 
\glt `He (the uncle) went with the girl [and] sat on the stone.'
\z 
 
\ea \label{ŽP.171}
\textit{tewenē pāne bē nīšāre seyāwe} \\ 
\gll tewenī-ē p=āne b-ē ∅-nīš-ā=re sey-ā=we \\ 
 stone\textsc{.f}\textsc{-indf} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}sit\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} rest\textsc{.prs}\textsc{-3pl}\textsc{=compl} \\ 
\glt `It was a wide stone. There was a stone there. They sat [on the stone] to rest.'
\z 
 
\ea \label{ŽP.172}
\textit{kināčekē māčo, āāāy} \\ 
\gll kināčē-ekē m-āč-o āāāy \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} \\ 
\glt `The girl said, ‘Oooh.’'
\z 
 
\ea \label{ŽP.173}
\textit{qisē kero māčo, āāāy} \\ 
\gll qisē-ē ∅-ker-o m-āč-o āāāy \\ 
 talk\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} \\ 
\glt `The girl spoke [and] said, ‘Oooh.’'
\z 
 
\ea \label{ŽP.174}
\textit{lāloekeš māčo, erē weɫā sirewā} \\ 
\gll lālo-eke=š m-āč-o erē weɫā sirewā \\ 
 maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes indeed be cured\textsc{.pst}\textsc{.3sg} \\ 
\glt `Her uncle said, ‘Indeed, she is cured.’'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.176}
\textit{tewejoēšā kero wextē miđyo duē dānē dēwē yošā īnā peštīšne mē řāre yošā werišne} \\ 
\gll tewejo-ē=šā ∅-ker-o wext-ē mi-đye-o duē dāne-ē dēw-ē yo=šā īnā-∅ peštī=š=ne m-ē řā=re yo=šā wer=iš=ne \\ 
 caution\textsc{-indf}\textsc{=3pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} two \textsc{clf}\textsc{-dir}\textsc{.pl} demon\textsc{.m}\textsc{-dir}\textsc{.pl} one\textsc{.m}\textsc{=3pl} deic\textsc{-3sg}\textsc{.m} back\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} road\textsc{.f}\textsc{=\textsc{post}} one\textsc{.m}\textsc{=3pl} front\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `He noticed [their presence]. He realised that two demons [were with them], one behind her [the girl], one in front of her.'
\z 
 
\ea \label{ŽP.177}
\textit{ī dēwēnē ī kināčēšā bestēne} \\ 
\gll ī dēw=ē=nē ī kināčē=šā bestē=ne \\ 
 dem.prox demon\textsc{.m}\textsc{=ez}.\textsc{attr}\textsc{=cop}\textsc{.3pl} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3pl} tie\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `It was the demons who had muted the girl.'
\z 
 
\ea \label{ŽP.181}
\textit{kināčekē māčo, āāāy āāāy} \\ 
\gll kināčē-ekē m-āč-o āāāy āāāy \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} \textsc{intj} \\ 
\glt `[Then] the girl said, ‘Oooh.’'
\z 
 
\ea \label{ŽP.183}
\textit{kināčekē māčo, āāāy} \\ 
\gll kināčē-ekē m-āč-o āāāy \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} \\ 
\glt `The girl said, ‘Ooooh.’'
\z 
 
\ea \label{ŽP.188}
\textit{pīr šelīyār je yānekew wēšāne} \\ 
\gll pīr šelīyār je yāne-eke=ū wē=šā=ne \\ 
 \textsc{pn} \textsc{pn} from house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar [was] in his\footnote{Lit. their.} house.'
\z 
 
\ea \label{ŽP.191}
\textit{miđyo dūtifɫene duē neferē āmē} \\ 
\gll mi-đye-o dūtifɫe=ne duē nefer-ē āmā-ē \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{=\textsc{post}} two person\textsc{.m}\textsc{-dir}\textsc{.pl} come\textsc{.pst}\textsc{-3pl} \\ 
\glt `He saw that two people were coming from Dutifla.'
\z 
 
\ea \label{ŽP.192}
\textit{dēwē keřetē milo peštīšāne keřetē mē weršāne} \\ 
\gll dēw-ē keřet-ē mi-l-o peštī=šā=ne keřet-ē m-ē wer=šā=ne \\ 
 demon\textsc{.m}\textsc{-dir}\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} back\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} front\textsc{.m}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt ` A demon would sometimes follow them, [and] sometimes go ahead of them.'
\z 
 
\ea \label{ŽP.193}
\textit{keřetē milo peštīšāne keřetē mē weršāne} \\ 
\gll keřet-ē mi-l-o peštī=šā=ne keřet-ē m-ē wer=šā=ne \\ 
 time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} back\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} front\textsc{.m}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `[A demon] would sometimes follow them, [and] sometimes go ahead of them.'
\z 
 
\ea \label{ŽP.194}
\textit{duē yerē keřetē xuřoš wene, bēheyā bo megēɫowe} \\ 
\gll duē yerē keřet-ē ∅-xuř-o=š wene bē- heyā ∅-b-o me-gēɫ-o=we \\ 
 two three time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}shout\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{povb}without decency \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{neg-}turn\textsc{-3sg}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) shouted at him (the demon) two or three times. [But] he (the demon) was indecent; he did not return.'
\z 
 
\ea \label{ŽP.198}
\textit{fermāwo, weɫā čūn ānā bēheyānī kerūt penū ʕāɫemī} \\ 
\gll ∅-fermāw-o weɫā čūn āne=ā bēheyā=nī ∅-ker-ū=t pen=ū ʕāɫem-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed as \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} shameless\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{2sg}} trouble\textsc{.m}\textsc{\textsc{=ez.gen}} world\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) said, ‘Indeed, since you are so indecent, I will disgrace you.’'
\z 
 
\ea \label{ŽP.200}
\textit{tewejoēš wene kero pēse gāw til bore e pā tenūrū dēwāre til bowew ginorewe. ānen māčāš pene tenūrū dēwā} \\ 
\gll tewejo-ē=š wene ∅-ker-o pēse gāw til b-o=re e=pā tenūr=ū dēw-ā=re til b-o=we=ū gin-o=re=we āne=n m-āč-ā=š pene tenūr=ū dēw-ā \\ 
 caution\textsc{-indf}\textsc{=3sg} \textsc{post} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} as\_if bull\textsc{.m} rolling be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=that oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} rolling be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}}=and fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}}\textsc{=compl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He (Pir Shaliyar) paid attention to him. He (the demon) rolled down like a cow. It is there. It is called Tanuru Dewa.'
\z 
 
\ea \label{ŽP.201}
\textit{tenūrū dēwā ētir pēse gāw til bowe e čā werāwerū girdū mexeleqeɫāw hewrāmānīwe dēweke til bowe ginore soɫāwe} \\ 
\gll tenūr=ū dēw-ā ētir pēse gāw til ∅-b-o=we e=čā werāwer=ū gird=ū mexeleqeɫā=ū hewrāmān-ī=we dēw-eke til ∅-b-o=we ∅-gin-o=re soɫāwe \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{disc.ptcl} as\_if bull\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=there in\_front\_of\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt ` Then it was as if a cow had rolled down; the [demon] rolled down and fell [into a place called] Solawa before all the people’s eyes.'
\z 
 
\ea \label{ŽP.204}
\textit{perso, jeryān čikonū? ī yānew kābrāy kēnū īne kēnū?} \\ 
\gll ∅-pers-o jeryān čiko=n=ū ī yāne=ū kābrā-ī kē=n=ū īne kē=n=ū \\ 
 \textsc{ind-}ask\textsc{.prs}\textsc{-3sg} story\textsc{.m} where\textsc{=cop}\textsc{.3sg}\textsc{.m}=and dem.prox house\textsc{.m}=and man\textsc{-obl}\textsc{.m} who\textsc{=cop}\textsc{.3sg}\textsc{.m}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} who\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He came to the village. He asked, ‘[what is the story]? Where is the house of the fellow? Who is he? Who is he?'
\z 
 
\ea \label{ŽP.206}
\textit{ā wextī lālo kināčē māčo, min, tāze pādšāy kerdenā wekēɫ} \\ 
\gll ā wext-ī lālo kināčē m-āč-o min tāze pādšā-ī kerde=nā wekēɫ \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} anyway king\textsc{.m}\textsc{-obl}\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m} \\ 
\glt `Then, the girl’s uncle said, ‘As for me, well, the king has put me in charge.\footnote{Lit. the king has made me advocate.}'
\z 
 
\ea \label{ŽP.207}
\textit{mišyo ī kināčē nīkāh kerū pey toū be mezebū wētāū} \\ 
\gll mišyo ī kināčē nīkāh ∅-ker-ū pey to=ū be mezeb=ū wē=tā=ū \\ 
 aux dem.prox girl\textsc{.dir}\textsc{.f} marriage\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} to \textsc{2sg}=and by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.213}
\textit{īnē be min mewero kināčē pādšāy,} \\ 
\gll īnē be min me-wer-o kināčē pādšā-ī \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} with \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg} girl\textsc{.dir}\textsc{.f} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `She will not eat with me [under the same roof], the king’s daughter.'
\z 
 
\ea \label{ŽP.214}
\textit{min āne nānmā xeplezeřātīnen} \\ 
\gll min āne nān=m=ā xeplezeřātīne=n \\ 
 \textsc{1sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Look, my meal is xaplazaratine.’'
\z 
 
\ea \label{ŽP.215}
\textit{āne zerātiš šānī miđo, zerāte sūrī} \\ 
\gll āne zerāt=iš šānī mi-đe-o zerāt=e sūrī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} corn\textsc{=3sg} show \textsc{ind-}give\textsc{.prs}\textsc{-3sg} corn\textsc{=ez}\textsc{.cmpd} Syrian \\ 
\glt `He (Pir Shaliyar) showed him some corn, [some] red corn.'
\z 
 
\ea \label{ŽP.216}
\textit{āne šānī mido,} \\ 
\gll āne šānī mi-de-o \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} show \textsc{ind-}give\textsc{.prs} \\ 
\glt `He showed him some [corn and]'
\z 
 
\ea \label{ŽP.217}
\textit{fermāwo, min īne nānmārēū} \\ 
\gll ∅-fermāw-o min īne nān=m=ā=rē=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}}=and \\ 
\glt `He said, ‘This is my meal.’'
\z 
 
\ea \label{ŽP.222}
\textit{māčā, īne čēšā?} \\ 
\gll m-āč-ā īne čēš=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the people) said, ‘What is this?’'
\z 
 
\ea \label{ŽP.223}
\textit{īnē ī qisē čēšene? čēš māčdē? māčā čēš?} \\ 
\gll īnē ī qisē čēše=ne čēš m-āč-dē m-āč-ā čēš \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} dem.prox talk\textsc{.dir}\textsc{.f} what\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} what \textsc{ind-}say\textsc{.prs}-\textsc{2pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} what \\ 
\glt `What is this talk? What are you saying? What are they saying?'
\z 
 
\ea \label{ŽP.225}
\textit{her lāloeke māčo, mišyo ī kināčē min meberūšo māreš biřūnew’} \\ 
\gll her lālo-eke m-āč-o mišyo ī kināčē min me-ber-ū=š=o māre=š ∅-biř-ūne=ū \\ 
 emph maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} aux dem.prox girl\textsc{.dir}\textsc{.f} \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} marraige\textsc{=3sg} \textsc{ind-}cut\textsc{.prs}\textsc{-\textsc{1sg}}=and \\ 
\glt ` The uncle kept saying, ‘I shall not take this girl back. I shall marry her [to Pir Shaliyar].’'
\z 
 
\ea \label{ŽP.226}
\textit{āđīč māčo, min metāwū ī kināčē wey kerū īne nānmā} \\ 
\gll āđ=īč m-āč-o min me-tāw-ū ī kināčē wey ∅-ker-ū īne nān=m=ā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox girl\textsc{.dir}\textsc{.f} raising \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `People … he (Pir Shaliyar) said, ‘I cannot take care of this girl. This is my food.’'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.228}
\textit{yomā āne tišit miđo} \\ 
\gll yo=mā āne tiš=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} molasses\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One of us will give [pomegranate] molasses.'
\z 
 
\ea \label{ŽP.236}
\textit{min hiɫošet miđo’} \\ 
\gll min hiɫoše=t mi-đe-o \\ 
 \textsc{1sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `[One of them said], ‘I will give you cracked wheat.’'
\z 
 
\ea \label{ŽP.242}
\textit{qey mekero, mādām desūr e pēsen pādšā řāzīe nīā ī ženī bilowe ī kināčē bilowe qey mekero!’} \\ 
\gll qey me-ker-o mādām desūr e=pēse=n pādšā řāzīe nīye=ā ī ženī bi-l-o=we ī kināčē bi-l-o=we qey me-ker-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} as\_long\_as task/order\textsc{.m} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} satisfied \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} dem.prox woman\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `There is no problem. Now that things have developed this way, [and] the king is not happy for the girl to return, it’s all right! [She can stay here]’'
\z 
 
\ea \label{ŽP.243}
\textit{pīr šelīyār fermāwo, tehāā weɫāhī īne šime desūrtā dāw wāttā čūnke āđē be diɫtā wāt řāstā wāt hetā xwāy dinyāš weše kerdēne her menowe kemīč mekero her zīyā kero} \\ 
\gll pīr šelīyār ∅-fermāw-o tehāā weɫāhī īne šime desūr=tā dā=ū wāt=tā čūnke āđē be diɫ=tā wāt řās=tā wāt hetā xwā-ī dinyā=š weš-e kerdē=ne her ∅-men-o=we kem=īč me-ker-o her zīyā ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} task/order\textsc{.m}=\textsc{2pl} give\textsc{.pst}=and say\textsc{.pst}=\textsc{2pl} since 3pl\textsc{.dir} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} truth=\textsc{2pl} say\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} small\textsc{=add} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar said, ‘Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] continues to exist; it will not be diminished; it will grow.'
\z 
 
\ea \label{ŽP.244}
\textit{ī bezme šime wāttāw ī čāštīū īne ī desūre šime ījrātā kerd hetā xwāy dinyāš weše kerdēne šime řāstā wāt be diɫtā wāt} \\ 
\gll ī bezm=e šime wāt=tā=ū ī čāštī=ū īne ī desūr=e šime ījrā=tā kerd hetā xwā-ī dinyā=š weš-e kerdē=ne šime řās=tā wāt be diɫ=tā wāt \\ 
 dem.prox jollification\textsc{.dir}\textsc{.m}\textsc{=dem} \textsc{2pl} say\textsc{.pst}=\textsc{2pl}=and dem.prox meal\textsc{.f}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox task/order\textsc{.m}\textsc{=dem} \textsc{2pl} performance=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} \\ 
\glt `[After] the things you said, and the meals which you prepared, so long as the world continues to exist, you are right [in so doing]; you did these [things] in good faith.'
\z 
 
\ea \label{ŽP.256}
\textit{merāsēmekew pīr šelīyārī bergozār kerāū} \\ 
\gll merāsēm-eke=ū pīr šelīyār-ī bergozār ∅-ker-ā=ū \\ 
 ceremony\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} holding \textsc{ind-}do\textsc{.prs}\textsc{-3pl}=and \\ 
\glt `They held Pir Shaliyar’s ceremony.'
\z 
 
\ea \label{ŽP.258}
\textit{ā wextīyekey lālo pīr šelīyār hīn lālo heremū pīr šelīyārī milowe xeber berowe pey pādšāy} \\ 
\gll ā wext-ī-ekey lālo pīr šelīyār hīn lālo hereme=ū pīr šelīyār-ī mi-l-o=we xeber ∅-ber-o=we pey pādšā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} maternal\_uncle\textsc{.m} \textsc{pn} \textsc{pn} \textsc{fill} maternal\_uncle\textsc{.m} wife\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=compl} to king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, Pir Shaliyar’s uncle, eh, Pir Shaliyar’s wife’s uncle, went back [to Egypt] and passed the news to the king.'
\z 
 
\ea \label{ŽP.259}
\textit{xeber berowe pey pādšāy māčo, hāɫū jereyānē e čīnen} \\ 
\gll xeber ∅-ber-o=we pey pādšā-ī m-āč-o hāɫ=ū jereyān-ē e=čīne=n \\ 
 news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=compl} to king\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} state\textsc{.m}\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He passed the news to the king [and] said, ‘The story is like this.’'
\z 
 
\ea \label{ŽP.262}
\textit{lue! to luewe hem wesāylīč bere pey ī pīyāyū ī kināčēw} \\ 
\gll ∅-lu-e to ∅-lu-e=we hem wesāyl=īč ∅-ber-e pey ī pīyā-ī=ū ī kināčē=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} both appliance\textsc{.m}\textsc{=add} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}=and dem.prox daughter\textsc{.dir}\textsc{.pl}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter'
\z 
 
\ea \label{ŽP.264}
\textit{lālo mēwe wesāyil mārowe pey kināčē pēse īsey māčmē teleš pey berā} \\ 
\gll lālo m-ē=we wesāyil m-ār-o=we pey kināčē pēse īse-ī m-āč-mē tele=š pey ∅-ber-ā \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} appliance\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=compl} to girl\textsc{.dir}\textsc{.f} as\_if now\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-1pl} wedding\_gift\textsc{=3sg} to \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \\ 
\glt `The uncle went back. He brought household equipment for the girl, just like nowadays when we say, ‘They [the bride’s family] take gifts for her [the bride].’'
\z 
 
\ea \label{ŽP.266}
\textit{ēti melowe āđīč melowe} \\ 
\gll ēti me-l-o=we āđ=īč me-l-o=we \\ 
 \textsc{disc.ptcl} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `After all, he didn’t go back [to Egypt]. He did not go back either.'
\z 
 
\ea \label{ŽP.267}
\textit{ānen īse šelīyāre sīyāwā} \\ 
\gll āne=n īse šelīyār=e sīyāw=ā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[His shrine] is over there now; [he is called] Shaliyar Siyaw.'
\z 
 
\ea \label{BP.3}
\textit{yānē yānē meseleyš serū īneyn yānē bābā xwāy dā, kuřēš peyā bī} \\ 
\gll yānē yānē mesele-ī=š ser=ū īne-ī=n yānē bābā xwā-ī dā kuř-ē=š peyā bī-∅ \\ 
 that\_means that\_means issue\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} that\_means grandfather\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} give\textsc{.pst} son\textsc{.m}\textsc{-indf}\textsc{=3sg} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `The point with his name is that God granted the grandfather [a son]. A boy  was born to him .\footnote{Lit. A son of him appeared.}'
\z 
 
\ea \label{BP.15}
\textit{hečkes xelēš bē, xelēš čāne sānēnē} \\ 
\gll hečkes xele-ē=š b-ē xele-ē=š č=ā=ne sānē=nē \\ 
 whoever wheat\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be.prs\textsc{-pstc}\textsc{.3sg} wheat\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} from=\textsc{dem.dist}\textsc{=\textsc{post}} take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Whoever had wheat, they would collect the wheat from him as tax.'
\z 
 
\ea \label{BP.16}
\textit{hečkes ħeywāniš bē řūenāneš čene sānēnē} \\ 
\gll hečkes ħeywān=iš b-ē řūenāne=š čene sānē=nē \\ 
 whoever animal\textsc{.dir}\textsc{.m}\textsc{=3sg} be.prs\textsc{-pstc}\textsc{.3sg} animal\_tax\textsc{=3sg} from take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Whoever had [domestic] animals, they would collect animal tax from him.'
\z 
 
\ea \label{BP.17}
\textit{ā pāse hetā hečkesē ženīš dē šū, bešepīyākew ēme begzāyē ēme} \\ 
\gll ā pāse hetā hečkes-ē ženī=š de-ē šū bešepīyā-eke=ū ēme begzāye-ē ēme \\ 
 \textsc{dem.dist} like even whoever\textsc{-indf} woman.f\textsc{=3sg} give.prs\textsc{-pstc}\textsc{.3sg} husband\textsc{.m} quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{1pl} nobleman\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} \\ 
\glt `[Things were] like this. Even when someone wanted to get his daughter married, he had to pay a quota  to our noblemen.'
\z 
 
\ea \label{BP.18}
\textit{eger ženīmā dē be bešepīyākew yo terī čīwēš zereriš sānē čene wāčēnēš sūrāne} \\ 
\gll eger ženī=mā de-ē be bešepīyā-eke=ū yo ter-ī čīw-ē=š zerer=iš sān-ē čene wāč-ēn-ē=š sūrāne \\ 
 if woman.f\textsc{=\textsc{1pl}} give.prs\textsc{-cond} to quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} thing\textsc{.m}\textsc{-indf}\textsc{=3sg} harm\textsc{.m}\textsc{=3sg} buy.prs\textsc{-pstc}\textsc{.3sg} from say.prs.\textsc{sbjv-}pstc\textsc{-3pl}\textsc{=3sg} \textsc{pn} \\ 
\glt `If one\footnote{Lit. we.} had married women from the quota of another [lord], then they [the lord of the bride's village] would charge him [the bride’s father]. This money is called Surana.'
\z 
 
\ea \label{BP.23}
\textit{āmēnē e čē čwār meʕmūrē penj meʕmūrēš kīyāsēnē} \\ 
\gll āmē=nē e=čē čwār meʕmūr-ē penj meʕmūr-ē=š kīyāsē=nē \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} in=here four officer\textsc{.m}\textsc{-dir}\textsc{.pl} five officer\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They came here. Four or five men would be sent'
\z 
 
\ea \label{BP.27}
\textit{jā luēnē berdenšāwe, luēnē berdenšāwe pey āġekew ʕērāqīšā} \\ 
\gll jā luē=nē berde=n=šā=we luē=nē berde=n=šā=we pey āġe-eke=ū ʕērāq-ī=šā \\ 
 afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} for agha\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3pl} \\ 
\glt `Then, they would go back and take the taxes to their aghas in Iraq.'
\z 
 
\ea \label{BP.28}
\textit{āġekew ʕērāqī berdenšāwe peyšāw} \\ 
\gll āġe-eke=ū ʕērāq-ī berde=n=šā=we pey=šā=ū \\ 
 agha\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} to\textsc{=3pl}=and \\ 
\glt `The aghas in Iraq, they took back [the taxes] to them.'
\z 
 
\ea \label{BP.38}
\textit{ī kābrēče je ʕērāqo āmēnē, wēregānew nīmeřonew seʕbne bexšnāyšā bexšnēnēwe} \\ 
\gll ī kābrā-ē=č=e je ʕērāq=o āmē=nē wēregā=ne=ū nīmeřo=ne=ū seʕbe=ne bexšnāy=šā bexšn-ēn-ē=we \\ 
 \textsc{dem.prox} man\textsc{-dir}\textsc{.pl}\textsc{=add}\textsc{=dem} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} evening\textsc{=\textsc{post}}=and noon\textsc{=\textsc{post}}=and morning.f\textsc{=\textsc{post}} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-indf}\textsc{=compl} \\ 
\glt `They (people) would donate [food] to the fellows (the tax collectors) who had come from Iraq, in the evenings, mornings, and at noon. '
\z 
 
\ea \label{BP.40}
\textit{her wēregāne pēse čāweɫ sempāšē bēnē, yoīšā dēnē duīšā dēnē yānew yoy ī meʕmūrīye ke ʕērāqo āmēnē, ā zemānū pīr šelīyārī bexšnāyšā bexšnēnē} \\ 
\gll her wēregā=ne pēse čāweɫ sempāš-ē b-ēn-ē yo-ī=šā de-ēn-ē duē-ī=šā de-ēn-ē yāne=ū yo-ī ī meʕmūr-ī=e ke ʕērāq=o āmē=nē ā zemān=ū pīr šelīyār-ī bexšnāy=šā bexšn-ēn-ē \\ 
 just evening\textsc{=\textsc{post}} like in\_the\_past poison\_sprayer\textsc{-dir}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} two\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} house\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} officer\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=dem} rel \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `As in the old days when there were people whose job was spraying pesticide on trees, [and] each would be hosted at the house of one [host], the officers who had come from Iraq would be fed in the time of Pir Shaliyar. '
\z 
 
\ea \label{BP.43}
\textit{yānew yo terī, her šewē her wēregāne ī mēmānē mēmānū yoy bīyēne} \\ 
\gll yāne=ū yo ter-ī her šewe-ē her wēregā=ne ī mēmān-ē mēmān=ū yo-ī bīyē=ne \\ 
 house\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} just night.f\textsc{-obl}.f just evening\textsc{=\textsc{post}} \textsc{dem.prox} guest\textsc{.m}\textsc{-dir}\textsc{.pl} guest\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `[likewise, one tax collector] would be the guest in the house of one [fellow from Hawraman]  each evening, each night.'
\z 
 
\ea \label{BP.45}
\textit{milā xeɫk pāne milo miđyo īne ā xerīkū genekārīnē ī kābrāy je āmēnē, ī meʕmūrē xerīkū genekārīnēw} \\ 
\gll mi-l-ā xeɫk p=ā=ne mi-l-o mi-đye-o īne=ā xerīk=ū genekārī=nē ī kābrā-ī je āmē=nē ī meʕmūr-ē xerīk=ū genekārī=nē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3pl} people\textsc{.m} at=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}look.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} man\textsc{-obl}\textsc{.m} from come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `People realised that these men were busy engaging in debauchery; these officers were busy engaing in debauchery.'
\z 
 
\ea \label{BP.46}
\textit{xeber yāwo be bābā xwāđāw ħemey ġeybī} \\ 
\gll xeber ∅-yāw-o be bābā xwāđā=ū ħeme=ī ġeybī \\ 
 news\textsc{.dir}\textsc{.m} \textsc{ind-}arrive.prs\textsc{-3sg} to grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible \\ 
\glt `The news reached Baba Khwada and Hama the invisible.'
\z 
 
\ea \label{BP.47}
\textit{māčā, ī īnē āmēnē melāwe’} \\ 
\gll m-āč-ā ī īnē āmē=nē me-l-ā=we \\ 
 \textsc{ind-}say.prs\textsc{-3pl} \textsc{dem.prox} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They (people) said, ‘These (tax collectors) have come here; they have not gone back yet.’'
\z 
 
\ea \label{BP.50}
\textit{e pāne tēkeɫī kerāw bezmē miđā wene} \\ 
\gll e=pāne tēkeɫī ∅-ker-ā=ū bezm-ē mi-đe-ā wene \\ 
 in=there mixture\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl}=and jollification\textsc{.dir}\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}give.prs\textsc{-3pl} at \\ 
\glt `They misbehaved. They made a big mess.'
\z 
 
\ea \label{BP.52}
\textit{dey ēšew herkes mēmāniš hen, ēšew herkes mēmāniš hen, mēmānekeš šewē witē, sereš biřo} \\ 
\gll dey ēšew herkes mēmān=iš hen-∅ ēšew herkes mēmān=iš hen-∅ mēmān-eke=š šewe-ē wit-ē sere=š ∅-biř-o \\ 
 \textsc{disc.ptcl} tonight everyone guest\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} tonight everyone guest\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} guest\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} night.f\textsc{-obl}.f sleep\textsc{.pst}\textsc{-3pl} head\textsc{.m}\textsc{=3sg} \textsc{sbjv-}cut.prs\textsc{-3sg} \\ 
\glt `‘Tonight, whoever has guests, when the guest sleeps at night, he shall decapitate him.'
\z 
 
\ea \label{BP.53}
\textit{mēmānekeyš sere biřo māđām īne kāršā bīyen} \\ 
\gll mēmān-ekey=š sere ∅-biř-o māđām īne kār=šā bīye=n \\ 
 guest\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} head\textsc{.m} \textsc{sbjv-}cut.prs\textsc{-3sg} as\_long\_as \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Since they (the tax collectors) are involved in this (debauchery), everyone should decapitate his guest.'
\z 
 
\ea \label{BP.57}
\textit{āđ māčo, konē ī meʕmūrē mine činnē wextēn? je hewrāmān bīyēnē āmēnē pey biřew bāre} \\ 
\gll āđ m-āč-o ko=nē ī meʕmūr-ē min=e činnē wext-ē=n je hewrāmān bīyē=nē āmē=nē pey biře=ū bāre \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} where\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1sg}\textsc{=dem} some\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} for portion\textsc{.m}=and profit\textsc{.m} \\ 
\glt `He (the agha) says, ‘where have my officers been during this time? They were in Hawraman; they had gone to collect tax.'
\z 
 
\ea \label{BP.60}
\textit{ēmeyč māčmē, ī meʕmūrē šime kīyāsēnē bā biřew bāremā dēw nāmēwe bārāšo} \\ 
\gll ēme=yč m-āč-mē ī meʕmūr-ē šime kīyāsē=nē ∅-b-ā biře=ū bāre=mā dā-ē=ū ne-āmā-ē=we b-ār-ā=š=o \\ 
 \textsc{1pl}\textsc{=add} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{ind-}be.prs\textsc{-3pl} portion\textsc{.m}=and profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}bring.prs\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `We will say, ‘We paid tax to the officers whom you had sent, [however,] they didn’t return to bring it back.'
\z 
 
\ea \label{BP.64}
\textit{jā īne qewɫ kerā yo} \\ 
\gll jā īne qewɫ ∅-ker-ā yo \\ 
 afterwards \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} arrangement\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} one\textsc{.m} \\ 
\glt `Then, they (the people of Hawraman) all agreed [about the plan]\footnote{Lit. they made their oaths one.}'
\z 
 
\ea \label{BP.73}
\textit{ewē mūsā ētir be mišār serešā biřā} \\ 
\gll ewē m-ūs-ā ētir be mišār sere=šā ∅-biř-ā \\ 
 3pl\textsc{.dir} \textsc{ind-}sleep.prs\textsc{-3pl} \textsc{disc.ptcl} to hand\_saw\textsc{.m} head\textsc{.m}\textsc{=3pl} \textsc{ind-}cut.prs\textsc{-3pl} \\ 
\glt `They (the guests) had slept. He decapitated them with a hand saw.'
\z 
 
\ea \label{BP.76}
\textit{miđyā īne be mišārī sereš biřyen nāmā nāmā nīyā} \\ 
\gll mi-đye-ā īne be mišār-ī sere=š biřye=n ne-āmā-∅ ne-āmā-∅ nīye=ā \\ 
 \textsc{ind-}look.prs\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} to hand\_saw\textsc{.m}\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=3sg} cut\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They noticed that the one who had decapitated his guests with a hand saw hadn’t come to the mosque. He hadn’t arrived. He wasn’t there.'
\z 
 
\ea \label{BP.85}
\textit{pāse qisēšā kerdēnē} \\ 
\gll pāse qisē-ē=šā kerdē=nē \\ 
 like talk.f\textsc{-dir}\textsc{.pl}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `The story is known like this.'
\z 
 
\ea \label{BP.86}
\textit{āđīč mē werew māčo, fermāwo, dā dey dey jārēne jimdēre’} \\ 
\gll āđ=īč m-ē wer=ew m-āč-o ∅-fermāw-o dā dey dey jār-ē=ne ∅-jim-dē=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}come.prs\textsc{.3sg} front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{disc.ptcl} \textsc{disc.ptcl} time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} \\ 
\glt `He (Baba Khwada) came forward [and] said, ‘Please, set off the way all of you.’'
\z 
 
\ea \label{BP.92}
\textit{biřew bārekešā qubūɫ kerdū girtšā tāwēɫū} \\ 
\gll biře=ū bāre-eke=šā qubūɫ kerd=ū girt=šā tāwēɫ=ū \\ 
 portion\textsc{.m}=and profit\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} accepted do\textsc{.pst}=and take\textsc{.pst}\textsc{=3pl} delivery\textsc{.m}=and \\ 
\glt `They accepted our taxes; they delivered them.'
\z 
 
\ea \label{BP.97}
\textit{seremā biřyē, īne sereū ēmew īne šimšērū šime} \\ 
\gll sere=mā biřī-ē īne sere=ū ēme=ū īne šimšēr=ū šime \\ 
 head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} \textsc{2pl} \\ 
\glt `We have decapitated them. Here are their heads and the agha’s swords.\footnote{Lit. These are our heads and these are the agha’s swords.}'
\z 
 
\ea \label{BP.99}
\textit{ewē šūnīšāre milā ētir her werew milā řārē} \\ 
\gll ewē šūnī=šā=re mi-l-ā ētir her wer=ew mi-l-ā řā=rē \\ 
 3pl\textsc{.dir} track.f\textsc{=3pl}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} \textsc{disc.ptcl} just front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} road.f\textsc{=\textsc{post}} \\ 
\glt `They (the people) followed them. They walked slowly.'
\z 
 
\ea \label{BP.104}
\textit{ewēč her serešā kumnānre} \\ 
\gll ewē=č her sere=šā kumnā=n=re \\ 
 3sg\textsc{.dir}.f\textsc{=add} just head\textsc{.m}\textsc{=3pl} bend\_down\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} \\ 
\glt `They (the people) nodded [in agreement]'
\z 
 
\ea \label{BP.105}
\textit{bābā xwāđāw ħemey ġeybīū ħemey wirdīkɫe her ginēnē weɫē īnē weɫēne ejošā girđ īnē šūnīšāwe} \\ 
\gll bābā xwāđā=ū ħeme=ī ġeybī=ū ħeme=ī wirdīkɫe her gin-ēn-ē weɫē īnē weɫē=ne ejo=šā girđ īnē šūnī=šā=we \\ 
 grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible=and \textsc{pn}=ez.\textsc{attr} small just fall.prs\textsc{-pstc}\textsc{-3pl} front \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} front\textsc{=\textsc{post}} guess.prs\textsc{=3pl} all \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} track.f\textsc{=3pl}\textsc{=compl} \\ 
\glt `Baba Khwada, Hama the invisible, and little Hama were in front, assuming that everyone was following them.'
\z 
 
\ea \label{BP.107}
\textit{wextē miđyo her ā yerē neferēnē ewēšā gird gēɫēnēwe} \\ 
\gll wext-ē mi-đye-o her ā yerē nefer-ē=nē ewēšā gird gēɫē=nē=we \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} just \textsc{dem.dist} three person\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} \\ 
\glt `They realised that there were only the three of them. All the rest had returned [to Hawraman]'
\z 
 
\ea \label{BP.108}
\textit{wextē miđyo her ā yerē neferēnē ewēšā gird gēɫēnēwe} \\ 
\gll wext-ē mi-đye-o her ā yerē nefer-ē=nē ewēšā gird gēɫē=nē=we \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} just \textsc{dem.dist} three person\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} \\ 
\glt ` They realised that there were only the three of them. All the rest had returned [to Hawraman].'
\z 
 
\ea \label{BP.109}
\textit{fermāwo, īne konē xeɫkeke čenū ēme bē?} \\ 
\gll ∅-fermāw-o īne ko=nē xeɫk-eke čenū ēme b-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} people\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} with\textsc{.ez.gen} \textsc{1pl} be.prs\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Baba Khwada) said, ‘Where are the people who were with us?'
\z 
 
\ea \label{BP.110}
\textit{ēme firē bēnmē. ṣew neferē bēnmē āmāymē řāre.} \\ 
\gll ēme fire-ē b-ēn-mē ṣew nefer-ē b-ēn-mē āmā-īmē řā=re \\ 
 \textsc{1pl} a\_lot\textsc{-pl} be.\textsc{prs}\textsc{-pstc}\textsc{-\textsc{1pl}} hundred person\textsc{.m}\textsc{-dir.pl} be.\textsc{prs}\textsc{-pstc}\textsc{-\textsc{1pl}} come\textsc{.pst}-\textsc{1pl} road.f\textsc{=\textsc{post}} \\ 
\glt `We were a large number. We were one hundred  people who came here.'
\z 
 
\ea \label{BP.111}
\textit{īne konē ewēšā? čēššā pene āmā?} \\ 
\gll īne ko=nē ewēšā čēš=šā pene āmā-∅ \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} what\textsc{=3pl} to come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `Where are they? What happened to them?'
\z 
 
\ea \label{BP.112}
\textit{īne konē ewē? čēššā pene āmā?} \\ 
\gll īne ko=nē ewē čēš=šā pene āmā-∅ \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} what\textsc{=3pl} to come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `Where are they? What happened to them?'
\z 
 
\ea \label{BP.115}
\textit{hewrāmī tā meřeber yānē hewrāmīyē tā meřeber āmē ētir girđ gēɫēnēwe čāge} \\ 
\gll hewrāmī tā meřeber yānē hewrāmī-ē tā meřeber āmā-ē ētir girđ gēɫē=nē=we č=āge \\ 
 Hawrami until \textsc{pn} that\_means Hawrami\textsc{-dir}\textsc{.pl} until \textsc{pn} come\textsc{.pst}\textsc{-3pl} \textsc{disc.ptcl} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} from=there \\ 
\glt `hawrāmī tā mařabar means the people of Hawraman came as far as Marabar. Then, they all returned to Hawraman from there.'
\z 
 
\ea \label{BP.116}
\textit{tenyā āđē luēnē} \\ 
\gll tenyā āđē luē=nē \\ 
 only 3pl\textsc{.dir} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Only they (Baba Khwada, Hama Ghaybi, and Hama Wirdikla) went [to Iraq]'
\z 
 
\ea \label{BP.117}
\textit{ētir āđē wātenšā, ēme diruē meyeymē} \\ 
\gll ētir āđē wāte=n=šā ēme diruē me-de-ymē \\ 
 \textsc{disc.ptcl} 3pl\textsc{.dir} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `They said, ‘We are not going to lie.'
\z 
 
\ea \label{BP.121}
\textit{māčmē ēme serū īney ī pīyēmā kuštēnē”’} \\ 
\gll m-āč-mē ēme ser=ū īney ī pīyē=mā kuštē=nē \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{1pl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox} man\textsc{.dir}\textsc{.pl}\textsc{=\textsc{1pl}} kill\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We say, ‘We have killed these men because of this.’'
\z 
 
\ea \label{BP.124}
\textit{ā wextī milārewe māčā, wiɫā ī meʕmūrū šime kīyāsēbēnētā pey īne, īnemā āne heke šime gerekbē dēmā penew} \\ 
\gll ā wext-ī mi-l-ā=re=we m-āč-ā wiɫā ī meʕmūr=ū šime kīyāsē b-ēn-ē=tā pey īne īne=mā āne heke šime gerek b-ē dā-ē=mā pene=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.\textsc{prs}\textsc{-3pl}\textsc{=\textsc{povb}}\textsc{=compl} \textsc{ind-}say.prs\textsc{-3pl} by\_God \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl}=\textsc{2pl} for \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1pl}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} if \textsc{2pl} necessary be.prs\textsc{-pstc}\textsc{.3sg} give\textsc{.pst}\textsc{-3pl}\textsc{=\textsc{1pl}} to=and \\ 
\glt `Then, they went [to the agha and] said, ‘Indeed, the officers whom you had sent to us, whatever [ taxation] you had asked for, we gave them,'
\z 
 
\ea \label{BP.125}
\textit{āđē nāmēw xerīkū genekārī bēnēw} \\ 
\gll āđē ne-āmā-ē=ū xerīk=ū genekārī b-ēn-ē=ū \\ 
 3pl\textsc{.dir} \textsc{neg-}come\textsc{.pst}\textsc{-3pl}=and busy\textsc{=ez.gen} debauchery\textsc{.m} be.prs\textsc{-pstc}\textsc{-3pl}=and \\ 
\glt `[however] they didn’t come back [to you]. They engaged in debauchery.'
\z 
 
\ea \label{BP.128}
\textit{īne serew ēmew īne šimšērū āġey} \\ 
\gll īne sere=ū ēme=ū īne šimšēr=ū āġe-ī \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Here are their heads and the agha’s swords.'
\z 
 
\ea \label{BP.131}
\textit{bābā xuāđāw ħemew ġeybī, āđē, āđīšā kerā zīnān’} \\ 
\gll bābā xuāđā=ū ħeme=ū ġeybī āđē āđīšā ∅-ker-ā zīnān \\ 
 grandfather\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez.gen} invisible 3pl\textsc{.dir} 3pl\textsc{.obl} \textsc{ind-}do.prs\textsc{-3pl} prison\textsc{.m} \\ 
\glt `Baba Khwada and Hama Ghaybi, they (agha’s men) put them  in prison.'
\z 
 
\ea \label{BP.139}
\textit{berā kerāšāne zīnāneke, wextē miđyā mezā nū če noʕē bīyen ħejījīyāyčšā girten xuɫkū hejījī} \\ 
\gll ∅-ber-ā ∅-ker-ā=šā=ne zīnān-eke wext-ē mi-đye-ā me-zān-ū če noʕ-ē bīye=n ħejījī-yā=yčšā girte=n xuɫk=ū hejīj-ī \\ 
 \textsc{ind-}take.prs\textsc{-3pl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=3pl}\textsc{=\textsc{povb}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-\textsc{1sg}} what kind/way\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} people\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They took [them] and put them in the jail, they realised that —they didn’t know for what reason— [but] a person from Hajij had also been arrested.'
\z 
 
\ea \label{BP.140}
\textit{āđīč her īnā je zīnānekene} \\ 
\gll āđ=īč her īnā-∅ je zīnān-eke=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} just deic\textsc{-3sg}\textsc{.m} from prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man from Hajij) was also in the prison. '
\z 
 
\ea \label{BP.141}
\textit{bāwe yo kerāšāwe yo, yerē neferē, neferē hejījīw duēū hewrāmānīyēū} \\ 
\gll b-ā=we yo ∅-ker-ā=šā=we yo yerē nefer-ē nefer-ē hejījī=ū duē=ū hewrāmānī-ē=ū \\ 
 be.prs\textsc{-3pl}\textsc{=compl} one\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=3pl}\textsc{=compl} one\textsc{.m} three person\textsc{.m}\textsc{-dir}\textsc{.pl} person\textsc{.m}\textsc{-indf} from\_Hajij=and two=and from\_Hewraman\textsc{-dir}\textsc{.pl}=and \\ 
\glt `They became one. They were put in the same cell , three people; one from Hajij and two from Hawraman.'
\z 
 
\ea \label{BP.148}
\textit{mekerāš zīnānī gēɫo šārekene nānšā pey peyā kero} \\ 
\gll me-ker-ā=š zīnān-ī ∅-gēɫ-o šār-eke=ne nān=šā pey peyā ∅-ker-o \\ 
 \textsc{neg-}do.prs\textsc{-3pl}\textsc{=3sg} prison\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}wander.prs\textsc{-3sg} city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} bread\textsc{.m}\textsc{=3pl} for visible \textsc{ind-}do.prs\textsc{-3sg} \\ 
\glt `They didn’t put him in jail. He would wander in town and find food for them.'
\z 
 
\ea \label{BP.149}
\textit{ētir māwāš fire bo yā kem bo her činne māngē bo her činne wextē bo wextē milo serū řūčinekey nānšā pey wizorew wextē miđyo hešpišā dēnēne} \\ 
\gll ētir māwā=š fire ∅-b-o yā kem ∅-b-o her činne mānge-ē ∅-b-o her činne wext-ē ∅-b-o wext-ē mi-l-o ser=ū řūčine-ekey nān=šā pey ∅-wiz-o=re=ū wext-ē mi-đye-o hešpišī-ā dē=nē=ne \\ 
 \textsc{disc.ptcl} period\textsc{.m}\textsc{=3sg} a\_lot \textsc{ind-}be.prs\textsc{-3sg} either little \textsc{ind-}be.prs\textsc{-3sg} just some month.f\textsc{-dir}\textsc{.pl} \textsc{ind-}be.prs\textsc{-3sg} just some time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}be.prs\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} on\textsc{=ez.gen} chimney\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} bread\textsc{.m}\textsc{=3pl} for \textsc{ind-}throw.prs\textsc{-3sg}\textsc{=\textsc{povb}}=and time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} \\ 
\glt `Their imprisonment lasted either a long time or short time. It took them months. Once, when he (Hama Wirdikla) went to the chimney to throw food for them, he realised that they had gotten lice.'
\z 
 
\ea \label{BP.151}
\textit{hešpišā dēnēne, čē hurmēzo pēse hēlyānī miročē nīšore} \\ 
\gll hešpišī-ā dē=nē=ne čē hur- m-ēz-o pēse hēlyānī miroče-ē ∅-nīš-o=re \\ 
 lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} here \textsc{pvb-}\textsc{ind-}rise.prs\textsc{-3sg} like nest ant\textsc{-dir}\textsc{.pl} \textsc{ind-}sit.prs\textsc{-3sg}\textsc{=\textsc{povb}} \\ 
\glt `They had gotten lice. The lice had grown up like ants. '
\z 
 
\ea \label{BP.154}
\textit{māčā, čī girewī āne miheme?’} \\ 
\gll m-āč-ā čī ∅-girew-ī āne miheme \\ 
 \textsc{ind-}say.prs\textsc{-3pl} why \textsc{ind-}cry.prs-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{pn} \\ 
\glt `They said, ‘Mihamad, why are you crying?’'
\z 
 
\ea \label{BP.155}
\textit{fermāwo āđīč, weɫā min pey šime girewū} \\ 
\gll ∅-fermāw-o āđ=īč weɫā min pey šime ∅-girew-ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} indeed \textsc{1sg} for \textsc{2pl} \textsc{ind-}cry.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘I’m crying for you.’'
\z 
 
\ea \label{BP.164}
\textit{gošiš nedārāyne ētir jā bowe xeber bārewe} \\ 
\gll goš=iš ne-dārā-ī=ne ētir jā b-o=we xeber b-ār-e=we \\ 
 ear\textsc{.m}\textsc{=3sg} \textsc{neg-}hold\textsc{.pst}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{disc.ptcl} afterwards \textsc{imp-}come.prs\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{imp-}bring.prs-\textsc{2sg}\textsc{=compl} \\ 
\glt `If they don’t listen to you, then, come back and let us know.'
\z 
 
\ea \label{BP.165}
\textit{wāče her kesī nān dānī, wāče, ēšew be xātirū xwāy toū īđū ī kiɫfetū yāneyte dūr ginewe čī šārī’} \\ 
\gll wāč-e her kes-ī nān dā=nī wāč-e ēšew be xātir=ū xwā-ī to=ū īđ=ū ī kiɫfet=ū yāne-ī=t=e dūr ∅-gin-e=we čī šār-ī \\ 
 say.prs-\textsc{imp-}\textsc{2sg} just person\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} say.prs-\textsc{imp-}\textsc{2sg} tonight to sake\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}=and 3sg.prox\textsc{.dir}\textsc{.m}=and \textsc{dem.prox} child\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} far \textsc{imp-}fall.prs-\textsc{2sg}\textsc{=compl} why city\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Tell whoever gave you food, say, “Tonight, for the sake of God, you and your household get away from this town.'
\z 
 
\ea \label{BP.167}
\textit{ā wextī milo mēwe herkesī goš dārābone, dūr kewtenwe je, kučē je šāreke meselen e čē luān tā milemārfā wēšū xānewādeš} \\ 
\gll ā wext-ī mi-l-o m-ē=we herkesī goš dārā b-one dūr kewte=n=we je kuč-ē je šār-eke meselen e=čē luā=n tā milemārfā wē=š=ū xānewāde=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}come.prs\textsc{.3sg}\textsc{=compl} whoever ear\textsc{.m} hold\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3sg} far fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} from little\textsc{-indf} from city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} for\_example in=here go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until \textsc{pn} \textsc{reflx}\textsc{=3sg}=and family\textsc{.m}\textsc{=3sg} \\ 
\glt `Then, he went [to town] and came back. Whoever might have listened to him had gone away from the city. For instance, they had gone together with their family as far as Milamarfa\footnote{name of a summer habitat.}.'
\z 
 
\ea \label{BP.169}
\textit{zīnāneke šokē kero} \\ 
\gll zīnān-eke šokē ∅-ker-o \\ 
 prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} shake \textsc{ind-}do.prs\textsc{-3sg} \\ 
\glt `He made the prison tremble.'
\z 
 
\ea \label{BP.170}
\textit{zīnāneke šokē kerow} \\ 
\gll zīnān-eke šokē ∅-ker-o=ū \\ 
 prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} shake \textsc{ind-}do.prs\textsc{-3sg}\textsc{=ez.gen} \\ 
\glt `He made the prison tremble.'
\z 
 
\ea \label{BP.172}
\textit{hejījīyeke, fermāwo hejījīyeke, be qurbānit bim sūre gīyān yekē tirī lē de leqē teriš de wene’} \\ 
\gll hejījī-eke ∅-fermāw-o hejījī-eke be qurbān=it ∅-b-im sūr=e gīyān yek-ē tir=ī lē ∅-de-e leq-ē ter=iš ∅-de-e wene \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} to sacrifice\textsc{=\textsc{2sg}} \textsc{sbjv-}be.prs\textsc{-\textsc{1sg}} red=ez\textsc{.cmpd} dear.\textsc{voc} one\textsc{-indf} other\textsc{=3sg} at \textsc{imp-}give.prs-\textsc{2sg} kick\textsc{-indf} another\textsc{=3sg} \textsc{imp-}give.prs-\textsc{2sg} at \\ 
\glt `The [prisoner from] Hajij said [to Baba Khwada], ‘May I be your sacrifice, you red-skin, kick it once more! Kick it once more!”'
\z 
 
\ea \label{BP.173}
\textit{bāzem dūbāre ī sefer wextē milowe dimā leqē māɫo zīnānekeyre zīnāneke yekser veɫādīm bo} \\ 
\gll bāzem dūbāre ī sefer wext-ē mi-l-o=we dimā leq-ē ∅-māɫ-o zīnān-ekey=re zīnān-eke yekser veɫādīm ∅-b-o \\ 
 again again \textsc{dem.prox} trip time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg}\textsc{=compl} afterwards kick\textsc{-indf} \textsc{ind-}smash.prs\textsc{-3sg} prison\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} totally collapsed \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `This time, when he went backward and kicked the prison [wall], the prison totally collapsed.'
\z 
 
\ea \label{BP.174}
\textit{zīnāneke yekser weɫādīm bo šāreke ētir ginowe serū āđēū} \\ 
\gll zīnān-eke yekser weɫādīm ∅-b-o šār-eke ētir ∅-gin-o=we ser=ū āđē=ū \\ 
 prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} totally collapsed \textsc{ind-}be.prs\textsc{-3sg} city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{disc.ptcl} \textsc{ind-}fall.prs\textsc{-3sg}\textsc{=compl} on\textsc{=ez.gen} 3pl\textsc{.dir}=and \\ 
\glt `The prison totally collapsed. The prison collapsed on them. '
\z 
 
\ea \label{BP.175}
\textit{āđē āzāyē bā mizyāre řū be hewrāmān} \\ 
\gll āđē āzā-ē ∅-b-ā mi-zye-ā=re řū be hewrāmān \\ 
 3pl\textsc{.dir} free\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} face to \textsc{pn} \\ 
\glt `They were liberated. They returned to Hawraman. '
\z 
 
\ea \label{BP.180}
\textit{šāreke šēwīo zīnānīyē mirexesē bā be āzāđī mēnēwe} \\ 
\gll šār-eke šēwye-o zīnānī-ē mirexes-ē ∅-b-ā be āzāđī m-ē-nē=we \\ 
 city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} be\_confused.prs\textsc{-3sg} prisoner\textsc{-dir}\textsc{.pl} released\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} to freedom\textsc{.m} \textsc{ind-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \\ 
\glt `There was turmoil in the city. The prisoners were liberated. They went back [to Hawraman] freely.'
\z 
 
\ea \label{BP.189}
\textit{āđēč miqdārē be enāzew seʕātē nīšāre dīyāršo} \\ 
\gll āđē=č miqdārē be enāze=ū seʕāt-ē ∅-nīš-ā=re dīyār=š=o \\ 
 3pl\textsc{.dir}\textsc{=add} a\_little to proportion\textsc{=ez.gen} hour\textsc{.m}\textsc{-indf} \textsc{ind-}sit.prs\textsc{-3pl}\textsc{=\textsc{povb}} region\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `They (Baba Khwada and Hama Wirdikla) waited for him for about an hour.'
\z 
 
\ea \label{BP.196}
\textit{ġeyb bīyen luān zemīnegere} \\ 
\gll ġeyb bīye=n luā=n zemīn-eke=re \\ 
 disappearance be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} land\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He had disappeared into the ground.'
\z 
 
\ea \label{BP.197}
\textit{ā wextī ētir āđē zānā ġeyb bīyen zemīnegere luān} \\ 
\gll ā wext-ī ētir āđē ∅-zān-ā ġeyb bīye=n zemīn-eke=re luā=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} 3pl\textsc{.dir} \textsc{ind-}know.prs\textsc{-3pl} disappearance be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} land\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, they realised that he had disappeared into the ground. '
\z 
 
\ea \label{BP.208}
\textit{īne dāstānū īney, īnē dāstānū bābāw pīr šelīyārīū ħemey ġeybī bīyēnē} \\ 
\gll īne dāstāne=ū īney īnē dāstāne=ū bābā=ū pīr šelīyār-ī=ū ħeme=ī ġeybī bīyē=nē \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story.f\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \textsc{pn}=ez.\textsc{attr} invisible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `This was the story of Pir Shaliyar’s grandfather and Hama Ghaybi.'
\z 
 
\ea \label{BP.209}
\textit{luēnē īnešā wene qomīyān} \\ 
\gll luē=nē īne=šā wene qomyā=n \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3pl} at happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] these things happened to them.'
\z 
 
\ea \label{BP.212}
\textit{īnē dāstānū āđī tā ewā} \\ 
\gll īnē dāstāne=ū āđī tā ewā \\ 
 \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} until there \\ 
\glt `This was his tale. [It was] until there.'
\z 
 

\ea \label{ŠJ.12}
\textit{tifengiš henū īneš henū āneš henū!} \\ 
\gll tifeng=iš hen-∅=ū īne=š hen-∅=ū āne=š hen-∅=ū \\ 
 gun\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist-3sg.m}=and \\ 
\glt `He has a gun, he has this, and he has that!'
\z 
 
\ea \label{ŠJ.19}
\textit{ānen pīyā!} \\ 
\gll āne=n pīyā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} man\textsc{.m} \\ 
\glt ` He is the man!’'
\z 
 
\ea \label{ŠJ.20}
\textit{āđīč seʕbē pīyā kīyāno māčo lodē bārdēš!} \\ 
\gll āđ=īč seʕbe-ē pīyā ∅-kīyān-o m-āč-o ∅-lo-dē b-ār-dē=š \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} morning.f\textsc{-obl}.f man\textsc{.m} \textsc{ind-}send.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}bring.prs-\textsc{2pl}\textsc{=3sg} \\ 
\glt `The next day, he (Jafir Sān) sent some men [and] said, ‘Go [and] bring him!'
\z 
 
\ea \label{ŠJ.28}
\textit{ʕāyetū īnsānē bēʕeqɫē tifengšā koɫēwe tifengšā bow} \\ 
\gll ʕāyet=ū īnsān-ē bēʕeqɫ-ē tifeng=šā koɫē=we tifeng=šā ∅-b-o=ū \\ 
 habit\textsc{=ez.gen} human\textsc{.m}\textsc{-dir}\textsc{.pl} silly\textsc{\textsc{-pl}} gun\textsc{=3pl} shoulder\textsc{=compl} gun\textsc{=3pl} \textsc{ind-}be.prs\textsc{-3sg}=and \\ 
\glt `According to the customs, the silly men [from Luhon] had guns;'
\z 
 
\ea \label{ŠJ.30}
\textit{māčā qurbān īne čiko bīyenī} \\ 
\gll m-āč-ā qurbān īne či=ko bīye=nī \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=where be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `They said, ‘Sir, where have you been?'
\z 
 
\ea \label{ŠJ.39}
\textit{čenū gum bīyey gumē bīyā serew řoē gumē bā mezānā bilāwe pey nođšey} \\ 
\gll čenū gum bīyey gum-ē bī-ā sere=ū řo-ē gum-ē ∅-b-ā me-zān-ā bi-l-ā=we pey nođše-ī \\ 
 with\textsc{.ez.gen} lost be\textsc{.inf} lost\textsc{-indf} be\textsc{.pst}\textsc{-3pl} \textsc{clf}\textsc{=ez.gen} day\textsc{.m}\textsc{-dir}\textsc{.pl} lost\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-3pl}\textsc{=compl} to \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[They got so scared that] they were lost. They were lost for one day. They didn’t know the way to go back to Nodsha.'
\z 
 
\ea \label{ŠJ.42}
\textit{dimāw serew řoē milāwe} \\ 
\gll dimā=ū sere=ū řo-ē mi-l-ā=we \\ 
 afterwards\textsc{=ez.gen} \textsc{clf}\textsc{=ez.gen} day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `After a day, they returned [to Nodsha].'
\z 
 
\ea \label{ŠJ.45}
\textit{weɫē bīyēnmē āro duē řoē ētir mezānmē koge milmē} \\ 
\gll weɫē bīyē=nmē āro duē řo-ē ētir me-zān-mē koge mi-l-mē \\ 
 front be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} today two day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{disc.ptcl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} where \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \\ 
\glt `We have spread out for two days; we didn’t know where we were going.’'
\z 
 
\ea \label{ŠJ.57}
\textit{tāteš tewēɫī bīyen šē ʕumere} \\ 
\gll tāte=š tewēɫe-ī bīye=n šē ʕumer-e \\ 
 father\textsc{.m}\textsc{=3sg} \textsc{pn}\textsc{-nmlz} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} sheikh\textsc{.m} \textsc{pn}\textsc{-def}\textsc{.dir} \\ 
\glt ` Sheikh Umer, his father, was from Tawela.'
\z 
 
\ea \label{ŠJ.64}
\textit{wiɫāxekā, berekey kero milone ew dīm wēš wizone tewēɫeke} \\ 
\gll wiɫāx-ekā bere-ekey ∅-ker-o mi-l-one ew dīm wē=š ∅-wiz-one tewēɫe-eke \\ 
 horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} door\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}do.prs\textsc{-3sg} \textsc{ind-}go.prs\textsc{-3sg} \textsc{dem.dist} side\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}throw.prs\textsc{-3sg} stable\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `He opened the door and went inside. He forced himself into the stable.'
\z 
 
\ea \label{ŠJ.65}
\textit{wēš wizone tewēɫeke} \\ 
\gll wē=š ∅-wiz-one tewēɫe-eke \\ 
 \textsc{reflx}\textsc{=3sg} \textsc{ind-}throw.prs\textsc{-3sg} stable\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `He forced himself into the table.'
\z 
 
\ea \label{ŠJ.66}
\textit{seʕbē wiɫāxdārē milā pey hīn lāw wiɫāxā tewēɫī wiɫāxē ber bārā bilā šūnū hermānērew kāērew hēzmīre} \\ 
\gll seʕbe-ē wiɫāxdār-ē mi-l-ā pey hīn lā=ū wiɫāx-ā tewēɫe-ī wiɫāx-ē ber ∅-bār-ā bi-l-ā šūn=ū hermānē=re=ū kāē=re=ū hēzmī=re \\ 
 morning.f\textsc{-obl}.f stableman\textsc{-dir}\textsc{.pl} \textsc{ind-}go.prs\textsc{-3pl} to \textsc{fill} at\_the\_place\_of\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-obl}\textsc{.pl} stable\textsc{.m}\textsc{-obl}\textsc{.m} horse/donkey\textsc{.m}\textsc{-dir}\textsc{.pl} out \textsc{sbjv-}load\textsc{.m}\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-3pl} after\textsc{=ez.gen} work\textsc{.obl}.f\textsc{=\textsc{post}}=and dry\_grass\textsc{.dir}\textsc{.pl}\textsc{=\textsc{post}}=and firewood.f\textsc{=\textsc{post}} \\ 
\glt ` In the morning, the horse grooms went to the horses in the stable. They [wanted] to take the horses out [and] go after their jobs, [collecting] straw, and wood.'
\z 
 
\ea \label{ŠJ.68}
\textit{kirž milā xeber miđā be še ʕusmānū tewēɫē} \\ 
\gll kirž mi-l-ā xeber mi-đe-ā be še ʕusmān=ū tewēɫē \\ 
 quick \textsc{ind-}go.prs\textsc{-3pl} news\textsc{.dir}\textsc{.m} \textsc{ind-}give.prs\textsc{-3pl} to sheikh \textsc{pn}\textsc{=ez.gen} \textsc{pn} \\ 
\glt `They went quickly [and] and passed the news to Sheikh Usman'
\z 
 
\ea \label{ŠJ.82}
\textit{ʕeybekeyčiš memeno} \\ 
\gll ʕeyb-eke=yč=iš me-men-o \\ 
 fault\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=add}\textsc{=3sg} \textsc{neg-}remain.prs\textsc{-3sg} \\ 
\glt `The defect had disappeared.'
\z 
 
\ea \label{ŠJ.85}
\textit{nīšārew qisē fermāwāw firīū} \\ 
\gll ∅-nīš-ā=re=ū qisē-ē ∅-fermāw-ā=ū firī=ū \\ 
 \textsc{ind-}sit.prs\textsc{-3pl}\textsc{=\textsc{povb}}=and talk.f\textsc{-dir}\textsc{.pl} \textsc{ind-}say.prs\textsc{-3pl}\textsc{=ez.gen} many\textsc{=ez.gen} \\ 
\glt `They sat and talked a lot'
\z 
 
\ea \label{ŠJ.91}
\textit{eger pīyāgerī bo, eger šēxyetī bo, eger nāmē bo řīk megino mentēqēwene de neferē bā de nānēšā bā yo mezāno řū we kē kero} \\ 
\gll eger pīyāgerī ∅-b-o eger šēxyetī ∅-b-o eger nāmē ∅-b-o řīk me-gin-o mentēqe-ēwe=ne de nefer-ē ∅-b-ā de nān-ē=šā ∅-b-ā yo me-zān-o řū=we kē ∅-ker-o \\ 
 if gentlemanhood \textsc{sbjv-}be.prs\textsc{-3sg} if sheikhhood \textsc{sbjv-}be.prs\textsc{-3sg} if name.f \textsc{sbjv-}be.prs\textsc{-3sg} arranged \textsc{neg-}fall.prs\textsc{-3sg} region\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} ten person\textsc{.m}\textsc{-indf} \textsc{sbjv-}be.prs\textsc{-3pl} ten bread\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3pl} \textsc{sbjv-}be.prs\textsc{-3pl} one\textsc{.m} \textsc{neg-}know.prs\textsc{-3sg} face=to who \textsc{sbjv-}do.prs\textsc{-3sg} \\ 
\glt `Say this job is about being a gentleman, being a sheikh, being well known, [still] it couldn’t be that if in a region there are ten people [worthy of ruling] who have power over ten families, one doesn’t know whom to plead with [when there is a problem]'
\z 
 
\ea \label{ŠJ.94}
\textit{šēx herčē fermāwištē fermāwebā šē ʕumerū bīyārē fermāwo peyšū} \\ 
\gll šēx herčē fermāwišt-ē fermāwe b-ā šē ʕumer=ū bīyārē ∅-fermāw-o pey=š=ū \\ 
 sheikh\textsc{.m} every speech\textsc{.m}\textsc{-dir}\textsc{.pl} say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3pl} sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} for\textsc{=3sg}=and \\ 
\glt `Whatever saying there was, Sheikh Umer told him'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{ŠJ.99}
\textit{kāke gīyān īnī ečānē bīyēnē} \\ 
\gll kāke gīyān īnī e=čāne-ē bīyē=nē \\ 
 brother.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.pl} in=like\_that\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Dear brother, they were like this.'
\z 
 
\ea \label{ŠJ.100}
\textit{šexsē ečānē bīyēnē} \\ 
\gll šexs-ē e=čāne-ē bīyē=nē \\ 
 saint\textsc{-dir}\textsc{.pl} in=like\_that\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They were such people.'
\z 
 
\ea \label{ŠJ.101}
\textit{dey āro be kē wāčī derđwārde bo je dinyāne kes nemenen} \\ 
\gll dey āro be kē ∅-wāč-ī derđwārde ∅-b-o je dinyā=ne kes ne-mene=n \\ 
 \textsc{disc.ptcl} today to who \textsc{sbjv-}say.prs.\textsc{sbjv-}\textsc{2sg} qualified\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} from world\textsc{.dir}.f\textsc{=\textsc{post}} person\textsc{.m} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Nowadays, who can you refer to in the world who would be qualified? No one has remained.'
\z 
 
\ea \label{ŠJ.105}
\textit{tenhā yo kuřēšā meneniš duktir Mādēh dinyāne řāš mebowe} \\ 
\gll tenhā yo kuř-ē=šā mene=n=iš duktir Mādēh dinyā=ne řā=š me-b-o=we \\ 
 not\_even one\textsc{.m} son/boy\textsc{.m}\textsc{-indf}\textsc{=3pl} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} physician\textsc{.m} \textsc{pn} world\textsc{.dir}.f\textsc{=\textsc{post}} road.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `Only a son is left of the chieftains, named Dr Madeh, who is not allowed here.'
\z 
 
\ea \label{ŠJ.108}
\textit{dinyāne řāš mebowe} \\ 
\gll dinyā=ne řā=š me-b-o=we \\ 
 world\textsc{.dir}.f\textsc{=\textsc{post}} road.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `[That’s why] he is not allowed here.'
\z 
 
\ea \label{KŠ.6}
\textit{tīrēš wene teqno} \\ 
\gll tīr-ē=š wene ∅-teqn-o \\ 
 bullet\textsc{.dir}\textsc{.m}\textsc{-indf}\textsc{=3sg} at \textsc{ind-}fire\textsc{.prs}\textsc{-3sg} \\ 
\glt `He fired a bullet at them.'
\z 
 
\ea \label{KŠ.12}
\textit{māčo, xās to čārenwīsnī, kināčēw pādšāyt īse pey kē niwīse? nāmēš pey kē?} \\ 
\gll m-āč-o xās to čārenwīs=nī kināčē=ū pādšā-ī=t īse pey kē niwīs-e nāmē=š pey kē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} fine \textsc{2sg} fortune\_teller\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} now to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} name\textsc{.f}\textsc{=3sg} to who \\ 
\glt `He (the king) said, ‘All right, [assuming that] you’re a fortune teller, whom have you considered [as the future husband] for the king’s daughter?\footnote{Lit. Whom did you write the king’s daughter for?} Who is she chosen for?'
\z 
 
\ea \label{KŠ.15}
\textit{ā wextīyekey, yewāšē wātiš pey, kināčēw pādšāyt pey kē niwīse?} \\ 
\gll ā wext-ī-ekey yewāšē wāt=iš pey kināčē=ū pādšā-ī=t pey kē niwīs-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} then say\textsc{.pst}\textsc{=3sg} to daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Then, he (the king) said, ‘Whom have you considered [as the future husband] for the king’s daughter?'
\z 
 
\ea \label{KŠ.17}
\textit{āđīč hukmiš dā wātiš, ženē leme peře mebo bo} \\ 
\gll āđ=īč hukm=iš dā wāt=iš ženī-ē leme peř-e me-b-o ∅-b-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} rule\textsc{.m}\textsc{=3sg} give\textsc{.pst} say\textsc{.pst}\textsc{=3sg} woman\textsc{.f}\textsc{-indf} stomach\textsc{.f} full\textsc{-f} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the king) issued an order [and] said, ‘No pregnant women should stay alive!’\footnote{Lit. No pregnant women should be.}'
\z 
 
\ea \label{KŠ.18}
\textit{benāxātirē īney āyeč bilo, hāmɫē bē} \\ 
\gll benāxātirē īney āye=č bi-l-o hāmɫē b-ē \\ 
 for\_the\_reason \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} 3sg\textsc{.f}\textsc{.dir}\textsc{=add} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} pregnant\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `The reason [for issuing this order] was that she (the shepherd’s wife), who was pregnant, would disappear.'
\z 
 
\ea \label{KŠ.23}
\textit{ā wextī zāwɫeke peyā bī šūnū peyā bīyeyšre eyākēš merđe} \\ 
\gll ā wext-ī zāwɫe-eke peyā bī-∅ šūn=ū peyā bīyey=š=re eđā-ekē=š merđ-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} child\textsc{-def}\textsc{.dir}\textsc{.m} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} after\textsc{=ez.gen} visible be\textsc{.inf}\textsc{=3sg}\textsc{=\textsc{post}} mother\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} die\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Well, the child was born. His mother died after his birth.'
\z 
 
\ea \label{KŠ.25}
\textit{zāwɫeke āmā luā bizū māmālēwe šotiš wārd} \\ 
\gll zāwɫe-eke āmā luā bize=ū māmā-lē-ēwe šot=iš wārd-∅ \\ 
 child\textsc{-def}\textsc{.dir}\textsc{.m} come\textsc{.pst}\textsc{.3sg} go\textsc{.pst}\textsc{.3sg} goat\textsc{.f}\textsc{=ez.gen} grandmother\textsc{.f}\textsc{-dim}\textsc{.f}\textsc{-indf} milk\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `The child went and drank milk from the goat of a grandmother.'
\z 
 
\ea \label{KŠ.26}
\textit{duē dehfē wārdiš} \\ 
\gll duē dehfe-ē wārd=iš \\ 
 two time\textsc{.f}\textsc{-dir}\textsc{.pl} eat\textsc{.pst}\textsc{=3sg} \\ 
\glt `He drank milk [from the goat] two times.'
\z 
 
\ea \label{KŠ.29}
\textit{āđīč wāt, dey xāsā min gino šūnī[šre]} \\ 
\gll āđ=īč wāt dey xās=ā min ∅-gin-o šūnī=š=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} track\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `He (the shepherd) said, ‘All right, I will look after her (the goat).’'
\z 
 
\ea \label{KŠ.31}
\textit{bizekē luā čērwe kuřekey memew bizekēš wārdū} \\ 
\gll bize-ekē luā čēr=we kuř-ekey meme=ū bize-ekē=š wārd=ū \\ 
 goat\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} go\textsc{.pst}\textsc{.3sg} under\textsc{=compl} son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} breast\textsc{.dir}\textsc{.m}\textsc{=ez.gen} goat\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} eat\textsc{.pst}=and \\ 
\glt `and that the goat went there. The boy fed from the goat’s udder.'
\z 
 
\ea \label{KŠ.40}
\textit{kināčēw pāđšāyč māčo, kuřū šuwāneyč bārdē ēme tāqetmā pey āney mē} \\ 
\gll kināčē=ū pāđšā-ī=č m-āč-o kuř=ū šuwāne-ī=č b-ār-dē ēme tāqet=mā pey āney m-ē \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} \textsc{1pl} persistence\textsc{.m}\textsc{=\textsc{1pl}} to \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `The king’s daughter said, ‘Fetch the shepherd’s son. He will entertain us.\footnote{Lit. Our patience comes by him.}’'
\z 
 
\ea \label{KŠ.43}
\textit{māčo be eyāšū tāteyšū āđīšā, bā īne menowe pey nokerū ēme epēne pey werdesīmā pey hermānberū yāneymā} \\ 
\gll m-āč-o be eyā=š=ū tāte-ī=š=ū āđīšā bā īne ∅-men-o=we pey noker=ū ēme epēne pey werdes-ī=mā pey hermānber=ū yāne-ī=mā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by mother\textsc{=3sg}=and father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and 3pl\textsc{.obl} \textsc{hort} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} to servant\textsc{.m}\textsc{=ez.gen} \textsc{1pl} here to attendant\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} to worker\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `She (the girl) said to her parents and others , ‘Let him stay here as our servant here, as our attendant, as our domestic worker.'
\z 
 
\ea \label{KŠ.52}
\textit{āđīč māčo, wāčdē bē melāqātīm} \\ 
\gll āđ=īč m-āč-o wāč-dē b-ē melāqātī=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} say\textsc{.prs}-\textsc{imp-}\textsc{2pl} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} visitor\textsc{.m}\textsc{=\textsc{1sg}} \\ 
\glt `He (the king) said, ‘Tell him to come to visit me.’'
\z 
 
\ea \label{KŠ.54}
\textit{āđīč milo čenīū duē yerē milo} \\ 
\gll āđ=īč mi-l-o čenī=ū duē yerē mi-l-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} with\textsc{=ez.gen} two three \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the shepherd’s son) too went, together with two, three [persons].'
\z 
 
\ea \label{KŠ.55}
\textit{āđ milo melāqātīw āđī} \\ 
\gll āđ mi-l-o melāqātī=ū āđī \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} visitor\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `He (the shepherd’s son) went to visit him (the king).'
\z 
 
\ea \label{KŠ.60}
\textit{yewāšē čāge mēwe yāne} \\ 
\gll yewāšē č=āge m-ē=we yāne \\ 
 then in=there \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} house\textsc{.m}\textsc{.dir}.sg \\ 
\glt `From there, he (the shepherd’s son) returned home. '
\z 
 
\ea \label{KŠ.63}
\textit{āđīč gēɫo āšpezxānekene kučē čīw menebo čāštū čīw wero} \\ 
\gll āđ=īč ∅-gēɫ-o āšpezxāne-eke=ne kuč-ē čīw mene b-o čāštī=ū čīw ∅-wer-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}wander\textsc{.prs}\textsc{-3sg} kitchen\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} little\textsc{-indf} thing\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} meal\textsc{.f}\textsc{\textsc{=ez.gen}} thing\textsc{.m} \textsc{ind-}eat\textsc{.prs}\textsc{-3sg} \\ 
\glt `He looked into the kitchen. There was some leftovers. He ate them. '
\z 
 
\ea \label{KŠ.65}
\textit{kināčekē hāyš bowe wextē miđyo āmāno lāw tāteyčšo mēw} \\ 
\gll kināčē-ekē hāy=š ∅-b-o=we wext-ē mi-đye-o āmā=n=o lā=ū tāte-ī=č=š=o m-ē=ū \\ 
 girl\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} awareness\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}=and \\ 
\glt `The [king’s] daughter became aware [of the coughing], and noticed that he (the shepherd’s son) had returned; he had returned from the place of her father.'
\z 
 
\ea \label{KŠ.73}
\textit{īnēč dīyāš pore bezeyīš āmā pore} \\ 
\gll īnē=č dīyā=š pore bezeyī=š āmā pore \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=add} look\textsc{.pst}\textsc{=3sg} at pity\textsc{.m}\textsc{=3sg} come\textsc{.pst}\textsc{.3sg} at \\ 
\glt `She looked at him. She took pity on him.'
\z 
 
\ea \label{KŠ.76}
\textit{īne wēš xeberiš nīyāw sewāđiš nīyā} \\ 
\gll īne wē=š xeber=iš nīye=ā=ū sewāđ=iš nīye=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{reflx}\textsc{=3sg} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and literacy\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He wasn’t aware [of the death plot] that \_\_ he was illiterate\_\_'
\z 
 
\ea \label{KŠ.79}
\textit{wēš hurmēzo hur yo minvīso kināčekē} \\ 
\gll wē=š hur-m-ēz-o hur yo mi-nvīs-o kināčē-ekē \\ 
 \textsc{reflx}\textsc{=3sg} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg} \textsc{povb} one\textsc{.m} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} girl\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} \\ 
\glt `The girl rose [and] wrote one [letter] herself. '
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.83}
\textit{xulāsew kelāmī qisēš čenū min kerdēw} \\ 
\gll xulāse=ū kelām-ī qisē=š čenū min kerd-ē=ū \\ 
 summary=and speech\textsc{-obl}\textsc{.m} talk\textsc{.dir}\textsc{.pl}\textsc{=3sg} with\textsc{\textsc{.ez}.gen} \textsc{1sg} do\textsc{.pst}\textsc{-3pl}=and \\ 
\glt `In short, he (the king) talked to me.'
\z 
 
\ea \label{KŠ.84}
\textit{īnā qāqezēčiš dānā pey} \\ 
\gll īne=ā qāqez-ē=č=iš dā=nā pey \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} letter\textsc{.m}\textsc{-indf}\textsc{=add}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to \\ 
\glt `Look, he has given me a letter.'
\z 
 
\ea \label{KŠ.91}
\textit{ētir ew čone āyā xeber zāno yā mezāno,} \\ 
\gll ētir ew č=o=ne āyā xeber ∅-zān-o yā me-zān-o \\ 
 \textsc{disc.ptcl} \textsc{dem.dist} in=\textsc{dem.dist}\textsc{=\textsc{post}} whether news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} or \textsc{neg-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `[It is not clear] whether he (the king) was aware of the wedding or not. '
\z 
 
\ea \label{KŠ.92}
\textit{yewāšē šūnū duē sāɫēre kināčē nīšo lemewe kuřē wīno} \\ 
\gll yewāšē šūn=ū duē sāɫe-ē=re kināčē ∅-nīš-o leme=we kuř-ē ∅-wīn-o \\ 
 then after\textsc{\textsc{=ez.gen}} two year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{.f} \textsc{ind-}sit\textsc{.prs}\textsc{-3sg} stomach\textsc{.f}\textsc{=compl} boy\textsc{.m}\textsc{-indf} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} \\ 
\glt `After two years, the girl became pregnant\footnote{Lit. She sits on belly.} [and] gave birth to a son.'
\z 
 
\ea \label{KŠ.93}
\textit{kuřē wīno yewāšē xeber berā māčā be xwā hīnēšā wer dē begzā, xānēšā wer dē āmē, āmēwe} \\ 
\gll kuř-ē ∅-wīn-o yewāšē xeber ∅-ber-ā m-āč-ā be xwā hīn-ē=šā wer dē begzā xān-ē=šā wer dē āmē āmē=we \\ 
 boy\textsc{.m}\textsc{-indf} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} then news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} by God\textsc{.m} \textsc{fill}\textsc{-dir}\textsc{.pl}\textsc{=3pl} out give\textsc{.pst}\textsc{.3pl} nobleman chief\textsc{.m}\textsc{-indf}\textsc{=3pl} out give\textsc{.pst}\textsc{.3pl} come\textsc{.pst}\textsc{.3pl} come\textsc{.pst}\textsc{.3pl}\textsc{=compl} \\ 
\glt `She gave birth to a son. Then, the news arrived that indeed the thingummy, the lord, the Khan, had been released [and] he had returned.'
\z 
 
\ea \label{KŠ.94}
\textit{yewāšē ewēč milā wero, xeɫk milo pāyšāwe} \\ 
\gll yewāšē ewē=č mi-l-ā ∅-wer-o xeɫk mi-l-o pā-ī=šā=we \\ 
 then 3pl\textsc{.dir}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{ind-}out\textsc{-3sg} people\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} foot\textsc{-obl}\textsc{.m}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `Then, they (the shepherd’s son’s family) went to the front [of the palace]; people welcomed them.'
\z 
 
\ea \label{KŠ.95}
\textit{kināčekē māčo be šūekeyš kuřekey,} \\ 
\gll kināčē-ekē m-āč-o be šū-ekey=š kuř-ekey \\ 
 girl\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by husband\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `The [king’s] daughter said to her husband, to the [shepherd’s] son,'
\z 
 
\ea \label{KŠ.97}
\textit{yewāšē neferēwe řemo weɫēwe māčo be xānī, be xwā āne to kārū xwāyt tefre dā,} \\ 
\gll yewāšē nefer-ēwe ∅-řem-o weɫē=we m-āč-o be xān-ī be xwā āne to kār=ū xwā-ī=t tefre dā \\ 
 then person\textsc{.m}\textsc{-indf} \textsc{ind-}run\textsc{.prs}\textsc{-3sg} front\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by chief\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2sg} task\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} evasion give\textsc{.pst} \\ 
\glt `Then, a person went forward and said to the Khan, ‘Indeed, you [tried to] evade God’s destiny.'
\z 
 
\ea \label{KŠ.98}
\textit{yārū tifengeket teqnā, ženū šuwāney kušī,} \\ 
\gll yārū tifeng-eke=t teqnā ženī=ū šuwāne-ī ∅-kuš-ī \\ 
 \textsc{fill} gun\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} shoot\textsc{.pst} woman\textsc{.f}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}kill\textsc{.prs}-\textsc{2sg} \\ 
\glt `You fired a bullet from your gun; [and] you [intended] to kill the shepherd’s wife.'
\z 
 
\ea \label{KŠ.99}
\textit{ā wextīyekey čārenwīsī niwīsebēš ā kināčē to pey kuřū šuwāney niwīsyene} \\ 
\gll ā wext-ī-ekey čārenwīs-ī niwīse b-ē=š ā kināčē to pey kuř=ū šuwāne-ī niwīsye=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} fortune\_teller\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{dem.dist} daughter\textsc{.dir}\textsc{.f} \textsc{2sg} to son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}.pass\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `Then, the fortune teller had allocated\footnote{Lit. written.} your daughter for the shepherd’s son. '
\z 
 
\ea \label{KŠ.100}
\textit{āne kuřū šuwāneyn serū esbekey to} \\ 
\gll āne kuř=ū šuwāne-ī=n ser=ū esb-ekey to \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} on\textsc{\textsc{=ez.gen}} horse\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{2sg} \\ 
\glt `Look, the shepherd’s son [is] on your horse.'
\z 
 
\ea \label{KŠ.101}
\textit{āne kināčekētne ženīšne} \\ 
\gll āne kināčē-ekē=t=ne ženī=š=ne \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.f} woman\textsc{.f}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `That is your daughter. She is his wife.'
\z 
 
\ea \label{ŽH.2}
\textit{eyām qisē kerdē wāčē, zemānū qedīmī pādšā hāřūne(l) řešīd bīyenū nā siɫtān mehmūw heyāsī žīr} \\ 
\gll eyā=m qisē-ē kerdē-ē wāč-ē zemān=ū qedīm-ī pādšā hāřūn-e(l) řešīd bīye=n=ū nā siɫtān mehmū=ū heyās=ī žīr \\ 
 mother\textsc{=1sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-3pl} say\textsc{.prs-}pstc\textsc{.3sg} time\textsc{.m}\textsc{=ez}\textsc{.gen} old\_time\textsc{-obl}\textsc{.m} king\textsc{.m} \textsc{pn}-of \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and no sultan\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez}.\textsc{attr} shrewd \\ 
\glt `My mother would narrate [tales]. She said, ‘In the olden days, there was Harun Rashid, the king. No, [there was] Sultan Mahmud and Hayas, the shrewd.'
\z 
 
\ea \label{ŽH.3}
\textit{wāčē řoē je řoā qāɫešā bo, serū mezān ū čēšī} \\ 
\gll wāč-ē řo-ē je řo-ā qāɫe=šā ∅-b-o sere=ū me-zān-ū čēš-ī \\ 
 say\textsc{.prs-}pstc\textsc{.3sg} day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} quarrel\textsc{\textsc{.f}}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} head\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what\textsc{-obl}\textsc{.m} \\ 
\glt `According to her (my mother),\footnote{Lit. She would say.}  Once they (Sultan Mahmud and Hayas) got into a quarrel with each other, about something I don’t know.'
\z 
 
\ea \label{ŽH.13}
\textit{ewīč māčo, min pilekānī mezānū binyewnew mezānū čēš māčī!’} \\ 
\gll ew=yč m-āč-o min pilekānī me-zān-ū bi-nye-ūne=ū me-zān-ū čēš m-āč-ī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg stair\textsc{\textsc{.f}} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} \textsc{sbjv-}put\textsc{.prs}\textsc{-1sg}=and \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what \textsc{ind-}say\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘I can’t set up a ladder and I don’t understand\footnote{Lit. know.}  what you are saying!’'
\z 
 
\ea \label{ŽH.19}
\textit{āđīč milo mizgī} \\ 
\gll āđ=īč mi-l-o mizgī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} mosque\textsc{.m} \\ 
\glt `He (Hayas) went to the mosque.'
\z 
 
\ea \label{ŽH.20}
\textit{milowe yāne. yerē kināčēš bā.} \\ 
\gll mi-l-o=we yāne yerē kināčē=š ∅-b-ā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m}\textsc{.dir}.sg three daughter\textsc{.dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `He (the man) returned home. He had three daughters.'
\z 
 
\ea \label{ŽH.21}
\textit{yerē kināčēš bā kābrāy} \\ 
\gll yerē kināčē=š bā kābrā-ī \\ 
 three daughter\textsc{.dir}\textsc{.pl}\textsc{=3sg} \textsc{hort} man\textsc{-obl}\textsc{.m} \\ 
\glt `The fellow had three daughters.'
\z 
 
\ea \label{ŽH.22}
\textit{ā wextīyekey, kināče gewrekēš māčo, tāte, her wēt āmānīwe?’} \\ 
\gll ā wext-ī-ekey kināčē=e gewre-ekē=š m-āč-o tāte her wē=t āmā=nī=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} big\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m} just \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `Then, his older daughter said, ‘Father, have you come back alone?’'
\z 
 
\ea \label{ŽH.27}
\textit{wātiš, tāte āne wāteniš, hekāyetē weš gīsnene, bā ī řē kučē keme kermē be hekāyetekeyt rāz biřmē} \\ 
\gll wāt=iš tāte āne wāte=n=iš hekāyet-ē weš-∅ ∅-gīsn-e=ne bā ī řē kuč-ē kem-e ∅-ker-mē be hekāyet-ekey=t rāz ∅-biř-mē \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} tale\textsc{.m}\textsc{-indf} well-m \textsc{imp-}start\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{hort} \textsc{dem.prox} road\textsc{.obl}\textsc{\textsc{.f}}.sg little\textsc{-indf} s\textsc{hort}\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} by tale\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{2sg}} tale\textsc{.m} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl} \\ 
\glt `She said, ‘Father, he had said, “Start [telling] a good tale, so that we [are entertained] and our way seems shorter. Let us finish your tale.”'
\z 
 
\ea \label{ŽH.30}
\textit{wātiš, qisēmā nekerdē āmāymē nizīkū dege,} \\ 
\gll wāt=iš qisē=mā ne-kerd-ē āmā-īmē nizīk=ū dege \\ 
 say\textsc{.pst}\textsc{=3sg} talk\textsc{.dir}\textsc{.pl}\textsc{=1pl} \textsc{neg-}do\textsc{.pst}\textsc{-3pl} come\textsc{.pst} \textsc{1pl} close\textsc{=ez}\textsc{.gen} village\textsc{.obl}\textsc{\textsc{.f}} \\ 
\glt `He said, ‘We didn’t talk [until] we got close to the village.'
\z 
 
\ea \label{ŽH.33}
\textit{wātiš, tāte, āne wāteniš, eger to xuɫkem mekerī, meberīm yāne milū mizgī’} \\ 
\gll wāt=iš tāte āne wāte=n=iš eger to xuɫke=m me-ker-ī me-ber-ī=m yāne mi-l-ū mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} if \textsc{2sg} invitation\textsc{=1sg} \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \textsc{neg-}take\textsc{.prs}-\textsc{2sg}\textsc{=1sg} house\textsc{.m}\textsc{.dir.sg} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} mosque\textsc{.m} \\ 
\glt `She said, ‘Father, he meant,\footnote{Lit. said.} “If you do not invite me and won’t take me home, I will go to the mosque.”'
\z 
 
\ea \label{ŽH.36}
\textit{duē řoē her mizgīne bē} \\ 
\gll duē řo-ē her mizgī=ne b-ē \\ 
 two day\textsc{.m}\textsc{-dir}\textsc{.pl} just mosque\textsc{.m}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He stayed in the mosque for two days.'
\z 
 
\ea \label{ŽH.37}
\textit{řo yerē āmāwe wātiš be kināčekē, řoɫe řefēqekem her mizgīne īnā mizgīne} \\ 
\gll řo yerē āmā=we wāt=iš be kināčē-ekē řoɫe řefēq-eke=m her mizgī=ne īnā-∅ mizgī=ne \\ 
 day\textsc{.m} three come\textsc{.pst}\textsc{.3sg}\textsc{=compl} say\textsc{.pst}\textsc{=3sg} to daughter\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} child.\textsc{voc} friend\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} just mosque\textsc{.m}\textsc{=\textsc{post}} deic\textsc{-3sg}\textsc{.m} mosque\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `On the third day, he (the father) returned [home and] said to the daughter, ‘Child, my friend is still in the mosque; he is at the mosque.’'
\z 
 
\ea \label{ŽH.41}
\textit{ā wextīyekey kināčekē beroš řāre māčo, be xwā īse kučēš čene werū.} \\ 
\gll ā wext-ī-ekey kināčē-ekē ∅-ber-o=š řā=re m-āč-o be xwā īse kuč-ē=š čene ∅-wer-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} now little\textsc{-indf}\textsc{=3sg} from \textsc{ind-}eat\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then, the girl set off on the way. She said, ‘Well, I shall eat some of it now.’'
\z 
 
\ea \label{ŽH.46}
\textit{ewīč milowe gēɫnowe kināče wičkilekēšā} \\ 
\gll ew=īč mi-l-o=we ∅-gēɫn-o=we kināčē=e wičkile-ekē=šā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}lead\textsc{.prs}\textsc{-3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3pl} \\ 
\glt `She, i.e. the girl, went back [and] narrated the story [to her sister];'
\z 
 
\ea \label{ŽH.48}
\textit{māčo, je māseket keč nekerđebo řāře wāɫē!’} \\ 
\gll m-āč-o je mās-eke=t keč ne-kerđe b-o řā=ře wāɫē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} from yoghurt\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} crooked \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} sister\textsc{\textsc{.f}} \\ 
\glt `She (the older sister) had said [to her younger sister], ‘Sister, could the yoghurt be reduced’'
\z 
 
\ea \label{ŽH.52}
\textit{duē řoē čā bo} \\ 
\gll duē řo-ē čā ∅-b-o \\ 
 two day\textsc{.m}\textsc{-dir}\textsc{.pl} there \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (Hayas) was there for two days.'
\z 
 
\ea \label{ŽH.56}
\textit{kināčēčmā gerekene binyeymē were’} \\ 
\gll kināčē=č=mā gerek-e=ne bi-nye-ymē were \\ 
 girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{=add}\textsc{=1pl} necessary\textsc{-f}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \textsc{sbjv-}put\textsc{.prs} \textsc{1pl} front \\ 
\glt `We want the daughter; we would like to put her in front [for marriage]'
\z 
 
\ea \label{ŽH.57}
\textit{ā wextīyekey āyeč māčo, weɫāhī tāte, dāwāw ēme yoymā kero’} \\ 
\gll ā wext-ī-ekey āye=č m-āč-o weɫāhī tāte dāwā=w ēme yo-ī=mā ∅-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{\textsc{.f}}\textsc{.dir}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed father\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{1pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then she (the eldest daughter) said, ‘Indeed, father, he is asking for the hand of one of us.’'
\z 
 
\ea \label{ŽH.62}
\textit{īne e čē čenū āġeyš jengšā bīyen, be ādizī lūān} \\ 
\gll īne e=čē čenū āġe-ī=š jeng=šā bīye=n be ādizī lūā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=here with.ez\textsc{.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} quarrel\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} with desperation go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (Hayas) got into a quarrel with his master. He had left [town] in desperation'
\z 
 
\ea \label{ŽH.64}
\textit{qerār pāse bo e čē xānū ewēgē āđ čenū āđī suɫtān mehmū bē hečkes bē qāɫešā bī hāčkes ženē xāse bāro bē ewēge āđ bilo bo mēmānšā} \\ 
\gll qerār pāse ∅-b-o e=čē xān=ū ewēge-ē āđ čenū āđī suɫtān mehmū b-ē hečkes b-ē qāɫe=šā bī-∅ hāčkes ženī-ē xās-e b-ār-o b-ē ewēge āđ bi-l-o ∅-b-o mēmān=šā \\ 
 arrangement like \textsc{ind-}be\textsc{.prs}\textsc{-3sg} in=here chief\textsc{.m}\textsc{=ez}\textsc{.gen} there\textsc{-obl}\textsc{\textsc{.f}} 3sg\textsc{.dir}\textsc{.m} with.ez\textsc{.gen} 3sg\textsc{.obl}\textsc{.m} sul \textsc{pn} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} whoever be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} quarrel\textsc{\textsc{.f}}\textsc{=3pl} be\textsc{.pst}\textsc{-3sg}\textsc{.m} whoever woman\textsc{\textsc{.f}}\textsc{-indf} fine\textsc{-f} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} there \textsc{3sg}\textsc{.dir}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} guest\textsc{.m}\textsc{=3pl} \\ 
\glt `In the old times, the rule was that — the khan of there, whether it was Sultan Mahmud, with whom he (Hayas) had got into a quarrel, or anyone else— whoever got married and took a beautiful wife, he (the king) would go [and] become their guest.'
\z 
 
\ea \label{ŽH.65}
\textit{yewāšē āđīč mārošo mēwe kināčēw luhonī bo teřū čermew īnīšā} \\ 
\gll yewāšē āđ=īč m-ār-o=š=o m-ē=we kināčē=ū luhon-ī ∅-b-o teř=ū čerme=ū īnīšā \\ 
 then 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} fresh=and white=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `He (Hayas) brought her (i.e. his wife). She was from Luhon; she was young, white, and so on.'
\z 
 
\ea \label{ŽH.67}
\textit{āđīč māčo, čiřdēš peym’} \\ 
\gll āđ=īč m-āč-o ∅-čiř-dē=š pey=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}call\textsc{.prs}\textsc{-2pl}\textsc{=3sg} for\textsc{=1sg} \\ 
\glt `He (the Sultan) said, ‘Call him for me.’'
\z 
 
\ea \label{ŽH.72}
\textit{yewāšē ēđīč wēš wezore milo pey yāneyšā} \\ 
\gll yewāšē ēđ=īč wē=š wez-o=re mi-l-o pey yāne-ī=šā \\ 
 then 3sg.prox\textsc{.dir}\textsc{=add} \textsc{reflx}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \\ 
\glt `Then, he (the sultan) prepared himself [and] went towards their (i.e. Hayas’s) house.'
\z 
 
\ea \label{ŽH.76}
\textit{milone yāne xeyrāmāyš kero} \\ 
\gll mi-l-one yāne xeyrāmāy=š ∅-ker-o \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} house\textsc{.m}\textsc{.dir}.sg welcoming\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the sultan) went inside [and] she welcomed him.'
\z 
 
\ea \label{ŽH.78}
\textit{āđīč māčo, min desūrim pēsnen pey wēt āmānā} \\ 
\gll āđ=īč m-āč-o min desūr=im pēsne=n pey wē=t āmā=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg task/order\textsc{.m}\textsc{=1sg} like\_this\textsc{=cop}\textsc{.3sg}\textsc{.m} for \textsc{reflx}\textsc{=\textsc{2sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \\ 
\glt `He said, ‘My task is like this! I have come for you.’'
\z 
 
\ea \label{ŽH.81}
\textit{milo yerē dānē hēɫē māro} \\ 
\gll mi-l-o yerē dāne-ē hēɫe-ē m-ār-o \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} three \textsc{clf}\textsc{-dir}\textsc{.pl} egg\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `She went [and] brought three eggs.'
\z 
 
\ea \label{ŽH.82}
\textit{xo hāɫīt bo hēɫē yānē kāmēnē} \\ 
\gll xo hāɫī=t ∅-b-o hēɫe-ē yānē kām-ē=nē \\ 
 dis.ptc understood\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} egg\textsc{.m}\textsc{-dir}\textsc{.pl} that\_means which\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Do you understand what eggs are?'
\z 
 
\ea \label{ŽH.85}
\textit{dānēšā her čerme menowe kināčekē} \\ 
\gll dāne-ē=šā her čerme men-o=we kināčē-ekē \\ 
 one\textsc{-indf}\textsc{=3pl} just white remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} girl\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \\ 
\glt `[and] one remained white'
\z 
 
\ea \label{ŽH.86}
\textit{māčo be tūwāɫekeyšā, īne čēšen?’} \\ 
\gll m-āč-o be tūwāɫe-ekey=šā īne čēše=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by shell\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what \\ 
\glt `She pointed to the (egg)shells [and] said, ‘what are these? \footnote{Lit. What is this}''
\z 
 
\ea \label{ŽH.87}
\textit{māčo, īnē māčmēšā hēɫe’} \\ 
\gll m-āč-o īnē m-āč-mē=šā hēɫe \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.3pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3pl} egg\textsc{.m} \\ 
\glt `He (the sultan) said, ‘We call these eggs.’'
\z 
 
\ea \label{ŽH.88}
\textit{māčo, īne bīye čin řengī?’} \\ 
\gll m-āč-o īne bī-e čin řengī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.pst} 3sg\textsc{\textsc{.f}} how\_many colour \\ 
\glt `She said, ‘How many colours have they become?’'
\z 
 
\ea \label{ŽH.89}
\textit{māčo, āne bīyē yerē řengī} \\ 
\gll m-āč-o āne bī-ē yerē řengī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.pst} 3pl three colour \\ 
\glt `He said, ‘They have turned into three colours’'
\z 
 
\ea \label{ŽH.95}
\textit{māčo, īne her hēɫe bī’} \\ 
\gll m-āč-o īne her hēɫe bī-∅ \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} just egg\textsc{.m} be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `He said, ‘This one was just an egg.’'
\z 
 
\ea \label{ŽH.100}
\textit{āne řengw ji berīš ferq kero} \\ 
\gll āne řengī=ū ji ber-ī=š ferq ∅-ker-o \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} colour=and from outside\textsc{-obl}\textsc{.m}\textsc{=3sg} difference \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `It [is] just the colour. They differ in appearance.'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.104}
\textit{ānā dersē ʕālit dānē} \\ 
\gll āne=ā ders-ē ʕāl=it dā-ānē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} lesson\textsc{.m}\textsc{-indf} good\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{-1sg} \\ 
\glt `You gave me a good lesson.'
\z 
 
\ea \label{ŽH.108}
\textit{mēwe, melowe yāne} \\ 
\gll m-ē=we me-l-o=we yāne \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m}\textsc{.dir}.sg \\ 
\glt `He returned [but] he didn’t go back home.'
\z 
 
\ea \label{ŽH.109}
\textit{melowe yāne, duē řoē pā degāne bo} \\ 
\gll me-l-o=we yāne duē řo-ē p=ā degā=ne ∅-b-o \\ 
 \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m} two day\textsc{.m}\textsc{-dir}\textsc{.pl} at\textsc{=dem}.dist village\textsc{.obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He didn’t go back home. He was [wandering] in the village for two days. '
\z 
 
\ea \label{ŽH.110}
\textit{šūnū duē řoeyre, kināčekē kīyāno lāw suɫtānī māčo, ēme pīyākemā kīyāstiš meʕmūrīyet nīyā, nāmāno’} \\ 
\gll šūn=ū duē řo-e-ī=re kināčē-ekē ∅-kīyān-o lā=ū suɫtān-ī m-āč-o ēme pīyā-eke=mā kīyāst-∅=iš meʕmūrīyet nīye=ā ne-āmā=n=o \\ 
 after\textsc{=ez}\textsc{.gen} two\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} sul\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} send\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=3sg} mission\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `After the second day, the girl (his wife) sent (someone) to the Sultan to tell [him], ‘He (the sultan) sent my husband on a mission, [however] he is not around; He hasn’t come back [yet].’'
\z 
 
\ea \label{ŽH.111}
\textit{māčo, āmāno īnā’} \\ 
\gll m-āč-o āmā=n=o īne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (the sultan) said, ‘He has returned; there he is.’'
\z 
 
\ea \label{ŽH.112}
\textit{māčo, āđīč čiřoš māčo, čī nelānīwe?’} \\ 
\gll m-āč-o āđ=īč ∅-čiř-o=š m-āč-o čī ne-lā=nī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `He (the sultan) called him (Hayas) [and] said, ‘Why don’t you go back (to your house)?’'
\z 
 
\ea \label{ŽH.116}
\textit{āđīč wāt, ā šēre minnā} \\ 
\gll āđ=īč wāt ā šēr=e min=nā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{dem.dist} lion\textsc{.m}\textsc{=dem} 1sg\textsc{=cop}\textsc{.1sg} \\ 
\glt `He (the sultan) said, ‘I am that lion.'
\z 
 
\ea \label{RE.2}
\textit{ēme eger zemānē yā qeymīēmā yā zemānū wēmā ženīmā ārdēbo} \\ 
\gll ēme eger zemān-ē yā qeymī-ē=mā yā zemān=ū wē=mā ženī=mā ārdē b-o \\ 
 \textsc{1pl} if time\textsc{.m}\textsc{-indf} either the\_elderly\textsc{-indf}\textsc{=\textsc{1pl}} either time\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1pl}} woman\textsc{.dir}\textsc{.f}\textsc{=\textsc{1pl}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f} be\textsc{.prs}\textsc{-3sg} \\ 
\glt ` Once, in the time of our elders or in our time, if one of us married,'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.12}
\textit{hewrāmīgerī řoē lūān kāriš pey kerden jīyāw xizmetīū} \\ 
\gll hewrāmīgerī řo-ē lūā=n kār=iš pey kerde=n jīyāw xizmet-ī=ū \\ 
 hewrami\_custom\textsc{.m} day\textsc{.m}\textsc{-dir}\textsc{.pl} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} task\textsc{.m}\textsc{=3sg} to do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} in\_place\_of service\textsc{.m}\textsc{-obl}\textsc{.m}=and \\ 
\glt `According to Hawrami [customs], he [the boy] would go and work for him (the father-in-law) for service.’'
\z 
 
\ea \label{RE.13}
\textit{řoē lūān nīštēnēre sere gāwēšā wārdenū} \\ 
\gll řo-ē lūā=n nīštē=nē=re sere gāw-ē=šā wārde=n=ū \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} sit\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} head\textsc{.m} bull\textsc{.m}\textsc{-indf}\textsc{=3pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They (the bridegroom’s) family would sit and eat a cow.'
\z 
 
\ea \label{RE.16}
\textit{ey pīyā ētir emīnū yāne bīyenū xizmetkār bīyen} \\ 
\gll ey pīyā ētir emīn=ū yāne bīye=n=ū xizmetkār bīye=n \\ 
 \textsc{dem.prox} man\textsc{.m} \textsc{disc.ptcl} confidant\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{.dir}.sg be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez.gen} servant\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was this man who was the family’s confidant. He was a servant.'
\z 
 
\ea \label{RE.18}
\textit{firmānbirđār bīyen, tā sāhībū ī kināčē wāčo, xeber bido wāčo, fiɫāne kār} \\ 
\gll firmānbirđār bīye=n tā sāhīb=ū ī kināčē ∅-wāč-o xeber bi-de-o ∅-wāč-o fiɫān=e kār \\ 
 obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until owner\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} news\textsc{.dir}\textsc{.m} \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} such\_and\_such=ez\textsc{.cmpd} task\textsc{.m} \\ 
\glt `He was obedient. The girl’s father\footnote{Lit. owner.} would say [to him] and ask him, ‘[Do] such-and-such tasks;'
\z 
 
\ea \label{RE.20}
\textit{lue kārekem pey kere} \\ 
\gll ∅-lu-e kār-eke=m pey ∅-ker-e \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} task\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} for \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Go do [this] task for me.'
\z 
 
\ea \label{RE.21}
\textit{ā wextīekey firmānbirđār bīyen tā ženekēš berdēne} \\ 
\gll ā wext-ī-ekey firmānbirđār bīye=n tā ženī-ekē=š berdē=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} take\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `He was at their service until the girl married.'
\z 
 
\ea \label{RE.24}
\textit{xeplezeřātīnē nīštēnēre nānešānē xeplēšā nīyēnēre} \\ 
\gll xeplezeřātīnē nīštē=nē=re nānešān-ē xeple-ē=šā nīyē=nē=re \\ 
 \textsc{pn} sit\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} basket\textsc{-indf} bread\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} \\ 
\glt `xaplazaratene,\footnote{A bread made of corn flour.} they would sit [together] and cook a basket of xaple.\footnote{A bread made of corn flour.}'
\z 
 
\ea \label{RE.26}
\textit{zemānēč ženekēš berdēne ētir čo dumā bīyēnē tēkeɫ} \\ 
\gll zemān-ē=č ženī-ekē=š berdē=ne ētir č=o dumā bīyē=nē tēkeɫ \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=add} woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} take\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} \textsc{disc.ptcl} in=\textsc{dem.dist} afterwards be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} mixed \\ 
\glt `When the brides were taken, they (the families) would become one.'
\z 
 
\ea \label{RE.31}
\textit{kināčekē qisē kero} \\ 
\gll kināčē-ekē qisē-ē ∅-ker-o \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} talk\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The girl talks [to the boy]'
\z 
 
\ea \label{RE.32}
\textit{kuřekeyč qisē kero} \\ 
\gll kuř-ekey=č qisē-ē ∅-ker-o \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} talk\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The boy, too, talks [to the girl]'
\z 
 
\ea \label{RE.49}
\textit{min kināčēt dēne kuřū fiɫāne kesī} \\ 
\gll min kināčē=t dē=ne kuř=ū fiɫān=e kes-ī \\ 
 \textsc{1sg} daughter\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}} son/boy\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and say to the lord], ‘I have given my\footnote{Lit. your.} daughter to the son of such-and-such person.'
\z 
 
\ea \label{RE.51}
\textit{ewīč wāten bā bizānmē} \\ 
\gll ew=īč wāte=n bā bi-zān-mē \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{hort} \textsc{sbjv-}know\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `He (the lord) would say, ‘Let me see.’\footnote{Lit. Let us know.}'
\z 
 
\ea \label{RE.52}
\textit{pīney bāyē māsīč bero} \\ 
\gll p=īney bāye-ē mās=īč ∅-ber-o \\ 
 for=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} pot\textsc{.m}\textsc{-dir}\textsc{.pl} yoghurt\textsc{.m}\textsc{=add} \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `For the reason that he (the girl’s father) would take a bowl of yoghurt [to him] too.'
\z 
 
\ea \label{RE.61}
\textit{sūrāne ānen} \\ 
\gll sūrāne āne=n \\ 
 \textsc{pn} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Surana was that.'
\z 
 
\ea \label{RE.67}
\textit{lūēnmē bāyē māsēmā berdēnē bē ewū} \\ 
\gll lūē=nmē bāye-ē mās-ē=mā berdē=nē bē ew=ū \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} pot\textsc{.m}\textsc{-dir}\textsc{.pl} yoghurt\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1pl}} take\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} without 3sg\textsc{.dir}\textsc{.m}=and \\ 
\glt `We would go [to the noblemen and] take them bowls of yoghurt free [of charge]'
\z 
 
\ea \label{RE.71}
\textit{āđē āmēnē tenekē girđ yo tenekēšā berden} \\ 
\gll āđē āmē=nē teneke-ē girđ yo teneke-ē=šā berde=n \\ 
 3pl\textsc{.dir} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} sack\textsc{.m}\textsc{-indf} all one\textsc{.m} sack\textsc{.m}\textsc{-indf}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the noblemen) would come [and] each take a sack [of wheat]'
\z 
 
\ea \label{ŽE.1}
\textit{dey čēgeyč qeyīm ʕāɫifšā kennenū} \\ 
\gll dey čēge=īč qeyīm ʕāɫif=šā kenne=n=ū \\ 
 \textsc{disc.ptcl} here\textsc{=add} old\_time grass\textsc{.dir}\textsc{.m}\textsc{=3pl} mow\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `In the past, they (people) mowed grass.'
\z 
 
\ea \label{ŽE.2}
\textit{ħeywānšā wey kerdenū} \\ 
\gll ħeywān=šā wey kerde=n=ū \\ 
 animal\textsc{.dir}\textsc{.m}\textsc{=3pl} raising do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They raised animals.'
\z 
 
\ea \label{ŽE.3}
\textit{hewārēšā wiškinēnē} \\ 
\gll hewār-ē=šā wiškinē=nē \\ 
 summar\_habitat\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3pl} ransack\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They scoured the summer habitats [searching for food etc.].'
\z 
 
\ea \label{ŽE.6}
\textit{āwīš nebīyēne} \\ 
\gll āwī=š ne-bīyē=ne \\ 
 water\textsc{.dir}\textsc{.f}\textsc{=3sg} \textsc{neg-}be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `It (the village) had no water.'
\z 
 
\ea \label{ŽE.9}
\textit{īse luēnē tā penjā čil šewē zim[sān], čil šewē pāyīz šīyen jā kočšā kerdeno} \\ 
\gll īse luē=nē tā penjā čil šewe-ē zimsān čil šewe-ē pāyīz šīye=n jā koč=šā kerde=n=o \\ 
 now go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} until fifty forty night\textsc{.f}\textsc{-dir}\textsc{.pl} winter\textsc{.m} forty night\textsc{.f}\textsc{-dir}\textsc{.pl} autumn go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} afterwards migration\textsc{.dir}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `They used to go [to the summer habitat] at this time of the year. They would migrate back [to the village] around 40 or 50 days after autumn.'
\z 
 
\ea \label{ŽE.12}
\textit{girđkes ʕāɫfiš kennenū} \\ 
\gll girđkes ʕāɫif=iš kenne=n=ū \\ 
 everyone grass\textsc{.dir}\textsc{.m}\textsc{=3sg} mow\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Everyone mowed grass'
\z 
 
\ea \label{ŽE.13}
\textit{ʕāɫfiš pētenū tā kočšā kerden tā ʕāɫifšā ārdeno} \\ 
\gll ʕāɫif=iš pēte=n=ū tā koč=šā kerde=n tā ʕāɫif=šā ārde=n=o \\ 
 grass\textsc{.dir}\textsc{.m}\textsc{=3sg} fold\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} until migration\textsc{.dir}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until grass\textsc{.dir}\textsc{.m}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `[and] made grass into stacks until when they returned [to the village]; until when they brought back grass '
\z 
 
\ea \label{ŽE.27}
\textit{de neferē serū jāmoɫē čāštīšā nīyēnere} \\ 
\gll de nefer-ē ser=ū jāmoɫē-ē čāštī=šā nīyē=ne=re \\ 
 ten person\textsc{.m}\textsc{-dir}\textsc{.pl} on\textsc{\textsc{=ez.gen}} small\_pot\textsc{-indf} meal\textsc{.obl}\textsc{.f}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f}\textsc{=\textsc{povb}} \\ 
\glt ` They would cook meals in a single small pot for ten persons.'
\z 
 
\ea \label{ŽE.29}
\textit{čemčekene girđ nobenobey čāštīšā wārdēne} \\ 
\gll čemče-eke=ne girđ nobenobey čāštī=šā wārdē=ne \\ 
 spoon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} all one\_by\_one meal\textsc{.obl}\textsc{.f}\textsc{=3pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `They would take turns to eat their meal using the spoon.'
\z 
 
\ea \label{ŽE.32}
\textit{yo sēfēš bīyēneū} \\ 
\gll yo sēfē=š bīyē=ne=ū \\ 
 one\textsc{.m} potato\textsc{.dir}\textsc{.f}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.f}=and \\ 
\glt `One had potatoes.'
\z 
 
\ea \label{ŽE.33}
\textit{luēnē eweɫū wextī lūēnē hēzmīšā biřyēnē} \\ 
\gll luē=nē eweɫ=ū wext-ī lūē=nē hēzmī=šā biřyē=nē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} first\textsc{\textsc{=ez.gen}} time\textsc{.m}\textsc{-obl}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} firewood\textsc{.dir}\textsc{.pl}\textsc{=3pl} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Early in the morning, they would go and chop wood.'
\z 
 
\ea \label{ŽE.34}
\textit{be derze hēzmīšā ārdēnē} \\ 
\gll be derze hēzmī=šā ārdē=nē \\ 
 by shoulder\textsc{.m} firewood\textsc{.dir}\textsc{.pl}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `They fetched firewood on their shoulders. '
\z 
 
\ea \label{ŽE.35}
\textit{be hesere hēzmīšā ārdēnē pey zimsānī} \\ 
\gll be hesere hēzmī=šā ārdē=nē pey zimsān-ī \\ 
 by mule\textsc{.f} firewood\textsc{.dir}\textsc{.pl}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} for winter\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `They fetched firewood for the winter on mules. '
\z 
 
\ea \label{ŽE.37}
\textit{hēzmī ētir zimsān āmān hurēsēnē yerē āwīr, āwīre kučēšā nīyēnēre bī kipē} \\ 
\gll hēzmī ētir zimsān āmā=n hur-ēsē=nē yerē āwīr āwīr=e kuč-ē=šā nīyē=nē=re bī-∅ kipē \\ 
 firewood\textsc{.f} \textsc{disc}.\textsc{ptcl} winter\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pvb-}rise\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} three fire\textsc{.m} fire\textsc{.m}\textsc{=dem} little\textsc{-dir}\textsc{.pl}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} be\textsc{.pst}\textsc{-3sg}\textsc{.m} sealed \\ 
\glt `Firewood …  Then when it was winter they would rise and start a small fire [and it would] damp down.'
\z 
 
\ea \label{ŽE.39}
\textit{čāštekēšā kerdēne be āwīrgākēū} \\ 
\gll čāštī-ekē=šā kerdē=ne be āwīrgā-ekē=ū \\ 
 meal\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} to fireplace\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}=and \\ 
\glt `They made the food on the hearth.'
\z 
 
\ea \label{ŽE.41}
\textit{desečirāwē bīyen desečirāweke čeqnānšā keɫekere jīyāw řošināyīšā} \\ 
\gll desečirāw-ē bīye=n desečirāw-eke čeqnā=n=šā keɫeke=re jīyāw řošināyī=šā \\ 
 hand\_lump\textsc{-dir}\textsc{.pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} hand\_lump\textsc{-def}\textsc{.dir}\textsc{.m} install\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} corner\textsc{=\textsc{post}} in\_place\_of light\textsc{.m}\textsc{=3pl} \\ 
\glt `There were hand-held oil lamps. They would install the hand-held oil lamps in a corner, in place of lights.'
\z 
 
\ea \label{ŽE.42}
\textit{čirāwe desīšā, čirāwekešā weš kerden bilā pey yānew pīr šelīyārī pey čīwī bārānšo} \\ 
\gll čirāw=e desī=šā čirāw-eke=šā weš kerde=n bi-l-ā pey yāne=ū pīr šelīyār-ī pey čīw-ī b-ār-ān=š=o \\ 
 lamp\textsc{.m}\textsc{=ez}\textsc{.cmpd} hand\textsc{=3pl} lamp\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} for house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} for thing\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `They lit the lamps to go the Pir Shaliyar’s shrine\footnote{Lit. house.} [and] to fetch things.'
\z 
 
\ea \label{ŽE.53}
\textit{īnēmā wārdēne} \\ 
\gll īnē=mā wārdē=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=1pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We have eaten this [meal]. '
\z 
 
\ea \label{ŽE.55}
\textit{pāse zāwlēšā wey kerdē} \\ 
\gll pāse zāwlē=šā wey kerd-ē \\ 
 like child\textsc{.dir}\textsc{.pl}\textsc{=3pl} raising do\textsc{.pst} 3pl \\ 
\glt `They raised children in this way. '
\z 
 
\ea \label{ŽE.62}
\textit{luēnā ħeywānim bīyen} \\ 
\gll luē=nā ħeywān=im bīye=n \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.\textsc{1sg}} animal\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I went … I had animals.'
\z 
 
\ea \label{ŽE.71}
\textit{pāne wešɫe wešɫeyī yānekoɫēšā kerden} \\ 
\gll p=āne wešɫe wešɫeyī yānekoɫē=šā kerde=n \\ 
 at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} well happy\_manner moving\_house\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They left the house in a happy manner.'
\z 
 
\ea \label{ŽE.76}
\textit{wāčē, weɫā bā kināčekē donmē’} \\ 
\gll wāč-ē weɫā bā kināčē-ekē ∅-don-mē \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} indeed \textsc{hort} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}talk\_to\textsc{-1pl} \\ 
\glt `She (the mother) would say, ‘Let us talk to the girl.’'
\z 
 
\ea \label{ŽE.78}
\textit{wāčē, dey lue duē yerē pīyē bēnē’} \\ 
\gll wāč-ē dey ∅-lu-e duē yerē pīyē b-ē-nē \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc.}\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} two three man\textsc{.dir}\textsc{.pl} \textsc{sbjv-}come\textsc{.prs}\textsc{-3pl} \\ 
\glt `She (the mother) would say, ‘Go! Let two, three men come.'
\z 
 
\ea \label{ŽE.79}
\textit{duē pīyē hījbīš kerā} \\ 
\gll duē pīyē hījbī=š ∅-ker-ā \\ 
 two man\textsc{.dir}\textsc{.pl} betrothal\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \\ 
\glt `Two men acted as intermediaries.'
\z 
 
\ea \label{ŽE.82}
\textit{wāčē, dey lue yānew wēt nīšere tā duē māngē ter yerē māngē ter} \\ 
\gll wāč-ē dey ∅-lu-e yāne=ū wē=t ∅-nīš-e=re tā duē mānge-ē ter yerē mānge-ē ter \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc}.\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} until two month\textsc{.f}\textsc{-dir}\textsc{.pl} other three month\textsc{.f}\textsc{-dir}\textsc{.pl} other \\ 
\glt `They (the family of the girl) would say [to the boy], ‘Go back home [and] wait\footnote{Lit. sit.} for two, three months.'
\z 
 
\ea \label{ŽM.1}
\textit{heštā sāɫē ʕemirmā} \\ 
\gll heštā sāɫe-ē ʕemr=m=ā \\ 
 eighty year\textsc{.f}\textsc{-dir}\textsc{.pl} age\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I am eighty years old.'
\z 
 
\ea \label{ŽM.2}
\textit{tātem duē ženīš ārdēnē} \\ 
\gll tāte=m duē ženī=š ārdē=nē \\ 
 father\textsc{.m}\textsc{=\textsc{1sg}} two woman\textsc{.dir}\textsc{.pl}\textsc{=3sg} bring\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `My father married two women.'
\z 
 
\ea \label{ŽM.7}
\textit{je kārēz biřyeywe pī heštā sāɫe, īse ħewt sāɫē} \\ 
\gll je kārēz biřyey=we p=ī heštā sāɫe=e īse ħewt sāɫe-ē \\ 
 from subterranean\_canal\textsc{.m} cut\textsc{.inf}\textsc{=compl} at=\textsc{dem.prox} eighty year\textsc{.f}\textsc{=dem} now seven year\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `Ever since [the time I was] digging canals until now that I am eighty years old [I have been working]'
\z 
 
\ea \label{ŽM.8}
\textit{īse ħewt sāɫēč āmānāwe serū īney dūkāndārnā} \\ 
\gll īse ħewt sāɫe-ē=č āmā=nā=we ser=ū īney dūkāndār=nā \\ 
 now seven year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} shopkeeper\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{ŽM.11}
\textit{čemēm korē bīyēnēw} \\ 
\gll čem-ē=m kor-ē bīyē=nē=ū \\ 
 eye\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} blind\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `My eyesight has deteriorated.\footnote{My eyes have gotten blind.} '
\z 
 
\ea \label{ŽM.12}
\textit{īne dāstānū minā} \\ 
\gll īne dāstāne=ū min=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{=ez.gen} \textsc{1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is my story.'
\z 
 
\ea \label{ŽM.14}
\textit{dizīm nekerđenū hīzīm nekerdenū girđkār nebīyenā} \\ 
\gll dizī=m ne-kerđe=n=ū hīzī=m ne-kerde=n=ū girđkār ne-bīye=nā \\ 
 theft\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and adultery\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and know\_all \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I have not committed theft or adultery. I was not a know-all person.'
\z 
 
\ea \label{ŽM.16}
\textit{dārāw īse elʕān hāɫēhāzirī dārāw penj kuřānāw yerē kināčā, heštū wēmū ženē īne de neferē} \\ 
\gll dārā=w īse elʕān hāɫēhāzir-ī dārā=w penj kuř-ā=nā=ū yerē kināčē-ā hešt=ū wē=m=ū ženī-ē īne de nefer-ē \\ 
 owner\textsc{.m}\textsc{=ez.gen} now now at\_the\_moment\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{=ez.gen} five son/boy\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1sg}}=and three daughter\textsc{.f}\textsc{-obl}\textsc{.pl} eight=and \textsc{reflx}\textsc{=\textsc{1sg}}=and woman\textsc{.f}\textsc{-indf} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} ten person\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `I have, at the moment, five sons and three daughters, [which makes] eight, [plus] myself and a wife, [making us] ten people.'
\z 
 
\ea \label{ŽM.20}
\textit{min zemānē zemāniš bīyen ženīm ārdēne girđū kārēw nebīyey bīyen} \\ 
\gll min zemān-ē zemān=iš bīye=n ženī=m ārdē=ne girđ=ū kār-ē=ū nebīyey bīye=n \\ 
 \textsc{1sg} time\textsc{.m}\textsc{-indf} time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} woman\textsc{.f}\textsc{=\textsc{1sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} all\textsc{=ez.gen} task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=ez.gen} poverty\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As for me, once I got married, there was poverty all around.'
\z 
 
\ea \label{ŽM.23}
\textit{luēnmē ferz tārān kārmā kerđen šiš māngē} \\ 
\gll luē=nmē ferz tārān kār=mā kerđe=n šiš mānge-ē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} guess\textsc{.m} \textsc{pn} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go, say, to Tehran and work for six months.'
\z 
 
\ea \label{ŽM.24}
\textit{luēnmē kirmān kārmā kerđen šiš māngē} \\ 
\gll luē=nmē kirmān kār=mā kerđe=n šiš mānge-ē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kerman and work for six months.'
\z 
 
\ea \label{ŽM.25}
\textit{jezīrew kīšīne kārmā kerđen šiš māngē} \\ 
\gll jezīre=w kīš-ī=ne kār=mā kerđe=n šiš mānge-ē \\ 
 island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kish Island and work for six months.'
\z 
 
\ea \label{ŽM.26}
\textit{ā wextīyekey ħewt no, yānze zāroɫēm bīyēnē} \\ 
\gll ā wext-ī-ekey ħewt no yānze zāroɫe-ē=m bīyē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} seven nine eleven child\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `I had seven or nine or eleven children.'
\z 
 
\ea \label{ŽM.27}
\textit{ī yānze zāroɫē ečē āsēnēm jīyā milū ī ženīme milū nebīyeywe} \\ 
\gll ī yānze zāroɫe-ē e=čē āsē=nē=m jīyā mil=ū ī ženī=m=e mil=ū nebīyey=we \\ 
 \textsc{dem.prox} eleven child\textsc{-dir}\textsc{.pl} in=here leave\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} separate on\textsc{=ez.gen} \textsc{dem.prox} woman\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} on\textsc{=ez.gen} poverty\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Because of destitution, [I had to] leave these eleven children to my wife.'
\z 
 
\ea \label{ŽM.28}
\textit{min ečāgene kārīger bīyenā šiš māngē xeberēm nezānān} \\ 
\gll min e=čāgene kārīger bīye=nā šiš mānge-ē xeber-ē=m ne-zānā=n \\ 
 \textsc{1sg} in=there labourer\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} six month\textsc{.f}\textsc{-dir}\textsc{.pl} news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` I was a worker there. I was unaware of them\footnote{Lit. I didn’t know their news.}  (my children) for six months,'
\z 
 
\ea \label{ŽM.30}
\textit{ā wextīyekey řūtē bīyēnē qūtē bīyēnē bīyenšā nebīyenšā xeberēm nezānēnē} \\ 
\gll ā wext-ī-ekey řūt-ē bīyē=nē qūt-ē bīyē=nē bīye=n=šā ne-bīye=n=šā xeber-ē=m ne-zānē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} naked\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} bare\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} news\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `[I was unaware if] they were without clothes; if they had [enough]; if they didn’t have [enough]. I didn't know their news.'
\z 
 
\ea \label{ŽM.33}
\textit{šesū šiš sāʕetē ēme be māšīn jā luēnmē binderʕebāsū} \\ 
\gll šes=ū šiš sāʕet-ē ēme be māšīn jā luē=nmē binderʕebās=ū \\ 
 sixty\textsc{=ez.gen} six hour\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} by car\textsc{.m} afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and \\ 
\glt `It took us sixty-three hours to reach Bandar Abbas by car.'
\z 
 
\ea \label{ŽM.35}
\textit{šiš māngē wēmā čā kārmā kerđen} \\ 
\gll šiš mānge-ē wē=mā čā kār=mā kerđe=n \\ 
 six month\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{reflx}\textsc{=\textsc{1pl}} there work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I worked there for [periods of] six months.'
\z 
 
\ea \label{ŽM.40}
\textit{ēme čāgene sāxtmānē sāznēnmē} \\ 
\gll ēme čāgene sāxtmān-ē sāznē=nmē \\ 
 \textsc{1pl} there building\textsc{.m}\textsc{-dir}\textsc{.pl} build\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `We built buildings there.'
\z 
 
\ea \label{ŽM.41}
\textit{kuweyt duē sāʕetē beynšā čenū kuweytī} \\ 
\gll kuweyt duē sāʕet-ē beyn=šā čenū kuweyt-ī \\ 
 \textsc{pn} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} between\textsc{=3pl} with\textsc{\textsc{.ez}.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Kuwait was two hours away from there.'
\z 
 
\ea \label{ŽM.42}
\textit{kuweyt, āđīč wēš kēšwerēwā kuweyt} \\ 
\gll kuweyt āđ=īč wē=š kēšwer-ēw=ā kuweyt \\ 
 \textsc{pn} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{reflx}\textsc{=3sg} country\textsc{-indf}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \\ 
\glt `Kuwait is a country on its own.'
\z 
 
\ea \label{ŽM.45}
\textit{ēme jezīre čenū ā jezīrey čeneš bēnmē duē sāʕetē beyniš bē} \\ 
\gll ēme jezīre čenū ā jezīre-ī čene=š b-ēn-mē duē sāʕet-ē beyn=iš b-ē \\ 
 \textsc{1pl} island with\textsc{\textsc{.ez}.gen} \textsc{dem.dist} island\textsc{-obl}\textsc{.m} at\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1pl}} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} distance\textsc{.m}\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `It (Kuwait) was two hours away from the island we were in.'
\z 
 
\ea \label{ŽM.49}
\textit{řoē sī penj timenē šesū yerē sāɫē čēweɫī kerū ētir meyānū sāɫū činē bē} \\ 
\gll řo-ē sī penj timen-ē šes=ū yerē sāɫe-ē čēweɫī ∅-ker-ū ētir me-zān-ū sāɫe-ū činē b-ē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} sixty\textsc{=ez.gen} three year\textsc{.f}\textsc{-dir}\textsc{.pl} ago \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc}.\textsc{ptcl} \textsc{neg-}know\textsc{.prs}\textsc{-\textsc{1sg}} year\textsc{.f}\textsc{-\textsc{1sg}} how\_many\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `A daily salary of thirty-five tomans, it was sixty-three years ago. I don’t know which year it was/'
\z 
 
\ea \label{ŽM.55}
\textit{ētir īse weʕzū hewrāmānī tā menteqē terē kučē xāsterā} \\ 
\gll ētir īse weʕze=ū hewrāmān-ī tā menteqe-ē ter-ē kuč-ē xās-ter=ā \\ 
 \textsc{disc}.\textsc{ptcl} now situation\textsc{.f}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} until region\textsc{-dir}\textsc{.pl} other\textsc{-dir}\textsc{.pl} little\textsc{-indf} fine\textsc{-cmpr}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The situation in Hawraman is a bit better than in other regions.'
\z 
 
\ea \label{ŽM.57}
\textit{ā wextīyekey ēme īne zinđigīš pene kermē be nebīyey} \\ 
\gll ā wext-ī-ekey ēme īne zinđigī=š pene ∅-ker-mē be nebīyey \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} life\textsc{.m}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1pl}} with poverty\textsc{.m} \\ 
\glt `Well, we live in poverty.'
\z 
 
\ea \label{ŽM.61}
\textit{herkes tāwo bilo koɫē bāro koɫē milwēnē koɫē duē milwēnē} \\ 
\gll herkes ∅-tāw-o bi-l-o koɫ-ē b-ār-o koɫ-ē milwēn-ē koɫ-ē duē milwēn-ē \\ 
 anyone \textsc{ind-}can\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} load\textsc{.m}\textsc{-indf} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3sg} load\textsc{.m}\textsc{-indf} million\textsc{.m}\textsc{-indf} load\textsc{.m}\textsc{-indf} two million\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `Anyone who can go [to the mountains] as a porter can earn one or two million (tomans) per load.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
