\chapter{All 55 examples of exist in these texts are found in the following sentences:}

\ea \label{ZB.29}
\textit{kučē qenū čāyū ā wextīekey, ne nān hen, ne čāy hen, ne čīw hen, ħīč} \\ 
\gll kuč-ē qen=ū čāy=ū ā wext-ī-ekey ne nān hen-∅ ne čāy hen-∅ ne čīw hen-∅ ħīč \\ 
 little\textsc{-indf} sugar\_cube=and tea\textsc{.m}=and \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m} nor tea\textsc{.m} exist\textsc{-3sg}\textsc{.m} nor thing\textsc{.m} exist\textsc{-3sg}\textsc{.m} nothing \\ 
\glt `[There was] some sugar and tea, well then, there was neither bread, nor water, nor anything else, [there was] nothing [left]'
\z 
 
\ea \label{ZB.60}
\textit{welē īse be qurʕān ne šēx menen, ne meɫā menen, ne kesē hen be weš dāđiš berī lā} \\ 
\gll welē īse be qurʕān ne šēx mene=n ne meɫā mene=n ne kes-ē hen-∅ be weš dāđ=iš ∅-ber-ī lā \\ 
 but now to \textsc{pn} neither sheikh\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} nor mullah\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} nor person\textsc{.m}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} to good yell\textsc{.m}\textsc{=3sg} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} at\_the\_place\_of \\ 
\glt `However, nowadays, I swear on the Quran, neither a [true] sheikh nor a [true] mullah remains, nor is there anyone one can take counsel with.'
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{ZQ.55}
\textit{fermāwo, dey āne ētir deseɫāt xwāy, pey her čē meyliš bē xwāy deseɫātē čānēš henē} \\ 
\gll ∅-fermāw-o dey āne ētir deseɫāt xwā-ī pey herčē meyl=iš b-ē xwā-ī deseɫāt-ē čāne-ē=š hen-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} power\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} to whatever willingness\textsc{=3sg} \textsc{sbjv-}come.prs\textsc{.3sg} God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m-dir.pl} such\textsc{-pl}\textsc{=3sg} \textsc{exist}\textsc{-3pl} \\ 
\glt ` The Shah said, ‘Well, then, this is in God’s hands. Whatever God wishes, he has the power [to exert]'
\z 
 
\ea \label{HB.2}
\textit{nānewēkēw šēh ʕelādīnī seʕbē hurmēzāwe xeber miđā māčā, xeber deydē be hezretū šēxī, be xwā xāneqāne ne nān hen, ne hārdī henē} \\ 
\gll nānewē-ekē=ū šēh ʕelādīn-ī seʕbe-ē hur-mi-ēz ā=we xeber mi-đe-ā mi-āč-ā xeber ∅-de-ydē be hezret=ū šēx-ī be xwā xāneqā=ne=ne nān hen-∅=ne hārdī hen-ē \\ 
 baker\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{.pl}=and sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} morning\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs} \textsc{dem.dist}\textsc{=compl} news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} news\textsc{.m} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} to his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}}=neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m}=nor flour\textsc{\textsc{.f}} exist\textsc{-3pl} \\ 
\glt `Sheikh Aladin’s bakers got up in the morning and passed on the news, saying, ‘Pass on the news to his highness the Sheikh. By God, there is neither bread nor flour in the monastery. '
\z 
 
\ea \label{HB.3}
\textit{ne nān henū ne hārdī henē} \\ 
\gll ne nān hen-∅=ū ne hārdī hen-ē \\ 
 no bread\textsc{.m} \textsc{exist}\textsc{-3sg}\textsc{.m}=and no flour\textsc{\textsc{.f}} \textsc{exist}\textsc{-3pl} \\ 
\glt `There is no bread. There is no flour;'
\z 
 
\ea \label{HB.4}
\textit{ħīčē nīyāw} \\ 
\gll ħīč-ē nīye=ā=ū \\ 
 nothing\textsc{-indf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `There is nothing [left]'
\z 
 
\ea \label{HB.29}
\textit{miđyo be ħīčē telklīfiš nīyā} \\ 
\gll mi-đye-o be ħīč-ē telklīf=iš nīye=ā \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} by nothing\textsc{-indf} option\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He realised that he had absolutely no options,'
\z 
 
\ea \label{HB.91}
\textit{bizāne neferē ter hen be zūwānū herī hāɫīš bo bejge to!} \\ 
\gll bi-zān-e nefer-ē ter hen-∅ be zūwān=ū her-ī hāɫī=š ∅-b-o bejge to \\ 
 \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} person\textsc{.m}\textsc{-dir}\textsc{.pl} other exist\textsc{-3sg}\textsc{.m} by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} understood\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `see if there is a person, apart from you, who can understand the language of donkeys!'
\z 
 
\ea \label{PM.17}
\textit{ānā ēme wēmā mehelēmā hen tūl} \\ 
\gll āne=ā ēme wē=mā mehele-ē=mā hen-∅ tūl \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{1pl} \textsc{reflx}\textsc{=1pl} neighbourhood\textsc{.m}\textsc{-indf}\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} long \\ 
\glt `There, I, myself, own a big neighbourhood.’\footnote{Lit. We, ourselves, have a big neighbourhood.}'
\z 
 
\ea \label{PM.20}
\textit{fermāwebē, fermāwo lodē e čā muɫkū ēme wēmā, herčēmā hen čāne ā firmānde wēš zāno heqū ēme kāmā} \\ 
\gll fermāwe b-ē ∅-fermāw-o ∅-lo-dē e=čā muɫk=ū ēme wē=mā herčē=mā hen-∅ č=ā=ne ā firmānde wē=š ∅-zān-o heq=ū ēme kām=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} in=there property\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{reflx}\textsc{=1pl} whatever\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} leader\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} right\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} which\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go! In that property of mine, whatever we possess there, your boss knows which portion [of land] is mine.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{DG.24}
\textit{řoē je řoā māčā pīyāē hen nāmēš šā ʕelādīnārēū} \\ 
\gll řo-ē je řo-ā m-āč-ā pīyā-ē hen-∅ nāmē=š šā ʕelādīn=ā=rē=ū \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} man\textsc{.m}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} name\textsc{\textsc{.f}}\textsc{=3sg} king\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `Once, people said, ‘There is a man called Aladin. '
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.40}
\textit{ewīč nimē māčo, nexeyr nimewnew mitmānim nīyā penet.} \\ 
\gll ew=yč ni-mi-ē mi-āč-o nexeyr ni-mi-e-wne=ū mitmān=im nīye=ā pene=t \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-1sg}=and trust\textsc{.m}\textsc{=1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} to\textsc{=\textsc{2sg}} \\ 
\glt ` Sibhan Agha\footnote{Lit. He.} did not go [to Jamsher Shah], saying, ‘No, I won’t come [to you]; I don’t trust you.’'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.86}
\textit{māčā xeɫkīč, ī pīyā šimeš pey persdē xo yānēw e čāneš nīyāw} \\ 
\gll m-āč-ā xeɫk=īč ī pīyā šime=š pey ∅-pers-dē xo yāne-ē=ū e=čāne=š nīye=ā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} people\textsc{.m}\textsc{=add} dem.prox man\textsc{.m} \textsc{2pl}\textsc{=3sg} for \textsc{ind-}ask\textsc{.prs}-\textsc{2pl} dis.ptc house\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} in=like\_that\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The people said, ‘This man whom you ask for, he does not have a proper house.'
\z 
 
\ea \label{ZP.87}
\textit{ženīū yāneš nīyārē īnīšā} \\ 
\gll ženī=ū yāne=š nīye=ā=rē īnīšā \\ 
 woman\textsc{.f}=and house\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He has no wife, nor a house, and so on.'
\z 
 
\ea \label{ZP.105}
\textit{min čīwēwim nīyārē lāyiqū ī pādšāzāye bo} \\ 
\gll min čīw-ēw=im nīye=ā=rē lāyiq=ū ī pādšāzāye ∅-b-o \\ 
 \textsc{1sg} thing\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} deserved\textsc{\textsc{=ez.gen}} dem.prox princess\textsc{.obl}\textsc{.f} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `I have nothing which would be up to the standards of this princess.'
\z 
 
\ea \label{ZP.117}
\textit{īđīč hīč minīč hīčim nīyārēw} \\ 
\gll īđ=īč hīč min=īč hīč=im nīye=ā=rē=ū \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} nothing \textsc{1sg}\textsc{=add} nothing\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `She [has] nothing, nor do I.'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ŽP.20}
\textit{yāgē hen māčāš nehāɫū pīr šelīyārā—} \\ 
\gll yāgē-ē hen-∅ m-āč-ā=š nehāɫ=ū pīr šelīyār-ā \\ 
 place\textsc{.f}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \\ 
\glt `There is a place called the valley of Pir Shaliyar.'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.163}
\textit{eger to ī kināčēte berdeū ī pīyā her kes hen} \\ 
\gll eger to ī kināčē=t=e berd-e=ū ī pīyā her kes hen-∅ \\ 
 if \textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} take\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m} emph person\textsc{.m} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `If you take the girl, [no matter] who this man (the doctor) is,'
\z 
 
\ea \label{ŽP.195}
\textit{mē ehā tāylē hen} \\ 
\gll m-ē ehā tāyle-ē hen-∅ \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{intj} tree\_name\textsc{-indf} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `He (the demon) kept walking. Have you noticed there is a tree?'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.242}
\textit{qey mekero, mādām desūr e pēsen pādšā řāzīe nīā ī ženī bilowe ī kināčē bilowe qey mekero!’} \\ 
\gll qey me-ker-o mādām desūr e=pēse=n pādšā řāzīe nīye=ā ī ženī bi-l-o=we ī kināčē bi-l-o=we qey me-ker-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} as\_long\_as task/order\textsc{.m} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} satisfied \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} dem.prox woman\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `There is no problem. Now that things have developed this way, [and] the king is not happy for the girl to return, it’s all right! [She can stay here]’'
\z 
 
\ea \label{BP.13}
\textit{to hošit nīyā, tāteyt miqdārēš hošā—} \\ 
\gll to hoš=it nīye=ā tāte-ī=t miqdārē=š hoš=ā \\ 
 \textsc{2sg} memory\textsc{.m}\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} a\_little\textsc{=3sg} memory\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `—you are not aware [of that period], [but] your father remembers it a bit—'
\z 
 
\ea \label{BP.52}
\textit{dey ēšew herkes mēmāniš hen, ēšew herkes mēmāniš hen, mēmānekeš šewē witē, sereš biřo} \\ 
\gll dey ēšew herkes mēmān=iš hen-∅ ēšew herkes mēmān=iš hen-∅ mēmān-eke=š šewe-ē wit-ē sere=š ∅-biř-o \\ 
 \textsc{disc.ptcl} tonight everyone guest\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} tonight everyone guest\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} guest\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} night.f\textsc{-obl}.f sleep\textsc{.pst}\textsc{-3pl} head\textsc{.m}\textsc{=3sg} \textsc{sbjv-}cut.prs\textsc{-3sg} \\ 
\glt `‘Tonight, whoever has guests, when the guest sleeps at night, he shall decapitate him.'
\z 
 
\ea \label{BP.76}
\textit{miđyā īne be mišārī sereš biřyen nāmā nāmā nīyā} \\ 
\gll mi-đye-ā īne be mišār-ī sere=š biřye=n ne-āmā-∅ ne-āmā-∅ nīye=ā \\ 
 \textsc{ind-}look.prs\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} to hand\_saw\textsc{.m}\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=3sg} cut\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They noticed that the one who had decapitated his guests with a hand saw hadn’t come to the mosque. He hadn’t arrived. He wasn’t there.'
\z 
 
\ea \label{BP.87}
\textit{jārēne jimdēre mexɫūq hečē henē mizgīne bā bilimdē pey ʕērāqī} \\ 
\gll jār-ē=ne ∅-jim-dē=re mexɫūq hečē hen-ē mizgī=ne ∅-b-ā bi-l-imdē pey ʕērāq-ī \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} people\textsc{.m} no\_matter\_how\_many exist\textsc{-3pl} mosque\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}/\textsc{2pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They all set off; every person who was in the mosque [said], ‘Let’s go to Iraq.'
\z 
 
\ea \label{BP.178}
\textit{āhā māčā īnā čū dīmne lāw beynū zāwerīnew tewēɫēne ħemey ġeybī hen} \\ 
\gll āhā m-āč-ā īnā-∅ č=ū dīm=ne lā=ū beyn=ū zāwer-ī=ne=ū tewēɫē=ne ħeme=ī ġeybī hen-∅ \\ 
 \textsc{intj} \textsc{ind-}say.prs\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{.3sg} side\textsc{.m}\textsc{=\textsc{post}} at\_the\_place\_of\textsc{=ez.gen} among\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{=\textsc{post}} \textsc{pn}=ez.\textsc{attr} invisible exist\textsc{-3sg}\textsc{.m} \\ 
\glt `It is said that, on this side of the border, there is the shrine of Hama the invisible between the villages of Zawar and Tawela.'
\z 
 
\ea \label{BP.184}
\textit{fermāwo, min kučē mūsū welē wesīyetēm hen kerū:} \\ 
\gll ∅-fermāw-o min kuč-ē m-ūs-ū welē wesīyet-ē=m hen-∅ ∅-ker-ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{1sg} little\textsc{-indf} \textsc{ind-}sleep.prs\textsc{-\textsc{1sg}} but will.\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} exist\textsc{-3sg}\textsc{.m} \textsc{ind-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, I will sleep a bit. However, I have a [last] will to give:'
\z 
 
\ea \label{BP.190}
\textit{miđyā her hāyš mebowe nīyā dērāw,} \\ 
\gll mi-đye-ā her hāy=š me-b-o=we nīye=ā dēr=ā=ū \\ 
 \textsc{ind-}look.prs\textsc{-3pl} just awareness\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} late\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They noticed that he didn’t wake up, he wasn’t there [and] it was getting late.'
\z 
 
\ea \label{BP.193}
\textit{deng mekero nīyā} \\ 
\gll deng me-ker-o nīye=ā \\ 
 noise\textsc{.m} \textsc{neg-}do.prs\textsc{-3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He didn’t make a sound. He wasn’t there.'
\z 
 
\ea \label{BP.194}
\textit{wextē ʕebē berzā kerāwe, miđyā kes nīyā čērišne} \\ 
\gll wext-ē ʕebā-ē berz=ā ∅-ker-ā=we mi-đye-ā kes nīye=ā čēr=iš=ne \\ 
 time\textsc{.m}\textsc{-indf} robe.f\textsc{-obl}.f high\textsc{=compl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=compl} \textsc{ind-}look.prs\textsc{-3pl} person\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} under\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `When they lifted the robe, they saw that no one was under it.'
\z 
 
\ea \label{ŠJ.12}
\textit{tifengiš henū īneš henū āneš henū!} \\ 
\gll tifeng=iš hen-∅=ū īne=š hen-∅=ū āne=š hen-∅=ū \\ 
 gun\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist-3sg.m}=and \\ 
\glt `He has a gun, he has this, and he has that!'
\z 
 
\ea \label{ŠJ.74}
\textit{fermāwo lodē lāš wāčdē kāriš be min nīyā} \\ 
\gll ∅-fermāw-o ∅-lo-dē lā=š wāč-dē kār=iš be min nīye=ā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} at\_the\_place\_of\textsc{=3sg} say.prs-\textsc{imp-}\textsc{2pl} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go to him [and] say, ‘His business is not with me.'
\z 
 
\ea \label{ŠJ.76}
\textit{hīč kārēš be min nīyā} \\ 
\gll hīč kār-ē=š be min nīye=ā \\ 
 nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He has no business with me.’'
\z 
 
\ea \label{ŠJ.77}
\textit{milā māčā qurbān šē ʕusmānī fermāwā kāriš be min nīyā} \\ 
\gll mi-l-ā m-āč-ā qurbān šē ʕusmān-ī fermāwā kār=iš be min nīye=ā \\ 
 \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] said [to Jafir Sān], ‘Sir, the sheikh said, “He has no business with me.'
\z 
 
\ea \label{ŠJ.86}
\textit{fermāwo qurbān hermāne be desū min nīyenew} \\ 
\gll ∅-fermāw-o qurbān hermāne be des=ū min nīye=ne=ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} work\textsc{.f} to hand\textsc{.m}\textsc{=ez.gen} \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `He (the sheikh) said, ‘Sir, the task [of ruling] is not in my hands;'
\z 
 
\ea \label{ŠJ.87}
\textit{be desū to nīyenew} \\ 
\gll be des=ū to nīye=ne=ū \\ 
 to hand\textsc{.m}\textsc{=ez.gen} \textsc{2sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `It is not in your hands either.'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{KŠ.76}
\textit{īne wēš xeberiš nīyāw sewāđiš nīyā} \\ 
\gll īne wē=š xeber=iš nīye=ā=ū sewāđ=iš nīye=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{reflx}\textsc{=3sg} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and literacy\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He wasn’t aware [of the death plot] that \_\_ he was illiterate\_\_'
\z 
 
\ea \label{ŽH.6}
\textit{wiɫātēwe termā hen ew dīmew ēme, pēsew zāwerīū tewēɫey} \\ 
\gll wiɫāt-ēwe ter=mā hen-∅ ew dīm=ew ēme pēse=ū zāwer-ī=ū tewēɫe-ī \\ 
 country\textsc{.m}\textsc{-indf} other\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} \textsc{dem.dist} side\textsc{.m}\textsc{=\textsc{post}} \textsc{1pl} like\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `There is another country on the other side of us; something like Zawar (in Iran) and Tawela (in Iraq) village.'
\z 
 
\ea \label{ŽH.110}
\textit{šūnū duē řoeyre, kināčekē kīyāno lāw suɫtānī māčo, ēme pīyākemā kīyāstiš meʕmūrīyet nīyā, nāmāno’} \\ 
\gll šūn=ū duē řo-e-ī=re kināčē-ekē ∅-kīyān-o lā=ū suɫtān-ī m-āč-o ēme pīyā-eke=mā kīyāst-∅=iš meʕmūrīyet nīye=ā ne-āmā=n=o \\ 
 after\textsc{=ez}\textsc{.gen} two\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} sul\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} send\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=3sg} mission\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `After the second day, the girl (his wife) sent (someone) to the Sultan to tell [him], ‘He (the sultan) sent my husband on a mission, [however] he is not around; He hasn’t come back [yet].’'
\z 
 
\ea \label{RE.34}
\textit{jūwābēš nīyā xeberiš nīyā tā tāte} \\ 
\gll jūwāb-ē=š nīye=ā xeber=iš nīye=ā tā tāte \\ 
 response\textsc{.m}\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} until father\textsc{.m} \\ 
\glt `[Then] there is no news until [the girl] says, ‘Father!’'
\z 
 
\ea \label{ŽE.83}
\textit{tā qewmekāmā wiš kermē qewmekā wiš kermē bizānmē ī qewmīmā řezāšāne yā řezāšā nīyene} \\ 
\gll tā qewm-ekā=mā wiš ∅-ker-mē qewm-ekā wiš ∅-ker-mē bi-zān-mē ī qewm-ī=mā řezā=šā=ne yā řezā=šā nīye=ne \\ 
 until relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=1pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} agreement\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} or agreement\textsc{.f}\textsc{=3pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We will inform our relatives. We will inform our relatives to see if they agree [to the  marriage] or not.'
\z 
 
\ea \label{ŽM.21}
\textit{ī mentēqew ēme nebīyeyš hen nebīyey} \\ 
\gll ī mentēqe=ū ēme nebīyey=š hen-∅ nebīyey \\ 
 \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} poverty\textsc{.m} \\ 
\glt `Our region is in poverty, [and] shortage.'
\z 
 
\ea \label{ŽM.56}
\textit{čēweɫ nebīyey bē ētir īseyč xū nīyā} \\ 
\gll čēweɫ nebīyey b-ē ētir īse=yč xū nīye=ā \\ 
 previously poverty\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{disc.ptcl} now\textsc{=add} \textsc{disc.ptcl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the past, there was poverty. But even nowadays, there is poverty.'
\z 
 
\ea \label{ŽM.59}
\textit{enāne ēme wārīyātēmā nīyā} \\ 
\gll enā=ne ēme wārīyāt-ē=mā nīye=ā \\ 
 otherwise\textsc{=\textsc{post}} \textsc{1pl} income\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `because we have no income.'
\z 
 
\ea \label{ŽM.60}
\textit{wārīyātmā nīyā kāsbīmā nīyā kārmā nīyā} \\ 
\gll wārīyāt=mā nīye=ā kāsbī=mā nīye=ā kār=mā nīye=ā \\ 
 income\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} business\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} work\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We have no income, no business, no jobs.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
