\chapter{All 549 examples of gen in these texts are found in the following sentences:}

\ea \label{ZB.1}
\textit{řoē je řoā yoe mē yoyč mē xizmetū šā ʕelādīnī, her ā šā ʕelādīne} \\ 
\gll řo-ē je řo-ā yo-e m-ē yo=yč m-ē xizmet=ū šā ʕelādīn-ī her ā šā ʕelādīn=e \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m}\textsc{-f} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} one\textsc{.m}\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} emph \textsc{dem.dist} king\textsc{.m} \textsc{pn}\textsc{=dem} \\ 
\glt `Once,\footnote{Lit. One day of days.} a woman and a man\footnote{Lit. One (f.) came, one (m.) came.} came to the service of šā ʕalādīn, the same šā ʕalādīn.'
\z 
 
\ea \label{ZB.7}
\textit{ginmē dewrū hāminī wehārī temāmīyo} \\ 
\gll ∅-gin-mē dewr=ū hāmin-ī wehār-ī temāmīye-o \\ 
 \textsc{ind-}fall\textsc{.prs}\textsc{-1pl} around\textsc{\textsc{=ez.gen}} summer\textsc{.m}\textsc{-obl}\textsc{.m} spring\textsc{.m}\textsc{-obl}\textsc{.m} finish\textsc{.prs}\textsc{-3sg} \\ 
\glt `It was\footnote{Lit. We fell} near the end of summer, [or] spring.'
\z 
 
\ea \label{ZB.11}
\textit{be ʕemrū xwā hewrē gino řū we āsmānī} \\ 
\gll be ʕemre=ū xwā hewr-ē ∅-gin-o řū=we āsmān-ī \\ 
 to order\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m} cloud\textsc{-indf} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} face=to sky\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `By the order of God, a cloud covered the sky.\footnote{Lit. a cloud fell on the sky.}'
\z 
 
\ea \label{ZB.12}
\textit{tejrebeš bo ʕemriš dewrū ṣewū wīs sāɫāne bo wīs sāɫē sī sāɫē pā dewrne} \\ 
\gll tejrebe=š ∅-b-o ʕemr=iš dewr=ū ṣew=ū wīs sāɫe-ā=ne ∅-b-o wīs sāɫe-ē sī sāɫe-ē p=ā dewr=ne \\ 
 experience\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} age\textsc{.m}\textsc{=3sg} around\textsc{\textsc{=ez.gen}} hundred=and twenty year\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} twenty year\textsc{.f}\textsc{-dir}\textsc{.pl} thirty year\textsc{.f}\textsc{-dir}\textsc{.pl} at=\textsc{dem.dist} around\textsc{=\textsc{post}} \\ 
\glt `He was experienced. He was around 120 years old, about 120 or 130,'
\z 
 
\ea \label{ZB.16}
\textit{ewēč māčo, wārānū čēšī} \\ 
\gll ewē=č m-āč-o wārān=ū čēš-ī \\ 
 3sg\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} rain\textsc{\textsc{=ez.gen}} what\textsc{-obl}\textsc{.m} \\ 
\glt `She said, ‘What rain?!’'
\z 
 
\ea \label{ZB.24}
\textit{šewē be ʕemreū xwāy ženēč nīšore duē zāroɫē wīno her ā šewē} \\ 
\gll šewe-ē be ʕemre=ū xwā-ī ženī-ē=č ∅-nīš-o=re duē zāroɫe-ē ∅-wīn-o her ā šewe-ē \\ 
 night\textsc{.f}\textsc{-obl}\textsc{.f} to order\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m} woman\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=add} \textsc{ind-}sit\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} two child\textsc{-dir}\textsc{.pl} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} emph \textsc{dem.dist} night\textsc{.f}\textsc{-obl}\textsc{.f} \\ 
\glt `During that very night, by the order of God, the wife gave birth to two babies.\footnote{Lit. She sat down [and] saw two babies.} '
\z 
 
\ea \label{ZB.25}
\textit{wizošāre dilū āwē} \\ 
\gll ∅-wiz-o=šā=re dil=ū āwē \\ 
 \textsc{ind-}throw\textsc{.prs}\textsc{-3sg=3pl}\textsc{=\textsc{povb}} inside\textsc{\textsc{=ez.gen}} water\textsc{.obl}\textsc{.f} \\ 
\glt `She delivered them in the water [which had amassed in the tent]'
\z 
 
\ea \label{ZB.28}
\textit{āwī luēnērew serū āne hārdīšā bīyēnēū} \\ 
\gll āwī luē=nē=re=ū ser=ū āne hārdī=šā bīyē=nē=ū \\ 
 water\textsc{.f} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}}=and on\textsc{\textsc{=ez.gen}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} flour\textsc{.f}\textsc{=3pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `The water had gone up to the level of the flour [sacks] that they had.'
\z 
 
\ea \label{ZB.36}
\textit{bero mizīoš qiřoɫū derē, hīnēwe meřū dārē} \\ 
\gll ∅-ber-o mi-zye-o=š qiřoɫ=ū dere-ē hīn-ēwe meře=ū dār-ē \\ 
 \textsc{ind-}take\textsc{.prs}\textsc{-3sg} \textsc{ind-}leave\textsc{.prs}\textsc{-3sg}\textsc{=3sg} hollow\textsc{.m}\textsc{\textsc{=ez.gen}} valley\textsc{.m}\textsc{-indf} \textsc{fill}\textsc{-indf} cave\textsc{.f}\textsc{\textsc{=ez.gen}} tree\textsc{.m}\textsc{-indf} \\ 
\glt `He took him (i.e., one of the baby boys), [and] put him in a hole in a tree, a thingummy, in the cave of a tree.'
\z 
 
\ea \label{ZB.37}
\textit{īne min īnem žinīyen lāw, fermāwištū šēxīn, šā ʕelādīnīwe} \\ 
\gll īne min īne=m žinīye=n lā=ū fermāwišt=ū šēx-ī=n šā ʕelādīn-ī=we \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{1sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} speech\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have heard this tale from —it is the Sheikh’s words— Shah Aladin.'
\z 
 
\ea \label{ZB.38}
\textit{ā wextīekey minyoš qiřoɫū beřū dārēweū} \\ 
\gll ā wext-ī-ekey mi-nye-o=š qiřoɫ=ū beřū dār-ēwe=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}put\textsc{.prs}\textsc{-3sg}\textsc{=3sg} hollow\textsc{.m}\textsc{\textsc{=ez.gen}} oak\textsc{.m} tree\textsc{.m}\textsc{-indf}=and \\ 
\glt `Anyhow, he put him in a tree hollow, in an oak tree.'
\z 
 
\ea \label{ZB.40}
\textit{milā ew kuřekey yoyšā berā ānešā zilterū ʕālter bo āđī berā} \\ 
\gll mi-l-ā ew kuř-ekey yo-ī=šā ∅-ber-ā āne=šā zil-ter=ū ʕāl-ter ∅-b-o āđī ∅-ber-ā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3pl} big\textsc{-cmpr}\textsc{\textsc{=ez.gen}} good\textsc{-cmpr} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} 3sg\textsc{.obl}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} \\ 
\glt `They went away [and took] that son. They took one of them (i.e., the boys), the one who was bigger and healthier; they took him.'
\z 
 
\ea \label{ZB.41}
\textit{ewī minyā qiřoɫū dārēwe} \\ 
\gll ewī mi-nye-ā qiřoɫ=ū dār-ēwe \\ 
 3sg\textsc{.obl}\textsc{.m} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} hollow\textsc{.m}\textsc{\textsc{=ez.gen}} tree\textsc{.m}\textsc{-indf} \\ 
\glt `They left him (i.e., the other one) in the hole in a tree.'
\z 
 
\ea \label{ZB.42}
\textit{čī mede xwāwinū ʕālemīyānī pāse īrāyešā bizēw menēne jīāy sāyqekewe} \\ 
\gll č=ī mede xwāwin=ū ʕālemīyān-ī pāse īrāye=š=ā bize-ēw menē=ne jīāy sāyqe-eke=we \\ 
 at=\textsc{dem.prox} time owner\textsc{\textsc{=ez.gen}} worlds\textsc{.m}\textsc{-obl}\textsc{.m} like volition\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} goat\textsc{.f}\textsc{-indf} remain\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} behind lightning\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `The will of the lord of all worlds was such that a goat was accidentally left behind [healthy] from the lightning [that caused the flood]'
\z 
 
\ea \label{ZB.45}
\textit{pey minyo serū nerekeyū, čērhur zāroɫeke šot wero} \\ 
\gll pey mi-nye-o ser=ū nere-ekey=ū čēr=hur zāroɫe-eke šot ∅-wer-o \\ 
 for \textsc{ind-}put\textsc{.prs}\textsc{-3sg} on\textsc{\textsc{=ez.gen}} oak\_tree\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{\textsc{=ez.gen}} under\textsc{=\textsc{post}} child\textsc{-def}\textsc{.dir}\textsc{.m} milk\textsc{.m} \textsc{ind-}eat\textsc{.prs}\textsc{-3sg} \\ 
\glt `[and] put its feet on the trunk of the oak tree [and] the baby drank [its] milk from below.'
\z 
 
\ea \label{ZB.48}
\textit{wextē miđyo kābrā qiřoɫū nerekeyne sere ber mēw milo řū zemīnīre} \\ 
\gll wext-ē mi-đye-o kābrā qiřoɫ=ū nere-ekey=ne sere ber m-ē=ū mi-l-o řū zemīn-ī=re \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} man hollow\textsc{.m}\textsc{\textsc{=ez.gen}} oat\_tree\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} head\textsc{.m} out \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{\textsc{=ez.gen}} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} face land\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `When he looked [at the tree], he noticed that someone had put his head out of the hollow of the oak tree and pulled himself back again.'
\z 
 
\ea \label{ZB.49}
\textit{milo wextē miđyo zāroleke bīyen peřū qiřoɫekey} \\ 
\gll mi-l-o wext-ē mi-đye-o zāroɫe-eke bīye=n peř=ū qiřoɫ-ekey \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} child\textsc{-def}\textsc{.dir}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{\textsc{=ez.gen}} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `The father went [closer and] noticed that the baby boy [had grown up so much that] he had filled the tree hollow.'
\z 
 
\ea \label{ZB.52}
\textit{hinū dilū qiřoɫū nerekey šiš kīloē čīđī zīyā bo} \\ 
\gll hin=ū dil=ū qiřoɫ=ū nere-ekey šiš kīlo-ē č=īđī zīyā ∅-b-o \\ 
 ez.pron\textsc{\textsc{=ez.gen}} inside\textsc{\textsc{=ez.gen}} hollow\textsc{.m}\textsc{\textsc{=ez.gen}} oat\_tree\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} six kilo\textsc{.m}\textsc{-indf} in=3sg\textsc{.obl}\textsc{.m} more \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The one raised in the tree hollow weighed more than six kilos.'
\z 
 
\ea \label{ZB.53}
\textit{kārū xwāy e čānen} \\ 
\gll kār=ū xwā-ī e=čāne=n \\ 
 task\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m} in=like\_that\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `God’s work is like this.'
\z 
 
\ea \label{ZB.58}
\textit{īne dāstānū gewregerīū ā mešeyxāne} \\ 
\gll īne dāstāne=ū gewregerī=ū ā mešeyxā=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{\textsc{=ez.gen}} greatness\textsc{\textsc{=ez.gen}} \textsc{dem.dist} sheikh\textsc{.obl}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `This is the story of the greatness of the sheikhs.'
\z 
 
\ea \label{ZQ.1}
\textit{ʕerzū xizmetīt kerū} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū \\ 
 petition\textsc{.m}\textsc{=ez.gen} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `Let me tell you!\footnote{Lit. May a petition be for your service.}'
\z 
 
\ea \label{ZQ.2}
\textit{hetā min žinīyenim pīyewe āmā xizmetū šā ʕelādīnī} \\ 
\gll hetā min žinīye=n=im pīye-we āmā xizmet=ū šā ʕelādīn-ī \\ 
 even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} man\textsc{.m} come\textsc{.pst}\textsc{.3sg} service\textsc{.m}\textsc{=ez.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have even heard that a man came to the sevice of Shah Aladin. '
\z 
 
\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.9}
\textit{vātiš, qurbān luāymē pār yā činnē sāɫē čēweɫter luāymē pey sārāɫū ērānī, koy sān} \\ 
\gll vāt=iš qurbān luā-ymē pār yā činnē sāɫe-ē čēweɫter luā-ymē pey sārāɫ=ū ērān-ī ko=y sān \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} go\textsc{.pst} \textsc{1pl} last\_year either some\textsc{.pl} year.f\textsc{-indf} ago go\textsc{.pst} \textsc{1pl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} where=ez.\textsc{attr} \textsc{pn} \\ 
\glt `He said, ‘Sir, we went, say, last year or a few years ago, we went to the Saral region in Iran, to the San mountain.'
\z 
 
\ea \label{ZQ.10}
\textit{čowe sārāɫū ērānīwe gēɫāymēwe} \\ 
\gll č=o=we sārāɫ=ū ērān-ī=we gēɫā-ymē=we \\ 
 in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=compl} turn\textsc{.pst} \textsc{1pl}\textsc{=compl} \\ 
\glt `On the way back, we returned from the Saral region of Iran.'
\z 
 
\ea \label{ZQ.11}
\textit{wextū ī jimāymārā āmāymēwe āmāymē kelēwe} \\ 
\gll wext=ū ī jimāy=mā=rā āmā-ymē=we āmā-īmē kel-ēwe \\ 
 time\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} move\textsc{.inf}\textsc{=\textsc{1pl}}\textsc{=\textsc{povb}} come\textsc{.pst} \textsc{1pl}\textsc{=compl} come\textsc{.pst} \textsc{1pl} mountain\textsc{-indf} \\ 
\glt `On our return (journey), we went\footnote{Lit. came.}  to a mountain. '
\z 
 
\ea \label{ZQ.13}
\textit{ženīem bē ženekēmīč bārū hemɫīe bē} \\ 
\gll ženīe=m b-ē ženī-ekē=m=īč bār=ū hemɫī-e b-ē \\ 
 woman.f\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs\textsc{.3sg} woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}}\textsc{=add} load\textsc{.m}\textsc{=ez.gen} pregnant\textsc{-f} \textsc{sbjv-}come.prs\textsc{.3sg} \\ 
\glt `I had a wife. My wife was pregnant.'
\z 
 
\ea \label{ZQ.16}
\textit{duē, be gīrmā wārd čenū ženīū čenū heywānīū dewɫetīū bē āwīū nāřihet e pāwe} \\ 
\gll duē be gīr=mā wārd čenū ženī=ū čenū heywān-ī=ū dewɫet-ī=ū bē āwī=ū nāřihet e=pāwe \\ 
 two by hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} woman.f=and with\textsc{.ez.gen} animal\textsc{.m}\textsc{-obl}\textsc{.m}=and animal\textsc{-obl}\textsc{.m}=and come\textsc{.imp}\textsc{.\textsc{2sg}} water.f=and sad in=there \\ 
\glt `I was stuck there with my wife, livestock, a lack of water, and a lack of comfort. '
\z 
 
\ea \label{ZQ.18}
\textit{āđmā tehwēɫ gērtū} \\ 
\gll āđ=mā tehwēɫ gērt=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=\textsc{1pl}} delivery\textsc{.m} grab\textsc{.pst}\textsc{=ez.gen} \\ 
\glt `We got hold of him.'
\z 
 
\ea \label{ZQ.21}
\textit{ēmeyč gīrmā wārd čenū her wī} \\ 
\gll ēme=yč gīr=mā wārd čenū her wē-ī \\ 
 \textsc{1pl}\textsc{=add} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} just two\textsc{-obl}\textsc{.m} \\ 
\glt `We were stuck with both babies.'
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{ZQ.27}
\textit{luāymēwe pey germīyānū ʕērāqī} \\ 
\gll luā-ymē=we pey germīyān=ū ʕērāq-ī \\ 
 go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We went back to the Garmiyan region of Iraq.'
\z 
 
\ea \label{ZQ.30}
\textit{dimāw šiš māngē fewtiš kerd lūā} \\ 
\gll dimā=ū šiš mānge-ē fewt=iš kerd-∅ lūā \\ 
 afterwards\textsc{=ez.gen} six month.f\textsc{-dir}\textsc{.pl} death\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{-3sg}\textsc{.m} go\textsc{.pst} \\ 
\glt `After six months, he passed away.'
\z 
 
\ea \label{ZQ.31}
\textit{ʕemrū xwāyš kerdū luā} \\ 
\gll ʕemre=ū kerd=ū luā \\ 
 order.f\textsc{=ez.gen} do\textsc{.pst}=and go\textsc{.pst}\textsc{.3sg} \\ 
\glt `He passed away.\footnote{Lit. He did the command of God.} '
\z 
 
\ea \label{ZQ.33}
\textit{vātiš, qurbān wehārē wāwey luāymēwe pey sārāɫū ērānī} \\ 
\gll vāt=iš qurbān wehār-ē wāwey luā-ymē=we pey sārāɫ=ū ērān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} spring\textsc{.m}\textsc{-indf} again go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘Sir, [The next] spring, again, we went back to the Saral region of Iran.'
\z 
 
\ea \label{ZQ.36}
\textit{luānē nizīkū qiřoɫekey bīyāwe} \\ 
\gll luā-ānē nizīk=ū qiřoɫ-ekey bī-ā=we \\ 
 go\textsc{.pst}\textsc{-\textsc{1sg}} close=\textsc{ez.gen} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} be\textsc{.pst}\textsc{-3pl}\textsc{=compl} \\ 
\glt `I went [and] got closer to the tree hollow.'
\z 
 
\ea \label{ZQ.38}
\textit{nizīk bīyāwe temāšem kerd wextē miđyew kuřeke bīyen peřū qiřoɫēwe, qiřoɫekey} \\ 
\gll nizīk bī-ā=we temāše=m kerd wext-ē mi-đye-ū kuř-eke bīye=n peř=ū qiřoɫ-ēwe qiřoɫ-ekey \\ 
 close be\textsc{.pst}\textsc{-3pl}\textsc{=compl} looking\textsc{=\textsc{1sg}} do\textsc{.pst} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{=ez.gen} hollow\textsc{.m}\textsc{-indf} hollow\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I got closer and looked inside. I noticed that the boy [had grown up so much that he] had filled a tree hollow, i.e. the tree hollow. '
\z 
 
\ea \label{ZQ.39}
\textit{wextē luānē werew temāšem kerd miđyew be ʕemrew xwāy e čī qiřolī| ēnnew čīčew ženē pīfē peydā bīyenē| ʕeyn pīfē} \\ 
\gll wext-ē luā-ānē wer=ew temāše=m kerd mi-đye-ū be ʕemre=ū xwā-ī ečī qiřoɫ-ī ēnne=w čīče=ū ženī-ē pīfē-ē peydā bīyē=nē ʕeyn pīfē \\ 
 time\textsc{.m}\textsc{-indf} go\textsc{.pst}\textsc{-\textsc{1sg}} front\textsc{.m}\textsc{=\textsc{post}} looking\textsc{=\textsc{1sg}} do\textsc{.pst} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} by order.f\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} in.\textsc{dem.prox} hollow\textsc{.m}\textsc{-obl}\textsc{.m} as\_much\_as\textsc{=ez.gen} nipple\textsc{=ez.gen} woman.f\textsc{-obl}.f burr.f\textsc{-dir}\textsc{.pl} visible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like balloon.f \\ 
\glt `When I got closer and looked inside, I noticed that, by the order of God, some burrs the size of a woman's nipple had appeared in the tree hollow, like balloons.'
\z 
 
\ea \label{ZQ.40}
\textit{popew pīfekēwe ēnnew memkew ženīe} \\ 
\gll pope=w pīfē-ekē=we ēnne=w memke=ū ženīe \\ 
 top\textsc{=ez.gen} burr.f\textsc{-def}\textsc{.obl}.f\textsc{=compl} as\_much\_as\textsc{=ez.gen} breast\textsc{.m}\textsc{=ez.gen} woman.f \\ 
\glt `The top of the burr was like a woman’s breast. '
\z 
 
\ea \label{ZQ.43}
\textit{popew pīfekēwe memkēweš weš kerden ʕeynū memkew ženī} \\ 
\gll pope=w pīfē-ekē=we memke-ēwe=š weš kerde=n ʕeyn=ū memke=ū ženī \\ 
 top\textsc{=ez.gen} burr.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{post}} breast\textsc{.m}\textsc{-indf}\textsc{=3sg} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} like\textsc{=ez.gen} breast\textsc{.m}\textsc{=ez.gen} woman.f \\ 
\glt `he had made a breast from the top of the burr, just like a woman’s breast.'
\z 
 
\ea \label{ZQ.45}
\textit{wextē miđyew kuř bīyen peřū qiřoɫē} \\ 
\gll wext-ē mi-đye-ū kuř bīye=n peř=ū qiřoɫ-ē \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} son/boy\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} full\textsc{=ez.gen} hollow\textsc{.m}\textsc{-indf} \\ 
\glt `When I looked, the boy had [grown up so much that he had] filled a tree hollow.'
\z 
 
\ea \label{ZQ.46}
\textit{wātim, beydē temāšew lutfū xwāy čenīnā’} \\ 
\gll wāt=im b-e-ydē temāše=ū lutf=ū xwā-ī čenīn=ā \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs-\textsc{2pl} looking\textsc{=ez.gen} grace\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} how\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I said [to the people], ‘Come and see, God’s grace is like this!'
\z 
 
\ea \label{ZQ.48}
\textit{wātim, beydē dilū āweđānīekey} \\ 
\gll wāt=im b-e-ydē dil=ū āweđānī-ekey \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs-\textsc{2pl} inside\textsc{=ez.gen} oasis\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I said, ‘Come to the meadow’'
\z 
 
\ea \label{ZQ.50}
\textit{wātim, beydē dilū āweđānīekey} \\ 
\gll wāt=im b-e-ydē dil=ū āweđānī-ekey \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} \textsc{sbjv-}come.prs-\textsc{2pl} inside\textsc{=ez.gen} oasis\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `I said, ‘By God, the son has grown up to the size of the tree hollow.’'
\z 
 
\ea \label{ZQ.53}
\textit{kuřmā berārdū berdmā čenū wēmāw wešīyē ʕejāyibū diɫwešīyē ʕejāyib} \\ 
\gll kuř=mā ber-ārd-∅=ū berd-∅=mā čenū wē=mā=ū wešī-ē ʕejāyib=ū diɫwešī-ē ʕejāyib \\ 
 son/boy\textsc{.m}\textsc{=\textsc{1pl}} out- bring\textsc{.pst}\textsc{-3sg}\textsc{.m}=and take\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=\textsc{1pl}} with\textsc{.ez.gen}=and happiness\textsc{.m}\textsc{-indf} unbelievable=and joy\textsc{.m}\textsc{-indf} unbelievable \\ 
\glt `We took out the son, and took him with us. [We were full of] unbelievable happiness; [full of] an unbelievable sense of joy. '
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{ZQ.56}
\textit{girđčē girđčē desū xwāy sehlārī} \\ 
\gll girđčē girđčē des=ū xwā-ī sehl=ā=rī \\ 
 everything everything hand\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Everything, everything, is easy for God.'
\z 
 

\ea \label{HB.1}
\textit{zemānē hezretū šēh ʕelādīnī šēh ʕelādīnī mertebeš ʕālī bo} \\ 
\gll zemān-ē hezret=ū šēh ʕelādīn-ī šēh ʕelādīn-ī mertebe=š ʕālī ∅-b-o \\ 
 time\textsc{.m}\textsc{-indf} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} rank\textsc{.m}\textsc{=3sg} eminent \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `In the time of Sheikh Aladin—Sheikh Aladin was eminent.'
\z 
 
\ea \label{HB.2}
\textit{nānewēkēw šēh ʕelādīnī seʕbē hurmēzāwe xeber miđā māčā, xeber deydē be hezretū šēxī, be xwā xāneqāne ne nān hen, ne hārdī henē} \\ 
\gll nānewē-ekē=ū šēh ʕelādīn-ī seʕbe-ē hur-mi-ēz ā=we xeber mi-đe-ā mi-āč-ā xeber ∅-de-ydē be hezret=ū šēx-ī be xwā xāneqā=ne=ne nān hen-∅=ne hārdī hen-ē \\ 
 baker\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{.pl}=and sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} morning\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs} \textsc{dem.dist}\textsc{=compl} news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} news\textsc{.m} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} to his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}}=neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m}=nor flour\textsc{\textsc{.f}} exist\textsc{-3pl} \\ 
\glt `Sheikh Aladin’s bakers got up in the morning and passed on the news, saying, ‘Pass on the news to his highness the Sheikh. By God, there is neither bread nor flour in the monastery. '
\z 
 
\ea \label{HB.5}
\textit{girdū her čē xeɫk īnā xāneqāne gird āwrān’} \\ 
\gll gird=ū herčē xeɫk īnā-∅ xāneqā=ne gird āwrā=n \\ 
 all\textsc{=ez}\textsc{.gen} every people\textsc{.m} deic\textsc{-3sg}\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}} all hungry\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Every person in the monastery is hungry.'
\z 
 
\ea \label{HB.6}
\textit{hezretū šēxīč fermāwo, dey konī?} \\ 
\gll hezret=ū šēx=yč ∅-fermāw-o dey ko=nī \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} where\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `His highness the Sheikh said, ‘Where are you?’'
\z 
 
\ea \label{HB.12}
\textit{bilo pey šeškū Bešāretī, āsāwekew šeškī} \\ 
\gll bi-l-o pey šešk=ū Bešāret-ī āsāw-eke=w šešk-ī \\ 
 \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} for \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} mill\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] go to Shashk Basharat\footnote{A village in Iraqi Kurdistan.}, to the mill of Shashk'
\z 
 
\ea \label{HB.15}
\textit{kābrāyč herbenew hereyn, milo herey gurj kerowew gino řā,} \\ 
\gll kābrā=īč herbene=ū her-e-ī=n mi-l-o her-e-ī gurj ∅-ker-o=we=ū ∅-gin-o řā \\ 
 man\textsc{=add} donkey\_breeder\textsc{.m}=and donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} alert \textsc{ind-}do\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}}\textsc{=ez}\textsc{.gen} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} road\textsc{\textsc{.f}} \\ 
\glt `The fellow who was a donkey breeder, went and prepared the donkey, [and] set off.'
\z 
 
\ea \label{HB.17}
\textit{miđyo herzretū šēxī nefermāwā bilue yānew āney bāre’} \\ 
\gll mi-đye-o herzret=ū šēx-ī ne-fermāwā bi-lue yāne=ū āney b-ār-e \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}say\textsc{.pst} \textsc{sbjv-}go\textsc{.imp}\textsc{.\textsc{2sg}} house=and \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `He noticed that his highness the Sheikh hadn’t told [him], ‘Go to the house of that (i.e. such-and-such) person and bring [back the flour]’'
\z 
 
\ea \label{HB.21}
\textit{min bilū āsāwū kē hārdī bārū} \\ 
\gll min bi-l-w āsāw=ū kē hārdī b-ār=ū \\ 
 1sg \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} mill\textsc{.m}\textsc{=ez}\textsc{.gen} who flour\textsc{\textsc{.f}} \textsc{sbjv-}bring\textsc{.prs}\textsc{=ez}\textsc{.gen} \\ 
\glt `To whose mill I should go and bring flour?'
\z 
 
\ea \label{HB.28}
\textit{xeylē rāziɫū diɫīš kerdēš} \\ 
\gll xeylē rāziɫ=ū diɫ-ī=š kerd-ē=š \\ 
 so much request\textsc{=ez}\textsc{.gen} heart\textsc{.m} \textsc{dem.prox}\textsc{=3sg} do\textsc{.pst} 3pl\textsc{=3sg} \\ 
\glt `He struggled a lot.'
\z 
 
\ea \label{HB.30}
\textit{firey lūān nēzīkū Šeškīū kewteno} \\ 
\gll firey lūā=n nēzīk=ū Šešk-ī=w kewte=n=o \\ 
 a lot go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} near\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}=and fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{compl}} \\ 
\glt `He had gone a long way [and] approached Shashk'
\z 
 
\ea \label{HB.33}
\textit{milmē gošew šārī, āwāyekey} \\ 
\gll mi-l-mē goše=w šār-ī āwāī-ekey \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-1pl} corner\textsc{=ez}\textsc{.gen} city\textsc{.m}\textsc{-obl}\textsc{.m} village\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `We will go to the city suburb, to the village. '
\z 
 
\ea \label{HB.36}
\textit{simerī kerā tūrekekew min} \\ 
\gll simerī ∅-ker-ā tūreke-eke=w min \\ 
 straw\textsc{\textsc{.f}} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} sack\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg \\ 
\glt `They (the family) will put [some] straw into my sack;'
\z 
 
\ea \label{HB.47}
\textit{ʕejāyib menow} \\ 
\gll ʕejāyib ∅-men-o=ū \\ 
 astonished \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=ez}\textsc{.gen} \\ 
\glt `The man was astonished.'
\z 
 
\ea \label{HB.48}
\textit{dāš wene hereš berd gošew šārū šeškī} \\ 
\gll dā=š wene her-e=š berd-∅ goše=w šār=ū šešk-ī \\ 
 give\textsc{.pst}\textsc{=3sg} \textsc{post} donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{=3sg} take\textsc{.pst}\textsc{-3sg}\textsc{.m} corner\textsc{=ez}\textsc{.gen} city\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He set off. He took the donkey to the suburb of Shashk city.'
\z 
 
\ea \label{HB.51}
\textit{wātiš, fermāwdē bēydē mēmānū minindē} \\ 
\gll wāt=iš ∅-fermāw-dē b-ē-ydē mēmān=ū min=indē \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{imp-}say\textsc{.prs}\textsc{-2pl} \textsc{imp-}come\textsc{.prs}\textsc{-2pl} guest\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg\textsc{=cop}\textsc{.2pl} \\ 
\glt `He said, ‘Please come! You are my guests.'
\z 
 
\ea \label{HB.52}
\textit{mēmānū minindē'} \\ 
\gll mēmān=ū min=indē \\ 
 guest\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg\textsc{=cop}\textsc{.2pl} \\ 
\glt `You are my guests.’'
\z 
 
\ea \label{HB.56}
\textit{hēɫeřuenīšā kerde pey sāhēbū hereyū pey herbenew hereyū} \\ 
\gll hēɫeřuenī=šā kerd-e pey sāhēb=ū her-e-ī=ū pey herbene=ū her-e-ī=ū \\ 
 fried\_egg\textsc{.dir}\textsc{\textsc{.f}}\textsc{=3pl} do\textsc{.pst}-f\textsc{.3sg} for owner\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} for donkey breeder\textsc{.m}=and donkey\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \\ 
\glt `They (the host’s family) cooked fried eggs for the owner of the donkey, for the donkey breeder.'
\z 
 
\ea \label{HB.60}
\textit{āmā āsāwekew Šeškī, kābrāke} \\ 
\gll āmā āsāw-eke=w Šešk-ī kābrā-eke \\ 
 come\textsc{.pst}\textsc{.3sg} mill\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} man\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `The fellow came to the Shashk mill.'
\z 
 
\ea \label{HB.61}
\textit{duē ferdē hārdīš dē qewū herekeywe peyšū} \\ 
\gll duē ferdē hārdī=š dā-ē qew=ū her-ekey=we pey=š=w \\ 
 two sack\textsc{.dir}\textsc{.pl} flour\textsc{.dir}\textsc{\textsc{.f}}\textsc{.pl}\textsc{=3sg} give\textsc{.pst}\textsc{-dir}\textsc{.pl} on\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} for\textsc{=3sg}=and \\ 
\glt `He put two sacks of flour on the back of the donkey'
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{HB.68}
\textit{milūwe, min činnē sāɫēn īnānē xizmetū šēxīne} \\ 
\gll mi-l-w=we min činnē sāɫe-ē=n īnā-ānē xizmet=ū šēx-ī=ne \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{compl}} 1sg some\textsc{.pl} year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-1sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I will go back! It’s been several years that I am at the service of the sheikh;'
\z 
 
\ea \label{HB.69}
\textit{herbeneš pey kerūnew renjberiš bīyenā} \\ 
\gll herbene=š pey ∅-ker-ūne=ū renjber=iš bīye=nā \\ 
 donkey breeder\textsc{.m}\textsc{=3sg} for \textsc{ind-}do\textsc{.prs}\textsc{-1sg}\textsc{=ez}\textsc{.gen} worker\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \\ 
\glt `I take care of the donkeys for him; I have been his worker.'
\z 
 
\ea \label{HB.70}
\textit{ēnnew herēm nezānān} \\ 
\gll ēnne=w her-ē=m ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Wasn’t I as much of a confidante as a donkey?' \footnote{Lit. didn’t I know as much as a donkey!}
\z 
 
\ea \label{HB.72}
\textit{milowew bārū herey wizorew} \\ 
\gll mi-l-o=we=ū bār=ū her-e-ī ∅-wiz-o=re=ū \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}}=and load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}-\textsc{2sg}\textsc{-obl}\textsc{.m} \textsc{ind-}throw\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}}=and \\ 
\glt `He went back and put down the donkey’s load.'
\z 
 
\ea \label{HB.73}
\textit{hārdā wizorew berū xāneqāyne lā nānewāyne} \\ 
\gll hārdī-ā ∅-wiz-o=re=ū ber=ū xāneqā-ī=ne lā nānewā-ī=ne \\ 
 flour\textsc{\textsc{.f}}\textsc{-obl}\textsc{.pl} \textsc{ind-}throw\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}}=\textsc{compl} front\textsc{=ez}\textsc{.gen} monastery\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} at\_the\_place\_of baker\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He put the flour [sacks] down in front of the monastery at the place of the baker.'
\z 
 
\ea \label{HB.75}
\textit{xeber midā hezretū šēxī} \\ 
\gll xeber mi-de-ā hezret=ū šēx=ī \\ 
 news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \\ 
\glt `They (the sheikh’s servants) passed on the news to his highness the Sheikh.'
\z 
 
\ea \label{HB.76}
\textit{māčā, qurbān wiɫā fiɫāne kes toryānū ānā} \\ 
\gll m-āč-ā qurbān wiɫā fiɫān=e kes toryā=n=ū āne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} by\_God such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m} get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \\ 
\glt `They said, ‘Sir, indeed, the fellow has got offended; there he is’'
\z 
 
\ea \label{HB.78}
\textit{hurmēzo hezretū šēxī pīyāy kīyāno} \\ 
\gll hur-m-ēz-o hezret=ū šēx=ī pīyā-ī ∅-kīyān-o \\ 
 \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} man\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then\footnote{Lit. He rose.}, his highness the Sheikh sent men [to the donkey breeder]'
\z 
 
\ea \label{HB.80}
\textit{čiřoš hezretū šēxī fermāwo, erē řoɫem čī toryānī?} \\ 
\gll ∅-čiř-o=š hezret=ū šēx=ī ∅-fermāw-o erē řoɫe=m čī toryā=nī? \\ 
 \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} child\textsc{.m}\textsc{=1sg} why get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt ` His highness the Sheikh summoned him, i.e. the donkey breeder, [and] said, ‘Oh son! why were you offended? '
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{HB.83}
\textit{ēnnew herē qābilīm nebē hāɫīm kerī!} \\ 
\gll ēnne=w her-ē qābilī=m ne-b-ē hāɫī=m ∅-ker-ī \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} deserved\textsc{=1sg} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} understood\textsc{=1sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-obl}\textsc{.m} \\ 
\glt `Wasn’t I as trustworthy as a donkey that you didn’t make me understand!'
\z 
 
\ea \label{HB.84}
\textit{ēnnew herē, īse her be herī wēš qisē kero min hīčim nezānān!’} \\ 
\gll ēnne=w her-ē īse her be her-ī wē=š qisē-ē ∅-ker-o min hīč=im ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} now both by donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As much as a donkey! now the donkey could speak even though it was a donkey, but I didn’t know a thing.'
\z 
 
\ea \label{HB.85}
\textit{hezretū šēxī fermāwo, dey xāsā bābā herā gā!} \\ 
\gll hezret=ū šēx=ī ∅-fermāw-o dey xās=ā bābā her-ā gā \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} grandfather\textsc{.m} donkey\textsc{.m}\textsc{-obl}\textsc{.pl} fornicate\textsc{.pst} \\ 
\glt `His highness the Sheikh said, ‘That is fine! God damn you!\footnote{Lit. May donkeys fornicate with your ancestors!}'
\z 
 
\ea \label{HB.86}
\textit{īse kē be zūwānū herī zāno bējge to?} \\ 
\gll īse kē be zūwān=ū her-ī ∅-zān-o bējge to \\ 
 now who by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `Now who can understand the language of donkeys except you?'
\z 
 
\ea \label{HB.87}
\textit{īse bējgem to kē be zuwānū herī zāno?} \\ 
\gll īse bējgem to kē be zuwān=ū her-ī ∅-zān-o \\ 
 now except\_for \textsc{2sg} who by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `Now, except for you, who can understand the language of donkeys?'
\z 
 
\ea \label{HB.88}
\textit{kē be zuwānū herī zāno bejgem to} \\ 
\gll kē be zuwān=ū her-ī ∅-zān-o bejgem to \\ 
 who by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `Who can understand the language of donkeys except you?'
\z 
 
\ea \label{HB.91}
\textit{bizāne neferē ter hen be zūwānū herī hāɫīš bo bejge to!} \\ 
\gll bi-zān-e nefer-ē ter hen-∅ be zūwān=ū her-ī hāɫī=š ∅-b-o bejge to \\ 
 \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} person\textsc{.m}\textsc{-dir}\textsc{.pl} other exist\textsc{-3sg}\textsc{.m} by language\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m} understood\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} except\_for \textsc{2sg} \\ 
\glt `see if there is a person, apart from you, who can understand the language of donkeys!'
\z 
 
\ea \label{PM.1}
\textit{ʕerzū xizmetīt kerū hetā min žinīyenim pāse mešhūr bīyen šewē šā ʕelādīnī, šewē peɫē merekuř peɫē merekuř mē šārezūr} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū hetā min žinīye=n=im pāse mešhūr bīye=n šewe-ē šā ʕelādīn-ī šewe-ē peɫ-ē merekuř peɫ-ē merekuř m-ē šārezūr \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} like famous be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} night.f\textsc{-obl}.f king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} night.f\textsc{-obl}.f \textsc{clf}\textsc{-indf} grasshopper \textsc{clf}\textsc{-indf} grasshopper \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{pn} \\ 
\glt `Let me tell you .\footnote{Lit. May a petition be for your service.} I have even heard (this). It is well known that a swarm of grasshoppers came to Sharazour one night.'
\z 
 
\ea \label{PM.3}
\textit{her yāgēwe mē terefū xwāywe mē} \\ 
\gll her yāgē-ēwe m-ē teref=ū xwā-ī=we m-ē \\ 
 just place.f\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `They came from every direction. They came from God.'
\z 
 
\ea \label{PM.5}
\textit{fermāndew merekuřekey, reʕīsekešā māčo, ēme bē ā šexsī metāwmē, e čē menmēwe zerer deymē čī mentēqey’} \\ 
\gll fermānde=ū merekuř-ekey reʕīs-eke=šā m-āč-o ēme bē ā šexs-ī me-tāw-mē e=čē ∅-men-mē=we zerer ∅-de-ymē č=ī mentēqe-ī \\ 
 leader\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} leader\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} in=here \textsc{sbjv-}stay\textsc{.prs}\textsc{-1pl}\textsc{=compl} harm\textsc{.m} \textsc{sbjv-}give\textsc{.prs}-\textsc{1pl} in=\textsc{dem.prox} region\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The head of the grasshoppers said, ‘Without [permission from] this person we cannot stay here and do harm to this region.'
\z 
 
\ea \label{PM.6}
\textit{nimāyinde kīyāno xizmetū šā ʕelādīnī, sāhēbū hīne gewrew merekuřekey} \\ 
\gll nimāyinde ∅-kīyān-o xizmet=ū šā ʕelādīn-ī sāhēb=ū hīne gewre=w merekuř-ekey \\ 
 representative\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{fill} head\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `He, i.e., the head of the grasshoppers, sent a representative to Shah Aladin’s presence;'
\z 
 
\ea \label{PM.8}
\textit{fermāwo, lue xizmetū āđī!} \\ 
\gll ∅-fermāw-o ∅-lu-e xizmet=ū āđī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `[and] said, ‘Go into his presence!'
\z 
 
\ea \label{PM.10}
\textit{nimāyindeke mē xizmetū šēxī} \\ 
\gll nimāyinde-eke m-ē xizmet=ū šēx-ī \\ 
 representative\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The messenger came into the Sheikh’s presence'
\z 
 
\ea \label{PM.13}
\textit{kīyāsenāšā xizmetū to} \\ 
\gll kīyāse=nā=šā xizmet=ū to \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=3pl} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} \\ 
\glt `I have been sent to your service.'
\z 
 
\ea \label{PM.14}
\textit{eger to ījāze miđey ēme e čē nehārē kermē ījāze be desū ton} \\ 
\gll eger to ījāze mi-đe-ī ēme e=čē nehār-ē ∅-ker-mē ījāze be des=ū to=n \\ 
 if \textsc{2sg} permission \textsc{ind-}give\textsc{.prs}-\textsc{2sg} \textsc{1pl} in=here lunch\textsc{.m}\textsc{-indf} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} permission to hand\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `It is in your hands whether to let us stay here for lunch.'
\z 
 
\ea \label{PM.15}
\textit{āmēnmē xizmetū to ījāzemā biđey} \\ 
\gll āmē=nmē xizmet=ū to ījāze=mā bi-đe-ī \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} permission\textsc{=1pl} \textsc{sbjv-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `I\footnote{Lit. We.} have come to your service so that you might permit us.'
\z 
 
\ea \label{PM.18}
\textit{muɫkēšā bīyebo čāne temāmū muɫkekey kuliš je zerāʕetne bīyebo āmābonew} \\ 
\gll muɫk-ē=šā bīye b-o č=ā=ne temām=ū muɫk-ekey kul=iš je zerāʕet=ne bīye b-o āmā b-one=ū \\ 
 property\textsc{.m}\textsc{-indf}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} in=\textsc{dem.dist}\textsc{=\textsc{post}} all\textsc{\textsc{=ez.gen}} property\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} all\textsc{=3sg} from agriculture\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `—īt appears that he\footnote{they} had a big property. The property might have been filled with crops; and they might have been ripe.'
\z 
 
\ea \label{PM.20}
\textit{fermāwebē, fermāwo lodē e čā muɫkū ēme wēmā, herčēmā hen čāne ā firmānde wēš zāno heqū ēme kāmā} \\ 
\gll fermāwe b-ē ∅-fermāw-o ∅-lo-dē e=čā muɫk=ū ēme wē=mā herčē=mā hen-∅ č=ā=ne ā firmānde wē=š ∅-zān-o heq=ū ēme kām=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} in=there property\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{reflx}\textsc{=1pl} whatever\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} leader\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} right\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} which\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go! In that property of mine, whatever we possess there, your boss knows which portion [of land] is mine.'
\z 
 
\ea \label{PM.21}
\textit{ēšew nehār kerdē, pey nehārū wētāw seʕbē xwāhāfīztā!} \\ 
\gll ēšew nehār ∅-ker-dē pey nehār=ū wē=tā=ū seʕbe-ē xwāhāfīz=tā \\ 
 tonight lunch\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2pl} for lunch\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and morning.f\textsc{-obl}.f goodbye\textsc{.m}=\textsc{2pl} \\ 
\glt `[Take it] as a camp tonight [and] eat from it as your lunch. [Then] Tomorrow, bid farewell!'
\z 
 
\ea \label{PM.23}
\textit{šewē merekuř nīšo milo mehelekew šēxī ʕelādīnī} \\ 
\gll šewe-ē merekuř ∅-nīš-o mi-l-o mehele-eke=ū šēx-ī ʕelādīn-ī \\ 
 night.f\textsc{-obl}.f grasshopper \textsc{ind-}sit\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} neighbourhood\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `The grasshoppers stayed\footnote{Lit. sat.} [there]. They attacked\footnote{Lit. sit on.} Sheikh Aladin’s neighbourhood.'
\z 
 
\ea \label{PM.28}
\textit{tāte řow bāwke řo verzīyer mē pey xizmetū šā ʕelādīnī} \\ 
\gll tāte řo=ū bāwke řo verzīyer m-ē pey xizmet=ū šā ʕelādīn-ī \\ 
 father\textsc{.m} onom\textsc{\textsc{=ez.gen}} fahter.\textsc{voc} onom farm\_labourer\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[Sheikh Aladin’s farm-labourer shouted], ‘Oh, dear God! The farm-labourer came into the presence of Shah Aladin.'
\z 
 
\ea \label{PM.30}
\textit{īne bē ījāzew to īne nekiryān} \\ 
\gll īne b-ē ījāze=ū to īne ne-kir-yā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} permission\textsc{\textsc{=ez.gen}} \textsc{2sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{neg-}do.pass-past.pass\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This couldn’t have been done without your permission.'
\z 
 
\ea \label{PM.31}
\textit{tenhā hinū ton, geɫēwew kesī zereriš neginān je šārezūrne} \\ 
\gll tenhā hin=ū to=n geɫā-ēwe=ū kes-ī zerer=iš ne-ginā=n je šārezūr=ne \\ 
 not\_even ez.pron\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} leaf\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} person\textsc{.m}\textsc{-obl}\textsc{.m} harm\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} \\ 
\glt `It’s all yours [otherwise] no one’s leaves have been harmed in Sharazour. '
\z 
 
\ea \label{PM.34}
\textit{nāřihet nebī āne meʕmūre bē terefū xwāywe āmābē} \\ 
\gll nāřihet ne-b-ī āne meʕmūre b-ē teref=ū xwā-ī=we āmā b-ē \\ 
 sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} agent be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Don’t worry! It was an agent [who] was sent  by God.'
\z 
 
\ea \label{PM.36}
\textit{min pewkī ījāzem dān hinekew ēme wero, bilo zerer je kesī negino} \\ 
\gll min pewkī ījāze=m dā=n hin-eke=ū ēme ∅-wer-o bi-l-o zerer je kes-ī ne-gin-o \\ 
 \textsc{1sg} that\_is\_why permission\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} ez.pron\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{sbjv-}eat\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} harm\textsc{.m} from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `I let them [eat our crop] so that they would eat our crop and go away, [but] not do harm to others.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.38}
\textit{girđū šārezūrī ēsāɫ merāʕetū ēme kero je pārī zīyāter pey toyčū pey ēmeyč,} \\ 
\gll girđ=ū šārezūr-ī ēsāɫ merāʕet=ū ēme ∅-ker-o je pār-ī zīyā-ter pey to=yč=ū pey ēme=yč \\ 
 all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} this\_year care\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} from last\_year\textsc{-obl}\textsc{.m} increased\textsc{-cmpr} for \textsc{2sg}\textsc{=add}=and for \textsc{1pl}\textsc{=add} \\ 
\glt `This year, all the people in Sharazour care for us— both you and I—, more than the previous year.'
\z 
 
 \ea \label{DG.1}
\textit{šā ʕelādīnū šā ʕosmān} \\ 
\gll šā ʕelādīn=ū šā ʕosmān \\ 
 king\textsc{.m} \textsc{pn}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{.m} \\ 
\glt `Shah Aladin and Shah Osman.'
\z 
 
\ea \label{DG.2}
\textit{šā ʕelādīn tātew šā ʕosmānī bīyen} \\ 
\gll šā ʕelādīn tāte=ū šā ʕosmān-ī bīye=n \\ 
 king\textsc{.m} \textsc{pn} father\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Shah Aladin was Shah Osman’s father.'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.5}
\textit{eger newešīēmā bīyebo duktur nebīyenū} \\ 
\gll eger newešī-ē=mā bīye b-o duktur ne-bīye=n=ū \\ 
 if illness\textsc{.m}\textsc{-indf}\textsc{=1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} physician\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \\ 
\glt `If there might have been an illness, there was no doctor. '
\z 
 
\ea \label{DG.6}
\textit{ā wextīekey lūēnmē e wāge duʕāw xeyrīš kerdēnew} \\ 
\gll ā wext-ī-ekey lūē=nmē e=wāge duʕā=ū xeyr-ī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} in=there prayer\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} goodness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `We would go there [i.e. to them and] they would bless us'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.11}
\textit{řoē je řoā yo mē xizmetiš, xizmetū šā ʕelādīnī} \\ 
\gll řo-ē je řo-ā yo m-ē xizmet=iš xizmet=ū šā ʕelādīn-ī \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{=3sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Once\footnote{Lit. One day of days.}, a man\footnote{Lit. one.}  came into his presence, to Shah Aladin'
\z 
 
\ea \label{DG.12}
\textit{birāš tūšū derdē mē,} \\ 
\gll birā=š tūš=ū derde-ē m-ē \\ 
 brother\textsc{.m}\textsc{=3sg} encounter\textsc{=ez}\textsc{.gen} illness\textsc{\textsc{.f}}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `His brother was suffering from an illness.  '
\z 
 
\ea \label{DG.13}
\textit{derdekēš, ā wextīekey, derdū gulī bo, dūrū gošdārā} \\ 
\gll derde-ekē=š ā wext-ī-ekey derde=ū gulī ∅-b-o dūr=ū gošdār-ā \\ 
 illness\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy \textsc{ind-}be\textsc{.prs}\textsc{-3sg} far\textsc{=ez}\textsc{.gen} listener\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `May it be far from the audience, his illness was leprosy.'
\z 
 
\ea \label{DG.14}
\textit{derdū gulī bo,} \\ 
\gll derde=ū gulī ∅-b-o \\ 
 illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `His illness was leprosy. '
\z 
 
\ea \label{DG.16}
\textit{kābrāyč werū řehmū wēšū milo pāɫ midowe birākeyšo} \\ 
\gll kābrā=īč wer=ū řehm=ū wē=š=ū mi-l-o pāɫ mi-de-o=we birā-ekey=š=o \\ 
 man\textsc{=add} out\_of\textsc{=ez}\textsc{.gen} mercy\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3sg}=and \textsc{ind-}go\textsc{.prs}\textsc{-3sg} side/flank\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `The fellow, out of mercy, went and supported his brother.'
\z 
 
\ea \label{DG.22}
\textit{birākeyš, derde, lāšeš derde kero āwī šāno derdū gulī bo} \\ 
\gll birā-ekey=š derde lāše=š derde ∅-ker-o āwī ∅-šān-o derde=ū gulī ∅-b-o \\ 
 brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} illness\textsc{\textsc{.f}} body\textsc{.m}\textsc{=3sg} illness\textsc{\textsc{.f}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} water\textsc{\textsc{.f}} \textsc{ind-}emit\textsc{.prs}\textsc{-3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `His brother had pain in his body; it would emit liquid\footnote{Lit. water.}. [The illness] was leprosy.'
\z 
 
\ea \label{DG.27}
\textit{lue wēt wize milū pāɫāyekā šā ʕelādīnīre} \\ 
\gll ∅-lu-e wē=t ∅-wiz-e mil=ū pāɫā-ekā šā ʕelādīn-ī=re \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}throw\textsc{.prs}-\textsc{2sg} on\textsc{=ez}\textsc{.gen} shoe\textsc{-def}\textsc{.obl}\textsc{.pl} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Go, throw yourself at the feet\footnote{Lit. shoes.} of Shah Aladin'
\z 
 
\ea \label{DG.30}
\textit{mēnē degāw činārīš māčā} \\ 
\gll m-ē-nē degā=ū činār-ī=š m-āč-ā \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} village\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \\ 
\glt `They came to a village called Chinar.'
\z 
 
\ea \label{DG.31}
\textit{mēnē e čā birākeyš bero teɫānēwene beynū činārāne} \\ 
\gll m-ē-nē e=čā birā-ekey=š ∅-ber-o teɫān-ēwe=ne beyn=ū činār-ā=ne \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} in=there brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}take\textsc{.prs}\textsc{-3sg} terrace\_farm\textsc{-indf}\textsc{=\textsc{post}} between\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-3pl}\textsc{=\textsc{post}} \\ 
\glt `They went to a terrace farm in between the the two sides of Chinar village. '
\z 
 
\ea \label{DG.32}
\textit{minyošrew wēš mē xāneqāw bīyārē} \\ 
\gll mi-nye-o=š=re=ū wē=š m-ē xāneqā=ū bīyārē \\ 
 \textsc{ind-}put\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=\textsc{povb}}=and \textsc{reflx}\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} monastery\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \\ 
\glt `He left his brother there, and went to the monastery.'
\z 
 
\ea \label{DG.33}
\textit{xāneqāw bīyārēne, ā wextīyekey dāđ kero} \\ 
\gll xāneqā=ū bīyārē=ne ā wext-ī-ekey dāđ ∅-ker-o \\ 
 monastery\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yell\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He gave a yell of despair at the monastery of Biyare.'
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DG.38}
\textit{ā wextīyekey, ētir, īse, min yerē sāɫē ī birāme tūšū ī derdēe āmānū} \\ 
\gll ā wext-ī-ekey ētir īse min yerē sāɫe-ē ī birā=m=e tūš=ū ī derde-ē=e āmā=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} encounter\textsc{=ez}\textsc{.gen} \textsc{dem.prox} illness\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=dem} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Now, as for me, it’s been three years that my brother is suffering from this illness. '
\z 
 
\ea \label{DG.42}
\textit{nizīkū āweyānīyē nimāzā ginmē} \\ 
\gll nizīk=ū āweyānī-ē ni-m-āz-ā ∅-gin-mē \\ 
 close\textsc{=ez}\textsc{.gen} habitat\textsc{\textsc{.m}}\textsc{-indf}\textsc{\textsc{.f}} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} sbjv\textsc{-f}all\textsc{.prs}\textsc{-1pl} \\ 
\glt `They don’t let us anywhere near a village.'
\z 
 
\ea \label{DG.43}
\textit{ī birāme derdeš, derdēš īnā po derdū gulīš māčā dāđē!} \\ 
\gll ī birā=m=e derde=š derde-ē=š īnā-∅ p==oderde=ū gulī=š m-āč-ā dāđ-ē \\ 
 \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} illness\textsc{\textsc{.f}}\textsc{=3sg} illness\textsc{\textsc{.f}}\textsc{-indf}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at=\textsc{=\textsc{post}}illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} yell\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `This brother of mine is ill; it is called leprosy; Please help!'
\z 
 
\ea \label{DG.46}
\textit{šēx mutewej bo, wextē miđyo derdū ī pīyāy lā xwāyū xwāwinū ʕāɫemīyānī mutewej bo, řānimonīš kero} \\ 
\gll šēx mutewej ∅-b-o wext-ē mi-đye-o derde=ū ī pīyā-ī lā xwā-ī=ū xwāwin=ū ʕāɫemīyān-ī mutewej ∅-b-o řānimonī=š ∅-ker-o \\ 
 sheikh\textsc{.m} aware \textsc{ind-}be\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} man\textsc{.m}\textsc{-obl}\textsc{.m} at\_the\_place\_of God\textsc{.m}\textsc{-obl}\textsc{.m}=and owner\textsc{=ez}\textsc{.gen} worlds\textsc{.m}\textsc{-obl}\textsc{.m} aware \textsc{ind-}be\textsc{.prs}\textsc{-3sg} guidance\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The Sheikh understood. When he looked at the man’s illness, the lord of all universes realised [what] his illness [was and] guided him (i.e. the Sheikh).'
\z 
 
\ea \label{DG.47}
\textit{winū kīseɫē sāwī be} \\ 
\gll winī=ū kīseɫī-ē ∅-sāw-ī be \\ 
 blood\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} tortoise\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{sbjv-}rub\textsc{.prs}-\textsc{2sg} to \\ 
\glt `[The Sheikh said to the man], ‘You may rub tortoise’s blood on …'
\z 
 
\ea \label{DG.49}
\textit{winū kīseɫē zāmdāre kere} \\ 
\gll winī=ū kīseɫī-ē zāmdār-e ∅-ker-e \\ 
 blood\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} tortoise\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} wounded\textsc{-f} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Make a tortoise wounded!'
\z 
 
\ea \label{DG.50}
\textit{winekēš sāwe pā qelbū pīyāyre} \\ 
\gll winī-ekē=š ∅-sāw-e p=ā qelb=ū pīyā-ī=re \\ 
 blood\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} by\textsc{=dem}.dist heart\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `[then] rub its blood on the heart of the [sick] man'
\z 
 
\ea \label{DG.53}
\textit{šā ʕelādīn īney īney māčo be, fermāwo be pīyā, hīnī birāw pīyākey} \\ 
\gll šā ʕelādīn īney īney m-āč-o be ∅-fermāw-o be pīyā hīn-ī birā=ū pīyā-ekey \\ 
 king\textsc{.m} \textsc{pn} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by man\textsc{.m} \textsc{fill}\textsc{-obl}\textsc{.m} brother\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Shah Aladin said this to the man, to the brother of the [sick] man.'
\z 
 
\ea \label{DG.54}
\textit{birāš īnārē degāw činārī, metāwo bilo āweyānī} \\ 
\gll birā=š īnā-∅=rē degā=ū činār-ī me-tāw-o bi-l-o āweyānī \\ 
 brother\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} village\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} habitat\textsc{\textsc{.m}} \\ 
\glt `His brother was in Chinar village. He wasn’t allowed to go the village '
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.60}
\textit{šūnū īneyre āhā lā xwāwe inne duʕāš meqboɫe bo} \\ 
\gll šūn=ū īney=re āhā lā xwā=we inne duʕā=š meqboɫ-e ∅-b-o \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} \textsc{intj} at\_the\_place\_of God\textsc{.m}\textsc{=\textsc{post}} so\_much prayer\textsc{\textsc{.f}}\textsc{=3sg} acceptable\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well then, his prayers were accepted by God.'
\z 
 
\ea \label{DG.61}
\textit{kīseɫēwe čā kīseɫā milo xizmetū šā ʕelādīnī} \\ 
\gll kīseɫī-ēwe č==ākīseɫī-ā mi-l-o xizmet=ū šā ʕelādīn-ī \\ 
 tortoise\textsc{\textsc{.f}}\textsc{-indf} from=\textsc{=cop}\textsc{.3sg}\textsc{.m}tortoise\textsc{\textsc{.f}}\textsc{-3pl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `A tortoise\footnote{Lit. A tortoise of those tortoises.}  came into the presence of Shah Aladin.'
\z 
 
\ea \label{DG.64}
\textit{ā wextīekey beɫē, min bilo ā memleketū wēšā, kiseɫīš čenē, ēme zāmdārē nekero ēme kečē nekero} \\ 
\gll ā wext-ī-ekey beɫē min bi-l-o ā memleket=ū wē=šā kiseɫī=š čenē ēme zāmdār-ē ne-ker-o ēme keč-ē ne-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yes \textsc{1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3pl} tortoise\textsc{\textsc{.f}}\textsc{=3sg} in \textsc{1pl} wounded\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `[The Sheikh said], ‘Yes!' '[The tortoise said] I [say] he [should] go to his region; there are tortoises there. He shall not injure us; he shall not make us crooked. '
\z 
 
\ea \label{DG.65}
\textit{berošo memliketū wēšā, kīseɫē zāmdār kero} \\ 
\gll ∅-ber-o=š=o memliket=ū wē=šā kīseɫī-ē zāmdār ∅-ker-o \\ 
 \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=\textsc{compl}} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3pl} tortoise\textsc{\textsc{.f}}\textsc{-indf} wounded \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He shall take him to his own region [and] injure tortoises from there.'
\z 
 
\ea \label{DG.68}
\textit{berešo memliketū wētā welē sermā deydē} \\ 
\gll ∅-ber-e=š=o memliket=ū wē=tā welē ser=mā ∅-de-ydē \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=\textsc{compl}} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=2pl} but head\textsc{.m}\textsc{=1pl} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} \\ 
\glt `Take him to your region, but pay us a visit.'
\z 
 
\ea \label{DG.73}
\textit{īnē dāstānū ī šēxāū ēme bīyen gewregerīšā} \\ 
\gll īnē dāstāne=ū ī šēx-ā=ū ēme bīye=n gewregerī=šā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{\textsc{.f}}\textsc{.3sg} story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} sheikh\textsc{.m}\textsc{-obl}\textsc{.pl}=\textsc{ez.gen} \textsc{1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} greatness\textsc{=3pl} \\ 
\glt `This is the story of our Sheikhs and their greatness.'
\z 
 

\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.2}
\textit{yošā be nāmēw sibhān āġeywe} \\ 
\gll yo=šā be nāmē=ū sibhān āġe-ī=we \\ 
 one\textsc{.m}\textsc{=3pl} by name\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `One of them [was] called Sibhan Agha.'
\z 
 
\ea \label{DP.3}
\textit{yošā be nāmēw jemšēr šāywe} \\ 
\gll yo=šā be nāmē=ū jemšēr šā-ī=we \\ 
 one\textsc{.m}\textsc{=3pl} by name\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `One of them [was] called Jamsher Agha.'
\z 
 
\ea \label{DP.9}
\textit{āmān lūān ā qeɫāw hewrāmānī e pāgewe nīštenre} \\ 
\gll āmā=n lūā=n ā qeɫā=w hewrāmān-ī e=pāgewe nīšte=n=re \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} in=there sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} \\ 
\glt `He came [here and] went to the citadel of Hawraman over there and settled there.'
\z 
 
\ea \label{DP.13}
\textit{dāwāw āġeyīš kerdēne} \\ 
\gll dāwā=w āġe-ī=š kerdē=ne \\ 
 demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} agha\textsc{.m}\textsc{-nmlz}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `He (i.e., Jamsher Shah) claimed governorship [over] Hawraman. '
\z 
 
\ea \label{DP.14}
\textit{ī memlikete desū āđīwe bīyen} \\ 
\gll ī memliket=e des=ū āđī=we bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=dem} hand\textsc{.m}\textsc{=ez}\textsc{.gen} 3sg\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This region was under Sibhan Agha’s control:'\footnote{Lit. This region was in his hands.}
\z 
 
\ea \label{DP.17}
\textit{ā wextīyekey īđīč āmān dāwāw pādšāgerīš kerdēnew} \\ 
\gll ā wext-ī-ekey īđ=yč āmā=n dāwā=w pādšāgerī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} kingship\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `He (i.e., Jamsher Shah) came [and] claimed kingship.'
\z 
 
\ea \label{DP.20}
\textit{wāteniš, min zīndigīū bābiɫ bābīm e čē bīyen} \\ 
\gll wāte=n=iš min zīndigī=ū bāb-iɫ bāb-ī=m e=čē bīye=n \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} life\textsc{.m}\textsc{=ez}\textsc{.gen} ancestor\textsc{.m}-of ancestor\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg} in=here be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` saying, ‘I, my ancestors, have all lived here.'
\z 
 
\ea \label{DP.21}
\textit{ī bāxū bāxātū hewrāmānī, memliketū hewrāmānī, xeɫkū hewrāmānī ā wextīyekey ēme hukimřewiš bīyēnmē} \\ 
\gll ī bāx=ū bāx-āt=ū hewrāmān-ī memliket=ū hewrāmān-ī xeɫk=ū hewrāmān ī ā wext-ī-ekey ēme hukimřew=iš bīyē=nmē \\ 
 \textsc{dem.prox} garden\textsc{.m}\textsc{=ez}\textsc{.gen} garden\textsc{.m}\textsc{-pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} people\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \textsc{dem.prox} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} governor\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `We have ruled over\footnote{Lit. We have been rulers.} these gardens of Hawraman, this region of Hawraman, [and] this people of Hawraman.'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.25}
\textit{šūnū āneyre beynē e čā bīyenū} \\ 
\gll šūn=ū āney=re beynē e=čā bīye=n=w \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} a\_while in=there be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Jamsher Shah was still there (i.e., in the citadel) afterwards.'
\z 
 
\ea \label{DP.28}
\textit{ī māɫū miɫke hinū āđī bīyenū desšo bīyenū} \\ 
\gll ī māɫ=w miɫk=e hin=ū āđī bīye=n=w des=š=o bīye=n=ū \\ 
 \textsc{dem.prox} property\textsc{.m}=and property\textsc{.m}\textsc{=dem} ez.pron\textsc{=ez}\textsc{.gen} 3sg\textsc{.obl}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and hand\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `These lands and properties here were his. They were under his control.'
\z 
 
\ea \label{DP.30}
\textit{šūnū āneyre wextē zāno hāmin āmāw gīriš kerd sibhān āġe milo āwekēū jemšēr āġey biřo, hesāryewe} \\ 
\gll šūn=ū āney=re wext-ē ∅-zān-o hāmin āmā=ū gīr=iš kerd sibhān āġe mi-l-o āwī-ekē=w jemšēr āġe-ī ∅- bi-ř-o hesār-ī=ewe \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} time\textsc{.m}\textsc{-indf} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} summer\textsc{.m} come\textsc{.pst}\textsc{.3sg}=and hook\textsc{.m}\textsc{=3sg} do\textsc{.pst} \textsc{pn} agha\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} water\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}=and \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}\textsc{sbjv-}cut\textsc{.prs}\textsc{-3sg} pond\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Then, as Sibhan Agha\footnote{Lit. he.} knew summer was coming [and] Jamsher Agha\footnote{Lit. he.} was stuck, Sibhan Agha went and cut off Jamsher Shah’s water supply [and led it to a] pond.'
\z 
 
\ea \label{DP.32}
\textit{āhekū gunjēw ārdēnešā ew dīmū qeɫē} \\ 
\gll āhek=ū gunj-ē=ū ārdē=ne=šā ew dīm=ū qeɫā-ē \\ 
 lime\textsc{.m}\textsc{=ez}\textsc{.gen} earthenware\_ pipe\textsc{.m}\textsc{-dir}\textsc{.pl}=and bring\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}\textsc{=3pl} \textsc{dem.dist} side\textsc{.m}\textsc{=ez}\textsc{.gen} castle\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \\ 
\glt `They took the water to the other side of the citadel.'
\z 
 
\ea \label{DP.35}
\textit{īzāfeyš wizo pā sextre māčāš qeɫāw hewrāmānī} \\ 
\gll īzāfe-ī=š wiz-o p=ā sext=re mi-āč-ā=š qeɫā=w hewrāmān-ī \\ 
 addition\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg} by\textsc{=dem}.dist mountain\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He would throw whatever was extra [of the snow] at that hard-to-access mountain called the citadel of Hawraman.'
\z 
 
\ea \label{DP.37}
\textit{be xātirū īney qesem wero qesemiš negino} \\ 
\gll be xātir=ū īney qesem ∅-wer-o qesem=iš ne-gin-o \\ 
 by sake\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} oath\textsc{.m} \textsc{sbjv-}out\textsc{-3sg} oath\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wanted to swear an oath using the sparrow. He did not want to swear falsely on himself.'
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.39}
\textit{bē bā nīšmēre seḷāhū wēmā kermē} \\ 
\gll bē bā ∅-nīš-mē=re seḷāh=ū wē=mā ∅-ker-mē \\ 
 come\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{hort} \textsc{sbjv-}sit\textsc{.prs}\textsc{-1pl}\textsc{=\textsc{povb}} solution\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `[He said to Sibhan Agha], ‘Let us sit together and come to a solution.’'
\z 
 
\ea \label{DP.45}
\textit{āđ kārū wēšū min hinū wēm} \\ 
\gll āđ kār=ū wē=š=ū min hin=ū wē=m \\ 
 3sg\textsc{.dir}\textsc{.m} task\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3sg}\textsc{=and} \textsc{1sg} \textsc{z.pron}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} \\ 
\glt `Then, he goes his way, and I go mine.'
\z 
\ea \label{DP.46}
\textit{ā wextīyekey wextē mēnē ā wextī herčē qisē kerā, herčē behsē kerā herčē hesābē kerā, qēxāw hewrāmānī herčē bā ā wextī isɫāhšā mebo} \\ 
\gll ā wext-ī-ekey wext-ē mi-ē-nē ā wext-ī herčē qisē-ē ∅-ker-ā herčē behse-ē ∅-ker-ā herčē hesāb-ē ∅-ker-ā qēxā=w hewrāmān-ī herčē ∅-b-ā ā wext-ī isɫāh=šā me-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} however\_much talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much discussion\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much account\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} chieftain\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} however\_much \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} amendment\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then Sibhan Agha and his men came [to Jamsher Shah]. No matter how much the headmen of Hawraman talked [to both of them], discussed the issue, and tried to mediate, they (i.e., Jamsher Shah and Sibhan Agha) disagreed.'
\z 
 
\ea \label{DP.49}
\textit{ī gīyānū qefesew sīney minne īnen āđe lūā} \\ 
\gll ī gīyān=ū qefese=ū sīne=y min=ne īne=n āđe lūā \\ 
 \textsc{dem.prox} soul\textsc{.m}\textsc{=ez}\textsc{.gen} cage\textsc{.m}=and chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{\textsc{.f}} go\textsc{.pst} \\ 
\glt ` The soul in my rib cage [that I swore an oath on] was this [sparrow]. It [just] flew away.'
\z 
 
\ea \label{DP.50}
\textit{ānen mirīčēɫēm wer dān, to xeyrū min siɫāmet} \\ 
\gll āne=n mirīčēɫe-ē=m wer dā=n to xeyr=ū min siɫāmet \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} small\_bird\textsc{.m}\textsc{-indf}\textsc{=1sg} out give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `’I have let the bird go away. Now, you follow your life, and I will follow mine!\footnote{Lit. Well-being for you and health for me.}’'
\z 
 
\ea \label{DP.53}
\textit{āđ bowe pādšā tā zemānēwe nāđir šā mē serū dewrī} \\ 
\gll āđ ∅-b-o=we pādšā tā zemān-ē-ewe nāđir šā m-ē ser=ū dewr-ī \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} king\textsc{.m} until time\textsc{.m}\textsc{-indf}\textsc{-indf} \textsc{pn} king\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} on\textsc{=ez}\textsc{.gen} period\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (i.e., Jamsher Shah) became the king up until the time of Nader Shah.'
\z 
 
\ea \label{DP.54}
\textit{nāđir šā mē serū dewrī her wī berkenārē kero} \\ 
\gll nāđir šā m-ē ser=ū dewr-ī her wē-ī berkenār-ē ∅-ker-o \\ 
 \textsc{pn} king\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} on\textsc{=ez}\textsc{.gen} period\textsc{.m}\textsc{-obl}\textsc{.m} both two\textsc{-obl}\textsc{.m} ousted\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Nader Shah came to power [and] expelled both [Jamsher Shah and Sibhan Agha]'
\z 
 
\ea \label{DP.55}
\textit{to xeyrū min siɫāmet} \\ 
\gll to xeyr=ū min siɫāmet \\ 
 \textsc{2sg} goodness\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{1sg} healthy \\ 
\glt `Follow your own life, and I will follow mine.'
\z 
 
\ea \label{DP.56}
\textit{hezār seɫewāt serū peyġemerī} \\ 
\gll hezār seɫewāt ser=ū peyġemer-ī \\ 
 thousand praise\textsc{.m} on\textsc{=ez}\textsc{.gen} prophet\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `A thousand prayers [be] upon the prophet.'
\z 

\ea \label{ZP.2}
\textit{kuřū sey meʕrūfīn} \\ 
\gll kuř=ū sey meʕrūf-ī=n \\ 
 son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is Say Maʕruf’s son.'
\z 
 
\ea \label{ZP.3}
\textit{čī degāw hewrāmānīne serāwinū degākēwe je eyē bīyen} \\ 
\gll č=ī degā=ū hewrāmān-ī=ne serāwin=ū degā-ekē=we je eyē bīye=n \\ 
 in=dem.prox village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} top\_of\_village\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} from mother\textsc{.obl}\textsc{.f} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was born in the village of Hawraman, at the top point of the village.'
\z 
 
\ea \label{ZP.6}
\textit{ā šexse īnā ew desū degāw hewrāmānīne māčāš pene pīr ħeme yoso māmoš bīyen} \\ 
\gll ā šexs=e īnā-∅ ew des=ū degā=ū hewrāmān-ī=ne m-āč-ā=š pene pīr ħeme yoso māmo=š bīye=n \\ 
 \textsc{dem.dist} saint\textsc{=dem} deic\textsc{-3sg}\textsc{.m} \textsc{dem.dist} hand\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn} \textsc{pn} \textsc{pn} paternal\_uncle\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The saint [whose grave is] on the other side of Hawraman is called Pir Hama Yoso. He is his (Pir Shaliyar’s) paternal uncle.'
\z 
 
\ea \label{ZP.8}
\textit{ētir wextē eđāw tāteš fewtšā kerdenū emreū xwāyšā kerdēne berdenšā lāw wēšā tā māwāyē yāwān pene} \\ 
\gll ētir wext-ē eđā=ū tāte=š fewt=šā kerde=n=ū emre=ū xwā-ī=šā kerdē=ne berde=n=šā lā=ū wē=šā tā māwā-yē yāwā=n pene \\ 
 \textsc{disc.ptcl} time\textsc{.m}\textsc{-indf} mother\textsc{.f}=and father\textsc{.m}\textsc{=3sg} death\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} command\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} until period\textsc{.m}\textsc{-indf} arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \\ 
\glt `When his parents passed away and departed this life\footnote{Lit. did the command of God.}, they (the family of his uncle) took him into their house [and raised him] until he grew up [into a young man]'
\z 
 
\ea \label{ZP.9}
\textit{bīyen be qerārē tāwāniš bilo šuwānū sāwāɫā kero šuwānū ħeywānī kero} \\ 
\gll bīye=n be qerār-ē tāwā=n=iš bi-l-o šuwāne=ū sāwāɫe-ā ∅-ker-o šuwāne=ū ħeywān-ī ∅-ker-o \\ 
 be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} with arrangement\textsc{-indf} can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} shepherd\textsc{.m}\textsc{\textsc{=ez.gen}} lamb\textsc{-obl}\textsc{.pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} shepherd\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He was settled. He could go and shepherd lambs, sheep, and [other domestic] animals.'
\z 
 
\ea \label{ZP.10}
\textit{kīyāsenšā māčāš pene nehāɫū pīr šelīyārī řūbeřū biɫbeřī} \\ 
\gll kīyāse=n=šā m-āč-ā=š pene nehāɫ=ū pīr šelīyār-ī řūbeřū=ū biɫbeř-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They sent him to [a place] called the valley of Pir Shaliyar; [it is] on the opposite side of Bilbar\footnote{A village close to Hawraman Takht}.'
\z 
 
\ea \label{ZP.12}
\textit{čenū heywānī, čenū heywānī, čenū bizeɫā milo pāgene mēwe wēregāne,} \\ 
\gll čenū heywān-ī čenū heywān-ī čenū bizeɫā mi-l-o pāgene m-ē=we wēregā=ne \\ 
 with\textsc{.ez.gen} animal\textsc{.m}\textsc{-obl}\textsc{.m} with\textsc{.ez.gen} animal\textsc{.m}\textsc{-obl}\textsc{.m} with\textsc{.ez.gen} goat.dim\textsc{.obl}\textsc{.pl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} there \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} evening\textsc{=\textsc{post}} \\ 
\glt `He would go there with animals, goats [and such] and would be back in the evenings.'
\z 
 
\ea \label{ZP.14}
\textit{kīyāsenšā pey beġā pey xizmetū hezretū ġosī} \\ 
\gll kīyāse=n=šā pey beġā pey xizmet=ū hezret=ū ġos-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} for \textsc{pn} for service\textsc{.m}\textsc{\textsc{=ez.gen}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[Then] they sent him to Baghdah, to the presence of his highness the Ghaws.'
\z 
 
\ea \label{ZP.18}
\textit{yāwān pene, hezretū ġosī fermāwān exī mistefā rejilun kāmilun} \\ 
\gll yāwā=n pene hezret=ū ġos-ī fermāwā=n exī mistefā rejilun kāmilun \\ 
 arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} my\_brother \textsc{pn} man complete \\ 
\glt `He became experienced (in his field). His Highness the Ghaws said, ‘My brother, Mustafa, is a well-rounded man.’'
\z 
 
\ea \label{ZP.19}
\textit{yānē yāwān pene tāwū hem šerīʕetū hem terīqet tāwū girđ kārēwe kero} \\ 
\gll yānē yāwā=n pene ∅-tāw-ū hem šerīʕet=ū hem terīqet ∅-tāw-ū girđ kār-ē=we ∅-ker-o \\ 
 that\_means arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} both sharia\_law\textsc{\textsc{=ez.gen}} both denomination \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} every task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{post}} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This meant that he had become experienced (in his field). He could practise Sharia law, lead an order (of dervishes), and carry out all tasks [appointed to him]'
\z 
 
\ea \label{ZP.20}
\textit{kīyāsenšo pey menteqew hewrāmānī} \\ 
\gll kīyāse=n=š=o pey menteqe=ū hewrāmān-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}\textsc{=compl} for region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (Ghaws) sent him back to the Hawraman region.'
\z 
 
\ea \label{ZP.21}
\textit{kīyāsenšo pey menteqew hewrāmānīw āmāno menteqew hewrāmānī} \\ 
\gll kīyāse=n=š=o pey menteqe=ū hewrāmān-ī=ū āmā=n=o menteqe=ū hewrāmān-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}\textsc{=compl} for region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (Ghaws) sent him back to the Hawraman region. He came to the Hawraman region.'
\z 
 
\ea \label{ZP.23}
\textit{e pā menteqew hewrāmānī ētir feqīyēwe bīyenū wenānšū} \\ 
\gll e=p=ā menteqe=ū hewrāmān-ī ētir feqī-ēwe bīye=n=ū wenā=n=š=ū \\ 
 in=by=\textsc{dem.dist} region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} theologian\textsc{.m}\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} read\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}=and \\ 
\glt `In this Hawraman region, well, he was a theologian; he was literate.'
\z 
 
\ea \label{ZP.25}
\textit{čā dimāy, kināčēw pādšāw misrī neweše gino} \\ 
\gll čā dimā-ī kināčē=ū pādšā=w misr-ī neweš-e ∅-gin-o \\ 
 there afterwards\textsc{-obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} ill\textsc{-f} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `After a while, the king of Egypt’s daughter fell sick.'
\z 
 
\ea \label{ZP.31}
\textit{māčo, māčā neferē īnārē je menteqew hewrāmānātī be nāmēw fiɫāne kesī} \\ 
\gll m-āč-o m-āč-ā nefer-ē īnā-∅=rē je menteqe=ū hewrāmān-āt-ī be nāmē=ū fiɫān=e kes-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} person\textsc{.m}\textsc{-indf} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} from region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} to name\textsc{.f}\textsc{\textsc{=ez.gen}} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `It was said [to him in the dream] that there is a person in the Hawram region called such-and-such.'
\z 
 
\ea \label{ZP.33}
\textit{ī kināčēw to dermānū derdīš īnā lā ī pīyāywe’} \\ 
\gll ī kināčē=ū to dermān=ū derd-ī=š īnā-∅ lā ī pīyā-ī=we \\ 
 dem.prox daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} \textsc{2sg} treatment\textsc{.m}\textsc{\textsc{=ez.gen}} illness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \\ 
\glt `‘This daughter of yours, the treatment for her illness lies with this man (i.e., Pir Shaliyar)'
\z 
 
\ea \label{ZP.36}
\textit{īne māčmēš pene šālīyāre sīyāw lālow kināčekēn} \\ 
\gll īne m-āč-mē=š pene šālīyār=e sīyāw lālo=ū kināčē-ekē=n \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn}\textsc{=dem} black maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This [person], whom we call Shaliyar Siya, is the maternal uncle of the [king’s] daughter.'
\z 
 
\ea \label{ZP.38}
\textit{kināčē mido desū lāloyš} \\ 
\gll kināčē mi-de-o des=ū lālo-ī=š \\ 
 girl\textsc{.obl}\textsc{.f} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{\textsc{=ez.gen}} maternal\_uncle\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `and delivered [his] daughter to her uncle.'
\z 
 
\ea \label{ZP.39}
\textit{māčo, ī kināčē bere degā be degāw} \\ 
\gll m-āč-o ī kināčē ∅-ber-e degā be degā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.obl}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `[and] said, ‘Take this girl village by village,'
\z 
 
\ea \label{ZP.41}
\textit{degā be degāw wirde wirde ārdēnešū āmān řūbeřū hewrāmānātī} \\ 
\gll degā be degā=ū wirde wirde ārdē=ne š=ū āmā=n řūbeřū=ū hewrāmān-āt-ī \\ 
 village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} little little bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} go\textsc{.prs}\textsc{\textsc{=ez.gen}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the uncle) led her village by village and came [to a place] next to the Hawraman River.'
\z 
 
\ea \label{ZP.44}
\textit{to eger ī kināčē berdet xizmetū ī pīyāy} \\ 
\gll to eger ī kināčē berd-e=t xizmet=ū ī pīyā-ī \\ 
 \textsc{2sg} if dem.prox daughter\textsc{.dir}\textsc{.f} take\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=\textsc{2sg}} service\textsc{.m}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `‘When you have taken this girl into the presence of this man'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.50}
\textit{tewenē īnāre beynū biɫbeřīnū silēnēne} \\ 
\gll tewenī-ē īnā-∅=re beyn=ū biɫbeř-ī=ne=ū silēnē=ne \\ 
 stone\textsc{.f}\textsc{-indf} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} among\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `There is a stone between Bilbar and Slen.'
\z 
 
\ea \label{ZP.51}
\textit{ētir kināče keřū lāɫe bo māčmēš pene tewenū sirewē} \\ 
\gll ētir kināčē=e keře=ū lāɫ-e ∅-b-o m-āč-mē=š pene tewenī=ū sirewē \\ 
 \textsc{disc.ptcl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} deaf\textsc{.f}=and mute\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} \\ 
\glt `Since the girl was deaf-mute, we call it the healing stone.'
\z 
 
\ea \label{ZP.52}
\textit{pāse māčāš pene tewenū sirewē kināčekē mārū serū tewenekē} \\ 
\gll pāse m-āč-ā=š pene tewenī=ū sirewē kināčē-ekē m-ār-ū ser=ū tewenī-ekē \\ 
 like \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-\textsc{1sg}} on\textsc{\textsc{=ez.gen}} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `It is called the healing stone. He (the uncle) took the girl to the top of the stone.'
\z 
 
\ea \label{ZP.55}
\textit{mārā serū tewenekēwe} \\ 
\gll m-ār-ā ser=ū tewenī-ekē=we \\ 
 \textsc{ind-}bring\textsc{.prs}\textsc{-3pl} on\textsc{\textsc{=ez.gen}} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `He (the uncle) took the girl to the top of the stone.'
\z 
 
\ea \label{ZP.56}
\textit{nīšāre čenū lāloekeyš} \\ 
\gll ∅-nīš-ā=re čenū lālo-ekey=š \\ 
 \textsc{ind-}sit\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} with\textsc{.ez.gen} maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `She sat [on the stone] with her uncle.'
\z 
 
\ea \label{ZP.59}
\textit{pokīš māčā pene tewenū sirewāy} \\ 
\gll pokī=š m-āč-ā pene tewenī=ū sirewāy \\ 
 that\_is\_why\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment.inf \\ 
\glt `That is why it is called the healing stone.'
\z 
 
\ea \label{ZP.61}
\textit{her ā dāđe kero, ētir māčo, erē weɫā sirewā’ hurmēzowe miđo wene māroš hesārū činārārū} \\ 
\gll her ā dāđ=e ∅-ker-o ētir m-āč-o erē weɫā sirewā hur-m-ēz-o=we mi-đe-o wene m-ār-o=š hesār=ū činār-ā=re=ū \\ 
 emph \textsc{dem.dist} yell\textsc{.m}\textsc{=dem} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes indeed be cured\textsc{.pst}\textsc{.3sg} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} at \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} pond\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \\ 
\glt `After he shouted [and] said, ‘Indeed she is healed’, then he rose, [and] set off, and took the girl to Hasar Chinara.'
\z 
 
\ea \label{ZP.62}
\textit{mēnē mizyā werāwerū degāw hewrāmānī} \\ 
\gll m-ē-nē mi-zye-ā werāwer=ū degā=ū hewrāmān-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They ended up at the opposite side of Hawraman village.'
\z 
 
\ea \label{ZP.63}
\textit{mizyo, mizyā werāwerū māčmēš pene dūtifɫe} \\ 
\gll mi-zye-o mi-zye-ā werāwer=ū m-āč-mē=š pene dūtifɫe \\ 
 \textsc{ind-}go\_out\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn} \\ 
\glt `[Then] they left [Hawraman village and went to a place which] we call Dutifla'
\z 
 
\ea \label{ZP.64}
\textit{mizyā werāwerū dūtifɫey} \\ 
\gll mi-zye-ā werāwer=ū dūtifɫey \\ 
 \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `[Then] they left Dutifla'
\z 
 
\ea \label{ZP.65}
\textit{pīr šelīyār ā vextī yānekew yānew tāteyšne bo, pū ser degāwe} \\ 
\gll pīr šelīyār ā vext-ī yāne-eke=ū yāne=ū tāte-ī=š=ne ∅-b-o p=ū ser degā=we \\ 
 \textsc{pn} \textsc{pn} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} house\textsc{.m}\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at=\textsc{dem.dist}\textsc{.3sg} on village\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar, well, he was [living] in his father’s house, [which was located] at the top of the village'
\z 
 
\ea \label{ZP.66}
\textit{čemiš čene bo wextē miđyo īnā kināčekē mē řārew čenū pīyēwe mēnē} \\ 
\gll čem=iš čene ∅-b-o wext-ē mi-đye-o īnā-∅ kināčē-ekē m-ē řā=re=ū čenū pīyā-ēwe m-ē-nē \\ 
 eye\textsc{=3sg} at \textsc{ind-}be\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} deic\textsc{-3sg}\textsc{.m} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} road\textsc{.f}\textsc{=\textsc{post}}=and with\textsc{.ez.gen} man\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \\ 
\glt `He (Pir Shaliyar) was staring at her [from afar]. He noticed she was coming [towards him] with a man.'
\z 
 
\ea \label{ZP.73}
\textit{īse māčmēš pene tenūrū dēwā} \\ 
\gll īse m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 now \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Now we call that place Tanūrū dēwā ‘the demons’ oven’'
\z 
 
\ea \label{ZP.74}
\textit{ā degākē ehā čew deso bīyēne duē kutē māčmēš pene tenūrū dēwā} \\ 
\gll ā degā-ekē ehā č=ew des=o bīyē=ne duē kut-ē m-āč-mē=š pene tenūr=ū dēw-ā \\ 
 \textsc{dem.dist} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{intj} in=\textsc{dem.dist} hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} two half\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `Do you see that the village is divided in half on the other side? We call that place Tanūrū dēwā.'
\z 
 
\ea \label{ZP.75}
\textit{mē ewā her gerūberū kināčekē lāšexeɫ mekero} \\ 
\gll m-ē ewā her gerūber=ū kināčē-ekē lāšexeɫ me-ker-o \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} there emph around\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} empty \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `They (the girl and her uncle) arrived there. He (the demon) would not leave the girl alone.'
\z 
 
\ea \label{ZP.76}
\textit{gerūberū kināčekē lāšexeɫ mekero} \\ 
\gll gerūber=ū kināčē-ekē lāšexeɫ me-ker-o \\ 
 around\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} empty \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the demon) would not leave the girl alone.'
\z 
 
\ea \label{ZP.77}
\textit{mē tenūrū dēwā} \\ 
\gll m-ē tenūr=ū dēw-ā \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `They arrived at Tanūrū dēwā.'
\z 
 
\ea \label{ZP.80}
\textit{werū čemāw girdū mexeleqeɫāyre til bowe ginore ā dere qoɫe} \\ 
\gll wer=ū čem-ā=ū gird=ū mexeleqeɫā-ī=re til ∅-b-o=we gin-o=re ā dere qoɫ=e \\ 
 out\_of\textsc{\textsc{=ez.gen}} eye\textsc{-3pl}\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl} fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{dem.dist} valley\textsc{.m} deep\textsc{=dem} \\ 
\glt `in front of people’s eyes, he (the demon) rolled down and fell into a deep valley.'
\z 
 
\ea \label{ZP.82}
\textit{kināčē ētir mēwe serū hāɫē weše bowew} \\ 
\gll kināčē ētir m-ē=we ser=ū hāɫe-ē weš-e ∅-b-o=we=ū \\ 
 girl\textsc{.obl}\textsc{.f} \textsc{disc.ptcl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} on\textsc{\textsc{=ez.gen}} state\textsc{.f}\textsc{-obl}\textsc{.f} well\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl}\textsc{\textsc{=ez.gen}} \\ 
\glt `Then, the girl was cured. She was healed.'
\z 
 
\ea \label{ZP.83}
\textit{mēnē dilū āweyānīū} \\ 
\gll m-ē-nē dil=ū āweyānī=ū \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} inside\textsc{\textsc{=ez.gen}} habitat\textsc{.m}=and \\ 
\glt `They (the girl and her uncle) went into the crowd'
\z 
 
\ea \label{ZP.84}
\textit{persā, ēme yānew fiɫāne kesīmā gerekā} \\ 
\gll ∅-pers-ā ēme yāne=ū fiɫān=e kes-ī=mā gerek=ā \\ 
 \textsc{ind-}ask\textsc{.prs}\textsc{-3pl} \textsc{1pl} house\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} necessary\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[and] asked [people], ‘We are looking for\footnote{Lit. we want.} the house of such-and-such a person.'
\z 
 
\ea \label{ZP.86}
\textit{māčā xeɫkīč, ī pīyā šimeš pey persdē xo yānēw e čāneš nīyāw} \\ 
\gll m-āč-ā xeɫk=īč ī pīyā šime=š pey ∅-pers-dē xo yāne-ē=ū e=čāne=š nīye=ā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} people\textsc{.m}\textsc{=add} dem.prox man\textsc{.m} \textsc{2pl}\textsc{=3sg} for \textsc{ind-}ask\textsc{.prs}-\textsc{2pl} dis.ptc house\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} in=like\_that\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The people said, ‘This man whom you ask for, he does not have a proper house.'
\z 
 
\ea \label{ZP.89}
\textit{kināčēw pādšāw fiɫān henānew miniš kerdenā be wekēɫ} \\ 
\gll kināčē=ū pādšā=w fiɫān hen-ā=ne=ū min=iš kerde=nā be wekēɫ \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such place\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \textsc{1sg}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to ad\textsc{voc}ate\textsc{.m} \\ 
\glt `She is the daughter of … the king of such-and-such place has put me in charge.\footnote{Lit. made me advocate.}'
\z 
 
\ea \label{ZP.93}
\textit{kināčēw pādšāy mārā lā pīr šelīyārī} \\ 
\gll kināčē=ū pādšā-ī m-ār-ā lā pīr šelīyār-ī \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl} at\_the\_place\_of \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They brought the king’s daughter to Pir Shaliyar.'
\z 
 
\ea \label{ZP.96}
\textit{počīš māčēnē zārātū pīr šelīyārī} \\ 
\gll počī=š m-āč-ēn-ē zārāt=ū pīr šelīyār-ī \\ 
 that\_is\_why\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-pstc}\textsc{-3pl} corn\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `That is why they (the farms) are called Pir Shaliyar’s farms.'
\z 
 
\ea \label{ZP.97}
\textit{ā wextī mārāš mē ʕerzū pīr šelīyārī kero lāloekeš} \\ 
\gll ā wext-ī m-ār-ā=š m-ē ʕerz=ū pīr šelīyār-ī ∅-ker-o lālo-eke=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \\ 
\glt `They (the people) brought her. They came (to Pir Shaliyar). The uncle said to Pir Shaliyar,'
\z 
 
\ea \label{ZP.100}
\textit{ī kināčē eger weše bīyowe mišyo māreš biřū pey to be mezebū wētāw} \\ 
\gll ī kināčē eger weš-e bīyē-o=we mišyo māre=š ∅-biř-ū pey to be mezeb=ū wē=tā=ū \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} if well\textsc{-f} be\textsc{.prs}.\textsc{cond}\textsc{.f}?\textsc{-3sg}\textsc{=compl} aux marraige\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for \textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `If the girl is healed, I shall marry her to you, according to your religion.'
\z 
 
\ea \label{ZP.102}
\textit{īne pādšāzāyēnew derdū min meweronew} \\ 
\gll īne pādšāzāyē=ne=ū derd=ū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}=and illness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She is no use to me.'
\z 
 
\ea \label{ZP.104}
\textit{īne pādšāzāyēne čenū min meweronew} \\ 
\gll īne pādšāzāyē=ne čenū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}} with\textsc{.ez.gen} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She will not eat with me [under the same roof].'
\z 
 
\ea \label{ZP.105}
\textit{min čīwēwim nīyārē lāyiqū ī pādšāzāye bo} \\ 
\gll min čīw-ēw=im nīye=ā=rē lāyiq=ū ī pādšāzāye ∅-b-o \\ 
 \textsc{1sg} thing\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} deserved\textsc{\textsc{=ez.gen}} dem.prox princess\textsc{.obl}\textsc{.f} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `I have nothing which would be up to the standards of this princess.'
\z 
 
\ea \label{ZP.106}
\textit{berešo. mubārekū sāhēbīš bo!’} \\ 
\gll ∅-ber-e=š=o mubārek=ū sāhēb-ī=š ∅-b-o \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=compl} blessed\textsc{\textsc{=ez.gen}} owner\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take her back. May she be happy with her owner (i.e., father)!'
\z 
 
\ea \label{ZP.114}
\textit{qisē kerā beynū wēšāne} \\ 
\gll qisē ∅-ker-ā beyn=ū wē=šā=ne \\ 
 talk\textsc{.dir}\textsc{.f} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} among\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `They talked among themselves.'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.120}
\textit{yoīč māčo, minīč beʕzē tiršū henārāt midonew’} \\ 
\gll yo=īč m-āč-o min=īč beʕzē tirš=ū henār-ā=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} some molasses\textsc{\textsc{=ez.gen}} pomegranate\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt ` Another said, ‘I will give you some pomegranate molasses.’'
\z 
 
\ea \label{ZP.123}
\textit{merāsēmekey řās kermēwew} \\ 
\gll merāsēm-ekey řās ∅-ker-mē=we=ū \\ 
 ceremony\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} true \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl}\textsc{=compl}\textsc{\textsc{=ez.gen}} \\ 
\glt `And carry out the ceremony.'
\z 
 
\ea \label{ZP.128}
\textit{weɫāhī īne šime māčdēšū īnā vistāwe řā pey min čūn šime be xātirū xwāytā kerd hetā xwā dinyāš weše kerdēne ī merāsēme her bo her meno’} \\ 
\gll weɫāhī īne šime m-āč-dē=š=ū īnā-∅ vist=tā=we řā pey min čūn šime be xātir=ū xwā-ī=tā kerd hetā xwā dinyā=š weš-e kerdē=ne ī merāsēm=e her ∅-b-o her ∅-men-o \\ 
 indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} \textsc{ind-}say\textsc{.prs}-\textsc{2pl}\textsc{=3sg}=and deic\textsc{-3sg}\textsc{.m} throw\textsc{.pst}=\textsc{2pl}\textsc{=compl} road\textsc{.f} for \textsc{1sg} because \textsc{2pl} to sake\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}be\textsc{.prs}\textsc{-3sg} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `Indeed, what you are saying and the way you organised [this ceremony] for me, since you did it in good faith, so long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ZP.129}
\textit{šūnū ēmeyčre her meno} \\ 
\gll šūn=ū ēme=yč=re her ∅-men-o \\ 
 after\textsc{\textsc{=ez.gen}} \textsc{1pl}\textsc{=add}\textsc{=\textsc{post}} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `It will continue to exist after us as well.'
\z 
 
\ea \label{ŽP.3}
\textit{māčāš pīr šelīyār yānē lūān pēwāw xwāywe} \\ 
\gll m-āč-ā=š pīr šelīyār yānē lūā=n pēwā=ū xwā=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{pn} \textsc{pn} that\_means go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} welcoming\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He is referred to as Pir Shaliyar, meaning that he has gone [and] embraced God.'
\z 
 
\ea \label{ŽP.6}
\textit{ānen īse īnā lāw hānew degē} \\ 
\gll āne=n īse īnā-∅ lā=ū hāne=ū degē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} water\_spring\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f} \\ 
\glt `That (his shrine) is on the other side of the village.'
\z 
 
\ea \label{ŽP.11}
\textit{milo lāw māmoyš māmo yosoy} \\ 
\gll mi-l-o lā=ū māmo-ī=š māmo yoso-ī \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3sg} paternal\_uncle \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He went to his uncle, Uncle Yoso.'
\z 
 
\ea \label{ŽP.13}
\textit{ētir milo lāw āđī} \\ 
\gll ētir mi-l-o lā=ū āđī \\ 
 \textsc{disc.ptcl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt ` Anyhow, he went to him.'
\z 
 
\ea \label{ŽP.16}
\textit{tāwo šuwānew bizeɫā kero, tāwo šuwāneū heywānī kero} \\ 
\gll ∅-tāw-o šuwāne=ū bizeɫā ∅-ker-o ∅-tāw-o šuwāne=ū heywān-ī ∅-ker-o \\ 
 \textsc{ind-}can\textsc{.prs}\textsc{-3sg} shepherd\textsc{.m}\textsc{\textsc{=ez.gen}} goat.dim\textsc{.obl}\textsc{.pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} shepherd\textsc{.m}=and animal\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He was able to tend goats; he was able to tend (domestic) animals.'
\z 
 
\ea \label{ŽP.17}
\textit{šuwānew heywānī bo,} \\ 
\gll šuwāne=ū heywān-ī ∅-b-o \\ 
 shepherd\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He was a shepherd.'
\z 
 
\ea \label{ŽP.19}
\textit{ānā ew dīmū qeɫāw kitāy} \\ 
\gll āne=ā ew dīm=ū qeɫā=ū kitā-ī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `—Look, it is [located] on the other side of the Kita castle'
\z 
 
\ea \label{ŽP.20}
\textit{yāgē hen māčāš nehāɫū pīr šelīyārā—} \\ 
\gll yāgē-ē hen-∅ m-āč-ā=š nehāɫ=ū pīr šelīyār-ā \\ 
 place\textsc{.f}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \\ 
\glt `There is a place called the valley of Pir Shaliyar.'
\z 
 
\ea \label{ŽP.22}
\textit{heremešū pīr heme yoso kīyānāš čenū bizā pey ewāgey} \\ 
\gll hereme=š=ū pīr heme yoso ∅-kīyān-ā=š čenū bize-ā pey ewāge-ī \\ 
 wife\textsc{.f}\textsc{=3sg}=and \textsc{pn} \textsc{pn} \textsc{pn} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} with\textsc{.ez.gen} goat\textsc{.f}\textsc{-obl}\textsc{.pl} to there\textsc{-obl}\textsc{.m} \\ 
\glt `Pir Hama Yoso and his wife sent him there [for shepherding] together with the goats.'
\z 
 
\ea \label{ŽP.27}
\textit{māčā, dey to metāwī -sey misefāw- kučēš yewē berī nehāɫū pīr šelīyārā šānīšā pižgīnīšā bā bizyā’} \\ 
\gll m-āč-ā dey to me-tāw-ī sey misefā=ū kuč-ē=š yewe-ē ∅-ber-ī nehāɫ=ū pīr šelīyār-ā ∅-šān-ī=šā ∅-pižgīn-ī=šā bā bi-zye-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{2sg} \textsc{neg-}can\textsc{.prs}-\textsc{2sg} \textsc{pn} \textsc{pn}=and little\textsc{-indf}\textsc{=3sg} barley\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \textsc{sbjv-}throw\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}scatter\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \textsc{hort} \textsc{sbjv-}grow\textsc{.prs}\textsc{-3pl} \\ 
\glt `They said, ‘You are not able, Say Mustafa, to take some barley seeds to the valley of Pir Shaliyar, [and] scatter them [on the field], [and] plant them, so that they grow.'
\z 
 
\ea \label{ŽP.28}
\textit{ētir xo āđ meselen hetīm bo, lāw māmoyšā bo řūeš mebo wāčo, yā metāwū yā tāwū} \\ 
\gll ētir xo āđ meselen hetīm ∅-b-o lā=ū māmo-ī=šā ∅-b-o řūe=š me-b-o ∅-wāč-o yā me-tāw-ū yā ∅-tāw-ū \\ 
 \textsc{disc.ptcl} dis.ptc 3sg\textsc{.dir}\textsc{.m} for\_example orphan \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} face\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-3sg} either \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} either \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Well, he was an orphan. He was at his uncle’s. He was ashamed to say, ‘I cannot, or I can.’'
\z 
 
\ea \label{ŽP.29}
\textit{werū mejbūrī yewāyč bero} \\ 
\gll wer=ū mejbūrī yewe-ā=yč ∅-ber-o \\ 
 out\_of\textsc{\textsc{=ez.gen}} obligation barley\textsc{.f}\textsc{-3pl}\textsc{=add} \textsc{ind-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt ` Out of obligation, he took the barley seeds, too.'
\z 
 
\ea \label{ŽP.31}
\textit{milo ew dīmū qeɫāw kitā} \\ 
\gll mi-l-o ew dīm=ū qeɫā=w kitā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `He went to the other side of the Kita castle.'
\z 
 
\ea \label{ŽP.34}
\textit{zengenekey minīyowe pāyīnū zemīnekey} \\ 
\gll zengene-ekey mi-nye-o=we pāyīn=ū zemīn-ekey \\ 
 hoe\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}put\textsc{.prs}\textsc{-3sg}\textsc{=compl} bottom\textsc{\textsc{=ez.gen}} land\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `He put the hoe at the south of the land.'
\z 
 
\ea \label{ŽP.43}
\textit{māčo bē dengē māmo yoso, dā lue šonū sey misefāyre bizāne řās māčo!} \\ 
\gll m-āč-o bē- deng-ē māmo yoso dā ∅-lu-e šūn=ū sey misefā-ī=re bi-zān-e řās m-āč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}without noise\textsc{.m}\textsc{-indf} paternal\_uncle \textsc{pn} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} after\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} truth \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \\ 
\glt `Uncle Yoso secretly said [to the servant], ‘Go after Say Mustafa, see if he is telling the truth!'
\z 
 
\ea \label{ŽP.48}
\textit{šuwānew bizāyč bo} \\ 
\gll šuwāne=ū bize-ā=yč ∅-b-o \\ 
 shepherd\textsc{.m}\textsc{\textsc{=ez.gen}} goat\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=add} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `And he is shepherding the goats.'
\z 
 
\ea \label{ŽP.57}
\textit{pīyāke mēwe ʕerzū heme yosoy kero māčo, weɫāhī qurbān min tūšū jeryānē e čīney āmānē} \\ 
\gll pīyā-eke m-ē=we ʕerz=ū heme yoso-ī ∅-ker-o m-āč-o weɫāhī qurbān min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sir.\textsc{voc} \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `The man returned [and said] to Hama Yoso, ‘Indeed, Sir! I have witnessed an event like this.'
\z 
 
\ea \label{ŽP.58}
\textit{min tūšū jeryānē e čīney āmānē} \\ 
\gll min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `I have witnessed an event like this.’'
\z 
 
\ea \label{ŽP.62}
\textit{wēregā pīr šelīyār mēwe sey misefā wextē mēwe, ʕerzū ženekēš kero māčo, bā ēšew īmti̩hānēš kermē lāw wēmāwe newsinmēš} \\ 
\gll wēregā pīr šelīyār m-ē=we sey misefā wext-ē m-ē=we ʕerz=ū ženī-ekē=š ∅-ker-o m-āč-o bā ēšew īmti̩hān-ē=š ∅-ker-mē lā=ū wē=mā=we ne-ūsin-mē=š \\ 
 evening \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{pn} \textsc{pn} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} tonight test\textsc{-indf}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=1pl}\textsc{=\textsc{post}} \textsc{neg-}put\_to\_sleep\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \\ 
\glt `Pir Shaliyar returned in the evening; when Say Mustafa returned, he (Hama Yoso) said to his wife, ‘Let’s test him tonight; let’s not put him to sleep next to us.’'
\z 
 
\ea \label{ŽP.72}
\textit{ʕerzū heremēš kero, fermāwo be heremēš weɫāhī ētir ēme metāwmē, sey misefā metāwmē hīč kārēš pene kermē īlā kīyānmēš bilo weno, bilo kīyānmēš weno} \\ 
\gll ʕerz=ū hereme-ē=š ∅-ker-o ∅-fermāw-o be hereme-ē=š weɫāhī ētir ēme me-tāw-mē sey misefā me-tāw-mē hīč kār-ē=š pene ∅-ker-mē īlā ∅-kīyān-mē=š bi-l-o ∅-wen-o bi-l-o ∅-kīyān-mē=š ∅-wen-o \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} indeed \textsc{disc.ptcl} \textsc{1pl} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} \textsc{pn} \textsc{pn} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} except \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}read\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}read\textsc{-3sg} \\ 
\glt `He (Pir Hama Yoso) said to his wife, ‘Indeed, we can no longer be helpful to Say Mustafa,\footnote{Lit. we can do nothing to him.} except that we send him to study; that we send him, to go [and] study.'
\z 
 
\ea \label{ŽP.80}
\textit{je her yāgē weno, čā dimāyē wextē miđyā ēnneš wenā qābilyetiš peyā kerd kīyānāš pey beġāy pey xizmetū, kīyānāš pey beġāy wenāy} \\ 
\gll je her yāgē-ē ∅-wen-o č=ā dimā-ē wext-ē mi-đye-ā ēnne=š wenā qābilyet=iš peyā kerd ∅-kīyān-ā=š pey beġā-ī pey xizmet=ū ∅-kīyān-ā=š pey beġā-ī wenāy \\ 
 from emph place\textsc{.f}\textsc{-indf} \textsc{ind-}read\textsc{-3sg} in=\textsc{dem.dist} afterwards\textsc{-obl}\textsc{.f} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} so\_much\textsc{=3sg} read\textsc{.pst} capability\textsc{.m}\textsc{=3sg} visible do\textsc{.pst} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn}\textsc{-obl}\textsc{.m} to service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn}\textsc{-obl}\textsc{.m} read.inf \\ 
\glt `Wherever he studied, after a while, they realised that he studied to the extent that he deserved to be sent to Baghdad into the presence of his highness [Khos]'
\z 
 
\ea \label{ŽP.82}
\textit{šūnū miqdārēre hezretū xosī fermāwo duē yerē īmtihānēš kero} \\ 
\gll šūn=ū miqdārē=re hezret=ū xos-ī ∅-fermāw-o duē yerē īmtihān-ē=š ∅-ker-o \\ 
 after\textsc{\textsc{=ez.gen}} a\_little\textsc{=\textsc{post}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} two three test\textsc{-dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `After a while, his Highness Khos said … He tested him two, three times.'
\z 
 
\ea \label{ŽP.83}
\textit{īmtihānēš kero, fermāwo, exī mistefā rejilun kāmilun yānē be ʕerebī wēšā pāseš fermāwān hezretū xosī} \\ 
\gll īmtihān-ē=š ∅-ker-o ∅-fermāw-o exī mistefā rejilun kāmilun yānē be ʕerebī wē=šā pāse=š fermāwā=n hezret=ū xos-ī \\ 
 test\textsc{-dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} my\_brother \textsc{pn} man complete that\_means by Arabic \textsc{reflx}\textsc{=3pl} this\textsc{=3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He tested him [and] said, ‘My brother Mustafa is a well-rounded man, which in their Arabic means, his Highness Khos said,'
\z 
 
\ea \label{ŽP.87}
\textit{mišyo bilowe pey menteqew wēšā xizmet be merđimī kero} \\ 
\gll mišyo bi-l-o=we pey menteqe=ū wē=šā xizmet be merđim-ī ∅-ker-o \\ 
 aux \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} to region\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} service\textsc{.m} by people\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He should go back to his region [and] and be of service to people.'
\z 
 
\ea \label{ŽP.91}
\textit{fermāwo hezretū xos, dey to mišyo tešrīf berī bilīwe, pey menteqew wētāw} \\ 
\gll ∅-fermāw-o hezret=ū xos dey to mišyo tešrīf ∅-ber-ī bi-l-ī=we pey menteqe=ū wē=tā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{disc.ptcl} \textsc{2sg} aux honouring \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} to region=and \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `His Highness Khos said, ‘You should leave [here and] return to your region.'
\z 
 
\ea \label{ŽP.92}
\textit{bilī čāy xizmetū mirīđū mensūb weš kerīw} \\ 
\gll bi-l-ī čā-ī xizmet=ū mirīđ=ū mensūb weš ∅-ker-ī=ū \\ 
 \textsc{sbjv-}go\textsc{.prs}-\textsc{2sg} there\textsc{-obl}\textsc{.m} service\textsc{.m}\textsc{\textsc{=ez.gen}} adherent=and appointees well \textsc{sbjv-}do\textsc{.prs}\textsc{-obl}\textsc{.m}=and \\ 
\glt `You should go there and be of good service to adherents and appointees, '
\z 
 
\ea \label{ŽP.97}
\textit{řāne mēnē pāne yā guzerēwene mesīrēwene čenū hejījīyē ginā pēwere} \\ 
\gll řā=ne m-ē-nē p=āne yā guzer-ēwe=ne mesīr-ēwe=ne čenū hejījī-ē ∅-gin-ā pēwere \\ 
 road\textsc{.f}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} either path\textsc{-indf}\textsc{=\textsc{post}} path\textsc{-indf}\textsc{=\textsc{post}} with\textsc{.ez.gen} from\_Hajij\textsc{-indf} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl} together \\ 
\glt `On the road, on a path, he ran into a person from Hajij.'
\z 
 
\ea \label{ŽP.100}
\textit{her mēnē mēnē mēnē nizīkū hejījī gināwe} \\ 
\gll her m-ē-nē m-ē-nē m-ē-nē nizīk=ū hejīj-ī ∅-gin-ā=we \\ 
 emph \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{post}} \\ 
\glt ` They kept walking. They arrived [at a place] close to Hajij.'
\z 
 
\ea \label{ŽP.101}
\textit{nizīkū hejījī gināwe miqdārē zeřšā pene bo} \\ 
\gll nizīk=ū hejīj-ī ∅-gin-ā=we miqdārē zeř=šā pene ∅-b-o \\ 
 close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl}\textsc{=compl} a\_little money\textsc{.m}\textsc{=3pl} with \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `They\footnote{Lit. He.} arrived [at a place] close to Hajij. They had some money with them\footnote{Lit. him.}.'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.103}
\textit{to lue zeřekey bere čenū wēt pey xerjīt lodē} \\ 
\gll to ∅-lu-e zeř-ekey ∅-ber-e čenū wē=t pey xerjī=t ∅-lo-dē \\ 
 \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} with\textsc{.ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} to consumption\textsc{=\textsc{2sg}} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} \\ 
\glt `You go, take the money with you for your expenses.'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.107}
\textit{be zor miđoš milū sey misefāyre zerey} \\ 
\gll be zor mi-đe-o=š mil=ū sey misefā-ī=re zer-e=y \\ 
 by force \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=3sg} on\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} money\textsc{.m}\textsc{-def}\textsc{=ez}.\textsc{attr} \\ 
\glt `He (the man from Hajij) did not take no for an answer\footnote{Lit. with force.} [and] gave him (Pir Shaliyar) the money.'
\z 
 
\ea \label{ŽP.110}
\textit{be kerāmāt kīyānošo zeřeš pey kīyānowew} \\ 
\gll be kerāmāt ∅-kīyān-o=š=o zeř-e=š pey ∅-kīyān-o=we=ū \\ 
 by supernatural\_power\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} money\textsc{.m}\textsc{-def}\textsc{=3sg} to \textsc{ind-}send\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} \\ 
\glt `Using his supernatural powers, he had sent it back; he had sent back the money.'
\z 
 
\ea \label{ŽP.116}
\textit{mēwe ētir ā zemāne pīr šelīyār mēwe menteqe des kero, jārē āne minyā māčāš meřū āsinī} \\ 
\gll m-ē=we ētir ā zemān=e pīr šelīyār m-ē=we menteqe des=t ∅-ker-o jārē āne mi-nye-ā m-āč-ā=š meře=ū āsin-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} once \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) returned [to Hawraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ‘Iron cave.’'
\z 
 
\ea \label{ŽP.117}
\textit{ētir hezretū xosī kīyāseno} \\ 
\gll ētir hezret=ū xos-ī kīyāse=n=o \\ 
 \textsc{disc.ptcl} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `Anyhow, His Highness Khos had sent him.'
\z 
 
\ea \label{ŽP.118}
\textit{lūānre meřū āsinī čilēš kēštēne} \\ 
\gll lūā=n=re meře=ū āsin-ī čilē=š kēštē=ne \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} forty\_days\_fasting\textsc{.f}\textsc{=3sg} pull\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `He went to the Maru Asini and fasted for forty days.'
\z 
 
\ea \label{ŽP.121}
\textit{e čā meřū āsnīne čilēš kēštēneū} \\ 
\gll e=čā meře=ū āsin-ī=ne čilē=š kēštē=ne=ū \\ 
 in=there cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} forty\_days\_fasting\textsc{.f}\textsc{=3sg} pull\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}}=and \\ 
\glt `He would fast rounds of forty days of fasting in that Iron Cave.'
\z 
 
\ea \label{ŽP.122}
\textit{lūān pāsārū leme čilēš kēštēneū} \\ 
\gll lūā=n pāsār=ū leme čilē=š kēštē=ne=ū \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{\textsc{=ez.gen}} \textsc{pn} forty\_days\_fasting\textsc{.f}\textsc{=3sg} pull\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}}=and \\ 
\glt `He would go to [a place called] Pasaru Lama ‘guardians of the belly’ and would fast for forty days,'
\z 
 
\ea \label{ŽP.124}
\textit{ā wextī xeberē yāwno hezretū xosī} \\ 
\gll ā wext-ī xeber-ē ∅-yāwn-o hezret=ū xos-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} news\textsc{.m}\textsc{-indf} \textsc{ind-}cause\_to\_arrive\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, the news reached His Highness Khos.'
\z 
 
\ea \label{ŽP.129}
\textit{hezretū xosī be mexfī pīyāy kīyāno menteqew hewrāmānātī} \\ 
\gll hezret=ū xos-ī be mexfī pīyā-ī ∅-kīyān-o menteqe=ū hewrāmān-āt-ī \\ 
 his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} by secret man\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} region=and \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} \\ 
\glt `His Highness Khos secretly sent a man to the Hawraman region.'
\z 
 
\ea \label{ŽP.131}
\textit{miđyā īnā meřū āsinī} \\ 
\gll mi-đye-ā īnā-∅ meře=ū āsin-ī \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `They realised that he was in the Iron Cave.'
\z 
 
\ea \label{ŽP.132}
\textit{pīyākew hezretū xosī milo meřū āsinī tūšiš mē} \\ 
\gll pīyā-eke=ū hezret=ū xos-ī mi-l-o meře=ū āsin-ī tūš=iš m-ē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} encounter\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `His Highness Khos’s man went [and] ran into him (Pir Shaliyar) in the Iron cave.'
\z 
 
\ea \label{ŽP.133}
\textit{meřū āsinī tūšiš mē} \\ 
\gll meře=ū āsin-ī tūš=iš m-ē \\ 
 cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} encounter\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `He ran into him (Pir Shaliyar) in the Iron Cave.'
\z 
 
\ea \label{ŽP.134}
\textit{xo āđ mezāno hezretū xosī kīyāseniš} \\ 
\gll xo āđ me-zān-o hezret=ū xos-ī kīyāse=n=iš \\ 
 dis.ptc 3sg\textsc{.dir}\textsc{.m} \textsc{neg-}know\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \\ 
\glt `Well, he (Pir Shaliyar) did not realise that His Highness Khos had sent him (the man)'
\z 
 
\ea \label{ŽP.141}
\textit{mē mēwe menteqew mēwe dilū hewrāmānīū yānešā yānešā e čā bo milo wā} \\ 
\gll m-ē m-ē=we menteqe=ū m-ē=we dil=ū hewrāmān-ī=ū yāne=šā yāne=šā e=čā ∅-b-o mi-l-o wā \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region=and \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} inside\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} house\textsc{.m}\textsc{=3pl} house\textsc{.m}\textsc{=3pl} in=there \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} there \\ 
\glt `He (Pir Shaliyar) returned to the region, returned to Hawraman. His\footnote{Lit. their.} house was there.'
\z 
 
\ea \label{ŽP.146}
\textit{pādšāū mīsrī kināčēš bo feleje bo} \\ 
\gll pādšā=ū mīsr-ī kināčē-ē=š ∅-b-o felej-e ∅-b-o \\ 
 king\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} daughter\textsc{.f}\textsc{-indf}\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} disabled\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The king of Egypt had a daughter [who] was disabled.'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.158}
\textit{lālošā, lālo heremū pīr šelyārīn} \\ 
\gll lālo=š=ā lālo hereme=ū pīr šelyār-ī=n \\ 
 maternal\_uncle\textsc{.m}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} maternal\_uncle\textsc{.m} wife\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `—He was her uncle. He was the uncle of Pir Shaliyar’s wife—'
\z 
 
\ea \label{ŽP.161}
\textit{metāwū serū textīwe wezūrew bilū ginū gelū kināčēw be peyū luāy řāreū degā be degā} \\ 
\gll me-tāw-ū ser=ū text-ī=we wez-ū=re=ū bi-l-ū ∅-gin-ū gel=ū kināčē=ū be pe-ī=ū luāy řā=re=ū degā be degā \\ 
 \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} on\textsc{\textsc{=ez.gen}} throne\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} climb\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{povb}}=and \textsc{sbjv-}go\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-\textsc{1sg}} with\textsc{\textsc{=ez.gen}} daughter\textsc{.dir}\textsc{.pl}=and by foot\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} go.inf road\textsc{.f}\textsc{=\textsc{post}}=and village\textsc{.obl}\textsc{.f} by village\textsc{.obl}\textsc{.f} \\ 
\glt `I cannot leave the throne [and] go [and] accompany the girl, [while] walking on foot on the roads [that go] village by village.'
\z 
 
\ea \label{ŽP.163}
\textit{eger to ī kināčēte berdeū ī pīyā her kes hen} \\ 
\gll eger to ī kināčē=t=e berd-e=ū ī pīyā her kes hen-∅ \\ 
 if \textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} take\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m} emph person\textsc{.m} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `If you take the girl, [no matter] who this man (the doctor) is,'
\z 
 
\ea \label{ŽP.164}
\textit{tāwāš ī duktire ī kināčēše weše kerdew} \\ 
\gll tāwā=š ī duktir=e ī kināčē=š=e weš-e kerd-e=ū \\ 
 can\textsc{.pst}\textsc{=3sg} dem.prox phy\textsc{=dem} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `If the doctor can cure the girl,'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.166}
\textit{lālo gino gelū kināčē māroš dey degā be degāw kēšwer be kēšwerū māroš silēne} \\ 
\gll lālo ∅-gin-o gel=ū kināčē m-ār-o=š dey degā be degā=ū kēšwer be kēšwer=ū m-ār-o=š silēne \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} with\textsc{\textsc{=ez.gen}} girl\textsc{.dir}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{disc.ptcl} village\textsc{.obl}\textsc{.f} by village\textsc{.f}=and country by country=and \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{pn} \\ 
\glt `The uncle set off on the road with the girl. He took her. He took her village by village and country by country until [they arrived] at Slen.'
\z 
 
\ea \label{ŽP.167}
\textit{silēne řeđē bā tewenē čā beynne bē beynū silēnē biɫberīne bē} \\ 
\gll silēne řeđ-ē bā tewenī-ē č=ā beyn=ne b-ē beyn=ū silēnē biɫber-ī=ne b-ē \\ 
 \textsc{pn} crossing\textsc{-dir}\textsc{.pl} \textsc{hort} stone\textsc{.f}\textsc{-indf} at=\textsc{dem.dist} between\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} between\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `They passed Slen. There was a stone in the road between Slen and Bilbar.'
\z 
 
\ea \label{ŽP.168}
\textit{leqemiš wāčēnē tewenū sirewāy} \\ 
\gll leqem=iš wāč-ēn-ē tewenī=ū sirewāy \\ 
 title\textsc{.m}\textsc{=3sg} say\textsc{.prs-pstc}-3pl stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment.inf \\ 
\glt `The stone is called ‘the healing stone.’'
\z 
 
\ea \label{ŽP.169}
\textit{mēnē čenū kināčekē serū tewenekēwe nīšāre} \\ 
\gll m-ē-nē čenū kināčē-ekē ser=ū tewenī-ekē=we ∅-nīš-ā=re \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} with\textsc{.ez.gen} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} on\textsc{\textsc{=ez.gen}} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} \textsc{ind-}sit\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \\ 
\glt `He (the uncle) went with the girl [and] sat on the stone.'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.185}
\textit{ēti nāmēšā nīyā tewenekē tewenū sirewāy”} \\ 
\gll ēti nāmē=šā nīyā tewenī-ekē tewenī=ū sirewāy \\ 
 \textsc{disc.ptcl} name\textsc{.f}\textsc{=3pl} put\textsc{.pst} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment.inf \\ 
\glt `That is why the stone is called Tawanu Sraway, ‘the healing stone.’'
\z 
 
\ea \label{ŽP.187}
\textit{milegā ā wextī mizyā werāwerū dūtifɫey} \\ 
\gll milegā ā wext-ī mi-zye-ā werāwer=ū dūtifɫey \\ 
 top\_of\_mountain \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `Then they came out in front of  Dutifla.'
\z 
 
\ea \label{ŽP.188}
\textit{pīr šelīyār je yānekew wēšāne} \\ 
\gll pīr šelīyār je yāne-eke=ū wē=šā=ne \\ 
 \textsc{pn} \textsc{pn} from house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar [was] in his\footnote{Lit. their.} house.'
\z 
 
\ea \label{ŽP.189}
\textit{hezretū sey misefāy čā miđyo xo kes ter mewīnoš} \\ 
\gll hezret=ū sey misefā-ī čā mi-đye-o xo kes ter me-wīn-o=š \\ 
 his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} there \textsc{ind-}look\textsc{.prs}\textsc{-3sg} dis.ptc person\textsc{.m} other \textsc{neg-}see\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \\ 
\glt `His Highness Say Mustafa watched [them] from above. No one else could see her.'
\z 
 
\ea \label{ŽP.196}
\textit{tenūrū māčāš pene tenūrū dēwā’} \\ 
\gll tenūr=ū m-āč-ā=š pene tenūr=ū dēw-ā \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-3pl} \\ 
\glt `They call [that place next to the tree] Tanuru Dewa ‘the demons’ oven.’'
\z 
 
\ea \label{ŽP.197}
\textit{mē henū tāylekey,} \\ 
\gll m-ē hen=ū tāyle-ekey \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} tree\_name\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `He (the demon) arrived next to the tree.'
\z 
 
\ea \label{ŽP.198}
\textit{fermāwo, weɫā čūn ānā bēheyānī kerūt penū ʕāɫemī} \\ 
\gll ∅-fermāw-o weɫā čūn āne=ā bēheyā=nī ∅-ker-ū=t pen=ū ʕāɫem-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed as \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} shameless\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{2sg}} trouble\textsc{.m}\textsc{\textsc{=ez.gen}} world\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) said, ‘Indeed, since you are so indecent, I will disgrace you.’'
\z 
 
\ea \label{ŽP.200}
\textit{tewejoēš wene kero pēse gāw til bore e pā tenūrū dēwāre til bowew ginorewe. ānen māčāš pene tenūrū dēwā} \\ 
\gll tewejo-ē=š wene ∅-ker-o pēse gāw til b-o=re e=pā tenūr=ū dēw-ā=re til b-o=we=ū gin-o=re=we āne=n m-āč-ā=š pene tenūr=ū dēw-ā \\ 
 caution\textsc{-indf}\textsc{=3sg} \textsc{post} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} as\_if bull\textsc{.m} rolling be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=that oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} rolling be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}}=and fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}}\textsc{=compl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He (Pir Shaliyar) paid attention to him. He (the demon) rolled down like a cow. It is there. It is called Tanuru Dewa.'
\z 
 
\ea \label{ŽP.201}
\textit{tenūrū dēwā ētir pēse gāw til bowe e čā werāwerū girdū mexeleqeɫāw hewrāmānīwe dēweke til bowe ginore soɫāwe} \\ 
\gll tenūr=ū dēw-ā ētir pēse gāw til ∅-b-o=we e=čā werāwer=ū gird=ū mexeleqeɫā=ū hewrāmān-ī=we dēw-eke til ∅-b-o=we ∅-gin-o=re soɫāwe \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{disc.ptcl} as\_if bull\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=there in\_front\_of\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt ` Then it was as if a cow had rolled down; the [demon] rolled down and fell [into a place called] Solawa before all the people’s eyes.'
\z 
 
\ea \label{ŽP.203}
\textit{kābrā lālow šābāre xātūnī mē dilū degēw} \\ 
\gll kābrā lālo=ū šābāre xātūn-ī m-ē dil=ū degē=ū \\ 
 man maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} lady\textsc{-obl}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} inside\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `The fellow, the uncle of Shahbara Khatun, went inside the village.'
\z 
 
\ea \label{ŽP.205}
\textit{muxteserū kelāmī, yāney yozāwew milā berāš xizmetū pīrīyū} \\ 
\gll muxteser=ū kelām-ī yāne-ī ∅-yoz-ā=we=ū mi-l-ā ∅-ber-ā=š xizmet=ū pīr-ī=ū \\ 
 summary\textsc{.m}\textsc{\textsc{=ez.gen}} speech\textsc{-obl}\textsc{.m} house\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}find\textsc{.prs}\textsc{-3pl}\textsc{=compl}\textsc{\textsc{=ez.gen}} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}=and \\ 
\glt ` To cut a long story short,\footnote{Lit. Gist of speech} they found the house. They (the people) went [and] took him to the presence of Pir [Shaliyar]'
\z 
 
\ea \label{ŽP.207}
\textit{mišyo ī kināčē nīkāh kerū pey toū be mezebū wētāū} \\ 
\gll mišyo ī kināčē nīkāh ∅-ker-ū pey to=ū be mezeb=ū wē=tā=ū \\ 
 aux dem.prox girl\textsc{.dir}\textsc{.f} marriage\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} to \textsc{2sg}=and by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.211}
\textit{mebārekū sāhēbīš bo} \\ 
\gll mebārek=ū sāhēb-ī=š ∅-b-o \\ 
 blessing\textsc{.m}\textsc{\textsc{=ez.gen}} owner\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `May she be happy with her owner (i.e., her father)'
\z 
 
\ea \label{ŽP.230}
\textit{yoīč mā yoīč māčo, kāke gīyān! minīč šimšū yerū heywānēt mido pene} \\ 
\gll yo=īč=mā yo=īč m-āč-o kāke gīyān min=īč šimšī=ū yeher=ū heywān-ē=t mi-de-o pene \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother.vob dear.\textsc{voc} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `One [said], One person said, ‘Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{ŽP.238}
\textit{yoīč māčo, minīč šimšū yeherit miđo} \\ 
\gll yo=īč m-āč-o min=īč šimšī=ū yeher=it mi-đe-o \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `One said, ‘I will give you spleen and liver.’'
\z 
 
\ea \label{ŽP.251}
\textit{kīyānāš serūpīrīwe qesābē bo māčāš tāyfew pādšāy} \\ 
\gll ∅-kīyān-ā=š serūpīrī=we qesāb-ē ∅-b-o m-āč-ā=š tāyfe=ū pādšā-ī \\ 
 \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{pn}\textsc{=\textsc{post}} butcher\textsc{.m}\textsc{-indf} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `They sent [one] to Saru Piri; there was a butcher who was from the king’s tribe.'
\z 
 
\ea \label{ŽP.252}
\textit{tāyfew pādšāyāš wāčēnē} \\ 
\gll tāyfe=ū pādšā-yā=š wāč-ēn-ē \\ 
 tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=3sg} say\textsc{.prs-pstc-3pl} \\ 
\glt `They (his tribe) are called the kings’ tribe.'
\z 
 
\ea \label{ŽP.256}
\textit{merāsēmekew pīr šelīyārī bergozār kerāū} \\ 
\gll merāsēm-eke=ū pīr šelīyār-ī bergozār ∅-ker-ā=ū \\ 
 ceremony\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} holding \textsc{ind-}do\textsc{.prs}\textsc{-3pl}=and \\ 
\glt `They held Pir Shaliyar’s ceremony.'
\z 
 
\ea \label{ŽP.257}
\textit{ženī be mezebū hewrāmānī māre biřā pey pīr šelīyārīū} \\ 
\gll ženī be mezeb=ū hewrāmān-ī māre ∅-biř-ā pey pīr šelīyār-ī=ū \\ 
 woman\textsc{.f} by religion\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} marraige \textsc{ind-}cut\textsc{.prs}\textsc{-3pl} to \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \\ 
\glt `They married the woman to Pir Shaliyar] according to] the religious customs in Hawraman.'
\z 
 
\ea \label{ŽP.258}
\textit{ā wextīyekey lālo pīr šelīyār hīn lālo heremū pīr šelīyārī milowe xeber berowe pey pādšāy} \\ 
\gll ā wext-ī-ekey lālo pīr šelīyār hīn lālo hereme=ū pīr šelīyār-ī mi-l-o=we xeber ∅-ber-o=we pey pādšā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} maternal\_uncle\textsc{.m} \textsc{pn} \textsc{pn} \textsc{fill} maternal\_uncle\textsc{.m} wife\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=compl} to king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, Pir Shaliyar’s uncle, eh, Pir Shaliyar’s wife’s uncle, went back [to Egypt] and passed the news to the king.'
\z 
 
\ea \label{ŽP.259}
\textit{xeber berowe pey pādšāy māčo, hāɫū jereyānē e čīnen} \\ 
\gll xeber ∅-ber-o=we pey pādšā-ī m-āč-o hāɫ=ū jereyān-ē e=čīne=n \\ 
 news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=compl} to king\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} state\textsc{.m}\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He passed the news to the king [and] said, ‘The story is like this.’'
\z 
 
\ea \label{ŽP.263}
\textit{ētir to lue meylū wētā’} \\ 
\gll ētir to ∅-lu-e meyl=ū wē=t=ā \\ 
 \textsc{disc.ptcl} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} willingness\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then it’s up to you [to stay in Hawraman]'
\z 
 
\ea \label{BP.1}
\textit{zemānū bābāw pīr šelīyārī, heke, nāmēš bābā xwāđā bīyen} \\ 
\gll zemān=ū bābā=ū pīr šelīyār-ī heke nāmē=š bābā xwāđā bīye=n \\ 
 time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} well name.f\textsc{=3sg} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the time of Pir Shaliyar’s grandfather —his name was Baba Khwada.'
\z 
 
\ea \label{BP.2}
\textit{bābāw pīr šelīyārī, nāmēš bābā xwāđā bīyen} \\ 
\gll bābā=ū pīr šelīyār-ī nāmē=š bābā xwāđā bīye=n \\ 
 grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} name.f\textsc{=3sg} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s grandfather, his name was Baba Khwada\footnote{Lit. Grandfather Khwada}. '
\z 
 
\ea \label{BP.3}
\textit{yānē yānē meseleyš serū īneyn yānē bābā xwāy dā, kuřēš peyā bī} \\ 
\gll yānē yānē mesele-ī=š ser=ū īne-ī=n yānē bābā xwā-ī dā kuř-ē=š peyā bī-∅ \\ 
 that\_means that\_means issue\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} that\_means grandfather\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} give\textsc{.pst} son\textsc{.m}\textsc{-indf}\textsc{=3sg} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `The point with his name is that God granted the grandfather [a son]. A boy  was born to him .\footnote{Lit. A son of him appeared.}'
\z 
 
\ea \label{BP.7}
\textit{tātew pīr šelīyārīn} \\ 
\gll tāte=ū pīr šelīyār-ī=n \\ 
 father\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is Pir Shaliyar’s father.'
\z 
 
\ea \label{BP.8}
\textit{tātew pīr šelīyārīn} \\ 
\gll tāte=ū pīr šelīyār-ī=n \\ 
 father\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is Pir Shaliyar’s father.'
\z 
 
\ea \label{BP.9}
\textit{ā zemān bābā xwāđā bīyen, ā zemān ī mexlūqū hewrāmānātū ēme reʕīsšā je ʕērāqew čā kurdewārīū ʕērāqīne bīyen} \\ 
\gll ā zemān bābā xwāđā bīye=n ā zemān ī mexlūq=ū hewrāmān-āt=ū ēme reʕīs=šā je ʕērāq=ew čā kurdewārī=ū ʕērāq-ī=ne bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} time\textsc{.m} \textsc{dem.prox} people\textsc{=ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{=ez.gen} \textsc{1pl} leader\textsc{.m}\textsc{=3pl} from \textsc{pn}\textsc{=\textsc{post}} there kurdish\_region\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Babab Khwada, the people of Hawraman were under Iraq’s rule, under the control of Kurdish regions in Iraq.'
\z 
 
\ea \label{BP.11}
\textit{zemānū bābāw pīr šelīyārī bābā xwāđāy āā je ʕērāqne ēme hukmū hewrāmānīšā kerden, biřew bārešā sānān} \\ 
\gll zemān=ū bābā=ū pīr šelīyār-ī bābā xwāđā-ī āā je ʕērāq=ne ēme hukm=ū hewrāmān-ī=šā kerde=n biře=ū bāre=šā sānā=n \\ 
 time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{intj} from \textsc{pn}\textsc{=\textsc{post}} \textsc{1pl} rule\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Pir Shaliyar’s grandfather, we were ruled by Iraq, [and] they would collect tax.'
\z 
 
\ea \label{BP.12}
\textit{biřew bārešā sānān, āmēnē pēse čāweɫī begzāyāw wēmā} \\ 
\gll biře=ū bāre=šā sānā=n āmē=nē pēse č=ā eweɫ-ī begzā-yā=ū wē=mā \\ 
 portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} nobleman\textsc{-obl}\textsc{.pl}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1pl}} \\ 
\glt `They would collect tax. They would come (here) [and] like our noblemen..'
\z 
 
\ea \label{BP.17}
\textit{ā pāse hetā hečkesē ženīš dē šū, bešepīyākew ēme begzāyē ēme} \\ 
\gll ā pāse hetā hečkes-ē ženī=š de-ē šū bešepīyā-eke=ū ēme begzāye-ē ēme \\ 
 \textsc{dem.dist} like even whoever\textsc{-indf} woman.f\textsc{=3sg} give.prs\textsc{-pstc}\textsc{.3sg} husband\textsc{.m} quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{1pl} nobleman\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} \\ 
\glt `[Things were] like this. Even when someone wanted to get his daughter married, he had to pay a quota  to our noblemen.'
\z 
 
\ea \label{BP.18}
\textit{eger ženīmā dē be bešepīyākew yo terī čīwēš zereriš sānē čene wāčēnēš sūrāne} \\ 
\gll eger ženī=mā de-ē be bešepīyā-eke=ū yo ter-ī čīw-ē=š zerer=iš sān-ē čene wāč-ēn-ē=š sūrāne \\ 
 if woman.f\textsc{=\textsc{1pl}} give.prs\textsc{-cond} to quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} thing\textsc{.m}\textsc{-indf}\textsc{=3sg} harm\textsc{.m}\textsc{=3sg} buy.prs\textsc{-pstc}\textsc{.3sg} from say.prs.\textsc{sbjv-}pstc\textsc{-3pl}\textsc{=3sg} \textsc{pn} \\ 
\glt `If one\footnote{Lit. we.} had married women from the quota of another [lord], then they [the lord of the bride's village] would charge him [the bride’s father]. This money is called Surana.'
\z 
 
\ea \label{BP.20}
\textit{jā ī dāstānē hinū bābāw pīr šelīyārīne} \\ 
\gll jā ī dāstāne-ē hin=ū bābā=ū pīr šelīyār-ī=ne \\ 
 afterwards \textsc{dem.prox} story.f\textsc{-obl}.f ez.pron\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `Anyhow, this story is about the grandfather of Pir Shaliyar.'
\z 
 
\ea \label{BP.21}
\textit{bābāw pīr šelīyārī hukmū bābāw pīr šelīyārī ā zemāne hewrāmān bīyen je ʕērāqo āmēnē hukimšā kerden} \\ 
\gll bābā=ū pīr šelīyār-ī hukm=ū bābā=ū pīr šelīyār-ī ā zemān=e hewrāmān bīye=n je ʕērāq=o āmē=nē hukim=šā kerde=n \\ 
 grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} rule\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} rule\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s grandfather, at the time of his religious rule, Hawraman was under the rule of Iraq.'
\z 
 
\ea \label{BP.24}
\textit{e čē biřew bārešā sānān je hewrāmānī pēse begzāyekāw ī beynū ēme} \\ 
\gll e=čē biře=ū bāre=šā sānā=n je hewrāmān-ī pēse begzāye-ekā=ū ī beyn=ū ēme \\ 
 in=here portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{-obl}\textsc{.m} like nobleman\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=ez.gen} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `[who] would collect taxes from Hawraman, [just] like our noblemen [in recent] times.'
\z 
 
\ea \label{BP.27}
\textit{jā luēnē berdenšāwe, luēnē berdenšāwe pey āġekew ʕērāqīšā} \\ 
\gll jā luē=nē berde=n=šā=we luē=nē berde=n=šā=we pey āġe-eke=ū ʕērāq-ī=šā \\ 
 afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} for agha\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3pl} \\ 
\glt `Then, they would go back and take the taxes to their aghas in Iraq.'
\z 
 
\ea \label{BP.28}
\textit{āġekew ʕērāqī berdenšāwe peyšāw} \\ 
\gll āġe-eke=ū ʕērāq-ī berde=n=šā=we pey=šā=ū \\ 
 agha\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} to\textsc{=3pl}=and \\ 
\glt `The aghas in Iraq, they took back [the taxes] to them.'
\z 
 
\ea \label{BP.31}
\textit{be enāzew wēš biřew bāre misānā, zemānū bābāw pīr šelīyārī, bābā xwāđāy} \\ 
\gll be enāze=ū wē=š biře=ū bāre mi-sān-ā zemān=ū bābā=ū pīr šelīyār-ī bābā xwāđā-ī \\ 
 to proportion\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg} portion\textsc{.m}=and profit\textsc{.m} \textsc{ind-}buy.prs\textsc{-3pl} time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `During the time of Pir Shaliyar’s grandfather, i.e. Baba Khwada, they (the tax collectors) collected tax in proportion to one’s wealth. '
\z 
 
\ea \label{BP.36}
\textit{je terefū ženū menū čīwe genekārī kerā nelāwe} \\ 
\gll je teref=ū ženī=ū men=ū čīw=e genekārī ∅-ker-ā ne-l-ā=we \\ 
 from side\textsc{=ez.gen} woman.f=and \textsc{rdp}=and thing\textsc{.m}\textsc{=dem} debauchery\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They wished to engage in debauchery with women and so on. They didn’t go back.'
\z 
 
\ea \label{BP.39}
\textit{yānew yoy nānšā wārden} \\ 
\gll yāne=ū yo-ī nān=šā wārde=n \\ 
 house\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m}\textsc{=3pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They would eat at the house of one [fellow at a time]'
\z 
 
\ea \label{BP.40}
\textit{her wēregāne pēse čāweɫ sempāšē bēnē, yoīšā dēnē duīšā dēnē yānew yoy ī meʕmūrīye ke ʕērāqo āmēnē, ā zemānū pīr šelīyārī bexšnāyšā bexšnēnē} \\ 
\gll her wēregā=ne pēse čāweɫ sempāš-ē b-ēn-ē yo-ī=šā de-ēn-ē duē-ī=šā de-ēn-ē yāne=ū yo-ī ī meʕmūr-ī=e ke ʕērāq=o āmē=nē ā zemān=ū pīr šelīyār-ī bexšnāy=šā bexšn-ēn-ē \\ 
 just evening\textsc{=\textsc{post}} like in\_the\_past poison\_sprayer\textsc{-dir}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} two\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} house\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} officer\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=dem} rel \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `As in the old days when there were people whose job was spraying pesticide on trees, [and] each would be hosted at the house of one [host], the officers who had come from Iraq would be fed in the time of Pir Shaliyar. '
\z 
 
\ea \label{BP.41}
\textit{yānew šime neferēšā kīyāsenū} \\ 
\gll yāne=ū šime nefer-ē=šā kīyāse=n=ū \\ 
 house\textsc{.m}\textsc{=ez.gen} \textsc{2pl} person\textsc{.m}\textsc{-indf}\textsc{=3pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `[For instance] One [tax collector] would be sent to your house;'
\z 
 
\ea \label{BP.42}
\textit{yānew ēme neferēšā kīyāsenū} \\ 
\gll yāne=ū ēme nefer-ē=šā kīyāse=n=ū \\ 
 house\textsc{.m}\textsc{=ez.gen} \textsc{1pl} person\textsc{.m}\textsc{-indf}\textsc{=3pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `One [tax collector] would be sent to our house;'
\z 
 
\ea \label{BP.43}
\textit{yānew yo terī, her šewē her wēregāne ī mēmānē mēmānū yoy bīyēne} \\ 
\gll yāne=ū yo ter-ī her šewe-ē her wēregā=ne ī mēmān-ē mēmān=ū yo-ī bīyē=ne \\ 
 house\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} just night.f\textsc{-obl}.f just evening\textsc{=\textsc{post}} \textsc{dem.prox} guest\textsc{.m}\textsc{-dir}\textsc{.pl} guest\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `[likewise, one tax collector] would be the guest in the house of one [fellow from Hawraman]  each evening, each night.'
\z 
 
\ea \label{BP.44}
\textit{mēmānū yoy bīyēne} \\ 
\gll mēmān=ū yo-ī bīyē=ne \\ 
 guest\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `They were each a guest of one person.'
\z 
 
\ea \label{BP.45}
\textit{milā xeɫk pāne milo miđyo īne ā xerīkū genekārīnē ī kābrāy je āmēnē, ī meʕmūrē xerīkū genekārīnēw} \\ 
\gll mi-l-ā xeɫk p=ā=ne mi-l-o mi-đye-o īne=ā xerīk=ū genekārī=nē ī kābrā-ī je āmē=nē ī meʕmūr-ē xerīk=ū genekārī=nē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3pl} people\textsc{.m} at=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}look.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} man\textsc{-obl}\textsc{.m} from come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `People realised that these men were busy engaging in debauchery; these officers were busy engaing in debauchery.'
\z 
 
\ea \label{BP.46}
\textit{xeber yāwo be bābā xwāđāw ħemey ġeybī} \\ 
\gll xeber ∅-yāw-o be bābā xwāđā=ū ħeme=ī ġeybī \\ 
 news\textsc{.dir}\textsc{.m} \textsc{ind-}arrive.prs\textsc{-3sg} to grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible \\ 
\glt `The news reached Baba Khwada and Hama the invisible.'
\z 
 
\ea \label{BP.49}
\textit{genekārī kerā čenū ženū menī} \\ 
\gll genekārī ∅-ker-ā čenū ženī=ū men-ī \\ 
 debauchery\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} with\textsc{.ez.gen} woman.f=and \textsc{rdp}\textsc{-obl}\textsc{.m} \\ 
\glt `They (the tax collectors) sexually assaulted women and so on. '
\z 
 
\ea \label{BP.51}
\textit{ħemey ġeybīw bābāw pīr šelīyārī, bābā xwāđā māčā be xeɫkekey,} \\ 
\gll ħeme=ī ġeybī=ū bābā=ū pīr šelīyār-ī bābā xwāđā m-āč-ā be xeɫk-ekey \\ 
 \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3pl} to people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Hama Ghaybi and Pir Shaliyar’s grandfather, i.e., Baba Khwada, said to the people, '
\z 
 
\ea \label{BP.54}
\textit{seʕbē girđtā beydē mizgī, ā mizgīyew hewrāmānī,} \\ 
\gll seʕbe-ē girđ=tā b-e-ydē mizgī ā mizgī=e=ū hewrāmān-ī \\ 
 morning.f\textsc{-obl}.f all=\textsc{2pl} \textsc{sbjv-}come.prs-\textsc{2pl} mosque\textsc{.m} \textsc{dem.dist} mosque\textsc{.m}\textsc{=dem}\textsc{=ez.gen}\textsc{-obl}\textsc{.m} \\ 
\glt `Tomorrow, all of you attend the mosque, that mosque of Hawraman.'
\z 
 
\ea \label{BP.56}
\textit{jā girđmā bilmē pey ʕērāqī pey lāw ī āġeyšā bilmērewo,} \\ 
\gll jā girđ=mā bi-l-mē pey ʕērāq-ī pey lā=ū ī āġe-ī=šā bi-l-mē=re=o \\ 
 afterwards all\textsc{=\textsc{1pl}} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} for at\_the\_place\_of\textsc{=ez.gen} \textsc{dem.prox} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}\textsc{=compl} \\ 
\glt `Then, we all go to Iraq, to their agha. We go down there.'
\z 
 
\ea \label{BP.61}
\textit{xerīkū genekārī bēnē} \\ 
\gll xerīk=ū genekārī b-ēn-ē \\ 
 busy\textsc{=ez.gen} debauchery\textsc{.m} be.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `They were busy engaging in debauchery.'
\z 
 
\ea \label{BP.63}
\textit{ēmeyč muxteserū kelāmī ā īney wāčmē,} \\ 
\gll ēme=yč muxteser=ū kelām-ī ā īney wāč-mē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} \textsc{dem.dist} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} say.prs.sbjv\textsc{-\textsc{1pl}} \\ 
\glt `As for us, to cut a long story short, we say this.'
\z 
 
\ea \label{BP.65}
\textit{qewɫ kerā yo, wēregā bābāw pīr šelīyārī xeberšā pene miđo} \\ 
\gll qewɫ ∅-ker-ā yo wēregā bābā=ū pīr šelīyār-ī xeber=šā pene mi-đe-o \\ 
 arrangement\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} one\textsc{.m} evening grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} news\textsc{.m}\textsc{=3pl} to \textsc{ind-}give.prs\textsc{-3sg} \\ 
\glt `They all agreed [about the plan]. In the evening Pir Shaliyar’s grandfather informed them'
\z 
 
\ea \label{BP.66}
\textit{īney māčā beynū wēšāne} \\ 
\gll īney m-āč-ā beyn=ū wē=šā=ne \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-3pl} among\textsc{=ez.gen} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `They said this to each other,'
\z 
 
\ea \label{BP.67}
\textit{gird kes mēmānū wēš sere biřo šewē} \\ 
\gll gird kes mēmān=ū wē=š sere ∅-biř-o šewe-ē \\ 
 all person\textsc{.m} guest\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg} head\textsc{.m} \textsc{sbjv-}cut.prs\textsc{-3sg} night.f\textsc{-obl}.f \\ 
\glt `‘Each person should decapitate his guest tonight.'
\z 
 
\ea \label{BP.70}
\textit{be xwā girđ kes wēregā milā yānew hečkesī bā čwār bā penj bā yewāšē girđkes mēmānū wēš sere biřo, yo be kārdē, yo be tirēštew} \\ 
\gll be xwā girđ kes wēregā mi-l-ā yāne=ū hečkes-ī ∅-b-ā čwār ∅-b-ā penj ∅-b-ā yewāšē girđkes mēmān=ū wē=š sere ∅- bi-ř-o yo be kārdī-ē yo be tirēšte=ū \\ 
 to God\textsc{.m} all person\textsc{.m} evening \textsc{ind-}go.prs\textsc{-3pl} house\textsc{.m}\textsc{=ez.gen} whoever\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3pl} four \textsc{ind-}be.prs\textsc{-3pl} five \textsc{ind-}be.prs\textsc{-3pl} then everyone guest\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg} head\textsc{.m} \textsc{ind-}\textsc{sbjv-}cut.prs\textsc{-3sg} one\textsc{.m} to knife.f\textsc{-obl}.f one\textsc{.m} to ax\textsc{.m}=and \\ 
\glt `So, each one went back [home] in the evening. No matter whose house they (the officers) were guests at; [no matter] if they were four or five [guests],  they  (the hosts) decapitated them during the night, one [did so] using knives, one [did so] using axes.'
\z 
 
\ea \label{BP.71}
\textit{neferē yāneš pī ser degāwe bo, tirēštew kārdū čīwiš mebo} \\ 
\gll nefer-ē yāne=š p=ī ser degā=we ∅-b-o tirēšte=ū kārdī=ū čīw=iš me-b-o \\ 
 person\textsc{.m}\textsc{-indf} house\textsc{.m}\textsc{=3sg} at=\textsc{dem.prox} on village.f\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3sg} ax\textsc{.m}=and knife.f\textsc{=ez.gen} thing\textsc{.m}\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg} \\ 
\glt `The house of one of the locals was at the top of the village. He had neither axe, nor knife, nor anything else. '
\z 
 
\ea \label{BP.90}
\textit{māčmē ī meʕmūrū šime āmē lāw ēme} \\ 
\gll m-āč-mē ī meʕmūr=ū šime āmā-ē lā=ū ēme \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} come\textsc{.pst}\textsc{-3pl} at\_the\_place\_of\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `We will say [to the agha], “your officers came to our region".'
\z 
 
\ea \label{BP.91}
\textit{biřew bāremā dēw} \\ 
\gll biře=ū bāre=mā dā-ē=ū \\ 
 portion\textsc{.m}\textsc{=ez.gen} profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \\ 
\glt `We paid them tax.'
\z 
 
\ea \label{BP.93}
\textit{xerīkū genekārī bēnē, nāmēwe} \\ 
\gll xerīk=ū genekārī b-ēn-ē ne-āmā-ē=we \\ 
 busy\textsc{=ez.gen} debauchery\textsc{.m} be.prs\textsc{-pstc}\textsc{-3pl} \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \\ 
\glt `[However] They fell into debauchery; They wouldn’t go back [to Iraq]'
\z 
 
\ea \label{BP.94}
\textit{nāmēwew xerīkū genekārīw kārī nāšerʕīšā kerdenū} \\ 
\gll ne-āmā-ē=we=ū xerīk=ū genekārī=ū kār=ī nāšerʕī=šā kerde=n=ū \\ 
 \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-indf}\textsc{=compl}=and busy\textsc{=ez.gen} debauchery\textsc{.m}=and task\textsc{.m}=ez.\textsc{attr} unlawful\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They didn’t go back [to Iraq]. They commited unlawful acts.'
\z 
 
\ea \label{BP.95}
\textit{ēmeyč muxteserū kelāmī, xo ēme diruē meyeymē} \\ 
\gll ēme=yč muxteser=ū kelām-ī xo ēme diruē me-de-ymē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} dis.ptc \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `As for us, to cut a long story short, we are not lying.'
\z 
 
\ea \label{BP.96}
\textit{ī meʕmūrū šimemā sere biřyēnē} \\ 
\gll ī meʕmūr=ū šime=mā sere biřyē=nē \\ 
 \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl}\textsc{=\textsc{1pl}} head\textsc{.m} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We have decapitated your officers.'
\z 
 
\ea \label{BP.97}
\textit{seremā biřyē, īne sereū ēmew īne šimšērū šime} \\ 
\gll sere=mā biřī-ē īne sere=ū ēme=ū īne šimšēr=ū šime \\ 
 head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} \textsc{2pl} \\ 
\glt `We have decapitated them. Here are their heads and the agha’s swords.\footnote{Lit. These are our heads and these are the agha’s swords.}'
\z 
 
\ea \label{BP.100}
\textit{āwetū xuɫkī īse ā wextīč xuɫk her tersezāɫ bo,} \\ 
\gll āđet=ū xuɫk-ī īse ā wext-ī=č xuɫk her tersezāɫ ∅-b-o \\ 
 habit\textsc{=ez.gen} people\textsc{.m}\textsc{-obl}\textsc{.m} now \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} people\textsc{.m} just coward \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `Just like people nowdays, in earlier times people were intimidated [by kings]. '
\z 
 
\ea \label{BP.102}
\textit{pādšā ezyetmā keronew misībetmā biđe ver} \\ 
\gll pādšā ezyet=mā ∅-ker-one=ū misībet=mā bi-đe-o ver \\ 
 king\textsc{.m} irritation\textsc{.m}\textsc{=\textsc{1pl}} \textsc{sbjv-}do.prs\textsc{-3sg}\textsc{=ez.gen} calamity\textsc{.m}\textsc{=\textsc{1pl}} \textsc{sbjv-}give.prs\textsc{-3sg} front \\ 
\glt `The king is going to upset us and give us a hard time.'
\z 
 
\ea \label{BP.105}
\textit{bābā xwāđāw ħemey ġeybīū ħemey wirdīkɫe her ginēnē weɫē īnē weɫēne ejošā girđ īnē šūnīšāwe} \\ 
\gll bābā xwāđā=ū ħeme=ī ġeybī=ū ħeme=ī wirdīkɫe her gin-ēn-ē weɫē īnē weɫē=ne ejo=šā girđ īnē šūnī=šā=we \\ 
 grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible=and \textsc{pn}=ez.\textsc{attr} small just fall.prs\textsc{-pstc}\textsc{-3pl} front \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} front\textsc{=\textsc{post}} guess.prs\textsc{=3pl} all \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} track.f\textsc{=3pl}\textsc{=compl} \\ 
\glt `Baba Khwada, Hama the invisible, and little Hama were in front, assuming that everyone was following them.'
\z 
 
\ea \label{BP.106}
\textit{wextē milā serū meřekēw meře beřī mēyāwe dimāwe} \\ 
\gll wext-ē mi-l-ā ser=ū meře-ekē=w meře beř-ī m-ē-ā=we dimā=we \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3pl} on\textsc{=ez.gen} cave.f\textsc{-def}\textsc{.obl}.f\textsc{=ez.gen} cave.f \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{-3pl}\textsc{=compl} back\textsc{=\textsc{post}} \\ 
\glt `When they reached the cave of Marabar, they looked behind [them].'
\z 
 
\ea \label{BP.109}
\textit{fermāwo, īne konē xeɫkeke čenū ēme bē?} \\ 
\gll ∅-fermāw-o īne ko=nē xeɫk-eke čenū ēme b-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} people\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} with\textsc{.ez.gen} \textsc{1pl} be.prs\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Baba Khwada) said, ‘Where are the people who were with us?'
\z 
 
\ea \label{BP.120}
\textit{jereyāniš kermē pey pādšāw ʕērāqī} \\ 
\gll jereyān=iš ∅-ker-mē pey pādšā=w ʕērāq-ī \\ 
 story\textsc{.m}\textsc{=3sg} \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}} for king\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We will tell the king the story.'
\z 
 
\ea \label{BP.121}
\textit{māčmē ēme serū īney ī pīyēmā kuštēnē”’} \\ 
\gll m-āč-mē ēme ser=ū īney ī pīyē=mā kuštē=nē \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{1pl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox} man\textsc{.dir}\textsc{.pl}\textsc{=\textsc{1pl}} kill\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We say, ‘We have killed these men because of this.’'
\z 
 
\ea \label{BP.122}
\textit{ā wextī reʕīsū milāre ʕērāq lāw āġekeyšā māčo seremā biřyēnē bābāw pīr šelīyārī bo, ā wextīyekey ħemey ġeybīū yoyč bo nāmē ħemey wirdīkɫe,} \\ 
\gll ā wext-ī reʕīs=ū mi-l-ā=re ʕērāq lā=ū āġe-ekey=šā m-āč-o sere=mā biřyē=nē bābā=ū pīr šelīyār-ī ∅-b-o ā wext-ī-ekey ħeme=ī ġeybī=ū yo=yč ∅-b-o nāmē ħeme=ī wirdīkɫe \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} leader\textsc{.m}\textsc{=ez.gen} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{pn} at\_the\_place\_of\textsc{=ez.gen} agha\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}say.prs\textsc{-3sg} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}=ez.\textsc{attr} invisible=and one\textsc{.m}\textsc{=add} \textsc{ind-}be.prs\textsc{-3sg} name.f \textsc{pn}=ez.\textsc{attr} small \\ 
\glt `The head of those who went to Iraq to the agha, saying, ‘We have decapitated [your men], was Pir Shaliyar’s grandfather, Baba Khwada. In addition [there were] Hama the invisible and Little Hama. '
\z 
 
\ea \label{BP.124}
\textit{ā wextī milārewe māčā, wiɫā ī meʕmūrū šime kīyāsēbēnētā pey īne, īnemā āne heke šime gerekbē dēmā penew} \\ 
\gll ā wext-ī mi-l-ā=re=we m-āč-ā wiɫā ī meʕmūr=ū šime kīyāsē b-ēn-ē=tā pey īne īne=mā āne heke šime gerek b-ē dā-ē=mā pene=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.\textsc{prs}\textsc{-3pl}\textsc{=\textsc{povb}}\textsc{=compl} \textsc{ind-}say.prs\textsc{-3pl} by\_God \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl}=\textsc{2pl} for \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1pl}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} if \textsc{2pl} necessary be.prs\textsc{-pstc}\textsc{.3sg} give\textsc{.pst}\textsc{-3pl}\textsc{=\textsc{1pl}} to=and \\ 
\glt `Then, they went [to the agha and] said, ‘Indeed, the officers whom you had sent to us, whatever [ taxation] you had asked for, we gave them,'
\z 
 
\ea \label{BP.125}
\textit{āđē nāmēw xerīkū genekārī bēnēw} \\ 
\gll āđē ne-āmā-ē=ū xerīk=ū genekārī b-ēn-ē=ū \\ 
 3pl\textsc{.dir} \textsc{neg-}come\textsc{.pst}\textsc{-3pl}=and busy\textsc{=ez.gen} debauchery\textsc{.m} be.prs\textsc{-pstc}\textsc{-3pl}=and \\ 
\glt `[however] they didn’t come back [to you]. They engaged in debauchery.'
\z 
 
\ea \label{BP.126}
\textit{nemēbē tēkeɫī kerēnē čenū ženū menīū} \\ 
\gll nemē b-ē tēkeɫī ker-ēn-ē čenū ženī=ū men-ī=ū \\ 
 grab\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be.prs\textsc{-pstc}\textsc{.3sg} mixture\textsc{.m} do.prs\textsc{-pstc}\textsc{-3pl} with\textsc{.ez.gen} woman.f=and \textsc{rdp}\textsc{-obl}\textsc{.m}=and \\ 
\glt `They fell into adultery with women and so on.'
\z 
 
\ea \label{BP.127}
\textit{ēmeyč muxteserū kelāmī seremā biřyēnē} \\ 
\gll ēme=yč muxteser=ū kelām-ī sere=mā biřyē=nē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `So, to cut a long story short, we decapitated them.'
\z 
 
\ea \label{BP.128}
\textit{īne serew ēmew īne šimšērū āġey} \\ 
\gll īne sere=ū ēme=ū īne šimšēr=ū āġe-ī \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Here are their heads and the agha’s swords.'
\z 
 
\ea \label{BP.131}
\textit{bābā xuāđāw ħemew ġeybī, āđē, āđīšā kerā zīnān’} \\ 
\gll bābā xuāđā=ū ħeme=ū ġeybī āđē āđīšā ∅-ker-ā zīnān \\ 
 grandfather\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez.gen} invisible 3pl\textsc{.dir} 3pl\textsc{.obl} \textsc{ind-}do.prs\textsc{-3pl} prison\textsc{.m} \\ 
\glt `Baba Khwada and Hama Ghaybi, they (agha’s men) put them  in prison.'
\z 
 
\ea \label{BP.132}
\textit{ħemey wirdīkɫeyč pēse īse māčā, heɫāy neyāwān bilūqū sinʕēš} \\ 
\gll ħeme=ī wirdīkɫe=yč pēse īse m-āč-ā heɫāy ne-yāwā=n bilūq=ū sinʕe-ē=š \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} like now \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} maturity\textsc{.m}\textsc{=ez.gen} age.f\textsc{-obl}.f\textsc{=3sg} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) said, ‘he hasn’t hit puberty yet.'
\z 
 
\ea \label{BP.139}
\textit{berā kerāšāne zīnāneke, wextē miđyā mezā nū če noʕē bīyen ħejījīyāyčšā girten xuɫkū hejījī} \\ 
\gll ∅-ber-ā ∅-ker-ā=šā=ne zīnān-eke wext-ē mi-đye-ā me-zān-ū če noʕ-ē bīye=n ħejījī-yā=yčšā girte=n xuɫk=ū hejīj-ī \\ 
 \textsc{ind-}take.prs\textsc{-3pl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=3pl}\textsc{=\textsc{povb}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-\textsc{1sg}} what kind/way\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} people\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They took [them] and put them in the jail, they realised that —they didn’t know for what reason— [but] a person from Hajij had also been arrested.'
\z 
 
\ea \label{BP.144}
\textit{milo pāne swāɫū nānī kero,} \\ 
\gll mi-l-o p=ā=ne swāɫ=ū nān-ī ∅-ker-o \\ 
 \textsc{ind-}go.prs\textsc{-3sg} at=\textsc{dem.dist}\textsc{=\textsc{post}} begging\textsc{=ez.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}do.prs\textsc{-3sg} \\ 
\glt `He would go there and beg for some food.'
\z 
 
\ea \label{BP.149}
\textit{ētir māwāš fire bo yā kem bo her činne māngē bo her činne wextē bo wextē milo serū řūčinekey nānšā pey wizorew wextē miđyo hešpišā dēnēne} \\ 
\gll ētir māwā=š fire ∅-b-o yā kem ∅-b-o her činne mānge-ē ∅-b-o her činne wext-ē ∅-b-o wext-ē mi-l-o ser=ū řūčine-ekey nān=šā pey ∅-wiz-o=re=ū wext-ē mi-đye-o hešpišī-ā dē=nē=ne \\ 
 \textsc{disc.ptcl} period\textsc{.m}\textsc{=3sg} a\_lot \textsc{ind-}be.prs\textsc{-3sg} either little \textsc{ind-}be.prs\textsc{-3sg} just some month.f\textsc{-dir}\textsc{.pl} \textsc{ind-}be.prs\textsc{-3sg} just some time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}be.prs\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} on\textsc{=ez.gen} chimney\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} bread\textsc{.m}\textsc{=3pl} for \textsc{ind-}throw.prs\textsc{-3sg}\textsc{=\textsc{povb}}=and time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} \\ 
\glt `Their imprisonment lasted either a long time or short time. It took them months. Once, when he (Hama Wirdikla) went to the chimney to throw food for them, he realised that they had gotten lice.'
\z 
 
\ea \label{BP.153}
\textit{serū řūčinekeywe des kero girewāy} \\ 
\gll ser=ū řūčine-ekey=we des ∅-ker-o girewāy \\ 
 on\textsc{=ez.gen} chimney\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} hand\textsc{.m} \textsc{ind-}do.prs\textsc{-3sg} cry\textsc{.inf} \\ 
\glt `He sarted to cry on the chimney.'
\z 
 
\ea \label{BP.161}
\textit{bizāne yānew kē pī činnē hāy wexte nānē čīwēšā dānī ārdenit pey ēme, ārdenit pey ēmew} \\ 
\gll bi-zān-e yāne=ū kē p=ī činnē hāy wext=e nān-ē čīw-ē=šā dā=nī ārde=n=it pey ēme ārde=n=it pey ēme=ū \\ 
 \textsc{sbjv-}know.prs-\textsc{2sg} house\textsc{.m}\textsc{=ez.gen} who at=\textsc{dem.prox} some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{=dem} meal\textsc{.m}\textsc{-indf} thing\textsc{.m}\textsc{-indf}\textsc{=3pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl}=and \\ 
\glt `see which household has given you food, which you brought to us, during this period!'
\z 
 
\ea \label{BP.165}
\textit{wāče her kesī nān dānī, wāče, ēšew be xātirū xwāy toū īđū ī kiɫfetū yāneyte dūr ginewe čī šārī’} \\ 
\gll wāč-e her kes-ī nān dā=nī wāč-e ēšew be xātir=ū xwā-ī to=ū īđ=ū ī kiɫfet=ū yāne-ī=t=e dūr ∅-gin-e=we čī šār-ī \\ 
 say.prs-\textsc{imp-}\textsc{2sg} just person\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} say.prs-\textsc{imp-}\textsc{2sg} tonight to sake\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}=and 3sg.prox\textsc{.dir}\textsc{.m}=and \textsc{dem.prox} child\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} far \textsc{imp-}fall.prs-\textsc{2sg}\textsc{=compl} why city\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Tell whoever gave you food, say, “Tonight, for the sake of God, you and your household get away from this town.'
\z 
 
\ea \label{BP.168}
\textit{ā wextī bābāw pīr šelīyārī milowe dūrē leqē māɫo zīnānekeyre} \\ 
\gll ā wext-ī bābā=ū pīr šelīyār-ī mi-l-o=we dūr-ē leq-ē ∅-māɫ-o zīnān-ekey=re \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs\textsc{-3sg}\textsc{=compl} far\textsc{-indf} kick\textsc{-indf} \textsc{ind-}smash.prs\textsc{-3sg} prison\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Then, Pir Shaliyar’s grandfather went backwards in his cell and kicked the wall of the prison.'
\z 
 
\ea \label{BP.170}
\textit{zīnāneke šokē kerow} \\ 
\gll zīnān-eke šokē ∅-ker-o=ū \\ 
 prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} shake \textsc{ind-}do.prs\textsc{-3sg}\textsc{=ez.gen} \\ 
\glt `He made the prison tremble.'
\z 
 
\ea \label{BP.174}
\textit{zīnāneke yekser weɫādīm bo šāreke ētir ginowe serū āđēū} \\ 
\gll zīnān-eke yekser weɫādīm ∅-b-o šār-eke ētir ∅-gin-o=we ser=ū āđē=ū \\ 
 prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} totally collapsed \textsc{ind-}be.prs\textsc{-3sg} city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{disc.ptcl} \textsc{ind-}fall.prs\textsc{-3sg}\textsc{=compl} on\textsc{=ez.gen} 3pl\textsc{.dir}=and \\ 
\glt `The prison totally collapsed. The prison collapsed on them. '
\z 
 
\ea \label{BP.177}
\textit{mārāšā mēnēwe, mēnē degā be degāw dē be dē mēnē} \\ 
\gll m-ār-ā=šā m-ē-nē=we m-ē-nē degā be degā=ū de be de m-ē-nē \\ 
 \textsc{ind-}bring.prs\textsc{-3pl}\textsc{=3pl} \textsc{ind-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{ind-}come.prs\textsc{-3pl} village\textsc{.obl}.f to village.f\textsc{=ez.gen} village to \textsc{ind-}come.prs\textsc{-3pl} \\ 
\glt `They went back. They went village by village. '
\z 
 
\ea \label{BP.178}
\textit{āhā māčā īnā čū dīmne lāw beynū zāwerīnew tewēɫēne ħemey ġeybī hen} \\ 
\gll āhā m-āč-ā īnā-∅ č=ū dīm=ne lā=ū beyn=ū zāwer-ī=ne=ū tewēɫē=ne ħeme=ī ġeybī hen-∅ \\ 
 \textsc{intj} \textsc{ind-}say.prs\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{.3sg} side\textsc{.m}\textsc{=\textsc{post}} at\_the\_place\_of\textsc{=ez.gen} among\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{=\textsc{post}} \textsc{pn}=ez.\textsc{attr} invisible exist\textsc{-3sg}\textsc{.m} \\ 
\glt `It is said that, on this side of the border, there is the shrine of Hama the invisible between the villages of Zawar and Tawela.'
\z 
 
\ea \label{BP.185}
\textit{šime luāydēwe min kuřēm je lemū ženekēmne, hāmīlēne} \\ 
\gll šime luā-ydē=we min kuř-ē=m je leme=ū ženī-ekē=m=ne hāmīlē=ne \\ 
 \textsc{2pl} go\textsc{.pst}-\textsc{2pl}\textsc{=compl} \textsc{1sg} son/boy\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} from stomach.f\textsc{=ez.gen} woman.f\textsc{-def}\textsc{.obl}.f\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} pregnant.f\textsc{=\textsc{post}} \\ 
\glt `You go back. My wife is pregnant with my son.\footnote{Lit. there is a son in my wife’s belly.} She is pregnant.'
\z 
 
\ea \label{BP.189}
\textit{āđēč miqdārē be enāzew seʕātē nīšāre dīyāršo} \\ 
\gll āđē=č miqdārē be enāze=ū seʕāt-ē ∅-nīš-ā=re dīyār=š=o \\ 
 3pl\textsc{.dir}\textsc{=add} a\_little to proportion\textsc{=ez.gen} hour\textsc{.m}\textsc{-indf} \textsc{ind-}sit.prs\textsc{-3pl}\textsc{=\textsc{povb}} region\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `They (Baba Khwada and Hama Wirdikla) waited for him for about an hour.'
\z 
 
\ea \label{BP.203}
\textit{ā wextī penj šiš neferē je hewrāmāno milā hesārē muweqetīyū ħemey ġeybī kēšā} \\ 
\gll ā wext-ī penj šiš nefer-ē je hewrāmān=o mi-l-ā hesār-ē muweqetī=ū ħeme=ī ġeybī kēšā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} five six person\textsc{.m}\textsc{-indf} from \textsc{pn}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} wall\textsc{-indf} temporary\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible pull\textsc{.pst} \\ 
\glt `Then five, six people went from Hawraman [and] raised a temporary wall around Hama Ghaybi’s shrine.'
\z 
 
\ea \label{BP.208}
\textit{īne dāstānū īney, īnē dāstānū bābāw pīr šelīyārīū ħemey ġeybī bīyēnē} \\ 
\gll īne dāstāne=ū īney īnē dāstāne=ū bābā=ū pīr šelīyār-ī=ū ħeme=ī ġeybī bīyē=nē \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story.f\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \textsc{pn}=ez.\textsc{attr} invisible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `This was the story of Pir Shaliyar’s grandfather and Hama Ghaybi.'
\z 
 
\ea \label{BP.211}
\textit{werū īney diruē neđāw werū īney je lāw xwāywe negināwe diro ħīčē suālē diruēšā neđēne} \\ 
\gll wer=ū īney diruē ne-đā=ū wer=ū īney je lā=ū xwā-ī=we ne-gin-ā=we diro ħīč-ē suāl-ē diruē=šā ne-đē=ne \\ 
 out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} lie.f \textsc{neg-}give\textsc{.pst}=and out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} from at\_the\_place\_of\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \textsc{neg-}fall.prs\textsc{-3pl}\textsc{=compl} lie nothing\textsc{-indf} question\textsc{-indf} lie.f\textsc{=3pl} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `In order not to lie, in order not to be liars in the presence of God, they didn’t tell a lie.'
\z 
 
\ea \label{BP.212}
\textit{īnē dāstānū āđī tā ewā} \\ 
\gll īnē dāstāne=ū āđī tā ewā \\ 
 \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} until there \\ 
\glt `This was his tale. [It was] until there.'
\z 
 

\ea \label{ŠJ.1}
\textit{zemānew jāfir sānū luhonī bo} \\ 
\gll zemāne=ū jāfir sān=ū luhon-ī ∅-b-o \\ 
 period\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `It was during the rule of Jafir Sān in Luhon.'
\z 
 
\ea \label{ŠJ.2}
\textit{šā ʕumerū bīyārē tāze āmābo bīyārē} \\ 
\gll šā ʕumer=ū bīyārē tāze āmā b-o bīyārē \\ 
 king\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn} anyway come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3sg} \textsc{pn} \\ 
\glt `Sheikh Umer from Biyare might have just arrived at Biyare.'
\z 
 
\ea \label{ŠJ.7}
\textit{be beqēwew tifengē hīnē tifengū ā wextī herčē bīyen pā kešāne gēɫān} \\ 
\gll be beq-ēwe=ū tifeng-ē hīn-ē tifeng=ū ā wext-ī herčē bīye=n p=ā keš-ā=ne gēɫā=n \\ 
 to partridge\textsc{.m}\textsc{-indf}=and gun\textsc{-indf} \textsc{fill}\textsc{-indf} gun\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} whatever be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at=\textsc{dem.dist} mountain\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} wander\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[He had] a partridge and a gun—of whatever type it was back then—[and] would wander in those mountains.'
\z 
 
\ea \label{ŠJ.9}
\textit{duē yerē mēnē pāne xeɫkū mentīqew hīnī jāfir sānū luhonī} \\ 
\gll duē yerē m-ē-nē p=ā=ne xeɫk=ū mentīqe=ū hīn-ī jāfir sān=ū luhon-ī \\ 
 two three \textsc{ind-}come.prs\textsc{-3pl} at=\textsc{dem.dist}\textsc{=\textsc{post}} people\textsc{.m}\textsc{=ez.gen} region\textsc{.m}\textsc{=ez.gen} \textsc{fill}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Two three men went\footnote{Lit. came} there [to Biyare]; they were from Jafir Sān’s region, from Luhon.'
\z 
 
\ea \label{ŠJ.13}
\textit{milāwe lāw jāfer sānū luhonī} \\ 
\gll mi-l-ā=we lā=ū jāfer sān=ū luhon-ī \\ 
 \textsc{ind-}go.prs\textsc{-3pl}\textsc{=compl} at\_the\_place\_of\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They went back to Jafir Sān in Luhon.'
\z 
 
\ea \label{ŠJ.15}
\textit{wātšā hīne šēh ʕumerū bīyārēn, jāfir sān} \\ 
\gll wāt=šā hīne šēh ʕumer=ū bīyārē=n jāfir sān \\ 
 say\textsc{.pst}\textsc{=3pl} \textsc{fill} sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \textsc{pn} \\ 
\glt `They said, ‘It’s Sheikh Umer from Biyare [who is acting like a chieftain]!’'
\z 
 
\ea \label{ŠJ.17}
\textit{vātiš qurbān seʕbē nemo hīn tifengēwew beqēwe} \\ 
\gll vāt=iš qurbān seʕbe-ē ∅-nem-o hīn tifeng-ēwe=ū beq-ēwe \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} morning.f\textsc{-obl}.f \textsc{ind-}lift.prs\textsc{-3sg} \textsc{fill} gun\textsc{-indf}\textsc{=ez.gen} partridge\textsc{.m}\textsc{-indf} \\ 
\glt `They said, ‘Sir, he lifts a gun and a partridge early in the morning.'
\z 
 
\ea \label{ŠJ.27}
\textit{wextū ʕesrī mēwe} \\ 
\gll wext=ū ʕesr-ī m-ē=we \\ 
 time\textsc{.m}\textsc{=ez.gen} evening\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{.3sg}\textsc{=compl} \\ 
\glt `In the evening, when he returned,'
\z 
 
\ea \label{ŠJ.28}
\textit{ʕāyetū īnsānē bēʕeqɫē tifengšā koɫēwe tifengšā bow} \\ 
\gll ʕāyet=ū īnsān-ē bēʕeqɫ-ē tifeng=šā koɫē=we tifeng=šā ∅-b-o=ū \\ 
 habit\textsc{=ez.gen} human\textsc{.m}\textsc{-dir}\textsc{.pl} silly\textsc{\textsc{-pl}} gun\textsc{=3pl} shoulder\textsc{=compl} gun\textsc{=3pl} \textsc{ind-}be.prs\textsc{-3sg}=and \\ 
\glt `According to the customs, the silly men [from Luhon] had guns;'
\z 
 
\ea \label{ŠJ.29}
\textit{wēšā be pīyāw jāfir sānī zānā} \\ 
\gll wē=šā be pīyā=ū jāfir sān-ī ∅-zān-ā \\ 
 \textsc{reflx}\textsc{=3pl} to man\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know.prs\textsc{-3pl} \\ 
\glt `They were boasting about being Jafir Sān’s men.'
\z 
 
\ea \label{ŠJ.31}
\textit{ēme xo mātiɫū tonīmē} \\ 
\gll ēme xo mātiɫ=ū to=nīmē \\ 
 \textsc{1pl} dis.ptc suspended\textsc{=ez.gen} \textsc{2sg}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `We have been waiting for you.'
\z 
 
\ea \label{ŠJ.32}
\textit{pey lāw to āmēnmēw} \\ 
\gll pey lā=ū to āmē=nmē=ū \\ 
 to at\_the\_place\_of\textsc{=ez.gen} \textsc{2sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}=and \\ 
\glt `We have come to you.’'
\z 
 
\ea \label{ŠJ.33}
\textit{fermāwo pīyew kēndē} \\ 
\gll ∅-fermāw-o pīye=w kē=ndē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} man\textsc{.m}\textsc{=ez.gen} who\textsc{=cop}.\textsc{2pl} \\ 
\glt `He said, ‘Whose men are you?’'
\z 
 
\ea \label{ŠJ.34}
\textit{māčā pīyew jāfir sānīnmē} \\ 
\gll m-āč-ā pīye=w jāfir sān-ī=nmē \\ 
 \textsc{ind-}say.prs\textsc{-3pl} man\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `They said, ‘We are Jafir Sān’s men.’'
\z 
 
\ea \label{ŠJ.37}
\textit{fermāwo hurbēzdē gumē bīdē gumē bīdē werū čemīmo!} \\ 
\gll ∅-fermāw-o hur- b-ēz-dē gum-ē ∅-b-īdē gum-ē ∅-b-īdē wer=ū čem-ī=m=o \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{pvb-}\textsc{imp-}rise.prs-\textsc{2pl} lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} out\_of\textsc{=ez.gen} eye\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} \\ 
\glt ` He said, ‘Get up and go away! Get out of my sight!'
\z 
 
\ea \label{ŠJ.39}
\textit{čenū gum bīyey gumē bīyā serew řoē gumē bā mezānā bilāwe pey nođšey} \\ 
\gll čenū gum bīyey gum-ē bī-ā sere=ū řo-ē gum-ē ∅-b-ā me-zān-ā bi-l-ā=we pey nođše-ī \\ 
 with\textsc{.ez.gen} lost be\textsc{.inf} lost\textsc{-indf} be\textsc{.pst}\textsc{-3pl} \textsc{clf}\textsc{=ez.gen} day\textsc{.m}\textsc{-dir}\textsc{.pl} lost\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-3pl}\textsc{=compl} to \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[They got so scared that] they were lost. They were lost for one day. They didn’t know the way to go back to Nodsha.'
\z 
 
\ea \label{ŠJ.42}
\textit{dimāw serew řoē milāwe} \\ 
\gll dimā=ū sere=ū řo-ē mi-l-ā=we \\ 
 afterwards\textsc{=ez.gen} \textsc{clf}\textsc{=ez.gen} day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `After a day, they returned [to Nodsha].'
\z 
 
\ea \label{ŠJ.51}
\textit{kirž māčo be wiɫā[xdār] ī pīyāw yāneyš} \\ 
\gll kirž m-āč-o be wiɫāxdār ī pīyā=ū yāne-ī=š \\ 
 quick \textsc{ind-}say.prs\textsc{-3sg} to stableman \textsc{dem.prox} man\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `Quickly, he said to the stableman, to the servant in his house,'
\z 
 
\ea \label{ŠJ.58}
\textit{šē ʕusmānū tewēɫīn tāteš} \\ 
\gll šē ʕusmān=ū tewēɫe-ī=n tāte=š \\ 
 sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-nmlz}\textsc{=cop}\textsc{.3sg}\textsc{.m} father\textsc{.m}\textsc{=3sg} \\ 
\glt `Sheikh Usman, his [Umer’s] father, was from Tawela.'
\z 
 
\ea \label{ŠJ.60}
\textit{zānī bārbenū wiɫāxāš čikowen?} \\ 
\gll ∅-zān-ī bārben=ū wiɫāx-ā=š či=ko=we=n \\ 
 \textsc{ind-}know.prs-\textsc{2sg} stableman\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=3sg} in=where\textsc{=\textsc{post}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `‘Do you know where the horses’ stable is?’'
\z 
 
\ea \label{ŠJ.62}
\textit{māčo berem pištū berew wiɫāxekā} \\ 
\gll m-āč-o ∅-ber-e=m pišt=ū bere=ū wiɫāx-ekā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}take.prs-\textsc{2sg}\textsc{=\textsc{1sg}} back\textsc{=ez.gen} door\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} \\ 
\glt `He (Jafir Sān) said, ‘Take me behind the door of the horses’ stable!’'
\z 
 
\ea \label{ŠJ.63}
\textit{yewāšē milo pištū berew wiɫāxekā,} \\ 
\gll yewāšē mi-l-o pišt=ū bere=ū wiɫāx-ekā \\ 
 then \textsc{ind-}go.prs\textsc{-3sg} back\textsc{=ez.gen} door\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} \\ 
\glt `He went behind the door of the horses’ stable.'
\z 
 
\ea \label{ŠJ.66}
\textit{seʕbē wiɫāxdārē milā pey hīn lāw wiɫāxā tewēɫī wiɫāxē ber bārā bilā šūnū hermānērew kāērew hēzmīre} \\ 
\gll seʕbe-ē wiɫāxdār-ē mi-l-ā pey hīn lā=ū wiɫāx-ā tewēɫe-ī wiɫāx-ē ber ∅-bār-ā bi-l-ā šūn=ū hermānē=re=ū kāē=re=ū hēzmī=re \\ 
 morning.f\textsc{-obl}.f stableman\textsc{-dir}\textsc{.pl} \textsc{ind-}go.prs\textsc{-3pl} to \textsc{fill} at\_the\_place\_of\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-obl}\textsc{.pl} stable\textsc{.m}\textsc{-obl}\textsc{.m} horse/donkey\textsc{.m}\textsc{-dir}\textsc{.pl} out \textsc{sbjv-}load\textsc{.m}\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-3pl} after\textsc{=ez.gen} work\textsc{.obl}.f\textsc{=\textsc{post}}=and dry\_grass\textsc{.dir}\textsc{.pl}\textsc{=\textsc{post}}=and firewood.f\textsc{=\textsc{post}} \\ 
\glt ` In the morning, the horse grooms went to the horses in the stable. They [wanted] to take the horses out [and] go after their jobs, [collecting] straw, and wood.'
\z 
 
\ea \label{ŠJ.68}
\textit{kirž milā xeber miđā be še ʕusmānū tewēɫē} \\ 
\gll kirž mi-l-ā xeber mi-đe-ā be še ʕusmān=ū tewēɫē \\ 
 quick \textsc{ind-}go.prs\textsc{-3pl} news\textsc{.dir}\textsc{.m} \textsc{ind-}give.prs\textsc{-3pl} to sheikh \textsc{pn}\textsc{=ez.gen} \textsc{pn} \\ 
\glt `They went quickly [and] and passed the news to Sheikh Usman'
\z 
 
\ea \label{ŠJ.71}
\textit{māčo qurbān hīn jāfir sānū luhonī īnā tewēɫew wiɫāxekāne} \\ 
\gll m-āč-o qurbān hīn jāfir sān=ū luhon-ī īnā-∅ tewēɫe=ū wiɫāx-ekā=ne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} \textsc{fill} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} deic\textsc{-3sg}\textsc{.m} stable\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=\textsc{post}} \\ 
\glt `He (the horse groom) said, ‘Sir, thingummy, Jafir Sān from Luhon is in the horses’ stable!’'
\z 
 
\ea \label{ŠJ.75}
\textit{kārū āđī īnā bīyārē, mišo bilo bīyārē} \\ 
\gll kār=ū āđī īnā-∅ bīyārē mišo bi-l-o bīyārē \\ 
 task\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} deic\textsc{-3sg}\textsc{.m} \textsc{pn} aux \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{pn} \\ 
\glt ` His business is in Biyare. He shall go to Biyare.'
\z 
 
\ea \label{ŠJ.79}
\textit{kārū āđī īnā ečā} \\ 
\gll kār=ū āđī īnā-∅ e=čā \\ 
 task\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} deic\textsc{-3sg}\textsc{.m} in=there \\ 
\glt `His business is taken care of there.”’'
\z 
 
\ea \label{ŠJ.83}
\textit{milo xizmetū šēh ʕumerī bīyārēū} \\ 
\gll mi-l-o xizmet=ū šēh ʕumer-ī bīyārē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3sg} service\textsc{.m}\textsc{=ez.gen} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{pn}=and \\ 
\glt `He went to the service of Sheikh Umer.'
\z 
 
\ea \label{ŠJ.85}
\textit{nīšārew qisē fermāwāw firīū} \\ 
\gll ∅-nīš-ā=re=ū qisē-ē ∅-fermāw-ā=ū firī=ū \\ 
 \textsc{ind-}sit.prs\textsc{-3pl}\textsc{=\textsc{povb}}=and talk.f\textsc{-dir}\textsc{.pl} \textsc{ind-}say.prs\textsc{-3pl}\textsc{=ez.gen} many\textsc{=ez.gen} \\ 
\glt `They sat and talked a lot'
\z 
 
\ea \label{ŠJ.86}
\textit{fermāwo qurbān hermāne be desū min nīyenew} \\ 
\gll ∅-fermāw-o qurbān hermāne be des=ū min nīye=ne=ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} work\textsc{.f} to hand\textsc{.m}\textsc{=ez.gen} \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `He (the sheikh) said, ‘Sir, the task [of ruling] is not in my hands;'
\z 
 
\ea \label{ŠJ.87}
\textit{be desū to nīyenew} \\ 
\gll be des=ū to nīye=ne=ū \\ 
 to hand\textsc{.m}\textsc{=ez.gen} \textsc{2sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `It is not in your hands either.'
\z 
 
\ea \label{ŠJ.88}
\textit{kārū ēme mānā īse īnā desū xwāyne} \\ 
\gll kār=ū ēme mānā īse īnā-∅ des=ū xwā-ī=ne \\ 
 task\textsc{.m}\textsc{=ez.gen} \textsc{1pl} that\_is now deic\textsc{-3sg}\textsc{.m} hand\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Our task now rests within God’s hands.'
\z 
 
\ea \label{ŠJ.91}
\textit{eger pīyāgerī bo, eger šēxyetī bo, eger nāmē bo řīk megino mentēqēwene de neferē bā de nānēšā bā yo mezāno řū we kē kero} \\ 
\gll eger pīyāgerī ∅-b-o eger šēxyetī ∅-b-o eger nāmē ∅-b-o řīk me-gin-o mentēqe-ēwe=ne de nefer-ē ∅-b-ā de nān-ē=šā ∅-b-ā yo me-zān-o řū=we kē ∅-ker-o \\ 
 if gentlemanhood \textsc{sbjv-}be.prs\textsc{-3sg} if sheikhhood \textsc{sbjv-}be.prs\textsc{-3sg} if name.f \textsc{sbjv-}be.prs\textsc{-3sg} arranged \textsc{neg-}fall.prs\textsc{-3sg} region\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} ten person\textsc{.m}\textsc{-indf} \textsc{sbjv-}be.prs\textsc{-3pl} ten bread\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3pl} \textsc{sbjv-}be.prs\textsc{-3pl} one\textsc{.m} \textsc{neg-}know.prs\textsc{-3sg} face=to who \textsc{sbjv-}do.prs\textsc{-3sg} \\ 
\glt `Say this job is about being a gentleman, being a sheikh, being well known, [still] it couldn’t be that if in a region there are ten people [worthy of ruling] who have power over ten families, one doesn’t know whom to plead with [when there is a problem]'
\z 
 
\ea \label{ŠJ.94}
\textit{šēx herčē fermāwištē fermāwebā šē ʕumerū bīyārē fermāwo peyšū} \\ 
\gll šēx herčē fermāwišt-ē fermāwe b-ā šē ʕumer=ū bīyārē ∅-fermāw-o pey=š=ū \\ 
 sheikh\textsc{.m} every speech\textsc{.m}\textsc{-dir}\textsc{.pl} say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3pl} sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} for\textsc{=3sg}=and \\ 
\glt `Whatever saying there was, Sheikh Umer told him'
\z 
 
\ea \label{ŠJ.95}
\textit{pēšīmān bowew je hāɫū wēšne memeno} \\ 
\gll pēšīmān b-o=we=ū je hāɫ=ū wē=š=ne me-men-o \\ 
 regretful be.prs\textsc{-3sg}\textsc{=\textsc{post}}=and from state\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{neg-}remain.prs\textsc{-3sg} \\ 
\glt `He (Jafir Sān) regretted it. He changed his ways.'
\z 
 
\ea \label{ŠJ.97}
\textit{ētir bo welhāsil bo muxlīsū āđī jāfir sānū luhonī} \\ 
\gll ētir ∅-b-o welhāsil ∅-b-o muxlīs=ū āđī jāfir sān=ū luhon-ī \\ 
 \textsc{disc.ptcl} \textsc{sbjv-}be.prs\textsc{-3sg} eventually \textsc{ind-}be.prs\textsc{-3sg} disciple\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Anyway, Jafir Sān from Luhon became his disciple.'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{ŠJ.102}
\textit{enāne hetā fermāwī ī mentēqew ʕērāqū ērān desū ānīšāwe bīyen} \\ 
\gll enā=ne hetā ∅-fermāw-ī ī mentēqe=ū ʕērāq=ū ērān des=ū ānīšā=we bīye=n \\ 
 otherwise\textsc{=\textsc{post}} even \textsc{ind-}say.prs-\textsc{2sg} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{=ez.gen} \textsc{pn} hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All these regions in Iraq and Iran were under their rule.'
\z 
 
\ea \label{ŠJ.103}
\textit{desū ānīšāwe bīyen} \\ 
\gll des=ū ānīšā=we bīye=n \\ 
 hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `It was under their rule.'
\z 
 
\ea \label{ŠJ.106}
\textit{čūn čī? werū čēšī?} \\ 
\gll čūn čī wer=ū čēš-ī \\ 
 because why out\_of\textsc{=ez.gen} what\textsc{-obl}\textsc{.m} \\ 
\glt `Why? For what reason?'
\z 
 
\ea \label{ŠJ.107}
\textit{pīyāw dinyēn, derđwārdew dinyēn.} \\ 
\gll pīyā=ū dinyē=n derđwārde=ū dinyē=n \\ 
 man\textsc{.m}\textsc{=ez.gen} world\textsc{.obl}.f\textsc{=cop}\textsc{.3sg}\textsc{.m} qualified\textsc{.m}\textsc{=ez.gen} world\textsc{.obl}.f\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is worldly. He is experienced in world affairs.'
\z 
 
\ea \label{KŠ.1}
\textit{ew wextī īnē her girđ řāzū eyemnē lāšo bīyenā eyām} \\ 
\gll ew wext-ī īnē her girđ řāz=ū eye=m=nē lā=š=o bīye=nā eyā=m \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox}\textsc{.3pl} just all tale\textsc{.m}\textsc{=ez.gen} mother\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3pl} at\_the\_place\_of\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} mother\textsc{=\textsc{1sg}} \\ 
\glt `Well, these are all my mother’s tales. I was with my mother.'
\z 
 
\ea \label{KŠ.3}
\textit{wāčē ħukmū ā wextī milo kešene} \\ 
\gll wāč-ē ħukm=ū ā wext-ī mi-l-o keš=ene \\ 
 say\textsc{.prs}\textsc{-pstc}\textsc{.3sg} ruler\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} mountain\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `According to her,\footnote{Lit. She said.} the king of time went to a mountain.'
\z 
 
\ea \label{KŠ.4}
\textit{milo serū hānēwe esɫeheš pene bo} \\ 
\gll mi-l-o ser=ū hāne-ēwe esɫehe=š pene ∅-b-o \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} on\textsc{=ez.gen} water\_spring\textsc{.m}\textsc{-indf} gun\textsc{.m}\textsc{=3sg} with \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He went to the edge of a water spring. He had a gun with him.'
\z 
 
\ea \label{KŠ.5}
\textit{miđyo werū wečā ejoš bāɫdārā dānšo} \\ 
\gll mi-đye-o wer=ū weč-ā ejo=š bāɫdār-ā dā=n=š=o \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} front\textsc{.m}\textsc{=ez.gen} branch\textsc{.m}\textsc{-3pl} guess\textsc{.prs}\textsc{=3sg} bird\textsc{.m}\textsc{-3pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `He realised that ... He thought he had shot birds on the branches of tree.'
\z 
 
\ea \label{KŠ.12}
\textit{māčo, xās to čārenwīsnī, kināčēw pādšāyt īse pey kē niwīse? nāmēš pey kē?} \\ 
\gll m-āč-o xās to čārenwīs=nī kināčē=ū pādšā-ī=t īse pey kē niwīs-e nāmē=š pey kē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} fine \textsc{2sg} fortune\_teller\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} now to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} name\textsc{.f}\textsc{=3sg} to who \\ 
\glt `He (the king) said, ‘All right, [assuming that] you’re a fortune teller, whom have you considered [as the future husband] for the king’s daughter?\footnote{Lit. Whom did you write the king’s daughter for?} Who is she chosen for?'
\z 
 
\ea \label{KŠ.13}
\textit{māčo, niwīsem pey kuřū šuwānī’} \\ 
\gll m-āč-o niwīs-e=m pey kuř=ū šuwāne-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} write\textsc{.pst}\textsc{-3sg}\textsc{.f}\textsc{=\textsc{1sg}} to son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the fortune teller) said, ‘I have chosen her for a shepherd’s son.’'
\z 
 
\ea \label{KŠ.15}
\textit{ā wextīyekey, yewāšē wātiš pey, kināčēw pādšāyt pey kē niwīse?} \\ 
\gll ā wext-ī-ekey yewāšē wāt=iš pey kināčē=ū pādšā-ī=t pey kē niwīs-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} then say\textsc{.pst}\textsc{=3sg} to daughter\textsc{.dir}\textsc{.pl}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} to who write\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Then, he (the king) said, ‘Whom have you considered [as the future husband] for the king’s daughter?'
\z 
 
\ea \label{KŠ.16}
\textit{wātiš, pey kuřū šuwāney’} \\ 
\gll wāt=iš pey kuř=ū šuwāne-ī \\ 
 say\textsc{.pst}\textsc{=3sg} to son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the fortune teller) said, ‘[I have chosen her] for the shepherd’s son.’'
\z 
 
\ea \label{KŠ.22}
\textit{mudetē šī pene, zāwɫēweš dī jā ʕemrew xwāyš kerde} \\ 
\gll mudet-ē šī pene zāwɫe-ēwe=š dī-∅ jā ʕemre=ū xwā-ī=š kerd-e \\ 
 while\textsc{-indf} go\textsc{.pst}\textsc{.3sg} to child\textsc{-indf}\textsc{=3sg} see\textsc{.pst}\textsc{-3sg}\textsc{.m} afterwards order\textsc{.f}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `A while passed. She gave birth to a child before she passed away.\footnote{Lit. She saw a child then did the command of God.}'
\z 
 
\ea \label{KŠ.23}
\textit{ā wextī zāwɫeke peyā bī šūnū peyā bīyeyšre eyākēš merđe} \\ 
\gll ā wext-ī zāwɫe-eke peyā bī-∅ šūn=ū peyā bīyey=š=re eđā-ekē=š merđ-e \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} child\textsc{-def}\textsc{.dir}\textsc{.m} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} after\textsc{=ez.gen} visible be\textsc{.inf}\textsc{=3sg}\textsc{=\textsc{post}} mother\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} die\textsc{.pst}\textsc{-3sg}\textsc{.f} \\ 
\glt `Well, the child was born. His mother died after his birth.'
\z 
 
\ea \label{KŠ.24}
\textit{her je sīnew eyākēš īstīfāde kerdē tā šuwāney heywāniš berd pey herdekey} \\ 
\gll her je sīne=w eđā-ekē=š īstīfāde kerdē-ē tā šuwāne-ī heywān=iš berd-∅ pey herd-ekey \\ 
 just from chest\textsc{.m}\textsc{=ez.gen} mother\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} use do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-3pl} until shepherd\textsc{.m}\textsc{-obl}\textsc{.m} animal\textsc{.m}\textsc{=3sg} take\textsc{.pst}\textsc{-3sg}\textsc{.m} to rocky\_mountain\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `He (the child) fed from his mother’s breast until the shepherd took the animals to the rocky mountain.'
\z 
 
\ea \label{KŠ.25}
\textit{zāwɫeke āmā luā bizū māmālēwe šotiš wārd} \\ 
\gll zāwɫe-eke āmā luā bize=ū māmā-lē-ēwe šot=iš wārd-∅ \\ 
 child\textsc{-def}\textsc{.dir}\textsc{.m} come\textsc{.pst}\textsc{.3sg} go\textsc{.pst}\textsc{.3sg} goat\textsc{.f}\textsc{=ez.gen} grandmother\textsc{.f}\textsc{-dim}\textsc{.f}\textsc{-indf} milk\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `The child went and drank milk from the goat of a grandmother.'
\z 
 
\ea \label{KŠ.27}
\textit{dehfew yerey māmālē šikātiš kerd šuwāney} \\ 
\gll dehfe=w yere-ī māmā-lē šikāt=iš kerd šuwāne-ī \\ 
 time\textsc{.f}\textsc{=ez.gen} third\textsc{-obl}\textsc{.m} grandmother\textsc{.f}\textsc{-dim}\textsc{.f} complaint\textsc{=3sg} do\textsc{.pst} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The third time, the grandmother complained to the shepherd.'
\z 
 
\ea \label{KŠ.31}
\textit{bizekē luā čērwe kuřekey memew bizekēš wārdū} \\ 
\gll bize-ekē luā čēr=we kuř-ekey meme=ū bize-ekē=š wārd=ū \\ 
 goat\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} go\textsc{.pst}\textsc{.3sg} under\textsc{=compl} son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} breast\textsc{.dir}\textsc{.m}\textsc{=ez.gen} goat\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} eat\textsc{.pst}=and \\ 
\glt `and that the goat went there. The boy fed from the goat’s udder.'
\z 
 
\ea \label{KŠ.32}
\textit{wextē miđyo hinū wēšne gēroš beġeɫ kero mārošo} \\ 
\gll wext-ē mi-đye-o hin=ū wē=š=ne ∅-gēr-o=š beġeɫ ∅-ker-o m-ār-o=š=o \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} ez.pron\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.f} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} embrace\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `When he saw that it was his son, he took him, cuddled him, and brought him back [home]'
\z 
 
\ea \label{KŠ.33}
\textit{māčāš pene kuřew šuwāney} \\ 
\gll m-āč-ā=š pene kuř-e=w šuwāne-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} with son/boy\textsc{.m}\textsc{-def}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He was called kuraw Shwanay ‘the shepherd’s son’'
\z 
 
\ea \label{KŠ.34}
\textit{kuřew šuwāney ēnne wešɫe bo,} \\ 
\gll kuř-e=w šuwāne-ī ēnne wešɫe ∅-b-o \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} so\_much adorable \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The shepherd’s son was adorable.'
\z 
 
\ea \label{KŠ.38}
\textit{ew wextīekey yewāšē tā bo be tāqetē xeɫkī, xeɫk beroš yānew wēš} \\ 
\gll ew wext-ī-ekey yewāšē tā ∅-b-o be tāqet-ē xeɫk-ī xeɫk ∅-ber-o=š yāne=ū wē=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} then until \textsc{ind-}be\textsc{.prs}\textsc{-3sg} by support\textsc{.m}\textsc{-indf} people\textsc{.m}\textsc{-obl}\textsc{.m} people\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=3sg} house\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg} \\ 
\glt `Well then, as he became an entertainer for people, people would take him into their own houses.'
\z 
 
\ea \label{KŠ.40}
\textit{kināčēw pāđšāyč māčo, kuřū šuwāneyč bārdē ēme tāqetmā pey āney mē} \\ 
\gll kināčē=ū pāđšā-ī=č m-āč-o kuř=ū šuwāne-ī=č b-ār-dē ēme tāqet=mā pey āney m-ē \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} \textsc{1pl} persistence\textsc{.m}\textsc{=\textsc{1pl}} to \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `The king’s daughter said, ‘Fetch the shepherd’s son. He will entertain us.\footnote{Lit. Our patience comes by him.}’'
\z 
 
\ea \label{KŠ.41}
\textit{kuřū šuwāney berā yānešā} \\ 
\gll kuř=ū šuwāne-ī ∅-ber-ā yāne=šā \\ 
 son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3pl} house\textsc{.m}\textsc{=3pl} \\ 
\glt `They (the king’s men) took the shepherd’s son to their (the king’s) house.'
\z 
 
\ea \label{KŠ.42}
\textit{kuřū šuwāney ēnne wešmekū weškelāmū juwānxās bo menowe} \\ 
\gll kuř=ū šuwāne-ī ēnne wešmek=ū weškelām=ū juwānxās ∅-b-o ∅-men-o=we \\ 
 son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} so\_much amusing=and eloquent=and gentlemanly \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `The shepherd’s son, as he was so amusing, eloquent, and gentlemanly, he stayed [at the king’s palace] '
\z 
 
\ea \label{KŠ.43}
\textit{māčo be eyāšū tāteyšū āđīšā, bā īne menowe pey nokerū ēme epēne pey werdesīmā pey hermānberū yāneymā} \\ 
\gll m-āč-o be eyā=š=ū tāte-ī=š=ū āđīšā bā īne ∅-men-o=we pey noker=ū ēme epēne pey werdes-ī=mā pey hermānber=ū yāne-ī=mā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by mother\textsc{=3sg}=and father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and 3pl\textsc{.obl} \textsc{hort} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} to servant\textsc{.m}\textsc{=ez.gen} \textsc{1pl} here to attendant\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} to worker\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `She (the girl) said to her parents and others , ‘Let him stay here as our servant here, as our attendant, as our domestic worker.'
\z 
 
\ea \label{KŠ.45}
\textit{hekē xerīkū nosebāɫxīš bo} \\ 
\gll hekē xerīk=ū nosebāɫx-ī=š ∅-b-o \\ 
 when busy\textsc{=ez.gen} teenager\textsc{.m}\textsc{-nmlz}\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He was about to be a teenager.'
\z 
 
\ea \label{KŠ.48}
\textit{berāšā dūrū mentēqew ewēgewe} \\ 
\gll ∅-ber-ā=šā dūr=ū mentēqe=ū ewēge=we \\ 
 \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3pl} far\textsc{=ez.gen} region\textsc{.m}\textsc{=ez.gen} there\textsc{=\textsc{post}} \\ 
\glt `They (the king and his attendants) were taken to a region far from their own.'
\z 
 
\ea \label{KŠ.49}
\textit{yewāšē e čāge kuře menowe e čē yānene bo henū nīyenū yāney} \\ 
\gll yewāšē e=čāge kuř-e ∅-men-o=we e=čē yāne=ne ∅-b-o hen=ū nīyen=ū yāne-ī \\ 
 then in=there son/boy\textsc{.m}\textsc{-def} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} in=here house\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} being\textsc{.m}=and non\_being\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The boy stayed there [in the palace] to assist with necessary things in the house.'
\z 
 
\ea \label{KŠ.50}
\textit{e čē nāmē minvīsā kesūkārū xānī kīyānāšre isfehān} \\ 
\gll e=čē nāme-ē mi-nvīs-ā kesūkār=ū xān-ī ∅-kīyān-ā=š=re isfehān \\ 
 in=here letter\textsc{.m}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3pl} relative\textsc{.m}\textsc{=ez.gen} chief\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt `Then, the king’s relatives wrote a letter [and] sent it to Isfahan [to the king]'
\z 
 
\ea \label{KŠ.51}
\textit{māčo, be xwā pāđšām seɫāmet kuřū šuwāney heke to kuštēū ā wextīyekey hukmit dāū īnīšā īse īnā yānetā īnīšā} \\ 
\gll m-āč-o be xwā pāđšā=m seɫāmet kuř=ū šuwāne-ī heke to kušt-ē=ū ā wext-ī-ekey hukm=it dā=ū īnīšā īse īnā-∅ yāne=tā īnīšā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} king\textsc{.m}\textsc{=\textsc{1sg}} healthy son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} if \textsc{2sg} kill\textsc{.pst}\textsc{-3pl}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} rule\textsc{.m}\textsc{=\textsc{2sg}} give\textsc{.pst}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} now \textsc{deic}\textsc{-3sg}\textsc{.m} house\textsc{.m}=\textsc{2pl} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `They said [in the letter], ‘Indeed, may my king be healthy, the shepherd’s son whom you ordered to be killed is now in your house and so on’'
\z 
 
\ea \label{KŠ.54}
\textit{āđīč milo čenīū duē yerē milo} \\ 
\gll āđ=īč mi-l-o čenī=ū duē yerē mi-l-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} with\textsc{=ez.gen} two three \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the shepherd’s son) too went, together with two, three [persons].'
\z 
 
\ea \label{KŠ.55}
\textit{āđ milo melāqātīw āđī} \\ 
\gll āđ mi-l-o melāqātī=ū āđī \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} visitor\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `He (the shepherd’s son) went to visit him (the king).'
\z 
 
\ea \label{KŠ.56}
\textit{yewāšē čāge, čogene šewē minwīso, māčo, ey kesūkārū min, ey henū nīyenū min, ey īnīšā ke mintā gereknā, tāyfew min heɫbetene ī kuře biřiš dāwe, be her jor ī kuřī beynne berdē.’} \\ 
\gll yewāšē čāge č=oge=ne šewe-ē mi-nwīs-o m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey īnīšā ke min=tā gerek=nā tāyfe=ū min heɫbete=ne ī kuř=e biř=iš dā=we be her jor ī kuř-ī beyn=ne ∅-ber-dē \\ 
 then there in=there\textsc{=\textsc{post}} night\textsc{.f}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} rel \textsc{1sg}=\textsc{2pl} necessary\textsc{=cop}\textsc{.\textsc{1sg}} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} by any means \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} between\textsc{=\textsc{post}} \textsc{ind-}take\textsc{.prs}-\textsc{2pl} \\ 
\glt `Then, one night he (the king) wrote [a letter], saying, ‘O my relatives, O my close family members , O those who want me, [are] from my tribe, [when you read this letter] it is obvious that this boy has arrived. By any means possible, kill him.'
\z 
 
\ea \label{KŠ.59}
\textit{yewāšē dewrū dimāw ʕīšāy kero, šewe kero wārāy wāro mē ew peřū degē} \\ 
\gll yewāšē dewr=ū dimā=ū ʕīšā-ī ∅-ker-o šewe ∅-ker-o wārāy ∅-wār-o m-ē ew peř=ū degē \\ 
 then around\textsc{\textsc{=ez.gen}} afterwards\textsc{\textsc{=ez.gen}} evening\_prayers\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} night\textsc{.f} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} rain\textsc{.adv} \textsc{ind-}rain\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f} \\ 
\glt `Then, it was either during the evening prayers or during the night that he arrived at the other side of the village while it was raining. '
\z 
 
\ea \label{KŠ.63}
\textit{āđīč gēɫo āšpezxānekene kučē čīw menebo čāštū čīw wero} \\ 
\gll āđ=īč ∅-gēɫ-o āšpezxāne-eke=ne kuč-ē čīw mene b-o čāštī=ū čīw ∅-wer-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}wander\textsc{.prs}\textsc{-3sg} kitchen\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} little\textsc{-indf} thing\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} meal\textsc{.f}\textsc{\textsc{=ez.gen}} thing\textsc{.m} \textsc{ind-}eat\textsc{.prs}\textsc{-3sg} \\ 
\glt `He looked into the kitchen. There was some leftovers. He ate them. '
\z 
 
\ea \label{KŠ.65}
\textit{kināčekē hāyš bowe wextē miđyo āmāno lāw tāteyčšo mēw} \\ 
\gll kināčē-ekē hāy=š ∅-b-o=we wext-ē mi-đye-o āmā=n=o lā=ū tāte-ī=č=š=o m-ē=ū \\ 
 girl\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} awareness\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}=and \\ 
\glt `The [king’s] daughter became aware [of the coughing], and noticed that he (the shepherd’s son) had returned; he had returned from the place of her father.'
\z 
 
\ea \label{KŠ.68}
\textit{kuřē juwānxās lāw tāteyšo mē} \\ 
\gll kuř-ē juwānxās lā=ū tāte-ī=š=o m-ē \\ 
 boy\textsc{.m}\textsc{-indf} gentlemanly at\_the\_place\_of\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `A beautiful boy had come from her father’s place.'
\z 
 
\ea \label{KŠ.72}
\textit{niwīsebēš, ey kesūkārū min, ey henū nīyenū min, heɫbetene ī kuřī biřiš dāwe, šime be wesīlēwe kuždēš, beynne berdēš!} \\ 
\gll niwīse b-ē=š ey kesūkār=ū min ey hen=ū nīyen=ū min heɫbete=ne ī kuř-ī biř=iš dā=we šime be wesīle-ēwe ∅-kuž-dē=š beyn=ne ∅-ber-dē=š \\ 
 write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} \textsc{2pl} by tool\textsc{.m}\textsc{-indf} \textsc{imp-}kill\textsc{.prs}-\textsc{2pl}\textsc{=3sg} between\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2pl}\textsc{=3sg} \\ 
\glt `He had written, ‘O my relatives, O my close family members, [when you read this letter] it is evident that this boy has arrived. Kill him with a thing, destroy him!'
\z 
 
\ea \label{KŠ.74}
\textit{miđyo kuřē noxetū juwānxāsū tāze yāwā penew} \\ 
\gll mi-đye-o kuř-ē noxet=ū juwānxās=ū tāze yāwā pene=ū \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} boy\textsc{.m}\textsc{-indf} just\_grown\_beard=and gentlemanly\textsc{\textsc{=ez.gen}} anyway arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m} to=and \\ 
\glt `She saw that he had just grown a beard; he was beautiful, and he had just hit puberty.'
\z 
 
\ea \label{KŠ.75}
\textit{lāw tātew āđēwe mēw} \\ 
\gll lā=ū tāte=ū āđē=we m-ē=ū \\ 
 at\_the\_place\_of\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{\textsc{=ez.gen}} 3sg\textsc{.obl}\textsc{.f}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}=and \\ 
\glt `He had returned from the place of her father and such.'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.81}
\textit{yewāšē seʕbē hurmēzāwe māčo, xeyrāmāyš kero merhebāyš kero māčo, tāteym hāɫū ēhwāɫtā?’} \\ 
\gll yewāšē seʕbe-ē hur- mi-ēz-ā=we m-āč-o xeyrāmāy=š ∅-ker-o merhebā-ī=š ∅-ker-o m-āč-o tāte-ī=m hāɫ=ū ēhwāɫ=tā \\ 
 then morning\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} welcoming\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} hello\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} state\textsc{.m}\textsc{\textsc{=ez.gen}} demeanour=\textsc{2pl} \\ 
\glt `The following day, they woke up. She said … she welcomed him (the shepherd’s son), greeted him [and] said, ‘How is my father doing?’'
\z 
 
\ea \label{KŠ.82}
\textit{māčo, tātet weš bēw hāɫū ēhwāɫiš xās bē} \\ 
\gll m-āč-o tāte=t weš b-ē=ū hāɫ=ū ēhwāɫ=iš xās b-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} father\textsc{.m}\textsc{=\textsc{2sg}} good be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and state\textsc{.m}\textsc{\textsc{=ez.gen}} demeanour\textsc{=3sg} fine be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (the shepherd’s son) said, ‘Your father was all right; he was feeling well.'
\z 
 
\ea \label{KŠ.83}
\textit{xulāsew kelāmī qisēš čenū min kerdēw} \\ 
\gll xulāse=ū kelām-ī qisē=š čenū min kerd-ē=ū \\ 
 summary=and speech\textsc{-obl}\textsc{.m} talk\textsc{.dir}\textsc{.pl}\textsc{=3sg} with\textsc{\textsc{.ez}.gen} \textsc{1sg} do\textsc{.pst}\textsc{-3pl}=and \\ 
\glt `In short, he (the king) talked to me.'
\z 
 
\ea \label{KŠ.85}
\textit{pey lāw šime ārdenmo} \\ 
\gll pey lā=ū šime ārde=n=m=o \\ 
 to at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=compl} \\ 
\glt `I have brought it for you.'
\z 
 
\ea \label{KŠ.92}
\textit{yewāšē šūnū duē sāɫēre kināčē nīšo lemewe kuřē wīno} \\ 
\gll yewāšē šūn=ū duē sāɫe-ē=re kināčē ∅-nīš-o leme=we kuř-ē ∅-wīn-o \\ 
 then after\textsc{\textsc{=ez.gen}} two year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{.f} \textsc{ind-}sit\textsc{.prs}\textsc{-3sg} stomach\textsc{.f}\textsc{=compl} boy\textsc{.m}\textsc{-indf} \textsc{ind-}see\textsc{.prs}\textsc{-3sg} \\ 
\glt `After two years, the girl became pregnant\footnote{Lit. She sits on belly.} [and] gave birth to a son.'
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{KŠ.97}
\textit{yewāšē neferēwe řemo weɫēwe māčo be xānī, be xwā āne to kārū xwāyt tefre dā,} \\ 
\gll yewāšē nefer-ēwe ∅-řem-o weɫē=we m-āč-o be xān-ī be xwā āne to kār=ū xwā-ī=t tefre dā \\ 
 then person\textsc{.m}\textsc{-indf} \textsc{ind-}run\textsc{.prs}\textsc{-3sg} front\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by chief\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2sg} task\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} evasion give\textsc{.pst} \\ 
\glt `Then, a person went forward and said to the Khan, ‘Indeed, you [tried to] evade God’s destiny.'
\z 
 
\ea \label{KŠ.98}
\textit{yārū tifengeket teqnā, ženū šuwāney kušī,} \\ 
\gll yārū tifeng-eke=t teqnā ženī=ū šuwāne-ī ∅-kuš-ī \\ 
 \textsc{fill} gun\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} shoot\textsc{.pst} woman\textsc{.f}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}kill\textsc{.prs}-\textsc{2sg} \\ 
\glt `You fired a bullet from your gun; [and] you [intended] to kill the shepherd’s wife.'
\z 
 
\ea \label{KŠ.99}
\textit{ā wextīyekey čārenwīsī niwīsebēš ā kināčē to pey kuřū šuwāney niwīsyene} \\ 
\gll ā wext-ī-ekey čārenwīs-ī niwīse b-ē=š ā kināčē to pey kuř=ū šuwāne-ī niwīsye=ne \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} fortune\_teller\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{dem.dist} daughter\textsc{.dir}\textsc{.f} \textsc{2sg} to son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} write\textsc{.pst}.pass\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `Then, the fortune teller had allocated\footnote{Lit. written.} your daughter for the shepherd’s son. '
\z 
 
\ea \label{KŠ.100}
\textit{āne kuřū šuwāneyn serū esbekey to} \\ 
\gll āne kuř=ū šuwāne-ī=n ser=ū esb-ekey to \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} on\textsc{\textsc{=ez.gen}} horse\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{2sg} \\ 
\glt `Look, the shepherd’s son [is] on your horse.'
\z 
 
\ea \label{KŠ.105}
\textit{kārū to řāsā, kes metāwoš tefre biđo} \\ 
\gll kār=ū to řās=ā kes me-tāw-o=š tefre bi-đe-o \\ 
 task\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} true\textsc{=cop}\textsc{.3sg}\textsc{.m} person\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-3sg}\textsc{=3sg} evasion \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `Your deed is right! No one can avoid destiny.'
\z 
 
\ea \label{ŽH.1}
\textit{min zāroɫe bēnē, lāw eyemo} \\ 
\gll min zāroɫe b-ēn-ē lā=ū eye=m=o \\ 
 \textsc{1sg} child be\textsc{.prs}\textsc{-pstc}\textsc{-1sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} mother\textsc{.obl}\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `I was a child. [I was] with my mother.'
\z 
 
\ea \label{ŽH.2}
\textit{eyām qisē kerdē wāčē, zemānū qedīmī pādšā hāřūne(l) řešīd bīyenū nā siɫtān mehmūw heyāsī žīr} \\ 
\gll eyā=m qisē-ē kerdē-ē wāč-ē zemān=ū qedīm-ī pādšā hāřūn-e(l) řešīd bīye=n=ū nā siɫtān mehmū=ū heyās=ī žīr \\ 
 mother\textsc{=1sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-3pl} say\textsc{.prs-}pstc\textsc{.3sg} time\textsc{.m}\textsc{=ez}\textsc{.gen} old\_time\textsc{-obl}\textsc{.m} king\textsc{.m} \textsc{pn}-of \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and no sultan\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez}.\textsc{attr} shrewd \\ 
\glt `My mother would narrate [tales]. She said, ‘In the olden days, there was Harun Rashid, the king. No, [there was] Sultan Mahmud and Hayas, the shrewd.'
\z 
 
\ea \label{ŽH.3}
\textit{wāčē řoē je řoā qāɫešā bo, serū mezān ū čēšī} \\ 
\gll wāč-ē řo-ē je řo-ā qāɫe=šā ∅-b-o sere=ū me-zān-ū čēš-ī \\ 
 say\textsc{.prs-}pstc\textsc{.3sg} day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} quarrel\textsc{\textsc{.f}}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} head\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what\textsc{-obl}\textsc{.m} \\ 
\glt `According to her (my mother),\footnote{Lit. She would say.}  Once they (Sultan Mahmud and Hayas) got into a quarrel with each other, about something I don’t know.'
\z 
 
\ea \label{ŽH.5}
\textit{e čē milo bilo pey ew dīmū kelī} \\ 
\gll e=čē mi-l-o bi-l-o pey ew dīm=ū kel-ī \\ 
 in=here \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} for \textsc{dem.dist} side\textsc{.m}\textsc{=ez}\textsc{.gen} mountain\textsc{-obl}\textsc{.m} \\ 
\glt `He went from here to the other side of the mountain.'
\z 
 
\ea \label{ŽH.6}
\textit{wiɫātēwe termā hen ew dīmew ēme, pēsew zāwerīū tewēɫey} \\ 
\gll wiɫāt-ēwe ter=mā hen-∅ ew dīm=ew ēme pēse=ū zāwer-ī=ū tewēɫe-ī \\ 
 country\textsc{.m}\textsc{-indf} other\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} \textsc{dem.dist} side\textsc{.m}\textsc{=\textsc{post}} \textsc{1pl} like\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `There is another country on the other side of us; something like Zawar (in Iran) and Tawela (in Iraq) village.'
\z 
 
\ea \label{ŽH.8}
\textit{ā wextīyekey yewāšē milo derey, tūšū neferēwe mē} \\ 
\gll ā wext-ī-ekey yewāšē mi-l-o dere-ī tūš=ū nefer-ēwe m-ē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{fill} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} valley\textsc{.m}\textsc{-obl}\textsc{.m} encounter\textsc{=ez}\textsc{.gen} person\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `Then, he went to a slope [and] ran into a person.'
\z 
 
\ea \label{ŽH.15}
\textit{milā řāhur, milā nizīkū degākē gināwe,} \\ 
\gll mi-l-ā řā=hur mi-l-ā nizīk=ū degā-ekē ∅-gin-ā=we \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} close\textsc{=ez}\textsc{.gen} village\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} ind\textsc{-f}all\textsc{.prs}\textsc{-3pl}\textsc{=compl} \\ 
\glt `They continued [along] the road. They got close to the village.'
\z 
 
\ea \label{ŽH.23}
\textit{māčo, ne wiɫā řoɫe, āmānē pāynū serekoē wāčmēš pene be zuwānū hewrāmīyānew wēmā derey} \\ 
\gll m-āč-o ne wiɫā řoɫe āmā-ānē pāyn=ū sereko-ē ∅-wāč-mē=š pene be zuwān=ū hewrāmī-yāne=ū wē=mā dere-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no by\_God child.\textsc{voc} come\textsc{.pst}\textsc{-1sg} bottom\textsc{=ez}\textsc{.gen} uphill\textsc{.m}\textsc{-indf} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-1pl}\textsc{=3sg} to by language\textsc{.m}\textsc{=ez}\textsc{.gen} Hawrami-vernacular\_of\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} valley\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘No, child, I went to the bottom of the slope —we call it ‘daray’ in our Hawrami language—'
\z 
 
\ea \label{ŽH.30}
\textit{wātiš, qisēmā nekerdē āmāymē nizīkū dege,} \\ 
\gll wāt=iš qisē=mā ne-kerd-ē āmā-īmē nizīk=ū dege \\ 
 say\textsc{.pst}\textsc{=3sg} talk\textsc{.dir}\textsc{.pl}\textsc{=1pl} \textsc{neg-}do\textsc{.pst}\textsc{-3pl} come\textsc{.pst} \textsc{1pl} close\textsc{=ez}\textsc{.gen} village\textsc{.obl}\textsc{\textsc{.f}} \\ 
\glt `He said, ‘We didn’t talk [until] we got close to the village.'
\z 
 
\ea \label{ŽH.40}
\textit{wātiš be wāɫe wičkilekēš ī māsī bere deš pā mēmān mizgī čenīw nānī’} \\ 
\gll wāt=iš be wāɫē=e wičkile-ekē=š ī mās-ī ∅-ber-e ∅-de-e=š p=ā mēmān mizgī čenī=ū nān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} by sister\textsc{\textsc{.f}}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} to\textsc{=dem}.dist guest\textsc{.m} mosque\textsc{.m} with\textsc{=ez}\textsc{.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said to her little daughter, ‘Take this yoghurt [and] together with [some] bread give it to the the guest in the mosque.’ '
\z 
 
\ea \label{ŽH.57}
\textit{ā wextīyekey āyeč māčo, weɫāhī tāte, dāwāw ēme yoymā kero’} \\ 
\gll ā wext-ī-ekey āye=č m-āč-o weɫāhī tāte dāwā=w ēme yo-ī=mā ∅-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg\textsc{\textsc{.f}}\textsc{.dir}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed father\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{1pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then she (the eldest daughter) said, ‘Indeed, father, he is asking for the hand of one of us.’'
\z 
 
\ea \label{ŽH.62}
\textit{īne e čē čenū āġeyš jengšā bīyen, be ādizī lūān} \\ 
\gll īne e=čē čenū āġe-ī=š jeng=šā bīye=n be ādizī lūā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=here with.ez\textsc{.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} quarrel\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} with desperation go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (Hayas) got into a quarrel with his master. He had left [town] in desperation'
\z 
 
\ea \label{ŽH.64}
\textit{qerār pāse bo e čē xānū ewēgē āđ čenū āđī suɫtān mehmū bē hečkes bē qāɫešā bī hāčkes ženē xāse bāro bē ewēge āđ bilo bo mēmānšā} \\ 
\gll qerār pāse ∅-b-o e=čē xān=ū ewēge-ē āđ čenū āđī suɫtān mehmū b-ē hečkes b-ē qāɫe=šā bī-∅ hāčkes ženī-ē xās-e b-ār-o b-ē ewēge āđ bi-l-o ∅-b-o mēmān=šā \\ 
 arrangement like \textsc{ind-}be\textsc{.prs}\textsc{-3sg} in=here chief\textsc{.m}\textsc{=ez}\textsc{.gen} there\textsc{-obl}\textsc{\textsc{.f}} 3sg\textsc{.dir}\textsc{.m} with.ez\textsc{.gen} 3sg\textsc{.obl}\textsc{.m} sul \textsc{pn} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} whoever be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} quarrel\textsc{\textsc{.f}}\textsc{=3pl} be\textsc{.pst}\textsc{-3sg}\textsc{.m} whoever woman\textsc{\textsc{.f}}\textsc{-indf} fine\textsc{-f} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} there \textsc{3sg}\textsc{.dir}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} guest\textsc{.m}\textsc{=3pl} \\ 
\glt `In the old times, the rule was that — the khan of there, whether it was Sultan Mahmud, with whom he (Hayas) had got into a quarrel, or anyone else— whoever got married and took a beautiful wife, he (the king) would go [and] become their guest.'
\z 
 
\ea \label{ŽH.65}
\textit{yewāšē āđīč mārošo mēwe kināčēw luhonī bo teřū čermew īnīšā} \\ 
\gll yewāšē āđ=īč m-ār-o=š=o m-ē=we kināčē=ū luhon-ī ∅-b-o teř=ū čerme=ū īnīšā \\ 
 then 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} fresh=and white=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `He (Hayas) brought her (i.e. his wife). She was from Luhon; she was young, white, and so on.'
\z 
 
\ea \label{ŽH.66}
\textit{māčā be siɫtān mehmūī, be xwā fiɫān heyās āmāno čīwē ečīnew ečāne ʕāliš ārden’} \\ 
\gll m-āč-ā be siɫtān mehmū-ī be xwā fiɫān heyās āmā=n=o čīw-ē e=čīne=w e=čāne ʕāl=iš ārde=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} by sultan\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} by God\textsc{.m} such\_and\_such \textsc{pn} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} thing\textsc{-indf} in=like\_this\textsc{=ez}\textsc{.gen} in=like\_that good\textsc{=3sg} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (people) said to Sultan Mahmud, ‘Such-and-such Hayas has returned. He has brought [with him] such a nice thinggummy (i.e. woman)  '
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.110}
\textit{šūnū duē řoeyre, kināčekē kīyāno lāw suɫtānī māčo, ēme pīyākemā kīyāstiš meʕmūrīyet nīyā, nāmāno’} \\ 
\gll šūn=ū duē řo-e-ī=re kināčē-ekē ∅-kīyān-o lā=ū suɫtān-ī m-āč-o ēme pīyā-eke=mā kīyāst-∅=iš meʕmūrīyet nīye=ā ne-āmā=n=o \\ 
 after\textsc{=ez}\textsc{.gen} two\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} sul\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} send\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=3sg} mission\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `After the second day, the girl (his wife) sent (someone) to the Sultan to tell [him], ‘He (the sultan) sent my husband on a mission, [however] he is not around; He hasn’t come back [yet].’'
\z 
 
\ea \label{ŽH.117}
\textit{kārēš pene kerdenā be mehremū wēm qubuɫim kerden.} \\ 
\gll kār-ē=š pene kerde=nā be mehrem=ū wē=m qubuɫ=im kerde=n \\ 
 task\textsc{.m}\textsc{-indf}\textsc{=3sg} to do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `She made  me such that I have accepted her as my own kindred. '
\z 
 
\ea \label{ŽH.119}
\textit{dey duʕāw seɫāmetīt} \\ 
\gll dey duʕā=ū seɫāmetī=t \\ 
 \textsc{disc.ptcl} prayer\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} health\textsc{=\textsc{2sg}} \\ 
\glt `That’s it. May good prayers  be for you.'
\z 



\ea \label{RE.2}
\textit{ēme eger zemānē yā qeymīēmā yā zemānū wēmā ženīmā ārdēbo} \\ 
\gll ēme eger zemān-ē yā qeymī-ē=mā yā zemān=ū wē=mā ženī=mā ārdē b-o \\ 
 \textsc{1pl} if time\textsc{.m}\textsc{-indf} either the\_elderly\textsc{-indf}\textsc{=\textsc{1pl}} either time\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1pl}} woman\textsc{.dir}\textsc{.f}\textsc{=\textsc{1pl}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f} be\textsc{.prs}\textsc{-3sg} \\ 
\glt ` Once, in the time of our elders or in our time, if one of us married,'
\z 
 
\ea \label{RE.3}
\textit{āmēnē lūēnmē werū berew kābrāy, hāwmāɫū wēmā} \\ 
\gll āmē=nē lūē=nmē wer=ū bere=ū kābrā-ī hāwmāɫ=ū wē=mā \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} front\textsc{.m}\textsc{\textsc{=ez.gen}} door\textsc{.m}\textsc{\textsc{=ez.gen}} man\textsc{-obl}\textsc{.m} neighbour\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1pl}} \\ 
\glt `they (our elderly family members) would go to the door of the fellow (the girl’s father), [who was] the neighbour.'
\z 
 
\ea \label{RE.6}
\textit{be bīyew nebīyeym zānī} \\ 
\gll be bīyey=w nebīyey=m ∅-zān-ī \\ 
 to resources\textsc{.m}\textsc{\textsc{=ez.gen}} s\textsc{hort}age\textsc{.m}\textsc{=\textsc{1sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} \\ 
\glt `You know about my resources and shortages. '
\z 
 
\ea \label{RE.7}
\textit{ʕerzū birāw wēm kerū ī kināčēte de pī kuřīme} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū ī kināčē=t=e ∅-de-e p=ī kuř-ī=m=e \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} brother\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} at=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \\ 
\glt `Let me tell you,\footnote{Lit. May a petition be at the service of my brother.} ‘Give your daughter to my son [in marriage]'
\z 
 
\ea \label{RE.8}
\textit{ā wextī eger xwāy īrāđeš bīyebo serū qisekē} \\ 
\gll ā wext-ī eger xwā-ī īrāđe=š bīye b-o ser=ū qisē-ekē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} if God\textsc{.m}\textsc{-obl}\textsc{.m} volition\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} on\textsc{\textsc{=ez.gen}} talk\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `Then, if God was willing over the subject,'
\z 
 
\ea \label{RE.9}
\textit{eger xwāw peyġemerī īrāđeš bīyebo serū qisekē} \\ 
\gll eger xwā=ū peyġemer-ī īrāđe=š bīye b-o ser=ū qisē-ekē \\ 
 if God\textsc{.m}\textsc{\textsc{=ez.gen}} prophet\textsc{.m}\textsc{-obl}\textsc{.m} volition\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} on\textsc{\textsc{=ez.gen}} talk\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `If God and the prophet were willing [that] the affair [could happen],'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.11}
\textit{ētir kābrā fers kere zemānū ā ānē hītē pāɫē bīyēnēū čwār gezē pārče bīyenū} \\ 
\gll ētir kābrā fers ∅-ker-e zemān=ū ā ānē hīt-ē pāɫā-ē bīyē=nē=ū čwār gez-ē pārče bīye=n=ū \\ 
 \textsc{disc.ptcl} man guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} \textsc{dem.dist}\textsc{.3pl} pair\textsc{-indf} shoe\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and four ell\textsc{.m}\textsc{-indf} textile\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Then, let us say,\footnote{Lit. imagine.} at the time, [the wedding gift] consisted of a pair of shoes [and] four gaz\footnote{A unit of measure equivalent to 72 cm.} of textile.'
\z 
 
\ea \label{RE.16}
\textit{ey pīyā ētir emīnū yāne bīyenū xizmetkār bīyen} \\ 
\gll ey pīyā ētir emīn=ū yāne bīye=n=ū xizmetkār bīye=n \\ 
 \textsc{dem.prox} man\textsc{.m} \textsc{disc.ptcl} confidant\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{.dir}.sg be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez.gen} servant\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was this man who was the family’s confidant. He was a servant.'
\z 
 
\ea \label{RE.18}
\textit{firmānbirđār bīyen, tā sāhībū ī kināčē wāčo, xeber bido wāčo, fiɫāne kār} \\ 
\gll firmānbirđār bīye=n tā sāhīb=ū ī kināčē ∅-wāč-o xeber bi-de-o ∅-wāč-o fiɫān=e kār \\ 
 obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until owner\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} news\textsc{.dir}\textsc{.m} \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} such\_and\_such=ez\textsc{.cmpd} task\textsc{.m} \\ 
\glt `He was obedient. The girl’s father\footnote{Lit. owner.} would say [to him] and ask him, ‘[Do] such-and-such tasks;'
\z 
 
\ea \label{RE.22}
\textit{šūnū ženekēčre ā wextīekey ētir hemāhengīšā weš bīyen} \\ 
\gll šūn=ū ženī-ekē=č=re ā wext-ī-ekey ētir hemāhengī=šā weš bīye=n \\ 
 after\textsc{=ez.gen} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} cooperation\textsc{.m}\textsc{=3pl} good be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `After the marriage, they (the couple's families) were on good terms.'
\z 
 
\ea \label{RE.23}
\textit{qisew mise nebīyen} \\ 
\gll qise=w mise ne-bīye=n \\ 
 talk\textsc{=ez.gen} \textsc{rdp} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was not a quarrel [between them]'
\z 
 
\ea \label{RE.33}
\textit{tāte bīyen māne āsāwū ewsey} \\ 
\gll tāte bīye=n māne āsāw=ū ewse-ī \\ 
 father\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} worn mill\textsc{.m}\textsc{=ez.gen} past\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The father is no longer the authority.\footnote{Lit. The father has become like the old mills of past times.}'
\z 
 
\ea \label{RE.41}
\textit{ī memliketū ēme ne welē wiɫātāre ētir tāte bīyen čūēw desū šuwāney} \\ 
\gll ī memliket=ū ēme ne welē wiɫāt-ā=re ētir tāte bīye=n čū-ē=w des=ū šuwāne-ī \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} no but region\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{disc.ptcl} father\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} stick\textsc{.m}\textsc{-indf}\textsc{=ez.gen} hand\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[It is not common] in our region, but elsewhere, the father has become like the stick in the shepherd’s hand.\footnote{Meaning that he has no power over whom his children will marry.}'
\z 
 
\ea \label{RE.42}
\textit{ēme zemānmā řisūmātū ewsāymā ečīne bīyen} \\ 
\gll ēme zemān=mā řisūmāt=ū ewsā-ī=mā e=čīne bīye=n \\ 
 \textsc{1pl} time\textsc{.m}\textsc{=\textsc{1pl}} traditions\textsc{.m}\textsc{=ez.gen} past\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} in=like be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the old times, our customs were like this. '
\z 
 
\ea \label{RE.48}
\textit{keɫešīrē gērten bāwšīšo qoqeqoqū epāgehur} \\ 
\gll keɫešīr-ē gērte=n bāwšī=š=o qoqeqoq=ū epāgehur \\ 
 rooster\textsc{.m}\textsc{-indf} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} emrace\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{onom}\textsc{=ez.gen} up\_there \\ 
\glt `He [The girl’s father] would grab a rooster in his arms [to present to the nobleman], [while the rooster] was crowing from this side to the top [of the village]'
\z 
 
\ea \label{RE.49}
\textit{min kināčēt dēne kuřū fiɫāne kesī} \\ 
\gll min kināčē=t dē=ne kuř=ū fiɫān=e kes-ī \\ 
 \textsc{1sg} daughter\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}} son/boy\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and say to the lord], ‘I have given my\footnote{Lit. your.} daughter to the son of such-and-such person.'
\z 
 
\ea \label{RE.59}
\textit{sūrānew ženāšā sānān} \\ 
\gll sūrāne=ū ženī-ā=šā sānā=n \\ 
 \textsc{pn}\textsc{=ez.gen} woman\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the noblemen) would grab surana (i.e. money in exchange for giving consent to marriage)'
\z 
 
\ea \label{RE.62}
\textit{ī memliketū ēme řisūmātū āwe zemānīš e čīne bīyen} \\ 
\gll ī memliket=ū ēme řisūmāt=ū āwe zemān-ī=š e=čīne bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} traditions\textsc{.m}\textsc{=ez.gen} that time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} in=like be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The traditions of earlier times were like this in our region.'
\z 
 
\ea \label{ŽE.7}
\textit{hīč āwū čīwiš nebīyen} \\ 
\gll hīč āwī=ū čīw=iš ne-bīye=n \\ 
 nothing water\textsc{.f}\textsc{\textsc{=ez.gen}} thing\textsc{.m}\textsc{=3sg} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` It had no water or anything.'
\z 
 
\ea \label{ŽE.13}
\textit{ʕāɫfiš pētenū tā kočšā kerden tā ʕāɫifšā ārdeno} \\ 
\gll ʕāɫif=iš pēte=n=ū tā koč=šā kerde=n tā ʕāɫif=šā ārde=n=o \\ 
 grass\textsc{.dir}\textsc{.m}\textsc{=3sg} fold\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} until migration\textsc{.dir}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until grass\textsc{.dir}\textsc{.m}\textsc{=3pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `[and] made grass into stacks until when they returned [to the village]; until when they brought back grass '
\z 
 
\ea \label{ŽE.18}
\textit{desū dīmšā gērten nimāšā kerdēne} \\ 
\gll des=ū dīm=šā gērte=n nimā=šā kerdē=ne \\ 
 hand\textsc{.m}\textsc{\textsc{=ez.gen}} face\textsc{.m}\textsc{=3pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} prayer\textsc{.f}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `They performed ablutions [and] prayed.'
\z 
 
\ea \label{ŽE.25}
\textit{yānew ētir ne yāne bīyenū} \\ 
\gll yāne=ū ētir ne yāne=e bīye=n=\\ 
 house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{disc.ptcl} no house\textsc{.m}\textsc{=ez}\textsc{.cmpd} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `As for houses, there was no [proper] house.'
\z 
 
\ea \label{ŽE.27}
\textit{de neferē serū jāmoɫē čāštīšā nīyēnere} \\ 
\gll de nefer-ē ser=ū jāmoɫē-ē čāštī=šā nīyē=ne=re \\ 
 ten person\textsc{.m}\textsc{-dir}\textsc{.pl} on\textsc{\textsc{=ez.gen}} small\_pot\textsc{-indf} meal\textsc{.obl}\textsc{.f}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f}\textsc{=\textsc{povb}} \\ 
\glt ` They would cook meals in a single small pot for ten persons.'
\z 
 
\ea \label{ŽE.28}
\textit{šūnū de neferekeyre her čemčēšā bīyen} \\ 
\gll šūn=ū de nefer-ekey=re her čemče-ē=šā bīye=n \\ 
 after\textsc{\textsc{=ez.gen}} ten person\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} emph spoon\textsc{.m}\textsc{-indf}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, there was only one spoon for the ten people.'
\z 
 
\ea \label{ŽE.33}
\textit{luēnē eweɫū wextī lūēnē hēzmīšā biřyēnē} \\ 
\gll luē=nē eweɫ=ū wext-ī lūē=nē hēzmī=šā biřyē=nē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} first\textsc{\textsc{=ez.gen}} time\textsc{.m}\textsc{-obl}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} firewood\textsc{.dir}\textsc{.pl}\textsc{=3pl} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Early in the morning, they would go and chop wood.'
\z 
 
\ea \label{ŽE.40}
\textit{her sikeɫ bīyen berqū čīw nebīyen} \\ 
\gll her sikeɫ bīye=n berq=ū čīw ne-bīye=n \\ 
 just ember\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} electricity\textsc{.m}\textsc{\textsc{=ez.gen}} thing\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was only embers [to light the house]. There was no electricity or anything.'
\z 
 
\ea \label{ŽE.42}
\textit{čirāwe desīšā, čirāwekešā weš kerden bilā pey yānew pīr šelīyārī pey čīwī bārānšo} \\ 
\gll čirāw=e desī=šā čirāw-eke=šā weš kerde=n bi-l-ā pey yāne=ū pīr šelīyār-ī pey čīw-ī b-ār-ān=š=o \\ 
 lamp\textsc{.m}\textsc{=ez}\textsc{.cmpd} hand\textsc{=3pl} lamp\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} for house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} for thing\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `They lit the lamps to go the Pir Shaliyar’s shrine\footnote{Lit. house.} [and] to fetch things.'
\z 
 
\ea \label{ŽE.43}
\textit{eđāw tātew min her beđbext bīyen} \\ 
\gll eđā=ū tāte=ū min her beđbext bīye=n \\ 
 mother\textsc{.f}=and father\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} emph poor be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `My parents were poor.'
\z 
 
\ea \label{ŽE.45}
\textit{jāmoɫēw čāštīš nīyēnere} \\ 
\gll jāmoɫē=ū čāštī=š nīyē=ne=re \\ 
 small\_pot\textsc{\textsc{=ez.gen}} meal\textsc{.f}\textsc{=3sg} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f}\textsc{=\textsc{povb}} \\ 
\glt `She [my mother] would cook food. '
\z 
 
\ea \label{ŽE.49}
\textit{wātenšā, xwā gīyān hemdū senā pey to’} \\ 
\gll wāte=n=šā xwā gīyān hemd=ū senā pey to \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} God\textsc{.m} dear.\textsc{voc} praise\textsc{\textsc{=ez.gen}} admiration for \textsc{2sg} \\ 
\glt `[They] the would say, ‘Dear God, may our praises be for you!’'
\z 
 
\ea \label{ŽE.50}
\textit{ī doīnete dēnmēū ī šelemīnēte dēnmē} \\ 
\gll ī doīne=t=e dē=nmē=ū ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{\textsc{=ez.gen}} \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this dwina. \footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{ŽE.54}
\textit{xwā hemdū senā pey to!’} \\ 
\gll xwā hemd=ū senā pey to \\ 
 God\textsc{.m} praise\textsc{\textsc{=ez.gen}} admiration for \textsc{2sg} \\ 
\glt `God, may our praises be for you!’'
\z 
 
\ea \label{ŽE.67}
\textit{āwū nānim dēnē nīyēnēm nām bēškekē} \\ 
\gll āwī=ū nān=im dē=nē nīyē=nē=m nām bēške-ekē \\ 
 water\textsc{.f}\textsc{\textsc{=ez.gen}} bread\textsc{.m}\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} inside cot\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \\ 
\glt `I gave them food [and] put them back in the cot.'
\z 
 
\ea \label{ŽE.74}
\textit{luēnī lāw ī kuřī wāčēnī, dā lue wāče eđēš bizāne ī kināčēše miđom pene yā ne’} \\ 
\gll lu-ēn-ī lā=ū ī kuř-ī wāč-ēn-ī dā ∅-lu-e ∅-wāč-e eđē=š bi-zān-e ī kināčē=š=e mi-đe-o=m pene yā ne \\ 
 go\textsc{.prs}\textsc{-pstc}-\textsc{2sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.prs}\textsc{-pstc}-\textsc{2sg} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}say\textsc{.prs-}.\textsc{2sg} mother\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=3sg}\textsc{=dem} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{1sg}} to either no \\ 
\glt ` You would go to this boy [seated next to the narrator] and would say, ‘Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!’'
\z 
 
\ea \label{ŽE.82}
\textit{wāčē, dey lue yānew wēt nīšere tā duē māngē ter yerē māngē ter} \\ 
\gll wāč-ē dey ∅-lu-e yāne=ū wē=t ∅-nīš-e=re tā duē mānge-ē ter yerē mānge-ē ter \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc}.\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} until two month\textsc{.f}\textsc{-dir}\textsc{.pl} other three month\textsc{.f}\textsc{-dir}\textsc{.pl} other \\ 
\glt `They (the family of the girl) would say [to the boy], ‘Go back home [and] wait\footnote{Lit. sit.} for two, three months.'
\z 
 
\ea \label{ŽM.3}
\textit{min kuřū duēnā} \\ 
\gll min kuř=ū duē=nā \\ 
 \textsc{1sg} son/boy\textsc{.m}\textsc{=ez.gen} second\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I am from the second one.'
\z 
 
\ea \label{ŽM.4}
\textit{šūnū āneyre, šūnū řāy eweɫ je kārīgerīwe serem beršīyen} \\ 
\gll šūn=ū āney=re šūn=ū řā=ī eweɫ je kārīgerī=we sere=m ber-šīye=n \\ 
 after\textsc{=ez.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} after\textsc{=ez.gen} road\textsc{.f}=ez.\textsc{attr} first from labourer\_job\textsc{=compl} head\textsc{.m}\textsc{=\textsc{1sg}} out- go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Well, I first learned\footnote{Lit. my head went out.}  to work as a labourer.'
\z 
 
\ea \label{ŽM.8}
\textit{īse ħewt sāɫēč āmānāwe serū īney dūkāndārnā} \\ 
\gll īse ħewt sāɫe-ē=č āmā=nā=we ser=ū īney dūkāndār=nā \\ 
 now seven year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} shopkeeper\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{ŽM.12}
\textit{īne dāstānū minā} \\ 
\gll īne dāstāne=ū min=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{=ez.gen} \textsc{1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is my story.'
\z 
 
\ea \label{ŽM.13}
\textit{eger tūlū ʕemrīm temāmū dinyēm wiškinān pey žīwāyū dinyē} \\ 
\gll eger tūl=ū ʕemr-ī=m temām=ū dinyē=m wiškinā=n pey žīwāy=ū dinyē \\ 
 if lenght\textsc{.m}\textsc{=ez.gen} age\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}} all\textsc{=ez.gen} world\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}} search\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to live\textsc{.inf}\textsc{=ez.gen} world\textsc{.obl}\textsc{.f} \\ 
\glt `Throughout my life, I have searched all around the world to make ends meet.'
\z 
 
\ea \label{ŽM.15}
\textit{ā wextīyekey ečī newʕe řū be werū wēmre řū be xwā řisqēm peyā kerđen} \\ 
\gll ā wext-ī-ekey ečī newʕ=e řū be wer=ū wē=m=re řū be xwā řisq-ē=m peyā kerđe=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} like\_this kind/way\textsc{=dem} face by front\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} face by God\textsc{.m} living\textsc{-indf}\textsc{=\textsc{1sg}} visible do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I have taken refuge in God and found living.'
\z 
 
\ea \label{ŽM.16}
\textit{dārāw īse elʕān hāɫēhāzirī dārāw penj kuřānāw yerē kināčā, heštū wēmū ženē īne de neferē} \\ 
\gll dārā=w īse elʕān hāɫēhāzir-ī dārā=w penj kuř-ā=nā=ū yerē kināčē-ā hešt=ū wē=m=ū ženī-ē īne de nefer-ē \\ 
 owner\textsc{.m}\textsc{=ez.gen} now now at\_the\_moment\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{=ez.gen} five son/boy\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1sg}}=and three daughter\textsc{.f}\textsc{-obl}\textsc{.pl} eight=and \textsc{reflx}\textsc{=\textsc{1sg}}=and woman\textsc{.f}\textsc{-indf} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} ten person\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `I have, at the moment, five sons and three daughters, [which makes] eight, [plus] myself and a wife, [making us] ten people.'
\z 
 
\ea \label{ŽM.17}
\textit{tā īse yāwānā be řehmū xwāywe} \\ 
\gll tā īse yāwā=nā be řehm=ū xwā-ī=we \\ 
 until now arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} by mercy\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have made\footnote{Lit. arrived.} it so far, thanks to God's blessing.'
\z 
 
\ea \label{ŽM.19}
\textit{ī menteqew ēme kemūkuřīn} \\ 
\gll ī menteqe=ū ēme kemūkořī=n \\ 
 \textsc{dem.prox} region\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Our region is in poverty.'
\z 
 
\ea \label{ŽM.20}
\textit{min zemānē zemāniš bīyen ženīm ārdēne girđū kārēw nebīyey bīyen} \\ 
\gll min zemān-ē zemān=iš bīye=n ženī=m ārdē=ne girđ=ū kār-ē=ū nebīyey bīye=n \\ 
 \textsc{1sg} time\textsc{.m}\textsc{-indf} time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} woman\textsc{.f}\textsc{=\textsc{1sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} all\textsc{=ez.gen} task\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=ez.gen} poverty\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As for me, once I got married, there was poverty all around.'
\z 
 
\ea \label{ŽM.21}
\textit{ī mentēqew ēme nebīyeyš hen nebīyey} \\ 
\gll ī mentēqe=ū ēme nebīyey=š hen-∅ nebīyey \\ 
 \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} poverty\textsc{.m} \\ 
\glt `Our region is in poverty, [and] shortage.'
\z 
 
\ea \label{ŽM.25}
\textit{jezīrew kīšīne kārmā kerđen šiš māngē} \\ 
\gll jezīre=w kīš-ī=ne kār=mā kerđe=n šiš mānge-ē \\ 
 island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kish Island and work for six months.'
\z 
 
\ea \label{ŽM.27}
\textit{ī yānze zāroɫē ečē āsēnēm jīyā milū ī ženīme milū nebīyeywe} \\ 
\gll ī yānze zāroɫe-ē e=čē āsē=nē=m jīyā mil=ū ī ženī=m=e mil=ū nebīyey=we \\ 
 \textsc{dem.prox} eleven child\textsc{-dir}\textsc{.pl} in=here leave\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} separate on\textsc{=ez.gen} \textsc{dem.prox} woman\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} on\textsc{=ez.gen} poverty\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Because of destitution, [I had to] leave these eleven children to my wife.'
\z 
 
\ea \label{ŽM.32}
\textit{ā wextīyekey ečāge hīč xeberēmā nebīyenū telefūnīčū žimārew telefūnī nebēw} \\ 
\gll ā wext-ī-ekey e=čāge hīč xeber-ē=mā ne-bīye=n=ū telefūn=īč=ū žimāre=ū telefūn-ī ne-b-ē=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} in=there nothing news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and telephone\textsc{.m}\textsc{=add}=and number\textsc{.m}\textsc{=ez.gen} telephone\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and \\ 
\glt `I\footnote{Lit. We.}  had no news [of my family] from there. There was no phone or phone number at the time.'
\z 
 
\ea \label{ŽM.33}
\textit{šesū šiš sāʕetē ēme be māšīn jā luēnmē binderʕebāsū} \\ 
\gll šes=ū šiš sāʕet-ē ēme be māšīn jā luē=nmē binderʕebās=ū \\ 
 sixty\textsc{=ez.gen} six hour\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} by car\textsc{.m} afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and \\ 
\glt `It took us sixty-three hours to reach Bandar Abbas by car.'
\z 
 
\ea \label{ŽM.36}
\textit{ī šiš mānge xeberū hīčīmā nezānān} \\ 
\gll ī šiš mānge=e xeber=ū hīč-ī=mā ne-zānā=n \\ 
 \textsc{dem.prox} six month\textsc{.f}\textsc{=dem} news\textsc{.m}\textsc{=ez.gen} nothing\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I didn’t have any updates [from my family] during those periods of six months.'
\z 
 
\ea \label{ŽM.38}
\textit{extīyārmā mištū ā fārsāne bīyen jezīrew kīšī} \\ 
\gll extīyār=mā mišt=ū ā fārs-ā=ne bīye=n jezīre=ū kīš-ī \\ 
 volition\textsc{.m}\textsc{=\textsc{1pl}} fist\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} Persian\textsc{-3pl}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We were under the rule of Persians at Kish island.'
\z 
 
\ea \label{ŽM.41}
\textit{kuweyt duē sāʕetē beynšā čenū kuweytī} \\ 
\gll kuweyt duē sāʕet-ē beyn=šā čenū kuweyt-ī \\ 
 \textsc{pn} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} between\textsc{=3pl} with\textsc{\textsc{.ez}.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Kuwait was two hours away from there.'
\z 
 
\ea \label{ŽM.43}
\textit{ā zemāne kuweyt je girđū kēšwerekā dewɫetmenter bē} \\ 
\gll ā zemān=e kuweyt je girđ=ū kēšwer-ekā dewɫetmen-ter b-ē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} from all\textsc{=ez.gen} country\textsc{-def}\textsc{.obl}\textsc{.pl} rich\textsc{-cmpr} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Back then, Kuwait was the richest of all countries.'
\z 
 
\ea \label{ŽM.45}
\textit{ēme jezīre čenū ā jezīrey čeneš bēnmē duē sāʕetē beyniš bē} \\ 
\gll ēme jezīre čenū ā jezīre-ī čene=š b-ēn-mē duē sāʕet-ē beyn=iš b-ē \\ 
 \textsc{1pl} island with\textsc{\textsc{.ez}.gen} \textsc{dem.dist} island\textsc{-obl}\textsc{.m} at\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1pl}} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} distance\textsc{.m}\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `It (Kuwait) was two hours away from the island we were in.'
\z 
 
\ea \label{ŽM.49}
\textit{řoē sī penj timenē šesū yerē sāɫē čēweɫī kerū ētir meyānū sāɫū činē bē} \\ 
\gll řo-ē sī penj timen-ē šes=ū yerē sāɫe-ē čēweɫī ∅-ker-ū ētir me-zān-ū sāɫe-ū činē b-ē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} sixty\textsc{=ez.gen} three year\textsc{.f}\textsc{-dir}\textsc{.pl} ago \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc}.\textsc{ptcl} \textsc{neg-}know\textsc{.prs}\textsc{-\textsc{1sg}} year\textsc{.f}\textsc{-\textsc{1sg}} how\_many\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `A daily salary of thirty-five tomans, it was sixty-three years ago. I don’t know which year it was/'
\z 
 
\ea \label{ŽM.53}
\textit{ʕerzū birāw wēm kerū āmēnmēwe ī mentēqe} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū āmē=nmē=we ī mentēqe \\ 
 petition\textsc{.m}\textsc{=ez.gen} brother\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \\ 
\glt `Let me tell you. I returned to this region.'
\z 
 
\ea \label{ŽM.54}
\textit{ētir īsey āmēnmē tā wē yāwēnmē be kemūkuřīū hewrāmānī} \\ 
\gll ētir īse-ī āmē=nmē tā wē yāwē=nmē be kemūkořī=ū hewrāmān-ī \\ 
 \textsc{disc}.\textsc{ptcl} now\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} until there arrive\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} by poverty\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have made it so far, [despite] poverty and destitution.'
\z 
 
\ea \label{ŽM.55}
\textit{ētir īse weʕzū hewrāmānī tā menteqē terē kučē xāsterā} \\ 
\gll ētir īse weʕze=ū hewrāmān-ī tā menteqe-ē ter-ē kuč-ē xās-ter=ā \\ 
 \textsc{disc}.\textsc{ptcl} now situation\textsc{.f}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} until region\textsc{-dir}\textsc{.pl} other\textsc{-dir}\textsc{.pl} little\textsc{-indf} fine\textsc{-cmpr}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The situation in Hawraman is a bit better than in other regions.'
\z 
 
