\chapter{All 102 examples of imp in these texts are found in the following sentences:}

\ea \label{ZQ.4}
\textit{nīšdērew fermāwāš čikowe meydē?} \\ 
\gll ∅-nīš-dē=re=ū fermāwā=š či=ko=we m-e-ydē \\ 
 \textsc{imp-}sit.prs-\textsc{2pl}\textsc{=\textsc{povb}}=and say\textsc{.pst}\textsc{=3sg} from=where\textsc{=compl} \textsc{ind-}come.prs-\textsc{2pl} \\ 
\glt ``Sit down!' the Shah said, ‘Where do you come from?’'
\z 
 
\ea \label{ZQ.6}
\textit{fermāwāš, xeyr beydē’} \\ 
\gll fermāwā=š xeyr b-e-ydē \\ 
 say\textsc{.pst}\textsc{=3sg} goodness\textsc{.m} \textsc{imp-}come.prs-\textsc{2pl} \\ 
\glt `The Shah said, ‘Welcome!’'
\z 
 
\ea \label{ZQ.16}
\textit{duē, be gīrmā wārd čenū ženīū čenū heywānīū dewɫetīū bē āwīū nāřihet e pāwe} \\ 
\gll duē be gīr=mā wārd čenū ženī=ū čenū heywān-ī=ū dewɫet-ī=ū bē āwī=ū nāřihet e=pāwe \\ 
 two by hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst} with\textsc{.ez.gen} woman.f=and with\textsc{.ez.gen} animal\textsc{.m}\textsc{-obl}\textsc{.m}=and animal\textsc{-obl}\textsc{.m}=and come\textsc{.imp}\textsc{.\textsc{2sg}} water.f=and sad in=there \\ 
\glt `I was stuck there with my wife, livestock, a lack of water, and a lack of comfort. '
\z 
 
\ea \label{HB.2}
\textit{nānewēkēw šēh ʕelādīnī seʕbē hurmēzāwe xeber miđā māčā, xeber deydē be hezretū šēxī, be xwā xāneqāne ne nān hen, ne hārdī henē} \\ 
\gll nānewē-ekē=ū šēh ʕelādīn-ī seʕbe-ē hur-mi-ēz ā=we xeber mi-đe-ā mi-āč-ā xeber ∅-de-ydē be hezret=ū šēx-ī be xwā xāneqā=ne=ne nān hen-∅=ne hārdī hen-ē \\ 
 baker\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{.pl}=and sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} morning\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs} \textsc{dem.dist}\textsc{=compl} news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} news\textsc{.m} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} to his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}}=neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m}=nor flour\textsc{\textsc{.f}} exist\textsc{-3pl} \\ 
\glt `Sheikh Aladin’s bakers got up in the morning and passed on the news, saying, ‘Pass on the news to his highness the Sheikh. By God, there is neither bread nor flour in the monastery. '
\z 
 
\ea \label{HB.11}
\textit{fermāwo, dā lodē wāčdē be fiɫāne kesī bilo herekey bero} \\ 
\gll ∅-fermāw-o dā ∅-lo-dē wāč-dē be fiɫān=e kes-ī bi-l-o her-ekey ∅-ber-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} \textsc{imp-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} say\textsc{.prs}.sbjv\textsc{-2pl} by such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the Sheikh) said [to the people around him], ‘Go and say to such-and-such a person that [he should] go [and] bring his donkey; '
\z 
 
\ea \label{HB.17}
\textit{miđyo herzretū šēxī nefermāwā bilue yānew āney bāre’} \\ 
\gll mi-đye-o herzret=ū šēx-ī ne-fermāwā bi-lue yāne=ū āney b-ār-e \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}say\textsc{.pst} \textsc{sbjv-}go\textsc{.imp}\textsc{.\textsc{2sg}} house=and \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `He noticed that his highness the Sheikh hadn’t told [him], ‘Go to the house of that (i.e. such-and-such) person and bring [back the flour]’'
\z 
 
\ea \label{HB.18}
\textit{yā zeřiš neyā pene, lue čiko bisāne bāre!’} \\ 
\gll yā zeř=iš ne-dā pene ∅-lu-e či=ko bi-sān-e b-ār-e \\ 
 or money\textsc{.m}\textsc{=3sg} \textsc{neg-}give\textsc{.pst} to \textsc{imp-}go.\textsc{imp-}\textsc{2sg} from=where \textsc{imp-}buy\textsc{.prs}-\textsc{2sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `Nor had he given him money, [saying], ‘Go [and] buy from a place\footnote{Lit. from where.} [and] bring [back the flour]’'
\z 
 
\ea \label{HB.43}
\textit{pekemā megino, to goš min dāre!’} \\ 
\gll peke=mā me-gin-o to goš min ∅-dār-e \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \textsc{2sg} ear\textsc{.m} 1sg \textsc{imp-}hold\textsc{.prs}-\textsc{2sg} \\ 
\glt `We should not be worried; listen to me!’'
\z 
 
\ea \label{HB.51}
\textit{wātiš, fermāwdē bēydē mēmānū minindē} \\ 
\gll wāt=iš ∅-fermāw-dē b-ē-ydē mēmān=ū min=indē \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{imp-}say\textsc{.prs}\textsc{-2pl} \textsc{imp-}come\textsc{.prs}\textsc{-2pl} guest\textsc{.m}\textsc{=ez}\textsc{.gen} 1sg\textsc{=cop}\textsc{.2pl} \\ 
\glt `He said, ‘Please come! You are my guests.'
\z 
 
\ea \label{HB.62}
\textit{vātiš, dey xuře vene bere} \\ 
\gll vāt=iš dey ∅-xuř-e vene ∅-ber-e \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}shout\textsc{.prs}-\textsc{2sg} \textsc{post} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `[and] said, ‘Command the donkey\footnote{Lit. shout at him.}; move him ; take him;'
\z 
 
\ea \label{HB.63}
\textit{bere’} \\ 
\gll ∅-ber-e \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt ` take him. '
\z 
 
\ea \label{HB.79}
\textit{māčo, dā lodē! fermāwo dā lodē bārdēš!’} \\ 
\gll m-āč-o dā ∅-lo-dē ∅-fermāw-o dā ∅-lo-dē b-ār-dē=š \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} \textsc{ind-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} \textsc{imp-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} \textsc{sbjv-}bring\textsc{.prs}\textsc{-2pl}\textsc{=3sg} \\ 
\glt `He said [to his men] ‘Go [and] bring him!’'
\z 
 
\ea \label{PM.8}
\textit{fermāwo, lue xizmetū āđī!} \\ 
\gll ∅-fermāw-o ∅-lu-e xizmet=ū āđī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt `[and] said, ‘Go into his presence!'
\z 
 
\ea \label{PM.9}
\textit{ījāze bizāne āđ ījāze miđo ēme e čē nehār kermē yām ne} \\ 
\gll ījāze bi-zān-e āđ ījāze mi-đe-o ēme e=čē nehār ∅-ker-mē yām ne \\ 
 permission \textsc{imp-}know\textsc{.prs}-\textsc{2sg} 3sg\textsc{.dir}\textsc{.m} permission \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \textsc{1pl} in=here lunch\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} or no \\ 
\glt `See if he lets us stay here for lunch or not'
\z 
 
\ea \label{PM.16}
\textit{šēxī ʕelādīnī fermāwo, lodērā šārezūr!} \\ 
\gll šēx-ī ʕelādīn-ī ∅-fermāw-o lo-dē=rā šārezūr \\ 
 sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} go\textsc{.prs}.\textsc{imp-}\textsc{2pl}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt `Sheikh Aladin said, ‘Go down to Sharazour!'
\z 
 
\ea \label{PM.20}
\textit{fermāwebē, fermāwo lodē e čā muɫkū ēme wēmā, herčēmā hen čāne ā firmānde wēš zāno heqū ēme kāmā} \\ 
\gll fermāwe b-ē ∅-fermāw-o ∅-lo-dē e=čā muɫk=ū ēme wē=mā herčē=mā hen-∅ č=ā=ne ā firmānde wē=š ∅-zān-o heq=ū ēme kām=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} in=there property\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{reflx}\textsc{=1pl} whatever\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{dem.dist} leader\textsc{.m} \textsc{reflx}\textsc{=3sg} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} right\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} which\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go! In that property of mine, whatever we possess there, your boss knows which portion [of land] is mine.'
\z 
 
\ea \label{PM.21}
\textit{ēšew nehār kerdē, pey nehārū wētāw seʕbē xwāhāfīztā!} \\ 
\gll ēšew nehār ∅-ker-dē pey nehār=ū wē=tā=ū seʕbe-ē xwāhāfīz=tā \\ 
 tonight lunch\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2pl} for lunch\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and morning.f\textsc{-obl}.f goodbye\textsc{.m}=\textsc{2pl} \\ 
\glt `[Take it] as a camp tonight [and] eat from it as your lunch. [Then] Tomorrow, bid farewell!'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.26}
\textit{to mišo ī birāyte šewēne bere} \\ 
\gll to mišo ī birā-ī=t=e šewe-ē=ne ∅-ber-e \\ 
 \textsc{2sg} aux \textsc{dem.prox} brother\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `You should take this brother of yours [there] at night.'
\z 
 
\ea \label{DG.27}
\textit{lue wēt wize milū pāɫāyekā šā ʕelādīnīre} \\ 
\gll ∅-lu-e wē=t ∅-wiz-e mil=ū pāɫā-ekā šā ʕelādīn-ī=re \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}throw\textsc{.prs}-\textsc{2sg} on\textsc{=ez}\textsc{.gen} shoe\textsc{-def}\textsc{.obl}\textsc{.pl} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Go, throw yourself at the feet\footnote{Lit. shoes.} of Shah Aladin'
\z 
 
\ea \label{DG.44}
\textit{īđīč fermāwo, lue čā nīšere,} \\ 
\gll īđ=īč ∅-fermāw-o ∅-lu-e čā ∅-nīš-e=re \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} there \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \\ 
\glt `Shah Aladin  said, ‘Go [and] sit there;'
\z 
 
\ea \label{DG.45}
\textit{lue xāneqā tā xeberit dewwe pene’} \\ 
\gll ∅-lu-e xāneqā tā xeber=it ∅-de-ū=we pene \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} monastery\textsc{.m} until news\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}give\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{compl}} to \\ 
\glt `Go to the monastery until I let you know.'
\z 
 
\ea \label{DG.49}
\textit{winū kīseɫē zāmdāre kere} \\ 
\gll winī=ū kīseɫī-ē zāmdār-e ∅-ker-e \\ 
 blood\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} tortoise\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} wounded\textsc{-f} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Make a tortoise wounded!'
\z 
 
\ea \label{DG.50}
\textit{winekēš sāwe pā qelbū pīyāyre} \\ 
\gll winī-ekē=š ∅-sāw-e p=ā qelb=ū pīyā-ī=re \\ 
 blood\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} by\textsc{=dem}.dist heart\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `[then] rub its blood on the heart of the [sick] man'
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.57}
\textit{lue xāneqā nān bere, bideš penew} \\ 
\gll ∅-lu-e xāneqā nān ∅-ber-e bi-de=š pene=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} monastery\textsc{.m} bread\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}give\textsc{.prs}\textsc{=3sg} to=and \\ 
\glt `Go to the monastery; take food [and] give it to him.'
\z 
 
\ea \label{DG.58}
\textit{ā wextīyekey, gēɫdē lodē penā bere be xwāy} \\ 
\gll ā wext-ī-ekey ∅-gēɫ-dē ∅-lo-dē penā ∅-ber-e be xwā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}wander\textsc{.prs}\textsc{-2pl} \textsc{imp-}go\textsc{.prs}\textsc{.imp}\textsc{-2pl} refuge \textsc{imp-}take\textsc{.prs}-\textsc{2sg} by God\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, go back [to your place]; Take refuge with God.'
\z 
 
\ea \label{DG.66}
\textit{ē.. dewāyekeš sāwe pore dā ēme kečē nebīmē} \\ 
\gll ē.. dewāī-eke=š ∅-sāw-e pore dā ēme keč-ē ne-b-īmē \\ 
 eh medication\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} at \textsc{hort} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs}-\textsc{1pl} \\ 
\glt `He shall rub the medicine on him; he shall not make us crooked.'
\z 
 
\ea \label{DG.68}
\textit{berešo memliketū wētā welē sermā deydē} \\ 
\gll ∅-ber-e=š=o memliket=ū wē=tā welē ser=mā ∅-de-ydē \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=\textsc{compl}} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=2pl} but head\textsc{.m}\textsc{=1pl} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} \\ 
\glt `Take him to your region, but pay us a visit.'
\z 
 
\ea \label{DP.24}
\textit{to luew čiko āmānī luewewe} \\ 
\gll to lue=ū či=ko āmā=nī lue=wewe \\ 
 \textsc{2sg} go\textsc{.imp}\textsc{.\textsc{2sg}}=and from=where come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} go\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `Go back to where you came from, go back!'
\z 
 
\ea \label{DP.39}
\textit{bē bā nīšmēre seḷāhū wēmā kermē} \\ 
\gll bē bā ∅-nīš-mē=re seḷāh=ū wē=mā ∅-ker-mē \\ 
 come\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{hort} \textsc{sbjv-}sit\textsc{.prs}\textsc{-1pl}\textsc{=\textsc{povb}} solution\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `[He said to Sibhan Agha], ‘Let us sit together and come to a solution.’'
\z 
 
\ea \label{DP.42}
\textit{māčo, ne bē!’} \\ 
\gll mi-āč-o ne ∅-bē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{imp-}come\textsc{.imp}\textsc{.\textsc{2sg}} \\ 
\glt `Jamsher Shah\footnote{Lit. He} said, ‘No [I won’t arrest you]. Come!’'
\z 
 
\ea \label{DP.43}
\textit{lue wāče bē!’} \\ 
\gll ∅-lue ∅-wāč-e ∅-bē \\ 
 \textsc{imp-}go\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}say\textsc{.prs-}\textsc{2sg} \textsc{imp-}come\textsc{.imp}\textsc{.\textsc{2sg}} \\ 
\glt `[He said to his men], ‘Go [and] tell him to come!’'
\z 
 
\ea \label{DP.51}
\textit{dā dey dewršā gērdē bā deġāɫet kerū} \\ 
\gll dā dey dewr=šā ∅-gēr-dē bā deġāɫet ∅-ker-w \\ 
 \textsc{hort} \textsc{disc.ptcl} round\textsc{=3pl} \textsc{imp-}take\textsc{.prs}\textsc{-2pl} \textsc{hort} intervention\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt ` Jamsher Shah\footnote{Lit. He.} said [to his men], ‘Encircle them so that I can defeat [Sibhan Agha].'
\z 
 
\ea \label{ZP.39}
\textit{māčo, ī kināčē bere degā be degāw} \\ 
\gll m-āč-o ī kināčē ∅-ber-e degā be degā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.obl}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `[and] said, ‘Take this girl village by village,'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.70}
\textit{fermāwo, gēɫdēwe bēheyāyā wāz bārdē čā zāyfē!’} \\ 
\gll ∅-fermāw-o ∅-gēɫ-dē=we bēheyā-yā wāz b-ār-dē č=ā zāyfe-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}wander\textsc{.prs}-\textsc{2pl}\textsc{=compl} shameless\textsc{-obl}\textsc{.pl} request \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} from=\textsc{dem.dist} woman\textsc{-obl}\textsc{.f} \\ 
\glt `He said, ‘Go back, you shameless [demons]! Let go of that poor girl!'
\z 
 
\ea \label{ZP.106}
\textit{berešo. mubārekū sāhēbīš bo!’} \\ 
\gll ∅-ber-e=š=o mubārek=ū sāhēb-ī=š ∅-b-o \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=compl} blessed\textsc{\textsc{=ez.gen}} owner\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take her back. May she be happy with her owner (i.e., father)!'
\z 
 
\ea \label{ŽP.18}
\textit{kīyānāš māčā, heywānekey bere} \\ 
\gll ∅-kīyān-ā=š m-āč-ā heywān-ekey ∅-ber-e \\ 
 \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} animal\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `They (his uncle’s family) sent him [shepherding]. They said, ‘Take the animals ….'
\z 
 
\ea \label{ŽP.21}
\textit{heywānekey bere pey ewā} \\ 
\gll heywān-ekey ∅-ber-e pey ewā \\ 
 animal\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to there \\ 
\glt `Take the animals there [the valley of Pir Shaliyar] '
\z 
 
\ea \label{ŽP.24}
\textit{milo mēweū māčā, hizmēč bāre’} \\ 
\gll mi-l-o m-ē=we=ū m-āč-ā hizmē=č b-ār-e \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl}=and \textsc{ind-}say\textsc{.prs}\textsc{-3pl} firewood\textsc{.f}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt ` He went [shepherding and] came back [home]. They said, ‘Bring firewood as well.’'
\z 
 
\ea \label{ŽP.43}
\textit{māčo bē dengē māmo yoso, dā lue šonū sey misefāyre bizāne řās māčo!} \\ 
\gll m-āč-o bē- deng-ē māmo yoso dā ∅-lu-e šūn=ū sey misefā-ī=re bi-zān-e řās m-āč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}without noise\textsc{.m}\textsc{-indf} paternal\_uncle \textsc{pn} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} after\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} truth \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \\ 
\glt `Uncle Yoso secretly said [to the servant], ‘Go after Say Mustafa, see if he is telling the truth!'
\z 
 
\ea \label{ŽP.95}
\textit{luewe} \\ 
\gll ∅-lu-e=we \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `Go back!’'
\z 
 
\ea \label{ŽP.103}
\textit{to lue zeřekey bere čenū wēt pey xerjīt lodē} \\ 
\gll to ∅-lu-e zeř-ekey ∅-ber-e čenū wē=t pey xerjī=t ∅-lo-dē \\ 
 \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} with\textsc{.ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} to consumption\textsc{=\textsc{2sg}} \textsc{imp-}go\textsc{.prs}.\textsc{imp-}\textsc{2pl} \\ 
\glt `You go, take the money with you for your expenses.'
\z 
 
\ea \label{ŽP.104}
\textit{zeřekey bere min, qey mekero} \\ 
\gll zeř-ekey ∅-ber-e min qey me-ker-o \\ 
 money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{1sg} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take the money. I will be fine.’'
\z 
 
\ea \label{ŽP.159}
\textit{to bo lue! ī kināčē bere} \\ 
\gll to b-o ∅-lu-e ī kināčē ∅-ber-e \\ 
 \textsc{2sg} \textsc{imp-}come\textsc{.prs}\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `‘Come [here and] go! Take the girl.'
\z 
 
\ea \label{ŽP.162}
\textit{dēhāt be dēhāt ī kināčē bere} \\ 
\gll dēhāt be dēhāt ī kināčē ∅-ber-e \\ 
 village\textsc{.f} by village\textsc{.f} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `Take the girl village by village.'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.212}
\textit{berešo qey mekeronew} \\ 
\gll ∅-ber-e=š=o qey me-ker-one=ū \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Take her back. It is all right!'
\z 
 
\ea \label{ŽP.262}
\textit{lue! to luewe hem wesāylīč bere pey ī pīyāyū ī kināčēw} \\ 
\gll ∅-lu-e to ∅-lu-e=we hem wesāyl=īč ∅-ber-e pey ī pīyā-ī=ū ī kināčē=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} both appliance\textsc{.m}\textsc{=add} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}=and dem.prox daughter\textsc{.dir}\textsc{.pl}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter'
\z 
 
\ea \label{ŽP.263}
\textit{ētir to lue meylū wētā’} \\ 
\gll ētir to ∅-lu-e meyl=ū wē=t=ā \\ 
 \textsc{disc.ptcl} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} willingness\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then it’s up to you [to stay in Hawraman]'
\z 
 
\ea \label{BP.77}
\textit{fermāwo, dā lodē bizāndē fiɫāne kes čī nāmā?} \\ 
\gll ∅-fermāw-o dā ∅-lo-dē bi-zān-dē fiɫān=e kes čī ne-āmā-∅ \\ 
 \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{sbjv-}know.prs-\textsc{2pl} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m} why \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `He (Baba Khwada) said, ‘Go [and] see\footnote{Lit. know} why such-and-such person hasn’t come [to the mosque]?'
\z 
 
\ea \label{BP.78}
\textit{wāče, bē dērā ēme milāy milmē pey ʕērāqī be peyē} \\ 
\gll wāč-e b-ē dēr=ā ēme mi-lāy mi-l-mē pey ʕērāq-ī be peyē \\ 
 say.prs-\textsc{imp-}\textsc{2sg} be.prs\textsc{-pstc}\textsc{.3sg} late\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} \textsc{ind-}go\textsc{.adv} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} to foot\textsc{.obl} \\ 
\glt `Tell him, “Come, it’s getting late. We are going to Iraq on foot.'
\z 
 
\ea \label{BP.86}
\textit{āđīč mē werew māčo, fermāwo, dā dey dey jārēne jimdēre’} \\ 
\gll āđ=īč m-ē wer=ew m-āč-o ∅-fermāw-o dā dey dey jār-ē=ne ∅-jim-dē=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}come.prs\textsc{.3sg} front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{disc.ptcl} \textsc{disc.ptcl} time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} \\ 
\glt `He (Baba Khwada) came forward [and] said, ‘Please, set off the way all of you.’'
\z 
 
\ea \label{BP.87}
\textit{jārēne jimdēre mexɫūq hečē henē mizgīne bā bilimdē pey ʕērāqī} \\ 
\gll jār-ē=ne ∅-jim-dē=re mexɫūq hečē hen-ē mizgī=ne ∅-b-ā bi-l-imdē pey ʕērāq-ī \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} people\textsc{.m} no\_matter\_how\_many exist\textsc{-3pl} mosque\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}/\textsc{2pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They all set off; every person who was in the mosque [said], ‘Let’s go to Iraq.'
\z 
 
\ea \label{BP.129}
\textit{āġe māčo, dey lodē berdē kerdēšā zīnān} \\ 
\gll āġe m-āč-o dey ∅-lo-dē ∅-ber-dē ∅-ker-dē=šā zīnān \\ 
 agha\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}take.prs-\textsc{2pl} \textsc{imp-}do.prs-\textsc{2pl}\textsc{=3pl} prison\textsc{.m} \\ 
\glt `The agha said [to his men], ‘Go [and] take them; put them in jail!’'
\z 
 
\ea \label{BP.130}
\textit{lodē kerdēšā zīnān} \\ 
\gll ∅-lo-dē ∅-ker-dē=šā zīnān \\ 
 \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}do.prs-\textsc{2pl}\textsc{=3pl} prison\textsc{.m} \\ 
\glt `Put them in jail.'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.162}
\textit{wāče, kučē dūr ginowe”} \\ 
\gll wāč-e kuč-ē dūr ∅-gin-o=we \\ 
 say.prs-\textsc{imp-}\textsc{2sg} little\textsc{-indf} far \textsc{ind-}fall.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `Tell [them], “Go to a remote place”'
\z 
 
\ea \label{BP.164}
\textit{gošiš nedārāyne ētir jā bowe xeber bārewe} \\ 
\gll goš=iš ne-dārā-ī=ne ētir jā b-o=we xeber b-ār-e=we \\ 
 ear\textsc{.m}\textsc{=3sg} \textsc{neg-}hold\textsc{.pst}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{disc.ptcl} afterwards \textsc{imp-}come.prs\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{imp-}bring.prs-\textsc{2sg}\textsc{=compl} \\ 
\glt `If they don’t listen to you, then, come back and let us know.'
\z 
 
\ea \label{BP.165}
\textit{wāče her kesī nān dānī, wāče, ēšew be xātirū xwāy toū īđū ī kiɫfetū yāneyte dūr ginewe čī šārī’} \\ 
\gll wāč-e her kes-ī nān dā=nī wāč-e ēšew be xātir=ū xwā-ī to=ū īđ=ū ī kiɫfet=ū yāne-ī=t=e dūr ∅-gin-e=we čī šār-ī \\ 
 say.prs-\textsc{imp-}\textsc{2sg} just person\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} say.prs-\textsc{imp-}\textsc{2sg} tonight to sake\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}=and 3sg.prox\textsc{.dir}\textsc{.m}=and \textsc{dem.prox} child\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} far \textsc{imp-}fall.prs-\textsc{2sg}\textsc{=compl} why city\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Tell whoever gave you food, say, “Tonight, for the sake of God, you and your household get away from this town.'
\z 
 
\ea \label{BP.166}
\textit{lue bīrūnā”} \\ 
\gll ∅-lu-e bīrūnā \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} outside \\ 
\glt `Go away!'
\z 
 
\ea \label{BP.172}
\textit{hejījīyeke, fermāwo hejījīyeke, be qurbānit bim sūre gīyān yekē tirī lē de leqē teriš de wene’} \\ 
\gll hejījī-eke ∅-fermāw-o hejījī-eke be qurbān=it ∅-b-im sūr=e gīyān yek-ē tir=ī lē ∅-de-e leq-ē ter=iš ∅-de-e wene \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} to sacrifice\textsc{=\textsc{2sg}} \textsc{sbjv-}be.prs\textsc{-\textsc{1sg}} red=ez\textsc{.cmpd} dear.\textsc{voc} one\textsc{-indf} other\textsc{=3sg} at \textsc{imp-}give.prs-\textsc{2sg} kick\textsc{-indf} another\textsc{=3sg} \textsc{imp-}give.prs-\textsc{2sg} at \\ 
\glt `The [prisoner from] Hajij said [to Baba Khwada], ‘May I be your sacrifice, you red-skin, kick it once more! Kick it once more!”'
\z 
 
\ea \label{BP.186}
\textit{kuřēmā bo nāmēš binyeydē miħemeđ} \\ 
\gll kuř-ē=mā ∅-b-o nāmē=š bi-nye-ydē miħemeđ \\ 
 boy\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{ind-}be.prs\textsc{-3sg} name.f\textsc{=3sg} \textsc{imp-}put.prs-\textsc{2pl} \textsc{pn} \\ 
\glt `We will have a son. Name him Mohammad.'
\z 
 
\ea \label{ŠJ.20}
\textit{āđīč seʕbē pīyā kīyāno māčo lodē bārdēš!} \\ 
\gll āđ=īč seʕbe-ē pīyā ∅-kīyān-o m-āč-o ∅-lo-dē b-ār-dē=š \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} morning.f\textsc{-obl}.f man\textsc{.m} \textsc{ind-}send.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}bring.prs-\textsc{2pl}\textsc{=3sg} \\ 
\glt `The next day, he (Jafir Sān) sent some men [and] said, ‘Go [and] bring him!'
\z 
 
\ea \label{ŠJ.21}
\textit{lodē bārdēš peym šē ʕumerī} \\ 
\gll ∅-lo-dē b-ār-dē=š pey=m šē ʕumer-ī \\ 
 \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}bring.prs-\textsc{2pl}\textsc{=3sg} for\textsc{=\textsc{1sg}} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Go [and] bring Sheikh Umer to me.’'
\z 
 
\ea \label{ŠJ.36}
\textit{āmēnmē jafir sānī fermāwān lodē} \\ 
\gll āmē=nmē jafir sān-ī fermāwā=n ∅-lo-dē \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \\ 
\glt `We have come here. Jafir San said, ‘Go [and bring him].’'
\z 
 
\ea \label{ŠJ.37}
\textit{fermāwo hurbēzdē gumē bīdē gumē bīdē werū čemīmo!} \\ 
\gll ∅-fermāw-o hur- b-ēz-dē gum-ē ∅-b-īdē gum-ē ∅-b-īdē wer=ū čem-ī=m=o \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{pvb-}\textsc{imp-}rise.prs-\textsc{2pl} lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} out\_of\textsc{=ez.gen} eye\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=\textsc{post}} \\ 
\glt ` He said, ‘Get up and go away! Get out of my sight!'
\z 
 
\ea \label{ŠJ.38}
\textit{gumē bīdē!!} \\ 
\gll gum-ē ∅-b-īdē \\ 
 lost\textsc{-indf} \textsc{imp-}be.prs-\textsc{2pl} \\ 
\glt `Go away!'
\z 
 
\ea \label{ŠJ.52}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `‘Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.53}
\textit{ā esbī zīnī kere peym} \\ 
\gll ā esb-ī zīnī ∅-ker-e pey=m \\ 
 \textsc{dem.dist} horse\textsc{-obl}\textsc{.m} saddle \textsc{imp-}do.prs-\textsc{2sg} for\textsc{=\textsc{1sg}} \\ 
\glt `Saddle up the horse for me!'
\z 
 
\ea \label{ŠJ.62}
\textit{māčo berem pištū berew wiɫāxekā} \\ 
\gll m-āč-o ∅-ber-e=m pišt=ū bere=ū wiɫāx-ekā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}take.prs-\textsc{2sg}\textsc{=\textsc{1sg}} back\textsc{=ez.gen} door\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} \\ 
\glt `He (Jafir Sān) said, ‘Take me behind the door of the horses’ stable!’'
\z 
 
\ea \label{ŠJ.74}
\textit{fermāwo lodē lāš wāčdē kāriš be min nīyā} \\ 
\gll ∅-fermāw-o ∅-lo-dē lā=š wāč-dē kār=iš be min nīye=ā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} at\_the\_place\_of\textsc{=3sg} say.prs-\textsc{imp-}\textsc{2pl} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go to him [and] say, ‘His business is not with me.'
\z 
 
\ea \label{KŠ.40}
\textit{kināčēw pāđšāyč māčo, kuřū šuwāneyč bārdē ēme tāqetmā pey āney mē} \\ 
\gll kināčē=ū pāđšā-ī=č m-āč-o kuř=ū šuwāne-ī=č b-ār-dē ēme tāqet=mā pey āney m-ē \\ 
 girl\textsc{.dir}\textsc{.f}\textsc{=ez.gen} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{imp-}bring\textsc{.prs}-\textsc{2pl} \textsc{1pl} persistence\textsc{.m}\textsc{=\textsc{1pl}} to \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `The king’s daughter said, ‘Fetch the shepherd’s son. He will entertain us.\footnote{Lit. Our patience comes by him.}’'
\z 
 
\ea \label{KŠ.52}
\textit{āđīč māčo, wāčdē bē melāqātīm} \\ 
\gll āđ=īč m-āč-o wāč-dē b-ē melāqātī=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} say\textsc{.prs}-\textsc{imp-}\textsc{2pl} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} visitor\textsc{.m}\textsc{=\textsc{1sg}} \\ 
\glt `He (the king) said, ‘Tell him to come to visit me.’'
\z 
 
\ea \label{KŠ.72}
\textit{niwīsebēš, ey kesūkārū min, ey henū nīyenū min, heɫbetene ī kuřī biřiš dāwe, šime be wesīlēwe kuždēš, beynne berdēš!} \\ 
\gll niwīse b-ē=š ey kesūkār=ū min ey hen=ū nīyen=ū min heɫbete=ne ī kuř-ī biř=iš dā=we šime be wesīle-ēwe ∅-kuž-dē=š beyn=ne ∅-ber-dē=š \\ 
 write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} \textsc{2pl} by tool\textsc{.m}\textsc{-indf} \textsc{imp-}kill\textsc{.prs}-\textsc{2pl}\textsc{=3sg} between\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2pl}\textsc{=3sg} \\ 
\glt `He had written, ‘O my relatives, O my close family members, [when you read this letter] it is evident that this boy has arrived. Kill him with a thing, destroy him!'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.86}
\textit{wātiš dey lue āney fiɫānīū fiɫānī čiře ī qewmāme bā bēnēwe qāqezekey wenā} \\ 
\gll wāt=iš dey ∅-lu-e āney fiɫānī=ū fiɫānī ∅-čiř-e ī qewm-ā=m=e bā b-ē-nē=we qāqez-ekey ∅-wen-ā \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} such\_and\_such\_person=and such\_and\_such\_person \textsc{imp-}call\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{hort} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}read\textsc{.prs}\textsc{-3pl} \\ 
\glt `She said, ‘Then go, and summon such-and-such, our relatives, so that they come and read the letter.’'
\z 
 
\ea \label{KŠ.87}
\textit{lue čiře wextē miđyo wāteniš, luāydē ā mejlistā girđ nīštīdēre hurneēzdēwe her čā māreš biřdē} \\ 
\gll ∅-lu-e ∅-čiř-e wext-ē mi-đye-o wāte=n=iš luā-ydē ā mejlis=tā girđ nīšt-īdē=re hur-ne-ēz-dē=we her čā māre=š ∅-biř-dē \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}call\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} go\textsc{.pst}-\textsc{2pl} \textsc{dem.dist} gathering\textsc{.m}=\textsc{2pl} all sit\textsc{.pst}-\textsc{2pl}\textsc{=\textsc{povb}} \textsc{pvb-}\textsc{neg-}rise\textsc{.prs}-\textsc{2pl}\textsc{=compl} just there marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2pl} \\ 
\glt `They (the relatives) went [to the palace]. They learned that he (the king) had said [in the letter], ‘Go and sit in that gathering. Then do not rise [from your seats] until you have married her to him.’'
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{ŽH.27}
\textit{wātiš, tāte āne wāteniš, hekāyetē weš gīsnene, bā ī řē kučē keme kermē be hekāyetekeyt rāz biřmē} \\ 
\gll wāt=iš tāte āne wāte=n=iš hekāyet-ē weš-∅ ∅-gīsn-e=ne bā ī řē kuč-ē kem-e ∅-ker-mē be hekāyet-ekey=t rāz ∅-biř-mē \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} tale\textsc{.m}\textsc{-indf} well-m \textsc{imp-}start\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{hort} \textsc{dem.prox} road\textsc{.obl}\textsc{\textsc{.f}}.sg little\textsc{-indf} s\textsc{hort}\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} by tale\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{2sg}} tale\textsc{.m} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl} \\ 
\glt `She said, ‘Father, he had said, “Start [telling] a good tale, so that we [are entertained] and our way seems shorter. Let us finish your tale.”'
\z 
 
\ea \label{ŽH.38}
\textit{vātiš, dey tāte dey bā čīwēweš pey bere!’} \\ 
\gll vāt=iš dey tāte dey bā čīw-ēwe=š pey ∅-ber-e \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} father\textsc{.m} \textsc{disc.ptcl} \textsc{hort} thing\textsc{-indf}\textsc{=3sg} for \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `She said, ‘Father, get him something.’'
\z 
 
\ea \label{ŽH.40}
\textit{wātiš be wāɫe wičkilekēš ī māsī bere deš pā mēmān mizgī čenīw nānī’} \\ 
\gll wāt=iš be wāɫē=e wičkile-ekē=š ī mās-ī ∅-ber-e ∅-de-e=š p=ā mēmān mizgī čenī=ū nān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} by sister\textsc{\textsc{.f}}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} to\textsc{=dem}.dist guest\textsc{.m} mosque\textsc{.m} with\textsc{=ez}\textsc{.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said to her little daughter, ‘Take this yoghurt [and] together with [some] bread give it to the the guest in the mosque.’ '
\z 
 
\ea \label{ŽH.44}
\textit{māčo, jā heweřekey bāro} \\ 
\gll m-āč-o jā heweř-ekey b-ār-e=o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} afterwards pot\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg}\textsc{=compl} \\ 
\glt `She said, ‘Bring the pot back afterwards.’'
\z 
 
\ea \label{ŽH.45}
\textit{māčo, wāče be dēdēt māngeš keče bē’} \\ 
\gll m-āč-o ∅-wāč-e be dēdē=t mānge=š keč-e b-ē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}say\textsc{.prs-}\textsc{2sg} to older\_sister\textsc{\textsc{.f}}\textsc{=\textsc{2sg}} month\textsc{\textsc{.f}}\textsc{=3sg} crooked\textsc{-f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Hayas) said, ‘Tell your sister\footnote{Lit. maternal aunt.} that her moon was incomplete .’'\footnote{A cryptic message which in this context means that the yoghurt has been fingered by the girl’s daughter}
\z 
 
\ea \label{ŽH.47}
\textit{māčo, mēmānekey wāt, wāče dēdēt} \\ 
\gll m-āč-o mēmān-ekey wāt ∅-wāč-e dēdē=t \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} guest\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} say\textsc{.pst} \textsc{imp-}say\textsc{.prs-}\textsc{2sg} older\_sister\textsc{\textsc{.f}}\textsc{=\textsc{2sg}} \\ 
\glt `She said, ‘The guest said, “Tell your sister [that her moon is incomplete]”'
\z 
 
\ea \label{ŽH.50}
\textit{ā wextīyekey, māčo be tāteyš, dā ā mēmānī bāre pey yāneymā!’} \\ 
\gll ā wext-ī-ekey m-āč-o be tāte-ī=š dā ā mēmān-ī b-ār-e pey yāne-ī=mā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{hort} \textsc{dem.dist} guest\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} \\ 
\glt `The, she (the older sister) said to her father, ‘Bring that guest home!’'
\z 
 
\ea \label{ŽH.53}
\textit{māčo, ī serū řīšīme kučē tāše’} \\ 
\gll m-āč-o ī sere=ū řīš-ī=m=e kuč-ē ∅-tāš-e \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox} head\textsc{.m}=and beard\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} little\textsc{-indf} \textsc{imp-}shave\textsc{.prs}-\textsc{2sg} \\ 
\glt `He (Hayas) said [to the girl], ‘shave my face a bit.’'
\z 
 
\ea \label{ŽH.67}
\textit{āđīč māčo, čiřdēš peym’} \\ 
\gll āđ=īč m-āč-o ∅-čiř-dē=š pey=m \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}call\textsc{.prs}\textsc{-2pl}\textsc{=3sg} for\textsc{=1sg} \\ 
\glt `He (the Sultan) said, ‘Call him for me.’'
\z 
 
\ea \label{ŽH.69}
\textit{māčo, bereš pey kurdistānī’} \\ 
\gll m-āč-o ∅-ber-e=š pey kurdistān-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said [to Hayas], ‘Take this to Kurdistan. '
\z 
 
\ea \label{ŽH.79}
\textit{ā wextī yewāšē māčo, dey tehemuɫ kere bā hēɫeřūenit pey kerū, hāɫīt kerū'} \\ 
\gll ā wext-ī yewāšē m-āč-o dey tehemuɫ ∅-ker-e bā hēɫeřūen=it pey ∅-ker-ū hāɫī=t ∅-ker-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} then \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} patience \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{hort} fried\_egg\textsc{=\textsc{2sg}} for \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} understood\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then she said, ‘All right, wait a little bit, let me cook you fried eggs, [and] make you understand.'
\z 
 
\ea \label{ŽH.94}
\textit{māčo, werešā} \\ 
\gll m-āč-o ∅-wer-e=šā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}eat\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \\ 
\glt `[and] said [to the sultan], ‘Eat them [one by one]!’'
\z 
 
\ea \label{ŽH.118}
\textit{luewe pey yāney!’} \\ 
\gll ∅-lu-e=we pey yāne-ī \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} for house\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[Now] go back home!'
\z 
 
\ea \label{RE.7}
\textit{ʕerzū birāw wēm kerū ī kināčēte de pī kuřīme} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū ī kināčē=t=e ∅-de-e p=ī kuř-ī=m=e \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} brother\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} at=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \\ 
\glt `Let me tell you,\footnote{Lit. May a petition be at the service of my brother.} ‘Give your daughter to my son [in marriage]'
\z 
 
\ea \label{RE.11}
\textit{ētir kābrā fers kere zemānū ā ānē hītē pāɫē bīyēnēū čwār gezē pārče bīyenū} \\ 
\gll ētir kābrā fers ∅-ker-e zemān=ū ā ānē hīt-ē pāɫā-ē bīyē=nē=ū čwār gez-ē pārče bīye=n=ū \\ 
 \textsc{disc.ptcl} man guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} \textsc{dem.dist}\textsc{.3pl} pair\textsc{-indf} shoe\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and four ell\textsc{.m}\textsc{-indf} textile\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Then, let us say,\footnote{Lit. imagine.} at the time, [the wedding gift] consisted of a pair of shoes [and] four gaz\footnote{A unit of measure equivalent to 72 cm.} of textile.'
\z 
 
\ea \label{RE.19}
\textit{lue hārdekā bārewe} \\ 
\gll ∅-lu-e hārdī-ekā b-ār-e=we \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} flour\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.pl} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg}\textsc{=compl} \\ 
\glt `Go bring the flour.'
\z 
 
\ea \label{RE.20}
\textit{lue kārekem pey kere} \\ 
\gll ∅-lu-e kār-eke=m pey ∅-ker-e \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} task\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} for \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Go do [this] task for me.'
\z 
 
\ea \label{RE.29}
\textit{kābrā melo wāčo, ženīm de} \\ 
\gll kābrā me-l-o ∅-wāč-o ženī=m ∅-de-e \\ 
 man \textsc{neg-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} woman\textsc{.f}\textsc{=\textsc{1sg}} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `The fellow wouldn’t go [to his parents and] say, ‘Find a girl for me!’\footnote{Lit. Give me a woman.}'
\z 
 
\ea \label{RE.55}
\textit{māčo, dey lue deš} \\ 
\gll m-āč-o dey ∅-lu-e ∅-de-e=š \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} \\ 
\glt `He (the nobleman) would say [to the girl’s father], ‘Go [and] give her’'
\z 
 
\ea \label{ŽE.74}
\textit{luēnī lāw ī kuřī wāčēnī, dā lue wāče eđēš bizāne ī kināčēše miđom pene yā ne’} \\ 
\gll lu-ēn-ī lā=ū ī kuř-ī wāč-ēn-ī dā ∅-lu-e ∅-wāč-e eđē=š bi-zān-e ī kināčē=š=e mi-đe-o=m pene yā ne \\ 
 go\textsc{.prs}\textsc{-pstc}-\textsc{2sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.prs}\textsc{-pstc}-\textsc{2sg} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}say\textsc{.prs-}.\textsc{2sg} mother\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=3sg}\textsc{=dem} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{1sg}} to either no \\ 
\glt ` You would go to this boy [seated next to the narrator] and would say, ‘Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!’'
\z 
 
\ea \label{ŽE.78}
\textit{wāčē, dey lue duē yerē pīyē bēnē’} \\ 
\gll wāč-ē dey ∅-lu-e duē yerē pīyē b-ē-nē \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc.}\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} two three man\textsc{.dir}\textsc{.pl} \textsc{sbjv-}come\textsc{.prs}\textsc{-3pl} \\ 
\glt `She (the mother) would say, ‘Go! Let two, three men come.'
\z 
 
\ea \label{ŽE.82}
\textit{wāčē, dey lue yānew wēt nīšere tā duē māngē ter yerē māngē ter} \\ 
\gll wāč-ē dey ∅-lu-e yāne=ū wē=t ∅-nīš-e=re tā duē mānge-ē ter yerē mānge-ē ter \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc}.\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} until two month\textsc{.f}\textsc{-dir}\textsc{.pl} other three month\textsc{.f}\textsc{-dir}\textsc{.pl} other \\ 
\glt `They (the family of the girl) would say [to the boy], ‘Go back home [and] wait\footnote{Lit. sit.} for two, three months.'
\z 
 
