\chapter{All 220 examples of neg in these texts are found in the following sentences:}

\ea \label{ZB.9}
\textit{kuřēš bā, kuřekēš germīyānne nimēniš čene} \\ 
\gll kuř-ē=š b-ā kuř-ekē=š germīyān=ne ni-m-ē-nē=iš čene \\ 
 son\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.prs}\textsc{-3pl} son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.pl}\textsc{=3sg} \textsc{pn}\textsc{=\textsc{post}} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-3pl}\textsc{=3sg} with \\ 
\glt `He had sons. His sons did not accompany him [;they stayed in] Garmiyan. '
\z 
 
\ea \label{ZB.15}
\textit{māčo, ēšew wāro tekbīrmā mekirīyo} \\ 
\gll m-āč-o ēšew ∅-wār-o tekbīr=mā me-kir-īye-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} tonight \textsc{ind-}rain\textsc{-3sg} solution\textsc{=1pl} \textsc{neg-}do.pass-pass\textsc{-3sg} \\ 
\glt `He said, ‘It’s going to rain tonight; We can’t handle (it)\footnote{Lit. Our solution won’t be done.} '
\z 
 
\ea \label{ZB.33}
\textit{yeženekē biřē pīyāke māčo, melāmā pey’} \\ 
\gll yeženī-ekē biřē pīyā-eke m-āč-o me-l-ā=mā pey \\ 
 woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} a\_while man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}go\textsc{.prs}\textsc{-3pl}\textsc{=1pl} for \\ 
\glt `The woman (or) the man said several times, ‘It's not possible for us [to take the babies]. \footnote{they don't go for us}’'
\z 
 
\ea \label{ZB.35}
\textit{mekiryoyč ī kuřī xinīkno} \\ 
\gll me-kir-īye-o=yč ī kuř-ī ∅-xinīkn-o \\ 
 \textsc{neg-}do.pass-pass\textsc{-3sg}\textsc{=add} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv}-strangle\textsc{.prs}\textsc{-3sg} \\ 
\glt `We should not strangle the [baby] boys.\footnote{Lit. It can't be done that we strangle the baby boys.}'
\z 
 
\ea \label{ZQ.12}
\textit{kelēwe dīmāndīmē bē fire tūl bē, dāresān bē āwīš nebē fire nāřihet} \\ 
\gll kel-ēwe dīmāndīm-ē b-ē fire tūl b-ē dāresān b-ē āwī=š ne-b-ē fire nāřihet \\ 
 mountain\textsc{-indf} wide\textsc{-indf} \textsc{sbjv-}come.prs\textsc{.3sg} a\_lot long be.prs\textsc{-pstc}\textsc{.3sg} woodland\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} water\textsc{.dir}.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-pstc}\textsc{.3sg} a\_lot sad \\ 
\glt `It was a big mountain; it was very high; it was a forest; [it was] very uncomfortable.'
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{HB.4}
\textit{ħīčē nīyāw} \\ 
\gll ħīč-ē nīye=ā=ū \\ 
 nothing\textsc{-indf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `There is nothing [left]'
\z 
 
\ea \label{HB.17}
\textit{miđyo herzretū šēxī nefermāwā bilue yānew āney bāre’} \\ 
\gll mi-đye-o herzret=ū šēx-ī ne-fermāwā bi-lue yāne=ū āney b-ār-e \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}say\textsc{.pst} \textsc{sbjv-}go\textsc{.imp}\textsc{.\textsc{2sg}} house=and \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `He noticed that his highness the Sheikh hadn’t told [him], ‘Go to the house of that (i.e. such-and-such) person and bring [back the flour]’'
\z 
 
\ea \label{HB.18}
\textit{yā zeřiš neyā pene, lue čiko bisāne bāre!’} \\ 
\gll yā zeř=iš ne-dā pene ∅-lu-e či=ko bi-sān-e b-ār-e \\ 
 or money\textsc{.m}\textsc{=3sg} \textsc{neg-}give\textsc{.pst} to \textsc{imp-}go.\textsc{imp-}\textsc{2sg} from=where \textsc{imp-}buy\textsc{.prs}-\textsc{2sg} \textsc{imp-}bring\textsc{.prs}-\textsc{2sg} \\ 
\glt `Nor had he given him money, [saying], ‘Go [and] buy from a place\footnote{Lit. from where.} [and] bring [back the flour]’'
\z 
 
\ea \label{HB.20}
\textit{newātim qurbān min bilū yānew kē hārđī bārū} \\ 
\gll ne-wāt=im qurbān min bi-l-w yāne=ū kē hārđī b-ār-w \\ 
 \textsc{neg-}say\textsc{.pst}\textsc{=1sg} sir.\textsc{voc} 1sg \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} house=and who flour\textsc{\textsc{.f}}\textsc{.obl} \textsc{sbjv-}bring\textsc{.prs}\textsc{-1sg} \\ 
\glt `I didn’t say, ‘Sir, to whose house should I go [and] bring flour!'
\z 
 
\ea \label{HB.22}
\textit{zeřīčiš neyānā bilū yāgē bisānū} \\ 
\gll zeř=yč=iš ne-yā=nā bi-l-w yāgē-ē bi-sān-w \\ 
 money\textsc{.m}\textsc{=add}\textsc{=3sg} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} place\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}buy\textsc{.prs}\textsc{-1sg} \\ 
\glt `He hasn’t given me money either so that I could go to a place [and] buy [flour]'
\z 
 
\ea \label{HB.26}
\textit{demē gerekiš bē bilo nezānē bilo yānew kē} \\ 
\gll dem-ē gerek=iš b-ē bi-l-o ne-zān-ē bi-l-o yāne=ū kē \\ 
 once\textsc{-indf} necessary\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{neg-}know\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} house=and who \\ 
\glt `At another point, he wanted to go [ahead and ask for flour but] he didn’t know whose house to [go to].'
\z 
 
\ea \label{HB.27}
\textit{nezānē bilo yānew kē, gīriš wārd} \\ 
\gll ne-zān-ē bi-l-o yāne=ū kē gīr=iš wārd \\ 
 \textsc{neg-}know\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} house=and who hook\textsc{.m}\textsc{=3sg} eat\textsc{.pst} \\ 
\glt `He didn’t know whose house to go to. He was stuck,'
\z 
 
\ea \label{HB.29}
\textit{miđyo be ħīčē telklīfiš nīyā} \\ 
\gll mi-đye-o be ħīč-ē telklīf=iš nīye=ā \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} by nothing\textsc{-indf} option\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He realised that he had absolutely no options,'
\z 
 
\ea \label{HB.31}
\textit{hereke āmāwe zuwāb wātiš, xefetit nebo!} \\ 
\gll her-eke āmā=we zuwāb wāt=iš xefet=it ne-b-o \\ 
 donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} come\textsc{.pst}\textsc{.3sg}\textsc{=compl} reply\textsc{.m} say\textsc{.pst}\textsc{=3sg} sorrow\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The donkey started to speak [and] said, ‘Don’t worry!'
\z 
 
\ea \label{HB.42}
\textit{pekemā megino ētir} \\ 
\gll peke=mā me-gin-o ētir \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} no\_longer \\ 
\glt `we should not be worried anymore'
\z 
 
\ea \label{HB.43}
\textit{pekemā megino, to goš min dāre!’} \\ 
\gll peke=mā me-gin-o to goš min ∅-dār-e \\ 
 strenght\textsc{\textsc{.f}}\textsc{=1pl} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \textsc{2sg} ear\textsc{.m} 1sg \textsc{imp-}hold\textsc{.prs}-\textsc{2sg} \\ 
\glt `We should not be worried; listen to me!’'
\z 
 
\ea \label{HB.46}
\textit{her tā īse qisēš nekerdēnēū} \\ 
\gll her tā īse qisē=š ne-kerdē=nē=w \\ 
 both until now talk\textsc{\textsc{.f}}\textsc{=3sg} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `Until then, the donkey hadn’t talked. '
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{HB.70}
\textit{ēnnew herēm nezānān} \\ 
\gll ēnne=w her-ē=m ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Wasn’t I as much of a confidante as a donkey?' \footnote{Lit. didn’t I know as much as a donkey!}
\z 
 
\ea \label{HB.71}
\textit{ī herī zūwāniš zānā min hīčim nezānān!} \\ 
\gll ī her-ī zūwān=iš zānā min hīč=im ne-zānā=n \\ 
 \textsc{dem.prox} donkey\textsc{.m}\textsc{-obl}\textsc{.m} language\textsc{.m}\textsc{=3sg} know\textsc{.pst}\textsc{.ptcp}\textsc{.m} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The donkey knew the [Sheikh’s] language; I didn’t know a thing!'
\z 
 
\ea \label{HB.83}
\textit{ēnnew herē qābilīm nebē hāɫīm kerī!} \\ 
\gll ēnne=w her-ē qābilī=m ne-b-ē hāɫī=m ∅-ker-ī \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} deserved\textsc{=1sg} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} understood\textsc{=1sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-obl}\textsc{.m} \\ 
\glt `Wasn’t I as trustworthy as a donkey that you didn’t make me understand!'
\z 
 
\ea \label{HB.84}
\textit{ēnnew herē, īse her be herī wēš qisē kero min hīčim nezānān!’} \\ 
\gll ēnne=w her-ē īse her be her-ī wē=š qisē-ē ∅-ker-o min hīč=im ne-zānā=n \\ 
 as\_much\_as\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-indf} now both by donkey\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `As much as a donkey! now the donkey could speak even though it was a donkey, but I didn’t know a thing.'
\z 
 
\ea \label{HB.92}
\textit{čīm hāɫī nekerdenī!} \\ 
\gll čī=m hāɫī ne-kerde=nī \\ 
 why\textsc{=1sg} understood \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Why haven’t I made you understand?'
\z 
 
\ea \label{HB.93}
\textit{hāɫīm nekerdenī řās, ānīšā} \\ 
\gll hāɫī=m ne-kerde=nī řās ānīšā \\ 
 understood\textsc{=1sg} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} really 3pl.dem\textsc{.obl} \\ 
\glt `Haven’t I really made you understand? and so on.'
\z 


\ea \label{PM.5}
\textit{fermāndew merekuřekey, reʕīsekešā māčo, ēme bē ā šexsī metāwmē, e čē menmēwe zerer deymē čī mentēqey’} \\ 
\gll fermānde=ū merekuř-ekey reʕīs-eke=šā m-āč-o ēme bē ā šexs-ī me-tāw-mē e=čē ∅-men-mē=we zerer ∅-de-ymē č=ī mentēqe-ī \\ 
 leader\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} leader\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} in=here \textsc{sbjv-}stay\textsc{.prs}\textsc{-1pl}\textsc{=compl} harm\textsc{.m} \textsc{sbjv-}give\textsc{.prs}-\textsc{1pl} in=\textsc{dem.prox} region\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The head of the grasshoppers said, ‘Without [permission from] this person we cannot stay here and do harm to this region.'
\z 
 
\ea \label{PM.22}
\textit{nemendēwe e čē} \\ 
\gll ne-men-dē=we e=čē \\ 
 \textsc{neg-}remain\textsc{.prs}-\textsc{2pl}\textsc{=compl} in=here \\ 
\glt `Don’t stay here!'
\z 
 
\ea \label{PM.24}
\textit{herčē bo seʕbē hurmēzāwe miđyā tenhā geɫēwe nemenen} \\ 
\gll herčē ∅-b-o seʕbe-ē hur- mi-ēz-ā=we mi-đye-ā tenhā geɫā-ēwe ne-mene=n \\ 
 whatever \textsc{ind-}be\textsc{.prs}\textsc{-3sg} morning.f\textsc{-obl}.f \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} not\_even leaf\textsc{.m}\textsc{-indf} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The next day,\footnote{Lit. When it became the next day.} They (i.e., people from Sheikh Aladin’s entourage) got up [and] noticed that nothing was left.'
\z 
 
\ea \label{PM.26}
\textit{ħīčiš nemenen merekuř nāmēš nemenyēne} \\ 
\gll ħīč=iš ne-mene=n merekuř nāmē=š ne-menyē=ne \\ 
 nothing\textsc{=3sg} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} grasshopper name.f\textsc{=3sg} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}.f\textsc{=\textsc{post}} \\ 
\glt `There was nothing left in it. The grasshoppers had disappeared.\footnote{Lit. The grasshopper, its name had not remained.}'
\z 
 
\ea \label{PM.30}
\textit{īne bē ījāzew to īne nekiryān} \\ 
\gll īne b-ē ījāze=ū to īne ne-kir-yā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} permission\textsc{\textsc{=ez.gen}} \textsc{2sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{neg-}do.pass-past.pass\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This couldn’t have been done without your permission.'
\z 
 
\ea \label{PM.31}
\textit{tenhā hinū ton, geɫēwew kesī zereriš neginān je šārezūrne} \\ 
\gll tenhā hin=ū to=n geɫā-ēwe=ū kes-ī zerer=iš ne-ginā=n je šārezūr=ne \\ 
 not\_even ez.pron\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} leaf\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} person\textsc{.m}\textsc{-obl}\textsc{.m} harm\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} \\ 
\glt `It’s all yours [otherwise] no one’s leaves have been harmed in Sharazour. '
\z 
 
\ea \label{PM.33}
\textit{fermāwo, nāřihet nebī!} \\ 
\gll ∅-fermāw-o nāřihet ne-b-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘May you not be worried!''
\z 
 
\ea \label{PM.34}
\textit{nāřihet nebī āne meʕmūre bē terefū xwāywe āmābē} \\ 
\gll nāřihet ne-b-ī āne meʕmūre b-ē teref=ū xwā-ī=we āmā b-ē \\ 
 sad \textsc{neg-}be\textsc{.prs}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} agent be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} side\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Don’t worry! It was an agent [who] was sent  by God.'
\z 
 
\ea \label{PM.35}
\textit{eger yerē řoē šārezūrne menēwe geɫāyē sewz šārezūrne nāzē} \\ 
\gll eger yerē řo-ē šārezūr=ne men-ē=we geɫā-ē sewz šārezūr=ne ne-āz-ē \\ 
 if three day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{pn}\textsc{=\textsc{post}} remain\textsc{.prs}\textsc{-cond}\textsc{.3pl}\textsc{=compl} leaf\textsc{.m}\textsc{-indf} green \textsc{pn}\textsc{=\textsc{post}} \textsc{neg-}let\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `If the grasshoppers had stayed in Sharazour for three days, they wouldn’t have left behind a single green leaf in Sharazour.'
\z 
 
\ea \label{PM.36}
\textit{min pewkī ījāzem dān hinekew ēme wero, bilo zerer je kesī negino} \\ 
\gll min pewkī ījāze=m dā=n hin-eke=ū ēme ∅-wer-o bi-l-o zerer je kes-ī ne-gin-o \\ 
 \textsc{1sg} that\_is\_why permission\textsc{=\textsc{1sg}} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} ez.pron\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{sbjv-}eat\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} harm\textsc{.m} from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `I let them [eat our crop] so that they would eat our crop and go away, [but] not do harm to others.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.39}
\textit{hīčkāmmā mihtājē mebīmē pīney} \\ 
\gll hīčkām=mā mihtāj-ē me-b-īmē p=īney \\ 
 no\_one\textsc{=1pl} needy\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs} \textsc{1pl} by=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \\ 
\glt `None of us would be in need of these [products] '
\z 
 
\ea \label{PM.40}
\textit{bilowe xemit nebo} \\ 
\gll bi-l-o=we xem=it ne-b-o \\ 
 \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} sorrow\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Go back; don’t worry’'
\z 
 
\ea \label{DG.5}
\textit{eger newešīēmā bīyebo duktur nebīyenū} \\ 
\gll eger newešī-ē=mā bīye b-o duktur ne-bīye=n=ū \\ 
 if illness\textsc{.m}\textsc{-indf}\textsc{=1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} physician\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \\ 
\glt `If there might have been an illness, there was no doctor. '
\z 
 
\ea \label{DG.23}
\textit{ā wextīyekey, nimāzā bilo āweyānī} \\ 
\gll ā wext-ī-ekey ni-m-āz-ā bi-l-o āweyānī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} habitat\textsc{\textsc{.m}} \\ 
\glt `People wouldn’t let him go to the village.'
\z 
 
\ea \label{DG.35}
\textit{nimāzā bilmē āweyānī} \\ 
\gll ni-m-āz-ā bi-l-mē āweyānī \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} habitat\textsc{\textsc{.f}} \\ 
\glt `we are not allowed to go to the village.'
\z 
 
\ea \label{DG.39}
\textit{nimāzā bilū āweyānī} \\ 
\gll ni-m-āz-ā bi-l-ū āweyānī \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} habitat\textsc{\textsc{.f}} \\ 
\glt `I am not allowed to go to the village.'
\z 
 
\ea \label{DG.42}
\textit{nizīkū āweyānīyē nimāzā ginmē} \\ 
\gll nizīk=ū āweyānī-ē ni-m-āz-ā ∅-gin-mē \\ 
 close\textsc{=ez}\textsc{.gen} habitat\textsc{\textsc{.m}}\textsc{-indf}\textsc{\textsc{.f}} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-3pl} sbjv\textsc{-f}all\textsc{.prs}\textsc{-1pl} \\ 
\glt `They don’t let us anywhere near a village.'
\z 
 
\ea \label{DG.54}
\textit{birāš īnārē degāw činārī, metāwo bilo āweyānī} \\ 
\gll birā=š īnā-∅=rē degā=ū činār-ī me-tāw-o bi-l-o āweyānī \\ 
 brother\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} village\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} habitat\textsc{\textsc{.m}} \\ 
\glt `His brother was in Chinar village. He wasn’t allowed to go the village '
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.64}
\textit{ā wextīekey beɫē, min bilo ā memleketū wēšā, kiseɫīš čenē, ēme zāmdārē nekero ēme kečē nekero} \\ 
\gll ā wext-ī-ekey beɫē min bi-l-o ā memleket=ū wē=šā kiseɫī=š čenē ēme zāmdār-ē ne-ker-o ēme keč-ē ne-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yes \textsc{1sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=3pl} tortoise\textsc{\textsc{.f}}\textsc{=3sg} in \textsc{1pl} wounded\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `[The Sheikh said], ‘Yes!' '[The tortoise said] I [say] he [should] go to his region; there are tortoises there. He shall not injure us; he shall not make us crooked. '
\z 
 
\ea \label{DG.66}
\textit{ē.. dewāyekeš sāwe pore dā ēme kečē nebīmē} \\ 
\gll ē.. dewāī-eke=š ∅-sāw-e pore dā ēme keč-ē ne-b-īmē \\ 
 eh medication\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{imp-}rub\textsc{.prs}-\textsc{2sg} at \textsc{hort} \textsc{1pl} crooked\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs}-\textsc{1pl} \\ 
\glt `He shall rub the medicine on him; he shall not make us crooked.'
\z 
 
\ea \label{DG.69}
\textit{milowe māngēš pene mešo mēwe} \\ 
\gll mi-l-o=we mānge-ē=š pene me-š-o m-ē=we \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} month\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=3sg} to \textsc{neg-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `He went back. A month passed [and] he came back.'
\z 
 
\ea \label{DP.16}
\textit{tā pādšā nebīyen āđ pādšāš bīyen} \\ 
\gll tā pādšā ne-bīye=n āđ pādšā=š bīye=n \\ 
 until king\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{.m} king\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was its king until there were no kings.'
\z 
 
\ea \label{DP.19}
\textit{sibhān āġey neyān deso} \\ 
\gll sibhān āġe-ī ne-dā=n des=o \\ 
 \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} hand\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Sibhan Agha did not hand over the governorship to him,'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.37}
\textit{be xātirū īney qesem wero qesemiš negino} \\ 
\gll be xātir=ū īney qesem ∅-wer-o qesem=iš ne-gin-o \\ 
 by sake\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} oath\textsc{.m} \textsc{sbjv-}out\textsc{-3sg} oath\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wanted to swear an oath using the sparrow. He did not want to swear falsely on himself.'
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.40}
\textit{ewīč nimē māčo, nexeyr nimewnew mitmānim nīyā penet.} \\ 
\gll ew=yč ni-mi-ē mi-āč-o nexeyr ni-mi-e-wne=ū mitmān=im nīye=ā pene=t \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-1sg}=and trust\textsc{.m}\textsc{=1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} to\textsc{=\textsc{2sg}} \\ 
\glt ` Sibhan Agha\footnote{Lit. He.} did not go [to Jamsher Shah], saying, ‘No, I won’t come [to you]; I don’t trust you.’'
\z 
 
\ea \label{DP.46}
\textit{ā wextīyekey wextē mēnē ā wextī herčē qisē kerā, herčē behsē kerā herčē hesābē kerā, qēxāw hewrāmānī herčē bā ā wextī isɫāhšā mebo} \\ 
\gll ā wext-ī-ekey wext-ē mi-ē-nē ā wext-ī herčē qisē-ē ∅-ker-ā herčē behse-ē ∅-ker-ā herčē hesāb-ē ∅-ker-ā qēxā=w hewrāmān-ī herčē ∅-b-ā ā wext-ī isɫāh=šā me-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} however\_much talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much discussion\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much account\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} chieftain\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} however\_much \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} amendment\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then Sibhan Agha and his men came [to Jamsher Shah]. No matter how much the headmen of Hawraman talked [to both of them], discussed the issue, and tried to mediate, they (i.e., Jamsher Shah and Sibhan Agha) disagreed.'
\z 
 
\ea \label{DP.47}
\textit{isɫāhšā mebonew nemo čolekekē weriš miđo} \\ 
\gll isɫāh=šā me-b-one=ū ∅-nem-o čoleke-ekē wer=iš mi-đe-o \\ 
 amendment\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.prs}\textsc{-3sg}=and \textsc{ind-}lift\textsc{.prs}\textsc{-3sg} sparrow\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} out\textsc{=3sg} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `They disagreed. Jamsher Shah\footnote{Lit. He.} took the sparrow [and] let it go.'
\z 
 
\ea \label{ZP.24}
\textit{kesī be noʕē pene nezānān} \\ 
\gll kesī be noʕ-ē pene ne-zānā=n \\ 
 no\_one to kind/way\textsc{-indf} to \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `No one really appreciated him.'
\z 
 
\ea \label{ZP.26}
\textit{neweše gino herčīw duktirū ī lāū ew lā berāš duktir hīč duktirē ʕelēješ mekero} \\ 
\gll neweš-e ∅-gin-o herčīw duktir=ū ī lā=ū ew lā ∅-ber-ā=š duktir hīč duktir-ē ʕelēje=š me-ker-o \\ 
 ill\textsc{-f} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} how\_often physician\textsc{.m}=and dem.prox at\_the\_place\_of=and \textsc{dem.dist} at\_the\_place\_of \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3sg} physician\textsc{.m} nothing physician\textsc{.m}\textsc{-indf} treatment\textsc{.f}\textsc{=3sg} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt ` No matter how often she was taken to doctors here and there, no doctor could cure her.'
\z 
 
\ea \label{ZP.29}
\textit{keřū lāɫe bo, hīč duktirē ʕelēješ mekero} \\ 
\gll keře=ū lāɫ-e ∅-b-o hīč duktir-ē ʕelēje=š me-ker-o \\ 
 deaf\textsc{.f}=and mute\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} nothing physician\textsc{.m}\textsc{-indf} treatment\textsc{.f}\textsc{=3sg} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `She was deaf-mute. No doctor could cure her.'
\z 
 
\ea \label{ZP.40}
\textit{ētir xo ā zemāne eɫbet wesīle nebīyen yā māšīn yā čīwē} \\ 
\gll ētir xo ā zemān=e eɫbet wesīle ne-bīye=n yā māšīn yā čīw-ē \\ 
 \textsc{disc.ptcl} dis.ptc \textsc{dem.dist} time\textsc{.m}\textsc{=dem} of\_course tool\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} or car\textsc{.m} or thing\textsc{-indf} \\ 
\glt `—Well, back then, indeed, there was no means of transport, be it a car or anything [else.]'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.48}
\textit{ī kināčē nārīwe ētir pey min her ēnne weše bowe’} \\ 
\gll ī kināčē ne-ār-ī=we ētir pey min her ēnne weš-e b-o=we \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} any\_more for \textsc{1sg} emph as\_much\_as well\textsc{-f} be\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `Please do not bring this daughter [of mine] back to me as long as she is cured.'
\z 
 
\ea \label{ZP.54}
\textit{čāweɫ tā jāwe nelābē pāne pēwīyē tewenekē tewenē zile bē} \\ 
\gll čāweɫ tā jāwe ne-lā b-ē p=āne pēwīye-ē tewenī-ekē tewenī-ē zil-e b-ē \\ 
 in\_the\_past until road \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-indf} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\_visible\textsc{.prs}\textsc{-pstc}\textsc{.3sg} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} stone\textsc{.f}\textsc{-indf} big\textsc{-f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `In the past, when there was no road, the stone was visible from afar. It was a big stone.'
\z 
 
\ea \label{ZP.75}
\textit{mē ewā her gerūberū kināčekē lāšexeɫ mekero} \\ 
\gll m-ē ewā her gerūber=ū kināčē-ekē lāšexeɫ me-ker-o \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} there emph around\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} empty \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `They (the girl and her uncle) arrived there. He (the demon) would not leave the girl alone.'
\z 
 
\ea \label{ZP.76}
\textit{gerūberū kināčekē lāšexeɫ mekero} \\ 
\gll gerūber=ū kināčē-ekē lāšexeɫ me-ker-o \\ 
 around\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} empty \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the demon) would not leave the girl alone.'
\z 
 
\ea \label{ZP.86}
\textit{māčā xeɫkīč, ī pīyā šimeš pey persdē xo yānēw e čāneš nīyāw} \\ 
\gll m-āč-ā xeɫk=īč ī pīyā šime=š pey ∅-pers-dē xo yāne-ē=ū e=čāne=š nīye=ā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} people\textsc{.m}\textsc{=add} dem.prox man\textsc{.m} \textsc{2pl}\textsc{=3sg} for \textsc{ind-}ask\textsc{.prs}-\textsc{2pl} dis.ptc house\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} in=like\_that\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The people said, ‘This man whom you ask for, he does not have a proper house.'
\z 
 
\ea \label{ZP.87}
\textit{ženīū yāneš nīyārē īnīšā} \\ 
\gll ženī=ū yāne=š nīye=ā=rē īnīšā \\ 
 woman\textsc{.f}=and house\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He has no wife, nor a house, and so on.'
\z 
 
\ea \label{ZP.102}
\textit{īne pādšāzāyēnew derdū min meweronew} \\ 
\gll īne pādšāzāyē=ne=ū derd=ū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}=and illness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She is no use to me.'
\z 
 
\ea \label{ZP.103}
\textit{min metāwū īne nigebānīye kerū} \\ 
\gll min me-tāw-ū īne nigebānī-e ∅-ker-ū \\ 
 \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} guardian\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt ` I cannot take care of her.'
\z 
 
\ea \label{ZP.104}
\textit{īne pādšāzāyēne čenū min meweronew} \\ 
\gll īne pādšāzāyē=ne čenū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}} with\textsc{.ez.gen} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She will not eat with me [under the same roof].'
\z 
 
\ea \label{ZP.105}
\textit{min čīwēwim nīyārē lāyiqū ī pādšāzāye bo} \\ 
\gll min čīw-ēw=im nīye=ā=rē lāyiq=ū ī pādšāzāye ∅-b-o \\ 
 \textsc{1sg} thing\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} deserved\textsc{\textsc{=ez.gen}} dem.prox princess\textsc{.obl}\textsc{.f} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `I have nothing which would be up to the standards of this princess.'
\z 
 
\ea \label{ZP.108}
\textit{wātenīčiš, mišyo neberūšo”} \\ 
\gll wāte=n=īč=iš mišyo ne-ber-ū=š=o \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=add}\textsc{=3sg} aux \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `He has said, “I shall not take her back.”'
\z 
 
\ea \label{ZP.109}
\textit{ētir min metāwū berūšo, hergīz meberūšo’} \\ 
\gll ētir min me-tāw-ū ∅-ber-ū=š=o hergīz me-ber-ū=š=o \\ 
 \textsc{disc.ptcl} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} never \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `I cannot take her back. I will never take her back.’'
\z 
 
\ea \label{ZP.111}
\textit{gīr wero mezāno čēš kero} \\ 
\gll gīr ∅-wer-o me-zān-o čēš ∅-ker-o \\ 
 hook\textsc{.m} \textsc{ind-}eat\textsc{.prs}\textsc{-3sg} \textsc{neg-}know\textsc{.prs}\textsc{-3sg} what \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He was in a dilemma. He did not know what to do.'
\z 
 
\ea \label{ZP.112}
\textit{yāneš mebo menziɫeš mebo čīwēš mebo} \\ 
\gll yāne=š me-b-o menziɫe=š me-b-o čīw-ē=š me-b-o \\ 
 house\textsc{.m}\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} house\textsc{.f}\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} thing\textsc{-indf}\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He didn’t have a proper house; he didn’t have anything.'
\z 
 
\ea \label{ZP.116}
\textit{māčo, meberūšo mišyo to māreš kerīnew”} \\ 
\gll m-āč-o me-ber-ū=š=o mišyo to māre=š ∅-ker-īne=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} aux \textsc{2sg} marraige\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs} \textsc{2sg}=and \\ 
\glt `He says, “I will not take her back; you shall marry her.”'
\z 
 
\ea \label{ZP.117}
\textit{īđīč hīč minīč hīčim nīyārēw} \\ 
\gll īđ=īč hīč min=īč hīč=im nīye=ā=rē=ū \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} nothing \textsc{1sg}\textsc{=add} nothing\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `She [has] nothing, nor do I.'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.124}
\textit{ētir tāze be nelowe dimāwew} \\ 
\gll ētir tāze be ne-l-o=we dimā=we=ū \\ 
 \textsc{disc.ptcl} anyway to \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} afterwards\textsc{=\textsc{post}}=and \\ 
\glt `After all, she shall not go back.'
\z 
 
\ea \label{ZP.130}
\textit{hetā xwāy dinyāš weše kerdēne ī merāsēme kem mekero} \\ 
\gll hetā xwā-ī dinyā=š weš-e kerdē=ne ī merāsēm=e kem me-ker-o \\ 
 as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} small \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `So long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ŽP.27}
\textit{māčā, dey to metāwī -sey misefāw- kučēš yewē berī nehāɫū pīr šelīyārā šānīšā pižgīnīšā bā bizyā’} \\ 
\gll m-āč-ā dey to me-tāw-ī sey misefā=ū kuč-ē=š yewe-ē ∅-ber-ī nehāɫ=ū pīr šelīyār-ā ∅-šān-ī=šā ∅-pižgīn-ī=šā bā bi-zye-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{2sg} \textsc{neg-}can\textsc{.prs}-\textsc{2sg} \textsc{pn} \textsc{pn}=and little\textsc{-indf}\textsc{=3sg} barley\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \textsc{sbjv-}throw\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}scatter\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \textsc{hort} \textsc{sbjv-}grow\textsc{.prs}\textsc{-3pl} \\ 
\glt `They said, ‘You are not able, Say Mustafa, to take some barley seeds to the valley of Pir Shaliyar, [and] scatter them [on the field], [and] plant them, so that they grow.'
\z 
 
\ea \label{ŽP.28}
\textit{ētir xo āđ meselen hetīm bo, lāw māmoyšā bo řūeš mebo wāčo, yā metāwū yā tāwū} \\ 
\gll ētir xo āđ meselen hetīm ∅-b-o lā=ū māmo-ī=šā ∅-b-o řūe=š me-b-o ∅-wāč-o yā me-tāw-ū yā ∅-tāw-ū \\ 
 \textsc{disc.ptcl} dis.ptc 3sg\textsc{.dir}\textsc{.m} for\_example orphan \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} face\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-3sg} either \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} either \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Well, he was an orphan. He was at his uncle’s. He was ashamed to say, ‘I cannot, or I can.’'
\z 
 
\ea \label{ŽP.49}
\textit{īneš netāwān kero} \\ 
\gll īne=š ne-tāwā=n ∅-ker-o \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He cannot do these [tasks].’'
\z 
 
\ea \label{ŽP.62}
\textit{wēregā pīr šelīyār mēwe sey misefā wextē mēwe, ʕerzū ženekēš kero māčo, bā ēšew īmti̩hānēš kermē lāw wēmāwe newsinmēš} \\ 
\gll wēregā pīr šelīyār m-ē=we sey misefā wext-ē m-ē=we ʕerz=ū ženī-ekē=š ∅-ker-o m-āč-o bā ēšew īmti̩hān-ē=š ∅-ker-mē lā=ū wē=mā=we ne-ūsin-mē=š \\ 
 evening \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{pn} \textsc{pn} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} tonight test\textsc{-indf}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=1pl}\textsc{=\textsc{post}} \textsc{neg-}put\_to\_sleep\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \\ 
\glt `Pir Shaliyar returned in the evening; when Say Mustafa returned, he (Hama Yoso) said to his wife, ‘Let’s test him tonight; let’s not put him to sleep next to us.’'
\z 
 
\ea \label{ŽP.72}
\textit{ʕerzū heremēš kero, fermāwo be heremēš weɫāhī ētir ēme metāwmē, sey misefā metāwmē hīč kārēš pene kermē īlā kīyānmēš bilo weno, bilo kīyānmēš weno} \\ 
\gll ʕerz=ū hereme-ē=š ∅-ker-o ∅-fermāw-o be hereme-ē=š weɫāhī ētir ēme me-tāw-mē sey misefā me-tāw-mē hīč kār-ē=š pene ∅-ker-mē īlā ∅-kīyān-mē=š bi-l-o ∅-wen-o bi-l-o ∅-kīyān-mē=š ∅-wen-o \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} indeed \textsc{disc.ptcl} \textsc{1pl} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} \textsc{pn} \textsc{pn} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} except \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}read\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}read\textsc{-3sg} \\ 
\glt `He (Pir Hama Yoso) said to his wife, ‘Indeed, we can no longer be helpful to Say Mustafa,\footnote{Lit. we can do nothing to him.} except that we send him to study; that we send him, to go [and] study.'
\z 
 
\ea \label{ŽP.75}
\textit{ētir mebo ēme kārēš pene kermē} \\ 
\gll ētir me-b-o ēme kār-ē=š pene ∅-ker-mē \\ 
 no\_longer \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{1pl} task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `We can no longer do anything for him.'
\z 
 
\ea \label{ŽP.98}
\textit{ētir herčī mēnē ā zemāne wesīlew māšīnū čīwī mebo} \\ 
\gll ētir herčī m-ē-nē ā zemān=e wesīle=ū māšīn=ū čīw-ī me-b-o \\ 
 \textsc{disc.ptcl} what\_ever \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} tool\textsc{.m}=and car\textsc{.m}=and thing\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well, in that period, there were no means of transportation, cars, etc.'
\z 
 
\ea \label{ŽP.99}
\textit{wesīlew swārī mebo be peyē mēnē řāre} \\ 
\gll wesīle=ū swārī me-b-o be peyē m-ē-nē řā=re \\ 
 tool\textsc{.m}=and driving \textsc{neg-}be\textsc{.prs}\textsc{-3sg} by foot\textsc{.obl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} road\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `There were no means of transportation. They (people) would set off on the road on foot.'
\z 
 
\ea \label{ŽP.104}
\textit{zeřekey bere min, qey mekero} \\ 
\gll zeř-ekey ∅-ber-e min qey me-ker-o \\ 
 money\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{1sg} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Take the money. I will be fine.’'
\z 
 
\ea \label{ŽP.105}
\textit{āđīč fermāwo, ne qey mekero bā lāw towe bo min xerjīm gerek nīyā milūwe’} \\ 
\gll āđ=īč ∅-fermāw-o ne qey me-ker-o bā lā=ū to=we ∅-b-o min xerjī=m gerek nīye=ā mi-l-ū=we \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{hort} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} consumption\textsc{=\textsc{1sg}} necessary \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) said, ‘No, it is all right! Let it be with you. I do not need money. I will go back.’'
\z 
 
\ea \label{ŽP.111}
\textit{zeřeš gerek mebo} \\ 
\gll zeř-e=š gerek me-b-o \\ 
 money\textsc{.m}\textsc{-def}\textsc{=3sg} necessary \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He did not need the money.'
\z 
 
\ea \label{ŽP.134}
\textit{xo āđ mezāno hezretū xosī kīyāseniš} \\ 
\gll xo āđ me-zān-o hezret=ū xos-ī kīyāse=n=iš \\ 
 dis.ptc 3sg\textsc{.dir}\textsc{.m} \textsc{neg-}know\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \\ 
\glt `Well, he (Pir Shaliyar) did not realise that His Highness Khos had sent him (the man)'
\z 
 
\ea \label{ŽP.144}
\textit{welī emāneketē xās xeɫk bāwiřiš pene mekero} \\ 
\gll welī emāneketē xās xeɫk bāwiř=iš pene me-ker-o \\ 
 but but fine people\textsc{.m} belief\textsc{=3sg} in \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `But people still did not believe in him [as a leader].'
\z 
 
\ea \label{ŽP.145}
\textit{miqdārē bāwiřiš pene kerā miqdārē mekerā} \\ 
\gll miqdārē bāwiř=iš pene ∅-ker-ā miqdārē me-ker-ā \\ 
 a\_little belief\textsc{=3sg} in \textsc{ind-}do\textsc{.prs}\textsc{-3pl} a\_little \textsc{neg-}do\textsc{.prs}\textsc{-3pl} \\ 
\glt `Some believed in him, some did not.'
\z 
 
\ea \label{ŽP.147}
\textit{ħīč dukturē metāwo wešeš kerowe} \\ 
\gll ħīč duktur-ē me-tāw-o weš-e=š ∅-ker-o=we \\ 
 nothing physician\textsc{.m}\textsc{-indf} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} well\textsc{-f}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \\ 
\glt `No doctor could cure her.'
\z 
 
\ea \label{ŽP.150}
\textit{qisē metāwo kero nitqiš bīnīyo} \\ 
\gll qisē-ē me-tāw-o ∅-ker-o nitq=iš bīnīye-o \\ 
 talk\textsc{.f}\textsc{-dir}\textsc{.pl} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} speech\textsc{=3sg} close\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `She was not able to speak. She had lost the ability to speak.\footnote{Lit. her speech was closed.}'
\z 
 
\ea \label{ŽP.152}
\textit{hīč dukturē metāwo wešeš kerowe} \\ 
\gll hīč duktur-ē me-tāw-o weš-e=š ∅-ker-o=we \\ 
 nothing physician\textsc{.m}\textsc{-indf} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} well\textsc{-f}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \\ 
\glt `No doctor could cure her.'
\z 
 
\ea \label{ŽP.161}
\textit{metāwū serū textīwe wezūrew bilū ginū gelū kināčēw be peyū luāy řāreū degā be degā} \\ 
\gll me-tāw-ū ser=ū text-ī=we wez-ū=re=ū bi-l-ū ∅-gin-ū gel=ū kināčē=ū be pe-ī=ū luāy řā=re=ū degā be degā \\ 
 \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} on\textsc{\textsc{=ez.gen}} throne\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} climb\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{povb}}=and \textsc{sbjv-}go\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-\textsc{1sg}} with\textsc{\textsc{=ez.gen}} daughter\textsc{.dir}\textsc{.pl}=and by foot\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} go.inf road\textsc{.f}\textsc{=\textsc{post}}=and village\textsc{.obl}\textsc{.f} by village\textsc{.obl}\textsc{.f} \\ 
\glt `I cannot leave the throne [and] go [and] accompany the girl, [while] walking on foot on the roads [that go] village by village.'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.189}
\textit{hezretū sey misefāy čā miđyo xo kes ter mewīnoš} \\ 
\gll hezret=ū sey misefā-ī čā mi-đye-o xo kes ter me-wīn-o=š \\ 
 his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} there \textsc{ind-}look\textsc{.prs}\textsc{-3sg} dis.ptc person\textsc{.m} other \textsc{neg-}see\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \\ 
\glt `His Highness Say Mustafa watched [them] from above. No one else could see her.'
\z 
 
\ea \label{ŽP.190}
\textit{āđīšā mewēno} \\ 
\gll āđīšā me-wēn-o \\ 
 3pl\textsc{.obl} \textsc{neg-}see\textsc{.prs}\textsc{-3sg} \\ 
\glt `He did not see them.'
\z 
 
\ea \label{ŽP.194}
\textit{duē yerē keřetē xuřoš wene, bēheyā bo megēɫowe} \\ 
\gll duē yerē keřet-ē ∅-xuř-o=š wene bē- heyā ∅-b-o me-gēɫ-o=we \\ 
 two three time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}shout\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{povb}without decency \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{neg-}turn\textsc{-3sg}\textsc{=compl} \\ 
\glt `He (Pir Shaliyar) shouted at him (the demon) two or three times. [But] he (the demon) was indecent; he did not return.'
\z 
 
\ea \label{ŽP.208}
\textit{xo min meberūšo} \\ 
\gll xo min me-ber-ū=š=o \\ 
 dis.ptc \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `I will not take her back.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.212}
\textit{berešo qey mekeronew} \\ 
\gll ∅-ber-e=š=o qey me-ker-one=ū \\ 
 \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Take her back. It is all right!'
\z 
 
\ea \label{ŽP.213}
\textit{īnē be min mewero kināčē pādšāy,} \\ 
\gll īnē be min me-wer-o kināčē pādšā-ī \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} with \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg} girl\textsc{.dir}\textsc{.f} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `She will not eat with me [under the same roof], the king’s daughter.'
\z 
 
\ea \label{ŽP.219}
\textit{xo ā wextī wēčiš yāneš her mebo} \\ 
\gll xo ā wext-ī wē=č=iš yāne=š her me-b-o \\ 
 dis.ptc \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=add}\textsc{=3sg} house\textsc{.m}\textsc{=3sg} emph \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He had no house at that time.'
\z 
 
\ea \label{ŽP.225}
\textit{her lāloeke māčo, mišyo ī kināčē min meberūšo māreš biřūnew’} \\ 
\gll her lālo-eke m-āč-o mišyo ī kināčē min me-ber-ū=š=o māre=š ∅-biř-ūne=ū \\ 
 emph maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} aux dem.prox girl\textsc{.dir}\textsc{.f} \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} marraige\textsc{=3sg} \textsc{ind-}cut\textsc{.prs}\textsc{-\textsc{1sg}}=and \\ 
\glt ` The uncle kept saying, ‘I shall not take this girl back. I shall marry her [to Pir Shaliyar].’'
\z 
 
\ea \label{ŽP.226}
\textit{āđīč māčo, min metāwū ī kināčē wey kerū īne nānmā} \\ 
\gll āđ=īč m-āč-o min me-tāw-ū ī kināčē wey ∅-ker-ū īne nān=m=ā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox girl\textsc{.dir}\textsc{.f} raising \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `People … he (Pir Shaliyar) said, ‘I cannot take care of this girl. This is my food.’'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.234}
\textit{qey mekero nimāymē peket gino} \\ 
\gll qey me-ker-o ni-m-āz-mē peke=t ∅-gin-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-1pl} strenght\textsc{.f}\textsc{=\textsc{2sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `It is all right. We will not let you worry about it.'
\z 
 
\ea \label{ŽP.235}
\textit{nimāymē peket gino} \\ 
\gll ni-m-āz-mē peke=t ∅-gin-o \\ 
 \textsc{neg-}\textsc{ind-}let\textsc{.prs}\textsc{-1pl} strenght\textsc{.f}\textsc{=\textsc{2sg}} \textsc{sbjv-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `We will not let you worry about it.'
\z 
 
\ea \label{ŽP.242}
\textit{qey mekero, mādām desūr e pēsen pādšā řāzīe nīā ī ženī bilowe ī kināčē bilowe qey mekero!’} \\ 
\gll qey me-ker-o mādām desūr e=pēse=n pādšā řāzīe nīye=ā ī ženī bi-l-o=we ī kināčē bi-l-o=we qey me-ker-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} as\_long\_as task/order\textsc{.m} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} satisfied \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} dem.prox woman\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `There is no problem. Now that things have developed this way, [and] the king is not happy for the girl to return, it’s all right! [She can stay here]’'
\z 
 
\ea \label{ŽP.243}
\textit{pīr šelīyār fermāwo, tehāā weɫāhī īne šime desūrtā dāw wāttā čūnke āđē be diɫtā wāt řāstā wāt hetā xwāy dinyāš weše kerdēne her menowe kemīč mekero her zīyā kero} \\ 
\gll pīr šelīyār ∅-fermāw-o tehāā weɫāhī īne šime desūr=tā dā=ū wāt=tā čūnke āđē be diɫ=tā wāt řās=tā wāt hetā xwā-ī dinyā=š weš-e kerdē=ne her ∅-men-o=we kem=īč me-ker-o her zīyā ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} task/order\textsc{.m}=\textsc{2pl} give\textsc{.pst}=and say\textsc{.pst}=\textsc{2pl} since 3pl\textsc{.dir} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} truth=\textsc{2pl} say\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} small\textsc{=add} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar said, ‘Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] continues to exist; it will not be diminished; it will grow.'
\z 
 
\ea \label{ŽP.245}
\textit{ī merāsēme her menowe hīčkeyč kem mekero her zīyāy kero īlāy žīyāyne bo zīyāyiš pene kero} \\ 
\gll ī merāsēm=e her ∅-men-o=we hīčke=yč kem me-ker-o her zīyāy ∅-ker-o īlāy žīyāy=ne ∅-b-o zīyāy=iš pene ∅-ker-o \\ 
 dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} nothing\textsc{=add} little \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increase \textsc{ind-}do\textsc{.prs}\textsc{-3sg} for\_sure increased\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} increased\textsc{=3sg} to \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This ceremony will be repeated, it won’t diminish. It will continue to grow. It will grow [into bigger ceremonies]'
\z 
 
\ea \label{ŽP.266}
\textit{ēti melowe āđīč melowe} \\ 
\gll ēti me-l-o=we āđ=īč me-l-o=we \\ 
 \textsc{disc.ptcl} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `After all, he didn’t go back [to Egypt]. He did not go back either.'
\z 
 
\ea \label{BP.13}
\textit{to hošit nīyā, tāteyt miqdārēš hošā—} \\ 
\gll to hoš=it nīye=ā tāte-ī=t miqdārē=š hoš=ā \\ 
 \textsc{2sg} memory\textsc{.m}\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} a\_little\textsc{=3sg} memory\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `—you are not aware [of that period], [but] your father remembers it a bit—'
\z 
 
\ea \label{BP.32}
\textit{melāwe, dēr milāwe, melāwe} \\ 
\gll me-l-ā=we dēr mi-l-ā=we me-l-ā=we \\ 
 \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} late \textsc{ind-}go.prs\textsc{-3pl}\textsc{=compl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `[This time] they (the tax collectors) didn’t go back. They went back late. They didn’t go back.'
\z 
 
\ea \label{BP.34}
\textit{tumez mekere, pewkī melāwe, epēgene gerekšān genekārī kerā} \\ 
\gll tumez me-ker-e pewkī me-l-ā=we epēgene gerek=šā=n genekārī ∅-ker-ā \\ 
 surprisingly \textsc{neg-}do.prs-\textsc{2sg} that\_is\_why \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} here necessary\textsc{=3pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} debauchery\textsc{.m} \textsc{sbjv-}do.prs\textsc{-3pl} \\ 
\glt `Ostensibly, they hadn’t gone back because they wished to engage in debauchery. '
\z 
 
\ea \label{BP.36}
\textit{je terefū ženū menū čīwe genekārī kerā nelāwe} \\ 
\gll je teref=ū ženī=ū men=ū čīw=e genekārī ∅-ker-ā ne-l-ā=we \\ 
 from side\textsc{=ez.gen} woman.f=and \textsc{rdp}=and thing\textsc{.m}\textsc{=dem} debauchery\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They wished to engage in debauchery with women and so on. They didn’t go back.'
\z 
 
\ea \label{BP.47}
\textit{māčā, ī īnē āmēnē melāwe’} \\ 
\gll m-āč-ā ī īnē āmē=nē me-l-ā=we \\ 
 \textsc{ind-}say.prs\textsc{-3pl} \textsc{dem.prox} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They (people) said, ‘These (tax collectors) have come here; they have not gone back yet.’'
\z 
 
\ea \label{BP.48}
\textit{ētir zimsānē melāwe biřew bārekey berāwe} \\ 
\gll ētir zimsān-ē me-l-ā=we biře=ū bāre-ekey ∅-ber-ā=we \\ 
 \textsc{disc.ptcl} winter\textsc{.m}\textsc{-indf} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} portion\textsc{.m}=and profit\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}take.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They didn’t go back one winter to take back the taxes.'
\z 
 
\ea \label{BP.58}
\textit{zū biřew bāre bisānā bēnēwe, īse čī nāmēwe} \\ 
\gll zū biře=ū bāre bi-sān-ā b-ē-nē=we īse čī ne-āmā-ē=we \\ 
 quick portion\textsc{.m}=and profit\textsc{.m} \textsc{sbjv-}buy.prs\textsc{-3pl} \textsc{sbjv-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} now why \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \\ 
\glt `They were supposed to collect taxes soon. Why haven’t they come back?'
\z 
 
\ea \label{BP.59}
\textit{čī nāmēnē?} \\ 
\gll čī ne-āmē=nē \\ 
 why \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Why haven’t they come back?'
\z 
 
\ea \label{BP.60}
\textit{ēmeyč māčmē, ī meʕmūrē šime kīyāsēnē bā biřew bāremā dēw nāmēwe bārāšo} \\ 
\gll ēme=yč m-āč-mē ī meʕmūr-ē šime kīyāsē=nē ∅-b-ā biře=ū bāre=mā dā-ē=ū ne-āmā-ē=we b-ār-ā=š=o \\ 
 \textsc{1pl}\textsc{=add} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{ind-}be.prs\textsc{-3pl} portion\textsc{.m}=and profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}bring.prs\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `We will say, ‘We paid tax to the officers whom you had sent, [however,] they didn’t return to bring it back.'
\z 
 
\ea \label{BP.71}
\textit{neferē yāneš pī ser degāwe bo, tirēštew kārdū čīwiš mebo} \\ 
\gll nefer-ē yāne=š p=ī ser degā=we ∅-b-o tirēšte=ū kārdī=ū čīw=iš me-b-o \\ 
 person\textsc{.m}\textsc{-indf} house\textsc{.m}\textsc{=3sg} at=\textsc{dem.prox} on village.f\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3sg} ax\textsc{.m}=and knife.f\textsc{=ez.gen} thing\textsc{.m}\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg} \\ 
\glt `The house of one of the locals was at the top of the village. He had neither axe, nor knife, nor anything else. '
\z 
 
\ea \label{BP.76}
\textit{miđyā īne be mišārī sereš biřyen nāmā nāmā nīyā} \\ 
\gll mi-đye-ā īne be mišār-ī sere=š biřye=n ne-āmā-∅ ne-āmā-∅ nīye=ā \\ 
 \textsc{ind-}look.prs\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} to hand\_saw\textsc{.m}\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=3sg} cut\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They noticed that the one who had decapitated his guests with a hand saw hadn’t come to the mosque. He hadn’t arrived. He wasn’t there.'
\z 
 
\ea \label{BP.77}
\textit{fermāwo, dā lodē bizāndē fiɫāne kes čī nāmā?} \\ 
\gll ∅-fermāw-o dā ∅-lo-dē bi-zān-dē fiɫān=e kes čī ne-āmā-∅ \\ 
 \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{sbjv-}know.prs-\textsc{2pl} such\_and\_such=ez\textsc{.cmpd} person\textsc{.m} why \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `He (Baba Khwada) said, ‘Go [and] see\footnote{Lit. know} why such-and-such person hasn’t come [to the mosque]?'
\z 
 
\ea \label{BP.89}
\textit{ētir ēme xo diruē meyeymē} \\ 
\gll ētir ēme xo diruē me-de-ymē \\ 
 \textsc{disc.ptcl} \textsc{1pl} dis.ptc lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `We are not going to lie.'
\z 
 
\ea \label{BP.93}
\textit{xerīkū genekārī bēnē, nāmēwe} \\ 
\gll xerīk=ū genekārī b-ēn-ē ne-āmā-ē=we \\ 
 busy\textsc{=ez.gen} debauchery\textsc{.m} be.prs\textsc{-pstc}\textsc{-3pl} \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \\ 
\glt `[However] They fell into debauchery; They wouldn’t go back [to Iraq]'
\z 
 
\ea \label{BP.94}
\textit{nāmēwew xerīkū genekārīw kārī nāšerʕīšā kerdenū} \\ 
\gll ne-āmā-ē=we=ū xerīk=ū genekārī=ū kār=ī nāšerʕī=šā kerde=n=ū \\ 
 \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-indf}\textsc{=compl}=and busy\textsc{=ez.gen} debauchery\textsc{.m}=and task\textsc{.m}=ez.\textsc{attr} unlawful\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They didn’t go back [to Iraq]. They commited unlawful acts.'
\z 
 
\ea \label{BP.95}
\textit{ēmeyč muxteserū kelāmī, xo ēme diruē meyeymē} \\ 
\gll ēme=yč muxteser=ū kelām-ī xo ēme diruē me-de-ymē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} dis.ptc \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `As for us, to cut a long story short, we are not lying.'
\z 
 
\ea \label{BP.117}
\textit{ētir āđē wātenšā, ēme diruē meyeymē} \\ 
\gll ētir āđē wāte=n=šā ēme diruē me-de-ymē \\ 
 \textsc{disc.ptcl} 3pl\textsc{.dir} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `They said, ‘We are not going to lie.'
\z 
 
\ea \label{BP.118}
\textit{řūē memenmērew je kesī metersmē īlā je xwāyne bo} \\ 
\gll řūē me-men-mē=re=ū je kes-ī me-ters-mē īlā je xwā-ī=ne ∅-b-o \\ 
 face\textsc{.obl} \textsc{neg-}remain.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}=and from person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\_afraid.prs\textsc{-\textsc{1pl}} except from God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{sbjv-}be.prs\textsc{-3sg} \\ 
\glt `We are not going to implore [the king]. We are not afraid of anyone except God.'
\z 
 
\ea \label{BP.123}
\textit{ħemey wirdīkɫe ā wextī, pēse īsey māčā heɫāy neyāwān sinʕe zīnāniš nekermē yā čīw yā čāne} \\ 
\gll ħeme=ī wirdīkɫe ā wext-ī pēse īse-ī m-āč-ā heɫāy ne-yāwā=n sinʕe zīnān=iš ne-ker-mē yā čīw yā č=ā=ne \\ 
 \textsc{pn}=ez.\textsc{attr} small \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} like now\textsc{-obl}\textsc{.m} \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} age.f prison\textsc{.m}\textsc{=3sg} \textsc{neg-}do.prs\textsc{-\textsc{1pl}} either thing\textsc{.m} either in=\textsc{dem.dist}\textsc{=\textsc{post}} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) will say, ‘Hama Wirdikla has not reached adulthood; we won’t imprison him, and so on.’'
\z 
 
\ea \label{BP.125}
\textit{āđē nāmēw xerīkū genekārī bēnēw} \\ 
\gll āđē ne-āmā-ē=ū xerīk=ū genekārī b-ēn-ē=ū \\ 
 3pl\textsc{.dir} \textsc{neg-}come\textsc{.pst}\textsc{-3pl}=and busy\textsc{=ez.gen} debauchery\textsc{.m} be.prs\textsc{-pstc}\textsc{-3pl}=and \\ 
\glt `[however] they didn’t come back [to you]. They engaged in debauchery.'
\z 
 
\ea \label{BP.132}
\textit{ħemey wirdīkɫeyč pēse īse māčā, heɫāy neyāwān bilūqū sinʕēš} \\ 
\gll ħeme=ī wirdīkɫe=yč pēse īse m-āč-ā heɫāy ne-yāwā=n bilūq=ū sinʕe-ē=š \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} like now \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} maturity\textsc{.m}\textsc{=ez.gen} age.f\textsc{-obl}.f\textsc{=3sg} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) said, ‘he hasn’t hit puberty yet.'
\z 
 
\ea \label{BP.133}
\textit{īneyšā mekerdē zīnān} \\ 
\gll īney=šā me-ker-dē zīnān \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=3pl} \textsc{neg-}do.prs-\textsc{2pl} prison\textsc{.m} \\ 
\glt `Do not put this one of them in prison.'
\z 
 
\ea \label{BP.134}
\textit{qānūn meweroš pene} \\ 
\gll qānūn me-wer-o=š pene \\ 
 law \textsc{neg-}eat.prs\textsc{-3sg}\textsc{=3sg} to \\ 
\glt `The law did not apply to him.'
\z 
 
\ea \label{BP.138}
\textit{hemey wirdīkɫey mekerā zīnān ewīšā kerā zīnān} \\ 
\gll heme=ī wirdīkɫe-ī me-ker-ā zīnān ewīšā ∅-ker-ā zīnān \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{-obl}\textsc{.m} \textsc{neg-}do.prs\textsc{-3pl} prison\textsc{.m} 3pl\textsc{.obl} \textsc{ind-}do.prs\textsc{-3pl} prison\textsc{.m} \\ 
\glt `They didn’t put Hama Wirdikla in jail. They put them (Baba Khwada and Hama Ghaybi) in jail.'
\z 
 
\ea \label{BP.139}
\textit{berā kerāšāne zīnāneke, wextē miđyā mezā nū če noʕē bīyen ħejījīyāyčšā girten xuɫkū hejījī} \\ 
\gll ∅-ber-ā ∅-ker-ā=šā=ne zīnān-eke wext-ē mi-đye-ā me-zān-ū če noʕ-ē bīye=n ħejījī-yā=yčšā girte=n xuɫk=ū hejīj-ī \\ 
 \textsc{ind-}take.prs\textsc{-3pl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=3pl}\textsc{=\textsc{povb}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-\textsc{1sg}} what kind/way\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} people\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They took [them] and put them in the jail, they realised that —they didn’t know for what reason— [but] a person from Hajij had also been arrested.'
\z 
 
\ea \label{BP.148}
\textit{mekerāš zīnānī gēɫo šārekene nānšā pey peyā kero} \\ 
\gll me-ker-ā=š zīnān-ī ∅-gēɫ-o šār-eke=ne nān=šā pey peyā ∅-ker-o \\ 
 \textsc{neg-}do.prs\textsc{-3pl}\textsc{=3sg} prison\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}wander.prs\textsc{-3sg} city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} bread\textsc{.m}\textsc{=3pl} for visible \textsc{ind-}do.prs\textsc{-3sg} \\ 
\glt `They didn’t put him in jail. He would wander in town and find food for them.'
\z 
 
\ea \label{BP.157}
\textit{ēme mezānmē tekbīrmā čēš bo īnīšā} \\ 
\gll ēme me-zān-mē tekbīr=mā čēš ∅-b-o īnīšā \\ 
 \textsc{1pl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} advice\textsc{=\textsc{1pl}} what \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `We don’t know what we should do, and so on.'
\z 
 
\ea \label{BP.164}
\textit{gošiš nedārāyne ētir jā bowe xeber bārewe} \\ 
\gll goš=iš ne-dārā-ī=ne ētir jā b-o=we xeber b-ār-e=we \\ 
 ear\textsc{.m}\textsc{=3sg} \textsc{neg-}hold\textsc{.pst}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{disc.ptcl} afterwards \textsc{imp-}come.prs\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{imp-}bring.prs-\textsc{2sg}\textsc{=compl} \\ 
\glt `If they don’t listen to you, then, come back and let us know.'
\z 
 
\ea \label{BP.190}
\textit{miđyā her hāyš mebowe nīyā dērāw,} \\ 
\gll mi-đye-ā her hāy=š me-b-o=we nīye=ā dēr=ā=ū \\ 
 \textsc{ind-}look.prs\textsc{-3pl} just awareness\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} late\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `They noticed that he didn’t wake up, he wasn’t there [and] it was getting late.'
\z 
 
\ea \label{BP.193}
\textit{deng mekero nīyā} \\ 
\gll deng me-ker-o nīye=ā \\ 
 noise\textsc{.m} \textsc{neg-}do.prs\textsc{-3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He didn’t make a sound. He wasn’t there.'
\z 
 
\ea \label{BP.194}
\textit{wextē ʕebē berzā kerāwe, miđyā kes nīyā čērišne} \\ 
\gll wext-ē ʕebā-ē berz=ā ∅-ker-ā=we mi-đye-ā kes nīye=ā čēr=iš=ne \\ 
 time\textsc{.m}\textsc{-indf} robe.f\textsc{-obl}.f high\textsc{=compl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=compl} \textsc{ind-}look.prs\textsc{-3pl} person\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} under\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `When they lifted the robe, they saw that no one was under it.'
\z 
 
\ea \label{BP.211}
\textit{werū īney diruē neđāw werū īney je lāw xwāywe negināwe diro ħīčē suālē diruēšā neđēne} \\ 
\gll wer=ū īney diruē ne-đā=ū wer=ū īney je lā=ū xwā-ī=we ne-gin-ā=we diro ħīč-ē suāl-ē diruē=šā ne-đē=ne \\ 
 out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} lie.f \textsc{neg-}give\textsc{.pst}=and out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} from at\_the\_place\_of\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \textsc{neg-}fall.prs\textsc{-3pl}\textsc{=compl} lie nothing\textsc{-indf} question\textsc{-indf} lie.f\textsc{=3pl} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `In order not to lie, in order not to be liars in the presence of God, they didn’t tell a lie.'
\z 
 
\ea \label{ŠJ.6}
\textit{seʕbē tā wēregā pāse desiš pey, des luā bīyen nemān} \\ 
\gll seʕbe-ē tā wēregā pāse des=iš pey des luā bīye=n ne-mān-∅ \\ 
 morning.f\textsc{-obl}.f until evening like hand\textsc{.m}\textsc{=3sg} to hand\textsc{.m} go\textsc{.pst}\textsc{.3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}remain\textsc{.pst}\textsc{.3sg}\textsc{-3sg}\textsc{.m} \\ 
\glt `He would disappear early in the morning until evening.'
\z 
 
\ea \label{ŠJ.25}
\textit{be xwā her nīšā čī nāmāwew,} \\ 
\gll be xwā her ∅-nīš-ā čī ne-āmā=we=ū \\ 
 to God\textsc{.m} just \textsc{ind-}sit.prs\textsc{-3pl} why \textsc{neg-}come\textsc{.pst}\textsc{.3sg}\textsc{=compl}=and \\ 
\glt `Well, they waited a while, [but] he didn’t come back.’'
\z 
 
\ea \label{ŠJ.26}
\textit{čī nāmāwew?} \\ 
\gll čī ne-āmā=we=ū \\ 
 why \textsc{neg-}come\textsc{.pst}\textsc{.3sg}\textsc{=compl}=and \\ 
\glt `[The men from Luhon said], ‘Why hasn’t he come back?'
\z 
 
\ea \label{ŠJ.39}
\textit{čenū gum bīyey gumē bīyā serew řoē gumē bā mezānā bilāwe pey nođšey} \\ 
\gll čenū gum bīyey gum-ē bī-ā sere=ū řo-ē gum-ē ∅-b-ā me-zān-ā bi-l-ā=we pey nođše-ī \\ 
 with\textsc{.ez.gen} lost be\textsc{.inf} lost\textsc{-indf} be\textsc{.pst}\textsc{-3pl} \textsc{clf}\textsc{=ez.gen} day\textsc{.m}\textsc{-dir}\textsc{.pl} lost\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-3pl}\textsc{=compl} to \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[They got so scared that] they were lost. They were lost for one day. They didn’t know the way to go back to Nodsha.'
\z 
 
\ea \label{ŠJ.45}
\textit{weɫē bīyēnmē āro duē řoē ētir mezānmē koge milmē} \\ 
\gll weɫē bīyē=nmē āro duē řo-ē ētir me-zān-mē koge mi-l-mē \\ 
 front be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} today two day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{disc.ptcl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} where \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \\ 
\glt `We have spread out for two days; we didn’t know where we were going.’'
\z 
 
\ea \label{ŠJ.74}
\textit{fermāwo lodē lāš wāčdē kāriš be min nīyā} \\ 
\gll ∅-fermāw-o ∅-lo-dē lā=š wāč-dē kār=iš be min nīye=ā \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} at\_the\_place\_of\textsc{=3sg} say.prs-\textsc{imp-}\textsc{2pl} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘Go to him [and] say, ‘His business is not with me.'
\z 
 
\ea \label{ŠJ.76}
\textit{hīč kārēš be min nīyā} \\ 
\gll hīč kār-ē=š be min nīye=ā \\ 
 nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He has no business with me.’'
\z 
 
\ea \label{ŠJ.77}
\textit{milā māčā qurbān šē ʕusmānī fermāwā kāriš be min nīyā} \\ 
\gll mi-l-ā m-āč-ā qurbān šē ʕusmān-ī fermāwā kār=iš be min nīye=ā \\ 
 \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] said [to Jafir Sān], ‘Sir, the sheikh said, “He has no business with me.'
\z 
 
\ea \label{ŠJ.82}
\textit{ʕeybekeyčiš memeno} \\ 
\gll ʕeyb-eke=yč=iš me-men-o \\ 
 fault\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=add}\textsc{=3sg} \textsc{neg-}remain.prs\textsc{-3sg} \\ 
\glt `The defect had disappeared.'
\z 
 
\ea \label{ŠJ.86}
\textit{fermāwo qurbān hermāne be desū min nīyenew} \\ 
\gll ∅-fermāw-o qurbān hermāne be des=ū min nīye=ne=ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} work\textsc{.f} to hand\textsc{.m}\textsc{=ez.gen} \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `He (the sheikh) said, ‘Sir, the task [of ruling] is not in my hands;'
\z 
 
\ea \label{ŠJ.87}
\textit{be desū to nīyenew} \\ 
\gll be des=ū to nīye=ne=ū \\ 
 to hand\textsc{.m}\textsc{=ez.gen} \textsc{2sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `It is not in your hands either.'
\z 
 
\ea \label{ŠJ.90}
\textit{xo řīk megino ī kāre pey girđ kesī bo} \\ 
\gll xo řīk me-gin-o ī kār=e pey girđ kes-ī ∅-b-o \\ 
 dis.ptc arranged \textsc{neg-}fall.prs\textsc{-3sg} \textsc{dem.prox} task\textsc{.m}\textsc{=dem} to all person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} \\ 
\glt `These (tasks) cannot be done by everyone.'
\z 
 
\ea \label{ŠJ.91}
\textit{eger pīyāgerī bo, eger šēxyetī bo, eger nāmē bo řīk megino mentēqēwene de neferē bā de nānēšā bā yo mezāno řū we kē kero} \\ 
\gll eger pīyāgerī ∅-b-o eger šēxyetī ∅-b-o eger nāmē ∅-b-o řīk me-gin-o mentēqe-ēwe=ne de nefer-ē ∅-b-ā de nān-ē=šā ∅-b-ā yo me-zān-o řū=we kē ∅-ker-o \\ 
 if gentlemanhood \textsc{sbjv-}be.prs\textsc{-3sg} if sheikhhood \textsc{sbjv-}be.prs\textsc{-3sg} if name.f \textsc{sbjv-}be.prs\textsc{-3sg} arranged \textsc{neg-}fall.prs\textsc{-3sg} region\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} ten person\textsc{.m}\textsc{-indf} \textsc{sbjv-}be.prs\textsc{-3pl} ten bread\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3pl} \textsc{sbjv-}be.prs\textsc{-3pl} one\textsc{.m} \textsc{neg-}know.prs\textsc{-3sg} face=to who \textsc{sbjv-}do.prs\textsc{-3sg} \\ 
\glt `Say this job is about being a gentleman, being a sheikh, being well known, [still] it couldn’t be that if in a region there are ten people [worthy of ruling] who have power over ten families, one doesn’t know whom to plead with [when there is a problem]'
\z 
 
\ea \label{ŠJ.95}
\textit{pēšīmān bowew je hāɫū wēšne memeno} \\ 
\gll pēšīmān b-o=we=ū je hāɫ=ū wē=š=ne me-men-o \\ 
 regretful be.prs\textsc{-3sg}\textsc{=\textsc{post}}=and from state\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{neg-}remain.prs\textsc{-3sg} \\ 
\glt `He (Jafir Sān) regretted it. He changed his ways.'
\z 
 
\ea \label{ŠJ.101}
\textit{dey āro be kē wāčī derđwārde bo je dinyāne kes nemenen} \\ 
\gll dey āro be kē ∅-wāč-ī derđwārde ∅-b-o je dinyā=ne kes ne-mene=n \\ 
 \textsc{disc.ptcl} today to who \textsc{sbjv-}say.prs.\textsc{sbjv-}\textsc{2sg} qualified\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} from world\textsc{.dir}.f\textsc{=\textsc{post}} person\textsc{.m} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Nowadays, who can you refer to in the world who would be qualified? No one has remained.'
\z 
 
\ea \label{ŠJ.104}
\textit{īse šūnīšā ne[menen]} \\ 
\gll īse šūnī=šā ne-mene=n \\ 
 now track.f\textsc{=3pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Now, nothing has remained of them.'
\z 
 
\ea \label{ŠJ.105}
\textit{tenhā yo kuřēšā meneniš duktir Mādēh dinyāne řāš mebowe} \\ 
\gll tenhā yo kuř-ē=šā mene=n=iš duktir Mādēh dinyā=ne řā=š me-b-o=we \\ 
 not\_even one\textsc{.m} son/boy\textsc{.m}\textsc{-indf}\textsc{=3pl} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} physician\textsc{.m} \textsc{pn} world\textsc{.dir}.f\textsc{=\textsc{post}} road.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `Only a son is left of the chieftains, named Dr Madeh, who is not allowed here.'
\z 
 
\ea \label{ŠJ.108}
\textit{dinyāne řāš mebowe} \\ 
\gll dinyā=ne řā=š me-b-o=we \\ 
 world\textsc{.dir}.f\textsc{=\textsc{post}} road.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `[That’s why] he is not allowed here.'
\z 
 
\ea \label{KŠ.17}
\textit{āđīč hukmiš dā wātiš, ženē leme peře mebo bo} \\ 
\gll āđ=īč hukm=iš dā wāt=iš ženī-ē leme peř-e me-b-o ∅-b-o \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} rule\textsc{.m}\textsc{=3sg} give\textsc{.pst} say\textsc{.pst}\textsc{=3sg} woman\textsc{.f}\textsc{-indf} stomach\textsc{.f} full\textsc{-f} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He (the king) issued an order [and] said, ‘No pregnant women should stay alive!’\footnote{Lit. No pregnant women should be.}'
\z 
 
\ea \label{KŠ.20}
\textit{ženū šuwāney pēkyā nekewte pene} \\ 
\gll ženī=ū šuwāne-ī pēkyā ne-kewt-e pene \\ 
 woman\textsc{.f}=and shepherd\textsc{.m}\textsc{-obl}\textsc{.m} smite\textsc{.pst}.pass\textsc{.3sg} \textsc{neg-}fall\textsc{.pst}\textsc{-3sg}\textsc{.f} with \\ 
\glt `They aimed an arrow at the shepherd’s wife, but the shooting did not take [i.e. hit] her.'
\z 
 
\ea \label{KŠ.21}
\textit{nekewte pene, luā kenāre} \\ 
\gll ne-kewt-e pene luā kenāre \\ 
 \textsc{neg-}fall\textsc{.pst}\textsc{-3sg}\textsc{.f} to go\textsc{.pst}\textsc{.3sg} corner \\ 
\glt `The shooting did not hit her. It (the arrow) went into a corner.'
\z 
 
\ea \label{KŠ.35}
\textit{eđāš mebonew tāteš šuwāne bo} \\ 
\gll eđā=š me-b-one=ū tāte=š šuwāne ∅-b-o \\ 
 mother\textsc{.f}\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg}=and father\textsc{.m}\textsc{=3sg} shepherd\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He did not have a mother. His father was a shepherd.'
\z 
 
\ea \label{KŠ.69}
\textit{māčo. zēɫeš nāmā hāyš kerowe} \\ 
\gll m-āč-o zēɫe=š ne-āmā hāy=š ∅-ker-o=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} heart\textsc{.m}\textsc{=3sg} \textsc{neg-}come\textsc{.pst}\textsc{.3sg} awareness\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `She said … ‘She was not inclined\footnote{Lit. Her heart did not come} to wake him up.’'
\z 
 
\ea \label{KŠ.76}
\textit{īne wēš xeberiš nīyāw sewāđiš nīyā} \\ 
\gll īne wē=š xeber=iš nīye=ā=ū sewāđ=iš nīye=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{reflx}\textsc{=3sg} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and literacy\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He wasn’t aware [of the death plot] that \_\_ he was illiterate\_\_'
\z 
 
\ea \label{KŠ.87}
\textit{lue čiře wextē miđyo wāteniš, luāydē ā mejlistā girđ nīštīdēre hurneēzdēwe her čā māreš biřdē} \\ 
\gll ∅-lu-e ∅-čiř-e wext-ē mi-đye-o wāte=n=iš luā-ydē ā mejlis=tā girđ nīšt-īdē=re hur-ne-ēz-dē=we her čā māre=š ∅-biř-dē \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}call\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} go\textsc{.pst}-\textsc{2pl} \textsc{dem.dist} gathering\textsc{.m}=\textsc{2pl} all sit\textsc{.pst}-\textsc{2pl}\textsc{=\textsc{povb}} \textsc{pvb-}\textsc{neg-}rise\textsc{.prs}-\textsc{2pl}\textsc{=compl} just there marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2pl} \\ 
\glt `They (the relatives) went [to the palace]. They learned that he (the king) had said [in the letter], ‘Go and sit in that gathering. Then do not rise [from your seats] until you have married her to him.’'
\z 
 
\ea \label{KŠ.91}
\textit{ētir ew čone āyā xeber zāno yā mezāno,} \\ 
\gll ētir ew č=o=ne āyā xeber ∅-zān-o yā me-zān-o \\ 
 \textsc{disc.ptcl} \textsc{dem.dist} in=\textsc{dem.dist}\textsc{=\textsc{post}} whether news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} or \textsc{neg-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `[It is not clear] whether he (the king) was aware of the wedding or not. '
\z 
 
\ea \label{KŠ.105}
\textit{kārū to řāsā, kes metāwoš tefre biđo} \\ 
\gll kār=ū to řās=ā kes me-tāw-o=š tefre bi-đe-o \\ 
 task\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{2sg} true\textsc{=cop}\textsc{.3sg}\textsc{.m} person\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-3sg}\textsc{=3sg} evasion \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `Your deed is right! No one can avoid destiny.'
\z 
 
\ea \label{ŽH.3}
\textit{wāčē řoē je řoā qāɫešā bo, serū mezān ū čēšī} \\ 
\gll wāč-ē řo-ē je řo-ā qāɫe=šā ∅-b-o sere=ū me-zān-ū čēš-ī \\ 
 say\textsc{.prs-}pstc\textsc{.3sg} day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} quarrel\textsc{\textsc{.f}}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} head\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what\textsc{-obl}\textsc{.m} \\ 
\glt `According to her (my mother),\footnote{Lit. She would say.}  Once they (Sultan Mahmud and Hayas) got into a quarrel with each other, about something I don’t know.'
\z 
 
\ea \label{ŽH.13}
\textit{ewīč māčo, min pilekānī mezānū binyewnew mezānū čēš māčī!’} \\ 
\gll ew=yč m-āč-o min pilekānī me-zān-ū bi-nye-ūne=ū me-zān-ū čēš m-āč-ī \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 1sg stair\textsc{\textsc{.f}} \textsc{neg-}know\textsc{.prs}\textsc{-1sg} \textsc{sbjv-}put\textsc{.prs}\textsc{-1sg}=and \textsc{neg-}know\textsc{.prs}\textsc{-1sg} what \textsc{ind-}say\textsc{.prs}-\textsc{2sg} \\ 
\glt `He said, ‘I can’t set up a ladder and I don’t understand\footnote{Lit. know.}  what you are saying!’'
\z 
 
\ea \label{ŽH.18}
\textit{māčo, eger to nebī, mizgī’} \\ 
\gll m-āč-o eger to ne-b-ī mizgī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} if \textsc{2sg} \textsc{neg-}be\textsc{.prs}-\textsc{2sg} mosque\textsc{.m} \\ 
\glt `He (Hayas) said, ‘If it is not you[r house, I will go to] the mosque.’'
\z 
 
\ea \label{ŽH.26}
\textit{māčo, nezānām} \\ 
\gll m-āč-o ne-zānā=m \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{neg-}know\textsc{.pst}\textsc{=1sg} \\ 
\glt `He (the man) said, ‘I didn’t understand [his point].’'
\z 
 
\ea \label{ŽH.28}
\textit{to nezānānt!’} \\ 
\gll to ne-zānā=n=t \\ 
 \textsc{2sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `[But] you didn’t understand [his point].’'
\z 
 
\ea \label{ŽH.30}
\textit{wātiš, qisēmā nekerdē āmāymē nizīkū dege,} \\ 
\gll wāt=iš qisē=mā ne-kerd-ē āmā-īmē nizīk=ū dege \\ 
 say\textsc{.pst}\textsc{=3sg} talk\textsc{.dir}\textsc{.pl}\textsc{=1pl} \textsc{neg-}do\textsc{.pst}\textsc{-3pl} come\textsc{.pst} \textsc{1pl} close\textsc{=ez}\textsc{.gen} village\textsc{.obl}\textsc{\textsc{.f}} \\ 
\glt `He said, ‘We didn’t talk [until] we got close to the village.'
\z 
 
\ea \label{ŽH.32}
\textit{wātiš, eger to nebī mizgī’} \\ 
\gll wāt=iš eger to ne-b-ī mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} if \textsc{2sg} \textsc{neg-}be\textsc{.prs}-\textsc{2sg} mosque\textsc{.m} \\ 
\glt `He (Hayas) said, ‘If it is not you[r house, I will go to] the mosque.’'
\z 
 
\ea \label{ŽH.33}
\textit{wātiš, tāte, āne wāteniš, eger to xuɫkem mekerī, meberīm yāne milū mizgī’} \\ 
\gll wāt=iš tāte āne wāte=n=iš eger to xuɫke=m me-ker-ī me-ber-ī=m yāne mi-l-ū mizgī \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} if \textsc{2sg} invitation\textsc{=1sg} \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \textsc{neg-}take\textsc{.prs}-\textsc{2sg}\textsc{=1sg} house\textsc{.m}\textsc{.dir.sg} \textsc{ind-}go\textsc{.prs}\textsc{-1sg} mosque\textsc{.m} \\ 
\glt `She said, ‘Father, he meant,\footnote{Lit. said.} “If you do not invite me and won’t take me home, I will go to the mosque.”'
\z 
 
\ea \label{ŽH.48}
\textit{māčo, je māseket keč nekerđebo řāře wāɫē!’} \\ 
\gll m-āč-o je mās-eke=t keč ne-kerđe b-o řā=ře wāɫē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} from yoghurt\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} crooked \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-3sg} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} sister\textsc{\textsc{.f}} \\ 
\glt `She (the older sister) had said [to her younger sister], ‘Sister, could the yoghurt be reduced’'
\z 
 
\ea \label{ŽH.70}
\textit{pīney bilo šewē neēwe} \\ 
\gll p=īney bi-l-o šewe-ē ne-ē=we \\ 
 for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{neg-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `The reason [was] so that he (Hayas) would go and not come back for the night. '
\z 
 
\ea \label{ŽH.71}
\textit{řākē dūrew inne dūre bo, pīney šewē biř nekero bēwe} \\ 
\gll řā-ekē dūr=ew inne dūr-e ∅-b-o p=īney šewe-ē biř ne-ker-o b-ē=we \\ 
 road\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} far\textsc{=\textsc{post}} so\_much far\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} piece \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `The distance was so long that he (Hayas) wouldn’t be able to return for the night.'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.108}
\textit{mēwe, melowe yāne} \\ 
\gll m-ē=we me-l-o=we yāne \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m}\textsc{.dir}.sg \\ 
\glt `He returned [but] he didn’t go back home.'
\z 
 
\ea \label{ŽH.109}
\textit{melowe yāne, duē řoē pā degāne bo} \\ 
\gll me-l-o=we yāne duē řo-ē p=ā degā=ne ∅-b-o \\ 
 \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m} two day\textsc{.m}\textsc{-dir}\textsc{.pl} at\textsc{=dem}.dist village\textsc{.obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `He didn’t go back home. He was [wandering] in the village for two days. '
\z 
 
\ea \label{ŽH.110}
\textit{šūnū duē řoeyre, kināčekē kīyāno lāw suɫtānī māčo, ēme pīyākemā kīyāstiš meʕmūrīyet nīyā, nāmāno’} \\ 
\gll šūn=ū duē řo-e-ī=re kināčē-ekē ∅-kīyān-o lā=ū suɫtān-ī m-āč-o ēme pīyā-eke=mā kīyāst-∅=iš meʕmūrīyet nīye=ā ne-āmā=n=o \\ 
 after\textsc{=ez}\textsc{.gen} two\textsc{-def}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} girl\textsc{.dir}\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{=ez}\textsc{.gen} sul\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} send\textsc{.pst}\textsc{-3sg}\textsc{.m}\textsc{=3sg} mission\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `After the second day, the girl (his wife) sent (someone) to the Sultan to tell [him], ‘He (the sultan) sent my husband on a mission, [however] he is not around; He hasn’t come back [yet].’'
\z 
 
\ea \label{ŽH.112}
\textit{māčo, āđīč čiřoš māčo, čī nelānīwe?’} \\ 
\gll m-āč-o āđ=īč ∅-čiř-o=š m-āč-o čī ne-lā=nī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} \\ 
\glt `He (the sultan) called him (Hayas) [and] said, ‘Why don’t you go back (to your house)?’'
\z 
 
\ea \label{RE.23}
\textit{qisew mise nebīyen} \\ 
\gll qise=w mise ne-bīye=n \\ 
 talk\textsc{=ez.gen} \textsc{rdp} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was not a quarrel [between them]'
\z 
 
\ea \label{RE.29}
\textit{kābrā melo wāčo, ženīm de} \\ 
\gll kābrā me-l-o ∅-wāč-o ženī=m ∅-de-e \\ 
 man \textsc{neg-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} woman\textsc{.f}\textsc{=\textsc{1sg}} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} \\ 
\glt `The fellow wouldn’t go [to his parents and] say, ‘Find a girl for me!’\footnote{Lit. Give me a woman.}'
\z 
 
\ea \label{RE.34}
\textit{jūwābēš nīyā xeberiš nīyā tā tāte} \\ 
\gll jūwāb-ē=š nīye=ā xeber=iš nīye=ā tā tāte \\ 
 response\textsc{.m}\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} until father\textsc{.m} \\ 
\glt `[Then] there is no news until [the girl] says, ‘Father!’'
\z 
 
\ea \label{RE.38}
\textit{međey mehr kerū,} \\ 
\gll me-đe-ī mehr ∅-ker-ū \\ 
 \textsc{neg-}give\textsc{.prs}-\textsc{2sg} mariage\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `If you don’t accept, I will marry [him]'
\z 
 
\ea \label{RE.50}
\textit{to qibūɫ kerī yā mekerī} \\ 
\gll to qibūɫ ∅-ker-ī yā me-ker-ī \\ 
 \textsc{2sg} accepted \textsc{ind-}do\textsc{.prs}-\textsc{2sg} either \textsc{neg-}do\textsc{.prs}-\textsc{2sg} \\ 
\glt `Will you consent to it or not?'
\z 
 
\ea \label{RE.64}
\textit{bē ā begā ēme netāwānmā ħīč kermē} \\ 
\gll bē ā beg-ā ēme ne-tāwā=n=mā ħīč ∅-ker-mē \\ 
 without \textsc{dem.dist} Beg\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{1pl} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1pl}} nothing \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1pl}} \\ 
\glt `Without them, we were not able to do anything.'
\z 
 
\ea \label{ŽE.5}
\textit{ħīč nebīyen} \\ 
\gll ħīč ne-bīye=n \\ 
 nothing \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was nothing [here].'
\z 
 
\ea \label{ŽE.6}
\textit{āwīš nebīyēne} \\ 
\gll āwī=š ne-bīyē=ne \\ 
 water\textsc{.dir}\textsc{.f}\textsc{=3sg} \textsc{neg-}be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `It (the village) had no water.'
\z 
 
\ea \label{ŽE.7}
\textit{hīč āwū čīwiš nebīyen} \\ 
\gll hīč āwī=ū čīw=iš ne-bīye=n \\ 
 nothing water\textsc{.f}\textsc{\textsc{=ez.gen}} thing\textsc{.m}\textsc{=3sg} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` It had no water or anything.'
\z 
 
\ea \label{ŽE.21}
\textit{āwē her kunene her netemāmīyēne} \\ 
\gll āwē her kune=ne her ne-temāmīyē=ne \\ 
 water\textsc{.obl}\textsc{.f} emph churn\textsc{=\textsc{post}} emph \textsc{neg-}finish\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `The water [was] in the clay pot. It would not finish [soon]'
\z 
 
\ea \label{ŽE.23}
\textit{čēgeyč āwī her nebīyēne bīnder} \\ 
\gll čēge=īč āwī her ne-bīyē=ne bīnder \\ 
 here\textsc{=add} water\textsc{.f} emph \textsc{neg-}be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} at\_all \\ 
\glt `There was no water here.'
\z 
 
\ea \label{ŽE.40}
\textit{her sikeɫ bīyen berqū čīw nebīyen} \\ 
\gll her sikeɫ bīye=n berq=ū čīw ne-bīye=n \\ 
 just ember\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} electricity\textsc{.m}\textsc{\textsc{=ez.gen}} thing\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was only embers [to light the house]. There was no electricity or anything.'
\z 
 
\ea \label{ŽE.56}
\textit{čāweɫī sextī nebīyen} \\ 
\gll č=ā eweɫ-ī sextī ne-bīye=n \\ 
 in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} difficulty\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the beginning there was no hardship.'
\z 
 
\ea \label{ŽE.57}
\textit{čāweɫī hīč nebīyen} \\ 
\gll č=ā eweɫ-ī hīč ne-bīye=n \\ 
 in=\textsc{dem.dist} first\textsc{-obl}\textsc{.m} nothing \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the beginning there was nothing.'
\z 
 
\ea \label{ŽE.58}
\textit{temāʕe nebīyēne} \\ 
\gll temāʕe ne-bīyē=ne \\ 
 greed\textsc{.f} \textsc{neg-}be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `There was no greed.'
\z 
 
\ea \label{ŽE.72}
\textit{pēse īsey nebīyen girđčē ʕāl bo weʕze hīn bo} \\ 
\gll pēse īse-ī ne-bīye=n girđčē ʕāl ∅-b-o weʕze hīn ∅-b-o \\ 
  like now\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} everything good \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} situation\textsc{.f} \textsc{fill} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `It wasn’t like nowadays, when everything is fine, and the situation is thingummy.'
\z 
 
\ea \label{ŽE.83}
\textit{tā qewmekāmā wiš kermē qewmekā wiš kermē bizānmē ī qewmīmā řezāšāne yā řezāšā nīyene} \\ 
\gll tā qewm-ekā=mā wiš ∅-ker-mē qewm-ekā wiš ∅-ker-mē bi-zān-mē ī qewm-ī=mā řezā=šā=ne yā řezā=šā nīye=ne \\ 
 until relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=1pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} agreement\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} or agreement\textsc{.f}\textsc{=3pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We will inform our relatives. We will inform our relatives to see if they agree [to the  marriage] or not.'
\z 
 
\ea \label{ŽE.85}
\textit{eger řezāšā nebīē neđēnē} \\ 
\gll eger řezā=šā ne-bī-ē ne-đe-ēn-ē \\ 
 if agreement\textsc{.f}\textsc{=3pl} \textsc{neg-}be\textsc{.pst}\textsc{-cond}\textsc{.3pl} \textsc{neg-}give\textsc{.prs}\textsc{-pstc-3pl} \\ 
\glt `If they didn’t agree, they wouldn’t give her.'
\z 
 

\ea \label{ŽM.14}
\textit{dizīm nekerđenū hīzīm nekerdenū girđkār nebīyenā} \\ 
\gll dizī=m ne-kerđe=n=ū hīzī=m ne-kerde=n=ū girđkār ne-bīye=nā \\ 
 theft\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and adultery\textsc{.dir}\textsc{.m}\textsc{=\textsc{1sg}} \textsc{neg-}do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and know\_all \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `I have not committed theft or adultery. I was not a know-all person.'
\z 
 
\ea \label{ŽM.28}
\textit{min ečāgene kārīger bīyenā šiš māngē xeberēm nezānān} \\ 
\gll min e=čāgene kārīger bīye=nā šiš mānge-ē xeber-ē=m ne-zānā=n \\ 
 \textsc{1sg} in=there labourer\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} six month\textsc{.f}\textsc{-dir}\textsc{.pl} news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` I was a worker there. I was unaware of them\footnote{Lit. I didn’t know their news.}  (my children) for six months,'
\z 
 
\ea \label{ŽM.30}
\textit{ā wextīyekey řūtē bīyēnē qūtē bīyēnē bīyenšā nebīyenšā xeberēm nezānēnē} \\ 
\gll ā wext-ī-ekey řūt-ē bīyē=nē qūt-ē bīyē=nē bīye=n=šā ne-bīye=n=šā xeber-ē=m ne-zānē=nē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} naked\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} bare\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} news\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=\textsc{1sg}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `[I was unaware if] they were without clothes; if they had [enough]; if they didn’t have [enough]. I didn't know their news.'
\z 
 
\ea \label{ŽM.32}
\textit{ā wextīyekey ečāge hīč xeberēmā nebīyenū telefūnīčū žimārew telefūnī nebēw} \\ 
\gll ā wext-ī-ekey e=čāge hīč xeber-ē=mā ne-bīye=n=ū telefūn=īč=ū žimāre=ū telefūn-ī ne-b-ē=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} in=there nothing news\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and telephone\textsc{.m}\textsc{=add}=and number\textsc{.m}\textsc{=ez.gen} telephone\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}=and \\ 
\glt `I\footnote{Lit. We.}  had no news [of my family] from there. There was no phone or phone number at the time.'
\z 
 
\ea \label{ŽM.36}
\textit{ī šiš mānge xeberū hīčīmā nezānān} \\ 
\gll ī šiš mānge=e xeber=ū hīč-ī=mā ne-zānā=n \\ 
 \textsc{dem.prox} six month\textsc{.f}\textsc{=dem} news\textsc{.m}\textsc{=ez.gen} nothing\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I didn’t have any updates [from my family] during those periods of six months.'
\z 
 
\ea \label{ŽM.49}
\textit{řoē sī penj timenē šesū yerē sāɫē čēweɫī kerū ētir meyānū sāɫū činē bē} \\ 
\gll řo-ē sī penj timen-ē šes=ū yerē sāɫe-ē čēweɫī ∅-ker-ū ētir me-zān-ū sāɫe-ū činē b-ē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} sixty\textsc{=ez.gen} three year\textsc{.f}\textsc{-dir}\textsc{.pl} ago \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc}.\textsc{ptcl} \textsc{neg-}know\textsc{.prs}\textsc{-\textsc{1sg}} year\textsc{.f}\textsc{-\textsc{1sg}} how\_many\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `A daily salary of thirty-five tomans, it was sixty-three years ago. I don’t know which year it was/'
\z 
 
\ea \label{ŽM.56}
\textit{čēweɫ nebīyey bē ētir īseyč xū nīyā} \\ 
\gll čēweɫ nebīyey b-ē ētir īse=yč xū nīye=ā \\ 
 previously poverty\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{disc.ptcl} now\textsc{=add} \textsc{disc.ptcl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the past, there was poverty. But even nowadays, there is poverty.'
\z 
 
\ea \label{ŽM.59}
\textit{enāne ēme wārīyātēmā nīyā} \\ 
\gll enā=ne ēme wārīyāt-ē=mā nīye=ā \\ 
 otherwise\textsc{=\textsc{post}} \textsc{1pl} income\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `because we have no income.'
\z 
 
\ea \label{ŽM.60}
\textit{wārīyātmā nīyā kāsbīmā nīyā kārmā nīyā} \\ 
\gll wārīyāt=mā nīye=ā kāsbī=mā nīye=ā kār=mā nīye=ā \\ 
 income\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} business\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} work\textsc{.m}\textsc{=\textsc{1pl}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We have no income, no business, no jobs.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
