\chapter{All 296 examples of pn in these texts are found in the following sentences:}

\ea \label{ZB.1}
\textit{řoē je řoā yoe mē yoyč mē xizmetū šā ʕelādīnī, her ā šā ʕelādīne} \\ 
\gll řo-ē je řo-ā yo-e m-ē yo=yč m-ē xizmet=ū šā ʕelādīn-ī her ā šā ʕelādīn=e \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m}\textsc{-f} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} one\textsc{.m}\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} emph \textsc{dem.dist} king\textsc{.m} \textsc{pn}\textsc{=dem} \\ 
\glt `Once,\footnote{Lit. One day of days.} a woman and a man\footnote{Lit. One (f.) came, one (m.) came.} came to the service of šā ʕalādīn, the same šā ʕalādīn.'
\z 
 
\ea \label{ZB.4}
\textit{mēnē qeredāx} \\ 
\gll m-ē-nē qeredāx \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{pn} \\ 
\glt `They (the man and his wife) came to Qaradagh.'
\z 
 
\ea \label{ZB.9}
\textit{kuřēš bā, kuřekēš germīyānne nimēniš čene} \\ 
\gll kuř-ē=š b-ā kuř-ekē=š germīyān=ne ni-m-ē-nē=iš čene \\ 
 son\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.prs}\textsc{-3pl} son/boy\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.pl}\textsc{=3sg} \textsc{pn}\textsc{=\textsc{post}} \textsc{neg-}\textsc{ind-}come\textsc{.prs}\textsc{-3pl}\textsc{=3sg} with \\ 
\glt `He had sons. His sons did not accompany him [;they stayed in] Garmiyan. '
\z 
 
\ea \label{ZB.37}
\textit{īne min īnem žinīyen lāw, fermāwištū šēxīn, šā ʕelādīnīwe} \\ 
\gll īne min īne=m žinīye=n lā=ū fermāwišt=ū šēx-ī=n šā ʕelādīn-ī=we \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{1sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} speech\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have heard this tale from —it is the Sheikh’s words— Shah Aladin.'
\z 
 
\ea \label{ZB.39}
\textit{kūč kerā milāwe pey germīyānī} \\ 
\gll kūč ∅-ker-ā mi-l-ā=we pey germīyān-ī \\ 
 migration\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{ind-}go\textsc{.prs}\textsc{-3pl}\textsc{=compl} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They (i.e., the family) migrated [and] went to Garmiyan.'
\z 
 
\ea \label{ZB.60}
\textit{welē īse be qurʕān ne šēx menen, ne meɫā menen, ne kesē hen be weš dāđiš berī lā} \\ 
\gll welē īse be qurʕān ne šēx mene=n ne meɫā mene=n ne kes-ē hen-∅ be weš dāđ=iš ∅-ber-ī lā \\ 
 but now to \textsc{pn} neither sheikh\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} nor mullah\textsc{.m} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} nor person\textsc{.m}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} to good yell\textsc{.m}\textsc{=3sg} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} at\_the\_place\_of \\ 
\glt `However, nowadays, I swear on the Quran, neither a [true] sheikh nor a [true] mullah remains, nor is there anyone one can take counsel with.'
\z 
 
\ea \label{ZQ.2}
\textit{hetā min žinīyenim pīyewe āmā xizmetū šā ʕelādīnī} \\ 
\gll hetā min žinīye=n=im pīye-we āmā xizmet=ū šā ʕelādīn-ī \\ 
 even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} man\textsc{.m} come\textsc{.pst}\textsc{.3sg} service\textsc{.m}\textsc{=ez.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have even heard that a man came to the sevice of Shah Aladin. '
\z 
 
\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.9}
\textit{vātiš, qurbān luāymē pār yā činnē sāɫē čēweɫter luāymē pey sārāɫū ērānī, koy sān} \\ 
\gll vāt=iš qurbān luā-ymē pār yā činnē sāɫe-ē čēweɫter luā-ymē pey sārāɫ=ū ērān-ī ko=y sān \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} go\textsc{.pst} \textsc{1pl} last\_year either some\textsc{.pl} year.f\textsc{-indf} ago go\textsc{.pst} \textsc{1pl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} where=ez.\textsc{attr} \textsc{pn} \\ 
\glt `He said, ‘Sir, we went, say, last year or a few years ago, we went to the Saral region in Iran, to the San mountain.'
\z 
 
\ea \label{ZQ.10}
\textit{čowe sārāɫū ērānīwe gēɫāymēwe} \\ 
\gll č=o=we sārāɫ=ū ērān-ī=we gēɫā-ymē=we \\ 
 in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=compl} turn\textsc{.pst} \textsc{1pl}\textsc{=compl} \\ 
\glt `On the way back, we returned from the Saral region of Iran.'
\z 
 
\ea \label{ZQ.27}
\textit{luāymēwe pey germīyānū ʕērāqī} \\ 
\gll luā-ymē=we pey germīyān=ū ʕērāq-ī \\ 
 go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We went back to the Garmiyan region of Iraq.'
\z 
 
\ea \label{ZQ.33}
\textit{vātiš, qurbān wehārē wāwey luāymēwe pey sārāɫū ērānī} \\ 
\gll vāt=iš qurbān wehār-ē wāwey luā-ymē=we pey sārāɫ=ū ērān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} spring\textsc{.m}\textsc{-indf} again go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘Sir, [The next] spring, again, we went back to the Saral region of Iran.'
\z 
 
\ea \label{HB.1}
\textit{zemānē hezretū šēh ʕelādīnī šēh ʕelādīnī mertebeš ʕālī bo} \\ 
\gll zemān-ē hezret=ū šēh ʕelādīn-ī šēh ʕelādīn-ī mertebe=š ʕālī ∅-b-o \\ 
 time\textsc{.m}\textsc{-indf} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} rank\textsc{.m}\textsc{=3sg} eminent \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `In the time of Sheikh Aladin—Sheikh Aladin was eminent.'
\z 
 
\ea \label{HB.2}
\textit{nānewēkēw šēh ʕelādīnī seʕbē hurmēzāwe xeber miđā māčā, xeber deydē be hezretū šēxī, be xwā xāneqāne ne nān hen, ne hārdī henē} \\ 
\gll nānewē-ekē=ū šēh ʕelādīn-ī seʕbe-ē hur-mi-ēz ā=we xeber mi-đe-ā mi-āč-ā xeber ∅-de-ydē be hezret=ū šēx-ī be xwā xāneqā=ne=ne nān hen-∅=ne hārdī hen-ē \\ 
 baker\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{.pl}=and sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} morning\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs} \textsc{dem.dist}\textsc{=compl} news\textsc{.m} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} news\textsc{.m} \textsc{imp-}give\textsc{.prs}\textsc{-2pl} to his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} by God\textsc{.m} monastery\textsc{.m}\textsc{=\textsc{post}}=neither bread\textsc{.m} exist\textsc{-3sg}\textsc{.m}=nor flour\textsc{\textsc{.f}} exist\textsc{-3pl} \\ 
\glt `Sheikh Aladin’s bakers got up in the morning and passed on the news, saying, ‘Pass on the news to his highness the Sheikh. By God, there is neither bread nor flour in the monastery. '
\z 
 
\ea \label{HB.12}
\textit{bilo pey šeškū Bešāretī, āsāwekew šeškī} \\ 
\gll bi-l-o pey šešk=ū Bešāret-ī āsāw-eke=w šešk-ī \\ 
 \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} for \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} mill\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] go to Shashk Basharat\footnote{A village in Iraqi Kurdistan.}, to the mill of Shashk'
\z 
 
\ea \label{HB.25}
\textit{demē gerekiš bē herekey gēɫnowe bēwe pey Bīyārē} \\ 
\gll dem-ē gerek=iš b-ē her-ekey ∅-gēɫn-o=we b-ē=we pey Bīyārē \\ 
 once\textsc{-indf} necessary\textsc{=3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}lead\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{=\textsc{compl}} to \textsc{pn} \\ 
\glt `At some point he wanted to lead back his donkey to go back to Biyare.\footnote{A village in Iraqi Kurdistan.}'
\z 
 
\ea \label{HB.30}
\textit{firey lūān nēzīkū Šeškīū kewteno} \\ 
\gll firey lūā=n nēzīk=ū Šešk-ī=w kewte=n=o \\ 
 a lot go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} near\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}=and fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{compl}} \\ 
\glt `He had gone a long way [and] approached Shashk'
\z 
 
\ea \label{HB.32}
\textit{ēme milmē Šešìk} \\ 
\gll ēme mi-l-mē Šešìk \\ 
 \textsc{1pl} \textsc{ind-}go\textsc{.prs}\textsc{-1pl} \textsc{pn} \\ 
\glt `We will go to Shashk.'
\z 
 
\ea \label{HB.48}
\textit{dāš wene hereš berd gošew šārū šeškī} \\ 
\gll dā=š wene her-e=š berd-∅ goše=w šār=ū šešk-ī \\ 
 give\textsc{.pst}\textsc{=3sg} \textsc{post} donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{=3sg} take\textsc{.pst}\textsc{-3sg}\textsc{.m} corner\textsc{=ez}\textsc{.gen} city\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He set off. He took the donkey to the suburb of Shashk city.'
\z 
 
\ea \label{HB.60}
\textit{āmā āsāwekew Šeškī, kābrāke} \\ 
\gll āmā āsāw-eke=w Šešk-ī kābrā-eke \\ 
 come\textsc{.pst}\textsc{.3sg} mill\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} man\textsc{-def}\textsc{.dir}\textsc{.m} \\ 
\glt `The fellow came to the Shashk mill.'
\z 
 
\ea \label{HB.65}
\textit{xeylēč menebē bilowe Bīyārē} \\ 
\gll xeylē=č mene b-ē bi-l-o=we Bīyārē \\ 
 so much\textsc{=add} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{pn} \\ 
\glt `He had a long way to go back to Biyare.'
\z 
 
\ea \label{HB.66}
\textit{xeylēč menebē bilowe Bīyārēw} \\ 
\gll xeylē=č mene b-ē bi-l-o=we Bīyārē=ū \\ 
 so much\textsc{=add} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{pn}=and \\ 
\glt `He had a long way to go back to Biyare.'
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{PM.1}
\textit{ʕerzū xizmetīt kerū hetā min žinīyenim pāse mešhūr bīyen šewē šā ʕelādīnī, šewē peɫē merekuř peɫē merekuř mē šārezūr} \\ 
\gll ʕerz=ū xizmet-ī=t ∅-ker-ū hetā min žinīye=n=im pāse mešhūr bīye=n šewe-ē šā ʕelādīn-ī šewe-ē peɫ-ē merekuř peɫ-ē merekuř m-ē šārezūr \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} service\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} even \textsc{1sg} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} like famous be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} night.f\textsc{-obl}.f king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} night.f\textsc{-obl}.f \textsc{clf}\textsc{-indf} grasshopper \textsc{clf}\textsc{-indf} grasshopper \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{pn} \\ 
\glt `Let me tell you .\footnote{Lit. May a petition be for your service.} I have even heard (this). It is well known that a swarm of grasshoppers came to Sharazour one night.'
\z 
 
\ea \label{PM.4}
\textit{mē šārezūr} \\ 
\gll m-ē šārezūr \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{pn} \\ 
\glt `They came to Sharazour.'
\z 
 
\ea \label{PM.6}
\textit{nimāyinde kīyāno xizmetū šā ʕelādīnī, sāhēbū hīne gewrew merekuřekey} \\ 
\gll nimāyinde ∅-kīyān-o xizmet=ū šā ʕelādīn-ī sāhēb=ū hīne gewre=w merekuř-ekey \\ 
 representative\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{fill} head\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `He, i.e., the head of the grasshoppers, sent a representative to Shah Aladin’s presence;'
\z 
 
\ea \label{PM.7}
\textit{nimāyindey kīyāno bīyārē} \\ 
\gll nimāyinde-ī ∅-kīyān-o bīyārē \\ 
 representative\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} \textsc{pn} \\ 
\glt `He sent a messenger to Biyare'
\z 
 
\ea \label{PM.16}
\textit{šēxī ʕelādīnī fermāwo, lodērā šārezūr!} \\ 
\gll šēx-ī ʕelādīn-ī ∅-fermāw-o lo-dē=rā šārezūr \\ 
 sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} go\textsc{.prs}.\textsc{imp-}\textsc{2pl}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt `Sheikh Aladin said, ‘Go down to Sharazour!'
\z 
 
\ea \label{PM.23}
\textit{šewē merekuř nīšo milo mehelekew šēxī ʕelādīnī} \\ 
\gll šewe-ē merekuř ∅-nīš-o mi-l-o mehele-eke=ū šēx-ī ʕelādīn-ī \\ 
 night.f\textsc{-obl}.f grasshopper \textsc{ind-}sit\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} neighbourhood\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `The grasshoppers stayed\footnote{Lit. sat.} [there]. They attacked\footnote{Lit. sit on.} Sheikh Aladin’s neighbourhood.'
\z 
 
\ea \label{PM.28}
\textit{tāte řow bāwke řo verzīyer mē pey xizmetū šā ʕelādīnī} \\ 
\gll tāte řo=ū bāwke řo verzīyer m-ē pey xizmet=ū šā ʕelādīn-ī \\ 
 father\textsc{.m} onom\textsc{\textsc{=ez.gen}} fahter.\textsc{voc} onom farm\_labourer\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[Sheikh Aladin’s farm-labourer shouted], ‘Oh, dear God! The farm-labourer came into the presence of Shah Aladin.'
\z 
 
\ea \label{PM.31}
\textit{tenhā hinū ton, geɫēwew kesī zereriš neginān je šārezūrne} \\ 
\gll tenhā hin=ū to=n geɫā-ēwe=ū kes-ī zerer=iš ne-ginā=n je šārezūr=ne \\ 
 not\_even ez.pron\textsc{\textsc{=ez.gen}} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} leaf\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} person\textsc{.m}\textsc{-obl}\textsc{.m} harm\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} \\ 
\glt `It’s all yours [otherwise] no one’s leaves have been harmed in Sharazour. '
\z 
 
\ea \label{PM.35}
\textit{eger yerē řoē šārezūrne menēwe geɫāyē sewz šārezūrne nāzē} \\ 
\gll eger yerē řo-ē šārezūr=ne men-ē=we geɫā-ē sewz šārezūr=ne ne-āz-ē \\ 
 if three day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{pn}\textsc{=\textsc{post}} remain\textsc{.prs}\textsc{-cond}\textsc{.3pl}\textsc{=compl} leaf\textsc{.m}\textsc{-indf} green \textsc{pn}\textsc{=\textsc{post}} \textsc{neg-}let\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `If the grasshoppers had stayed in Sharazour for three days, they wouldn’t have left behind a single green leaf in Sharazour.'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.38}
\textit{girđū šārezūrī ēsāɫ merāʕetū ēme kero je pārī zīyāter pey toyčū pey ēmeyč,} \\ 
\gll girđ=ū šārezūr-ī ēsāɫ merāʕet=ū ēme ∅-ker-o je pār-ī zīyā-ter pey to=yč=ū pey ēme=yč \\ 
 all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} this\_year care\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} from last\_year\textsc{-obl}\textsc{.m} increased\textsc{-cmpr} for \textsc{2sg}\textsc{=add}=and for \textsc{1pl}\textsc{=add} \\ 
\glt `This year, all the people in Sharazour care for us— both you and I—, more than the previous year.'
\z 
 
\ea \label{PM.41}
\textit{wextēwe šārezūr mēnēw wesīle yāwo penīū} \\ 
\gll wext-ēwe šārezūr m-ē-nē=ū wesīle ∅-yāw-o penī=ū \\ 
 time\textsc{.m}\textsc{-indf} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{-3pl}=and tool\textsc{.m} \textsc{ind-}arrive\textsc{.prs}\textsc{-3sg} at=and \\ 
\glt `When they came back from Sharazour, the products were ready for harvesting  '
\z 
 
\ea \label{PM.42}
\textit{šēxī ʕelādīn hemīšey bīyārēwe girđčē zīyā kero} \\ 
\gll šēx-ī ʕelādīn hemīše-ī bīyārē=we girđčē zīyā ∅-ker-o \\ 
 sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn} always\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{=\textsc{post}} everything increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Sheikh Aladin was used to having more agricultural products in Biyare.'
\z 
 
\ea \label{PM.44}
\textit{werzīyereke māčo, weɫāhī šēxī ʕelādīne řāsiš fermāwān} \\ 
\gll werzīyer-eke m-āč-o weɫāhī šēx-ī ʕelādīn=e řās=iš fermāwā=n \\ 
 farm\_labourer\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sheikh\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn}\textsc{=dem} truth\textsc{=3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The farm-labourer said; ‘Indeed, Sheikh Aladin was right.'
\z 
 
 \ea \label{DG.1}
\textit{šā ʕelādīnū šā ʕosmān} \\ 
\gll šā ʕelādīn=ū šā ʕosmān \\ 
 king\textsc{.m} \textsc{pn}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{.m} \\ 
\glt `Shah Aladin and Shah Osman.'
\z 
 
\ea \label{DG.2}
\textit{šā ʕelādīn tātew šā ʕosmānī bīyen} \\ 
\gll šā ʕelādīn tāte=ū šā ʕosmān-ī bīye=n \\ 
 king\textsc{.m} \textsc{pn} father\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Shah Aladin was Shah Osman’s father.'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.11}
\textit{řoē je řoā yo mē xizmetiš, xizmetū šā ʕelādīnī} \\ 
\gll řo-ē je řo-ā yo m-ē xizmet=iš xizmet=ū šā ʕelādīn-ī \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} one\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} service\textsc{.m}\textsc{=3sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Once\footnote{Lit. One day of days.}, a man\footnote{Lit. one.}  came into his presence, to Shah Aladin'
\z 
 
\ea \label{DG.24}
\textit{řoē je řoā māčā pīyāē hen nāmēš šā ʕelādīnārēū} \\ 
\gll řo-ē je řo-ā m-āč-ā pīyā-ē hen-∅ nāmē=š šā ʕelādīn=ā=rē=ū \\ 
 day\textsc{.m}\textsc{-dir}\textsc{.pl} from day\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} man\textsc{.m}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} name\textsc{\textsc{.f}}\textsc{=3sg} king\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `Once, people said, ‘There is a man called Aladin. '
\z 
 
\ea \label{DG.25}
\textit{īnā yāneš bīyārēū} \\ 
\gll īnā-∅ yāne=š bīyārē=ū \\ 
 deic\textsc{-3sg}\textsc{.m} house\textsc{=3sg} \textsc{pn}=and \\ 
\glt `He lives in Biyare.'\footnote{Lit. His house is in Biyare.}
\z 
 
\ea \label{DG.27}
\textit{lue wēt wize milū pāɫāyekā šā ʕelādīnīre} \\ 
\gll ∅-lu-e wē=t ∅-wiz-e mil=ū pāɫā-ekā šā ʕelādīn-ī=re \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}throw\textsc{.prs}-\textsc{2sg} on\textsc{=ez}\textsc{.gen} shoe\textsc{-def}\textsc{.obl}\textsc{.pl} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Go, throw yourself at the feet\footnote{Lit. shoes.} of Shah Aladin'
\z 
 
\ea \label{DG.29}
\textit{šewē, de, de řoē bo, yā pāngze řoē bo yā wīs řoē bo, je qeredāxo māroš tā heɫebje} \\ 
\gll šewe-ē de de řo-ē ∅-b-o yā pāngze řo-ē ∅-b-o yā wīs řo-ē ∅-b-o je qeredāx=o m-ār-o=š tā heɫebje \\ 
 night\textsc{\textsc{.f}}\textsc{-indf} ten ten day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} or fifteeen day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} or twenty day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} from \textsc{pn}\textsc{=\textsc{post}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} until \textsc{pn} \\ 
\glt `Whether it was one night, ten days, fifteen days, or twenty days, he took his brother from Qaradax to Halabja.'
\z 
 
\ea \label{DG.30}
\textit{mēnē degāw činārīš māčā} \\ 
\gll m-ē-nē degā=ū činār-ī=š m-āč-ā \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} village\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \\ 
\glt `They came to a village called Chinar.'
\z 
 
\ea \label{DG.31}
\textit{mēnē e čā birākeyš bero teɫānēwene beynū činārāne} \\ 
\gll m-ē-nē e=čā birā-ekey=š ∅-ber-o teɫān-ēwe=ne beyn=ū činār-ā=ne \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} in=there brother\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}take\textsc{.prs}\textsc{-3sg} terrace\_farm\textsc{-indf}\textsc{=\textsc{post}} between\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-3pl}\textsc{=\textsc{post}} \\ 
\glt `They went to a terrace farm in between the the two sides of Chinar village. '
\z 
 
\ea \label{DG.32}
\textit{minyošrew wēš mē xāneqāw bīyārē} \\ 
\gll mi-nye-o=š=re=ū wē=š m-ē xāneqā=ū bīyārē \\ 
 \textsc{ind-}put\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=\textsc{povb}}=and \textsc{reflx}\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} monastery\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \\ 
\glt `He left his brother there, and went to the monastery.'
\z 
 
\ea \label{DG.33}
\textit{xāneqāw bīyārēne, ā wextīyekey dāđ kero} \\ 
\gll xāneqā=ū bīyārē=ne ā wext-ī-ekey dāđ ∅-ker-o \\ 
 monastery\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} yell\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He gave a yell of despair at the monastery of Biyare.'
\z 
 
\ea \label{DG.53}
\textit{šā ʕelādīn īney īney māčo be, fermāwo be pīyā, hīnī birāw pīyākey} \\ 
\gll šā ʕelādīn īney īney m-āč-o be ∅-fermāw-o be pīyā hīn-ī birā=ū pīyā-ekey \\ 
 king\textsc{.m} \textsc{pn} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by man\textsc{.m} \textsc{fill}\textsc{-obl}\textsc{.m} brother\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Shah Aladin said this to the man, to the brother of the [sick] man.'
\z 
 
\ea \label{DG.54}
\textit{birāš īnārē degāw činārī, metāwo bilo āweyānī} \\ 
\gll birā=š īnā-∅=rē degā=ū činār-ī me-tāw-o bi-l-o āweyānī \\ 
 brother\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} village\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} habitat\textsc{\textsc{.m}} \\ 
\glt `His brother was in Chinar village. He wasn’t allowed to go the village '
\z 
 
\ea \label{DG.61}
\textit{kīseɫēwe čā kīseɫā milo xizmetū šā ʕelādīnī} \\ 
\gll kīseɫī-ēwe č==ākīseɫī-ā mi-l-o xizmet=ū šā ʕelādīn-ī \\ 
 tortoise\textsc{\textsc{.f}}\textsc{-indf} from=\textsc{=cop}\textsc{.3sg}\textsc{.m}tortoise\textsc{\textsc{.f}}\textsc{-3pl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} service\textsc{.m}\textsc{=ez}\textsc{.gen} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `A tortoise\footnote{Lit. A tortoise of those tortoises.}  came into the presence of Shah Aladin.'
\z 
 
\ea \label{DG.67}
\textit{šā ʕelādīnīč fermāwo birākem wextē īnē zānā min gunākē māy mē milimre} \\ 
\gll šā ʕelādīn-ī=č ∅-fermāw-o birā-eke=m wext-ē īnē ∅-zān-ā min gunā-ekē m-āy m-ē mil=im=re \\ 
 king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} time\textsc{.m}\textsc{-indf} \textsc{dem.prox}\textsc{.3pl} \textsc{ind-}know\textsc{.prs}\textsc{-3pl} \textsc{1sg} sin\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}come\textsc{.pst}\textsc{.adv} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} shoulder\textsc{.m}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `Shah Aladin said, ‘Brother!—as they (i.e. the tortoises) know [about the medicine] I won’t shoulder the burden of injuring them.'
\z 
 
\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.2}
\textit{yošā be nāmēw sibhān āġeywe} \\ 
\gll yo=šā be nāmē=ū sibhān āġe-ī=we \\ 
 one\textsc{.m}\textsc{=3pl} by name\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `One of them [was] called Sibhan Agha.'
\z 
 
\ea \label{DP.3}
\textit{yošā be nāmēw jemšēr šāywe} \\ 
\gll yo=šā be nāmē=ū jemšēr šā-ī=we \\ 
 one\textsc{.m}\textsc{=3pl} by name\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn} king\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `One of them [was] called Jamsher Agha.'
\z 
 
\ea \label{DP.4}
\textit{jemšēr šā je dūrew āmān} \\ 
\gll jemšēr šā je dūr=ew āmā=n \\ 
 \textsc{pn} king\textsc{.m} from far\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Jamsher Shah came [to Hawraman] from far away.'
\z 
 
\ea \label{DP.9}
\textit{āmān lūān ā qeɫāw hewrāmānī e pāgewe nīštenre} \\ 
\gll āmā=n lūā=n ā qeɫā=w hewrāmān-ī e=pāgewe nīšte=n=re \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} in=there sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} \\ 
\glt `He came [here and] went to the citadel of Hawraman over there and settled there.'
\z 
 
\ea \label{DP.12}
\textit{ā wextī ī jemšēr šā, sibhān āġey čew deso bīyen} \\ 
\gll ā wext-ī ī jemšēr šā sibhān āġe-ī č=ew des=o bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} \textsc{pn} king\textsc{.m} \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} in\textsc{=dem}.dist hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Jamsher Shah, eh … Sibhan Agha was on the other side [of the village].'
\z 
 
\ea \label{DP.18}
\textit{gerekiš bīyen hewrāmānī dāwā kero, dāgīr kero} \\ 
\gll gerek=iš bīye=n hewrāmān-ī dāwā ∅-ker-o dāgīr ∅-ker-o \\ 
 necessary\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} demand\textsc{\textsc{.f}} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} occupied \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wanted to occupy Hawraman.'
\z 
 
\ea \label{DP.19}
\textit{sibhān āġey neyān deso} \\ 
\gll sibhān āġe-ī ne-dā=n des=o \\ 
 \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} hand\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Sibhan Agha did not hand over the governorship to him,'
\z 
 
\ea \label{DP.21}
\textit{ī bāxū bāxātū hewrāmānī, memliketū hewrāmānī, xeɫkū hewrāmānī ā wextīyekey ēme hukimřewiš bīyēnmē} \\ 
\gll ī bāx=ū bāx-āt=ū hewrāmān-ī memliket=ū hewrāmān-ī xeɫk=ū hewrāmān ī ā wext-ī-ekey ēme hukimřew=iš bīyē=nmē \\ 
 \textsc{dem.prox} garden\textsc{.m}\textsc{=ez}\textsc{.gen} garden\textsc{.m}\textsc{-pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} people\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \textsc{dem.prox} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} governor\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `We have ruled over\footnote{Lit. We have been rulers.} these gardens of Hawraman, this region of Hawraman, [and] this people of Hawraman.'
\z 
 
\ea \label{DP.26}
\textit{sibhān āġey ā wextīyekey āġe bīyen ħīn bīyen} \\ 
\gll sibhān āġe-ī ā wext-ī-ekey āġe bīye=n ħīn bīye=n \\ 
 \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} agha\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{fill} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Sibhan Agha was thingummy, the ruler.'
\z 
 
\ea \label{DP.27}
\textit{sibhān āġe čā āġegerīš kerdenū} \\ 
\gll sibhān āġe čā āġegerī=š kerde=n=w \\ 
 \textsc{pn} agha\textsc{.m} there governship\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Sibhan Agha governed over there.'
\z 
 
\ea \label{DP.30}
\textit{šūnū āneyre wextē zāno hāmin āmāw gīriš kerd sibhān āġe milo āwekēū jemšēr āġey biřo, hesāryewe} \\ 
\gll šūn=ū āney=re wext-ē ∅-zān-o hāmin āmā=ū gīr=iš kerd sibhān āġe mi-l-o āwī-ekē=w jemšēr āġe-ī ∅- bi-ř-o hesār-ī=ewe \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} time\textsc{.m}\textsc{-indf} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} summer\textsc{.m} come\textsc{.pst}\textsc{.3sg}=and hook\textsc{.m}\textsc{=3sg} do\textsc{.pst} \textsc{pn} agha\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} water\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}=and \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}\textsc{sbjv-}cut\textsc{.prs}\textsc{-3sg} pond\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Then, as Sibhan Agha\footnote{Lit. he.} knew summer was coming [and] Jamsher Agha\footnote{Lit. he.} was stuck, Sibhan Agha went and cut off Jamsher Shah’s water supply [and led it to a] pond.'
\z 
 
\ea \label{DP.35}
\textit{īzāfeyš wizo pā sextre māčāš qeɫāw hewrāmānī} \\ 
\gll īzāfe-ī=š wiz-o p=ā sext=re mi-āč-ā=š qeɫā=w hewrāmān-ī \\ 
 addition\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg} by\textsc{=dem}.dist mountain\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} castle\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He would throw whatever was extra [of the snow] at that hard-to-access mountain called the citadel of Hawraman.'
\z 
 
\ea \label{DP.46}
\textit{ā wextīyekey wextē mēnē ā wextī herčē qisē kerā, herčē behsē kerā herčē hesābē kerā, qēxāw hewrāmānī herčē bā ā wextī isɫāhšā mebo} \\ 
\gll ā wext-ī-ekey wext-ē mi-ē-nē ā wext-ī herčē qisē-ē ∅-ker-ā herčē behse-ē ∅-ker-ā herčē hesāb-ē ∅-ker-ā qēxā=w hewrāmān-ī herčē ∅-b-ā ā wext-ī isɫāh=šā me-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} however\_much talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much discussion\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} however\_much account\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} chieftain\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} however\_much \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} amendment\textsc{.m}\textsc{=3pl} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then Sibhan Agha and his men came [to Jamsher Shah]. No matter how much the headmen of Hawraman talked [to both of them], discussed the issue, and tried to mediate, they (i.e., Jamsher Shah and Sibhan Agha) disagreed.'
\z 
 
\ea \label{DP.53}
\textit{āđ bowe pādšā tā zemānēwe nāđir šā mē serū dewrī} \\ 
\gll āđ ∅-b-o=we pādšā tā zemān-ē-ewe nāđir šā m-ē ser=ū dewr-ī \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} king\textsc{.m} until time\textsc{.m}\textsc{-indf}\textsc{-indf} \textsc{pn} king\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} on\textsc{=ez}\textsc{.gen} period\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (i.e., Jamsher Shah) became the king up until the time of Nader Shah.'
\z 
 
\ea \label{DP.54}
\textit{nāđir šā mē serū dewrī her wī berkenārē kero} \\ 
\gll nāđir šā m-ē ser=ū dewr-ī her wē-ī berkenār-ē ∅-ker-o \\ 
 \textsc{pn} king\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} on\textsc{=ez}\textsc{.gen} period\textsc{.m}\textsc{-obl}\textsc{.m} both two\textsc{-obl}\textsc{.m} ousted\textsc{-dir}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Nader Shah came to power [and] expelled both [Jamsher Shah and Sibhan Agha]'
\z 
 
\ea \label{ZP.1}
\textit{ʕerz kerā pīr šelīyār, nāmēš sey misefān} \\ 
\gll ʕerz ∅-ker-ā pīr šelīyār nāmē=š sey misefā=n \\ 
 petition\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{pn} \textsc{pn} name\textsc{.f}\textsc{=3sg} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `It is said that Pir Shaliyar, his [real] name was Say Mustafa.'
\z 
 
\ea \label{ZP.2}
\textit{kuřū sey meʕrūfīn} \\ 
\gll kuř=ū sey meʕrūf-ī=n \\ 
 son/boy\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is Say Maʕruf’s son.'
\z 
 
\ea \label{ZP.3}
\textit{čī degāw hewrāmānīne serāwinū degākēwe je eyē bīyen} \\ 
\gll č=ī degā=ū hewrāmān-ī=ne serāwin=ū degā-ekē=we je eyē bīye=n \\ 
 in=dem.prox village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} top\_of\_village\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} from mother\textsc{.obl}\textsc{.f} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was born in the village of Hawraman, at the top point of the village.'
\z 
 
\ea \label{ZP.6}
\textit{ā šexse īnā ew desū degāw hewrāmānīne māčāš pene pīr ħeme yoso māmoš bīyen} \\ 
\gll ā šexs=e īnā-∅ ew des=ū degā=ū hewrāmān-ī=ne m-āč-ā=š pene pīr ħeme yoso māmo=š bīye=n \\ 
 \textsc{dem.dist} saint\textsc{=dem} deic\textsc{-3sg}\textsc{.m} \textsc{dem.dist} hand\textsc{.m}\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn} \textsc{pn} \textsc{pn} paternal\_uncle\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The saint [whose grave is] on the other side of Hawraman is called Pir Hama Yoso. He is his (Pir Shaliyar’s) paternal uncle.'
\z 
 
\ea \label{ZP.10}
\textit{kīyāsenšā māčāš pene nehāɫū pīr šelīyārī řūbeřū biɫbeřī} \\ 
\gll kīyāse=n=šā m-āč-ā=š pene nehāɫ=ū pīr šelīyār-ī řūbeřū=ū biɫbeř-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They sent him to [a place] called the valley of Pir Shaliyar; [it is] on the opposite side of Bilbar\footnote{A village close to Hawraman Takht}.'
\z 
 
\ea \label{ZP.14}
\textit{kīyāsenšā pey beġā pey xizmetū hezretū ġosī} \\ 
\gll kīyāse=n=šā pey beġā pey xizmet=ū hezret=ū ġos-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} for \textsc{pn} for service\textsc{.m}\textsc{\textsc{=ez.gen}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[Then] they sent him to Baghdah, to the presence of his highness the Ghaws.'
\z 
 
\ea \label{ZP.18}
\textit{yāwān pene, hezretū ġosī fermāwān exī mistefā rejilun kāmilun} \\ 
\gll yāwā=n pene hezret=ū ġos-ī fermāwā=n exī mistefā rejilun kāmilun \\ 
 arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} my\_brother \textsc{pn} man complete \\ 
\glt `He became experienced (in his field). His Highness the Ghaws said, ‘My brother, Mustafa, is a well-rounded man.’'
\z 
 
\ea \label{ZP.20}
\textit{kīyāsenšo pey menteqew hewrāmānī} \\ 
\gll kīyāse=n=š=o pey menteqe=ū hewrāmān-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}\textsc{=compl} for region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (Ghaws) sent him back to the Hawraman region.'
\z 
 
\ea \label{ZP.21}
\textit{kīyāsenšo pey menteqew hewrāmānīw āmāno menteqew hewrāmānī} \\ 
\gll kīyāse=n=š=o pey menteqe=ū hewrāmān-ī=ū āmā=n=o menteqe=ū hewrāmān-ī \\ 
 send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}\textsc{=compl} for region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (Ghaws) sent him back to the Hawraman region. He came to the Hawraman region.'
\z 
 
\ea \label{ZP.23}
\textit{e pā menteqew hewrāmānī ētir feqīyēwe bīyenū wenānšū} \\ 
\gll e=p=ā menteqe=ū hewrāmān-ī ētir feqī-ēwe bīye=n=ū wenā=n=š=ū \\ 
 in=by=\textsc{dem.dist} region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} theologian\textsc{.m}\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} read\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}=and \\ 
\glt `In this Hawraman region, well, he was a theologian; he was literate.'
\z 
 
\ea \label{ZP.25}
\textit{čā dimāy, kināčēw pādšāw misrī neweše gino} \\ 
\gll čā dimā-ī kināčē=ū pādšā=w misr-ī neweš-e ∅-gin-o \\ 
 there afterwards\textsc{-obl}\textsc{.m} daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} ill\textsc{-f} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `After a while, the king of Egypt’s daughter fell sick.'
\z 
 
\ea \label{ZP.31}
\textit{māčo, māčā neferē īnārē je menteqew hewrāmānātī be nāmēw fiɫāne kesī} \\ 
\gll m-āč-o m-āč-ā nefer-ē īnā-∅=rē je menteqe=ū hewrāmān-āt-ī be nāmē=ū fiɫān=e kes-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} person\textsc{.m}\textsc{-indf} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} from region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} to name\textsc{.f}\textsc{\textsc{=ez.gen}} such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `It was said [to him in the dream] that there is a person in the Hawram region called such-and-such.'
\z 
 
\ea \label{ZP.34}
\textit{meger berī kīyānīš pey hewrāmānī} \\ 
\gll meger ∅-ber-ī ∅-kīyān-ī=š pey hewrāmān-ī \\ 
 if\_only \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}send\textsc{.prs}-\textsc{2sg}\textsc{=3sg} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Maybe if you take her to him.'
\z 
 
\ea \label{ZP.36}
\textit{īne māčmēš pene šālīyāre sīyāw lālow kināčekēn} \\ 
\gll īne m-āč-mē=š pene šālīyār=e sīyāw lālo=ū kināčē-ekē=n \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn}\textsc{=dem} black maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This [person], whom we call Shaliyar Siya, is the maternal uncle of the [king’s] daughter.'
\z 
 
\ea \label{ZP.41}
\textit{degā be degāw wirde wirde ārdēnešū āmān řūbeřū hewrāmānātī} \\ 
\gll degā be degā=ū wirde wirde ārdē=ne š=ū āmā=n řūbeřū=ū hewrāmān-āt-ī \\ 
 village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} little little bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} go\textsc{.prs}\textsc{\textsc{=ez.gen}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the uncle) led her village by village and came [to a place] next to the Hawraman River.'
\z 
 
\ea \label{ZP.50}
\textit{tewenē īnāre beynū biɫbeřīnū silēnēne} \\ 
\gll tewenī-ē īnā-∅=re beyn=ū biɫbeř-ī=ne=ū silēnē=ne \\ 
 stone\textsc{.f}\textsc{-indf} deic\textsc{-3sg}\textsc{.m}\textsc{=\textsc{povb}} among\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `There is a stone between Bilbar and Slen.'
\z 
 
\ea \label{ZP.61}
\textit{her ā dāđe kero, ētir māčo, erē weɫā sirewā’ hurmēzowe miđo wene māroš hesārū činārārū} \\ 
\gll her ā dāđ=e ∅-ker-o ētir m-āč-o erē weɫā sirewā hur-m-ēz-o=we mi-đe-o wene m-ār-o=š hesār=ū činār-ā=re=ū \\ 
 emph \textsc{dem.dist} yell\textsc{.m}\textsc{=dem} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes indeed be cured\textsc{.pst}\textsc{.3sg} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} at \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} pond\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \\ 
\glt `After he shouted [and] said, ‘Indeed she is healed’, then he rose, [and] set off, and took the girl to Hasar Chinara.'
\z 
 
\ea \label{ZP.62}
\textit{mēnē mizyā werāwerū degāw hewrāmānī} \\ 
\gll m-ē-nē mi-zye-ā werāwer=ū degā=ū hewrāmān-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They ended up at the opposite side of Hawraman village.'
\z 
 
\ea \label{ZP.63}
\textit{mizyo, mizyā werāwerū māčmēš pene dūtifɫe} \\ 
\gll mi-zye-o mi-zye-ā werāwer=ū m-āč-mē=š pene dūtifɫe \\ 
 \textsc{ind-}go\_out\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn} \\ 
\glt `[Then] they left [Hawraman village and went to a place which] we call Dutifla'
\z 
 
\ea \label{ZP.64}
\textit{mizyā werāwerū dūtifɫey} \\ 
\gll mi-zye-ā werāwer=ū dūtifɫey \\ 
 \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `[Then] they left Dutifla'
\z 
 
\ea \label{ZP.65}
\textit{pīr šelīyār ā vextī yānekew yānew tāteyšne bo, pū ser degāwe} \\ 
\gll pīr šelīyār ā vext-ī yāne-eke=ū yāne=ū tāte-ī=š=ne ∅-b-o p=ū ser degā=we \\ 
 \textsc{pn} \textsc{pn} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} house\textsc{.m}\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at=\textsc{dem.dist}\textsc{.3sg} on village\textsc{.f}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar, well, he was [living] in his father’s house, [which was located] at the top of the village'
\z 
 
\ea \label{ZP.78}
\textit{e čā pīr šelīyār nezerēš wene kero} \\ 
\gll e=čā pīr šelīyār nezer-ē=š wene ∅-ker-o \\ 
 in=there \textsc{pn} \textsc{pn} vision\textsc{-indf}\textsc{=3sg} at \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `There, Pir Shaliyar cast an evil eye on him (the demon)'
\z 
 
\ea \label{ZP.92}
\textit{meʕeřefīš kerā pey lā pīr šelīyārī} \\ 
\gll meʕeřefī=š ∅-ker-ā pey lā pīr šelīyār-ī \\ 
 introduction\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} for at\_the\_place\_of \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They introduced them to Pir Shaliyar.'
\z 
 
\ea \label{ZP.93}
\textit{kināčēw pādšāy mārā lā pīr šelīyārī} \\ 
\gll kināčē=ū pādšā-ī m-ār-ā lā pīr šelīyār-ī \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl} at\_the\_place\_of \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They brought the king’s daughter to Pir Shaliyar.'
\z 
 
\ea \label{ZP.94}
\textit{ā wextī ēme xeplezeřetīnē bēnē tā pāyīz} \\ 
\gll ā wext-ī ēme xeplezeřetīnē b-ēn-ē tā pāyīz \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{1pl} \textsc{pn}\textsc{.dir}\textsc{.pl} be\textsc{.prs}\textsc{-pstc}\textsc{-3pl} until autumn \\ 
\glt `Back then, there were farms of black corn\footnote{A type of corn. Also a type of bread made from corn flour.} until autumn.'
\z 
 
\ea \label{ZP.95}
\textit{šāmne ā tome pīr šelīyārī wēš ārden je šāmne bē} \\ 
\gll šām=ne ā tom=e pīr šelīyār-ī wē=š ārde=n je šām=ne b-ē \\ 
 \textsc{pn}\textsc{=\textsc{post}} \textsc{dem.dist} seed\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{reflx}\textsc{=3sg} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `From Damascus, Pir Shaliyar had brought its seeds. He was in Damascus (for a while).'
\z 
 
\ea \label{ZP.96}
\textit{počīš māčēnē zārātū pīr šelīyārī} \\ 
\gll počī=š m-āč-ēn-ē zārāt=ū pīr šelīyār-ī \\ 
 that\_is\_why\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-pstc}\textsc{-3pl} corn\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `That is why they (the farms) are called Pir Shaliyar’s farms.'
\z 
 
\ea \label{ZP.97}
\textit{ā wextī mārāš mē ʕerzū pīr šelīyārī kero lāloekeš} \\ 
\gll ā wext-ī m-ār-ā=š m-ē ʕerz=ū pīr šelīyār-ī ∅-ker-o lālo-eke=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \\ 
\glt `They (the people) brought her. They came (to Pir Shaliyar). The uncle said to Pir Shaliyar,'
\z 
 
\ea \label{ZP.126}
\textit{ā wextī pīr šelīyār fermāwo, šime īne māčdē řāstā wāt} \\ 
\gll ā wext-ī pīr šelīyār ∅-fermāw-o šime īne m-āč-dē řās=tā wāt \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{2pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} \\ 
\glt `Then Pir Shaliyar said, ‘What you are saying is right.’\footnote{Lit. You said this; you told the truth.}'
\z 
 
\ea \label{ŽP.1}
\textit{nāmē heqetīneš sey misefān} \\ 
\gll nāmē=ī heqetīne=š sey misefā=n \\ 
 name\textsc{.f}\textsc{=ez}.\textsc{attr} real\textsc{=3sg} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `His real name is Say Mustafa'
\z 
 
\ea \label{ŽP.2}
\textit{nāmē wēš pīr šelīyāre sey misefān} \\ 
\gll nāmē=ī wē=š pīr šelīyār-e sey misefā=n \\ 
 name\textsc{.f}\textsc{=ez}.\textsc{attr} \textsc{reflx}\textsc{=3sg} \textsc{pn} \textsc{pn}\textsc{-def} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s name is [initially] Say Mustafa.'
\z 
 
\ea \label{ŽP.3}
\textit{māčāš pīr šelīyār yānē lūān pēwāw xwāywe} \\ 
\gll m-āč-ā=š pīr šelīyār yānē lūā=n pēwā=ū xwā=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{pn} \textsc{pn} that\_means go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} welcoming\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He is referred to as Pir Shaliyar, meaning that he has gone [and] embraced God.'
\z 
 
\ea \label{ŽP.4}
\textit{pewkīmā leqemšā miđā māčāš pīr šelīyār} \\ 
\gll pewkī=mā leqem=šā mi-đe-ā m-āč-ā=š pīr šelīyār \\ 
 that\_is\_why\textsc{=1pl} title\textsc{.m}\textsc{=3pl} \textsc{ind-}give\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{pn} \textsc{pn} \\ 
\glt `That is why he’s been given this title [and] is called Pir Shaliyar.'
\z 
 
\ea \label{ŽP.5}
\textit{tāteš nāmēš pīr bāyēɫā} \\ 
\gll tāte=š nāmē=š pīr bāyēɫ=ā \\ 
 father\textsc{.m}\textsc{=3sg} name\textsc{.f}\textsc{=3sg} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `His father’s name\footnote{Lit. His father, his name.} was Pir Bayel.'
\z 
 
\ea \label{ŽP.8}
\textit{tāteš fewt kero. pīr bāyēɫ fewt kero.} \\ 
\gll tāte=š fewt ∅-ker-o pīr bāyēɫ fewt ∅-ker-o \\ 
 father\textsc{.m}\textsc{=3sg} death\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{pn} \textsc{pn} death\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `His father passed away. Pir Bayel passed away.'
\z 
 
\ea \label{ŽP.10}
\textit{pīr šelīyār hetīm bo} \\ 
\gll pīr šelīyār hetīm ∅-b-o \\ 
 \textsc{pn} \textsc{pn} orphan \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt ` Pir Shaliyar was an orphan.'
\z 
 
\ea \label{ŽP.11}
\textit{milo lāw māmoyš māmo yosoy} \\ 
\gll mi-l-o lā=ū māmo-ī=š māmo yoso-ī \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3sg} paternal\_uncle \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He went to his uncle, Uncle Yoso.'
\z 
 
\ea \label{ŽP.12}
\textit{māmo yoso ā pīr heme yoso māmošā} \\ 
\gll māmo yoso ā pīr heme yoso māmo=š=ā \\ 
 paternal\_uncle \textsc{pn} \textsc{dem.dist} \textsc{pn} \textsc{pn} \textsc{pn} paternal\_uncle\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Uncle Yoso, [otherwise called] Pir Hama Yoso, was his uncle.'
\z 
 
\ea \label{ŽP.19}
\textit{ānā ew dīmū qeɫāw kitāy} \\ 
\gll āne=ā ew dīm=ū qeɫā=ū kitā-ī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `—Look, it is [located] on the other side of the Kita castle'
\z 
 
\ea \label{ŽP.20}
\textit{yāgē hen māčāš nehāɫū pīr šelīyārā—} \\ 
\gll yāgē-ē hen-∅ m-āč-ā=š nehāɫ=ū pīr šelīyār-ā \\ 
 place\textsc{.f}\textsc{-indf} exist\textsc{-3sg}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \\ 
\glt `There is a place called the valley of Pir Shaliyar.'
\z 
 
\ea \label{ŽP.22}
\textit{heremešū pīr heme yoso kīyānāš čenū bizā pey ewāgey} \\ 
\gll hereme=š=ū pīr heme yoso ∅-kīyān-ā=š čenū bize-ā pey ewāge-ī \\ 
 wife\textsc{.f}\textsc{=3sg}=and \textsc{pn} \textsc{pn} \textsc{pn} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} with\textsc{.ez.gen} goat\textsc{.f}\textsc{-obl}\textsc{.pl} to there\textsc{-obl}\textsc{.m} \\ 
\glt `Pir Hama Yoso and his wife sent him there [for shepherding] together with the goats.'
\z 
 
\ea \label{ŽP.27}
\textit{māčā, dey to metāwī -sey misefāw- kučēš yewē berī nehāɫū pīr šelīyārā šānīšā pižgīnīšā bā bizyā’} \\ 
\gll m-āč-ā dey to me-tāw-ī sey misefā=ū kuč-ē=š yewe-ē ∅-ber-ī nehāɫ=ū pīr šelīyār-ā ∅-šān-ī=šā ∅-pižgīn-ī=šā bā bi-zye-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{2sg} \textsc{neg-}can\textsc{.prs}-\textsc{2sg} \textsc{pn} \textsc{pn}=and little\textsc{-indf}\textsc{=3sg} barley\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \textsc{sbjv-}throw\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}scatter\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \textsc{hort} \textsc{sbjv-}grow\textsc{.prs}\textsc{-3pl} \\ 
\glt `They said, ‘You are not able, Say Mustafa, to take some barley seeds to the valley of Pir Shaliyar, [and] scatter them [on the field], [and] plant them, so that they grow.'
\z 
 
\ea \label{ŽP.31}
\textit{milo ew dīmū qeɫāw kitā} \\ 
\gll mi-l-o ew dīm=ū qeɫā=w kitā \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `He went to the other side of the Kita castle.'
\z 
 
\ea \label{ŽP.43}
\textit{māčo bē dengē māmo yoso, dā lue šonū sey misefāyre bizāne řās māčo!} \\ 
\gll m-āč-o bē- deng-ē māmo yoso dā ∅-lu-e šūn=ū sey misefā-ī=re bi-zān-e řās m-āč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}without noise\textsc{.m}\textsc{-indf} paternal\_uncle \textsc{pn} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} after\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} truth \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \\ 
\glt `Uncle Yoso secretly said [to the servant], ‘Go after Say Mustafa, see if he is telling the truth!'
\z 
 
\ea \label{ŽP.57}
\textit{pīyāke mēwe ʕerzū heme yosoy kero māčo, weɫāhī qurbān min tūšū jeryānē e čīney āmānē} \\ 
\gll pīyā-eke m-ē=we ʕerz=ū heme yoso-ī ∅-ker-o m-āč-o weɫāhī qurbān min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sir.\textsc{voc} \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `The man returned [and said] to Hama Yoso, ‘Indeed, Sir! I have witnessed an event like this.'
\z 
 
\ea \label{ŽP.62}
\textit{wēregā pīr šelīyār mēwe sey misefā wextē mēwe, ʕerzū ženekēš kero māčo, bā ēšew īmti̩hānēš kermē lāw wēmāwe newsinmēš} \\ 
\gll wēregā pīr šelīyār m-ē=we sey misefā wext-ē m-ē=we ʕerz=ū ženī-ekē=š ∅-ker-o m-āč-o bā ēšew īmti̩hān-ē=š ∅-ker-mē lā=ū wē=mā=we ne-ūsin-mē=š \\ 
 evening \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{pn} \textsc{pn} time\textsc{.m}\textsc{-indf} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{hort} tonight test\textsc{-indf}\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=1pl}\textsc{=\textsc{post}} \textsc{neg-}put\_to\_sleep\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \\ 
\glt `Pir Shaliyar returned in the evening; when Say Mustafa returned, he (Hama Yoso) said to his wife, ‘Let’s test him tonight; let’s not put him to sleep next to us.’'
\z 
 
\ea \label{ŽP.71}
\textit{pīr heme yoso wēš milo temāšeš kero miđyo āne īnā hewāwe zāno īne ētir kārī xwāyīn} \\ 
\gll pīr heme yoso wē=š mi-l-o temāše=š ∅-ker-o mi-đye-o āne īnā-∅ hewā=we ∅-zān-o īne ētir kār-ī xwāyī=n \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{reflx}\textsc{=3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} looking\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} task\textsc{.m}\textsc{-obl}\textsc{.m} divine\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Hama Yoso went [and] looked at him; he noticed that he (Pir Shaliyar) was aloft in the air [up until a certain height] and came to realise that that was God’s work.'
\z 
 
\ea \label{ŽP.72}
\textit{ʕerzū heremēš kero, fermāwo be heremēš weɫāhī ētir ēme metāwmē, sey misefā metāwmē hīč kārēš pene kermē īlā kīyānmēš bilo weno, bilo kīyānmēš weno} \\ 
\gll ʕerz=ū hereme-ē=š ∅-ker-o ∅-fermāw-o be hereme-ē=š weɫāhī ētir ēme me-tāw-mē sey misefā me-tāw-mē hīč kār-ē=š pene ∅-ker-mē īlā ∅-kīyān-mē=š bi-l-o ∅-wen-o bi-l-o ∅-kīyān-mē=š ∅-wen-o \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} indeed \textsc{disc.ptcl} \textsc{1pl} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} \textsc{pn} \textsc{pn} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} except \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}read\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}read\textsc{-3sg} \\ 
\glt `He (Pir Hama Yoso) said to his wife, ‘Indeed, we can no longer be helpful to Say Mustafa,\footnote{Lit. we can do nothing to him.} except that we send him to study; that we send him, to go [and] study.'
\z 
 
\ea \label{ŽP.74}
\textit{māčo min īnem dī be sey misefāywe} \\ 
\gll m-āč-o min īne=m dī-∅ be sey misefā-ī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} see\textsc{.pst}\textsc{-3sg}\textsc{.m} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘I witnessed this about Say Mustafa.’'
\z 
 
\ea \label{ŽP.80}
\textit{je her yāgē weno, čā dimāyē wextē miđyā ēnneš wenā qābilyetiš peyā kerd kīyānāš pey beġāy pey xizmetū, kīyānāš pey beġāy wenāy} \\ 
\gll je her yāgē-ē ∅-wen-o č=ā dimā-ē wext-ē mi-đye-ā ēnne=š wenā qābilyet=iš peyā kerd ∅-kīyān-ā=š pey beġā-ī pey xizmet=ū ∅-kīyān-ā=š pey beġā-ī wenāy \\ 
 from emph place\textsc{.f}\textsc{-indf} \textsc{ind-}read\textsc{-3sg} in=\textsc{dem.dist} afterwards\textsc{-obl}\textsc{.f} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3pl} so\_much\textsc{=3sg} read\textsc{.pst} capability\textsc{.m}\textsc{=3sg} visible do\textsc{.pst} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn}\textsc{-obl}\textsc{.m} to service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to \textsc{pn}\textsc{-obl}\textsc{.m} read.inf \\ 
\glt `Wherever he studied, after a while, they realised that he studied to the extent that he deserved to be sent to Baghdad into the presence of his highness [Khos]'
\z 
 
\ea \label{ŽP.82}
\textit{šūnū miqdārēre hezretū xosī fermāwo duē yerē īmtihānēš kero} \\ 
\gll šūn=ū miqdārē=re hezret=ū xos-ī ∅-fermāw-o duē yerē īmtihān-ē=š ∅-ker-o \\ 
 after\textsc{\textsc{=ez.gen}} a\_little\textsc{=\textsc{post}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} two three test\textsc{-dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `After a while, his Highness Khos said … He tested him two, three times.'
\z 
 
\ea \label{ŽP.83}
\textit{īmtihānēš kero, fermāwo, exī mistefā rejilun kāmilun yānē be ʕerebī wēšā pāseš fermāwān hezretū xosī} \\ 
\gll īmtihān-ē=š ∅-ker-o ∅-fermāw-o exī mistefā rejilun kāmilun yānē be ʕerebī wē=šā pāse=š fermāwā=n hezret=ū xos-ī \\ 
 test\textsc{-dir}\textsc{.pl}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} my\_brother \textsc{pn} man complete that\_means by Arabic \textsc{reflx}\textsc{=3pl} this\textsc{=3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He tested him [and] said, ‘My brother Mustafa is a well-rounded man, which in their Arabic means, his Highness Khos said,'
\z 
 
\ea \label{ŽP.84}
\textit{misefā yāwān pene tāwo je šerīʕetīč īstīfāde kero je terēqetīč} \\ 
\gll misefā yāwā=n pene ∅-tāw-o je šerīʕet-ī=č īstīfāde ∅-ker-o je terēqet-ī=č \\ 
 \textsc{pn} arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \textsc{ind-}can\textsc{.prs}\textsc{-3sg} from sharia\_law\textsc{-obl}\textsc{.m}\textsc{=add} use \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} from denomination\textsc{-obl}\textsc{.m}\textsc{=add} \\ 
\glt `Mustafa has got [knowledgeable] enough that he can both practise Sharia law and lead an order (of dervishes).'
\z 
 
\ea \label{ŽP.86}
\textit{exī mistefā rejilun kāmilun} \\ 
\gll exī mistefā rejilun kāmilun \\ 
 my\_brother \textsc{pn} man complete \\ 
\glt `My brother Mustafa is a well-rounded man. '
\z 
 
\ea \label{ŽP.91}
\textit{fermāwo hezretū xos, dey to mišyo tešrīf berī bilīwe, pey menteqew wētāw} \\ 
\gll ∅-fermāw-o hezret=ū xos dey to mišyo tešrīf ∅-ber-ī bi-l-ī=we pey menteqe=ū wē=tā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{disc.ptcl} \textsc{2sg} aux honouring \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} to region=and \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `His Highness Khos said, ‘You should leave [here and] return to your region.'
\z 
 
\ea \label{ŽP.100}
\textit{her mēnē mēnē mēnē nizīkū hejījī gināwe} \\ 
\gll her m-ē-nē m-ē-nē m-ē-nē nizīk=ū hejīj-ī ∅-gin-ā=we \\ 
 emph \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{post}} \\ 
\glt ` They kept walking. They arrived [at a place] close to Hajij.'
\z 
 
\ea \label{ŽP.101}
\textit{nizīkū hejījī gināwe miqdārē zeřšā pene bo} \\ 
\gll nizīk=ū hejīj-ī ∅-gin-ā=we miqdārē zeř=šā pene ∅-b-o \\ 
 close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3pl}\textsc{=compl} a\_little money\textsc{.m}\textsc{=3pl} with \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `They\footnote{Lit. He.} arrived [at a place] close to Hajij. They had some money with them\footnote{Lit. him.}.'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.107}
\textit{be zor miđoš milū sey misefāyre zerey} \\ 
\gll be zor mi-đe-o=š mil=ū sey misefā-ī=re zer-e=y \\ 
 by force \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=3sg} on\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} money\textsc{.m}\textsc{-def}\textsc{=ez}.\textsc{attr} \\ 
\glt `He (the man from Hajij) did not take no for an answer\footnote{Lit. with force.} [and] gave him (Pir Shaliyar) the money.'
\z 
 
\ea \label{ŽP.113}
\textit{miqdārēš āđī be sey misefāy ārdeniš} \\ 
\gll miqdārē=š āđī be sey misefā-ī ārde=n=iš \\ 
 a\_little\textsc{=3sg} 3sg\textsc{.obl}\textsc{.m} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \\ 
\glt `The money which he had made Say Mustafa take.'
\z 
 
\ea \label{ŽP.116}
\textit{mēwe ētir ā zemāne pīr šelīyār mēwe menteqe des kero, jārē āne minyā māčāš meřū āsinī} \\ 
\gll m-ē=we ētir ā zemān=e pīr šelīyār m-ē=we menteqe des=t ∅-ker-o jārē āne mi-nye-ā m-āč-ā=š meře=ū āsin-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} once \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) returned [to Hawraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ‘Iron cave.’'
\z 
 
\ea \label{ŽP.117}
\textit{ētir hezretū xosī kīyāseno} \\ 
\gll ētir hezret=ū xos-ī kīyāse=n=o \\ 
 \textsc{disc.ptcl} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `Anyhow, His Highness Khos had sent him.'
\z 
 
\ea \label{ŽP.120}
\textit{čilexānēš bīyenū} \\ 
\gll čilexāne-ē=š bīye=n=\\ 
 \textsc{pn}\textsc{-indf}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `It (the cave) was a place for forty days of fasting.'
\z 
 
\ea \label{ŽP.122}
\textit{lūān pāsārū leme čilēš kēštēneū} \\ 
\gll lūā=n pāsār=ū leme čilē=š kēštē=ne=ū \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{\textsc{=ez.gen}} \textsc{pn} forty\_days\_fasting\textsc{.f}\textsc{=3sg} pull\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}}=and \\ 
\glt `He would go to [a place called] Pasaru Lama ‘guardians of the belly’ and would fast for forty days,'
\z 
 
\ea \label{ŽP.123}
\textit{ā čilexāne čilexāne lemē} \\ 
\gll ā čilexāne čilexāne lemē \\ 
 \textsc{dem.dist} \textsc{pn} \textsc{pn} \textsc{pn}\textsc{.obl}\textsc{.f} \\ 
\glt `[he would fast] in that place of fasting [called] Chilaxana Lame.'
\z 
 
\ea \label{ŽP.124}
\textit{ā wextī xeberē yāwno hezretū xosī} \\ 
\gll ā wext-ī xeber-ē ∅-yāwn-o hezret=ū xos-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} news\textsc{.m}\textsc{-indf} \textsc{ind-}cause\_to\_arrive\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, the news reached His Highness Khos.'
\z 
 
\ea \label{ŽP.126}
\textit{māčā, qurbān to desūrēt dā be sey misefāy bilowe šerēʕetū terēqet ders wāčo} \\ 
\gll m-āč-ā qurbān to desūr-ē=t dā be sey misefā-ī bi-l-o=we šerēʕet=ū terēqet ders ∅-wāč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} \textsc{2sg} task/order\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{hort} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} sharia\_law=and denomination lesson\textsc{.m} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} \\ 
\glt `They (the gossips) said, ‘Sir, you ordered Say Mustafa to go back [to Hawraman and] teach Sharia law and [the ways of] orders (of dervishes).'
\z 
 
\ea \label{ŽP.129}
\textit{hezretū xosī be mexfī pīyāy kīyāno menteqew hewrāmānātī} \\ 
\gll hezret=ū xos-ī be mexfī pīyā-ī ∅-kīyān-o menteqe=ū hewrāmān-āt-ī \\ 
 his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} by secret man\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3sg} region=and \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} \\ 
\glt `His Highness Khos secretly sent a man to the Hawraman region.'
\z 
 
\ea \label{ŽP.132}
\textit{pīyākew hezretū xosī milo meřū āsinī tūšiš mē} \\ 
\gll pīyā-eke=ū hezret=ū xos-ī mi-l-o meře=ū āsin-ī tūš=iš m-ē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} encounter\textsc{=3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \\ 
\glt `His Highness Khos’s man went [and] ran into him (Pir Shaliyar) in the Iron cave.'
\z 
 
\ea \label{ŽP.134}
\textit{xo āđ mezāno hezretū xosī kīyāseniš} \\ 
\gll xo āđ me-zān-o hezret=ū xos-ī kīyāse=n=iš \\ 
 dis.ptc 3sg\textsc{.dir}\textsc{.m} \textsc{neg-}know\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \\ 
\glt `Well, he (Pir Shaliyar) did not realise that His Highness Khos had sent him (the man)'
\z 
 
\ea \label{ŽP.141}
\textit{mē mēwe menteqew mēwe dilū hewrāmānīū yānešā yānešā e čā bo milo wā} \\ 
\gll m-ē m-ē=we menteqe=ū m-ē=we dil=ū hewrāmān-ī=ū yāne=šā yāne=šā e=čā ∅-b-o mi-l-o wā \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region=and \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} inside\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{\textsc{=ez.gen}} house\textsc{.m}\textsc{=3pl} house\textsc{.m}\textsc{=3pl} in=there \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} there \\ 
\glt `He (Pir Shaliyar) returned to the region, returned to Hawraman. His\footnote{Lit. their.} house was there.'
\z 
 
\ea \label{ŽP.146}
\textit{pādšāū mīsrī kināčēš bo feleje bo} \\ 
\gll pādšā=ū mīsr-ī kināčē-ē=š ∅-b-o felej-e ∅-b-o \\ 
 king\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} daughter\textsc{.f}\textsc{-indf}\textsc{=3sg} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} disabled\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The king of Egypt had a daughter [who] was disabled.'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.157}
\textit{pādšāyč māčo be lālo kināčekē be šālyāre sīyāwīš} \\ 
\gll pādšā=yč m-āč-o be lālo kināčē-ekē be šālyār=e sīyāw-ī=š \\ 
 king\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f} by \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `The king said to [his] daughter’s uncle, to Shaliyar Syaw'
\z 
 
\ea \label{ŽP.158}
\textit{lālošā, lālo heremū pīr šelyārīn} \\ 
\gll lālo=š=ā lālo hereme=ū pīr šelyār-ī=n \\ 
 maternal\_uncle\textsc{.m}\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} maternal\_uncle\textsc{.m} wife\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `—He was her uncle. He was the uncle of Pir Shaliyar’s wife—'
\z 
 
\ea \label{ŽP.166}
\textit{lālo gino gelū kināčē māroš dey degā be degāw kēšwer be kēšwerū māroš silēne} \\ 
\gll lālo ∅-gin-o gel=ū kināčē m-ār-o=š dey degā be degā=ū kēšwer be kēšwer=ū m-ār-o=š silēne \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} with\textsc{\textsc{=ez.gen}} girl\textsc{.dir}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{disc.ptcl} village\textsc{.obl}\textsc{.f} by village\textsc{.f}=and country by country=and \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{pn} \\ 
\glt `The uncle set off on the road with the girl. He took her. He took her village by village and country by country until [they arrived] at Slen.'
\z 
 
\ea \label{ŽP.167}
\textit{silēne řeđē bā tewenē čā beynne bē beynū silēnē biɫberīne bē} \\ 
\gll silēne řeđ-ē bā tewenī-ē č=ā beyn=ne b-ē beyn=ū silēnē biɫber-ī=ne b-ē \\ 
 \textsc{pn} crossing\textsc{-dir}\textsc{.pl} \textsc{hort} stone\textsc{.f}\textsc{-indf} at=\textsc{dem.dist} between\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} between\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `They passed Slen. There was a stone in the road between Slen and Bilbar.'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.182}
\textit{pīr šelīyār tewejoekey kero} \\ 
\gll pīr šelīyār tewejo-ekey ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} caution\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar paid attention to her.'
\z 
 
\ea \label{ŽP.186}
\textit{mēnē, mēnē pēse māčmē biɫbeřū hurwezā milegāw} \\ 
\gll m-ē-nē m-ē-nē pēse m-āč-mē biɫbeř=ū hur-wez-ā milegā=ū \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} as\_if \textsc{ind-}say\textsc{.prs}\textsc{-1pl} \textsc{pn}=and \textsc{pvb-}climb\textsc{.prs}\textsc{-3pl} top\_of\_mountain=and \\ 
\glt `They (the girl and her uncle) came [to the direction of Hawraman]; they climbed a peak of the mountain'
\z 
 
\ea \label{ŽP.187}
\textit{milegā ā wextī mizyā werāwerū dūtifɫey} \\ 
\gll milegā ā wext-ī mi-zye-ā werāwer=ū dūtifɫey \\ 
 top\_of\_mountain \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go\_out\textsc{.prs}\textsc{-3pl} in\_front\_of\textsc{\textsc{=ez.gen}} \textsc{pn} \\ 
\glt `Then they came out in front of  Dutifla.'
\z 
 
\ea \label{ŽP.188}
\textit{pīr šelīyār je yānekew wēšāne} \\ 
\gll pīr šelīyār je yāne-eke=ū wē=šā=ne \\ 
 \textsc{pn} \textsc{pn} from house\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `Pir Shaliyar [was] in his\footnote{Lit. their.} house.'
\z 
 
\ea \label{ŽP.189}
\textit{hezretū sey misefāy čā miđyo xo kes ter mewīnoš} \\ 
\gll hezret=ū sey misefā-ī čā mi-đye-o xo kes ter me-wīn-o=š \\ 
 his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} there \textsc{ind-}look\textsc{.prs}\textsc{-3sg} dis.ptc person\textsc{.m} other \textsc{neg-}see\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \\ 
\glt `His Highness Say Mustafa watched [them] from above. No one else could see her.'
\z 
 
\ea \label{ŽP.191}
\textit{miđyo dūtifɫene duē neferē āmē} \\ 
\gll mi-đye-o dūtifɫe=ne duē nefer-ē āmā-ē \\ 
 \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{=\textsc{post}} two person\textsc{.m}\textsc{-dir}\textsc{.pl} come\textsc{.pst}\textsc{-3pl} \\ 
\glt `He saw that two people were coming from Dutifla.'
\z 
 
\ea \label{ŽP.201}
\textit{tenūrū dēwā ētir pēse gāw til bowe e čā werāwerū girdū mexeleqeɫāw hewrāmānīwe dēweke til bowe ginore soɫāwe} \\ 
\gll tenūr=ū dēw-ā ētir pēse gāw til ∅-b-o=we e=čā werāwer=ū gird=ū mexeleqeɫā=ū hewrāmān-ī=we dēw-eke til ∅-b-o=we ∅-gin-o=re soɫāwe \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{disc.ptcl} as\_if bull\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=there in\_front\_of\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt ` Then it was as if a cow had rolled down; the [demon] rolled down and fell [into a place called] Solawa before all the people’s eyes.'
\z 
 
\ea \label{ŽP.203}
\textit{kābrā lālow šābāre xātūnī mē dilū degēw} \\ 
\gll kābrā lālo=ū šābāre xātūn-ī m-ē dil=ū degē=ū \\ 
 man maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} lady\textsc{-obl}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} inside\textsc{\textsc{=ez.gen}} village\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `The fellow, the uncle of Shahbara Khatun, went inside the village.'
\z 
 
\ea \label{ŽP.205}
\textit{muxteserū kelāmī, yāney yozāwew milā berāš xizmetū pīrīyū} \\ 
\gll muxteser=ū kelām-ī yāne-ī ∅-yoz-ā=we=ū mi-l-ā ∅-ber-ā=š xizmet=ū pīr-ī=ū \\ 
 summary\textsc{.m}\textsc{\textsc{=ez.gen}} speech\textsc{-obl}\textsc{.m} house\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}find\textsc{.prs}\textsc{-3pl}\textsc{=compl}\textsc{\textsc{=ez.gen}} \textsc{ind-}go\textsc{.prs}\textsc{-3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3sg} service\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}=and \\ 
\glt ` To cut a long story short,\footnote{Lit. Gist of speech} they found the house. They (the people) went [and] took him to the presence of Pir [Shaliyar]'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.214}
\textit{min āne nānmā xeplezeřātīnen} \\ 
\gll min āne nān=m=ā xeplezeřātīne=n \\ 
 \textsc{1sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Look, my meal is xaplazaratine.’'
\z 
 
\ea \label{ŽP.241}
\textit{čāštekē hiɫošīnekē kermē} \\ 
\gll čāštī-ekē hiɫošīne-ekē ∅-ker-mē \\ 
 meal\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{pn}\textsc{-def}\textsc{.obl}\textsc{.f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \\ 
\glt `Let us make the meal called Hiloshina.'
\z 
 
\ea \label{ŽP.243}
\textit{pīr šelīyār fermāwo, tehāā weɫāhī īne šime desūrtā dāw wāttā čūnke āđē be diɫtā wāt řāstā wāt hetā xwāy dinyāš weše kerdēne her menowe kemīč mekero her zīyā kero} \\ 
\gll pīr šelīyār ∅-fermāw-o tehāā weɫāhī īne šime desūr=tā dā=ū wāt=tā čūnke āđē be diɫ=tā wāt řās=tā wāt hetā xwā-ī dinyā=š weš-e kerdē=ne her ∅-men-o=we kem=īč me-ker-o her zīyā ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} task/order\textsc{.m}=\textsc{2pl} give\textsc{.pst}=and say\textsc{.pst}=\textsc{2pl} since 3pl\textsc{.dir} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} truth=\textsc{2pl} say\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} small\textsc{=add} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar said, ‘Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] continues to exist; it will not be diminished; it will grow.'
\z 
 
\ea \label{ŽP.250}
\textit{kīyānāš kābrāyē serūpīrī qesāb bo} \\ 
\gll ∅-kīyān-ā=š kābrā-ē serūpīrī qesāb ∅-b-o \\ 
 \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} man\textsc{-indf} \textsc{pn} butcher\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `They sent for a fellow from Saru Piri, who was a butcher. They fetched him.'
\z 
 
\ea \label{ŽP.251}
\textit{kīyānāš serūpīrīwe qesābē bo māčāš tāyfew pādšāy} \\ 
\gll ∅-kīyān-ā=š serūpīrī=we qesāb-ē ∅-b-o m-āč-ā=š tāyfe=ū pādšā-ī \\ 
 \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg} \textsc{pn}\textsc{=\textsc{post}} butcher\textsc{.m}\textsc{-indf} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `They sent [one] to Saru Piri; there was a butcher who was from the king’s tribe.'
\z 
 
\ea \label{ŽP.256}
\textit{merāsēmekew pīr šelīyārī bergozār kerāū} \\ 
\gll merāsēm-eke=ū pīr šelīyār-ī bergozār ∅-ker-ā=ū \\ 
 ceremony\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} holding \textsc{ind-}do\textsc{.prs}\textsc{-3pl}=and \\ 
\glt `They held Pir Shaliyar’s ceremony.'
\z 
 
\ea \label{ŽP.257}
\textit{ženī be mezebū hewrāmānī māre biřā pey pīr šelīyārīū} \\ 
\gll ženī be mezeb=ū hewrāmān-ī māre ∅-biř-ā pey pīr šelīyār-ī=ū \\ 
 woman\textsc{.f} by religion\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} marraige \textsc{ind-}cut\textsc{.prs}\textsc{-3pl} to \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \\ 
\glt `They married the woman to Pir Shaliyar] according to] the religious customs in Hawraman.'
\z 
 
\ea \label{ŽP.258}
\textit{ā wextīyekey lālo pīr šelīyār hīn lālo heremū pīr šelīyārī milowe xeber berowe pey pādšāy} \\ 
\gll ā wext-ī-ekey lālo pīr šelīyār hīn lālo hereme=ū pīr šelīyār-ī mi-l-o=we xeber ∅-ber-o=we pey pādšā-ī \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} maternal\_uncle\textsc{.m} \textsc{pn} \textsc{pn} \textsc{fill} maternal\_uncle\textsc{.m} wife\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{ind-}take\textsc{.prs}\textsc{-3sg}\textsc{=compl} to king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Then, Pir Shaliyar’s uncle, eh, Pir Shaliyar’s wife’s uncle, went back [to Egypt] and passed the news to the king.'
\z 
 
\ea \label{ŽP.267}
\textit{ānen īse šelīyāre sīyāwā} \\ 
\gll āne=n īse šelīyār=e sīyāw=ā \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} now \textsc{pn}\textsc{=ez}\textsc{.cmpd} black\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[His shrine] is over there now; [he is called] Shaliyar Siyaw.'
\z 
 
\ea \label{BP.1}
\textit{zemānū bābāw pīr šelīyārī, heke, nāmēš bābā xwāđā bīyen} \\ 
\gll zemān=ū bābā=ū pīr šelīyār-ī heke nāmē=š bābā xwāđā bīye=n \\ 
 time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} well name.f\textsc{=3sg} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the time of Pir Shaliyar’s grandfather —his name was Baba Khwada.'
\z 
 
\ea \label{BP.2}
\textit{bābāw pīr šelīyārī, nāmēš bābā xwāđā bīyen} \\ 
\gll bābā=ū pīr šelīyār-ī nāmē=š bābā xwāđā bīye=n \\ 
 grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} name.f\textsc{=3sg} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s grandfather, his name was Baba Khwada\footnote{Lit. Grandfather Khwada}. '
\z 
 
\ea \label{BP.5}
\textit{pīr bāyiɫiš peyā bīyen, pīr bāyiɫ seyīē bēɫā} \\ 
\gll pīr bāyiɫ=iš peyā bīye=n pīr bāyiɫ seyīē bēɫ=ā \\ 
 \textsc{pn} \textsc{pn}\textsc{=3sg} visible be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \textsc{pn} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `His (the son’s) name was Pir Bayil. Pir Bayil’s [real name] is Say Bel.'
\z 
 
\ea \label{BP.6}
\textit{pīr bāyiɫ seyīē bēɫā} \\ 
\gll pīr bāyiɫ seyīē bēɫ=ā \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Bayil’s [real name] is Say Bel.'
\z 
 
\ea \label{BP.7}
\textit{tātew pīr šelīyārīn} \\ 
\gll tāte=ū pīr šelīyār-ī=n \\ 
 father\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is Pir Shaliyar’s father.'
\z 
 
\ea \label{BP.8}
\textit{tātew pīr šelīyārīn} \\ 
\gll tāte=ū pīr šelīyār-ī=n \\ 
 father\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He is Pir Shaliyar’s father.'
\z 
 
\ea \label{BP.9}
\textit{ā zemān bābā xwāđā bīyen, ā zemān ī mexlūqū hewrāmānātū ēme reʕīsšā je ʕērāqew čā kurdewārīū ʕērāqīne bīyen} \\ 
\gll ā zemān bābā xwāđā bīye=n ā zemān ī mexlūq=ū hewrāmān-āt=ū ēme reʕīs=šā je ʕērāq=ew čā kurdewārī=ū ʕērāq-ī=ne bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} time\textsc{.m} \textsc{dem.prox} people\textsc{=ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{=ez.gen} \textsc{1pl} leader\textsc{.m}\textsc{=3pl} from \textsc{pn}\textsc{=\textsc{post}} there kurdish\_region\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Babab Khwada, the people of Hawraman were under Iraq’s rule, under the control of Kurdish regions in Iraq.'
\z 
 
\ea \label{BP.11}
\textit{zemānū bābāw pīr šelīyārī bābā xwāđāy āā je ʕērāqne ēme hukmū hewrāmānīšā kerden, biřew bārešā sānān} \\ 
\gll zemān=ū bābā=ū pīr šelīyār-ī bābā xwāđā-ī āā je ʕērāq=ne ēme hukm=ū hewrāmān-ī=šā kerde=n biře=ū bāre=šā sānā=n \\ 
 time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{intj} from \textsc{pn}\textsc{=\textsc{post}} \textsc{1pl} rule\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Pir Shaliyar’s grandfather, we were ruled by Iraq, [and] they would collect tax.'
\z 
 
\ea \label{BP.18}
\textit{eger ženīmā dē be bešepīyākew yo terī čīwēš zereriš sānē čene wāčēnēš sūrāne} \\ 
\gll eger ženī=mā de-ē be bešepīyā-eke=ū yo ter-ī čīw-ē=š zerer=iš sān-ē čene wāč-ēn-ē=š sūrāne \\ 
 if woman.f\textsc{=\textsc{1pl}} give.prs\textsc{-cond} to quota\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} thing\textsc{.m}\textsc{-indf}\textsc{=3sg} harm\textsc{.m}\textsc{=3sg} buy.prs\textsc{-pstc}\textsc{.3sg} from say.prs.\textsc{sbjv-}pstc\textsc{-3pl}\textsc{=3sg} \textsc{pn} \\ 
\glt `If one\footnote{Lit. we.} had married women from the quota of another [lord], then they [the lord of the bride's village] would charge him [the bride’s father]. This money is called Surana.'
\z 
 
\ea \label{BP.20}
\textit{jā ī dāstānē hinū bābāw pīr šelīyārīne} \\ 
\gll jā ī dāstāne-ē hin=ū bābā=ū pīr šelīyār-ī=ne \\ 
 afterwards \textsc{dem.prox} story.f\textsc{-obl}.f ez.pron\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `Anyhow, this story is about the grandfather of Pir Shaliyar.'
\z 
 
\ea \label{BP.21}
\textit{bābāw pīr šelīyārī hukmū bābāw pīr šelīyārī ā zemāne hewrāmān bīyen je ʕērāqo āmēnē hukimšā kerden} \\ 
\gll bābā=ū pīr šelīyār-ī hukm=ū bābā=ū pīr šelīyār-ī ā zemān=e hewrāmān bīye=n je ʕērāq=o āmē=nē hukim=šā kerde=n \\ 
 grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} rule\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} rule\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Shaliyar’s grandfather, at the time of his religious rule, Hawraman was under the rule of Iraq.'
\z 
 
\ea \label{BP.24}
\textit{e čē biřew bārešā sānān je hewrāmānī pēse begzāyekāw ī beynū ēme} \\ 
\gll e=čē biře=ū bāre=šā sānā=n je hewrāmān-ī pēse begzāye-ekā=ū ī beyn=ū ēme \\ 
 in=here portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{-obl}\textsc{.m} like nobleman\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=ez.gen} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `[who] would collect taxes from Hawraman, [just] like our noblemen [in recent] times.'
\z 
 
\ea \label{BP.27}
\textit{jā luēnē berdenšāwe, luēnē berdenšāwe pey āġekew ʕērāqīšā} \\ 
\gll jā luē=nē berde=n=šā=we luē=nē berde=n=šā=we pey āġe-eke=ū ʕērāq-ī=šā \\ 
 afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} for agha\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=3pl} \\ 
\glt `Then, they would go back and take the taxes to their aghas in Iraq.'
\z 
 
\ea \label{BP.28}
\textit{āġekew ʕērāqī berdenšāwe peyšāw} \\ 
\gll āġe-eke=ū ʕērāq-ī berde=n=šā=we pey=šā=ū \\ 
 agha\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=compl} to\textsc{=3pl}=and \\ 
\glt `The aghas in Iraq, they took back [the taxes] to them.'
\z 
 
\ea \label{BP.31}
\textit{be enāzew wēš biřew bāre misānā, zemānū bābāw pīr šelīyārī, bābā xwāđāy} \\ 
\gll be enāze=ū wē=š biře=ū bāre mi-sān-ā zemān=ū bābā=ū pīr šelīyār-ī bābā xwāđā-ī \\ 
 to proportion\textsc{=ez.gen} \textsc{reflx}\textsc{=3sg} portion\textsc{.m}=and profit\textsc{.m} \textsc{ind-}buy.prs\textsc{-3pl} time\textsc{.m}\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `During the time of Pir Shaliyar’s grandfather, i.e. Baba Khwada, they (the tax collectors) collected tax in proportion to one’s wealth. '
\z 
 
\ea \label{BP.38}
\textit{ī kābrēče je ʕērāqo āmēnē, wēregānew nīmeřonew seʕbne bexšnāyšā bexšnēnēwe} \\ 
\gll ī kābrā-ē=č=e je ʕērāq=o āmē=nē wēregā=ne=ū nīmeřo=ne=ū seʕbe=ne bexšnāy=šā bexšn-ēn-ē=we \\ 
 \textsc{dem.prox} man\textsc{-dir}\textsc{.pl}\textsc{=add}\textsc{=dem} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} evening\textsc{=\textsc{post}}=and noon\textsc{=\textsc{post}}=and morning.f\textsc{=\textsc{post}} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-indf}\textsc{=compl} \\ 
\glt `They (people) would donate [food] to the fellows (the tax collectors) who had come from Iraq, in the evenings, mornings, and at noon. '
\z 
 
\ea \label{BP.40}
\textit{her wēregāne pēse čāweɫ sempāšē bēnē, yoīšā dēnē duīšā dēnē yānew yoy ī meʕmūrīye ke ʕērāqo āmēnē, ā zemānū pīr šelīyārī bexšnāyšā bexšnēnē} \\ 
\gll her wēregā=ne pēse čāweɫ sempāš-ē b-ēn-ē yo-ī=šā de-ēn-ē duē-ī=šā de-ēn-ē yāne=ū yo-ī ī meʕmūr-ī=e ke ʕērāq=o āmē=nē ā zemān=ū pīr šelīyār-ī bexšnāy=šā bexšn-ēn-ē \\ 
 just evening\textsc{=\textsc{post}} like in\_the\_past poison\_sprayer\textsc{-dir}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} two\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} house\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} officer\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=dem} rel \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `As in the old days when there were people whose job was spraying pesticide on trees, [and] each would be hosted at the house of one [host], the officers who had come from Iraq would be fed in the time of Pir Shaliyar. '
\z 
 
\ea \label{BP.46}
\textit{xeber yāwo be bābā xwāđāw ħemey ġeybī} \\ 
\gll xeber ∅-yāw-o be bābā xwāđā=ū ħeme=ī ġeybī \\ 
 news\textsc{.dir}\textsc{.m} \textsc{ind-}arrive.prs\textsc{-3sg} to grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible \\ 
\glt `The news reached Baba Khwada and Hama the invisible.'
\z 
 
\ea \label{BP.51}
\textit{ħemey ġeybīw bābāw pīr šelīyārī, bābā xwāđā māčā be xeɫkekey,} \\ 
\gll ħeme=ī ġeybī=ū bābā=ū pīr šelīyār-ī bābā xwāđā m-āč-ā be xeɫk-ekey \\ 
 \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3pl} to people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Hama Ghaybi and Pir Shaliyar’s grandfather, i.e., Baba Khwada, said to the people, '
\z 
 
\ea \label{BP.56}
\textit{jā girđmā bilmē pey ʕērāqī pey lāw ī āġeyšā bilmērewo,} \\ 
\gll jā girđ=mā bi-l-mē pey ʕērāq-ī pey lā=ū ī āġe-ī=šā bi-l-mē=re=o \\ 
 afterwards all\textsc{=\textsc{1pl}} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} for at\_the\_place\_of\textsc{=ez.gen} \textsc{dem.prox} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}\textsc{=compl} \\ 
\glt `Then, we all go to Iraq, to their agha. We go down there.'
\z 
 
\ea \label{BP.57}
\textit{āđ māčo, konē ī meʕmūrē mine činnē wextēn? je hewrāmān bīyēnē āmēnē pey biřew bāre} \\ 
\gll āđ m-āč-o ko=nē ī meʕmūr-ē min=e činnē wext-ē=n je hewrāmān bīyē=nē āmē=nē pey biře=ū bāre \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} where\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1sg}\textsc{=dem} some\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} for portion\textsc{.m}=and profit\textsc{.m} \\ 
\glt `He (the agha) says, ‘where have my officers been during this time? They were in Hawraman; they had gone to collect tax.'
\z 
 
\ea \label{BP.62}
\textit{je hewrāmānne kārī nāšerʕī kerēnē.’} \\ 
\gll je hewrāmān=ne kār=ī nāšerʕī ker-ēn-ē \\ 
 from \textsc{pn}\textsc{=\textsc{post}} task\textsc{.m}=ez.\textsc{attr} unlawful do.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `They committed unlawful acts in Hawraman.’'
\z 
 
\ea \label{BP.65}
\textit{qewɫ kerā yo, wēregā bābāw pīr šelīyārī xeberšā pene miđo} \\ 
\gll qewɫ ∅-ker-ā yo wēregā bābā=ū pīr šelīyār-ī xeber=šā pene mi-đe-o \\ 
 arrangement\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} one\textsc{.m} evening grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} news\textsc{.m}\textsc{=3pl} to \textsc{ind-}give.prs\textsc{-3sg} \\ 
\glt `They all agreed [about the plan]. In the evening Pir Shaliyar’s grandfather informed them'
\z 
 
\ea \label{BP.78}
\textit{wāče, bē dērā ēme milāy milmē pey ʕērāqī be peyē} \\ 
\gll wāč-e b-ē dēr=ā ēme mi-lāy mi-l-mē pey ʕērāq-ī be peyē \\ 
 say.prs-\textsc{imp-}\textsc{2sg} be.prs\textsc{-pstc}\textsc{.3sg} late\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} \textsc{ind-}go\textsc{.adv} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} to foot\textsc{.obl} \\ 
\glt `Tell him, “Come, it’s getting late. We are going to Iraq on foot.'
\z 
 
\ea \label{BP.79}
\textit{milāymē pey ʕērāqī milāymē ī xeberī bermēwe} \\ 
\gll mi-lā-ymē pey ʕērāq-ī mi-lā-ymē ī xeber-ī ∅-ber-mē=we \\ 
 \textsc{ind-}go.prs \textsc{1pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs \textsc{1pl} \textsc{dem.prox} news\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}take.prs\textsc{-\textsc{1pl}}\textsc{=compl} \\ 
\glt `We are going to Iraq. We are going [there] to pass this news [to the agha in Iraq]”'
\z 
 
\ea \label{BP.87}
\textit{jārēne jimdēre mexɫūq hečē henē mizgīne bā bilimdē pey ʕērāqī} \\ 
\gll jār-ē=ne ∅-jim-dē=re mexɫūq hečē hen-ē mizgī=ne ∅-b-ā bi-l-imdē pey ʕērāq-ī \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} people\textsc{.m} no\_matter\_how\_many exist\textsc{-3pl} mosque\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}/\textsc{2pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They all set off; every person who was in the mosque [said], ‘Let’s go to Iraq.'
\z 
 
\ea \label{BP.88}
\textit{bilmēre wā ʕērāqī} \\ 
\gll bi-l-mē=re wā ʕērāq-ī \\ 
 \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} there \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Let’s go there, to Iraq. '
\z 
 
\ea \label{BP.98}
\textit{ā wextī ēti seʕbe milā milāne mizgī, mizyāre beynē gināre řā beynē bilā pey ʕērāqī jemʕē bā ħemey ġeybīū bābā xuāđā ginā weɫēwe} \\ 
\gll ā wext-ī ēti seʕbe mi-l-ā mi-l-ān=e mizgī mi- zye-ā=re beynē ∅-gin-ā=re řā beynē bi-l-ā pey ʕērāq-ī jemʕ-ē ∅-b-ā ħeme=ī ġeybī=ū bābā xuāđā ∅-gin-ā weɫē=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} morning.f \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}go.prs\textsc{-3pl}=drct mosque\textsc{.m} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} a\_while \textsc{ind-}fall.prs\textsc{-3pl}\textsc{=\textsc{povb}} road.f a\_while \textsc{sbjv-}go.prs\textsc{-3pl} for \textsc{pn}\textsc{-obl}\textsc{.m} added\textsc{-pstc}\textsc{.3sg} \textsc{ind-}be.prs\textsc{-3pl} \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m} \textsc{pn} \textsc{ind-}fall.prs\textsc{-3pl} front\textsc{=\textsc{post}} \\ 
\glt `So, in the morning they went to the mosque. They went a little way towards Iraq. They all gathered. Hama Ghaybi and Baba Khwada went in front of them.'
\z 
 
\ea \label{BP.101}
\textit{māčā ēme čī bilmēre ʕērāq wēmā kermē zīnānū} \\ 
\gll m-āč-ā ēme čī bi-l-mē=re ʕērāq wē=mā ∅-ker-mē zīnān=ū \\ 
 \textsc{ind-}say.prs\textsc{-3pl} \textsc{1pl} why \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} \textsc{pn} \textsc{reflx}\textsc{=\textsc{1pl}} \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}} prison\textsc{.m}=and \\ 
\glt `They (the people) said, ‘Why should we go to Iraq and put ourselves in   prison? '
\z 
 
\ea \label{BP.105}
\textit{bābā xwāđāw ħemey ġeybīū ħemey wirdīkɫe her ginēnē weɫē īnē weɫēne ejošā girđ īnē šūnīšāwe} \\ 
\gll bābā xwāđā=ū ħeme=ī ġeybī=ū ħeme=ī wirdīkɫe her gin-ēn-ē weɫē īnē weɫē=ne ejo=šā girđ īnē šūnī=šā=we \\ 
 grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible=and \textsc{pn}=ez.\textsc{attr} small just fall.prs\textsc{-pstc}\textsc{-3pl} front \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} front\textsc{=\textsc{post}} guess.prs\textsc{=3pl} all \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} track.f\textsc{=3pl}\textsc{=compl} \\ 
\glt `Baba Khwada, Hama the invisible, and little Hama were in front, assuming that everyone was following them.'
\z 
 
\ea \label{BP.106}
\textit{wextē milā serū meřekēw meře beřī mēyāwe dimāwe} \\ 
\gll wext-ē mi-l-ā ser=ū meře-ekē=w meře beř-ī m-ē-ā=we dimā=we \\ 
 time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3pl} on\textsc{=ez.gen} cave.f\textsc{-def}\textsc{.obl}.f\textsc{=ez.gen} cave.f \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{-3pl}\textsc{=compl} back\textsc{=\textsc{post}} \\ 
\glt `When they reached the cave of Marabar, they looked behind [them].'
\z 
 
\ea \label{BP.113}
\textit{fermāwo ħemey ġeybī, āy wiɫāhī hewrāmī tā meřeber} \\ 
\gll ∅-fermāw-o ħeme=ī ġeybī āy wiɫāhī hewrāmī tā meřeber \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{pn}=ez.\textsc{attr} invisible \textsc{intj} indeed Hawrami until \textsc{pn} \\ 
\glt `Hama Ghaybi said, ‘Well, By God, hawrāmī tā mařabar'
\z 
 
\ea \label{BP.114}
\textit{tā meřeber āmē girđ gēɫēnēwe} \\ 
\gll tā meřeber āmā-ē girđ gēɫē=nē=we \\ 
 until \textsc{pn} come\textsc{.pst}\textsc{-3pl} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} \\ 
\glt `[which means] they came as far as Marabar. [Then] they returned.'
\z 
 
\ea \label{BP.115}
\textit{hewrāmī tā meřeber yānē hewrāmīyē tā meřeber āmē ētir girđ gēɫēnēwe čāge} \\ 
\gll hewrāmī tā meřeber yānē hewrāmī-ē tā meřeber āmā-ē ētir girđ gēɫē=nē=we č=āge \\ 
 Hawrami until \textsc{pn} that\_means Hawrami\textsc{-dir}\textsc{.pl} until \textsc{pn} come\textsc{.pst}\textsc{-3pl} \textsc{disc.ptcl} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} from=there \\ 
\glt `hawrāmī tā mařabar means the people of Hawraman came as far as Marabar. Then, they all returned to Hawraman from there.'
\z 
 
\ea \label{BP.120}
\textit{jereyāniš kermē pey pādšāw ʕērāqī} \\ 
\gll jereyān=iš ∅-ker-mē pey pādšā=w ʕērāq-ī \\ 
 story\textsc{.m}\textsc{=3sg} \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}} for king\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We will tell the king the story.'
\z 
 
\ea \label{BP.122}
\textit{ā wextī reʕīsū milāre ʕērāq lāw āġekeyšā māčo seremā biřyēnē bābāw pīr šelīyārī bo, ā wextīyekey ħemey ġeybīū yoyč bo nāmē ħemey wirdīkɫe,} \\ 
\gll ā wext-ī reʕīs=ū mi-l-ā=re ʕērāq lā=ū āġe-ekey=šā m-āč-o sere=mā biřyē=nē bābā=ū pīr šelīyār-ī ∅-b-o ā wext-ī-ekey ħeme=ī ġeybī=ū yo=yč ∅-b-o nāmē ħeme=ī wirdīkɫe \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} leader\textsc{.m}\textsc{=ez.gen} \textsc{ind-}go.prs\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{pn} at\_the\_place\_of\textsc{=ez.gen} agha\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}say.prs\textsc{-3sg} head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}=ez.\textsc{attr} invisible=and one\textsc{.m}\textsc{=add} \textsc{ind-}be.prs\textsc{-3sg} name.f \textsc{pn}=ez.\textsc{attr} small \\ 
\glt `The head of those who went to Iraq to the agha, saying, ‘We have decapitated [your men], was Pir Shaliyar’s grandfather, Baba Khwada. In addition [there were] Hama the invisible and Little Hama. '
\z 
 
\ea \label{BP.123}
\textit{ħemey wirdīkɫe ā wextī, pēse īsey māčā heɫāy neyāwān sinʕe zīnāniš nekermē yā čīw yā čāne} \\ 
\gll ħeme=ī wirdīkɫe ā wext-ī pēse īse-ī m-āč-ā heɫāy ne-yāwā=n sinʕe zīnān=iš ne-ker-mē yā čīw yā č=ā=ne \\ 
 \textsc{pn}=ez.\textsc{attr} small \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} like now\textsc{-obl}\textsc{.m} \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} age.f prison\textsc{.m}\textsc{=3sg} \textsc{neg-}do.prs\textsc{-\textsc{1pl}} either thing\textsc{.m} either in=\textsc{dem.dist}\textsc{=\textsc{post}} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) will say, ‘Hama Wirdikla has not reached adulthood; we won’t imprison him, and so on.’'
\z 
 
\ea \label{BP.131}
\textit{bābā xuāđāw ħemew ġeybī, āđē, āđīšā kerā zīnān’} \\ 
\gll bābā xuāđā=ū ħeme=ū ġeybī āđē āđīšā ∅-ker-ā zīnān \\ 
 grandfather\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez.gen} invisible 3pl\textsc{.dir} 3pl\textsc{.obl} \textsc{ind-}do.prs\textsc{-3pl} prison\textsc{.m} \\ 
\glt `Baba Khwada and Hama Ghaybi, they (agha’s men) put them  in prison.'
\z 
 
\ea \label{BP.132}
\textit{ħemey wirdīkɫeyč pēse īse māčā, heɫāy neyāwān bilūqū sinʕēš} \\ 
\gll ħeme=ī wirdīkɫe=yč pēse īse m-āč-ā heɫāy ne-yāwā=n bilūq=ū sinʕe-ē=š \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} like now \textsc{ind-}say.prs\textsc{-3pl} yet \textsc{neg-}arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} maturity\textsc{.m}\textsc{=ez.gen} age.f\textsc{-obl}.f\textsc{=3sg} \\ 
\glt `As [it is] nowadays, they (the agha’s attendants) said, ‘he hasn’t hit puberty yet.'
\z 
 
\ea \label{BP.138}
\textit{hemey wirdīkɫey mekerā zīnān ewīšā kerā zīnān} \\ 
\gll heme=ī wirdīkɫe-ī me-ker-ā zīnān ewīšā ∅-ker-ā zīnān \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{-obl}\textsc{.m} \textsc{neg-}do.prs\textsc{-3pl} prison\textsc{.m} 3pl\textsc{.obl} \textsc{ind-}do.prs\textsc{-3pl} prison\textsc{.m} \\ 
\glt `They didn’t put Hama Wirdikla in jail. They put them (Baba Khwada and Hama Ghaybi) in jail.'
\z 
 
\ea \label{BP.139}
\textit{berā kerāšāne zīnāneke, wextē miđyā mezā nū če noʕē bīyen ħejījīyāyčšā girten xuɫkū hejījī} \\ 
\gll ∅-ber-ā ∅-ker-ā=šā=ne zīnān-eke wext-ē mi-đye-ā me-zān-ū če noʕ-ē bīye=n ħejījī-yā=yčšā girte=n xuɫk=ū hejīj-ī \\ 
 \textsc{ind-}take.prs\textsc{-3pl} \textsc{ind-}do.prs\textsc{-3pl}\textsc{=3pl}\textsc{=\textsc{povb}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-\textsc{1sg}} what kind/way\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} people\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They took [them] and put them in the jail, they realised that —they didn’t know for what reason— [but] a person from Hajij had also been arrested.'
\z 
 
\ea \label{BP.143}
\textit{ħemey wirdīkɫeyč mičkile bo, wirdīkɫe bo pāne girđ seferē milo pā šārne gēɫo} \\ 
\gll ħeme=ī wirdīkɫe=yč mičkile ∅-b-o wirdīkɫe ∅-b-o p=ā=ne girđ sefer-ē mi-l-o p=ā šār=ne ∅-gēɫ-o \\ 
 \textsc{pn}=ez.\textsc{attr} small\textsc{=add} small \textsc{ind-}be.prs\textsc{-3sg} small \textsc{ind-}be.prs\textsc{-3sg} at=\textsc{dem.dist}\textsc{=\textsc{post}} all trip\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} at=\textsc{dem.dist} city\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}wander.prs\textsc{-3sg} \\ 
\glt `Hama Wirdikla was small and young. He would constantly go around in the city.'
\z 
 
\ea \label{BP.154}
\textit{māčā, čī girewī āne miheme?’} \\ 
\gll m-āč-ā čī ∅-girew-ī āne miheme \\ 
 \textsc{ind-}say.prs\textsc{-3pl} why \textsc{ind-}cry.prs-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{pn} \\ 
\glt `They said, ‘Mihamad, why are you crying?’'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.167}
\textit{ā wextī milo mēwe herkesī goš dārābone, dūr kewtenwe je, kučē je šāreke meselen e čē luān tā milemārfā wēšū xānewādeš} \\ 
\gll ā wext-ī mi-l-o m-ē=we herkesī goš dārā b-one dūr kewte=n=we je kuč-ē je šār-eke meselen e=čē luā=n tā milemārfā wē=š=ū xānewāde=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}come.prs\textsc{.3sg}\textsc{=compl} whoever ear\textsc{.m} hold\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3sg} far fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} from little\textsc{-indf} from city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} for\_example in=here go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until \textsc{pn} \textsc{reflx}\textsc{=3sg}=and family\textsc{.m}\textsc{=3sg} \\ 
\glt `Then, he went [to town] and came back. Whoever might have listened to him had gone away from the city. For instance, they had gone together with their family as far as Milamarfa\footnote{name of a summer habitat.}.'
\z 
 
\ea \label{BP.168}
\textit{ā wextī bābāw pīr šelīyārī milowe dūrē leqē māɫo zīnānekeyre} \\ 
\gll ā wext-ī bābā=ū pīr šelīyār-ī mi-l-o=we dūr-ē leq-ē ∅-māɫ-o zīnān-ekey=re \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs\textsc{-3sg}\textsc{=compl} far\textsc{-indf} kick\textsc{-indf} \textsc{ind-}smash.prs\textsc{-3sg} prison\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Then, Pir Shaliyar’s grandfather went backwards in his cell and kicked the wall of the prison.'
\z 
 
\ea \label{BP.175}
\textit{āđē āzāyē bā mizyāre řū be hewrāmān} \\ 
\gll āđē āzā-ē ∅-b-ā mi-zye-ā=re řū be hewrāmān \\ 
 3pl\textsc{.dir} free\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} face to \textsc{pn} \\ 
\glt `They were liberated. They returned to Hawraman. '
\z 
 
\ea \label{BP.178}
\textit{āhā māčā īnā čū dīmne lāw beynū zāwerīnew tewēɫēne ħemey ġeybī hen} \\ 
\gll āhā m-āč-ā īnā-∅ č=ū dīm=ne lā=ū beyn=ū zāwer-ī=ne=ū tewēɫē=ne ħeme=ī ġeybī hen-∅ \\ 
 \textsc{intj} \textsc{ind-}say.prs\textsc{-3pl} deic\textsc{-3sg}\textsc{.m} in=\textsc{dem.dist}\textsc{.3sg} side\textsc{.m}\textsc{=\textsc{post}} at\_the\_place\_of\textsc{=ez.gen} among\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}}=and \textsc{pn}\textsc{=\textsc{post}} \textsc{pn}=ez.\textsc{attr} invisible exist\textsc{-3sg}\textsc{.m} \\ 
\glt `It is said that, on this side of the border, there is the shrine of Hama the invisible between the villages of Zawar and Tawela.'
\z 
 
\ea \label{BP.179}
\textit{mēnē bēnēwe pey hewrāmānī, mēnē bēnēwe pey hewrāmānī ētir zīnānne mirexesē bāw āzāē bāw} \\ 
\gll m-ē-nē b-ē-nē=we pey hewrāmān-ī m-ē-nē b-ē-nē=we pey hewrāmān-ī ētir zīnān=ne mirexes-ē b-ā=ū āzā-ē b-ā=ū \\ 
 \textsc{ind-}come.prs\textsc{-3pl} \textsc{sbjv-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{-3pl} \textsc{sbjv-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} prison\textsc{.m}\textsc{=\textsc{post}} released\textsc{\textsc{-pl}} be.prs\textsc{-3pl}=and free\textsc{\textsc{-pl}} be.prs\textsc{-3pl}=and \\ 
\glt `They were on their way to Hawraman. They were on their way to Hawraman. They were liberated from the prison. They got free.'
\z 
 
\ea \label{BP.181}
\textit{mēnē e čā ħemey ġeybī fermāwo, fermāwo, min mānīyēnmē e čē bā kučē nīšmēre} \\ 
\gll m-ē-nē e=čā ħeme=ī ġeybī ∅-fermāw-o ∅-fermāw-o min mānīyē=nmē e=čē bā kuč-ē ∅-nīš-mē=re \\ 
 \textsc{ind-}come.prs\textsc{-3pl} in=there \textsc{pn}=ez.\textsc{attr} invisible \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} \textsc{1sg} tired\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} in=here \textsc{hort} little\textsc{-indf} \textsc{sbjv-}sit.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}} \\ 
\glt `They went [to the region between Zawar and Tawela]. Hama Ghaybi said, ‘I, we are tired. Let us rest\footnote{Lit. sit.} a little bit here.'
\z 
 
\ea \label{BP.186}
\textit{kuřēmā bo nāmēš binyeydē miħemeđ} \\ 
\gll kuř-ē=mā ∅-b-o nāmē=š bi-nye-ydē miħemeđ \\ 
 boy\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} \textsc{ind-}be.prs\textsc{-3sg} name.f\textsc{=3sg} \textsc{imp-}put.prs-\textsc{2pl} \textsc{pn} \\ 
\glt `We will have a son. Name him Mohammad.'
\z 
 
\ea \label{BP.192}
\textit{čiřāš pene, hurbēzowe dērā mihemeđ} \\ 
\gll ∅-čiř-ā=š pene hur- b-ēz-o=we dēr=ā mihemeđ \\ 
 \textsc{ind-}call.prs\textsc{-3pl}\textsc{=3sg} to \textsc{pvb-}\textsc{sbjv-}rise.prs\textsc{-3sg}\textsc{=compl} late\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \\ 
\glt `They shouted at him, ‘Get up Mohammad, it’s late!’'
\z 
 
\ea \label{BP.201}
\textit{mēnēwe hewrāmān} \\ 
\gll m-ē-nē=we hewrāmān \\ 
 \textsc{ind-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{pn} \\ 
\glt `They went back to Hawaram.'
\z 
 
\ea \label{BP.203}
\textit{ā wextī penj šiš neferē je hewrāmāno milā hesārē muweqetīyū ħemey ġeybī kēšā} \\ 
\gll ā wext-ī penj šiš nefer-ē je hewrāmān=o mi-l-ā hesār-ē muweqetī=ū ħeme=ī ġeybī kēšā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} five six person\textsc{.m}\textsc{-indf} from \textsc{pn}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} wall\textsc{-indf} temporary\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible pull\textsc{.pst} \\ 
\glt `Then five, six people went from Hawraman [and] raised a temporary wall around Hama Ghaybi’s shrine.'
\z 
 
\ea \label{BP.205}
\textit{mēnēwe čā dimāy, ženū ħemey ġeybī hāmīlē bo kuřē wīno,} \\ 
\gll m-ē-nē=we čā dimā-ī ženī=ū ħeme=ī ġeybī hāmīlē ∅-b-o kuř-ē ∅-wīn-o \\ 
 \textsc{ind-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} there afterwards\textsc{-obl}\textsc{.m} woman.f=and \textsc{pn}=ez.\textsc{attr} invisible pregnant.f \textsc{ind-}be.prs\textsc{-3sg} son/boy\textsc{.m}\textsc{-indf} \textsc{ind-}see.prs\textsc{-3sg} \\ 
\glt `After they came back [to Hawraman], Hama Ghaybi’s wife, who was pregnant, gave birth to a boy [Lit. She sees a boy].'
\z 
 
\ea \label{BP.206}
\textit{kuřey nāmē minyā mihemeđ} \\ 
\gll kuř-e-ī nāmē=ī mi-nye-ā mihemeđ \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{-obl}\textsc{.m} name.f=ez.\textsc{attr} \textsc{ind-}put.prs\textsc{-3pl} \textsc{pn} \\ 
\glt `They named the son Muhammad.'
\z 
 
\ea \label{BP.207}
\textit{īse čiřāš pene mihemeđe wirdīkɫe} \\ 
\gll īse ∅-čiř-ā=š pene mihemeđ=e wirdīkɫe \\ 
 now \textsc{ind-}call.prs\textsc{-3pl}\textsc{=3sg} to \textsc{pn}=ez\textsc{.cmpd} small \\ 
\glt `Nowadays, he is called Little  Muhammad.'
\z 
 
\ea \label{BP.208}
\textit{īne dāstānū īney, īnē dāstānū bābāw pīr šelīyārīū ħemey ġeybī bīyēnē} \\ 
\gll īne dāstāne=ū īney īnē dāstāne=ū bābā=ū pīr šelīyār-ī=ū ħeme=ī ġeybī bīyē=nē \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story.f\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \textsc{pn}=ez.\textsc{attr} invisible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `This was the story of Pir Shaliyar’s grandfather and Hama Ghaybi.'
\z 
 
\ea \label{ŠJ.1}
\textit{zemānew jāfir sānū luhonī bo} \\ 
\gll zemāne=ū jāfir sān=ū luhon-ī ∅-b-o \\ 
 period\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `It was during the rule of Jafir Sān in Luhon.'
\z 
 
\ea \label{ŠJ.2}
\textit{šā ʕumerū bīyārē tāze āmābo bīyārē} \\ 
\gll šā ʕumer=ū bīyārē tāze āmā b-o bīyārē \\ 
 king\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn} anyway come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3sg} \textsc{pn} \\ 
\glt `Sheikh Umer from Biyare might have just arrived at Biyare.'
\z 
 
\ea \label{ŠJ.3}
\textit{tāze āmābo bīyārē} \\ 
\gll tāze āmā b-o bīyārē \\ 
 anyway come\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3sg} \textsc{pn} \\ 
\glt `He might have just come to Biyare.'
\z 
 
\ea \label{ŠJ.9}
\textit{duē yerē mēnē pāne xeɫkū mentīqew hīnī jāfir sānū luhonī} \\ 
\gll duē yerē m-ē-nē p=ā=ne xeɫk=ū mentīqe=ū hīn-ī jāfir sān=ū luhon-ī \\ 
 two three \textsc{ind-}come.prs\textsc{-3pl} at=\textsc{dem.dist}\textsc{=\textsc{post}} people\textsc{.m}\textsc{=ez.gen} region\textsc{.m}\textsc{=ez.gen} \textsc{fill}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Two three men went\footnote{Lit. came} there [to Biyare]; they were from Jafir Sān’s region, from Luhon.'
\z 
 
\ea \label{ŠJ.10}
\textit{mēnē māčā šē ʕumer čikon} \\ 
\gll m-ē-nē m-āč-ā šē ʕumer čiko=n \\ 
 \textsc{ind-}come.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sheikh\textsc{.m} \textsc{pn} where\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] said, ‘Where is Sheikh Umer?’'
\z 
 
\ea \label{ŠJ.13}
\textit{milāwe lāw jāfer sānū luhonī} \\ 
\gll mi-l-ā=we lā=ū jāfer sān=ū luhon-ī \\ 
 \textsc{ind-}go.prs\textsc{-3pl}\textsc{=compl} at\_the\_place\_of\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `They went back to Jafir Sān in Luhon.'
\z 
 
\ea \label{ŠJ.14}
\textit{māčā qurbān to čī bīyenī pīyā, čī bīyenī jāfir sān?} \\ 
\gll m-āč-ā qurbān to čī bīye=nī pīyā čī bīye=nī jāfir sān \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{2sg} why be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} man\textsc{.m} why be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{pn} \textsc{pn} \\ 
\glt `[and] said, ‘Sir, how have you become such a man? How have you turned into the Jafir Sān (that you are)?'
\z 
 
\ea \label{ŠJ.15}
\textit{wātšā hīne šēh ʕumerū bīyārēn, jāfir sān} \\ 
\gll wāt=šā hīne šēh ʕumer=ū bīyārē=n jāfir sān \\ 
 say\textsc{.pst}\textsc{=3pl} \textsc{fill} sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \textsc{pn} \\ 
\glt `They said, ‘It’s Sheikh Umer from Biyare [who is acting like a chieftain]!’'
\z 
 
\ea \label{ŠJ.16}
\textit{vātiš čenīnā jāfir sān?} \\ 
\gll vāt=iš čenīn=ā jāfir sān \\ 
 say\textsc{.pst}\textsc{=3sg} how\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \textsc{pn} \\ 
\glt `Jafir Sān said, ‘What is he like?’'
\z 
 
\ea \label{ŠJ.21}
\textit{lodē bārdēš peym šē ʕumerī} \\ 
\gll ∅-lo-dē b-ār-dē=š pey=m šē ʕumer-ī \\ 
 \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}bring.prs-\textsc{2pl}\textsc{=3sg} for\textsc{=\textsc{1sg}} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Go [and] bring Sheikh Umer to me.’'
\z 
 
\ea \label{ŠJ.22}
\textit{mēnē bīyārē persā māčā šē ʕumer čikon?} \\ 
\gll m-ē-nē bīyārē ∅-pers-ā m-āč-ā šē ʕumer čiko=n \\ 
 \textsc{ind-}come.prs\textsc{-3pl} \textsc{pn} \textsc{ind-}ask.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sheikh\textsc{.m} \textsc{pn} where\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went to Biyare [and] asked, ‘Where is Sheikh Umer?’'
\z 
 
\ea \label{ŠJ.29}
\textit{wēšā be pīyāw jāfir sānī zānā} \\ 
\gll wē=šā be pīyā=ū jāfir sān-ī ∅-zān-ā \\ 
 \textsc{reflx}\textsc{=3pl} to man\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know.prs\textsc{-3pl} \\ 
\glt `They were boasting about being Jafir Sān’s men.'
\z 
 
\ea \label{ŠJ.34}
\textit{māčā pīyew jāfir sānīnmē} \\ 
\gll m-āč-ā pīye=w jāfir sān-ī=nmē \\ 
 \textsc{ind-}say.prs\textsc{-3pl} man\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `They said, ‘We are Jafir Sān’s men.’'
\z 
 
\ea \label{ŠJ.36}
\textit{āmēnmē jafir sānī fermāwān lodē} \\ 
\gll āmē=nmē jafir sān-ī fermāwā=n ∅-lo-dē \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \\ 
\glt `We have come here. Jafir San said, ‘Go [and bring him].’'
\z 
 
\ea \label{ŠJ.39}
\textit{čenū gum bīyey gumē bīyā serew řoē gumē bā mezānā bilāwe pey nođšey} \\ 
\gll čenū gum bīyey gum-ē bī-ā sere=ū řo-ē gum-ē ∅-b-ā me-zān-ā bi-l-ā=we pey nođše-ī \\ 
 with\textsc{.ez.gen} lost be\textsc{.inf} lost\textsc{-indf} be\textsc{.pst}\textsc{-3pl} \textsc{clf}\textsc{=ez.gen} day\textsc{.m}\textsc{-dir}\textsc{.pl} lost\textsc{\textsc{-pl}} \textsc{ind-}be.prs\textsc{-3pl} \textsc{neg-}know.prs\textsc{-3pl} \textsc{sbjv-}go.prs\textsc{-3pl}\textsc{=compl} to \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[They got so scared that] they were lost. They were lost for one day. They didn’t know the way to go back to Nodsha.'
\z 
 
\ea \label{ŠJ.43}
\textit{jafir sān māčo čī āmēndē? čī dēr āmēndē?} \\ 
\gll jafir sān m-āč-o čī āmē=ndē čī dēr āmē=ndē \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} why come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}.\textsc{2pl} why late come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}.\textsc{2pl} \\ 
\glt `Jāfir Sān said, ‘Why have you come back [now]? Why have you come back late?’'
\z 
 
\ea \label{ŠJ.46}
\textit{šewē jafir sān sere minyowe wermiš gino pore} \\ 
\gll šewe-ē jafir sān sere mi-nye-o=we werm=iš ∅-gin-o pore \\ 
 night.f\textsc{-obl}.f \textsc{pn} \textsc{pn} head\textsc{.m} \textsc{ind-}put.prs\textsc{-3sg}\textsc{=compl} dream\textsc{=3sg} \textsc{ind-}fall.prs\textsc{-3sg} at \\ 
\glt `At night, Jafir Sān went to sleep. He had a dream.'
\z 
 
\ea \label{ŠJ.56}
\textit{mē tewēɫē,} \\ 
\gll m-ē tewēɫe-ē \\ 
 \textsc{ind-}come.prs\textsc{.3sg} \textsc{pn}\textsc{-obl}.f \\ 
\glt `and went to Tawela.'
\z 
 
\ea \label{ŠJ.57}
\textit{tāteš tewēɫī bīyen šē ʕumere} \\ 
\gll tāte=š tewēɫe-ī bīye=n šē ʕumer-e \\ 
 father\textsc{.m}\textsc{=3sg} \textsc{pn}\textsc{-nmlz} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} sheikh\textsc{.m} \textsc{pn}\textsc{-def}\textsc{.dir} \\ 
\glt ` Sheikh Umer, his father, was from Tawela.'
\z 
 
\ea \label{ŠJ.58}
\textit{šē ʕusmānū tewēɫīn tāteš} \\ 
\gll šē ʕusmān=ū tewēɫe-ī=n tāte=š \\ 
 sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-nmlz}\textsc{=cop}\textsc{.3sg}\textsc{.m} father\textsc{.m}\textsc{=3sg} \\ 
\glt `Sheikh Usman, his [Umer’s] father, was from Tawela.'
\z 
 
\ea \label{ŠJ.59}
\textit{ā wextī mē milo, jāfir sān māčo be pīyākeyš} \\ 
\gll ā wext-ī m-ē mi-l-o jāfir sān m-āč-o be pīyā-ekey=š \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{.3sg} \textsc{ind-}go.prs\textsc{-3sg} \textsc{pn} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} to man\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `Jafir Sān went [to Tawela]. He said to his servant,'
\z 
 
\ea \label{ŠJ.67}
\textit{miđyā jāfir sān īnā kunjēw tewēɫekeyne} \\ 
\gll mi-đye-ā jāfir sān īnā-∅ kunj-ēw tewēɫe-ekey=ne \\ 
 \textsc{ind-}look.prs\textsc{-3pl} \textsc{pn} \textsc{pn} deic\textsc{-3sg}\textsc{.m} corner\textsc{-indf} \textsc{pn}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `They realised that Jafir Sān was in a corner within the stable.'
\z 
 
\ea \label{ŠJ.68}
\textit{kirž milā xeber miđā be še ʕusmānū tewēɫē} \\ 
\gll kirž mi-l-ā xeber mi-đe-ā be še ʕusmān=ū tewēɫē \\ 
 quick \textsc{ind-}go.prs\textsc{-3pl} news\textsc{.dir}\textsc{.m} \textsc{ind-}give.prs\textsc{-3pl} to sheikh \textsc{pn}\textsc{=ez.gen} \textsc{pn} \\ 
\glt `They went quickly [and] and passed the news to Sheikh Usman'
\z 
 
\ea \label{ŠJ.71}
\textit{māčo qurbān hīn jāfir sānū luhonī īnā tewēɫew wiɫāxekāne} \\ 
\gll m-āč-o qurbān hīn jāfir sān=ū luhon-ī īnā-∅ tewēɫe=ū wiɫāx-ekā=ne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} \textsc{fill} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} deic\textsc{-3sg}\textsc{.m} stable\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=\textsc{post}} \\ 
\glt `He (the horse groom) said, ‘Sir, thingummy, Jafir Sān from Luhon is in the horses’ stable!’'
\z 
 
\ea \label{ŠJ.75}
\textit{kārū āđī īnā bīyārē, mišo bilo bīyārē} \\ 
\gll kār=ū āđī īnā-∅ bīyārē mišo bi-l-o bīyārē \\ 
 task\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} deic\textsc{-3sg}\textsc{.m} \textsc{pn} aux \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{pn} \\ 
\glt ` His business is in Biyare. He shall go to Biyare.'
\z 
 
\ea \label{ŠJ.77}
\textit{milā māčā qurbān šē ʕusmānī fermāwā kāriš be min nīyā} \\ 
\gll mi-l-ā m-āč-ā qurbān šē ʕusmān-ī fermāwā kār=iš be min nīye=ā \\ 
 \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] said [to Jafir Sān], ‘Sir, the sheikh said, “He has no business with me.'
\z 
 
\ea \label{ŠJ.78}
\textit{mišo bilo pey hīnī bīyārē, bilo bīyārē} \\ 
\gll mišo bi-l-o pey hīn-ī bīyārē bi-l-o bīyārē \\ 
 aux \textsc{sbjv-}go.prs\textsc{-3sg} to \textsc{fill}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{pn} \\ 
\glt `He has to go to Thingummy, to Biyare.'
\z 
 
\ea \label{ŠJ.80}
\textit{tā jimore milo bīyārē čā} \\ 
\gll tā ∅-jim-o=re mi-l-o bīyārē čā \\ 
 until \textsc{ind-}move.prs\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{ind-}go.prs\textsc{-3sg} \textsc{pn} there \\ 
\glt `He (Jafir Sān) set off to go to Biyare from there.'
\z 
 
\ea \label{ŠJ.83}
\textit{milo xizmetū šēh ʕumerī bīyārēū} \\ 
\gll mi-l-o xizmet=ū šēh ʕumer-ī bīyārē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3sg} service\textsc{.m}\textsc{=ez.gen} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{pn}=and \\ 
\glt `He went to the service of Sheikh Umer.'
\z 
 
\ea \label{ŠJ.94}
\textit{šēx herčē fermāwištē fermāwebā šē ʕumerū bīyārē fermāwo peyšū} \\ 
\gll šēx herčē fermāwišt-ē fermāwe b-ā šē ʕumer=ū bīyārē ∅-fermāw-o pey=š=ū \\ 
 sheikh\textsc{.m} every speech\textsc{.m}\textsc{-dir}\textsc{.pl} say\textsc{.pst}\textsc{.ptcp}\textsc{.m} be.prs\textsc{-3pl} sheikh\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} for\textsc{=3sg}=and \\ 
\glt `Whatever saying there was, Sheikh Umer told him'
\z 
 
\ea \label{ŠJ.97}
\textit{ētir bo welhāsil bo muxlīsū āđī jāfir sānū luhonī} \\ 
\gll ētir ∅-b-o welhāsil ∅-b-o muxlīs=ū āđī jāfir sān=ū luhon-ī \\ 
 \textsc{disc.ptcl} \textsc{sbjv-}be.prs\textsc{-3sg} eventually \textsc{ind-}be.prs\textsc{-3sg} disciple\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Anyway, Jafir Sān from Luhon became his disciple.'
\z 
 
\ea \label{ŠJ.102}
\textit{enāne hetā fermāwī ī mentēqew ʕērāqū ērān desū ānīšāwe bīyen} \\ 
\gll enā=ne hetā ∅-fermāw-ī ī mentēqe=ū ʕērāq=ū ērān des=ū ānīšā=we bīye=n \\ 
 otherwise\textsc{=\textsc{post}} even \textsc{ind-}say.prs-\textsc{2sg} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{=ez.gen} \textsc{pn} hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All these regions in Iraq and Iran were under their rule.'
\z 
 
\ea \label{ŠJ.105}
\textit{tenhā yo kuřēšā meneniš duktir Mādēh dinyāne řāš mebowe} \\ 
\gll tenhā yo kuř-ē=šā mene=n=iš duktir Mādēh dinyā=ne řā=š me-b-o=we \\ 
 not\_even one\textsc{.m} son/boy\textsc{.m}\textsc{-indf}\textsc{=3pl} remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} physician\textsc{.m} \textsc{pn} world\textsc{.dir}.f\textsc{=\textsc{post}} road.f\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg}\textsc{=compl} \\ 
\glt `Only a son is left of the chieftains, named Dr Madeh, who is not allowed here.'
\z 
 
\ea \label{KŠ.14}
\textit{māčmēš pene suɫtān mehmūđ bī} \\ 
\gll m-āč-mē=š pene suɫtān mehmūđ bī-∅ \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-\textsc{1pl}}\textsc{=3sg} to sultan \textsc{pn} be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `[Let’s say] his (the king's) name was Sultan Mahmud.'
\z 
 
\ea \label{KŠ.47}
\textit{yewāšē hīn hukim mēnē gērāšā berāšā pey menteqē pēse esfehānī} \\ 
\gll yewāšē hīn hukim m-ē-nē ∅-gēr-ā=šā ∅-ber-ā=šā pey menteqe-ē pēse esfehān-ī \\ 
 then \textsc{fill} rule\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3pl} \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3pl} to region\textsc{-indf} like \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Later, there was an order [according to which] he (the king) was arrested and taken to a region like Isfahan.'
\z 
 
\ea \label{KŠ.50}
\textit{e čē nāmē minvīsā kesūkārū xānī kīyānāšre isfehān} \\ 
\gll e=čē nāme-ē mi-nvīs-ā kesūkār=ū xān-ī ∅-kīyān-ā=š=re isfehān \\ 
 in=here letter\textsc{.m}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3pl} relative\textsc{.m}\textsc{=ez.gen} chief\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}send\textsc{.prs}\textsc{-3pl}\textsc{=3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt `Then, the king’s relatives wrote a letter [and] sent it to Isfahan [to the king]'
\z 
 
\ea \label{ŽH.2}
\textit{eyām qisē kerdē wāčē, zemānū qedīmī pādšā hāřūne(l) řešīd bīyenū nā siɫtān mehmūw heyāsī žīr} \\ 
\gll eyā=m qisē-ē kerdē-ē wāč-ē zemān=ū qedīm-ī pādšā hāřūn-e(l) řešīd bīye=n=ū nā siɫtān mehmū=ū heyās=ī žīr \\ 
 mother\textsc{=1sg} talk\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{-3pl} say\textsc{.prs-}pstc\textsc{.3sg} time\textsc{.m}\textsc{=ez}\textsc{.gen} old\_time\textsc{-obl}\textsc{.m} king\textsc{.m} \textsc{pn}-of \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and no sultan\textsc{.m} \textsc{pn}=and \textsc{pn}\textsc{=ez}.\textsc{attr} shrewd \\ 
\glt `My mother would narrate [tales]. She said, ‘In the olden days, there was Harun Rashid, the king. No, [there was] Sultan Mahmud and Hayas, the shrewd.'
\z 
 
\ea \label{ŽH.6}
\textit{wiɫātēwe termā hen ew dīmew ēme, pēsew zāwerīū tewēɫey} \\ 
\gll wiɫāt-ēwe ter=mā hen-∅ ew dīm=ew ēme pēse=ū zāwer-ī=ū tewēɫe-ī \\ 
 country\textsc{.m}\textsc{-indf} other\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} \textsc{dem.dist} side\textsc{.m}\textsc{=\textsc{post}} \textsc{1pl} like\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `There is another country on the other side of us; something like Zawar (in Iran) and Tawela (in Iraq) village.'
\z 
 
\ea \label{ŽH.7}
\textit{milo pey zāwerī e čē} \\ 
\gll mi-l-o pey zāwer-ī e=čē \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg} for \textsc{pn}\textsc{-obl}\textsc{.m} in=here \\ 
\glt `From here he (Hayas) went to Zawar. '
\z 
 
\ea \label{ŽH.12}
\textit{milā řāhur māčo, heyās māčoš pene, kāke gīyān pilekānēwe binye bā bilmē!’} \\ 
\gll mi-l-ā řā=hur m-āč-o heyās m-āč-o=š pene kāke gīyān pilekānī-ēwe bi-nye-e bā bi-l-mē \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to brother.vob dear.\textsc{voc} stair\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}put\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \\ 
\glt `They set off. He, said, [i.e.] Hayas said to him, ‘Dear brother, set up a ladder for us to move !’'\footnote{A cryptic message meaning ‘tell a story’.}
\z 
 
\ea \label{ŽH.64}
\textit{qerār pāse bo e čē xānū ewēgē āđ čenū āđī suɫtān mehmū bē hečkes bē qāɫešā bī hāčkes ženē xāse bāro bē ewēge āđ bilo bo mēmānšā} \\ 
\gll qerār pāse ∅-b-o e=čē xān=ū ewēge-ē āđ čenū āđī suɫtān mehmū b-ē hečkes b-ē qāɫe=šā bī-∅ hāčkes ženī-ē xās-e b-ār-o b-ē ewēge āđ bi-l-o ∅-b-o mēmān=šā \\ 
 arrangement like \textsc{ind-}be\textsc{.prs}\textsc{-3sg} in=here chief\textsc{.m}\textsc{=ez}\textsc{.gen} there\textsc{-obl}\textsc{\textsc{.f}} 3sg\textsc{.dir}\textsc{.m} with.ez\textsc{.gen} 3sg\textsc{.obl}\textsc{.m} sul \textsc{pn} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} whoever be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} quarrel\textsc{\textsc{.f}}\textsc{=3pl} be\textsc{.pst}\textsc{-3sg}\textsc{.m} whoever woman\textsc{\textsc{.f}}\textsc{-indf} fine\textsc{-f} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg} there \textsc{3sg}\textsc{.dir}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} guest\textsc{.m}\textsc{=3pl} \\ 
\glt `In the old times, the rule was that — the khan of there, whether it was Sultan Mahmud, with whom he (Hayas) had got into a quarrel, or anyone else— whoever got married and took a beautiful wife, he (the king) would go [and] become their guest.'
\z 
 
\ea \label{ŽH.65}
\textit{yewāšē āđīč mārošo mēwe kināčēw luhonī bo teřū čermew īnīšā} \\ 
\gll yewāšē āđ=īč m-ār-o=š=o m-ē=we kināčē=ū luhon-ī ∅-b-o teř=ū čerme=ū īnīšā \\ 
 then 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} fresh=and white=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `He (Hayas) brought her (i.e. his wife). She was from Luhon; she was young, white, and so on.'
\z 
 
\ea \label{ŽH.66}
\textit{māčā be siɫtān mehmūī, be xwā fiɫān heyās āmāno čīwē ečīnew ečāne ʕāliš ārden’} \\ 
\gll m-āč-ā be siɫtān mehmū-ī be xwā fiɫān heyās āmā=n=o čīw-ē e=čīne=w e=čāne ʕāl=iš ārde=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} by sultan\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} by God\textsc{.m} such\_and\_such \textsc{pn} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} thing\textsc{-indf} in=like\_this\textsc{=ez}\textsc{.gen} in=like\_that good\textsc{=3sg} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (people) said to Sultan Mahmud, ‘Such-and-such Hayas has returned. He has brought [with him] such a nice thinggummy (i.e. woman)  '
\z 
 
\ea \label{ŽH.69}
\textit{māčo, bereš pey kurdistānī’} \\ 
\gll m-āč-o ∅-ber-e=š pey kurdistān-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}take\textsc{.prs}-\textsc{2sg}\textsc{=3sg} for \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said [to Hayas], ‘Take this to Kurdistan. '
\z 
 
\ea \label{RE.24}
\textit{xeplezeřātīnē nīštēnēre nānešānē xeplēšā nīyēnēre} \\ 
\gll xeplezeřātīnē nīštē=nē=re nānešān-ē xeple-ē=šā nīyē=nē=re \\ 
 \textsc{pn} sit\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} basket\textsc{-indf} bread\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} \\ 
\glt `xaplazaratene,\footnote{A bread made of corn flour.} they would sit [together] and cook a basket of xaple.\footnote{A bread made of corn flour.}'
\z 
 
\ea \label{RE.58}
\textit{ā wextīyekey āwyānešā sānān} \\ 
\gll ā wext-ī-ekey āwyāne=šā sānā=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{pn}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, they (the noblemen) would ask for awyana (i.e. the money in exchange for giving consent to marriage)'
\z 
 
\ea \label{RE.59}
\textit{sūrānew ženāšā sānān} \\ 
\gll sūrāne=ū ženī-ā=šā sānā=n \\ 
 \textsc{pn}\textsc{=ez.gen} woman\textsc{.f}\textsc{-obl}\textsc{.pl}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the noblemen) would grab surana (i.e. money in exchange for giving consent to marriage)'
\z 
 
\ea \label{RE.60}
\textit{sūrāne yānē mišo bārī to zeř dey jā min māčū erē} \\ 
\gll sūrāne yānē mišo ∅- b-ār-ī to zeř ∅-de-ī jā min m-āč-ū erē \\ 
 \textsc{pn} that\_means aux \textsc{ind-}\textsc{sbjv-}bring\textsc{.prs}-\textsc{2sg} \textsc{2sg} money\textsc{.m} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} afterwards \textsc{1sg} \textsc{ind-}say\textsc{.prs}\textsc{-\textsc{1sg}} yes \\ 
\glt `Surana means that you should give me money, then I will say, ‘Yes [to the marriage]'
\z 
 
\ea \label{RE.61}
\textit{sūrāne ānen} \\ 
\gll sūrāne āne=n \\ 
 \textsc{pn} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Surana was that.'
\z 
 
\ea \label{ŽE.42}
\textit{čirāwe desīšā, čirāwekešā weš kerden bilā pey yānew pīr šelīyārī pey čīwī bārānšo} \\ 
\gll čirāw=e desī=šā čirāw-eke=šā weš kerde=n bi-l-ā pey yāne=ū pīr šelīyār-ī pey čīw-ī b-ār-ān=š=o \\ 
 lamp\textsc{.m}\textsc{=ez}\textsc{.cmpd} hand\textsc{=3pl} lamp\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3pl} for house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} for thing\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}bring\textsc{.prs}\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `They lit the lamps to go the Pir Shaliyar’s shrine\footnote{Lit. house.} [and] to fetch things.'
\z 
 
\ea \label{ŽE.50}
\textit{ī doīnete dēnmēū ī šelemīnēte dēnmē} \\ 
\gll ī doīne=t=e dē=nmē=ū ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{\textsc{=ez.gen}} \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this dwina. \footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{ŽE.51}
\textit{ī šelemīnēte dēnmē} \\ 
\gll ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `you gave us this Shalamina. \footnote{Dish made of cracked wheat and turnip.}'
\z 
 
\ea \label{ŽE.52}
\textit{ī xeplēte dēnmē} \\ 
\gll ī xeplē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this xapla '
\z 
 
\ea \label{ŽM.23}
\textit{luēnmē ferz tārān kārmā kerđen šiš māngē} \\ 
\gll luē=nmē ferz tārān kār=mā kerđe=n šiš mānge-ē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} guess\textsc{.m} \textsc{pn} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go, say, to Tehran and work for six months.'
\z 
 
\ea \label{ŽM.24}
\textit{luēnmē kirmān kārmā kerđen šiš māngē} \\ 
\gll luē=nmē kirmān kār=mā kerđe=n šiš mānge-ē \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kerman and work for six months.'
\z 
 
\ea \label{ŽM.25}
\textit{jezīrew kīšīne kārmā kerđen šiš māngē} \\ 
\gll jezīre=w kīš-ī=ne kār=mā kerđe=n šiš mānge-ē \\ 
 island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} work\textsc{.m}\textsc{=\textsc{1pl}} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} six month\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `We would go to Kish Island and work for six months.'
\z 
 
\ea \label{ŽM.33}
\textit{šesū šiš sāʕetē ēme be māšīn jā luēnmē binderʕebāsū} \\ 
\gll šes=ū šiš sāʕet-ē ēme be māšīn jā luē=nmē binderʕebās=ū \\ 
 sixty\textsc{=ez.gen} six hour\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1pl} by car\textsc{.m} afterwards go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and \\ 
\glt `It took us sixty-three hours to reach Bandar Abbas by car.'
\z 
 
\ea \label{ŽM.34}
\textit{luēnmē čārekū kerdenmā ew des} \\ 
\gll luē=nmē čārek=ū kerde=n=mā ew des \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{dem.dist} hand\textsc{.m} \\ 
\glt `We went to Charak and went to the other side.'
\z 
 
\ea \label{ŽM.38}
\textit{extīyārmā mištū ā fārsāne bīyen jezīrew kīšī} \\ 
\gll extīyār=mā mišt=ū ā fārs-ā=ne bīye=n jezīre=ū kīš-ī \\ 
 volition\textsc{.m}\textsc{=\textsc{1pl}} fist\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} Persian\textsc{-3pl}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} island\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `We were under the rule of Persians at Kish island.'
\z 
 
\ea \label{ŽM.39}
\textit{imrīkāī āmēnē ečā kārmā pey kerdēnē} \\ 
\gll imrīkāī āmē=nē e=čā kār=mā pey kerdē=nē \\ 
 \textsc{pn} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} in=there work\textsc{.m}\textsc{=\textsc{1pl}} to do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Americans came [there]. We had them work [for us].'
\z 
 
\ea \label{ŽM.41}
\textit{kuweyt duē sāʕetē beynšā čenū kuweytī} \\ 
\gll kuweyt duē sāʕet-ē beyn=šā čenū kuweyt-ī \\ 
 \textsc{pn} two hour\textsc{.m}\textsc{-dir}\textsc{.pl} between\textsc{=3pl} with\textsc{\textsc{.ez}.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Kuwait was two hours away from there.'
\z 
 
\ea \label{ŽM.42}
\textit{kuweyt, āđīč wēš kēšwerēwā kuweyt} \\ 
\gll kuweyt āđ=īč wē=š kēšwer-ēw=ā kuweyt \\ 
 \textsc{pn} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{reflx}\textsc{=3sg} country\textsc{-indf}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pn} \\ 
\glt `Kuwait is a country on its own.'
\z 
 
\ea \label{ŽM.43}
\textit{ā zemāne kuweyt je girđū kēšwerekā dewɫetmenter bē} \\ 
\gll ā zemān=e kuweyt je girđ=ū kēšwer-ekā dewɫetmen-ter b-ē \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} from all\textsc{=ez.gen} country\textsc{-def}\textsc{.obl}\textsc{.pl} rich\textsc{-cmpr} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Back then, Kuwait was the richest of all countries.'
\z 
 
\ea \label{ŽM.44}
\textit{je tārānī kuweyt dewɫetmenter bē} \\ 
\gll je tārān-ī kuweyt dewɫetmen-ter b-ē \\ 
 from \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{pn} rich\textsc{-cmpr} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `Kuwait was more affluent than Tehran.'
\z 
 
\ea \label{ŽM.46}
\textit{řoē sī penj timenē kār kerēnmē} \\ 
\gll řo-ē sī penj timen-ē kār ker-ēn-mē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} task\textsc{.m} do\textsc{.prs}\textsc{-pstc}\textsc{-\textsc{1pl}} \\ 
\glt `I worked for a daily salary of thirty-five tomans. '
\z 
 
\ea \label{ŽM.47}
\textit{penj řoē čāne min řemezānim gērten} \\ 
\gll penj řo-ē č=ā=ne min řemezān=im gērte=n \\ 
 five day\textsc{.m}\textsc{-indf} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{1sg} \textsc{pn}\textsc{=\textsc{1sg}} grab\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[while there] I fasted five days during Ramadan.'
\z 
 
\ea \label{ŽM.49}
\textit{řoē sī penj timenē šesū yerē sāɫē čēweɫī kerū ētir meyānū sāɫū činē bē} \\ 
\gll řo-ē sī penj timen-ē šes=ū yerē sāɫe-ē čēweɫī ∅-ker-ū ētir me-zān-ū sāɫe-ū činē b-ē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} sixty\textsc{=ez.gen} three year\textsc{.f}\textsc{-dir}\textsc{.pl} ago \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc}.\textsc{ptcl} \textsc{neg-}know\textsc{.prs}\textsc{-\textsc{1sg}} year\textsc{.f}\textsc{-\textsc{1sg}} how\_many\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `A daily salary of thirty-five tomans, it was sixty-three years ago. I don’t know which year it was/'
\z 
 
\ea \label{ŽM.52}
\textit{luēnmē benenū āmēnmēwew} \\ 
\gll luē=nmē benen=ū āmē=nmē=we=ū \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \textsc{pn}=and come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl}=and \\ 
\glt `We would go to Banan (the summer habitat) and return (home).'
\z 
 
\ea \label{ŽM.54}
\textit{ētir īsey āmēnmē tā wē yāwēnmē be kemūkuřīū hewrāmānī} \\ 
\gll ētir īse-ī āmē=nmē tā wē yāwē=nmē be kemūkořī=ū hewrāmān-ī \\ 
 \textsc{disc}.\textsc{ptcl} now\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} until there arrive\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} by poverty\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have made it so far, [despite] poverty and destitution.'
\z 
 
\ea \label{ŽM.55}
\textit{ētir īse weʕzū hewrāmānī tā menteqē terē kučē xāsterā} \\ 
\gll ētir īse weʕze=ū hewrāmān-ī tā menteqe-ē ter-ē kuč-ē xās-ter=ā \\ 
 \textsc{disc}.\textsc{ptcl} now situation\textsc{.f}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} until region\textsc{-dir}\textsc{.pl} other\textsc{-dir}\textsc{.pl} little\textsc{-indf} fine\textsc{-cmpr}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The situation in Hawraman is a bit better than in other regions.'
\z 
 
