\chapter{All 228 examples of prox in these texts are found in the following sentences:}

\ea \label{ZB.35}
\textit{mekiryoyč ī kuřī xinīkno} \\ 
\gll me-kir-īye-o=yč ī kuř-ī ∅-xinīkn-o \\ 
 \textsc{neg-}do.pass-pass\textsc{-3sg}\textsc{=add} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv}-strangle\textsc{.prs}\textsc{-3sg} \\ 
\glt `We should not strangle the [baby] boys.\footnote{Lit. It can't be done that we strangle the baby boys.}'
\z 
 
\ea \label{ZB.37}
\textit{īne min īnem žinīyen lāw, fermāwištū šēxīn, šā ʕelādīnīwe} \\ 
\gll īne min īne=m žinīye=n lā=ū fermāwišt=ū šēx-ī=n šā ʕelādīn-ī=we \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{1sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} hear\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} at\_the\_place\_of\textsc{\textsc{=ez.gen}} speech\textsc{.m}\textsc{\textsc{=ez.gen}} sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `I have heard this tale from —it is the Sheikh’s words— Shah Aladin.'
\z 
 
\ea \label{ZB.42}
\textit{čī mede xwāwinū ʕālemīyānī pāse īrāyešā bizēw menēne jīāy sāyqekewe} \\ 
\gll č=ī mede xwāwin=ū ʕālemīyān-ī pāse īrāye=š=ā bize-ēw menē=ne jīāy sāyqe-eke=we \\ 
 at=\textsc{dem.prox} time owner\textsc{\textsc{=ez.gen}} worlds\textsc{.m}\textsc{-obl}\textsc{.m} like volition\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} goat\textsc{.f}\textsc{-indf} remain\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} behind lightning\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `The will of the lord of all worlds was such that a goat was accidentally left behind [healthy] from the lightning [that caused the flood]'
\z 
 
\ea \label{ZB.54}
\textit{ī bizēče powe āwēzāne bo mē e čē ēđ čeknošū milo lā yāgekēš} \\ 
\gll ī bizē=č=e p=o=we āwēzān-e ∅-b-o m-ē e=čē ēđ čekn-o=š=ū mi-l-o lā yāgē-ekē=š \\ 
 \textsc{dem.prox} goat\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=dem} at=\textsc{dem.dist}\textsc{=compl} hanging\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} in=here 3sg.prox\textsc{.dir} suckle\textsc{.prs}\textsc{-3sg}\textsc{=3sg}=and \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of place\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=3sg} \\ 
\glt `The goat would stay in a hanging position, the child would feed from its udder, and it would go back to its place.'
\z 
 
\ea \label{ZB.58}
\textit{īne dāstānū gewregerīū ā mešeyxāne} \\ 
\gll īne dāstāne=ū gewregerī=ū ā mešeyxā=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{\textsc{=ez.gen}} greatness\textsc{\textsc{=ez.gen}} \textsc{dem.dist} sheikh\textsc{.obl}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `This is the story of the greatness of the sheikhs.'
\z 
 
\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.11}
\textit{wextū ī jimāymārā āmāymēwe āmāymē kelēwe} \\ 
\gll wext=ū ī jimāy=mā=rā āmā-ymē=we āmā-īmē kel-ēwe \\ 
 time\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} move\textsc{.inf}\textsc{=\textsc{1pl}}\textsc{=\textsc{povb}} come\textsc{.pst} \textsc{1pl}\textsc{=compl} come\textsc{.pst} \textsc{1pl} mountain\textsc{-indf} \\ 
\glt `On our return (journey), we went\footnote{Lit. came.}  to a mountain. '
\z 
 
\ea \label{ZQ.25}
\textit{vātim, xo hinū meney nīyā metāwmēšū gīrmā werden težnāyene epē wext bē sočmē nāřehetīyene’} \\ 
\gll vāt=im xo hin=ū meney nīye=ā me-tāw-mē=š=ū gīr=mā werde=n težnāy=ene e=p=ē wext b-ē ∅-soč-mē nāřehetī=ene \\ 
 say\textsc{.pst}\textsc{=\textsc{1sg}} dis.ptc ez.pron\textsc{=ez.gen} remain\textsc{.inf} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}can.prs\textsc{-\textsc{1pl}}\textsc{=3sg}\textsc{=ez.gen} hook\textsc{.m}\textsc{=\textsc{1pl}} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} thirst\textsc{=\textsc{post}} in.\textsc{dem.prox} time\textsc{.m} \textsc{sbjv-}come.prs\textsc{.3sg} \textsc{ind-}burn.prs\textsc{-\textsc{1pl}} sadness\textsc{=\textsc{post}} \\ 
\glt `I said [to myself], ‘Well, he is not going to live. We can’t cope with him. We are stuck. We were about to burn from thirst and from sorrow [from leaving behind one of the sons]’'
\z 
 
\ea \label{ZQ.32}
\textit{āy bāwke řo, āy tāte řo īđīč luāw ewīč luā} \\ 
\gll āy bāwke řo āy tāte řo īđ=īč luā=ū ew=īč luā \\ 
 \textsc{intj} fahter.\textsc{voc} onom \textsc{intj} father\textsc{.m} onom 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg}=and 3sg\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg} \\ 
\glt `Oh dear God\footnote{The expressions bāwka řo and tāta řo are used to mourn someone’s death, or to express distress.}, he (the child whom I took with me) died; he (the child whom I left behind) died too!\footnote{Lit. This one went; that one, too, went.} '
\z 
 
\ea \label{ZQ.39}
\textit{wextē luānē werew temāšem kerd miđyew be ʕemrew xwāy e čī qiřolī| ēnnew čīčew ženē pīfē peydā bīyenē| ʕeyn pīfē} \\ 
\gll wext-ē luā-ānē wer=ew temāše=m kerd mi-đye-ū be ʕemre=ū xwā-ī ečī qiřoɫ-ī ēnne=w čīče=ū ženī-ē pīfē-ē peydā bīyē=nē ʕeyn pīfē \\ 
 time\textsc{.m}\textsc{-indf} go\textsc{.pst}\textsc{-\textsc{1sg}} front\textsc{.m}\textsc{=\textsc{post}} looking\textsc{=\textsc{1sg}} do\textsc{.pst} \textsc{ind-}look.prs\textsc{-\textsc{1sg}} by order.f\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} in.\textsc{dem.prox} hollow\textsc{.m}\textsc{-obl}\textsc{.m} as\_much\_as\textsc{=ez.gen} nipple\textsc{=ez.gen} woman.f\textsc{-obl}.f burr.f\textsc{-dir}\textsc{.pl} visible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} like balloon.f \\ 
\glt `When I got closer and looked inside, I noticed that, by the order of God, some burrs the size of a woman's nipple had appeared in the tree hollow, like balloons.'
\z 
 
\ea \label{ZQ.42}
\textit{xwāy deseɫātē čāneš bīyen ī pīfēše weše kerdēne} \\ 
\gll xwā-ī deseɫāt=ē čāne=š bīye=n ī pīfē=š=e weš-e kerdē=ne \\ 
 God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m}=ez.\textsc{attr} like\_that\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox} burr.f\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `God’s work was such: he had made this burr; '
\z 
 
\ea \label{ZQ.44}
\textit{ī kuře her memeš wārden her memes wārden} \\ 
\gll ī kuř=e her meme=š wārde=n her meme=š wārde=n \\ 
 \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} just breast\textsc{.dir}\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} just breast\textsc{.dir}\textsc{.m}\textsc{=3sg} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The boy had kept breast-feeding [from the tree]; he had kept breast-feeding .'
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{HB.23}
\textit{ī gīre čēš bī min wārd?’} \\ 
\gll ī gīr=e čēš bī-∅ min wārd \\ 
 \textsc{dem.prox} hook\textsc{.m}\textsc{=dem} what be\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{1sg} eat\textsc{.pst} \\ 
\glt `What is this situation that I am caught in?’'
\z 
 
\ea \label{HB.28}
\textit{xeylē rāziɫū diɫīš kerdēš} \\ 
\gll xeylē rāziɫ=ū diɫ-ī=š kerd-ē=š \\ 
 so much request\textsc{=ez}\textsc{.gen} heart\textsc{.m} \textsc{dem.prox}\textsc{=3sg} do\textsc{.pst} 3pl\textsc{=3sg} \\ 
\glt `He struggled a lot.'
\z 
 
\ea \label{HB.44}
\textit{hereke mēwe juwāb ī qisā kero} \\ 
\gll her-eke m-ē=we juwāb ī qisē-ā ∅-ker-o \\ 
 donkey\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} reply\textsc{.m} \textsc{dem.prox} talk\textsc{\textsc{.f}}\textsc{-obl}\textsc{.pl} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The donkey started to speak; it spoke.'
\z 
 
\ea \label{HB.67}
\textit{vātiš, be xwā kābrāy, be xwā min luānēwe bīyārē ī bārū herīme vistre dehfē ter min lā šēxīwe menīšū} \\ 
\gll vāt=iš be xwā kābrā-ī be xwā min luā-ānē=we bīyārē ī bār=ū her-ī=m=e vist=re dehfe=ē ter min lā šēx-ī=we me-nīš-w \\ 
 say\textsc{.pst}\textsc{=3sg} by God\textsc{.m} man\textsc{-obl}\textsc{.m} by God\textsc{.m} 1sg go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{-1sg}\textsc{=\textsc{compl}} \textsc{pn} \textsc{dem.prox} load\textsc{.m}\textsc{=ez}\textsc{.gen} donkey\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} throw\textsc{.pst}\textsc{=\textsc{povb}} time\textsc{\textsc{.f}}\textsc{=ez}.\textsc{attr} other 1sg at\_the\_place\_of sheikh\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{neg-}sit\textsc{.prs}\textsc{-1sg} \\ 
\glt `The fellow, i.e. the donkey breeder, said, ‘By God, when I go back to Biyare [and] unload the donkey’s load, I won’t sit together with the sheikh anymore!'
\z 
 
\ea \label{HB.71}
\textit{ī herī zūwāniš zānā min hīčim nezānān!} \\ 
\gll ī her-ī zūwān=iš zānā min hīč=im ne-zānā=n \\ 
 \textsc{dem.prox} donkey\textsc{.m}\textsc{-obl}\textsc{.m} language\textsc{.m}\textsc{=3sg} know\textsc{.pst}\textsc{.ptcp}\textsc{.m} 1sg nothing\textsc{=1sg} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The donkey knew the [Sheikh’s] language; I didn’t know a thing!'
\z 
 
\ea \label{HB.90}
\textit{dā dey hurbēze ī mexlūqī girdiš gēɫeš pone} \\ 
\gll dā dey hur-b-ēz-e ī mexlūq-ī gird=iš gēɫ-e=š pone \\ 
 \textsc{hort} \textsc{disc.ptcl} \textsc{pvb-}\textsc{sbjv-}rise\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} people\textsc{-obl}\textsc{.m} all\textsc{=3sg} wander\textsc{.prs}-\textsc{2sg}\textsc{=3sg} at \\ 
\glt `Come on, get up and search among all these people,'
\z 
 
\ea \label{PM.5}
\textit{fermāndew merekuřekey, reʕīsekešā māčo, ēme bē ā šexsī metāwmē, e čē menmēwe zerer deymē čī mentēqey’} \\ 
\gll fermānde=ū merekuř-ekey reʕīs-eke=šā m-āč-o ēme bē ā šexs-ī me-tāw-mē e=čē ∅-men-mē=we zerer ∅-de-ymē č=ī mentēqe-ī \\ 
 leader\textsc{.m}\textsc{\textsc{=ez.gen}} grasshopper\textsc{-def}\textsc{.obl}\textsc{.m} leader\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-obl}\textsc{.m} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} in=here \textsc{sbjv-}stay\textsc{.prs}\textsc{-1pl}\textsc{=compl} harm\textsc{.m} \textsc{sbjv-}give\textsc{.prs}-\textsc{1pl} in=\textsc{dem.prox} region\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The head of the grasshoppers said, ‘Without [permission from] this person we cannot stay here and do harm to this region.'
\z 
 
\ea \label{PM.29}
\textit{māčo qurbān īne čēšit kerdū?} \\ 
\gll m-āč-o qurbān īne čēš=it kerd=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=\textsc{2sg}} do\textsc{.pst}=and \\ 
\glt `They said, ‘Sir, what have you done?'
\z 
 
\ea \label{PM.30}
\textit{īne bē ījāzew to īne nekiryān} \\ 
\gll īne b-ē ījāze=ū to īne ne-kir-yā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} permission\textsc{\textsc{=ez.gen}} \textsc{2sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{neg-}do.pass-past.pass\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This couldn’t have been done without your permission.'
\z 
 
\ea \label{PM.32}
\textit{īne čī e pēse?} \\ 
\gll īne čī e=pēse \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} why in=like \\ 
\glt `Why is this situation like this?'
\z 
 
\ea \label{PM.37}
\textit{ēme girdū šārezūrī zāno ēsāɫ miɫkekemā nemenenū zerāʕetmā nīyāw benjānīmā nīyenew xīyārmā nīyāw genmīmā nīyenew īne nīyāw ūne nīyāw ūne nīyā} \\ 
\gll ēme gird=ū šārezūr-ī ∅-zān-o ēsāɫ miɫk-eke=mā ne-mene=n=ū zerāʕet=mā nīye=ā=ū benjānī=mā nīye=ne=ū xīyār=mā nīye=ā=ū genmī=mā nīye=ne=ū īne nīye=ā=ū ūne nīye=ā=ū ūne nīye=ā \\ 
 \textsc{1pl} all\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} this\_year property\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1pl} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and agriculture\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and tomato.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and cucmber\textsc{.m}\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and wheat.f\textsc{=1pl} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \textsc{dem.dist} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Everyone in Sharazour knows that we don’t have much cultivated land this year; we don’t have much agriculture; we don’t have cucumbers; we don’t have tomatoes; we don’t have wheat; This [product] is not available; that [product] is not available; that [product] is not available.'
\z 
 
\ea \label{PM.39}
\textit{hīčkāmmā mihtājē mebīmē pīney} \\ 
\gll hīčkām=mā mihtāj-ē me-b-īmē p=īney \\ 
 no\_one\textsc{=1pl} needy\textsc{\textsc{-pl}} \textsc{neg-}be\textsc{.prs} \textsc{1pl} by=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \\ 
\glt `None of us would be in need of these [products] '
\z 
 
\ea \label{PM.45}
\textit{min īse ī nīštenāre īsrāhetīč kerū} \\ 
\gll min īse ī nīšte=nā=re īsrāhet=īč ∅-ker-ū \\ 
 \textsc{1sg} now \textsc{dem.prox} sit\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=\textsc{povb}} rest\textsc{.m}\textsc{=add} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Now I am seated and I am resting.'
\z 
 
\ea \label{PM.47}
\textit{ēđ čowe mē peym māro} \\ 
\gll ēđ č=o=we m-ē pey=m m-ār-o \\ 
 3sg.prox\textsc{.dir} in=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for\textsc{=\textsc{1sg}} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} \\ 
\glt `From this direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.51}
\textit{īne heɫbet xwāy meylišā īne} \\ 
\gll īne heɫbet xwā-ī meyl=š=ā īne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} obviously God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \\ 
\glt `This is indeed God’s will'
\z 
 
\ea \label{PM.52}
\textit{īne heɫbet xwāy meylšā} \\ 
\gll īne heɫbet xwā-ī meyl=š=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} obviously God\textsc{.m}\textsc{-obl}\textsc{.m} willingness\textsc{=3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is indeed God’s will’'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.8}
\textit{zemānē je zemāniš bīyen āđē fers kere je šēxī ɫyāʕdīnīwe tā āro āđē šēxū ī memleketī bīyēnē} \\ 
\gll zemān-ē je zemān=iš bīye=n āđē fers ∅-ker-e je šēx=ī ɫyāʕdīn-ī=we tā āro āđē šēx=ū ī memleket-ī bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} from time\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3pl\textsc{.dir} guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} from sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{pn}\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} until today 3pl\textsc{.dir} sheikh\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `From a long time ago, let’s say from the period of Sheikh Ghayasadin until today, they have been Sheikhs in our region.'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerū tā milo kursānū tā milo tā wā, tirkmen sārā āđē ī memliketšā gēɫnān be dīndārī} \\ 
\gll ʕerz=it ∅-ker-ū tā mi-l-o kursān=ū tā mi-l-o tā wā tirkmen sārā āđē ī memliket=šā gēɫnā=n be dīndārī \\ 
 petition\textsc{.m}\textsc{=\textsc{2sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m}\textsc{=ez}\textsc{.gen} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} until there \textsc{pn} \textsc{pn} 3pl\textsc{.dir} \textsc{dem.prox} country\textsc{.m}\textsc{=3pl} lead\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region by piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.26}
\textit{to mišo ī birāyte šewēne bere} \\ 
\gll to mišo ī birā-ī=t=e šewe-ē=ne ∅-ber-e \\ 
 \textsc{2sg} aux \textsc{dem.prox} brother\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `You should take this brother of yours [there] at night.'
\z 
 
\ea \label{DG.34}
\textit{māčo, seweqet bo, min birāyēwim henū īne wāʕzemnew īne kārmāw īne wāʕzemne} \\ 
\gll m-āč-o seweqe=t ∅-b-o min birā-ēw=im hen-∅=ū īne wāʕze=m=ne=ū īne kār=m=ā=ū īne wāʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sacrifice\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{1sg} brother\textsc{.m}\textsc{-indf}\textsc{=1sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} situation\textsc{\textsc{.f}}\textsc{=1sg}\textsc{=\textsc{cop.3sg\textsc{.f}}} \\ 
\glt `He said [to Shah Aladin], ‘May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this; '
\z 
 
\ea \label{DG.38}
\textit{ā wextīyekey, ētir, īse, min yerē sāɫē ī birāme tūšū ī derdēe āmānū} \\ 
\gll ā wext-ī-ekey ētir īse min yerē sāɫe-ē ī birā=m=e tūš=ū ī derde-ē=e āmā=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} encounter\textsc{=ez}\textsc{.gen} \textsc{dem.prox} illness\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=dem} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Now, as for me, it’s been three years that my brother is suffering from this illness. '
\z 
 
\ea \label{DG.43}
\textit{ī birāme derdeš, derdēš īnā po derdū gulīš māčā dāđē!} \\ 
\gll ī birā=m=e derde=š derde-ē=š īnā-∅ p==oderde=ū gulī=š m-āč-ā dāđ-ē \\ 
 \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} illness\textsc{\textsc{.f}}\textsc{=3sg} illness\textsc{\textsc{.f}}\textsc{-indf}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at=\textsc{=\textsc{post}}illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} leprosy\textsc{=3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} yell\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `This brother of mine is ill; it is called leprosy; Please help!'
\z 
 
\ea \label{DG.44}
\textit{īđīč fermāwo, lue čā nīšere,} \\ 
\gll īđ=īč ∅-fermāw-o ∅-lu-e čā ∅-nīš-e=re \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} there \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \\ 
\glt `Shah Aladin  said, ‘Go [and] sit there;'
\z 
 
\ea \label{DG.46}
\textit{šēx mutewej bo, wextē miđyo derdū ī pīyāy lā xwāyū xwāwinū ʕāɫemīyānī mutewej bo, řānimonīš kero} \\ 
\gll šēx mutewej ∅-b-o wext-ē mi-đye-o derde=ū ī pīyā-ī lā xwā-ī=ū xwāwin=ū ʕāɫemīyān-ī mutewej ∅-b-o řānimonī=š ∅-ker-o \\ 
 sheikh\textsc{.m} aware \textsc{ind-}be\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} man\textsc{.m}\textsc{-obl}\textsc{.m} at\_the\_place\_of God\textsc{.m}\textsc{-obl}\textsc{.m}=and owner\textsc{=ez}\textsc{.gen} worlds\textsc{.m}\textsc{-obl}\textsc{.m} aware \textsc{ind-}be\textsc{.prs}\textsc{-3sg} guidance\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `The Sheikh understood. When he looked at the man’s illness, the lord of all universes realised [what] his illness [was and] guided him (i.e. the Sheikh).'
\z 
 
\ea \label{DG.51}
\textit{ī pose āne gino, pos māro nāwe} \\ 
\gll ī pos=e āne ∅-gin-o pos m-ār-o nāwe \\ 
 \textsc{dem.prox} skin\textsc{.m}\textsc{=dem} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} ind\textsc{-f}all\textsc{.prs}\textsc{-3sg} skin\textsc{.m} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg} inside \\ 
\glt `his skin will fall off, and [a new] skin will replace it'
\z 
 
\ea \label{DG.53}
\textit{šā ʕelādīn īney īney māčo be, fermāwo be pīyā, hīnī birāw pīyākey} \\ 
\gll šā ʕelādīn īney īney m-āč-o be ∅-fermāw-o be pīyā hīn-ī birā=ū pīyā-ekey \\ 
 king\textsc{.m} \textsc{pn} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by man\textsc{.m} \textsc{fill}\textsc{-obl}\textsc{.m} brother\textsc{.m}\textsc{=ez}\textsc{.gen} man\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \\ 
\glt `Shah Aladin said this to the man, to the brother of the [sick] man.'
\z 
 
\ea \label{DG.55}
\textit{ā wextīyekey īne vīyero, qerār bo} \\ 
\gll ā wext-ī-ekey īne vīyer-o qerār ∅-b-o \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} pass\textsc{.prs}\textsc{-3sg} arrangement \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Then, this was settled. The deal was ...'
\z 
 
\ea \label{DG.56}
\textit{fermāwo hošit nešowe bilewe īne derdū ī birāw tone} \\ 
\gll ∅-fermāw-o hoš=it ne-š-o=we bi-l-e=we īne derde=ū ī birā=ū to=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} intelligence\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{compl}} \textsc{imp-}go\textsc{.prs}-\textsc{2sg}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} illness\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} brother\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `Shah Aladin said, ‘Don’t forget! Go back [to your brother]. This is the medicine for your brother’'
\z 
 
\ea \label{DG.59}
\textit{gāhes pī qisē mine xwā řehm kero, weš bo} \\ 
\gll gāhes p=ī qisē-ē min=e xwā řehm ∅-ker-o weš ∅-b-o \\ 
 maybe by\textsc{=dem}.prox talk\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{1sg}\textsc{=dem} God\textsc{.m} mercy\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} well \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Through my intervention may God have mercy on him so that he gets well.'
\z 
 
\ea \label{DG.60}
\textit{šūnū īneyre āhā lā xwāwe inne duʕāš meqboɫe bo} \\ 
\gll šūn=ū īney=re āhā lā xwā=we inne duʕā=š meqboɫ-e ∅-b-o \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{post}} \textsc{intj} at\_the\_place\_of God\textsc{.m}\textsc{=\textsc{post}} so\_much prayer\textsc{\textsc{.f}}\textsc{=3sg} acceptable\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well then, his prayers were accepted by God.'
\z 
 
\ea \label{DG.67}
\textit{šā ʕelādīnīč fermāwo birākem wextē īnē zānā min gunākē māy mē milimre} \\ 
\gll šā ʕelādīn-ī=č ∅-fermāw-o birā-eke=m wext-ē īnē ∅-zān-ā min gunā-ekē m-āy m-ē mil=im=re \\ 
 king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} time\textsc{.m}\textsc{-indf} \textsc{dem.prox}\textsc{.3pl} \textsc{ind-}know\textsc{.prs}\textsc{-3pl} \textsc{1sg} sin\textsc{\textsc{.f}}\textsc{-def}\textsc{.dir}\textsc{\textsc{.f}} \textsc{ind-}come\textsc{.pst}\textsc{.adv} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} shoulder\textsc{.m}\textsc{=1sg}\textsc{=\textsc{post}} \\ 
\glt `Shah Aladin said, ‘Brother!—as they (i.e. the tortoises) know [about the medicine] I won’t shoulder the burden of injuring them.'
\z 
 
\ea \label{DG.73}
\textit{īnē dāstānū ī šēxāū ēme bīyen gewregerīšā} \\ 
\gll īnē dāstāne=ū ī šēx-ā=ū ēme bīye=n gewregerī=šā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{\textsc{.f}}\textsc{.3sg} story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} sheikh\textsc{.m}\textsc{-obl}\textsc{.pl}=\textsc{ez.gen} \textsc{1pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} greatness\textsc{=3pl} \\ 
\glt `This is the story of our Sheikhs and their greatness.'
\z 
 

\ea \label{DP.1}
\textit{dāstānū ī hewrāmānū ēmeyče duē pādšē e čē bā} \\ 
\gll dāstāne=ū ī hewrāmān=ū ēme=īč=e duē pādšā-ē e=čē ∅-b-ā \\ 
 story\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} \textsc{dem.prox} \textsc{pn}\textsc{=ez}\textsc{.gen} \textsc{1pl}\textsc{=add}\textsc{=dem} two king\textsc{.m}\textsc{-dir}\textsc{.pl} in=here \textsc{ind-}be\textsc{.prs}\textsc{-3pl} \\ 
\glt `As for the tales of this Hawraman region of ours, there used to be two kings here.'
\z 
 
\ea \label{DP.5}
\textit{āmān gerekiš bīyen ī memliketī dāgīr kero pey wēš} \\ 
\gll āmā=n gerek=iš bīye=n ī memliket-ī dāgīr ∅-ker-o pey wē=š \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} necessary\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.prox} country\textsc{.m}\textsc{-obl}\textsc{.m} occupied \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} for \textsc{reflx}\textsc{=3sg} \\ 
\glt `He wanted to occupy this region [and control it] for his own interest.'
\z 
 
\ea \label{DP.12}
\textit{ā wextī ī jemšēr šā, sibhān āġey čew deso bīyen} \\ 
\gll ā wext-ī ī jemšēr šā sibhān āġe-ī č=ew des=o bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} \textsc{pn} king\textsc{.m} \textsc{pn} agha\textsc{.m}\textsc{-obl}\textsc{.m} in\textsc{=dem}.dist hand\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Jamsher Shah, eh … Sibhan Agha was on the other side [of the village].'
\z 
 
\ea \label{DP.14}
\textit{ī memlikete desū āđīwe bīyen} \\ 
\gll ī memliket=e des=ū āđī=we bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=dem} hand\textsc{.m}\textsc{=ez}\textsc{.gen} 3sg\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This region was under Sibhan Agha’s control:'\footnote{Lit. This region was in his hands.}
\z 
 
\ea \label{DP.17}
\textit{ā wextīyekey īđīč āmān dāwāw pādšāgerīš kerdēnew} \\ 
\gll ā wext-ī-ekey īđ=yč āmā=n dāwā=w pādšāgerī=š kerdē=ne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} demand\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} kingship\textsc{.m}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}}=and \\ 
\glt `He (i.e., Jamsher Shah) came [and] claimed kingship.'
\z 
 
\ea \label{DP.21}
\textit{ī bāxū bāxātū hewrāmānī, memliketū hewrāmānī, xeɫkū hewrāmānī ā wextīyekey ēme hukimřewiš bīyēnmē} \\ 
\gll ī bāx=ū bāx-āt=ū hewrāmān-ī memliket=ū hewrāmān-ī xeɫk=ū hewrāmān ī ā wext-ī-ekey ēme hukimřew=iš bīyē=nmē \\ 
 \textsc{dem.prox} garden\textsc{.m}\textsc{=ez}\textsc{.gen} garden\textsc{.m}\textsc{-pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} country\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} people\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{pn} \textsc{dem.prox} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} governor\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `We have ruled over\footnote{Lit. We have been rulers.} these gardens of Hawraman, this region of Hawraman, [and] this people of Hawraman.'
\z 
 
\ea \label{DP.22}
\textit{ēme, min īney međo desū towe} \\ 
\gll ēme min īney me-đe-o des=ū to=we \\ 
 \textsc{1pl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{neg-}give\textsc{.prs}\textsc{-3sg} hand\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `I will not hand over these to you.'
\z 
 
\ea \label{DP.23}
\textit{min īney dew deso towe ey wēm kowe bilū} \\ 
\gll min īney ∅-de-ū des=o to=we ey wē=m ko=we bi-l-w \\ 
 \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}give\textsc{.prs}\textsc{-1sg} hand\textsc{.m}\textsc{=\textsc{post}} \textsc{2sg}\textsc{=\textsc{post}} \textsc{intj} \textsc{reflx}\textsc{=1sg} where\textsc{=\textsc{post}} \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} \\ 
\glt `If I hand over these to you, where should I go myself?'
\z 
 
\ea \label{DP.28}
\textit{ī māɫū miɫke hinū āđī bīyenū desšo bīyenū} \\ 
\gll ī māɫ=w miɫk=e hin=ū āđī bīye=n=w des=š=o bīye=n=ū \\ 
 \textsc{dem.prox} property\textsc{.m}=and property\textsc{.m}\textsc{=dem} ez.pron\textsc{=ez}\textsc{.gen} 3sg\textsc{.obl}\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and hand\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `These lands and properties here were his. They were under his control.'
\z 
 
\ea \label{DP.37}
\textit{be xātirū īney qesem wero qesemiš negino} \\ 
\gll be xātir=ū īney qesem ∅-wer-o qesem=iš ne-gin-o \\ 
 by sake\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} oath\textsc{.m} \textsc{sbjv-}out\textsc{-3sg} oath\textsc{.m}\textsc{=3sg} \textsc{neg-}fall\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wanted to swear an oath using the sparrow. He did not want to swear falsely on himself.'
\z 
 
\ea \label{DP.38}
\textit{wāteniš, min tā diɫī ī gīyāne īnā qefesū sīney minne diɫī beyim čāđī nīyā} \\ 
\gll wāte=n=iš min tā diɫ=ī ī gīyān=e īnā=ā qefes=ū sīne=y min=ne diɫ=ī bey=im č=āđī nīye=ā \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} \textsc{1sg} until heart\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{dem.prox} soul\textsc{.m}\textsc{=dem} deic\textsc{=cop}\textsc{.3sg}\textsc{.m} cage\textsc{.m}\textsc{=ez}\textsc{.gen} chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} heart\textsc{.m}\textsc{=ez}.\textsc{attr} bad\textsc{=1sg} from\textsc{=3sg}\textsc{.obl}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He said, ‘I swear on the soul that is in my chest, I am not angry with him\footnote{Lit. My bad heart is not at him} (i.e., Sibhan Agha)'
\z 
 
\ea \label{DP.48}
\textit{māčo, dey min īney pīneym wāten} \\ 
\gll mi-āč-o dey min īney p=īney=m wāte=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=1sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` Jamsher Shah said [to Sibhan Agha], ‘When I swore an oath, I swore on this sparrow;'
\z 
 
\ea \label{DP.49}
\textit{ī gīyānū qefesew sīney minne īnen āđe lūā} \\ 
\gll ī gīyān=ū qefese=ū sīne=y min=ne īne=n āđe lūā \\ 
 \textsc{dem.prox} soul\textsc{.m}\textsc{=ez}\textsc{.gen} cage\textsc{.m}=and chest\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{1sg}\textsc{=\textsc{post}} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} 3sg\textsc{.dir}\textsc{\textsc{.f}} go\textsc{.pst} \\ 
\glt ` The soul in my rib cage [that I swore an oath on] was this [sparrow]. It [just] flew away.'
\z 
 
\ea \label{ZP.3}
\textit{čī degāw hewrāmānīne serāwinū degākēwe je eyē bīyen} \\ 
\gll č=ī degā=ū hewrāmān-ī=ne serāwin=ū degā-ekē=we je eyē bīye=n \\ 
 in=dem.prox village\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} top\_of\_village\textsc{\textsc{=ez.gen}} village\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{post}} from mother\textsc{.obl}\textsc{.f} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He was born in the village of Hawraman, at the top point of the village.'
\z 
 
\ea \label{ZP.26}
\textit{neweše gino herčīw duktirū ī lāū ew lā berāš duktir hīč duktirē ʕelēješ mekero} \\ 
\gll neweš-e ∅-gin-o herčīw duktir=ū ī lā=ū ew lā ∅-ber-ā=š duktir hīč duktir-ē ʕelēje=š me-ker-o \\ 
 ill\textsc{-f} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} how\_often physician\textsc{.m}=and dem.prox at\_the\_place\_of=and \textsc{dem.dist} at\_the\_place\_of \textsc{ind-}take\textsc{.prs}\textsc{-3pl}\textsc{=3sg} physician\textsc{.m} nothing physician\textsc{.m}\textsc{-indf} treatment\textsc{.f}\textsc{=3sg} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt ` No matter how often she was taken to doctors here and there, no doctor could cure her.'
\z 
 
\ea \label{ZP.33}
\textit{ī kināčēw to dermānū derdīš īnā lā ī pīyāywe’} \\ 
\gll ī kināčē=ū to dermān=ū derd-ī=š īnā-∅ lā ī pīyā-ī=we \\ 
 dem.prox daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} \textsc{2sg} treatment\textsc{.m}\textsc{\textsc{=ez.gen}} illness\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} deic\textsc{-3sg}\textsc{.m} at\_the\_place\_of dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \\ 
\glt `‘This daughter of yours, the treatment for her illness lies with this man (i.e., Pir Shaliyar)'
\z 
 
\ea \label{ZP.36}
\textit{īne māčmēš pene šālīyāre sīyāw lālow kināčekēn} \\ 
\gll īne m-āč-mē=š pene šālīyār=e sīyāw lālo=ū kināčē-ekē=n \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to \textsc{pn}\textsc{=dem} black maternal\_uncle\textsc{.m}\textsc{\textsc{=ez.gen}} girl\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This [person], whom we call Shaliyar Siya, is the maternal uncle of the [king’s] daughter.'
\z 
 
\ea \label{ZP.39}
\textit{māčo, ī kināčē bere degā be degāw} \\ 
\gll m-āč-o ī kināčē ∅-ber-e degā be degā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.obl}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} village\textsc{.obl}\textsc{.f} to village\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `[and] said, ‘Take this girl village by village,'
\z 
 
\ea \label{ZP.44}
\textit{to eger ī kināčē berdet xizmetū ī pīyāy} \\ 
\gll to eger ī kināčē berd-e=t xizmet=ū ī pīyā-ī \\ 
 \textsc{2sg} if dem.prox daughter\textsc{.dir}\textsc{.f} take\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=\textsc{2sg}} service\textsc{.m}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `‘When you have taken this girl into the presence of this man'
\z 
 
\ea \label{ZP.45}
\textit{ī pīyāy tāwāš ī kināčēše kerde be qerārē wešeš kerdewe} \\ 
\gll ī pīyā-ī tāwā=š ī kināčē=š=e kerd-e be qerār-ē weš-e=š kerd-e=we \\ 
 dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m} can\textsc{.pst}\textsc{=3sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} do\textsc{.pst}\textsc{-f}\textsc{.3sg} to arrangement\textsc{-indf} well\textsc{-f}\textsc{=3sg} do\textsc{.pst}\textsc{-f}\textsc{.3sg}\textsc{=compl} \\ 
\glt `[and if] the man has been able to cure the girl'
\z 
 
\ea \label{ZP.48}
\textit{ī kināčē nārīwe ētir pey min her ēnne weše bowe’} \\ 
\gll ī kināčē ne-ār-ī=we ētir pey min her ēnne weš-e b-o=we \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} any\_more for \textsc{1sg} emph as\_much\_as well\textsc{-f} be\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `Please do not bring this daughter [of mine] back to me as long as she is cured.'
\z 
 
\ea \label{ZP.67}
\textit{tewejo kero wextē miđyo ī ženīū ī pīyā mēnē řāře duē dēwē īnē dewršāne} \\ 
\gll tewejo ∅-ker-o wext-ē mi-đye-o ī ženī=ū ī pīyā m-ē-nē řā=ře duē dēw-ē īnē dewr=šā=ne \\ 
 caution \textsc{ind-}do\textsc{.prs}\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} dem.prox woman\textsc{.f}=and dem.prox man\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{-3pl} road\textsc{.f}\textsc{=\textsc{post}} two demon\textsc{.m}\textsc{-dir}\textsc{.pl} dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} around\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `He noticed that when these two persons were on their way, they were surrounded by two demons.'
\z 
 
\ea \label{ZP.86}
\textit{māčā xeɫkīč, ī pīyā šimeš pey persdē xo yānēw e čāneš nīyāw} \\ 
\gll m-āč-ā xeɫk=īč ī pīyā šime=š pey ∅-pers-dē xo yāne-ē=ū e=čāne=š nīye=ā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} people\textsc{.m}\textsc{=add} dem.prox man\textsc{.m} \textsc{2pl}\textsc{=3sg} for \textsc{ind-}ask\textsc{.prs}-\textsc{2pl} dis.ptc house\textsc{.m}\textsc{-indf}\textsc{\textsc{=ez.gen}} in=like\_that\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `The people said, ‘This man whom you ask for, he does not have a proper house.'
\z 
 
\ea \label{ZP.87}
\textit{ženīū yāneš nīyārē īnīšā} \\ 
\gll ženī=ū yāne=š nīye=ā=rē īnīšā \\ 
 woman\textsc{.f}=and house\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He has no wife, nor a house, and so on.'
\z 
 
\ea \label{ZP.88}
\textit{ā wextīekey ʕerz kerā, dey ēme ī kināčē jeryāniš e čīnew} \\ 
\gll ā wext-ī-ekey ʕerz ∅-ker-ā dey ēme ī kināčē jeryān=iš e=čīne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} petition\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{1pl} \textsc{dem.prox} girl\textsc{.f} story\textsc{.m}\textsc{=3sg} in=like=and \\ 
\glt `Then they (her uncle) said, ‘This girl’s story is like this.'
\z 
 
\ea \label{ZP.90}
\textit{mišo māreš biřū pey ī šexsīw ī šexse eger ī kināčēše weše kerde,} \\ 
\gll mišo māre=š ∅-biř-ū pey ī šexs-ī=ū ī šexs=e eger ī kināčē=š=e weš-e kerd-e \\ 
 aux marriage\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for dem.prox saint\textsc{-obl}\textsc{.m}=and dem.prox saint\textsc{=dem} if dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f} \\ 
\glt `I shall marry her to this person [on the condition that] this person has healed this girl.'
\z 
 
\ea \label{ZP.91}
\textit{ī kināčē weše bīyēnew īnīšā’} \\ 
\gll ī kināčē weš-e bīyē=ne=ū īnīšā \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} well\textsc{-f} be\textsc{.ptcp}\textsc{.pl}\textsc{=\textsc{post}}=and dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `The girl is now healed, and so on.'
\z 
 
\ea \label{ZP.98}
\textit{māčo, min īneyā jeryānim e čīnenū} \\ 
\gll m-āč-o min īne=ā jeryān=im e=čīne=n=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} story\textsc{.m}\textsc{=\textsc{1sg}} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He said, ‘I— look, my story is like this.'
\z 
 
\ea \label{ZP.100}
\textit{ī kināčē eger weše bīyowe mišyo māreš biřū pey to be mezebū wētāw} \\ 
\gll ī kināčē eger weš-e bīyē-o=we mišyo māre=š ∅-biř-ū pey to be mezeb=ū wē=tā=ū \\ 
 dem.prox girl\textsc{.obl}\textsc{.f} if well\textsc{-f} be\textsc{.prs}.\textsc{cond}\textsc{.f}?\textsc{-3sg}\textsc{=compl} aux marraige\textsc{=3sg} \textsc{sbjv-}cut\textsc{.prs}\textsc{-\textsc{1sg}} for \textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `If the girl is healed, I shall marry her to you, according to your religion.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.102}
\textit{īne pādšāzāyēnew derdū min meweronew} \\ 
\gll īne pādšāzāyē=ne=ū derd=ū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}=and illness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She is no use to me.'
\z 
 
\ea \label{ZP.103}
\textit{min metāwū īne nigebānīye kerū} \\ 
\gll min me-tāw-ū īne nigebānī-e ∅-ker-ū \\ 
 \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} guardian\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt ` I cannot take care of her.'
\z 
 
\ea \label{ZP.104}
\textit{īne pādšāzāyēne čenū min meweronew} \\ 
\gll īne pādšāzāyē=ne čenū min me-wer-one=ū \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}} with\textsc{.ez.gen} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `She is a princess. She will not eat with me [under the same roof].'
\z 
 
\ea \label{ZP.105}
\textit{min čīwēwim nīyārē lāyiqū ī pādšāzāye bo} \\ 
\gll min čīw-ēw=im nīye=ā=rē lāyiq=ū ī pādšāzāye ∅-b-o \\ 
 \textsc{1sg} thing\textsc{-indf}\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}} deserved\textsc{\textsc{=ez.gen}} dem.prox princess\textsc{.obl}\textsc{.f} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `I have nothing which would be up to the standards of this princess.'
\z 
 
\ea \label{ZP.115}
\textit{māčā, dey īne ī pīyā ī ženīče ārdēne} \\ 
\gll m-āč-ā dey īne ī pīyā ī ženī=č=e ārdē=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox man\textsc{.m} dem.prox woman\textsc{.dir}f\textsc{=add}\textsc{=dem} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `[and] said, ‘Well, this man has brought this woman [here]'
\z 
 
\ea \label{ZP.117}
\textit{īđīč hīč minīč hīčim nīyārēw} \\ 
\gll īđ=īč hīč min=īč hīč=im nīye=ā=rē=ū \\ 
 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} nothing \textsc{1sg}\textsc{=add} nothing\textsc{=\textsc{1sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `She [has] nothing, nor do I.'
\z 
 
\ea \label{ZP.126}
\textit{ā wextī pīr šelīyār fermāwo, šime īne māčdē řāstā wāt} \\ 
\gll ā wext-ī pīr šelīyār ∅-fermāw-o šime īne m-āč-dē řās=tā wāt \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{2pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} \\ 
\glt `Then Pir Shaliyar said, ‘What you are saying is right.’\footnote{Lit. You said this; you told the truth.}'
\z 
 
\ea \label{ZP.128}
\textit{weɫāhī īne šime māčdēšū īnā vistāwe řā pey min čūn šime be xātirū xwāytā kerd hetā xwā dinyāš weše kerdēne ī merāsēme her bo her meno’} \\ 
\gll weɫāhī īne šime m-āč-dē=š=ū īnā-∅ vist=tā=we řā pey min čūn šime be xātir=ū xwā-ī=tā kerd hetā xwā dinyā=š weš-e kerdē=ne ī merāsēm=e her ∅-b-o her ∅-men-o \\ 
 indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} \textsc{ind-}say\textsc{.prs}-\textsc{2pl}\textsc{=3sg}=and deic\textsc{-3sg}\textsc{.m} throw\textsc{.pst}=\textsc{2pl}\textsc{=compl} road\textsc{.f} for \textsc{1sg} because \textsc{2pl} to sake\textsc{.m}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}be\textsc{.prs}\textsc{-3sg} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg} \\ 
\glt `Indeed, what you are saying and the way you organised [this ceremony] for me, since you did it in good faith, so long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ZP.130}
\textit{hetā xwāy dinyāš weše kerdēne ī merāsēme kem mekero} \\ 
\gll hetā xwā-ī dinyā=š weš-e kerdē=ne ī merāsēm=e kem me-ker-o \\ 
 as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox ceremony\textsc{.m}\textsc{=dem} small \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `So long as the world continues to exist, this ceremony will keep on being held [each year]'
\z 
 
\ea \label{ŽP.44}
\textit{ī yewēše šānēnē pī zemīnne} \\ 
\gll ī yewe-ē=š=e šānē=nē p=ī zemīn=ne \\ 
 dem.prox barley\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3sg}\textsc{=dem} scatter\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} at=dem.prox land\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Has he planted the seeds in the field?'
\z 
 
\ea \label{ŽP.45}
\textit{yā ī yewēše zāxēre kerdēnē?} \\ 
\gll yā ī yewe-ē=š=e zāxēre kerdē=nē \\ 
 either dem.prox barley\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=3sg}\textsc{=dem} sotrage do\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Or has he [instead] stored the seeds [for himself]'
\z 
 
\ea \label{ŽP.49}
\textit{īneš netāwān kero} \\ 
\gll īne=š ne-tāwā=n ∅-ker-o \\ 
 dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{neg-}can\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He cannot do these [tasks].’'
\z 
 
\ea \label{ŽP.71}
\textit{pīr heme yoso wēš milo temāšeš kero miđyo āne īnā hewāwe zāno īne ētir kārī xwāyīn} \\ 
\gll pīr heme yoso wē=š mi-l-o temāše=š ∅-ker-o mi-đye-o āne īnā-∅ hewā=we ∅-zān-o īne ētir kār-ī xwāyī=n \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{reflx}\textsc{=3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} looking\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} task\textsc{.m}\textsc{-obl}\textsc{.m} divine\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Hama Yoso went [and] looked at him; he noticed that he (Pir Shaliyar) was aloft in the air [up until a certain height] and came to realise that that was God’s work.'
\z 
 
\ea \label{ŽP.74}
\textit{māčo min īnem dī be sey misefāywe} \\ 
\gll m-āč-o min īne=m dī-∅ be sey misefā-ī=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1sg}} see\textsc{.pst}\textsc{-3sg}\textsc{.m} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘I witnessed this about Say Mustafa.’'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.135}
\textit{suāleš čene kero, māčo, čī āmānīwe pey īnīšā?’} \\ 
\gll suāle=š čene ∅-ker-o m-āč-o čī āmā=nī=we pey īnīšā \\ 
 question\textsc{.f}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} why come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}}\textsc{=compl} to dem.prox\textsc{.obl}\textsc{.3pl} \\ 
\glt `He asked him [and] said, ‘Why have you come here?\footnote{Lit. for these.}’'
\z 
 
\ea \label{ŽP.138}
\textit{māčoš lāwe māčo min pey īney āmānā āmānāw} \\ 
\gll m-āč-o=š lā=we m-āč-o min pey īney āmā=nā āmā=nā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} to dem.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}=and \\ 
\glt `He said to him, ‘I have come for this [reason]’'
\z 
 
\ea \label{ŽP.154}
\textit{wermekešne māčāš pene menteqew hewrāmānātī pīyāwē hen tāwo ī kināčē to weše kerowe} \\ 
\gll werm-eke=š=ne m-āč-ā=š pene menteqe=ū hewrāmān-āt-ī pīyā-ē hen-∅ ∅-tāw-o ī kināčē to weš-e ∅-ker-o=we \\ 
 sleep\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} to region=\textsc{ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{-obl}\textsc{.m} man\textsc{.m}\textsc{-indf} \textsc{exist}\textsc{-3sg}\textsc{.m} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `In the dream, it was said to him that there was a man in the Hawraman region who could cure his daughter.'
\z 
 
\ea \label{ŽP.156}
\textit{māčā, ī kināčē to tāwo weše kerowe’} \\ 
\gll m-āč-ā ī kināčē to ∅-tāw-o weš-e ∅-ker-o=we \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{2sg} \textsc{ind-}can\textsc{.prs}\textsc{-3sg} well\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `It was said [to him in the dream], ‘He can cure this daughter of yours.’'
\z 
 
\ea \label{ŽP.159}
\textit{to bo lue! ī kināčē bere} \\ 
\gll to b-o ∅-lu-e ī kināčē ∅-ber-e \\ 
 \textsc{2sg} \textsc{imp-}come\textsc{.prs}\textsc{.imp}\textsc{.\textsc{2sg}} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `‘Come [here and] go! Take the girl.'
\z 
 
\ea \label{ŽP.162}
\textit{dēhāt be dēhāt ī kināčē bere} \\ 
\gll dēhāt be dēhāt ī kināčē ∅-ber-e \\ 
 village\textsc{.f} by village\textsc{.f} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `Take the girl village by village.'
\z 
 
\ea \label{ŽP.163}
\textit{eger to ī kināčēte berdeū ī pīyā her kes hen} \\ 
\gll eger to ī kināčē=t=e berd-e=ū ī pīyā her kes hen-∅ \\ 
 if \textsc{2sg} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} take\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} dem.prox man\textsc{.m} emph person\textsc{.m} exist\textsc{-3sg}\textsc{.m} \\ 
\glt `If you take the girl, [no matter] who this man (the doctor) is,'
\z 
 
\ea \label{ŽP.164}
\textit{tāwāš ī duktire ī kināčēše weše kerdew} \\ 
\gll tāwā=š ī duktir=e ī kināčē=š=e weš-e kerd-e=ū \\ 
 can\textsc{.pst}\textsc{=3sg} dem.prox phy\textsc{=dem} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3sg}\textsc{=dem} well\textsc{-f} do\textsc{.pst} 3sg\textsc{.f}\textsc{\textsc{=ez.gen}} \\ 
\glt `If the doctor can cure the girl,'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.175}
\textit{tumez mekere ī āy wextē čo jimāyre yāwā gerūberū silēnēū nāwēū e pāgene nizīkē bāwe pīr šelīyār e čē xeber zāno} \\ 
\gll tumez me-ker-e ī āy wext-ē č=o ∅-jim-ā=re ∅-yāw-ā gerūber=ū silēnē=ū nāwē=ū e=pāgene nizīk-ē b-ā=we pīr šelīyār e=čē xeber ∅-zān-o \\ 
 surprisingly \textsc{neg-}do\textsc{.prs}-\textsc{2sg} dem.prox \textsc{intj} time\textsc{.m}\textsc{-indf} in=\textsc{dem.dist} \textsc{ind-}move\textsc{.prs}\textsc{-3pl}\textsc{=\textsc{povb}} \textsc{ind-}arrive\textsc{.prs}\textsc{-3pl} around\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{.obl}\textsc{.f}\textsc{\textsc{=ez.gen}} in=there close\textsc{\textsc{-pl}} be\textsc{.prs}\textsc{-3pl}\textsc{=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `It turned out\footnote{Lit. You wouldn’t say/see!} that he (Pir Shaliyar) became aware of [the sound] ‘Oh’ [that the girl said] while they were walking and were getting close to somewhere between Slen and Naw.'
\z 
 
\ea \label{ŽP.177}
\textit{ī dēwēnē ī kināčēšā bestēne} \\ 
\gll ī dēw=ē=nē ī kināčē=šā bestē=ne \\ 
 dem.prox demon\textsc{.m}\textsc{=ez}.\textsc{attr}\textsc{=cop}\textsc{.3pl} dem.prox girl\textsc{.dir}\textsc{.f}\textsc{=3pl} tie\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `It was the demons who had muted the girl.'
\z 
 
\ea \label{ŽP.178}
\textit{nitqšā girtēne ī due dēwene} \\ 
\gll nitq=šā girtē=ne ī due dēwe=ne \\ 
 speech\textsc{=3pl} grab\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} dem.prox two\textsc{.f} demon\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `They had muted her speech, these two demons.'
\z 
 
\ea \label{ŽP.204}
\textit{perso, jeryān čikonū? ī yānew kābrāy kēnū īne kēnū?} \\ 
\gll ∅-pers-o jeryān čiko=n=ū ī yāne=ū kābrā-ī kē=n=ū īne kē=n=ū \\ 
 \textsc{ind-}ask\textsc{.prs}\textsc{-3sg} story\textsc{.m} where\textsc{=cop}\textsc{.3sg}\textsc{.m}=and dem.prox house\textsc{.m}=and man\textsc{-obl}\textsc{.m} who\textsc{=cop}\textsc{.3sg}\textsc{.m}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} who\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `He came to the village. He asked, ‘[what is the story]? Where is the house of the fellow? Who is he? Who is he?'
\z 
 
\ea \label{ŽP.207}
\textit{mišyo ī kināčē nīkāh kerū pey toū be mezebū wētāū} \\ 
\gll mišyo ī kināčē nīkāh ∅-ker-ū pey to=ū be mezeb=ū wē=tā=ū \\ 
 aux dem.prox girl\textsc{.dir}\textsc{.f} marriage\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} to \textsc{2sg}=and by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.213}
\textit{īnē be min mewero kināčē pādšāy,} \\ 
\gll īnē be min me-wer-o kināčē pādšā-ī \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} with \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg} girl\textsc{.dir}\textsc{.f} king\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `She will not eat with me [under the same roof], the king’s daughter.'
\z 
 
\ea \label{ŽP.217}
\textit{fermāwo, min īne nānmārēū} \\ 
\gll ∅-fermāw-o min īne nān=m=ā=rē=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{povb}}=and \\ 
\glt `He said, ‘This is my meal.’'
\z 
 
\ea \label{ŽP.222}
\textit{māčā, īne čēšā?} \\ 
\gll m-āč-ā īne čēš=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They (the people) said, ‘What is this?’'
\z 
 
\ea \label{ŽP.223}
\textit{īnē ī qisē čēšene? čēš māčdē? māčā čēš?} \\ 
\gll īnē ī qisē čēše=ne čēš m-āč-dē m-āč-ā čēš \\ 
 dem.prox\textsc{.dir}\textsc{.f}\textsc{.3sg} dem.prox talk\textsc{.dir}\textsc{.f} what\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} what \textsc{ind-}say\textsc{.prs}-\textsc{2pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} what \\ 
\glt `What is this talk? What are you saying? What are they saying?'
\z 
 
\ea \label{ŽP.225}
\textit{her lāloeke māčo, mišyo ī kināčē min meberūšo māreš biřūnew’} \\ 
\gll her lālo-eke m-āč-o mišyo ī kināčē min me-ber-ū=š=o māre=š ∅-biř-ūne=ū \\ 
 emph maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} aux dem.prox girl\textsc{.dir}\textsc{.f} \textsc{1sg} \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} marraige\textsc{=3sg} \textsc{ind-}cut\textsc{.prs}\textsc{-\textsc{1sg}}=and \\ 
\glt ` The uncle kept saying, ‘I shall not take this girl back. I shall marry her [to Pir Shaliyar].’'
\z 
 
\ea \label{ŽP.226}
\textit{āđīč māčo, min metāwū ī kināčē wey kerū īne nānmā} \\ 
\gll āđ=īč m-āč-o min me-tāw-ū ī kināčē wey ∅-ker-ū īne nān=m=ā \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox girl\textsc{.dir}\textsc{.f} raising \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `People … he (Pir Shaliyar) said, ‘I cannot take care of this girl. This is my food.’'
\z 
 
\ea \label{ŽP.242}
\textit{qey mekero, mādām desūr e pēsen pādšā řāzīe nīā ī ženī bilowe ī kināčē bilowe qey mekero!’} \\ 
\gll qey me-ker-o mādām desūr e=pēse=n pādšā řāzīe nīye=ā ī ženī bi-l-o=we ī kināčē bi-l-o=we qey me-ker-o \\ 
 limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} as\_long\_as task/order\textsc{.m} in=like\textsc{=cop}\textsc{.3sg}\textsc{.m} king\textsc{.m} satisfied \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} dem.prox woman\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} dem.prox girl\textsc{.dir}\textsc{.f} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `There is no problem. Now that things have developed this way, [and] the king is not happy for the girl to return, it’s all right! [She can stay here]’'
\z 
 
\ea \label{ŽP.243}
\textit{pīr šelīyār fermāwo, tehāā weɫāhī īne šime desūrtā dāw wāttā čūnke āđē be diɫtā wāt řāstā wāt hetā xwāy dinyāš weše kerdēne her menowe kemīč mekero her zīyā kero} \\ 
\gll pīr šelīyār ∅-fermāw-o tehāā weɫāhī īne šime desūr=tā dā=ū wāt=tā čūnke āđē be diɫ=tā wāt řās=tā wāt hetā xwā-ī dinyā=š weš-e kerdē=ne her ∅-men-o=we kem=īč me-ker-o her zīyā ∅-ker-o \\ 
 \textsc{pn} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{intj} indeed dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{2pl} task/order\textsc{.m}=\textsc{2pl} give\textsc{.pst}=and say\textsc{.pst}=\textsc{2pl} since 3pl\textsc{.dir} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} truth=\textsc{2pl} say\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} small\textsc{=add} \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increased \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `Pir Shaliyar said, ‘Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] continues to exist; it will not be diminished; it will grow.'
\z 
 
\ea \label{ŽP.244}
\textit{ī bezme šime wāttāw ī čāštīū īne ī desūre šime ījrātā kerd hetā xwāy dinyāš weše kerdēne šime řāstā wāt be diɫtā wāt} \\ 
\gll ī bezm=e šime wāt=tā=ū ī čāštī=ū īne ī desūr=e šime ījrā=tā kerd hetā xwā-ī dinyā=š weš-e kerdē=ne šime řās=tā wāt be diɫ=tā wāt \\ 
 dem.prox jollification\textsc{.dir}\textsc{.m}\textsc{=dem} \textsc{2pl} say\textsc{.pst}=\textsc{2pl}=and dem.prox meal\textsc{.f}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox task/order\textsc{.m}\textsc{=dem} \textsc{2pl} performance=\textsc{2pl} do\textsc{.pst} as\_long\_as God\textsc{.m}\textsc{-obl}\textsc{.m} world\textsc{.dir}\textsc{.f}\textsc{=3sg} well\textsc{-f} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \textsc{2pl} truth=\textsc{2pl} say\textsc{.pst} by heart\textsc{.m}=\textsc{2pl} say\textsc{.pst} \\ 
\glt `[After] the things you said, and the meals which you prepared, so long as the world continues to exist, you are right [in so doing]; you did these [things] in good faith.'
\z 
 
\ea \label{ŽP.245}
\textit{ī merāsēme her menowe hīčkeyč kem mekero her zīyāy kero īlāy žīyāyne bo zīyāyiš pene kero} \\ 
\gll ī merāsēm=e her ∅-men-o=we hīčke=yč kem me-ker-o her zīyāy ∅-ker-o īlāy žīyāy=ne ∅-b-o zīyāy=iš pene ∅-ker-o \\ 
 dem.prox ceremony\textsc{.m}\textsc{=dem} emph \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} nothing\textsc{=add} little \textsc{neg-}do\textsc{.prs}\textsc{-3sg} emph increase \textsc{ind-}do\textsc{.prs}\textsc{-3sg} for\_sure increased\textsc{=\textsc{post}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} increased\textsc{=3sg} to \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `This ceremony will be repeated, it won’t diminish. It will continue to grow. It will grow [into bigger ceremonies]'
\z 
 
\ea \label{ŽP.262}
\textit{lue! to luewe hem wesāylīč bere pey ī pīyāyū ī kināčēw} \\ 
\gll ∅-lu-e to ∅-lu-e=we hem wesāyl=īč ∅-ber-e pey ī pīyā-ī=ū ī kināčē=ū \\ 
 \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg}\textsc{=\textsc{post}} both appliance\textsc{.m}\textsc{=add} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} to dem.prox man\textsc{.m}\textsc{-obl}\textsc{.m}=and dem.prox daughter\textsc{.dir}\textsc{.pl}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter'
\z 
 
\ea \label{BP.3}
\textit{yānē yānē meseleyš serū īneyn yānē bābā xwāy dā, kuřēš peyā bī} \\ 
\gll yānē yānē mesele-ī=š ser=ū īne-ī=n yānē bābā xwā-ī dā kuř-ē=š peyā bī-∅ \\ 
 that\_means that\_means issue\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} that\_means grandfather\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} give\textsc{.pst} son\textsc{.m}\textsc{-indf}\textsc{=3sg} visible be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `The point with his name is that God granted the grandfather [a son]. A boy  was born to him .\footnote{Lit. A son of him appeared.}'
\z 
 
\ea \label{BP.9}
\textit{ā zemān bābā xwāđā bīyen, ā zemān ī mexlūqū hewrāmānātū ēme reʕīsšā je ʕērāqew čā kurdewārīū ʕērāqīne bīyen} \\ 
\gll ā zemān bābā xwāđā bīye=n ā zemān ī mexlūq=ū hewrāmān-āt=ū ēme reʕīs=šā je ʕērāq=ew čā kurdewārī=ū ʕērāq-ī=ne bīye=n \\ 
 \textsc{dem.dist} time\textsc{.m} grandfather\textsc{.m} \textsc{pn} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{dem.dist} time\textsc{.m} \textsc{dem.prox} people\textsc{=ez.gen} \textsc{pn}\textsc{\textsc{-pl}}\textsc{=ez.gen} \textsc{1pl} leader\textsc{.m}\textsc{=3pl} from \textsc{pn}\textsc{=\textsc{post}} there kurdish\_region\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `At the time of Babab Khwada, the people of Hawraman were under Iraq’s rule, under the control of Kurdish regions in Iraq.'
\z 
 
\ea \label{BP.20}
\textit{jā ī dāstānē hinū bābāw pīr šelīyārīne} \\ 
\gll jā ī dāstāne-ē hin=ū bābā=ū pīr šelīyār-ī=ne \\ 
 afterwards \textsc{dem.prox} story.f\textsc{-obl}.f ez.pron\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `Anyhow, this story is about the grandfather of Pir Shaliyar.'
\z 
 
\ea \label{BP.24}
\textit{e čē biřew bārešā sānān je hewrāmānī pēse begzāyekāw ī beynū ēme} \\ 
\gll e=čē biře=ū bāre=šā sānā=n je hewrāmān-ī pēse begzāye-ekā=ū ī beyn=ū ēme \\ 
 in=here portion\textsc{.m}=and profit\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn}\textsc{-obl}\textsc{.m} like nobleman\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=ez.gen} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `[who] would collect taxes from Hawraman, [just] like our noblemen [in recent] times.'
\z 
 
\ea \label{BP.38}
\textit{ī kābrēče je ʕērāqo āmēnē, wēregānew nīmeřonew seʕbne bexšnāyšā bexšnēnēwe} \\ 
\gll ī kābrā-ē=č=e je ʕērāq=o āmē=nē wēregā=ne=ū nīmeřo=ne=ū seʕbe=ne bexšnāy=šā bexšn-ēn-ē=we \\ 
 \textsc{dem.prox} man\textsc{-dir}\textsc{.pl}\textsc{=add}\textsc{=dem} from \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} evening\textsc{=\textsc{post}}=and noon\textsc{=\textsc{post}}=and morning.f\textsc{=\textsc{post}} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-indf}\textsc{=compl} \\ 
\glt `They (people) would donate [food] to the fellows (the tax collectors) who had come from Iraq, in the evenings, mornings, and at noon. '
\z 
 
\ea \label{BP.40}
\textit{her wēregāne pēse čāweɫ sempāšē bēnē, yoīšā dēnē duīšā dēnē yānew yoy ī meʕmūrīye ke ʕērāqo āmēnē, ā zemānū pīr šelīyārī bexšnāyšā bexšnēnē} \\ 
\gll her wēregā=ne pēse čāweɫ sempāš-ē b-ēn-ē yo-ī=šā de-ēn-ē duē-ī=šā de-ēn-ē yāne=ū yo-ī ī meʕmūr-ī=e ke ʕērāq=o āmē=nē ā zemān=ū pīr šelīyār-ī bexšnāy=šā bexšn-ēn-ē \\ 
 just evening\textsc{=\textsc{post}} like in\_the\_past poison\_sprayer\textsc{-dir}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl} one\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} two\textsc{-obl}\textsc{.m}\textsc{=3pl} give.prs\textsc{-pstc}\textsc{-3pl} house\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox} officer\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=dem} rel \textsc{pn}\textsc{=\textsc{post}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} distribute\textsc{.inf}\textsc{=3pl} distribute.prs\textsc{-pstc}\textsc{-3pl} \\ 
\glt `As in the old days when there were people whose job was spraying pesticide on trees, [and] each would be hosted at the house of one [host], the officers who had come from Iraq would be fed in the time of Pir Shaliyar. '
\z 
 
\ea \label{BP.43}
\textit{yānew yo terī, her šewē her wēregāne ī mēmānē mēmānū yoy bīyēne} \\ 
\gll yāne=ū yo ter-ī her šewe-ē her wēregā=ne ī mēmān-ē mēmān=ū yo-ī bīyē=ne \\ 
 house\textsc{.m}\textsc{=ez.gen} one\textsc{.m} other\textsc{-obl}\textsc{.m} just night.f\textsc{-obl}.f just evening\textsc{=\textsc{post}} \textsc{dem.prox} guest\textsc{.m}\textsc{-dir}\textsc{.pl} guest\textsc{.m}\textsc{=ez.gen} one\textsc{.m}\textsc{-obl}\textsc{.m} be\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `[likewise, one tax collector] would be the guest in the house of one [fellow from Hawraman]  each evening, each night.'
\z 
 
\ea \label{BP.45}
\textit{milā xeɫk pāne milo miđyo īne ā xerīkū genekārīnē ī kābrāy je āmēnē, ī meʕmūrē xerīkū genekārīnēw} \\ 
\gll mi-l-ā xeɫk p=ā=ne mi-l-o mi-đye-o īne=ā xerīk=ū genekārī=nē ī kābrā-ī je āmē=nē ī meʕmūr-ē xerīk=ū genekārī=nē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3pl} people\textsc{.m} at=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}look.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} man\textsc{-obl}\textsc{.m} from come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `People realised that these men were busy engaging in debauchery; these officers were busy engaing in debauchery.'
\z 
 
\ea \label{BP.47}
\textit{māčā, ī īnē āmēnē melāwe’} \\ 
\gll m-āč-ā ī īnē āmē=nē me-l-ā=we \\ 
 \textsc{ind-}say.prs\textsc{-3pl} \textsc{dem.prox} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They (people) said, ‘These (tax collectors) have come here; they have not gone back yet.’'
\z 
 
\ea \label{BP.53}
\textit{mēmānekeyš sere biřo māđām īne kāršā bīyen} \\ 
\gll mēmān-ekey=š sere ∅-biř-o māđām īne kār=šā bīye=n \\ 
 guest\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3sg} head\textsc{.m} \textsc{sbjv-}cut.prs\textsc{-3sg} as\_long\_as \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} task\textsc{.m}\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Since they (the tax collectors) are involved in this (debauchery), everyone should decapitate his guest.'
\z 
 
\ea \label{BP.56}
\textit{jā girđmā bilmē pey ʕērāqī pey lāw ī āġeyšā bilmērewo,} \\ 
\gll jā girđ=mā bi-l-mē pey ʕērāq-ī pey lā=ū ī āġe-ī=šā bi-l-mē=re=o \\ 
 afterwards all\textsc{=\textsc{1pl}} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}} for \textsc{pn}\textsc{-obl}\textsc{.m} for at\_the\_place\_of\textsc{=ez.gen} \textsc{dem.prox} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}go.prs\textsc{-\textsc{1pl}}\textsc{=\textsc{povb}}\textsc{=compl} \\ 
\glt `Then, we all go to Iraq, to their agha. We go down there.'
\z 
 
\ea \label{BP.57}
\textit{āđ māčo, konē ī meʕmūrē mine činnē wextēn? je hewrāmān bīyēnē āmēnē pey biřew bāre} \\ 
\gll āđ m-āč-o ko=nē ī meʕmūr-ē min=e činnē wext-ē=n je hewrāmān bīyē=nē āmē=nē pey biře=ū bāre \\ 
 3sg\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} where\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{1sg}\textsc{=dem} some\textsc{.pl} time\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} from \textsc{pn} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} for portion\textsc{.m}=and profit\textsc{.m} \\ 
\glt `He (the agha) says, ‘where have my officers been during this time? They were in Hawraman; they had gone to collect tax.'
\z 
 
\ea \label{BP.60}
\textit{ēmeyč māčmē, ī meʕmūrē šime kīyāsēnē bā biřew bāremā dēw nāmēwe bārāšo} \\ 
\gll ēme=yč m-āč-mē ī meʕmūr-ē šime kīyāsē=nē ∅-b-ā biře=ū bāre=mā dā-ē=ū ne-āmā-ē=we b-ār-ā=š=o \\ 
 \textsc{1pl}\textsc{=add} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{ind-}be.prs\textsc{-3pl} portion\textsc{.m}=and profit\textsc{.m}\textsc{=\textsc{1pl}} give\textsc{.pst}\textsc{-3pl}=and \textsc{neg-}come\textsc{.pst}\textsc{-3pl}\textsc{=compl} \textsc{sbjv-}bring.prs\textsc{-3pl}\textsc{=3sg}\textsc{=compl} \\ 
\glt `We will say, ‘We paid tax to the officers whom you had sent, [however,] they didn’t return to bring it back.'
\z 
 
\ea \label{BP.63}
\textit{ēmeyč muxteserū kelāmī ā īney wāčmē,} \\ 
\gll ēme=yč muxteser=ū kelām-ī ā īney wāč-mē \\ 
 \textsc{1pl}\textsc{=add} summary\textsc{.m}\textsc{=ez.gen} speech\textsc{-obl}\textsc{.m} \textsc{dem.dist} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} say.prs.sbjv\textsc{-\textsc{1pl}} \\ 
\glt `As for us, to cut a long story short, we say this.'
\z 
 
\ea \label{BP.64}
\textit{jā īne qewɫ kerā yo} \\ 
\gll jā īne qewɫ ∅-ker-ā yo \\ 
 afterwards \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} arrangement\textsc{.m} \textsc{ind-}do.prs\textsc{-3pl} one\textsc{.m} \\ 
\glt `Then, they (the people of Hawraman) all agreed [about the plan]\footnote{Lit. they made their oaths one.}'
\z 
 
\ea \label{BP.66}
\textit{īney māčā beynū wēšāne} \\ 
\gll īney m-āč-ā beyn=ū wē=šā=ne \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-3pl} among\textsc{=ez.gen} \textsc{reflx}\textsc{=3pl}\textsc{=\textsc{post}} \\ 
\glt `They said this to each other,'
\z 
 
\ea \label{BP.68}
\textit{ēme milmē īney māčmē} \\ 
\gll ēme mi-l-mē īney m-āč-mē \\ 
 \textsc{1pl} \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \\ 
\glt `Then, we will go [to Iraq] and say this (i.e. what we agreed on)'
\z 
 
\ea \label{BP.69}
\textit{īney māčmē.’} \\ 
\gll īney m-āč-mē \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \\ 
\glt `We will say this.’'
\z 
 
\ea \label{BP.71}
\textit{neferē yāneš pī ser degāwe bo, tirēštew kārdū čīwiš mebo} \\ 
\gll nefer-ē yāne=š p=ī ser degā=we ∅-b-o tirēšte=ū kārdī=ū čīw=iš me-b-o \\ 
 person\textsc{.m}\textsc{-indf} house\textsc{.m}\textsc{=3sg} at=\textsc{dem.prox} on village.f\textsc{=\textsc{post}} \textsc{ind-}be.prs\textsc{-3sg} ax\textsc{.m}=and knife.f\textsc{=ez.gen} thing\textsc{.m}\textsc{=3sg} \textsc{neg-}be.prs\textsc{-3sg} \\ 
\glt `The house of one of the locals was at the top of the village. He had neither axe, nor knife, nor anything else. '
\z 
 
\ea \label{BP.76}
\textit{miđyā īne be mišārī sereš biřyen nāmā nāmā nīyā} \\ 
\gll mi-đye-ā īne be mišār-ī sere=š biřye=n ne-āmā-∅ ne-āmā-∅ nīye=ā \\ 
 \textsc{ind-}look.prs\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} to hand\_saw\textsc{.m}\textsc{-obl}\textsc{.m} head\textsc{.m}\textsc{=3sg} cut\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{neg-}come\textsc{.pst}\textsc{-3sg}\textsc{.m} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They noticed that the one who had decapitated his guests with a hand saw hadn’t come to the mosque. He hadn’t arrived. He wasn’t there.'
\z 
 
\ea \label{BP.79}
\textit{milāymē pey ʕērāqī milāymē ī xeberī bermēwe} \\ 
\gll mi-lā-ymē pey ʕērāq-ī mi-lā-ymē ī xeber-ī ∅-ber-mē=we \\ 
 \textsc{ind-}go.prs \textsc{1pl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}go.prs \textsc{1pl} \textsc{dem.prox} news\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}take.prs\textsc{-\textsc{1pl}}\textsc{=compl} \\ 
\glt `We are going to Iraq. We are going [there] to pass this news [to the agha in Iraq]”'
\z 
 
\ea \label{BP.90}
\textit{māčmē ī meʕmūrū šime āmē lāw ēme} \\ 
\gll m-āč-mē ī meʕmūr=ū šime āmā-ē lā=ū ēme \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} come\textsc{.pst}\textsc{-3pl} at\_the\_place\_of\textsc{=ez.gen} \textsc{1pl} \\ 
\glt `We will say [to the agha], “your officers came to our region".'
\z 
 
\ea \label{BP.96}
\textit{ī meʕmūrū šimemā sere biřyēnē} \\ 
\gll ī meʕmūr=ū šime=mā sere biřyē=nē \\ 
 \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl}\textsc{=\textsc{1pl}} head\textsc{.m} cut\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We have decapitated your officers.'
\z 
 
\ea \label{BP.97}
\textit{seremā biřyē, īne sereū ēmew īne šimšērū šime} \\ 
\gll sere=mā biřī-ē īne sere=ū ēme=ū īne šimšēr=ū šime \\ 
 head\textsc{.m}\textsc{=\textsc{1pl}} cut\textsc{.pst}\textsc{-3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} \textsc{2pl} \\ 
\glt `We have decapitated them. Here are their heads and the agha’s swords.\footnote{Lit. These are our heads and these are the agha’s swords.}'
\z 
 
\ea \label{BP.105}
\textit{bābā xwāđāw ħemey ġeybīū ħemey wirdīkɫe her ginēnē weɫē īnē weɫēne ejošā girđ īnē šūnīšāwe} \\ 
\gll bābā xwāđā=ū ħeme=ī ġeybī=ū ħeme=ī wirdīkɫe her gin-ēn-ē weɫē īnē weɫē=ne ejo=šā girđ īnē šūnī=šā=we \\ 
 grandfather\textsc{.m} \textsc{pn}\textsc{=ez.gen} \textsc{pn}=ez.\textsc{attr} invisible=and \textsc{pn}=ez.\textsc{attr} small just fall.prs\textsc{-pstc}\textsc{-3pl} front \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} front\textsc{=\textsc{post}} guess.prs\textsc{=3pl} all \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} track.f\textsc{=3pl}\textsc{=compl} \\ 
\glt `Baba Khwada, Hama the invisible, and little Hama were in front, assuming that everyone was following them.'
\z 
 
\ea \label{BP.109}
\textit{fermāwo, īne konē xeɫkeke čenū ēme bē?} \\ 
\gll ∅-fermāw-o īne ko=nē xeɫk-eke čenū ēme b-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} people\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} with\textsc{.ez.gen} \textsc{1pl} be.prs\textsc{-pstc}\textsc{.3sg} \\ 
\glt `He (Baba Khwada) said, ‘Where are the people who were with us?'
\z 
 
\ea \label{BP.111}
\textit{īne konē ewēšā? čēššā pene āmā?} \\ 
\gll īne ko=nē ewēšā čēš=šā pene āmā-∅ \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} what\textsc{=3pl} to come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `Where are they? What happened to them?'
\z 
 
\ea \label{BP.112}
\textit{īne konē ewē? čēššā pene āmā?} \\ 
\gll īne ko=nē ewē čēš=šā pene āmā-∅ \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} where\textsc{=cop}\textsc{.3pl} 3pl\textsc{.dir} what\textsc{=3pl} to come\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `Where are they? What happened to them?'
\z 
 
\ea \label{BP.119}
\textit{milmē īney qomīyān,} \\ 
\gll mi-l-mē īney qomyā=n \\ 
 \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `We are going [there since] this has happened. '
\z 
 
\ea \label{BP.121}
\textit{māčmē ēme serū īney ī pīyēmā kuštēnē”’} \\ 
\gll m-āč-mē ēme ser=ū īney ī pīyē=mā kuštē=nē \\ 
 \textsc{ind-}say.prs\textsc{-\textsc{1pl}} \textsc{1pl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox} man\textsc{.dir}\textsc{.pl}\textsc{=\textsc{1pl}} kill\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `We say, ‘We have killed these men because of this.’'
\z 
 
\ea \label{BP.124}
\textit{ā wextī milārewe māčā, wiɫā ī meʕmūrū šime kīyāsēbēnētā pey īne, īnemā āne heke šime gerekbē dēmā penew} \\ 
\gll ā wext-ī mi-l-ā=re=we m-āč-ā wiɫā ī meʕmūr=ū šime kīyāsē b-ēn-ē=tā pey īne īne=mā āne heke šime gerek b-ē dā-ē=mā pene=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}go.\textsc{prs}\textsc{-3pl}\textsc{=\textsc{povb}}\textsc{=compl} \textsc{ind-}say.prs\textsc{-3pl} by\_God \textsc{dem.prox} officer\textsc{.m}\textsc{=ez.gen} \textsc{2pl} send\textsc{.pst}\textsc{.ptcp}\textsc{.pl} be.prs\textsc{-pstc}\textsc{-3pl}=\textsc{2pl} for \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=\textsc{1pl}} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} if \textsc{2pl} necessary be.prs\textsc{-pstc}\textsc{.3sg} give\textsc{.pst}\textsc{-3pl}\textsc{=\textsc{1pl}} to=and \\ 
\glt `Then, they went [to the agha and] said, ‘Indeed, the officers whom you had sent to us, whatever [ taxation] you had asked for, we gave them,'
\z 
 
\ea \label{BP.128}
\textit{īne serew ēmew īne šimšērū āġey} \\ 
\gll īne sere=ū ēme=ū īne šimšēr=ū āġe-ī \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} head\textsc{.m}\textsc{=ez.gen} \textsc{1pl}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} sword\textsc{.m}\textsc{=ez.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Here are their heads and the agha’s swords.'
\z 
 
\ea \label{BP.133}
\textit{īneyšā mekerdē zīnān} \\ 
\gll īney=šā me-ker-dē zīnān \\ 
 \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=3pl} \textsc{neg-}do.prs-\textsc{2pl} prison\textsc{.m} \\ 
\glt `Do not put this one of them in prison.'
\z 
 
\ea \label{BP.150}
\textit{ī due nefere hešpišā dēnēne, gīršā wārden} \\ 
\gll ī due nefer=e hešpišī-ā dē=nē=ne gīr=šā wārde=n \\ 
 \textsc{dem.prox} two.f person\textsc{.m}\textsc{=dem} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} hook\textsc{.m}\textsc{=3pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `These two persons (Baba Khwada and Hama Ghaybi) had gotten lice. They had gotten stuck.'
\z 
 
\ea \label{BP.157}
\textit{ēme mezānmē tekbīrmā čēš bo īnīšā} \\ 
\gll ēme me-zān-mē tekbīr=mā čēš ∅-b-o īnīšā \\ 
 \textsc{1pl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} advice\textsc{=\textsc{1pl}} what \textsc{ind-}be.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `We don’t know what we should do, and so on.'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.161}
\textit{bizāne yānew kē pī činnē hāy wexte nānē čīwēšā dānī ārdenit pey ēme, ārdenit pey ēmew} \\ 
\gll bi-zān-e yāne=ū kē p=ī činnē hāy wext=e nān-ē čīw-ē=šā dā=nī ārde=n=it pey ēme ārde=n=it pey ēme=ū \\ 
 \textsc{sbjv-}know.prs-\textsc{2sg} house\textsc{.m}\textsc{=ez.gen} who at=\textsc{dem.prox} some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{=dem} meal\textsc{.m}\textsc{-indf} thing\textsc{.m}\textsc{-indf}\textsc{=3pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{2sg}} for \textsc{1pl}=and \\ 
\glt `see which household has given you food, which you brought to us, during this period!'
\z 
 
\ea \label{BP.165}
\textit{wāče her kesī nān dānī, wāče, ēšew be xātirū xwāy toū īđū ī kiɫfetū yāneyte dūr ginewe čī šārī’} \\ 
\gll wāč-e her kes-ī nān dā=nī wāč-e ēšew be xātir=ū xwā-ī to=ū īđ=ū ī kiɫfet=ū yāne-ī=t=e dūr ∅-gin-e=we čī šār-ī \\ 
 say.prs-\textsc{imp-}\textsc{2sg} just person\textsc{.m}\textsc{-obl}\textsc{.m} bread\textsc{.m} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} say.prs-\textsc{imp-}\textsc{2sg} tonight to sake\textsc{.m}\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}=and 3sg.prox\textsc{.dir}\textsc{.m}=and \textsc{dem.prox} child\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}}\textsc{=dem} far \textsc{imp-}fall.prs-\textsc{2sg}\textsc{=compl} why city\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `Tell whoever gave you food, say, “Tonight, for the sake of God, you and your household get away from this town.'
\z 
 
\ea \label{BP.173}
\textit{bāzem dūbāre ī sefer wextē milowe dimā leqē māɫo zīnānekeyre zīnāneke yekser veɫādīm bo} \\ 
\gll bāzem dūbāre ī sefer wext-ē mi-l-o=we dimā leq-ē ∅-māɫ-o zīnān-ekey=re zīnān-eke yekser veɫādīm ∅-b-o \\ 
 again again \textsc{dem.prox} trip time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg}\textsc{=compl} afterwards kick\textsc{-indf} \textsc{ind-}smash.prs\textsc{-3sg} prison\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} totally collapsed \textsc{ind-}be.prs\textsc{-3sg} \\ 
\glt `This time, when he went backward and kicked the prison [wall], the prison totally collapsed.'
\z 
 
\ea \label{BP.208}
\textit{īne dāstānū īney, īnē dāstānū bābāw pīr šelīyārīū ħemey ġeybī bīyēnē} \\ 
\gll īne dāstāne=ū īney īnē dāstāne=ū bābā=ū pīr šelīyār-ī=ū ħeme=ī ġeybī bīyē=nē \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story.f\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} grandfather\textsc{.m}\textsc{=ez.gen} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m}=and \textsc{pn}=ez.\textsc{attr} invisible be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `This was the story of Pir Shaliyar’s grandfather and Hama Ghaybi.'
\z 
 
\ea \label{BP.209}
\textit{luēnē īnešā wene qomīyān} \\ 
\gll luē=nē īne=šā wene qomyā=n \\ 
 go\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3pl} at happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] these things happened to them.'
\z 
 
\ea \label{BP.210}
\textit{ī kāršā āmān milo} \\ 
\gll ī kār=šā āmā=n mil=o \\ 
 \textsc{dem.prox} task\textsc{.m}\textsc{=3pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} on\textsc{=\textsc{post}} \\ 
\glt `These things happened to them.'
\z 
 
\ea \label{BP.211}
\textit{werū īney diruē neđāw werū īney je lāw xwāywe negināwe diro ħīčē suālē diruēšā neđēne} \\ 
\gll wer=ū īney diruē ne-đā=ū wer=ū īney je lā=ū xwā-ī=we ne-gin-ā=we diro ħīč-ē suāl-ē diruē=šā ne-đē=ne \\ 
 out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} lie.f \textsc{neg-}give\textsc{.pst}=and out\_of\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} from at\_the\_place\_of\textsc{=ez.gen} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=compl} \textsc{neg-}fall.prs\textsc{-3pl}\textsc{=compl} lie nothing\textsc{-indf} question\textsc{-indf} lie.f\textsc{=3pl} \textsc{neg-}give\textsc{.pst}\textsc{.ptcp}.f\textsc{=cop}\textsc{.3sg}.f \\ 
\glt `In order not to lie, in order not to be liars in the presence of God, they didn’t tell a lie.'
\z 
 
\ea \label{BP.212}
\textit{īnē dāstānū āđī tā ewā} \\ 
\gll īnē dāstāne=ū āđī tā ewā \\ 
 \textsc{dem.prox}\textsc{.dir}.f\textsc{.3sg} story.f\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} until there \\ 
\glt `This was his tale. [It was] until there.'
\z 
 

\ea \label{ŠJ.12}
\textit{tifengiš henū īneš henū āneš henū!} \\ 
\gll tifeng=iš hen-∅=ū īne=š hen-∅=ū āne=š hen-∅=ū \\ 
 gun\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist}\textsc{-3sg}\textsc{.m}=and \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=3sg} \textsc{exist-3sg.m}=and \\ 
\glt `He has a gun, he has this, and he has that!'
\z 
 
\ea \label{ŠJ.30}
\textit{māčā qurbān īne čiko bīyenī} \\ 
\gll m-āč-ā qurbān īne či=ko bīye=nī \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=where be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `They said, ‘Sir, where have you been?'
\z 
 
\ea \label{ŠJ.48}
\textit{perčī nāqāyme miđyo bīyen ženī ēhwāɫiš fāřyān} \\ 
\gll per- č=ī nāqāym=e mi-đye-o bīye=n ženī ēhwāɫ=iš fāřyā=n \\ 
 \textsc{ptcl}- at=\textsc{dem.prox} sudden\textsc{=dem} \textsc{ind-}look.prs\textsc{-3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} woman.f demeanour\textsc{=3sg} change\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All of a sudden, he realised that he had turned into a woman; his demeanour had changed;'
\z 
 
\ea \label{ŠJ.51}
\textit{kirž māčo be wiɫā[xdār] ī pīyāw yāneyš} \\ 
\gll kirž m-āč-o be wiɫāxdār ī pīyā=ū yāne-ī=š \\ 
 quick \textsc{ind-}say.prs\textsc{-3sg} to stableman \textsc{dem.prox} man\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \\ 
\glt `Quickly, he said to the stableman, to the servant in his house,'
\z 
 
\ea \label{ŠJ.90}
\textit{xo řīk megino ī kāre pey girđ kesī bo} \\ 
\gll xo řīk me-gin-o ī kār=e pey girđ kes-ī ∅-b-o \\ 
 dis.ptc arranged \textsc{neg-}fall.prs\textsc{-3sg} \textsc{dem.prox} task\textsc{.m}\textsc{=dem} to all person\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} \\ 
\glt `These (tasks) cannot be done by everyone.'
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{ŠJ.99}
\textit{kāke gīyān īnī ečānē bīyēnē} \\ 
\gll kāke gīyān īnī e=čāne-ē bīyē=nē \\ 
 brother.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.pl} in=like\_that\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Dear brother, they were like this.'
\z 
 
\ea \label{ŠJ.102}
\textit{enāne hetā fermāwī ī mentēqew ʕērāqū ērān desū ānīšāwe bīyen} \\ 
\gll enā=ne hetā ∅-fermāw-ī ī mentēqe=ū ʕērāq=ū ērān des=ū ānīšā=we bīye=n \\ 
 otherwise\textsc{=\textsc{post}} even \textsc{ind-}say.prs-\textsc{2sg} \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{=ez.gen} \textsc{pn} hand\textsc{.m}\textsc{=ez.gen} 3pl.dem\textsc{.obl}\textsc{=compl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All these regions in Iraq and Iran were under their rule.'
\z 
 
\ea \label{KŠ.1}
\textit{ew wextī īnē her girđ řāzū eyemnē lāšo bīyenā eyām} \\ 
\gll ew wext-ī īnē her girđ řāz=ū eye=m=nē lā=š=o bīye=nā eyā=m \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{dem.prox}\textsc{.3pl} just all tale\textsc{.m}\textsc{=ez.gen} mother\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3pl} at\_the\_place\_of\textsc{=3sg}\textsc{=\textsc{post}} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} mother\textsc{=\textsc{1sg}} \\ 
\glt `Well, these are all my mother’s tales. I was with my mother.'
\z 
 
\ea \label{KŠ.18}
\textit{benāxātirē īney āyeč bilo, hāmɫē bē} \\ 
\gll benāxātirē īney āye=č bi-l-o hāmɫē b-ē \\ 
 for\_the\_reason \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} 3sg\textsc{.f}\textsc{.dir}\textsc{=add} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} pregnant\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `The reason [for issuing this order] was that she (the shepherd’s wife), who was pregnant, would disappear.'
\z 
 
\ea \label{KŠ.28}
\textit{wātiš, qurbān herkey gele berī epāne min ī bizeme čekyo} \\ 
\gll wāt=iš qurbān herkey gele ∅-ber-ī epāne min ī bize=m=e ∅-čekye-o \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} each\_time herd\textsc{.m} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} there \textsc{1sg} \textsc{dem.prox} goat\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{ind-}drain\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `[and] said, ‘Sir, whenever you take the herd there, my goat[‘s milk] is drained.'
\z 
 
\ea \label{KŠ.43}
\textit{māčo be eyāšū tāteyšū āđīšā, bā īne menowe pey nokerū ēme epēne pey werdesīmā pey hermānberū yāneymā} \\ 
\gll m-āč-o be eyā=š=ū tāte-ī=š=ū āđīšā bā īne ∅-men-o=we pey noker=ū ēme epēne pey werdes-ī=mā pey hermānber=ū yāne-ī=mā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by mother\textsc{=3sg}=and father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and 3pl\textsc{.obl} \textsc{hort} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}remain\textsc{.prs}\textsc{-3sg}\textsc{=compl} to servant\textsc{.m}\textsc{=ez.gen} \textsc{1pl} here to attendant\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} to worker\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `She (the girl) said to her parents and others , ‘Let him stay here as our servant here, as our attendant, as our domestic worker.'
\z 
 
\ea \label{KŠ.46}
\textit{ā wextīyekey xeɫk milo māčo be bābeyšū īnīšā māčā, be xwā pāse dīyārā} \\ 
\gll ā wext-ī-ekey xeɫk mi-l-o m-āč-o be bābe-ī=š=ū īnīšā m-āč-ā be xwā pāse dīyār=ā \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} people\textsc{.m} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} by God\textsc{.m} like apparent\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then people went [and] said to his father and so on, ‘Indeed, it seems that [he is well-integrated into the king’s family]'
\z 
 
\ea \label{KŠ.51}
\textit{māčo, be xwā pāđšām seɫāmet kuřū šuwāney heke to kuštēū ā wextīyekey hukmit dāū īnīšā īse īnā yānetā īnīšā} \\ 
\gll m-āč-o be xwā pāđšā=m seɫāmet kuř=ū šuwāne-ī heke to kušt-ē=ū ā wext-ī-ekey hukm=it dā=ū īnīšā īse īnā-∅ yāne=tā īnīšā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by God\textsc{.m} king\textsc{.m}\textsc{=\textsc{1sg}} healthy son/boy\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} if \textsc{2sg} kill\textsc{.pst}\textsc{-3pl}\textsc{=ez.gen} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} rule\textsc{.m}\textsc{=\textsc{2sg}} give\textsc{.pst}=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} now \textsc{deic}\textsc{-3sg}\textsc{.m} house\textsc{.m}=\textsc{2pl} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `They said [in the letter], ‘Indeed, may my king be healthy, the shepherd’s son whom you ordered to be killed is now in your house and so on’'
\z 
 
\ea \label{KŠ.56}
\textit{yewāšē čāge, čogene šewē minwīso, māčo, ey kesūkārū min, ey henū nīyenū min, ey īnīšā ke mintā gereknā, tāyfew min heɫbetene ī kuře biřiš dāwe, be her jor ī kuřī beynne berdē.’} \\ 
\gll yewāšē čāge č=oge=ne šewe-ē mi-nwīs-o m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey īnīšā ke min=tā gerek=nā tāyfe=ū min heɫbete=ne ī kuř=e biř=iš dā=we be her jor ī kuř-ī beyn=ne ∅-ber-dē \\ 
 then there in=there\textsc{=\textsc{post}} night\textsc{.f}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} rel \textsc{1sg}=\textsc{2pl} necessary\textsc{=cop}\textsc{.\textsc{1sg}} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} by any means \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} between\textsc{=\textsc{post}} \textsc{ind-}take\textsc{.prs}-\textsc{2pl} \\ 
\glt `Then, one night he (the king) wrote [a letter], saying, ‘O my relatives, O my close family members , O those who want me, [are] from my tribe, [when you read this letter] it is obvious that this boy has arrived. By any means possible, kill him.'
\z 
 
\ea \label{KŠ.57}
\textit{īnē dilē nāmekeyne minvīso} \\ 
\gll īnē dilē nāme-ekey=ne mi-nvīs-o \\ 
 \textsc{dem.prox}\textsc{.3pl} inside letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \\ 
\glt `He wrote this inside the letter [and gave it to the shepherd’s son].'
\z 
 
\ea \label{KŠ.72}
\textit{niwīsebēš, ey kesūkārū min, ey henū nīyenū min, heɫbetene ī kuřī biřiš dāwe, šime be wesīlēwe kuždēš, beynne berdēš!} \\ 
\gll niwīse b-ē=š ey kesūkār=ū min ey hen=ū nīyen=ū min heɫbete=ne ī kuř-ī biř=iš dā=we šime be wesīle-ēwe ∅-kuž-dē=š beyn=ne ∅-ber-dē=š \\ 
 write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} \textsc{2pl} by tool\textsc{.m}\textsc{-indf} \textsc{imp-}kill\textsc{.prs}-\textsc{2pl}\textsc{=3sg} between\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2pl}\textsc{=3sg} \\ 
\glt `He had written, ‘O my relatives, O my close family members, [when you read this letter] it is evident that this boy has arrived. Kill him with a thing, destroy him!'
\z 
 
\ea \label{KŠ.73}
\textit{īnēč dīyāš pore bezeyīš āmā pore} \\ 
\gll īnē=č dīyā=š pore bezeyī=š āmā pore \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=add} look\textsc{.pst}\textsc{=3sg} at pity\textsc{.m}\textsc{=3sg} come\textsc{.pst}\textsc{.3sg} at \\ 
\glt `She looked at him. She took pity on him.'
\z 
 
\ea \label{KŠ.76}
\textit{īne wēš xeberiš nīyāw sewāđiš nīyā} \\ 
\gll īne wē=š xeber=iš nīye=ā=ū sewāđ=iš nīye=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{reflx}\textsc{=3sg} news\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and literacy\textsc{.m}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He wasn’t aware [of the death plot] that \_\_ he was illiterate\_\_'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{KŠ.84}
\textit{īnā qāqezēčiš dānā pey} \\ 
\gll īne=ā qāqez-ē=č=iš dā=nā pey \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} letter\textsc{.m}\textsc{-indf}\textsc{=add}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to \\ 
\glt `Look, he has given me a letter.'
\z 
 
\ea \label{KŠ.86}
\textit{wātiš dey lue āney fiɫānīū fiɫānī čiře ī qewmāme bā bēnēwe qāqezekey wenā} \\ 
\gll wāt=iš dey ∅-lu-e āney fiɫānī=ū fiɫānī ∅-čiř-e ī qewm-ā=m=e bā b-ē-nē=we qāqez-ekey ∅-wen-ā \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} such\_and\_such\_person=and such\_and\_such\_person \textsc{imp-}call\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{hort} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}read\textsc{.prs}\textsc{-3pl} \\ 
\glt `She said, ‘Then go, and summon such-and-such, our relatives, so that they come and read the letter.’'
\z 
 
\ea \label{KŠ.96}
\textit{ā esbū tāteyme biznere zīnīš kere to suwār be, bā min bew pey ser ī kuřīče binye dānit} \\ 
\gll ā esb=ū tāte-ī=m=e bi-zn-e=re zīnī=š ∅-ker-e to suwār ∅-b-e bā min b-e-ū pey ser ī kuř=īč=e bi-nye-e dān=it \\ 
 \textsc{dem.dist} horse\textsc{\textsc{=ez.gen}} father\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{imp-}take\_out\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} saddle\textsc{=3sg} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{2sg} riding \textsc{imp-}be\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{1sg} \textsc{sbjv-}come\textsc{.prs}\textsc{-\textsc{1sg}} to on \textsc{dem.prox} boy\textsc{.m}\textsc{=add}\textsc{=dem} \textsc{imp-}put\textsc{.prs}-\textsc{2sg} hip\textsc{.m}\textsc{=\textsc{2sg}} \\ 
\glt `‘Take my father’s horse out [of the stable], mount it [and] then I will climb [and] put the son on your hip.'
\z 
 
\ea \label{ŽH.27}
\textit{wātiš, tāte āne wāteniš, hekāyetē weš gīsnene, bā ī řē kučē keme kermē be hekāyetekeyt rāz biřmē} \\ 
\gll wāt=iš tāte āne wāte=n=iš hekāyet-ē weš-∅ ∅-gīsn-e=ne bā ī řē kuč-ē kem-e ∅-ker-mē be hekāyet-ekey=t rāz ∅-biř-mē \\ 
 say\textsc{.pst}\textsc{=3sg} father\textsc{.m} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} tale\textsc{.m}\textsc{-indf} well-m \textsc{imp-}start\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{hort} \textsc{dem.prox} road\textsc{.obl}\textsc{\textsc{.f}}.sg little\textsc{-indf} s\textsc{hort}\textsc{-f} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} by tale\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=\textsc{2sg}} tale\textsc{.m} \textsc{sbjv-}cut\textsc{.prs}\textsc{-1pl} \\ 
\glt `She said, ‘Father, he had said, “Start [telling] a good tale, so that we [are entertained] and our way seems shorter. Let us finish your tale.”'
\z 
 
\ea \label{ŽH.40}
\textit{wātiš be wāɫe wičkilekēš ī māsī bere deš pā mēmān mizgī čenīw nānī’} \\ 
\gll wāt=iš be wāɫē=e wičkile-ekē=š ī mās-ī ∅-ber-e ∅-de-e=š p=ā mēmān mizgī čenī=ū nān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} by sister\textsc{\textsc{.f}}\textsc{=ez}\textsc{.cmpd} little\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}}\textsc{=3sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} to\textsc{=dem}.dist guest\textsc{.m} mosque\textsc{.m} with\textsc{=ez}\textsc{.gen} bread\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[and] said to her little daughter, ‘Take this yoghurt [and] together with [some] bread give it to the the guest in the mosque.’ '
\z 
 
\ea \label{ŽH.42}
\textit{kučēš čene werū ī māsī’} \\ 
\gll kuč-ē=š čene ∅-wer-ū ī mās-ī \\ 
 little\textsc{-indf}\textsc{=3sg} with \textsc{ind-}eat\textsc{.prs}\textsc{-1sg} \textsc{dem.prox} yoghurt\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `I shall eat some of it.'
\z 
 
\ea \label{ŽH.53}
\textit{māčo, ī serū řīšīme kučē tāše’} \\ 
\gll m-āč-o ī sere=ū řīš-ī=m=e kuč-ē ∅-tāš-e \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox} head\textsc{.m}=and beard\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1sg}\textsc{=dem} little\textsc{-indf} \textsc{imp-}shave\textsc{.prs}-\textsc{2sg} \\ 
\glt `He (Hayas) said [to the girl], ‘shave my face a bit.’'
\z 
 
\ea \label{ŽH.62}
\textit{īne e čē čenū āġeyš jengšā bīyen, be ādizī lūān} \\ 
\gll īne e=čē čenū āġe-ī=š jeng=šā bīye=n be ādizī lūā=n \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=here with.ez\textsc{.gen} agha\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} quarrel\textsc{=3pl} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} with desperation go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (Hayas) got into a quarrel with his master. He had left [town] in desperation'
\z 
 
\ea \label{ŽH.65}
\textit{yewāšē āđīč mārošo mēwe kināčēw luhonī bo teřū čermew īnīšā} \\ 
\gll yewāšē āđ=īč m-ār-o=š=o m-ē=we kināčē=ū luhon-ī ∅-b-o teř=ū čerme=ū īnīšā \\ 
 then 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg}\textsc{=compl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} fresh=and white=and \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} \\ 
\glt `He (Hayas) brought her (i.e. his wife). She was from Luhon; she was young, white, and so on.'
\z 
 
\ea \label{ŽH.70}
\textit{pīney bilo šewē neēwe} \\ 
\gll p=īney bi-l-o šewe-ē ne-ē=we \\ 
 for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} \textsc{neg-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `The reason [was] so that he (Hayas) would go and not come back for the night. '
\z 
 
\ea \label{ŽH.71}
\textit{řākē dūrew inne dūre bo, pīney šewē biř nekero bēwe} \\ 
\gll řā-ekē dūr=ew inne dūr-e ∅-b-o p=īney šewe-ē biř ne-ker-o b-ē=we \\ 
 road\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} far\textsc{=\textsc{post}} so\_much far\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg} night\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}} piece \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \\ 
\glt `The distance was so long that he (Hayas) wouldn’t be able to return for the night.'
\z 
 
\ea \label{ŽH.72}
\textit{yewāšē ēđīč wēš wezore milo pey yāneyšā} \\ 
\gll yewāšē ēđ=īč wē=š wez-o=re mi-l-o pey yāne-ī=šā \\ 
 then 3sg.prox\textsc{.dir}\textsc{=add} \textsc{reflx}\textsc{=3sg} throw\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} for house\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} \\ 
\glt `Then, he (the sultan) prepared himself [and] went towards their (i.e. Hayas’s) house.'
\z 
 
\ea \label{ŽH.86}
\textit{māčo be tūwāɫekeyšā, īne čēšen?’} \\ 
\gll m-āč-o be tūwāɫe-ekey=šā īne čēše=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by shell\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=3pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what \\ 
\glt `She pointed to the (egg)shells [and] said, ‘what are these? \footnote{Lit. What is this}''
\z 
 
\ea \label{ŽH.87}
\textit{māčo, īnē māčmēšā hēɫe’} \\ 
\gll m-āč-o īnē m-āč-mē=šā hēɫe \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.3pl} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3pl} egg\textsc{.m} \\ 
\glt `He (the sultan) said, ‘We call these eggs.’'
\z 
 
\ea \label{ŽH.88}
\textit{māčo, īne bīye čin řengī?’} \\ 
\gll m-āč-o īne bī-e čin řengī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} be\textsc{.pst} 3sg\textsc{\textsc{.f}} how\_many colour \\ 
\glt `She said, ‘How many colours have they become?’'
\z 
 
\ea \label{ŽH.95}
\textit{māčo, īne her hēɫe bī’} \\ 
\gll m-āč-o īne her hēɫe bī-∅ \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} just egg\textsc{.m} be\textsc{.pst}\textsc{-3sg}\textsc{.m} \\ 
\glt `He said, ‘This one was just an egg.’'
\z 
 
\ea \label{ŽH.111}
\textit{māčo, āmāno īnā’} \\ 
\gll m-āč-o āmā=n=o īne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `He (the sultan) said, ‘He has returned; there he is.’'
\z 
 
\ea \label{RE.4}
\textit{bābe gīyān ī kināčēte yā tātemā lūān yā māmomā yā kesūkārmā} \\ 
\gll bābe gīyān ī kināčē=t=e yā tāte=mā lūā=n yā māmo=mā yā kesūkār=mā \\ 
 father.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} either father\textsc{.m}\textsc{=\textsc{1pl}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} either paternal\_uncle\textsc{=\textsc{1pl}} either relative\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `They would say, ‘Dear brother,\footnote{Lit. father.} this daughter of yours \—either our father would go [to the girl’s father], or our uncle, or a relative of ours—'
\z 
 
\ea \label{RE.7}
\textit{ʕerzū birāw wēm kerū ī kināčēte de pī kuřīme} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū ī kināčē=t=e ∅-de-e p=ī kuř-ī=m=e \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} brother\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} \textsc{imp-}give\textsc{.prs}-\textsc{2sg} at=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=dem} \\ 
\glt `Let me tell you,\footnote{Lit. May a petition be at the service of my brother.} ‘Give your daughter to my son [in marriage]'
\z 
 
\ea \label{RE.16}
\textit{ey pīyā ētir emīnū yāne bīyenū xizmetkār bīyen} \\ 
\gll ey pīyā ētir emīn=ū yāne bīye=n=ū xizmetkār bīye=n \\ 
 \textsc{dem.prox} man\textsc{.m} \textsc{disc.ptcl} confidant\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{.dir}.sg be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez.gen} servant\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was this man who was the family’s confidant. He was a servant.'
\z 
 
\ea \label{RE.18}
\textit{firmānbirđār bīyen, tā sāhībū ī kināčē wāčo, xeber bido wāčo, fiɫāne kār} \\ 
\gll firmānbirđār bīye=n tā sāhīb=ū ī kināčē ∅-wāč-o xeber bi-de-o ∅-wāč-o fiɫān=e kār \\ 
 obedient be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} until owner\textsc{.m}\textsc{=ez.gen} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} news\textsc{.dir}\textsc{.m} \textsc{sbjv-}give\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} such\_and\_such=ez\textsc{.cmpd} task\textsc{.m} \\ 
\glt `He was obedient. The girl’s father\footnote{Lit. owner.} would say [to him] and ask him, ‘[Do] such-and-such tasks;'
\z 
 
\ea \label{RE.41}
\textit{ī memliketū ēme ne welē wiɫātāre ētir tāte bīyen čūēw desū šuwāney} \\ 
\gll ī memliket=ū ēme ne welē wiɫāt-ā=re ētir tāte bīye=n čū-ē=w des=ū šuwāne-ī \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} no but region\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{disc.ptcl} father\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} stick\textsc{.m}\textsc{-indf}\textsc{=ez.gen} hand\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[It is not common] in our region, but elsewhere, the father has become like the stick in the shepherd’s hand.\footnote{Meaning that he has no power over whom his children will marry.}'
\z 
 
\ea \label{RE.52}
\textit{pīney bāyē māsīč bero} \\ 
\gll p=īney bāye-ē mās=īč ∅-ber-o \\ 
 for=\textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} pot\textsc{.m}\textsc{-dir}\textsc{.pl} yoghurt\textsc{.m}\textsc{=add} \textsc{sbjv-}take\textsc{.prs}\textsc{-3sg} \\ 
\glt `For the reason that he (the girl’s father) would take a bowl of yoghurt [to him] too.'
\z 
 
\ea \label{RE.56}
\textit{zemānē ēme ī begēmā bīyēnē} \\ 
\gll zemān-ē ēme ī beg-ē=mā bīyē=nē \\ 
 time\textsc{.m}\textsc{-indf} \textsc{1pl} \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Once we had these noblemen [in our region]'
\z 
 
\ea \label{RE.57}
\textit{ī begē qibūɫ kerā} \\ 
\gll ī beg-ē qibūɫ ∅-ker-ā \\ 
 \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf} accepted \textsc{sbjv-}do\textsc{.prs}\textsc{-3pl} \\ 
\glt `[After] the noblemen should consent [to the marriage]'
\z 
 
\ea \label{RE.62}
\textit{ī memliketū ēme řisūmātū āwe zemānīš e čīne bīyen} \\ 
\gll ī memliket=ū ēme řisūmāt=ū āwe zemān-ī=š e=čīne bīye=n \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} traditions\textsc{.m}\textsc{=ez.gen} that time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} in=like be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The traditions of earlier times were like this in our region.'
\z 
 
\ea \label{RE.63}
\textit{ī begēmā bīyēnē, begē} \\ 
\gll ī beg-ē=mā bīyē=nē beg-ē \\ 
 \textsc{dem.prox} Beg\textsc{.m}\textsc{-indf}\textsc{=\textsc{1pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} Beg\textsc{.m}\textsc{-indf} \\ 
\glt `We had these noblemen.'
\z 
 
\ea \label{ŽE.4}
\textit{ī degā toš vīnī čoɫe bīyēne} \\ 
\gll ī degā to=š ∅-vīn-ī čoɫ-e bīyē=ne \\ 
 \textsc{dem.prox} village\textsc{.obl}\textsc{\textsc{.f}} \textsc{2sg}\textsc{=3sg} \textsc{ind-}see\textsc{.prs}-\textsc{2sg} deserted-\textsc{f} be\textsc{.ptcp}\textsc{\textsc{.f}}\textsc{=cop}\textsc{.3sg}\textsc{\textsc{.f}} \\ 
\glt `This village, which you see, was deserted.'
\z 
 
\ea \label{ŽE.50}
\textit{ī doīnete dēnmēū ī šelemīnēte dēnmē} \\ 
\gll ī doīne=t=e dē=nmē=ū ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{\textsc{=ez.gen}} \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this dwina. \footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{ŽE.51}
\textit{ī šelemīnēte dēnmē} \\ 
\gll ī šelemīnē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `you gave us this Shalamina. \footnote{Dish made of cracked wheat and turnip.}'
\z 
 
\ea \label{ŽE.52}
\textit{ī xeplēte dēnmē} \\ 
\gll ī xeplē=t=e dē=nmē \\ 
 \textsc{dem.prox} \textsc{pn}\textsc{=\textsc{2sg}}\textsc{=dem} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `You gave us this xapla '
\z 
 
\ea \label{ŽE.53}
\textit{īnēmā wārdēne} \\ 
\gll īnē=mā wārdē=ne \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.f}\textsc{.3sg}\textsc{=1pl} eat\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We have eaten this [meal]. '
\z 
 
\ea \label{ŽE.74}
\textit{luēnī lāw ī kuřī wāčēnī, dā lue wāče eđēš bizāne ī kināčēše miđom pene yā ne’} \\ 
\gll lu-ēn-ī lā=ū ī kuř-ī wāč-ēn-ī dā ∅-lu-e ∅-wāč-e eđē=š bi-zān-e ī kināčē=š=e mi-đe-o=m pene yā ne \\ 
 go\textsc{.prs}\textsc{-pstc}-\textsc{2sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} say\textsc{.prs}\textsc{-pstc}-\textsc{2sg} \textsc{hort} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}say\textsc{.prs-}.\textsc{2sg} mother\textsc{.obl}\textsc{.f}\textsc{=3sg} \textsc{sbjv-}know\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=3sg}\textsc{=dem} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{1sg}} to either no \\ 
\glt ` You would go to this boy [seated next to the narrator] and would say, ‘Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!’'
\z 
 
\ea \label{ŽE.75}
\textit{kuřekeyč ē māčī, kināčekēt miđey pī kuřī?’} \\ 
\gll kuř-ekey=č ē m-āč-ī kināčē-ekē=t mi-đe-ī p=ī kuř-ī \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} come\textsc{.prs}\textsc{.pst}c\textsc{.3sg} \textsc{ind-}say\textsc{.prs}-\textsc{2sg} daughter\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `The boy comes. [Addressing her grand son] you would say [to her mother], ‘Will you give your daughter to this boy [in marriage]?’'
\z 
 
\ea \label{ŽE.80}
\textit{ēnē hījbīš kerē wāčē, ī kināčē miđey pī kuřī?’} \\ 
\gll ēnē hījbī=š ker-ē wāč-ē ī kināčē mi-đe-ī p=ī kuř-ī \\ 
 come\textsc{.prs}\textsc{.pstc}\textsc{.3pl} betrothal\textsc{=3sg} do\textsc{.prs}\textsc{-pstc} say\textsc{.prs-pstc}\textsc{.3sg} \textsc{dem.prox} girl\textsc{.obl}\textsc{.f} \textsc{ind-}give\textsc{.prs}-\textsc{2sg} to=\textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `They would come [and] act as intermediary to ask for the hand of the girl. They would say, ‘Will you give this girl to this boy?’'
\z 
 
\ea \label{ŽE.83}
\textit{tā qewmekāmā wiš kermē qewmekā wiš kermē bizānmē ī qewmīmā řezāšāne yā řezāšā nīyene} \\ 
\gll tā qewm-ekā=mā wiš ∅-ker-mē qewm-ekā wiš ∅-ker-mē bi-zān-mē ī qewm-ī=mā řezā=šā=ne yā řezā=šā nīye=ne \\ 
 until relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=1pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} relative\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl} memory\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=1pl} agreement\textsc{.f}\textsc{=3pl}\textsc{=\textsc{post}} or agreement\textsc{.f}\textsc{=3pl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `We will inform our relatives. We will inform our relatives to see if they agree [to the  marriage] or not.'
\z 
 
\ea \label{ŽM.7}
\textit{je kārēz biřyeywe pī heštā sāɫe, īse ħewt sāɫē} \\ 
\gll je kārēz biřyey=we p=ī heštā sāɫe=e īse ħewt sāɫe-ē \\ 
 from subterranean\_canal\textsc{.m} cut\textsc{.inf}\textsc{=compl} at=\textsc{dem.prox} eighty year\textsc{.f}\textsc{=dem} now seven year\textsc{.f}\textsc{-dir}\textsc{.pl} \\ 
\glt `Ever since [the time I was] digging canals until now that I am eighty years old [I have been working]'
\z 
 
\ea \label{ŽM.8}
\textit{īse ħewt sāɫēč āmānāwe serū īney dūkāndārnā} \\ 
\gll īse ħewt sāɫe-ē=č āmā=nā=we ser=ū īney dūkāndār=nā \\ 
 now seven year\textsc{.f}\textsc{-dir}\textsc{.pl}\textsc{=add} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} on\textsc{=ez.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} shopkeeper\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{ŽM.12}
\textit{īne dāstānū minā} \\ 
\gll īne dāstāne=ū min=ā \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} story\textsc{.f}\textsc{=ez.gen} \textsc{1sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `This is my story.'
\z 
 
\ea \label{ŽM.16}
\textit{dārāw īse elʕān hāɫēhāzirī dārāw penj kuřānāw yerē kināčā, heštū wēmū ženē īne de neferē} \\ 
\gll dārā=w īse elʕān hāɫēhāzir-ī dārā=w penj kuř-ā=nā=ū yerē kināčē-ā hešt=ū wē=m=ū ženī-ē īne de nefer-ē \\ 
 owner\textsc{.m}\textsc{=ez.gen} now now at\_the\_moment\textsc{-obl}\textsc{.m} owner\textsc{.m}\textsc{=ez.gen} five son/boy\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1sg}}=and three daughter\textsc{.f}\textsc{-obl}\textsc{.pl} eight=and \textsc{reflx}\textsc{=\textsc{1sg}}=and woman\textsc{.f}\textsc{-indf} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} ten person\textsc{.m}\textsc{-dir}\textsc{.pl} \\ 
\glt `I have, at the moment, five sons and three daughters, [which makes] eight, [plus] myself and a wife, [making us] ten people.'
\z 
 
\ea \label{ŽM.19}
\textit{ī menteqew ēme kemūkuřīn} \\ 
\gll ī menteqe=ū ēme kemūkořī=n \\ 
 \textsc{dem.prox} region\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Our region is in poverty.'
\z 
 
\ea \label{ŽM.21}
\textit{ī mentēqew ēme nebīyeyš hen nebīyey} \\ 
\gll ī mentēqe=ū ēme nebīyey=š hen-∅ nebīyey \\ 
 \textsc{dem.prox} region\textsc{.m}\textsc{=ez.gen} \textsc{1pl} poverty\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} poverty\textsc{.m} \\ 
\glt `Our region is in poverty, [and] shortage.'
\z 
 
\ea \label{ŽM.27}
\textit{ī yānze zāroɫē ečē āsēnēm jīyā milū ī ženīme milū nebīyeywe} \\ 
\gll ī yānze zāroɫe-ē e=čē āsē=nē=m jīyā mil=ū ī ženī=m=e mil=ū nebīyey=we \\ 
 \textsc{dem.prox} eleven child\textsc{-dir}\textsc{.pl} in=here leave\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{1sg}} separate on\textsc{=ez.gen} \textsc{dem.prox} woman\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} on\textsc{=ez.gen} poverty\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Because of destitution, [I had to] leave these eleven children to my wife.'
\z 
 
\ea \label{ŽM.36}
\textit{ī šiš mānge xeberū hīčīmā nezānān} \\ 
\gll ī šiš mānge=e xeber=ū hīč-ī=mā ne-zānā=n \\ 
 \textsc{dem.prox} six month\textsc{.f}\textsc{=dem} news\textsc{.m}\textsc{=ez.gen} nothing\textsc{-obl}\textsc{.m}\textsc{=\textsc{1pl}} \textsc{neg-}know\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `I didn’t have any updates [from my family] during those periods of six months.'
\z 
 
\ea \label{ŽM.51}
\textit{wextē āmēnmēwe ī mentēqe ētir čē musteqerē bīyēnmē} \\ 
\gll wext-ē āmē=nmē=we ī mentēqe ētir čē musteqer-ē bīyē=nmē \\ 
 time\textsc{.m}\textsc{-indf} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \textsc{disc}.\textsc{ptcl} here settled\textsc{\textsc{-pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `When I returned to this region, I settled down here.'
\z 
 
\ea \label{ŽM.53}
\textit{ʕerzū birāw wēm kerū āmēnmēwe ī mentēqe} \\ 
\gll ʕerz=ū birā=ū wē=m ∅-ker-ū āmē=nmē=we ī mentēqe \\ 
 petition\textsc{.m}\textsc{=ez.gen} brother\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{1sg}} \textsc{sbjv-}do\textsc{.prs}\textsc{-\textsc{1sg}} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \\ 
\glt `Let me tell you. I returned to this region.'
\z 
 
\ea \label{ŽM.57}
\textit{ā wextīyekey ēme īne zinđigīš pene kermē be nebīyey} \\ 
\gll ā wext-ī-ekey ēme īne zinđigī=š pene ∅-ker-mē be nebīyey \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{1pl} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} life\textsc{.m}\textsc{=3sg} with \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1pl}} with poverty\textsc{.m} \\ 
\glt `Well, we live in poverty.'
\z 
 
