\chapter{All 122 examples of ptcl in these texts are found in the following sentences:}

\ea \label{ZB.17}
\textit{māčo, be xwā wāro! ānā hewrē sīyāw pise četrē kewt āsmān’} \\ 
\gll m-āč-o be xwā ∅-wār-o āne=ā hewr-ē sīyāw pise četre-ē kewt-∅ āsmān \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} to God\textsc{.m} \textsc{ind-}rain\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} cloud\textsc{-indf} black like umbrella\textsc{-obl}\textsc{.f} fall\textsc{.pst}\textsc{-3sg}\textsc{.m} sky\textsc{.m} \\ 
\glt ` He said, ‘By God, it will rain. There, a black cloud has covered the sky like an umbrella!’'
\z 
 
\ea \label{ZB.46}
\textit{sāɫē ter mēnēwe, māčo, bilmē bizānmē tekbīrī ētir řāwēžmā čēš bo’ āđē, wiškesāɫīn?} \\ 
\gll sāɫe-ē ter m-ē-nē=we m-āč-o bi-l-mē bi-zān-mē tekbīr-ī ētir řāwēž=mā čēš ∅-b-o āđē wiškesāɫī=n \\ 
 year\textsc{.f}\textsc{-dir}\textsc{.pl} other \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \textsc{sbjv-}know\textsc{.prs}\textsc{-1pl} advice\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} conjecture\textsc{=1pl} what \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} 3pl\textsc{.dir} drought\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The following year, they went back. He (the father) said, ‘We shall go [and] see what is ahead of us with this period of drought.’'
\z 
 
\ea \label{ZQ.8}
\textit{fermāwāš, dā dey řoɫe čēš māčī’} \\ 
\gll fermāwā=š dā dey řoɫe čēš m-āč-ī \\ 
 say\textsc{.pst}\textsc{=3sg} give\textsc{.pst} \textsc{disc.ptcl} child.\textsc{voc} what \textsc{ind-}say.prs-\textsc{2sg} \\ 
\glt `The Shah said, ‘Go on son, What is it ?\footnote{Lit. What are you saying?}’'
\z 
 
\ea \label{ZQ.55}
\textit{fermāwo, dey āne ētir deseɫāt xwāy, pey her čē meyliš bē xwāy deseɫātē čānēš henē} \\ 
\gll ∅-fermāw-o dey āne ētir deseɫāt xwā-ī pey herčē meyl=iš b-ē xwā-ī deseɫāt-ē čāne-ē=š hen-ē \\ 
 \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} power\textsc{.m} God\textsc{.m}\textsc{-obl}\textsc{.m} to whatever willingness\textsc{=3sg} \textsc{sbjv-}come.prs\textsc{.3sg} God\textsc{.m}\textsc{-obl}\textsc{.m} power\textsc{.m-dir.pl} such\textsc{-pl}\textsc{=3sg} \textsc{exist}\textsc{-3pl} \\ 
\glt ` The Shah said, ‘Well, then, this is in God’s hands. Whatever God wishes, he has the power [to exert]'
\z 
 
\ea \label{HB.6}
\textit{hezretū šēxīč fermāwo, dey konī?} \\ 
\gll hezret=ū šēx=yč ∅-fermāw-o dey ko=nī \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} where\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `His highness the Sheikh said, ‘Where are you?’'
\z 
 
\ea \label{HB.62}
\textit{vātiš, dey xuře vene bere} \\ 
\gll vāt=iš dey ∅-xuř-e vene ∅-ber-e \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}shout\textsc{.prs}-\textsc{2sg} \textsc{post} \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `[and] said, ‘Command the donkey\footnote{Lit. shout at him.}; move him ; take him;'
\z 
 
\ea \label{HB.76}
\textit{māčā, qurbān wiɫā fiɫāne kes toryānū ānā} \\ 
\gll m-āč-ā qurbān wiɫā fiɫān=e kes toryā=n=ū āne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} by\_God such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m} get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \\ 
\glt `They said, ‘Sir, indeed, the fellow has got offended; there he is’'
\z 
 
\ea \label{HB.80}
\textit{čiřoš hezretū šēxī fermāwo, erē řoɫem čī toryānī?} \\ 
\gll ∅-čiř-o=š hezret=ū šēx=ī ∅-fermāw-o erē řoɫe=m čī toryā=nī? \\ 
 \textsc{ind-}call\textsc{.prs}\textsc{-3sg}\textsc{=3sg} his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} child\textsc{.m}\textsc{=1sg} why get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt ` His highness the Sheikh summoned him, i.e. the donkey breeder, [and] said, ‘Oh son! why were you offended? '
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{HB.85}
\textit{hezretū šēxī fermāwo, dey xāsā bābā herā gā!} \\ 
\gll hezret=ū šēx=ī ∅-fermāw-o dey xās=ā bābā her-ā gā \\ 
 his\_highness\textsc{.m}\textsc{=ez}\textsc{.gen} sheikh\textsc{.m}\textsc{=ez}.\textsc{attr} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} grandfather\textsc{.m} donkey\textsc{.m}\textsc{-obl}\textsc{.pl} fornicate\textsc{.pst} \\ 
\glt `His highness the Sheikh said, ‘That is fine! God damn you!\footnote{Lit. May donkeys fornicate with your ancestors!}'
\z 
 
\ea \label{HB.89}
\textit{ētir to čī ādiz bīyenī?} \\ 
\gll ētir to čī ādiz bīye=nī \\ 
 \textsc{disc.ptcl} \textsc{2sg} why upset be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `Then why have you become upset?'
\z 
 
\ea \label{HB.90}
\textit{dā dey hurbēze ī mexlūqī girdiš gēɫeš pone} \\ 
\gll dā dey hur-b-ēz-e ī mexlūq-ī gird=iš gēɫ-e=š pone \\ 
 \textsc{hort} \textsc{disc.ptcl} \textsc{pvb-}\textsc{sbjv-}rise\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} people\textsc{-obl}\textsc{.m} all\textsc{=3sg} wander\textsc{.prs}-\textsc{2sg}\textsc{=3sg} at \\ 
\glt `Come on, get up and search among all these people,'
\z 
 
\ea \label{PM.17}
\textit{ānā ēme wēmā mehelēmā hen tūl} \\ 
\gll āne=ā ēme wē=mā mehele-ē=mā hen-∅ tūl \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{1pl} \textsc{reflx}\textsc{=1pl} neighbourhood\textsc{.m}\textsc{-indf}\textsc{=1pl} exist\textsc{-3sg}\textsc{.m} long \\ 
\glt `There, I, myself, own a big neighbourhood.’\footnote{Lit. We, ourselves, have a big neighbourhood.}'
\z 
 
\ea \label{PM.49}
\textit{fermāwo dey ētir min čī nāřihet bīyā!} \\ 
\gll ∅-fermāw-o dey ētir min čī nāřihet bī-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{disc.ptcl} \textsc{1sg} why sad become\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `So, why did I get worried at all?'
\z 
 
\ea \label{DG.4}
\textit{šūnū īney yerē kuřēš bīyēnē, ētir ī memlekete tā milo kursān, tā milo beġā, ī šā ʕelādīne ā wextīekey xwā pāyebiɫindiš keroneū xwā ā wextīekey ʕefweš keroneū ēme čērū himāyeyš bīyēnmē} \\ 
\gll šūn=ū īney yerē kuř-ē=š bīyē=nē ētir ī memleket=e tā mi-l-o kursān tā mi-l-o beġā, ī šā ʕelādīn=e ā wext-ī-ekey xwā pāyebiɫind=iš ∅-ker-one=xwā ā wext-ī-ekey ʕefwe=š ∅-ker-one=ū ēme čēr=ū himāye-ī=š bīyē=nmē \\ 
 after\textsc{=ez}\textsc{.gen} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} three son/boy\textsc{.m}\textsc{-dir}\textsc{.pl}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m}\textsc{=dem} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} until \textsc{ind-}go\textsc{.prs}\textsc{-3sg} \textsc{pn}\textsc{.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn}\textsc{=dem} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} God\textsc{.m} grand\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} pardon\textsc{=3sg} \textsc{sbjv-}do\textsc{.prs}\textsc{-3sg}=and \textsc{1pl} under\textsc{=ez}\textsc{.gen} support\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Shah Osman [well rather] Shah Aladin, in this region and far beyond until Sanandaj and eh Baghdad.'
\z 
 
\ea \label{DG.38}
\textit{ā wextīyekey, ētir, īse, min yerē sāɫē ī birāme tūšū ī derdēe āmānū} \\ 
\gll ā wext-ī-ekey ētir īse min yerē sāɫe-ē ī birā=m=e tūš=ū ī derde-ē=e āmā=n=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl} \textsc{dem.prox} brother\textsc{.m}\textsc{=1sg}\textsc{=dem} encounter\textsc{=ez}\textsc{.gen} \textsc{dem.prox} illness\textsc{\textsc{.f}}\textsc{-obl}\textsc{\textsc{.f}}\textsc{=dem} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Now, as for me, it’s been three years that my brother is suffering from this illness. '
\z 
 
\ea \label{DG.70}
\textit{kuře wextē miđyo ānā čā šāzāyēwe} \\ 
\gll kuř-e wext-ē mi-đye-o āne=ā čā šāzā-ēwe \\ 
 son/boy\textsc{.m}\textsc{-def}\textsc{.dir} time\textsc{.m}\textsc{-indf} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} there prince\textsc{.m}\textsc{-indf} \\ 
\glt `The [sick] boy has become like a prince.'
\z 
 
\ea \label{DP.48}
\textit{māčo, dey min īney pīneym wāten} \\ 
\gll mi-āč-o dey min īney p=īney=m wāte=n \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} \textsc{dem.prox}\textsc{.obl}\textsc{.m}\textsc{.3sg} for\textsc{=dem}.prox\textsc{.obl}\textsc{.m}\textsc{.3sg}\textsc{=1sg} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt ` Jamsher Shah said [to Sibhan Agha], ‘When I swore an oath, I swore on this sparrow;'
\z 
 
\ea \label{DP.51}
\textit{dā dey dewršā gērdē bā deġāɫet kerū} \\ 
\gll dā dey dewr=šā ∅-gēr-dē bā deġāɫet ∅-ker-w \\ 
 \textsc{hort} \textsc{disc.ptcl} round\textsc{=3pl} \textsc{imp-}take\textsc{.prs}\textsc{-2pl} \textsc{hort} intervention\textsc{.m} \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt ` Jamsher Shah\footnote{Lit. He.} said [to his men], ‘Encircle them so that I can defeat [Sibhan Agha].'
\z 
 
\ea \label{ZP.8}
\textit{ētir wextē eđāw tāteš fewtšā kerdenū emreū xwāyšā kerdēne berdenšā lāw wēšā tā māwāyē yāwān pene} \\ 
\gll ētir wext-ē eđā=ū tāte=š fewt=šā kerde=n=ū emre=ū xwā-ī=šā kerdē=ne berde=n=šā lā=ū wē=šā tā māwā-yē yāwā=n pene \\ 
 \textsc{disc.ptcl} time\textsc{.m}\textsc{-indf} mother\textsc{.f}=and father\textsc{.m}\textsc{=3sg} death\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} command\textsc{.f}\textsc{\textsc{=ez.gen}} God\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} take\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} at\_the\_place\_of\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} until period\textsc{.m}\textsc{-indf} arrive\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to \\ 
\glt `When his parents passed away and departed this life\footnote{Lit. did the command of God.}, they (the family of his uncle) took him into their house [and raised him] until he grew up [into a young man]'
\z 
 
\ea \label{ZP.15}
\textit{čāgene ētir derziš wenānū bīyen be feqī} \\ 
\gll čāgene ētir derz=iš wenā=n=ū bīye=n be feqī \\ 
 there \textsc{disc.ptcl} lesson\textsc{.m}\textsc{=3sg} read\textsc{.pst}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to theologian\textsc{.m} \\ 
\glt `There, he studied and became a theologian.'
\z 
 
\ea \label{ZP.23}
\textit{e pā menteqew hewrāmānī ētir feqīyēwe bīyenū wenānšū} \\ 
\gll e=p=ā menteqe=ū hewrāmān-ī ētir feqī-ēwe bīye=n=ū wenā=n=š=ū \\ 
 in=by=\textsc{dem.dist} region\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} theologian\textsc{.m}\textsc{-indf} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{\textsc{=ez.gen}} read\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg}=and \\ 
\glt `In this Hawraman region, well, he was a theologian; he was literate.'
\z 
 
\ea \label{ZP.35}
\textit{ētir ewīč wēš pādšā bo} \\ 
\gll ētir ew=yč wē=š pādšā ∅-b-o \\ 
 \textsc{disc.ptcl} 3sg\textsc{.dir}\textsc{.m} \textsc{reflx}\textsc{=3sg} king\textsc{.m} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well, he was a king.'
\z 
 
\ea \label{ZP.40}
\textit{ētir xo ā zemāne eɫbet wesīle nebīyen yā māšīn yā čīwē} \\ 
\gll ētir xo ā zemān=e eɫbet wesīle ne-bīye=n yā māšīn yā čīw-ē \\ 
 \textsc{disc.ptcl} dis.ptc \textsc{dem.dist} time\textsc{.m}\textsc{=dem} of\_course tool\textsc{.m} \textsc{neg-}be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} or car\textsc{.m} or thing\textsc{-indf} \\ 
\glt `—Well, back then, indeed, there was no means of transport, be it a car or anything [else.]'
\z 
 
\ea \label{ZP.51}
\textit{ētir kināče keřū lāɫe bo māčmēš pene tewenū sirewē} \\ 
\gll ētir kināčē=e keře=ū lāɫ-e ∅-b-o m-āč-mē=š pene tewenī=ū sirewē \\ 
 \textsc{disc.ptcl} daughter\textsc{.dir}\textsc{.pl}\textsc{=ez}\textsc{.cmpd} deaf\textsc{.f}=and mute\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-1pl}\textsc{=3sg} to stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment\textsc{.f} \\ 
\glt `Since the girl was deaf-mute, we call it the healing stone.'
\z 
 
\ea \label{ZP.61}
\textit{her ā dāđe kero, ētir māčo, erē weɫā sirewā’ hurmēzowe miđo wene māroš hesārū činārārū} \\ 
\gll her ā dāđ=e ∅-ker-o ētir m-āč-o erē weɫā sirewā hur-m-ēz-o=we mi-đe-o wene m-ār-o=š hesār=ū činār-ā=re=ū \\ 
 emph \textsc{dem.dist} yell\textsc{.m}\textsc{=dem} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} yes indeed be cured\textsc{.pst}\textsc{.3sg} \textsc{pvb-}\textsc{ind-}rise\textsc{.prs}\textsc{-3sg}\textsc{=compl} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} at \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} pond\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \\ 
\glt `After he shouted [and] said, ‘Indeed she is healed’, then he rose, [and] set off, and took the girl to Hasar Chinara.'
\z 
 
\ea \label{ZP.82}
\textit{kināčē ētir mēwe serū hāɫē weše bowew} \\ 
\gll kināčē ētir m-ē=we ser=ū hāɫe-ē weš-e ∅-b-o=we=ū \\ 
 girl\textsc{.obl}\textsc{.f} \textsc{disc.ptcl} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} on\textsc{\textsc{=ez.gen}} state\textsc{.f}\textsc{-obl}\textsc{.f} well\textsc{-f} \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=compl}\textsc{\textsc{=ez.gen}} \\ 
\glt `Then, the girl was cured. She was healed.'
\z 
 
\ea \label{ZP.88}
\textit{ā wextīekey ʕerz kerā, dey ēme ī kināčē jeryāniš e čīnew} \\ 
\gll ā wext-ī-ekey ʕerz ∅-ker-ā dey ēme ī kināčē jeryān=iš e=čīne=ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} petition\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{1pl} \textsc{dem.prox} girl\textsc{.f} story\textsc{.m}\textsc{=3sg} in=like=and \\ 
\glt `Then they (her uncle) said, ‘This girl’s story is like this.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.109}
\textit{ētir min metāwū berūšo, hergīz meberūšo’} \\ 
\gll ētir min me-tāw-ū ∅-ber-ū=š=o hergīz me-ber-ū=š=o \\ 
 \textsc{disc.ptcl} \textsc{1sg} \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{sbjv-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} never \textsc{neg-}take\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=3sg}\textsc{=compl} \\ 
\glt `I cannot take her back. I will never take her back.’'
\z 
 
\ea \label{ZP.115}
\textit{māčā, dey īne ī pīyā ī ženīče ārdēne} \\ 
\gll m-āč-ā dey īne ī pīyā ī ženī=č=e ārdē=ne \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} dem.prox man\textsc{.m} dem.prox woman\textsc{.dir}f\textsc{=add}\textsc{=dem} bring\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \\ 
\glt `[and] said, ‘Well, this man has brought this woman [here]'
\z 
 
\ea \label{ZP.118}
\textit{xeɫkekeyč māčo, dey birā! to ħīčit nīyā min nīme kewčew hiɫošet midonew’} \\ 
\gll xeɫk-ekey=č m-āč-o dey birā to ħīč=it nīye=ā min nīme kewče=w hiɫoše=t mi-de-one=ū \\ 
 people\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{2sg} nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} half\textsc{.m} six\_kilos\textsc{\textsc{=ez.gen}} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}\textsc{\textsc{=ez.gen}} \\ 
\glt `[One among] the crowd said, ‘Brother! [Now that] you have nothing, I will give you about six kilos of cracked wheat [for the wedding ceremony]'
\z 
 
\ea \label{ZP.119}
\textit{yoīč māčo, dey minīč beʕzē pīyāzīt midonew’} \\ 
\gll yo=īč m-āč-o dey min=īč beʕzē pīyāz-ī=t mi-de-one=ū \\ 
 one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg}\textsc{=add} some onion\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Another said, ‘I will give you some onions.’'
\z 
 
\ea \label{ZP.121}
\textit{kābrāyēč māčo, dey birā, minīč serē heywānit midew} \\ 
\gll kābrā-ē=č m-āč-o dey birā min=īč ser-ē heywān=it mi-de-ū \\ 
 man\textsc{-indf}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother\textsc{.m} \textsc{1sg}\textsc{=add} \textsc{clf}\textsc{-indf} animal\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `A man said, ‘Brother, I will give you an animal.’'
\z 
 
\ea \label{ZP.124}
\textit{ētir tāze be nelowe dimāwew} \\ 
\gll ētir tāze be ne-l-o=we dimā=we=ū \\ 
 \textsc{disc.ptcl} anyway to \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} afterwards\textsc{=\textsc{post}}=and \\ 
\glt `After all, she shall not go back.'
\z 
 
\ea \label{ŽP.13}
\textit{ētir milo lāw āđī} \\ 
\gll ētir mi-l-o lā=ū āđī \\ 
 \textsc{disc.ptcl} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} 3sg\textsc{.obl}\textsc{.m} \\ 
\glt ` Anyhow, he went to him.'
\z 
 
\ea \label{ŽP.19}
\textit{ānā ew dīmū qeɫāw kitāy} \\ 
\gll āne=ā ew dīm=ū qeɫā=ū kitā-ī \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \textsc{dem.dist} side\textsc{.m}\textsc{\textsc{=ez.gen}} castle\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `—Look, it is [located] on the other side of the Kita castle'
\z 
 
\ea \label{ŽP.27}
\textit{māčā, dey to metāwī -sey misefāw- kučēš yewē berī nehāɫū pīr šelīyārā šānīšā pižgīnīšā bā bizyā’} \\ 
\gll m-āč-ā dey to me-tāw-ī sey misefā=ū kuč-ē=š yewe-ē ∅-ber-ī nehāɫ=ū pīr šelīyār-ā ∅-šān-ī=šā ∅-pižgīn-ī=šā bā bi-zye-ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} \textsc{2sg} \textsc{neg-}can\textsc{.prs}-\textsc{2sg} \textsc{pn} \textsc{pn}=and little\textsc{-indf}\textsc{=3sg} barley\textsc{.f}\textsc{-obl}\textsc{.f} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} valley\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.pl} \textsc{sbjv-}throw\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{sbjv-}scatter\textsc{.prs}-\textsc{2sg}\textsc{=3pl} \textsc{hort} \textsc{sbjv-}grow\textsc{.prs}\textsc{-3pl} \\ 
\glt `They said, ‘You are not able, Say Mustafa, to take some barley seeds to the valley of Pir Shaliyar, [and] scatter them [on the field], [and] plant them, so that they grow.'
\z 
 
\ea \label{ŽP.28}
\textit{ētir xo āđ meselen hetīm bo, lāw māmoyšā bo řūeš mebo wāčo, yā metāwū yā tāwū} \\ 
\gll ētir xo āđ meselen hetīm ∅-b-o lā=ū māmo-ī=šā ∅-b-o řūe=š me-b-o ∅-wāč-o yā me-tāw-ū yā ∅-tāw-ū \\ 
 \textsc{disc.ptcl} dis.ptc 3sg\textsc{.dir}\textsc{.m} for\_example orphan \textsc{ind-}be\textsc{.prs}\textsc{-3sg} at\_the\_place\_of\textsc{\textsc{=ez.gen}} paternal\_uncle\textsc{-obl}\textsc{.m}\textsc{=3pl} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} face\textsc{=3sg} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-3sg} either \textsc{neg-}can\textsc{.prs}\textsc{-\textsc{1sg}} either \textsc{ind-}can\textsc{.prs}\textsc{-\textsc{1sg}} \\ 
\glt `Well, he was an orphan. He was at his uncle’s. He was ashamed to say, ‘I cannot, or I can.’'
\z 
 
\ea \label{ŽP.71}
\textit{pīr heme yoso wēš milo temāšeš kero miđyo āne īnā hewāwe zāno īne ētir kārī xwāyīn} \\ 
\gll pīr heme yoso wē=š mi-l-o temāše=š ∅-ker-o mi-đye-o āne īnā-∅ hewā=we ∅-zān-o īne ētir kār-ī xwāyī=n \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{reflx}\textsc{=3sg} \textsc{ind-}go\textsc{.prs}\textsc{-3sg} looking\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}look\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} deic\textsc{-3sg}\textsc{.m} air\textsc{=\textsc{post}} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{disc.ptcl} task\textsc{.m}\textsc{-obl}\textsc{.m} divine\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Pir Hama Yoso went [and] looked at him; he noticed that he (Pir Shaliyar) was aloft in the air [up until a certain height] and came to realise that that was God’s work.'
\z 
 
\ea \label{ŽP.72}
\textit{ʕerzū heremēš kero, fermāwo be heremēš weɫāhī ētir ēme metāwmē, sey misefā metāwmē hīč kārēš pene kermē īlā kīyānmēš bilo weno, bilo kīyānmēš weno} \\ 
\gll ʕerz=ū hereme-ē=š ∅-ker-o ∅-fermāw-o be hereme-ē=š weɫāhī ētir ēme me-tāw-mē sey misefā me-tāw-mē hīč kār-ē=š pene ∅-ker-mē īlā ∅-kīyān-mē=š bi-l-o ∅-wen-o bi-l-o ∅-kīyān-mē=š ∅-wen-o \\ 
 petition\textsc{.m}\textsc{\textsc{=ez.gen}} wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} by wife\textsc{.f}\textsc{-obl}\textsc{.f}\textsc{=3sg} indeed \textsc{disc.ptcl} \textsc{1pl} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} \textsc{pn} \textsc{pn} \textsc{neg-}can\textsc{.prs}\textsc{-1pl} nothing task\textsc{.m}\textsc{-indf}\textsc{=3sg} to \textsc{sbjv-}do\textsc{.prs}\textsc{-1pl} except \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}read\textsc{-3sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg} \textsc{sbjv-}send\textsc{.prs}\textsc{-1pl}\textsc{=3sg} \textsc{sbjv-}read\textsc{-3sg} \\ 
\glt `He (Pir Hama Yoso) said to his wife, ‘Indeed, we can no longer be helpful to Say Mustafa,\footnote{Lit. we can do nothing to him.} except that we send him to study; that we send him, to go [and] study.'
\z 
 
\ea \label{ŽP.91}
\textit{fermāwo hezretū xos, dey to mišyo tešrīf berī bilīwe, pey menteqew wētāw} \\ 
\gll ∅-fermāw-o hezret=ū xos dey to mišyo tešrīf ∅-ber-ī bi-l-ī=we pey menteqe=ū wē=tā=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{disc.ptcl} \textsc{2sg} aux honouring \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} \textsc{sbjv-}go\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=compl} to region=and \textsc{reflx}=\textsc{2pl}=and \\ 
\glt `His Highness Khos said, ‘You should leave [here and] return to your region.'
\z 
 
\ea \label{ŽP.98}
\textit{ētir herčī mēnē ā zemāne wesīlew māšīnū čīwī mebo} \\ 
\gll ētir herčī m-ē-nē ā zemān=e wesīle=ū māšīn=ū čīw-ī me-b-o \\ 
 \textsc{disc.ptcl} what\_ever \textsc{ind-}come\textsc{.prs}\textsc{-3pl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} tool\textsc{.m}=and car\textsc{.m}=and thing\textsc{-obl}\textsc{.m} \textsc{neg-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Well, in that period, there were no means of transportation, cars, etc.'
\z 
 
\ea \label{ŽP.102}
\textit{hejījīyeke māčo, dey ī min īnā nizīkū hejījī kotenāwe} \\ 
\gll hejījī-eke m-āč-o dey ī min īnā-∅ nizīk=ū hejīj-ī kote=nā=we \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dem.prox \textsc{1sg} deic\textsc{-3sg}\textsc{.m} close\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} fall\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}}\textsc{=compl} \\ 
\glt `The man from Hajij said, ‘I have arrived [at a place] close to Hajij.’'
\z 
 
\ea \label{ŽP.116}
\textit{mēwe ētir ā zemāne pīr šelīyār mēwe menteqe des kero, jārē āne minyā māčāš meřū āsinī} \\ 
\gll m-ē=we ētir ā zemān=e pīr šelīyār m-ē=we menteqe des=t ∅-ker-o jārē āne mi-nye-ā m-āč-ā=š meře=ū āsin-ī \\ 
 \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m}\textsc{=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} region hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} once \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{ind-}put\textsc{.prs}\textsc{-3pl} \textsc{ind-}say\textsc{.prs}\textsc{-3pl}\textsc{=3sg} cave\textsc{.f}\textsc{\textsc{=ez.gen}} iron\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) returned [to Hawraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ‘Iron cave.’'
\z 
 
\ea \label{ŽP.117}
\textit{ētir hezretū xosī kīyāseno} \\ 
\gll ētir hezret=ū xos-ī kīyāse=n=o \\ 
 \textsc{disc.ptcl} his\_highness\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m} send\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=compl} \\ 
\glt `Anyhow, His Highness Khos had sent him.'
\z 
 
\ea \label{ŽP.143}
\textit{milowe yānew ētir pāne nāmēwe peyā kero} \\ 
\gll mi-l-o=we yāne=ū ētir p=āne nāmē-ewe peyā ∅-ker-o \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} house\textsc{.m}=and \textsc{disc.ptcl} at=\textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} name\textsc{.f}\textsc{-indf} visible \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \\ 
\glt `He went there (to Hawraman) and got a bit famous.\footnote{Lit. He found a name.}'
\z 
 
\ea \label{ŽP.166}
\textit{lālo gino gelū kināčē māroš dey degā be degāw kēšwer be kēšwerū māroš silēne} \\ 
\gll lālo ∅-gin-o gel=ū kināčē m-ār-o=š dey degā be degā=ū kēšwer be kēšwer=ū m-ār-o=š silēne \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} with\textsc{\textsc{=ez.gen}} girl\textsc{.dir}\textsc{.f} \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{disc.ptcl} village\textsc{.obl}\textsc{.f} by village\textsc{.f}=and country by country=and \textsc{ind-}bring\textsc{.prs}\textsc{-3sg}\textsc{=3sg} \textsc{pn} \\ 
\glt `The uncle set off on the road with the girl. He took her. He took her village by village and country by country until [they arrived] at Slen.'
\z 
 
\ea \label{ŽP.185}
\textit{ēti nāmēšā nīyā tewenekē tewenū sirewāy”} \\ 
\gll ēti nāmē=šā nīyā tewenī-ekē tewenī=ū sirewāy \\ 
 \textsc{disc.ptcl} name\textsc{.f}\textsc{=3pl} put\textsc{.pst} stone\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f} stone\textsc{.f}\textsc{\textsc{=ez.gen}} treatment.inf \\ 
\glt `That is why the stone is called Tawanu Sraway, ‘the healing stone.’'
\z 
 
\ea \label{ŽP.198}
\textit{fermāwo, weɫā čūn ānā bēheyānī kerūt penū ʕāɫemī} \\ 
\gll ∅-fermāw-o weɫā čūn āne=ā bēheyā=nī ∅-ker-ū=t pen=ū ʕāɫem-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed as \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} shameless\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}}\textsc{=\textsc{2sg}} trouble\textsc{.m}\textsc{\textsc{=ez.gen}} world\textsc{-obl}\textsc{.m} \\ 
\glt `He (Pir Shaliyar) said, ‘Indeed, since you are so indecent, I will disgrace you.’'
\z 
 
\ea \label{ŽP.201}
\textit{tenūrū dēwā ētir pēse gāw til bowe e čā werāwerū girdū mexeleqeɫāw hewrāmānīwe dēweke til bowe ginore soɫāwe} \\ 
\gll tenūr=ū dēw-ā ētir pēse gāw til ∅-b-o=we e=čā werāwer=ū gird=ū mexeleqeɫā=ū hewrāmān-ī=we dēw-eke til ∅-b-o=we ∅-gin-o=re soɫāwe \\ 
 oven\textsc{.m}\textsc{\textsc{=ez.gen}} demon\textsc{.m}\textsc{-obl}\textsc{.pl} \textsc{disc.ptcl} as\_if bull\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} in=there in\_front\_of\textsc{\textsc{=ez.gen}} all\textsc{\textsc{=ez.gen}} people\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} demon\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} rolling \textsc{ind-}be\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{post}} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg}\textsc{=\textsc{povb}} \textsc{pn} \\ 
\glt ` Then it was as if a cow had rolled down; the [demon] rolled down and fell [into a place called] Solawa before all the people’s eyes.'
\z 
 
\ea \label{ŽP.210}
\textit{pīr šelīyārīč fermāwo, dey min kābrāyēnā āhā xeplāzeřetīnī werūneū īne pādšāzāyēneū derdū min meweronew} \\ 
\gll pīr šelīyār=īč ∅-fermāw-o dey min kābrā-ē=nā āhā xeplāzeřetīnī ∅-wer-ūne=ū īne pādšāzāyē=ne=ū derde=ū min me-wer-one=ū \\ 
 \textsc{pn} \textsc{pn}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf}\textsc{=cop}\textsc{.\textsc{1sg}} \textsc{intj} \textsc{pn} \textsc{ind-}eat\textsc{.prs}\textsc{-\textsc{1sg}}=and dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} princess\textsc{=\textsc{post}}\textsc{\textsc{=ez.gen}} illness\textsc{.f}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{neg-}eat\textsc{.prs}\textsc{-3sg}=and \\ 
\glt `Pir Shaliyar said, ‘Well, I am a fellow —look—who eats xaplazaratine (a bread made of corn flour’. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{ŽP.220}
\textit{ā wextī māčā, dey qurbān} \\ 
\gll ā wext-ī m-āč-ā dey qurbān \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} sir.\textsc{voc} \\ 
\glt `Then, they (people) said, Oh, Sir!’'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.260}
\textit{pādšā fermāwo, dey mubārekeš bo} \\ 
\gll pādšā ∅-fermāw-o dey mubārek-e=š ∅-b-o \\ 
 king\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} blessed\textsc{-f}\textsc{=3sg} \textsc{sbjv-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `The king said, ‘May he be blessed with happiness.'
\z 
 
\ea \label{ŽP.263}
\textit{ētir to lue meylū wētā’} \\ 
\gll ētir to ∅-lu-e meyl=ū wē=t=ā \\ 
 \textsc{disc.ptcl} \textsc{2sg} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} willingness\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then it’s up to you [to stay in Hawraman]'
\z 
 
\ea \label{ŽP.266}
\textit{ēti melowe āđīč melowe} \\ 
\gll ēti me-l-o=we āđ=īč me-l-o=we \\ 
 \textsc{disc.ptcl} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{neg-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} \\ 
\glt `After all, he didn’t go back [to Egypt]. He did not go back either.'
\z 
 
\ea \label{BP.45}
\textit{milā xeɫk pāne milo miđyo īne ā xerīkū genekārīnē ī kābrāy je āmēnē, ī meʕmūrē xerīkū genekārīnēw} \\ 
\gll mi-l-ā xeɫk p=ā=ne mi-l-o mi-đye-o īne=ā xerīk=ū genekārī=nē ī kābrā-ī je āmē=nē ī meʕmūr-ē xerīk=ū genekārī=nē=ū \\ 
 \textsc{ind-}go.prs\textsc{-3pl} people\textsc{.m} at=\textsc{dem.dist}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3sg} \textsc{ind-}look.prs\textsc{-3sg} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} man\textsc{-obl}\textsc{.m} from come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \textsc{dem.prox} officer\textsc{.m}\textsc{-dir}\textsc{.pl} busy\textsc{=ez.gen} debauchery\textsc{.m}\textsc{=cop}\textsc{.3pl}=and \\ 
\glt `People realised that these men were busy engaging in debauchery; these officers were busy engaing in debauchery.'
\z 
 
\ea \label{BP.48}
\textit{ētir zimsānē melāwe biřew bārekey berāwe} \\ 
\gll ētir zimsān-ē me-l-ā=we biře=ū bāre-ekey ∅-ber-ā=we \\ 
 \textsc{disc.ptcl} winter\textsc{.m}\textsc{-indf} \textsc{neg-}go.prs\textsc{-3pl}\textsc{=compl} portion\textsc{.m}=and profit\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}take.prs\textsc{-3pl}\textsc{=compl} \\ 
\glt `They didn’t go back one winter to take back the taxes.'
\z 
 
\ea \label{BP.52}
\textit{dey ēšew herkes mēmāniš hen, ēšew herkes mēmāniš hen, mēmānekeš šewē witē, sereš biřo} \\ 
\gll dey ēšew herkes mēmān=iš hen-∅ ēšew herkes mēmān=iš hen-∅ mēmān-eke=š šewe-ē wit-ē sere=š ∅-biř-o \\ 
 \textsc{disc.ptcl} tonight everyone guest\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} tonight everyone guest\textsc{.m}\textsc{=3sg} exist\textsc{-3sg}\textsc{.m} guest\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} night.f\textsc{-obl}.f sleep\textsc{.pst}\textsc{-3pl} head\textsc{.m}\textsc{=3sg} \textsc{sbjv-}cut.prs\textsc{-3sg} \\ 
\glt `‘Tonight, whoever has guests, when the guest sleeps at night, he shall decapitate him.'
\z 
 
\ea \label{BP.73}
\textit{ewē mūsā ētir be mišār serešā biřā} \\ 
\gll ewē m-ūs-ā ētir be mišār sere=šā ∅-biř-ā \\ 
 3pl\textsc{.dir} \textsc{ind-}sleep.prs\textsc{-3pl} \textsc{disc.ptcl} to hand\_saw\textsc{.m} head\textsc{.m}\textsc{=3pl} \textsc{ind-}cut.prs\textsc{-3pl} \\ 
\glt `They (the guests) had slept. He decapitated them with a hand saw.'
\z 
 
\ea \label{BP.86}
\textit{āđīč mē werew māčo, fermāwo, dā dey dey jārēne jimdēre’} \\ 
\gll āđ=īč m-ē wer=ew m-āč-o ∅-fermāw-o dā dey dey jār-ē=ne ∅-jim-dē=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}come.prs\textsc{.3sg} front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}say.prs\textsc{-3sg} \textsc{ind-}say.prs\textsc{-3sg} give\textsc{.pst} \textsc{disc.ptcl} \textsc{disc.ptcl} time\textsc{.m}\textsc{-indf}\textsc{=\textsc{post}} \textsc{imp-}move.prs-\textsc{2pl}\textsc{=\textsc{povb}} \\ 
\glt `He (Baba Khwada) came forward [and] said, ‘Please, set off the way all of you.’'
\z 
 
\ea \label{BP.89}
\textit{ētir ēme xo diruē meyeymē} \\ 
\gll ētir ēme xo diruē me-de-ymē \\ 
 \textsc{disc.ptcl} \textsc{1pl} dis.ptc lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `We are not going to lie.'
\z 
 
\ea \label{BP.98}
\textit{ā wextī ēti seʕbe milā milāne mizgī, mizyāre beynē gināre řā beynē bilā pey ʕērāqī jemʕē bā ħemey ġeybīū bābā xuāđā ginā weɫēwe} \\ 
\gll ā wext-ī ēti seʕbe mi-l-ā mi-l-ān=e mizgī mi- zye-ā=re beynē ∅-gin-ā=re řā beynē bi-l-ā pey ʕērāq-ī jemʕ-ē ∅-b-ā ħeme=ī ġeybī=ū bābā xuāđā ∅-gin-ā weɫē=we \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} morning.f \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}go.prs\textsc{-3pl}=drct mosque\textsc{.m} \textsc{ind-}go\_out.prs\textsc{-3pl}\textsc{=\textsc{povb}} a\_while \textsc{ind-}fall.prs\textsc{-3pl}\textsc{=\textsc{povb}} road.f a\_while \textsc{sbjv-}go.prs\textsc{-3pl} for \textsc{pn}\textsc{-obl}\textsc{.m} added\textsc{-pstc}\textsc{.3sg} \textsc{ind-}be.prs\textsc{-3pl} \textsc{pn}=ez.\textsc{attr} invisible=and grandfather\textsc{.m} \textsc{pn} \textsc{ind-}fall.prs\textsc{-3pl} front\textsc{=\textsc{post}} \\ 
\glt `So, in the morning they went to the mosque. They went a little way towards Iraq. They all gathered. Hama Ghaybi and Baba Khwada went in front of them.'
\z 
 
\ea \label{BP.99}
\textit{ewē šūnīšāre milā ētir her werew milā řārē} \\ 
\gll ewē šūnī=šā=re mi-l-ā ētir her wer=ew mi-l-ā řā=rē \\ 
 3pl\textsc{.dir} track.f\textsc{=3pl}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} \textsc{disc.ptcl} just front\textsc{.m}\textsc{=\textsc{post}} \textsc{ind-}go.prs\textsc{-3pl} road.f\textsc{=\textsc{post}} \\ 
\glt `They (the people) followed them. They walked slowly.'
\z 
 
\ea \label{BP.115}
\textit{hewrāmī tā meřeber yānē hewrāmīyē tā meřeber āmē ētir girđ gēɫēnēwe čāge} \\ 
\gll hewrāmī tā meřeber yānē hewrāmī-ē tā meřeber āmā-ē ētir girđ gēɫē=nē=we č=āge \\ 
 Hawrami until \textsc{pn} that\_means Hawrami\textsc{-dir}\textsc{.pl} until \textsc{pn} come\textsc{.pst}\textsc{-3pl} \textsc{disc.ptcl} all turn\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=compl} from=there \\ 
\glt `hawrāmī tā mařabar means the people of Hawraman came as far as Marabar. Then, they all returned to Hawraman from there.'
\z 
 
\ea \label{BP.117}
\textit{ētir āđē wātenšā, ēme diruē meyeymē} \\ 
\gll ētir āđē wāte=n=šā ēme diruē me-de-ymē \\ 
 \textsc{disc.ptcl} 3pl\textsc{.dir} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} \textsc{1pl} lie.f \textsc{neg-}give.prs \textsc{1pl} \\ 
\glt `They said, ‘We are not going to lie.'
\z 
 
\ea \label{BP.129}
\textit{āġe māčo, dey lodē berdē kerdēšā zīnān} \\ 
\gll āġe m-āč-o dey ∅-lo-dē ∅-ber-dē ∅-ker-dē=šā zīnān \\ 
 agha\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} \textsc{imp-}go.prs.\textsc{imp-}\textsc{2pl} \textsc{imp-}take.prs-\textsc{2pl} \textsc{imp-}do.prs-\textsc{2pl}\textsc{=3pl} prison\textsc{.m} \\ 
\glt `The agha said [to his men], ‘Go [and] take them; put them in jail!’'
\z 
 
\ea \label{BP.149}
\textit{ētir māwāš fire bo yā kem bo her činne māngē bo her činne wextē bo wextē milo serū řūčinekey nānšā pey wizorew wextē miđyo hešpišā dēnēne} \\ 
\gll ētir māwā=š fire ∅-b-o yā kem ∅-b-o her činne mānge-ē ∅-b-o her činne wext-ē ∅-b-o wext-ē mi-l-o ser=ū řūčine-ekey nān=šā pey ∅-wiz-o=re=ū wext-ē mi-đye-o hešpišī-ā dē=nē=ne \\ 
 \textsc{disc.ptcl} period\textsc{.m}\textsc{=3sg} a\_lot \textsc{ind-}be.prs\textsc{-3sg} either little \textsc{ind-}be.prs\textsc{-3sg} just some month.f\textsc{-dir}\textsc{.pl} \textsc{ind-}be.prs\textsc{-3sg} just some time\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{ind-}be.prs\textsc{-3sg} time\textsc{.m}\textsc{-indf} \textsc{ind-}go.prs\textsc{-3sg} on\textsc{=ez.gen} chimney\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} bread\textsc{.m}\textsc{=3pl} for \textsc{ind-}throw.prs\textsc{-3sg}\textsc{=\textsc{povb}}=and time\textsc{.m}\textsc{-indf} \textsc{ind-}look.prs\textsc{-3sg} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} \\ 
\glt `Their imprisonment lasted either a long time or short time. It took them months. Once, when he (Hama Wirdikla) went to the chimney to throw food for them, he realised that they had gotten lice.'
\z 
 
\ea \label{BP.156}
\textit{ā dey hešpišā dēndēne īse ēme tekbīrmā čēš bo čēš kermēū} \\ 
\gll =ā dey hešpišī-ā dē=ndē=ne īse ēme tekbīr=mā čēš ∅-b-o čēš ∅-ker-mē=ū \\ 
=\textsc{ptcl} \textsc{disc.ptcl} lice.f\textsc{-obl}\textsc{.pl} give\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}.\textsc{2pl}\textsc{=\textsc{post}} now \textsc{1pl} advice\textsc{=\textsc{1pl}} what \textsc{sbjv-}be.prs\textsc{-3sg} what \textsc{sbjv-}do.prs\textsc{-\textsc{1pl}}=and \\ 
\glt `Look, you have gotten lice. What is the plan now for us? What should we do?'
\z 
 
\ea \label{BP.160}
\textit{bābā xuāđā fermāwo, dey eger to ādiznī, eger to ādiznī pey ēmew nāřihetnī pey ēme, lue gēɫe pī šārre} \\ 
\gll bābā xuāđā ∅-fermāw-o dey eger to ādiz=nī eger to ādiz=nī pey ēme=ū nāřihet=nī pey ēme ∅-lu-e ∅-gēɫ-e p=ī šār=re \\ 
 grandfather\textsc{.m} \textsc{pn} \textsc{ind-}say.prs\textsc{-3sg} \textsc{disc.ptcl} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} if \textsc{2sg} upset\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl}=and sad\textsc{=cop}\textsc{.\textsc{2sg}} for \textsc{1pl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{ind-}wander.prs-\textsc{2sg} at=\textsc{dem.prox} city\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `Baba Khwada said, ‘If you are upset; if you are worried about us; if you are sad about us, go [and] wander in this town! '
\z 
 
\ea \label{BP.164}
\textit{gošiš nedārāyne ētir jā bowe xeber bārewe} \\ 
\gll goš=iš ne-dārā-ī=ne ētir jā b-o=we xeber b-ār-e=we \\ 
 ear\textsc{.m}\textsc{=3sg} \textsc{neg-}hold\textsc{.pst}-\textsc{2sg}\textsc{=\textsc{povb}} \textsc{disc.ptcl} afterwards \textsc{imp-}come.prs\textsc{.imp}\textsc{.\textsc{2sg}}\textsc{=compl} news\textsc{.dir}\textsc{.m} \textsc{imp-}bring.prs-\textsc{2sg}\textsc{=compl} \\ 
\glt `If they don’t listen to you, then, come back and let us know.'
\z 
 
\ea \label{BP.174}
\textit{zīnāneke yekser weɫādīm bo šāreke ētir ginowe serū āđēū} \\ 
\gll zīnān-eke yekser weɫādīm ∅-b-o šār-eke ētir ∅-gin-o=we ser=ū āđē=ū \\ 
 prison\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} totally collapsed \textsc{ind-}be.prs\textsc{-3sg} city\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{disc.ptcl} \textsc{ind-}fall.prs\textsc{-3sg}\textsc{=compl} on\textsc{=ez.gen} 3pl\textsc{.dir}=and \\ 
\glt `The prison totally collapsed. The prison collapsed on them. '
\z 
 
\ea \label{BP.179}
\textit{mēnē bēnēwe pey hewrāmānī, mēnē bēnēwe pey hewrāmānī ētir zīnānne mirexesē bāw āzāē bāw} \\ 
\gll m-ē-nē b-ē-nē=we pey hewrāmān-ī m-ē-nē b-ē-nē=we pey hewrāmān-ī ētir zīnān=ne mirexes-ē b-ā=ū āzā-ē b-ā=ū \\ 
 \textsc{ind-}come.prs\textsc{-3pl} \textsc{sbjv-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}come.prs\textsc{-3pl} \textsc{sbjv-}come.prs\textsc{.3sg}\textsc{-3pl}\textsc{=compl} for \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} prison\textsc{.m}\textsc{=\textsc{post}} released\textsc{\textsc{-pl}} be.prs\textsc{-3pl}=and free\textsc{\textsc{-pl}} be.prs\textsc{-3pl}=and \\ 
\glt `They were on their way to Hawraman. They were on their way to Hawraman. They were liberated from the prison. They got free.'
\z 
 
\ea \label{BP.197}
\textit{ā wextī ētir āđē zānā ġeyb bīyen zemīnegere luān} \\ 
\gll ā wext-ī ētir āđē ∅-zān-ā ġeyb bīye=n zemīn-eke=re luā=n \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{disc.ptcl} 3pl\textsc{.dir} \textsc{ind-}know.prs\textsc{-3pl} disappearance be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} land\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{post}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then, they realised that he had disappeared into the ground. '
\z 
 
\ea \label{ŠJ.5}
\textit{-her čā bīyen welē ētir tāze peyđā bīyenū bīyen pīyāw bīyen šēxū mēšhūr bīyen} \\ 
\gll her čā bīye=n welē ētir tāze peyđā bīye=n=ū bīye=n pīyā=ū bīye=n šēx=ū mēšhūr bīye=n \\ 
 emph there be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} but \textsc{disc.ptcl} anyway visible be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} man\textsc{.m}=and be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} sheikh\textsc{.m}=and famous be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `—Well, he was already there [in Biyare], but he had just come out as a gentleman, as a sheikh, he had just become famous.'
\z 
 
\ea \label{ŠJ.45}
\textit{weɫē bīyēnmē āro duē řoē ētir mezānmē koge milmē} \\ 
\gll weɫē bīyē=nmē āro duē řo-ē ētir me-zān-mē koge mi-l-mē \\ 
 front be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} today two day\textsc{.m}\textsc{-dir}\textsc{.pl} \textsc{disc.ptcl} \textsc{neg-}know.prs\textsc{-\textsc{1pl}} where \textsc{ind-}go.prs\textsc{-\textsc{1pl}} \\ 
\glt `We have spread out for two days; we didn’t know where we were going.’'
\z 
 
\ea \label{ŠJ.48}
\textit{perčī nāqāyme miđyo bīyen ženī ēhwāɫiš fāřyān} \\ 
\gll per- č=ī nāqāym=e mi-đye-o bīye=n ženī ēhwāɫ=iš fāřyā=n \\ 
 \textsc{ptcl}- at=\textsc{dem.prox} sudden\textsc{=dem} \textsc{ind-}look.prs\textsc{-3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} woman.f demeanour\textsc{=3sg} change\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `All of a sudden, he realised that he had turned into a woman; his demeanour had changed;'
\z 
 
\ea \label{ŠJ.50}
\textit{kirž ħāɫīš bo zāno be wēš ētir} \\ 
\gll kirž ħāɫī=š ∅-b-o ∅-zān-o be wē=š ētir \\ 
 quick understood\textsc{=3sg} \textsc{sbjv-}be.prs\textsc{-3sg} \textsc{ind-}know.prs\textsc{-3sg} to \textsc{reflx}\textsc{=3sg} \textsc{disc.ptcl} \\ 
\glt `He understood quickly [what had happened]; He knew about it.'
\z 
 
\ea \label{ŠJ.97}
\textit{ētir bo welhāsil bo muxlīsū āđī jāfir sānū luhonī} \\ 
\gll ētir ∅-b-o welhāsil ∅-b-o muxlīs=ū āđī jāfir sān=ū luhon-ī \\ 
 \textsc{disc.ptcl} \textsc{sbjv-}be.prs\textsc{-3sg} eventually \textsc{ind-}be.prs\textsc{-3sg} disciple\textsc{.m}\textsc{=ez.gen} 3sg\textsc{.obl}\textsc{.m} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `Anyway, Jafir Sān from Luhon became his disciple.'
\z 
 
\ea \label{ŠJ.101}
\textit{dey āro be kē wāčī derđwārde bo je dinyāne kes nemenen} \\ 
\gll dey āro be kē ∅-wāč-ī derđwārde ∅-b-o je dinyā=ne kes ne-mene=n \\ 
 \textsc{disc.ptcl} today to who \textsc{sbjv-}say.prs.\textsc{sbjv-}\textsc{2sg} qualified\textsc{.m} \textsc{sbjv-}be.prs\textsc{-3sg} from world\textsc{.dir}.f\textsc{=\textsc{post}} person\textsc{.m} \textsc{neg-}remain\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Nowadays, who can you refer to in the world who would be qualified? No one has remained.'
\z 
 
\ea \label{KŠ.29}
\textit{āđīč wāt, dey xāsā min gino šūnī[šre]} \\ 
\gll āđ=īč wāt dey xās=ā min ∅-gin-o šūnī=š=re \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst} \textsc{disc.ptcl} fine\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1sg} \textsc{ind-}fall\textsc{.prs}\textsc{-3sg} track\textsc{.f}\textsc{=3sg}\textsc{=\textsc{post}} \\ 
\glt `He (the shepherd) said, ‘All right, I will look after her (the goat).’'
\z 
 
\ea \label{KŠ.84}
\textit{īnā qāqezēčiš dānā pey} \\ 
\gll īne=ā qāqez-ē=č=iš dā=nā pey \\ 
 \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} letter\textsc{.m}\textsc{-indf}\textsc{=add}\textsc{=3sg} give\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to \\ 
\glt `Look, he has given me a letter.'
\z 
 
\ea \label{KŠ.86}
\textit{wātiš dey lue āney fiɫānīū fiɫānī čiře ī qewmāme bā bēnēwe qāqezekey wenā} \\ 
\gll wāt=iš dey ∅-lu-e āney fiɫānī=ū fiɫānī ∅-čiř-e ī qewm-ā=m=e bā b-ē-nē=we qāqez-ekey ∅-wen-ā \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{dem.dist}\textsc{.obl}\textsc{.m}\textsc{.3sg} such\_and\_such\_person=and such\_and\_such\_person \textsc{imp-}call\textsc{.prs}-\textsc{2sg} \textsc{dem.prox} relative\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{hort} \textsc{sbjv-}come\textsc{.prs}\textsc{.3sg}\textsc{-3pl}\textsc{=compl} letter\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{sbjv-}read\textsc{.prs}\textsc{-3pl} \\ 
\glt `She said, ‘Then go, and summon such-and-such, our relatives, so that they come and read the letter.’'
\z 
 
\ea \label{KŠ.91}
\textit{ētir ew čone āyā xeber zāno yā mezāno,} \\ 
\gll ētir ew č=o=ne āyā xeber ∅-zān-o yā me-zān-o \\ 
 \textsc{disc.ptcl} \textsc{dem.dist} in=\textsc{dem.dist}\textsc{=\textsc{post}} whether news\textsc{.dir}\textsc{.m} \textsc{ind-}know\textsc{.prs}\textsc{-3sg} or \textsc{neg-}know\textsc{.prs}\textsc{-3sg} \\ 
\glt `[It is not clear] whether he (the king) was aware of the wedding or not. '
\z 
 
\ea \label{ŽH.11}
\textit{māčo, dey bilmē minīč mewt čene} \\ 
\gll m-āč-o dey bi-l-mē min=īč m-e-ū=t čene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} 1sg\textsc{=add} \textsc{ind-}come\textsc{.prs}\textsc{-1sg}\textsc{=\textsc{2sg}} with \\ 
\glt `He (Hayas) said, ‘Let’s go, I am also coming with you.’'
\z 
 
\ea \label{ŽH.29}
\textit{wātiš, dey ētir?’} \\ 
\gll wāt=iš dey ētir \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} else \\ 
\glt `She said, ‘What else?’'
\z 
 
\ea \label{ŽH.38}
\textit{vātiš, dey tāte dey bā čīwēweš pey bere!’} \\ 
\gll vāt=iš dey tāte dey bā čīw-ēwe=š pey ∅-ber-e \\ 
 say\textsc{.pst}\textsc{=3sg} \textsc{disc.ptcl} father\textsc{.m} \textsc{disc.ptcl} \textsc{hort} thing\textsc{-indf}\textsc{=3sg} for \textsc{imp-}take\textsc{.prs}-\textsc{2sg} \\ 
\glt `She said, ‘Father, get him something.’'
\z 
 
\ea \label{ŽH.58}
\textit{māčo, dey} \\ 
\gll m-āč-o dey \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \\ 
\glt `He said, ‘Go ahead!’'
\z 
 
\ea \label{ŽH.59}
\textit{māčo, dey min šūyš kerū pene’} \\ 
\gll m-āč-o dey min šū-ī=š ∅-ker-ū pene \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} 1sg husband\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{=3sg} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} to \\ 
\glt `She said, ‘I will marry him.'
\z 
 
\ea \label{ŽH.79}
\textit{ā wextī yewāšē māčo, dey tehemuɫ kere bā hēɫeřūenit pey kerū, hāɫīt kerū'} \\ 
\gll ā wext-ī yewāšē m-āč-o dey tehemuɫ ∅-ker-e bā hēɫeřūen=it pey ∅-ker-ū hāɫī=t ∅-ker-ū \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} then \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} patience \textsc{imp-}do\textsc{.prs}-\textsc{2sg} \textsc{hort} fried\_egg\textsc{=\textsc{2sg}} for \textsc{sbjv-}do\textsc{.prs}\textsc{-1sg} understood\textsc{=\textsc{2sg}} \textsc{ind-}do\textsc{.prs}\textsc{-1sg} \\ 
\glt `Then she said, ‘All right, wait a little bit, let me cook you fried eggs, [and] make you understand.'
\z 
 
\ea \label{ŽH.80}
\textit{māčo, dey} \\ 
\gll m-āč-o dey \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \\ 
\glt `He said, ‘Go ahead!’'
\z 
 
\ea \label{ŽH.103}
\textit{āđīč māčoš pene, ānā be mehremū wēm qebūɫim kerdī! ānā qerār desit neteknūtčū} \\ 
\gll āđ=īč m-āč-o=š pene āne=ā be mehrem=ū wē=m qebūɫ=im kerd-ī āne=ā qerār des=it ne-tekn--ū=č=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} by kindred\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1sg} accepted\textsc{=1sg} do\textsc{.pst}-\textsc{2sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} arrangement hand\textsc{.m}\textsc{=\textsc{2sg}} \textsc{neg-}touch\_sb\textsc{.prs}\textsc{-1sg}\textsc{=add}=and \\ 
\glt `He (the sultan) told her, ‘That’s it, I have accepted you as my kindred. I swear [in this moment] that I won’t touch you.'
\z 
 
\ea \label{ŽH.104}
\textit{ānā dersē ʕālit dānē} \\ 
\gll āne=ā ders-ē ʕāl=it dā-ānē \\ 
 \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} lesson\textsc{.m}\textsc{-indf} good\textsc{=\textsc{2sg}} give\textsc{.pst}\textsc{-1sg} \\ 
\glt `You gave me a good lesson.'
\z 
 
\ea \label{ŽH.119}
\textit{dey duʕāw seɫāmetīt} \\ 
\gll dey duʕā=ū seɫāmetī=t \\ 
 \textsc{disc.ptcl} prayer\textsc{\textsc{.f}}\textsc{=ez}\textsc{.gen} health\textsc{=\textsc{2sg}} \\ 
\glt `That’s it. May good prayers  be for you.'
\z 



\ea \label{RE.5}
\textit{āmān lūān wāteniš, birā gīyān! min xū wēt zānī hāwmāɫitnā} \\ 
\gll āmā=n lūā=n wāte=n=iš birā gīyān min xū wē=t ∅-zān-ī hāwmāɫ=it=nā \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} brother\textsc{.m} dear.\textsc{voc} \textsc{1sg} \textsc{disc.ptcl} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} neighbour\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `They would go [and] say, ‘Dear brother, you know [well] that I am your neighbour.'
\z 
 
\ea \label{RE.10}
\textit{ewīč wāten qey čēšī kero, ānā nīyām lā pāɫāket welē be qewrū be hurmet} \\ 
\gll ew=īč wāte=n qey čēš-ī ∅-ker-o āne=ā nīyā=m lā pāɫā-eke=t welē be qewr=ū be hurmet \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} limit what\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} put\textsc{.pst}\textsc{=\textsc{1sg}} at\_the\_place\_of shoe\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=\textsc{2sg}} but to appreciation\textsc{.m}\textsc{\textsc{=ez.gen}} to respect\textsc{.m} \\ 
\glt `Then, he (the girl’s father) would say, ‘She is yours,\footnote{Lit. I have put [her] on your shoes.} but [only treat us] with appreciation and respect.'
\z 
 
\ea \label{RE.11}
\textit{ētir kābrā fers kere zemānū ā ānē hītē pāɫē bīyēnēū čwār gezē pārče bīyenū} \\ 
\gll ētir kābrā fers ∅-ker-e zemān=ū ā ānē hīt-ē pāɫā-ē bīyē=nē=ū čwār gez-ē pārče bīye=n=ū \\ 
 \textsc{disc.ptcl} man guess\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg} time\textsc{.m}\textsc{=ez.gen} \textsc{dem.dist} \textsc{dem.dist}\textsc{.3pl} pair\textsc{-indf} shoe\textsc{-indf} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl}=and four ell\textsc{.m}\textsc{-indf} textile\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `Then, let us say,\footnote{Lit. imagine.} at the time, [the wedding gift] consisted of a pair of shoes [and] four gaz\footnote{A unit of measure equivalent to 72 cm.} of textile.'
\z 
 
\ea \label{RE.16}
\textit{ey pīyā ētir emīnū yāne bīyenū xizmetkār bīyen} \\ 
\gll ey pīyā ētir emīn=ū yāne bīye=n=ū xizmetkār bīye=n \\ 
 \textsc{dem.prox} man\textsc{.m} \textsc{disc.ptcl} confidant\textsc{.m}\textsc{=ez.gen} house\textsc{.m}\textsc{.dir}.sg be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez.gen} servant\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `There was this man who was the family’s confidant. He was a servant.'
\z 
 
\ea \label{RE.22}
\textit{šūnū ženekēčre ā wextīekey ētir hemāhengīšā weš bīyen} \\ 
\gll šūn=ū ženī-ekē=č=re ā wext-ī-ekey ētir hemāhengī=šā weš bīye=n \\ 
 after\textsc{=ez.gen} woman\textsc{.f}\textsc{-def}\textsc{.obl}\textsc{.f}\textsc{=add}\textsc{=\textsc{post}} \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m} \textsc{disc.ptcl} cooperation\textsc{.m}\textsc{=3pl} good be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `After the marriage, they (the couple's families) were on good terms.'
\z 
 
\ea \label{RE.26}
\textit{zemānēč ženekēš berdēne ētir čo dumā bīyēnē tēkeɫ} \\ 
\gll zemān-ē=č ženī-ekē=š berdē=ne ētir č=o dumā bīyē=nē tēkeɫ \\ 
 time\textsc{.m}\textsc{-indf}\textsc{=add} woman\textsc{.f}\textsc{-def}\textsc{.dir}\textsc{.f}\textsc{=3sg} take\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=\textsc{post}} \textsc{disc.ptcl} in=\textsc{dem.dist} afterwards be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} mixed \\ 
\glt `When the brides were taken, they (the families) would become one.'
\z 
 
\ea \label{RE.27}
\textit{ētir īse zemān fāřyān hāɫīt bo?} \\ 
\gll ētir īse zemān fāřyā=n hāɫī=t ∅-b-o \\ 
 \textsc{disc.ptcl} now time\textsc{.m} change\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} understood\textsc{=\textsc{2sg}} \textsc{ind-}be\textsc{.prs}\textsc{-3sg} \\ 
\glt `Now, things have changed. You see?'
\z 
 
\ea \label{RE.41}
\textit{ī memliketū ēme ne welē wiɫātāre ētir tāte bīyen čūēw desū šuwāney} \\ 
\gll ī memliket=ū ēme ne welē wiɫāt-ā=re ētir tāte bīye=n čū-ē=w des=ū šuwāne-ī \\ 
 \textsc{dem.prox} country\textsc{.m}\textsc{=ez.gen} \textsc{1pl} no but region\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}} \textsc{disc.ptcl} father\textsc{.m} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} stick\textsc{.m}\textsc{-indf}\textsc{=ez.gen} hand\textsc{.m}\textsc{=ez.gen} shepherd\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `[It is not common] in our region, but elsewhere, the father has become like the stick in the shepherd’s hand.\footnote{Meaning that he has no power over whom his children will marry.}'
\z 
 
\ea \label{RE.55}
\textit{māčo, dey lue deš} \\ 
\gll m-āč-o dey ∅-lu-e ∅-de-e=š \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} \textsc{imp-}give\textsc{.prs}-\textsc{2sg}\textsc{=3sg} \\ 
\glt `He (the nobleman) would say [to the girl’s father], ‘Go [and] give her’'
\z 
 
\ea \label{ŽE.1}
\textit{dey čēgeyč qeyīm ʕāɫifšā kennenū} \\ 
\gll dey čēge=īč qeyīm ʕāɫif=šā kenne=n=ū \\ 
 \textsc{disc.ptcl} here\textsc{=add} old\_time grass\textsc{.dir}\textsc{.m}\textsc{=3pl} mow\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `In the past, they (people) mowed grass.'
\z 
 
\ea \label{ŽE.11}
\textit{ētir pāgene girđkes ħeywāndārīš kerdenū} \\ 
\gll ētir pāgene girđkes ħeywāndārī=š kerde=n=ū \\ 
 \textsc{disc.ptcl} there everyone animal\_husbandry\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `There, everyone tended animals.'
\z 
 
\ea \label{ŽE.14}
\textit{ētir zimsān āmān} \\ 
\gll ētir zimsān āmā=n \\ 
 \textsc{disc.ptcl} winter\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Then the winter came.'
\z 
 
\ea \label{ŽE.22}
\textit{her firē bīyēne ētir e pāse} \\ 
\gll her fire-ē bīyē=ne ētir e=pāse \\ 
 emph a\_lot\textsc{-f} be\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3sg}\textsc{.f} \textsc{disc.ptcl} in=like \\ 
\glt `It (the water) was a lot. Well, [it was] like this.'
\z 
 
\ea \label{ŽE.25}
\textit{yānew ētir ne yāne bīyenū} \\ 
\gll yāne=ū ētir ne yāne=e bīye=n=\\ 
 house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{disc.ptcl} no house\textsc{.m}\textsc{=ez}\textsc{.cmpd} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `As for houses, there was no [proper] house.'
\z 
 
\ea \label{ŽE.30}
\textit{ētir yo xeleš bīyenū} \\ 
\gll ētir yo xele=š bīye=n=\\ 
 \textsc{disc.ptcl} one\textsc{.m} wheat\textsc{.m}\textsc{=3sg} be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}=and \\ 
\glt `A [person] had wheat [as agricultural product].'
\z 
 
\ea \label{ŽE.37}
\textit{hēzmī ētir zimsān āmān hurēsēnē yerē āwīr, āwīre kučēšā nīyēnēre bī kipē} \\ 
\gll hēzmī ētir zimsān āmā=n hur-ēsē=nē yerē āwīr āwīr=e kuč-ē=šā nīyē=nē=re bī-∅ kipē \\ 
 firewood\textsc{.f} \textsc{disc}.\textsc{ptcl} winter\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{pvb-}rise\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} three fire\textsc{.m} fire\textsc{.m}\textsc{=dem} little\textsc{-dir}\textsc{.pl}\textsc{=3pl} put\textsc{.pst}\textsc{.ptcp}\textsc{.f}\textsc{=cop}\textsc{.3pl}\textsc{=\textsc{povb}} be\textsc{.pst}\textsc{-3sg}\textsc{.m} sealed \\ 
\glt `Firewood …  Then when it was winter they would rise and start a small fire [and it would] damp down.'
\z 
 
\ea \label{ŽE.38}
\textit{nīyānšāre ētir, čāyšā kerden be nānekeyū} \\ 
\gll nīyā=n=šā=re ētir čāy=šā kerde=n be nān-ekey=ū \\ 
 put\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl}\textsc{=\textsc{povb}} \textsc{disc}.\textsc{ptcl} tea\textsc{.m}\textsc{=3pl} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} to bread\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.m}=and \\ 
\glt `The started [the fire].  Then, they made tea [to be served] with the food.'
\z 
 
\ea \label{ŽE.78}
\textit{wāčē, dey lue duē yerē pīyē bēnē’} \\ 
\gll wāč-ē dey ∅-lu-e duē yerē pīyē b-ē-nē \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc.}\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} two three man\textsc{.dir}\textsc{.pl} \textsc{sbjv-}come\textsc{.prs}\textsc{-3pl} \\ 
\glt `She (the mother) would say, ‘Go! Let two, three men come.'
\z 
 
\ea \label{ŽE.82}
\textit{wāčē, dey lue yānew wēt nīšere tā duē māngē ter yerē māngē ter} \\ 
\gll wāč-ē dey ∅-lu-e yāne=ū wē=t ∅-nīš-e=re tā duē mānge-ē ter yerē mānge-ē ter \\ 
 say\textsc{.prs-pstc}\textsc{.3sg} \textsc{disc}.\textsc{ptcl} \textsc{imp-}go.\textsc{imp-}\textsc{2sg} house\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{imp-}sit\textsc{.prs}-\textsc{2sg}\textsc{=\textsc{povb}} until two month\textsc{.f}\textsc{-dir}\textsc{.pl} other three month\textsc{.f}\textsc{-dir}\textsc{.pl} other \\ 
\glt `They (the family of the girl) would say [to the boy], ‘Go back home [and] wait\footnote{Lit. sit.} for two, three months.'
\z 
 
\ea \label{ŽE.84}
\textit{ētir ēnēwe eger řezāšā bīyē kināčē dēnē} \\ 
\gll ētir ēnē=we eger řezā=šā bī-ē kināčē de-ēn-ē \\ 
 \textsc{disc}.\textsc{ptcl} come\textsc{.prs}\textsc{.pst}c\textsc{.3pl}\textsc{=compl} if agreement\textsc{.f}\textsc{=3pl} be\textsc{.pst}\textsc{-cond}\textsc{.3pl} girl\textsc{.obl}\textsc{.f} give\textsc{.prs}\textsc{-pstc-3pl} \\ 
\glt `They (the relatives) would come [to the family of the girl] and if they agreed to it, they would give the girl in marriage.'
\z 
 
\ea \label{ŽM.49}
\textit{řoē sī penj timenē šesū yerē sāɫē čēweɫī kerū ētir meyānū sāɫū činē bē} \\ 
\gll řo-ē sī penj timen-ē šes=ū yerē sāɫe-ē čēweɫī ∅-ker-ū ētir me-zān-ū sāɫe-ū činē b-ē \\ 
 day\textsc{.m}\textsc{-indf} thirty five \textsc{pn}\textsc{-indf} sixty\textsc{=ez.gen} three year\textsc{.f}\textsc{-dir}\textsc{.pl} ago \textsc{ind-}do\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc}.\textsc{ptcl} \textsc{neg-}know\textsc{.prs}\textsc{-\textsc{1sg}} year\textsc{.f}\textsc{-\textsc{1sg}} how\_many\textsc{.f} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \\ 
\glt `A daily salary of thirty-five tomans, it was sixty-three years ago. I don’t know which year it was/'
\z 
 
\ea \label{ŽM.51}
\textit{wextē āmēnmēwe ī mentēqe ētir čē musteqerē bīyēnmē} \\ 
\gll wext-ē āmē=nmē=we ī mentēqe ētir čē musteqer-ē bīyē=nmē \\ 
 time\textsc{.m}\textsc{-indf} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}}\textsc{=compl} \textsc{dem.prox} region\textsc{.m} \textsc{disc}.\textsc{ptcl} here settled\textsc{\textsc{-pl}} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} \\ 
\glt `When I returned to this region, I settled down here.'
\z 
 
\ea \label{ŽM.54}
\textit{ētir īsey āmēnmē tā wē yāwēnmē be kemūkuřīū hewrāmānī} \\ 
\gll ētir īse-ī āmē=nmē tā wē yāwē=nmē be kemūkořī=ū hewrāmān-ī \\ 
 \textsc{disc}.\textsc{ptcl} now\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} until there arrive\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.\textsc{1pl}} by poverty\textsc{.m}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `I have made it so far, [despite] poverty and destitution.'
\z 
 
\ea \label{ŽM.55}
\textit{ētir īse weʕzū hewrāmānī tā menteqē terē kučē xāsterā} \\ 
\gll ētir īse weʕze=ū hewrāmān-ī tā menteqe-ē ter-ē kuč-ē xās-ter=ā \\ 
 \textsc{disc}.\textsc{ptcl} now situation\textsc{.f}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} until region\textsc{-dir}\textsc{.pl} other\textsc{-dir}\textsc{.pl} little\textsc{-indf} fine\textsc{-cmpr}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `The situation in Hawraman is a bit better than in other regions.'
\z 
 
\ea \label{ŽM.56}
\textit{čēweɫ nebīyey bē ētir īseyč xū nīyā} \\ 
\gll čēweɫ nebīyey b-ē ētir īse=yč xū nīye=ā \\ 
 previously poverty\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg} \textsc{disc.ptcl} now\textsc{=add} \textsc{disc.ptcl} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `In the past, there was poverty. But even nowadays, there is poverty.'
\z 
 
\ea \label{ŽM.62}
\textit{herkesīč metāwo pēse min ke nelābū ētir āne hīčē derāmeđēš nīyā} \\ 
\gll herkes=īč me-tāw-o pēse min ke ne-lā b-ū ētir āne hīč-ē derāmeđ-ē=š nīye=ā \\ 
 anyone\textsc{=add} \textsc{neg-}can\textsc{.prs}\textsc{-3sg} like \textsc{1sg} rel \textsc{neg-}go\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-\textsc{1sg}} \textsc{disc.ptcl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} nothing\textsc{-indf} income\textsc{-indf}\textsc{=3sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
\ea \label{ŽM.63}
\textit{epāse ētir} \\ 
\gll e=pāse ētir \\ 
 in=like \textsc{disc.ptcl} \\ 
\glt ` Well, (the situation is) like this.'
\z 
 
\end{sloppypar}