\chapter{All 50 examples of voc in these texts are found in the following sentences:}

\ea \label{ZQ.5}
\textit{vātiš, qurbān xuɫkū fiɫāne yāgēnmē degēweš nāmē dā pī pāyīn꞊ū ērāqīre'} \\ 
\gll vāt=iš qurbān xuɫk=ū fiɫān=e yāgē=nmē degā-ēwe=š nāmē dā p=ī pāyīn=ū ʕērāq-ī=re \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} people\textsc{.m}\textsc{=ez.gen} such\_and\_such=ez\textsc{.cmpd} place.f\textsc{=cop}\textsc{.\textsc{1pl}} village.f\textsc{-indf}\textsc{=3sg} name.f give\textsc{.pst} at=\textsc{dem.prox} bottom\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `He (the man) said, ‘Sir, we are from such-and-such a place. He gave him the name of a village in the south of Iraq.'
\z 
 
\ea \label{ZQ.7}
\textit{ʕerziš kerđ, qurbān bā dāstānēt pey gēɫnūwe, dāstāne} \\ 
\gll ʕerz=iš kerđ qurbān ∅-b-ā dāstāne-ē=t pey ∅-gēɫn-ū=we dāstāne \\ 
 petition\textsc{.m}\textsc{=3sg} do\textsc{.pst} sir.\textsc{voc} \textsc{ind-}be.prs\textsc{-3pl} story.f\textsc{-indf}\textsc{=\textsc{2sg}} to \textsc{sbjv-}lead.prs\textsc{-\textsc{1sg}}\textsc{=compl} story.f \\ 
\glt `He said, ‘Sir, Let me tell you a story, the story!’'
\z 
 
\ea \label{ZQ.8}
\textit{fermāwāš, dā dey řoɫe čēš māčī’} \\ 
\gll fermāwā=š dā dey řoɫe čēš m-āč-ī \\ 
 say\textsc{.pst}\textsc{=3sg} give\textsc{.pst} \textsc{disc.ptcl} child.\textsc{voc} what \textsc{ind-}say.prs-\textsc{2sg} \\ 
\glt `The Shah said, ‘Go on son, What is it ?\footnote{Lit. What are you saying?}’'
\z 
 
\ea \label{ZQ.9}
\textit{vātiš, qurbān luāymē pār yā činnē sāɫē čēweɫter luāymē pey sārāɫū ērānī, koy sān} \\ 
\gll vāt=iš qurbān luā-ymē pār yā činnē sāɫe-ē čēweɫter luā-ymē pey sārāɫ=ū ērān-ī ko=y sān \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} go\textsc{.pst} \textsc{1pl} last\_year either some\textsc{.pl} year.f\textsc{-indf} ago go\textsc{.pst} \textsc{1pl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} where=ez.\textsc{attr} \textsc{pn} \\ 
\glt `He said, ‘Sir, we went, say, last year or a few years ago, we went to the Saral region in Iran, to the San mountain.'
\z 
 
\ea \label{ZQ.26}
\textit{yošā berd nīyāmne qiřoɫe dāreke qurbān!} \\ 
\gll yo=šā berd-∅ nīyā=m=ne qiřoɫ=e dār-eke qurbān \\ 
 one\textsc{.m}\textsc{=3pl} take\textsc{.pst}\textsc{-3sg}\textsc{.m} put\textsc{.pst}\textsc{=\textsc{1sg}}\textsc{=\textsc{povb}} hollow\textsc{.m}=ez\textsc{.cmpd} tree\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} sir.\textsc{voc} \\ 
\glt `Sir, I took one [and] put him in the tree hollow. '
\z 
 
\ea \label{ZQ.32}
\textit{āy bāwke řo, āy tāte řo īđīč luāw ewīč luā} \\ 
\gll āy bāwke řo āy tāte řo īđ=īč luā=ū ew=īč luā \\ 
 \textsc{intj} fahter.\textsc{voc} onom \textsc{intj} father\textsc{.m} onom 3sg.prox\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg}=and 3sg\textsc{.dir}\textsc{.m}\textsc{=add} go\textsc{.pst}\textsc{.3sg} \\ 
\glt `Oh dear God\footnote{The expressions bāwka řo and tāta řo are used to mourn someone’s death, or to express distress.}, he (the child whom I took with me) died; he (the child whom I left behind) died too!\footnote{Lit. This one went; that one, too, went.} '
\z 
 
\ea \label{ZQ.33}
\textit{vātiš, qurbān wehārē wāwey luāymēwe pey sārāɫū ērānī} \\ 
\gll vāt=iš qurbān wehār-ē wāwey luā-ymē=we pey sārāɫ=ū ērān-ī \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} spring\textsc{.m}\textsc{-indf} again go\textsc{.pst} \textsc{1pl}\textsc{=compl} to \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘Sir, [The next] spring, again, we went back to the Saral region of Iran.'
\z 
 
\ea \label{ZQ.54}
\textit{vātiš, qurbān! qesem pā xwāy toš e pī lāyiqetī weš kerdenī e īnen ārdenim xizmetū wēt juānē pēse ī girdū juwānā’} \\ 
\gll vāt=iš qurbān qesem p=ā xwā-ī to=š ep=ī lāyiqetī weš kerde=nī e īne=n ārde=n=im xizmet=ū wē=t juān-ē b-ē pēse ī gird=ū juān-ā \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} oath\textsc{.m} at=\textsc{dem.dist} God\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{2sg}\textsc{=3sg} in=\textsc{dem.prox} worthiness\textsc{.m} good do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} emph \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg}\textsc{=cop}\textsc{.3sg}\textsc{.m} bring\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} \textsc{reflx}\textsc{=\textsc{2sg}} youth\textsc{.m}\textsc{-indf} be.prs\textsc{-pstc}\textsc{.3sg} like \textsc{dem.prox} all\textsc{=ez.gen} youth\textsc{.m}\textsc{-obl}\textsc{.pl} \\ 
\glt `He said, ‘Sir, I swear to the God, who has made you with such virtuosity, that he [the boy] is this [man here] that I have brought to your service; [he is] a young man like all the young men. '
\z 
 
\ea \label{HB.20}
\textit{newātim qurbān min bilū yānew kē hārđī bārū} \\ 
\gll ne-wāt=im qurbān min bi-l-w yāne=ū kē hārđī b-ār-w \\ 
 \textsc{neg-}say\textsc{.pst}\textsc{=1sg} sir.\textsc{voc} 1sg \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg} house=and who flour\textsc{\textsc{.f}}\textsc{.obl} \textsc{sbjv-}bring\textsc{.prs}\textsc{-1sg} \\ 
\glt `I didn’t say, ‘Sir, to whose house should I go [and] bring flour!'
\z 
 
\ea \label{HB.76}
\textit{māčā, qurbān wiɫā fiɫāne kes toryānū ānā} \\ 
\gll m-āč-ā qurbān wiɫā fiɫān=e kes toryā=n=ū āne=ā \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} by\_God such\_and\_such\textsc{=ez}\textsc{.cmpd} person\textsc{.m} get offended\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg}=\textsc{ptcl} \\ 
\glt `They said, ‘Sir, indeed, the fellow has got offended; there he is’'
\z 
 
\ea \label{HB.82}
\textit{āđīč māčo, dey qurbān min ānne sāɫēn īnā xizmetū tone!} \\ 
\gll āđ=yč m-āč-o dey qurbān min ānne sāɫe-ē=n īnā-∅ xizmet=ū to=ne \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} sir.\textsc{voc} 1sg some year\textsc{\textsc{.f}}\textsc{-dir}\textsc{.pl}\textsc{=cop}\textsc{.3sg}\textsc{.m} deic\textsc{-3sg}\textsc{.m} service\textsc{.m}\textsc{=ez}\textsc{.gen} \textsc{2sg}\textsc{=\textsc{post}} \\ 
\glt `He said, ‘Oh sir, I have been at your service for so many years!'
\z 
 
\ea \label{PM.11}
\textit{māčo, qurbān ēme tāyfēnmē āmēnmēre,} \\ 
\gll m-āč-o qurbān ēme tāyfe-ē=nmē āmē=nmē=re \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{1pl} tirbe\textsc{.m}\textsc{-indf}\textsc{=cop}\textsc{.1pl} come\textsc{.pst}\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.1pl}\textsc{=\textsc{povb}} \\ 
\glt `[and] said, ‘Sir, we are a family [who] have come here.'
\z 
 
\ea \label{PM.28}
\textit{tāte řow bāwke řo verzīyer mē pey xizmetū šā ʕelādīnī} \\ 
\gll tāte řo=ū bāwke řo verzīyer m-ē pey xizmet=ū šā ʕelādīn-ī \\ 
 father\textsc{.m} onom\textsc{\textsc{=ez.gen}} fahter.\textsc{voc} onom farm\_labourer\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg} for service\textsc{.m}\textsc{\textsc{=ez.gen}} king\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} \\ 
\glt `[Sheikh Aladin’s farm-labourer shouted], ‘Oh, dear God! The farm-labourer came into the presence of Shah Aladin.'
\z 
 
\ea \label{PM.29}
\textit{māčo qurbān īne čēšit kerdū?} \\ 
\gll m-āč-o qurbān īne čēš=it kerd=ū \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} what\textsc{=\textsc{2sg}} do\textsc{.pst}=and \\ 
\glt `They said, ‘Sir, what have you done?'
\z 
 
\ea \label{ZP.47}
\textit{to wekēɫū min be, be mezebū wēšā ber her noʕē mezebšā her noʕ hen māreš biře peyš nārīšo} \\ 
\gll to wekēɫ=ū min ∅-b-e be mezeb=ū wē=šā ber her noʕ-ē mezeb=šā her noʕ hen-∅ māre=š ∅-biř-e pey=š ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}=and \textsc{1sg} \textsc{imp-}be\textsc{.prs}-\textsc{2sg} to religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} out emph kind/way\textsc{-indf} religion\textsc{=3pl} emph kind/way exist\textsc{-3sg}\textsc{.m} marriage\textsc{=3sg} \textsc{imp-}cut\textsc{.prs}-\textsc{2sg} to\textsc{=3sg} \textsc{neg-}bring\textsc{.prs}\textsc{-2sg}\textsc{=3sg}\textsc{=compl} \\ 
\glt `[then] be my advocate [and] marry the girl to him according to their religious beliefs, whatever their religion is. May you not bring her back.'
\z 
 
\ea \label{ZP.89}
\textit{kināčēw pādšāw fiɫān henānew miniš kerdenā be wekēɫ} \\ 
\gll kināčē=ū pādšā=w fiɫān hen-ā=ne=ū min=iš kerde=nā be wekēɫ \\ 
 daughter\textsc{.dir}\textsc{.pl}\textsc{\textsc{=ez.gen}} king\textsc{.m}\textsc{\textsc{=ez.gen}} such\_and\_such place\textsc{.m}\textsc{-obl}\textsc{.pl}\textsc{=\textsc{post}}=and \textsc{1sg}\textsc{=3sg} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} to ad\textsc{voc}ate\textsc{.m} \\ 
\glt `She is the daughter of … the king of such-and-such place has put me in charge.\footnote{Lit. made me advocate.}'
\z 
 
\ea \label{ZP.99}
\textit{pādšāy kerdenā wekēɫū} \\ 
\gll pādšā-ī kerde=nā wekēɫ=ū \\ 
 king\textsc{.m}\textsc{-obl}\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m}=and \\ 
\glt `The king has put me in charge.'
\z 
 
\ea \label{ZP.101}
\textit{āđīč fermāwo, dey xo bābe gīyān, min īne nānmārēū} \\ 
\gll āđ=īč ∅-fermāw-o dey xo bābe gīyān min īne nān=m=ā=rē=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} dis.ptc father.\textsc{voc} dear.\textsc{voc} \textsc{1sg} dem.prox\textsc{.dir}\textsc{.m}\textsc{.3sg} meal\textsc{.m}\textsc{=\textsc{1sg}}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=\textsc{post}}=and \\ 
\glt `He (Pir Shaliyar) said, ‘dear brother! Look, this is my food (i.e., bread made of corn flour)'
\z 
 
\ea \label{ZP.107}
\textit{ewīč māčo, nexeyre lāloekeš min tāze pādšā kerdenā wekēɫū} \\ 
\gll ew=īč m-āč-o nexeyre lālo-eke=š min tāze pādšā kerde=nā wekēɫ=ū \\ 
 3sg\textsc{.dir}\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no maternal\_uncle\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=3sg} \textsc{1sg} anyway king\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m}=and \\ 
\glt ` The uncle said, ‘No! I—the king has put me in charge.'
\z 
 
\ea \label{ŽP.57}
\textit{pīyāke mēwe ʕerzū heme yosoy kero māčo, weɫāhī qurbān min tūšū jeryānē e čīney āmānē} \\ 
\gll pīyā-eke m-ē=we ʕerz=ū heme yoso-ī ∅-ker-o m-āč-o weɫāhī qurbān min tūš=ū jeryān-ē e=čīne-ī āmā-ānē \\ 
 man\textsc{.m}\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}come\textsc{.prs}\textsc{.3sg}\textsc{=compl} petition\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{ind-}do\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} indeed sir.\textsc{voc} \textsc{1sg} encounter\textsc{\textsc{=ez.gen}} story\textsc{.m}\textsc{-indf} in=like\textsc{-obl}\textsc{.m} come\textsc{.pst}\textsc{-\textsc{1sg}} \\ 
\glt `The man returned [and said] to Hama Yoso, ‘Indeed, Sir! I have witnessed an event like this.'
\z 
 
\ea \label{ŽP.126}
\textit{māčā, qurbān to desūrēt dā be sey misefāy bilowe šerēʕetū terēqet ders wāčo} \\ 
\gll m-āč-ā qurbān to desūr-ē=t dā be sey misefā-ī bi-l-o=we šerēʕet=ū terēqet ders ∅-wāč-o \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3pl} sir.\textsc{voc} \textsc{2sg} task/order\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{hort} by \textsc{pn} \textsc{pn}\textsc{-obl}\textsc{.m} \textsc{sbjv-}go\textsc{.prs}\textsc{-3sg}\textsc{=compl} sharia\_law=and denomination lesson\textsc{.m} \textsc{sbjv-}say\textsc{.prs}.sbjv\textsc{-3sg} \\ 
\glt `They (the gossips) said, ‘Sir, you ordered Say Mustafa to go back [to Hawraman and] teach Sharia law and [the ways of] orders (of dervishes).'
\z 
 
\ea \label{ŽP.165}
\textit{to wekēɫū min be be mezebū wēšā ī kināčešā pey nīkāh kerew nārīšo} \\ 
\gll to wekēɫ=ū min be be mezeb=ū wē=šā ī kināče=šā pey nīkāh ∅-ker-e=ū ne-ār-ī=š=o \\ 
 \textsc{2sg} ad\textsc{voc}ate\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} by by religion\textsc{\textsc{=ez.gen}} \textsc{reflx}\textsc{=3pl} dem.prox daughter\textsc{.f}\textsc{=3pl} to marriage\textsc{.m} \textsc{imp-}do\textsc{.prs}-\textsc{2sg}\textsc{\textsc{=ez.gen}} \textsc{neg-}bring\textsc{.prs}\textsc{-obl}\textsc{.m}\textsc{=3sg}\textsc{=compl} \\ 
\glt `Be my advocate [and] marry the girl to him [according to the customs of] their religion.'
\z 
 
\ea \label{ŽP.206}
\textit{ā wextī lālo kināčē māčo, min, tāze pādšāy kerdenā wekēɫ} \\ 
\gll ā wext-ī lālo kināčē m-āč-o min tāze pādšā-ī kerde=nā wekēɫ \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} maternal\_uncle\textsc{.m} girl\textsc{.dir}\textsc{.f} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{1sg} anyway king\textsc{.m}\textsc{-obl}\textsc{.m} do\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{1sg}} ad\textsc{voc}ate\textsc{.m} \\ 
\glt `Then, the girl’s uncle said, ‘As for me, well, the king has put me in charge.\footnote{Lit. the king has made me advocate.}'
\z 
 
\ea \label{ŽP.220}
\textit{ā wextī māčā, dey qurbān} \\ 
\gll ā wext-ī m-āč-ā dey qurbān \\ 
 \textsc{dem.dist} time\textsc{.m}\textsc{-obl}\textsc{.m} \textsc{ind-}say\textsc{.prs}\textsc{-3pl} \textsc{disc.ptcl} sir.\textsc{voc} \\ 
\glt `Then, they (people) said, Oh, Sir!’'
\z 
 
\ea \label{ŽP.227}
\textit{xeɫkīč māčo, dey kāke gīyān qey mekero to īse hīčit nīyā ēme, yomā āne hiɫošet miđo} \\ 
\gll xeɫk=īč m-āč-o dey kāke gīyān qey me-ker-o to īse hīč=it nīye=ā ēme yo=mā āne hiɫoše=t mi-đe-o \\ 
 people\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{disc.ptcl} brother.vob dear.\textsc{voc} limit \textsc{neg-}do\textsc{.prs}\textsc{-3sg} \textsc{2sg} now nothing\textsc{=\textsc{2sg}} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \textsc{1pl} one\textsc{.m}\textsc{=1pl} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} cracked\_wheat\textsc{.dir}\textsc{.f}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} \\ 
\glt `The people said, ‘Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{ŽP.230}
\textit{yoīč mā yoīč māčo, kāke gīyān! minīč šimšū yerū heywānēt mido pene} \\ 
\gll yo=īč=mā yo=īč m-āč-o kāke gīyān min=īč šimšī=ū yeher=ū heywān-ē=t mi-de-o pene \\ 
 one\textsc{.m}\textsc{=add}\textsc{=1pl} one\textsc{.m}\textsc{=add} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} brother.vob dear.\textsc{voc} \textsc{1sg}\textsc{=add} spleen\textsc{.f}\textsc{\textsc{=ez.gen}} liver\textsc{.m}\textsc{\textsc{=ez.gen}} animal\textsc{.m}\textsc{-indf}\textsc{=\textsc{2sg}} \textsc{ind-}give\textsc{.prs}\textsc{-3sg} to \\ 
\glt `One [said], One person said, ‘Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{BP.172}
\textit{hejījīyeke, fermāwo hejījīyeke, be qurbānit bim sūre gīyān yekē tirī lē de leqē teriš de wene’} \\ 
\gll hejījī-eke ∅-fermāw-o hejījī-eke be qurbān=it ∅-b-im sūr=e gīyān yek-ē tir=ī lē ∅-de-e leq-ē ter=iš ∅-de-e wene \\ 
 from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} \textsc{ind-}say.prs\textsc{-3sg} from\_Hajij\textsc{-def}\textsc{.dir}\textsc{.m} to sacrifice\textsc{=\textsc{2sg}} \textsc{sbjv-}be.prs\textsc{-\textsc{1sg}} red=ez\textsc{.cmpd} dear.\textsc{voc} one\textsc{-indf} other\textsc{=3sg} at \textsc{imp-}give.prs-\textsc{2sg} kick\textsc{-indf} another\textsc{=3sg} \textsc{imp-}give.prs-\textsc{2sg} at \\ 
\glt `The [prisoner from] Hajij said [to Baba Khwada], ‘May I be your sacrifice, you red-skin, kick it once more! Kick it once more!”'
\z 
 
\ea \label{ŠJ.14}
\textit{māčā qurbān to čī bīyenī pīyā, čī bīyenī jāfir sān?} \\ 
\gll m-āč-ā qurbān to čī bīye=nī pīyā čī bīye=nī jāfir sān \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{2sg} why be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} man\textsc{.m} why be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \textsc{pn} \textsc{pn} \\ 
\glt `[and] said, ‘Sir, how have you become such a man? How have you turned into the Jafir Sān (that you are)?'
\z 
 
\ea \label{ŠJ.17}
\textit{vātiš qurbān seʕbē nemo hīn tifengēwew beqēwe} \\ 
\gll vāt=iš qurbān seʕbe-ē ∅-nem-o hīn tifeng-ēwe=ū beq-ēwe \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} morning.f\textsc{-obl}.f \textsc{ind-}lift.prs\textsc{-3sg} \textsc{fill} gun\textsc{-indf}\textsc{=ez.gen} partridge\textsc{.m}\textsc{-indf} \\ 
\glt `They said, ‘Sir, he lifts a gun and a partridge early in the morning.'
\z 
 
\ea \label{ŠJ.30}
\textit{māčā qurbān īne čiko bīyenī} \\ 
\gll m-āč-ā qurbān īne či=ko bīye=nī \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.m}\textsc{.3sg} in=where be\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.\textsc{2sg}} \\ 
\glt `They said, ‘Sir, where have you been?'
\z 
 
\ea \label{ŠJ.44}
\textit{māčo qurbān! be xwā xuřyeymēš venī} \\ 
\gll m-āč-o qurbān be xwā xuřye-ymē=š venī \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} to God\textsc{.m} shout\textsc{.pst} \textsc{1pl}\textsc{=3sg} at \\ 
\glt `They said, ‘Sir, indeed, [as] he shouted at us.'
\z 
 
\ea \label{ŠJ.69}
\textit{māčā qurbān wāqēʕe e čīne qomyān} \\ 
\gll m-āč-ā qurbān wāqēʕe e=čīne qomyā=n \\ 
 \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} event\textsc{.m} in=like happen\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `[and] said, ‘Sir, a story like this has happened!'
\z 
 
\ea \label{ŠJ.71}
\textit{māčo qurbān hīn jāfir sānū luhonī īnā tewēɫew wiɫāxekāne} \\ 
\gll m-āč-o qurbān hīn jāfir sān=ū luhon-ī īnā-∅ tewēɫe=ū wiɫāx-ekā=ne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} \textsc{fill} \textsc{pn} \textsc{pn}\textsc{=ez.gen} \textsc{pn}\textsc{-obl}\textsc{.m} deic\textsc{-3sg}\textsc{.m} stable\textsc{.m}\textsc{=ez.gen} horse/donkey\textsc{.m}\textsc{-def}\textsc{.obl}\textsc{.pl}\textsc{=\textsc{post}} \\ 
\glt `He (the horse groom) said, ‘Sir, thingummy, Jafir Sān from Luhon is in the horses’ stable!’'
\z 
 
\ea \label{ŠJ.73}
\textit{māčo weɫā qurbān} \\ 
\gll m-āč-o weɫā qurbān \\ 
 \textsc{ind-}say.prs\textsc{-3sg} indeed sir.\textsc{voc} \\ 
\glt `He (the horse groom) said, ‘Indeed, sir [it’s true].’'
\z 
 
\ea \label{ŠJ.77}
\textit{milā māčā qurbān šē ʕusmānī fermāwā kāriš be min nīyā} \\ 
\gll mi-l-ā m-āč-ā qurbān šē ʕusmān-ī fermāwā kār=iš be min nīye=ā \\ 
 \textsc{ind-}go.prs\textsc{-3pl} \textsc{ind-}say.prs\textsc{-3pl} sir.\textsc{voc} sheikh\textsc{.m} \textsc{pn}\textsc{-obl}\textsc{.m} say\textsc{.pst} task\textsc{.m}\textsc{=3sg} to \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=cop}\textsc{.3sg}\textsc{.m} \\ 
\glt `They went [and] said [to Jafir Sān], ‘Sir, the sheikh said, “He has no business with me.'
\z 
 
\ea \label{ŠJ.86}
\textit{fermāwo qurbān hermāne be desū min nīyenew} \\ 
\gll ∅-fermāw-o qurbān hermāne be des=ū min nīye=ne=ū \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} work\textsc{.f} to hand\textsc{.m}\textsc{=ez.gen} \textsc{1sg} \textsc{\textsc{neg.}exist}\textsc{=\textsc{post}}=and \\ 
\glt `He (the sheikh) said, ‘Sir, the task [of ruling] is not in my hands;'
\z 
 
\ea \label{ŠJ.96}
\textit{fermāwo qurbān} \\ 
\gll ∅-fermāw-o qurbān \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} \\ 
\glt `[and] said, ‘Sir!’ '
\z 
 
\ea \label{ŠJ.98}
\textit{māčo qurbān šerte bo tā min henā sāɫē duē deħfē yerē deħfē bew xizmetū merāʕetū er desīčim bilo ī xāneqāw tobe kerūne} \\ 
\gll m-āč-o qurbān šerte ∅-b-o tā min hen-ā sāɫe-ē duē deħfe-ē yerē deħfe-ē b-e-ū xizmet=ū merāʕet=ū er des=īč=im bi-l-o ī xāneqā=ū tobe ∅-ker-ūne \\ 
 \textsc{ind-}say.prs\textsc{-3sg} sir.\textsc{voc} condition \textsc{sbjv-}be.prs\textsc{-3sg} until \textsc{1sg} exist\textsc{-\textsc{1sg}} year.f\textsc{-indf} two time.f\textsc{-dir}\textsc{.pl} three time.f\textsc{-dir}\textsc{.pl} \textsc{sbjv-}come.prs\textsc{-\textsc{1sg}} service\textsc{.m}\textsc{=ez.gen} care\textsc{.m}\textsc{=ez.gen} if hand\textsc{.m}\textsc{=add}\textsc{=\textsc{1sg}} \textsc{sbjv-}go.prs\textsc{-3sg} \textsc{dem.prox} monastery\textsc{.m}=and repentance \textsc{sbjv-}do.prs\textsc{-\textsc{1sg}} \\ 
\glt `He said, ‘Sir, I swear that as long as I live\footnote{Lit. I exist}, I [will] come here to your service in this monastery two, three times per year. If I can,\footnote{lit. if my hands go} I shall also repent [and ask for atonement].'
\z 
 
\ea \label{ŠJ.99}
\textit{kāke gīyān īnī ečānē bīyēnē} \\ 
\gll kāke gīyān īnī e=čāne-ē bīyē=nē \\ 
 brother.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox}\textsc{.dir}\textsc{.pl} in=like\_that\textsc{-dir}\textsc{.pl} be\textsc{.ptcp}\textsc{.pl}\textsc{=cop}\textsc{.3pl} \\ 
\glt `Dear brother, they were like this.'
\z 
 
\ea \label{KŠ.28}
\textit{wātiš, qurbān herkey gele berī epāne min ī bizeme čekyo} \\ 
\gll wāt=iš qurbān herkey gele ∅-ber-ī epāne min ī bize=m=e ∅-čekye-o \\ 
 say\textsc{.pst}\textsc{=3sg} sir.\textsc{voc} each\_time herd\textsc{.m} \textsc{sbjv-}take\textsc{.prs}-\textsc{2sg} there \textsc{1sg} \textsc{dem.prox} goat\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} \textsc{ind-}drain\textsc{.prs}.pass\textsc{-3sg} \\ 
\glt `[and] said, ‘Sir, whenever you take the herd there, my goat[‘s milk] is drained.'
\z 
 
\ea \label{KŠ.56}
\textit{yewāšē čāge, čogene šewē minwīso, māčo, ey kesūkārū min, ey henū nīyenū min, ey īnīšā ke mintā gereknā, tāyfew min heɫbetene ī kuře biřiš dāwe, be her jor ī kuřī beynne berdē.’} \\ 
\gll yewāšē čāge č=oge=ne šewe-ē mi-nwīs-o m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey īnīšā ke min=tā gerek=nā tāyfe=ū min heɫbete=ne ī kuř=e biř=iš dā=we be her jor ī kuř-ī beyn=ne ∅-ber-dē \\ 
 then there in=there\textsc{=\textsc{post}} night\textsc{.f}\textsc{-indf} \textsc{ind-}write\textsc{.prs}\textsc{-3sg} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.prox}\textsc{.obl}\textsc{.3pl} rel \textsc{1sg}=\textsc{2pl} necessary\textsc{=cop}\textsc{.\textsc{1sg}} tirbe\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} by any means \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} between\textsc{=\textsc{post}} \textsc{ind-}take\textsc{.prs}-\textsc{2pl} \\ 
\glt `Then, one night he (the king) wrote [a letter], saying, ‘O my relatives, O my close family members , O those who want me, [are] from my tribe, [when you read this letter] it is obvious that this boy has arrived. By any means possible, kill him.'
\z 
 
\ea \label{KŠ.72}
\textit{niwīsebēš, ey kesūkārū min, ey henū nīyenū min, heɫbetene ī kuřī biřiš dāwe, šime be wesīlēwe kuždēš, beynne berdēš!} \\ 
\gll niwīse b-ē=š ey kesūkār=ū min ey hen=ū nīyen=ū min heɫbete=ne ī kuř-ī biř=iš dā=we šime be wesīle-ēwe ∅-kuž-dē=š beyn=ne ∅-ber-dē=š \\ 
 write\textsc{.pst}\textsc{.ptcp}\textsc{.m} be\textsc{.prs}\textsc{-pstc}\textsc{.3sg}\textsc{=3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{-obl}\textsc{.m} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} \textsc{2pl} by tool\textsc{.m}\textsc{-indf} \textsc{imp-}kill\textsc{.prs}-\textsc{2pl}\textsc{=3sg} between\textsc{=\textsc{post}} \textsc{imp-}take\textsc{.prs}-\textsc{2pl}\textsc{=3sg} \\ 
\glt `He had written, ‘O my relatives, O my close family members, [when you read this letter] it is evident that this boy has arrived. Kill him with a thing, destroy him!'
\z 
 
\ea \label{KŠ.80}
\textit{māčo, ey kesūkārū min, ey henū nīyenū min, ey āne īnāy hīnne heɫbetene ī kuře biřiš dāwe fīlfor ī kināčēme peyš māre kerdē} \\ 
\gll m-āč-o ey kesūkār=ū min ey hen=ū nīyen=ū min ey āne īnā-ī hīn=ne heɫbete=ne ī kuř=e biř=iš dā=we fīlfor ī kināčē=m=e pey=š māre ∅-ker-dē \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{voc} relative\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} being\textsc{.m}=and non\_being\textsc{.m}\textsc{\textsc{=ez.gen}} \textsc{1sg} \textsc{voc} \textsc{dem.dist}\textsc{.dir}\textsc{.m}\textsc{.3sg} \textsc{deic}-\textsc{2sg} \textsc{fill}\textsc{=\textsc{post}} of\_course\textsc{=\textsc{post}} \textsc{dem.prox} son/boy\textsc{.m}\textsc{=dem} piece\textsc{=3sg} give\textsc{.pst}\textsc{=compl} immediately \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{1sg}}\textsc{=dem} to\textsc{=3sg} marraige \textsc{imp-}do\textsc{.prs}-\textsc{2pl} \\ 
\glt `She wrote,\footnote{Lit. said.} ‘‘O my relatives, O my close family members, O those who are in the thingummy, [when you read this letter] it is obvious that this boy has arrived. Immediately marry my daughter to him.’'
\z 
 
\ea \label{ŽH.12}
\textit{milā řāhur māčo, heyās māčoš pene, kāke gīyān pilekānēwe binye bā bilmē!’} \\ 
\gll mi-l-ā řā=hur m-āč-o heyās m-āč-o=š pene kāke gīyān pilekānī-ēwe bi-nye-e bā bi-l-mē \\ 
 \textsc{ind-}go\textsc{.prs}\textsc{-3pl} road\textsc{\textsc{.f}}\textsc{=\textsc{post}} \textsc{ind-}say\textsc{.prs}\textsc{-3sg} \textsc{pn} \textsc{ind-}say\textsc{.prs}\textsc{-3sg}\textsc{=3sg} to brother.vob dear.\textsc{voc} stair\textsc{\textsc{.f}}\textsc{-indf} \textsc{sbjv-}put\textsc{.prs}-\textsc{2sg} \textsc{hort} \textsc{sbjv-}go\textsc{.prs}\textsc{-1pl} \\ 
\glt `They set off. He, said, [i.e.] Hayas said to him, ‘Dear brother, set up a ladder for us to move !’'\footnote{A cryptic message meaning ‘tell a story’.}
\z 
 
\ea \label{ŽH.23}
\textit{māčo, ne wiɫā řoɫe, āmānē pāynū serekoē wāčmēš pene be zuwānū hewrāmīyānew wēmā derey} \\ 
\gll m-āč-o ne wiɫā řoɫe āmā-ānē pāyn=ū sereko-ē ∅-wāč-mē=š pene be zuwān=ū hewrāmī-yāne=ū wē=mā dere-ī \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} no by\_God child.\textsc{voc} come\textsc{.pst}\textsc{-1sg} bottom\textsc{=ez}\textsc{.gen} uphill\textsc{.m}\textsc{-indf} \textsc{ind-}say\textsc{.prs}.sbjv\textsc{-1pl}\textsc{=3sg} to by language\textsc{.m}\textsc{=ez}\textsc{.gen} Hawrami-vernacular\_of\textsc{=ez}\textsc{.gen} \textsc{reflx}\textsc{=1pl} valley\textsc{.m}\textsc{-obl}\textsc{.m} \\ 
\glt `He (the man) said, ‘No, child, I went to the bottom of the slope —we call it ‘daray’ in our Hawrami language—'
\z 
 
\ea \label{ŽH.37}
\textit{řo yerē āmāwe wātiš be kināčekē, řoɫe řefēqekem her mizgīne īnā mizgīne} \\ 
\gll řo yerē āmā=we wāt=iš be kināčē-ekē řoɫe řefēq-eke=m her mizgī=ne īnā-∅ mizgī=ne \\ 
 day\textsc{.m} three come\textsc{.pst}\textsc{.3sg}\textsc{=compl} say\textsc{.pst}\textsc{=3sg} to daughter\textsc{\textsc{.f}}\textsc{-def}\textsc{.obl}\textsc{\textsc{.f}} child.\textsc{voc} friend\textsc{-def}\textsc{.dir}\textsc{.m}\textsc{=1sg} just mosque\textsc{.m}\textsc{=\textsc{post}} deic\textsc{-3sg}\textsc{.m} mosque\textsc{.m}\textsc{=\textsc{post}} \\ 
\glt `On the third day, he (the father) returned [home and] said to the daughter, ‘Child, my friend is still in the mosque; he is at the mosque.’'
\z 
 
\ea \label{ŽH.113}
\textit{māčo, qurbān čī bilūwe! min činnē hāy wextē bāxēm bāxēm ārd ber} \\ 
\gll m-āč-o qurbān čī bi-l-ū=we min činnē hāy wext-ē bāx-ē=m bāx-ē=m ārd-∅ ber \\ 
 \textsc{ind-}say\textsc{.prs}\textsc{-3sg} sir.\textsc{voc} why \textsc{sbjv-}go\textsc{.prs}\textsc{-1sg}\textsc{=compl} 1sg some\textsc{.pl} \textsc{intj} time\textsc{.m}\textsc{-indf} garden\textsc{.m}\textsc{-indf}\textsc{=1sg} garden\textsc{.m}\textsc{-indf}\textsc{=1sg} bring\textsc{.pst}\textsc{-3sg}\textsc{.m} product \\ 
\glt `He (Hayas) said, ‘Sir, why should I go back (home)? After a long time my garden has produced [fruit]'
\z 
 
\ea \label{RE.4}
\textit{bābe gīyān ī kināčēte yā tātemā lūān yā māmomā yā kesūkārmā} \\ 
\gll bābe gīyān ī kināčē=t=e yā tāte=mā lūā=n yā māmo=mā yā kesūkār=mā \\ 
 father.\textsc{voc} dear.\textsc{voc} \textsc{dem.prox} daughter\textsc{.obl}\textsc{.f}\textsc{=\textsc{2sg}}\textsc{=dem} either father\textsc{.m}\textsc{=\textsc{1pl}} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} either paternal\_uncle\textsc{=\textsc{1pl}} either relative\textsc{.m}\textsc{=\textsc{1pl}} \\ 
\glt `They would say, ‘Dear brother,\footnote{Lit. father.} this daughter of yours \—either our father would go [to the girl’s father], or our uncle, or a relative of ours—'
\z 
 
\ea \label{RE.5}
\textit{āmān lūān wāteniš, birā gīyān! min xū wēt zānī hāwmāɫitnā} \\ 
\gll āmā=n lūā=n wāte=n=iš birā gīyān min xū wē=t ∅-zān-ī hāwmāɫ=it=nā \\ 
 come\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} go\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m} say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3sg} brother\textsc{.m} dear.\textsc{voc} \textsc{1sg} \textsc{disc.ptcl} \textsc{reflx}\textsc{=\textsc{2sg}} \textsc{ind-}know\textsc{.prs}-\textsc{2sg} neighbour\textsc{=\textsc{2sg}}\textsc{=cop}\textsc{.\textsc{1sg}} \\ 
\glt `They would go [and] say, ‘Dear brother, you know [well] that I am your neighbour.'
\z 
 
\ea \label{ŽE.49}
\textit{wātenšā, xwā gīyān hemdū senā pey to’} \\ 
\gll wāte=n=šā xwā gīyān hemd=ū senā pey to \\ 
 say\textsc{.pst}\textsc{.ptcp}\textsc{.m}\textsc{=cop}\textsc{.3sg}\textsc{.m}\textsc{=3pl} God\textsc{.m} dear.\textsc{voc} praise\textsc{\textsc{=ez.gen}} admiration for \textsc{2sg} \\ 
\glt `[They] the would say, ‘Dear God, may our praises be for you!’'
\z 
 
